### 2024-09-16 paper/paper.tex
some prose

% -*- fill-column: 80; -*-
\documentclass[sigconf,review,anonymous,screen,acmsmall]{acmart}
\usepackage{mathpartir}
\usepackage{tikz-cd}
\usepackage{lipsum}
\usepackage{enumitem}
\usepackage{wrapfig}
\usepackage{fancyvrb}
\usepackage{xspace}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{stmaryrd}
\usepackage{scalerel}
\usepackage{tikz}
\usetikzlibrary{automata, positioning, arrows, fit}

\usepackage{pdfpages}

\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
\newcommand{\cat}[1]{\mathbf{#1}}
\newcommand{\lto}{\multimap}
\newcommand{\tol}{\mathrel{\rotatebox[origin=c]{180}{$\lto$}}}
\newcommand{\String}{\Sigma^{*}}
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\SemAct}{\mathbf{SemAct}}
\newcommand{\Gr}{\mathbf{Gr}}
\newcommand{\Grammar}{\mathbf{Gr}}
\newcommand{\Type}{\mathbf{Type}}
\newcommand{\Prop}{\mathbf{Prop}}
\newcommand{\Bool}{\mathbf{Bool}}
\newcommand{\true}{\mathbf{true}}
\newcommand{\false}{\mathbf{false}}
\newcommand{\nat}{\mathbb{N}}
\newcommand{\theoryname}{Grammar Type Theory\xspace}
\newcommand{\theoryabbv}{GramTT\xspace}

\newcommand{\gluedNL}{{\mathcal G}_S}
\newcommand{\gluedNLUniv}{{\mathcal G}_{S,i}}
\newcommand{\gluedL}{{\mathcal G}_L}

\newcommand{\amp}{\mathrel{\&}}
\newcommand{\pair}{\amp}
\DeclareMathOperator*{\bigamp}{\scalerel*{\&}{\bigoplus}}

\newcommand{\simulsubst}[2]{#1\{#2\}}
\newcommand{\subst}[3]{\simulsubst {#1} {#2/#3}}
\newcommand{\letin}[3]{\mathsf{let}\, #1 = #2 \, \mathsf{in}\, #3}
\newcommand{\lamb}[2]{\lambda #1.\, #2}
\newcommand{\lamblto}[2]{\lambda^{{\lto}} #1.\, #2}
\newcommand{\lambtol}[2]{\lambda^{{\tol}} #1.\, #2}
\newcommand{\dlamb}[2]{\overline{\lambda} #1.\, #2}
\newcommand{\app}[2]{#1 \, #2}
\newcommand{\applto}[2]{#1 \mathop{{}^{\lto}} #2}
\newcommand{\apptol}[2]{#1 \mathop{{}^{\tol}} #2}
\newcommand{\PiTy}[3]{\textstyle\prod (#1 : #2). #3}
\newcommand{\SigTy}[3]{\textstyle\sum (#1 : #2). #3}
\newcommand{\LinPiTy}[3]{\textstyle\bigamp (#1 : #2). #3}
\newcommand{\LinSigTy}[3]{\textstyle\bigoplus (#1 : #2). #3}
\newcommand{\GrTy}{\mathsf{Gr}}

\newcommand{\ctxwff}[1]{#1 \,\, \mathsf{ok}}
\newcommand{\ctxwffjdg}[2]{#1 \vdash #2 \,\, \mathsf{type}}
\newcommand{\linctxwff}[2]{#1 \vdash #2 \,\, \mathsf{ok}}
\newcommand{\linctxwffjdg}[2]{#1 \vdash #2 \,\, \mathsf{linear}}

\newif\ifdraft
\drafttrue
\newcommand{\steven}[1]{\ifdraft{\color{orange}[{\bf Steven says}: #1]}\fi}
\renewcommand{\max}[1]{\ifdraft{\color{blue}[{\bf Max says}: #1]}\fi}
\newcommand{\pedro}[1]{\ifdraft{\color{red}[{\bf Pedro says}: #1]}\fi}
\newcommand{\pipe}{\,|\,}

\begin{document}

\pagestyle{plain}

\pagebreak

\title{A Dependent Lambek Calculus for Intrinsic Verification of Parsers and Formal Grammar Theory}

\author{Steven Schaefer}
\affiliation{\department{Electrical Engineering and Computer Science}
  \institution{University of Michigan}
  \country{USA}
}
\email{stschaef@umich.edu}

\author{Max S. New}
\affiliation{
  \department{Electrical Engineering and Computer Science}
  \institution{University of Michigan}
  \country{USA}
}
\email{maxsnew@umich.edu}

\author{Pedro H. Azevedo de Amorim}
\affiliation{
  \department{Department of Computer Science}
  \institution{University of Oxford}
  \country{UK}
}
\email{pedro.azevedo.de.amorim@cs.ox.ac.uk}

\makeatletter
\let\@authorsaddresses\@empty
\makeatother

\begin{abstract}
  We propose \theoryname~(\theoryabbv) a new type-theoretic calculus
  for the specification of formal grammars and intrinsic verification
  of parsers. \theoryabbv is a dependent non-commutative
  linear/non-linear type theory where the linear types can be viewed
  as formal grammars, and linear terms can be viewed as parse
  transformers, of which parsers are a special case. We formalize this
  interpretation by giving a denotational semantics of the theory in a
  novel categorical semantics of formal grammars in a category of
  families indexed by strings.

  To demonstrate the usefulness of the calculus for formal grammar
  theory, we show that existing grammar formalisms --- e.g.\ regular
  expressions, context free expressions, and their associated automata \max{which ones do
    we actually show?} \steven{regular expressions, $\mu$-regular expressions
    (or context free expressions), and different automata classes (finite,
    pushdown, Turing)} --- can be expressed as linear types in
  \theoryabbv. We give simple characterizations of the regular and
  context-free grammars as expressible by restricting which
  connectives of \theoryabbv are allowed, a form of type theoretic
  characterization of these grammar classes. Further, we demonstrate
  some central proofs of formal grammar theory can be intrinsically
  verified: \emph{strong} equivalence between regular expressions and
  NFAs is expressed by a syntactic \emph{isomorphism} definable in
  \theoryabbv.

  Finally, since \theoryabbv is a syntactic calculus, in addition to
  the ``standard'' semantics of linear types as grammars, we
  demonstrate two non-standard semantic interpretations. One is a
  semantics where linear types are instead interpreted as
  \emph{semantic actions}, rather than grammars, which are more common
  in application than formal grammars proper. Lastly we use a
  non-standard semantics to prove a \emph{canonicity} metatheorem that
  shows that the interpretation of types as grammars is canonical: the
  type-theoretic notion of a parse agrees with the expected semantic
  notion.
\end{abstract}

\maketitle

\section{Introduction}
\label{sec:intro}

Parsing is central to computing. Verification of parsers is a crucial
task in many verified applications....There is a great deal of
literature on specific parsing formalisms and proving them correct,
but of course each formalism is restrictive and must be given a new
formal treatement of its correctness.

% TODO: more about why a *framework* is useful

To address these issues with verified parsing, we have designed a
\emph{domain-specific type theory} which supports the definition of grammar
formalisms and \emph{intrinsic verification} of parser implementations. The type
theory is based on a \emph{dependent non-commutative linear type theory}, which
we call the \emph{dependent Lambek calculus}. In this type theory, the linear
types are interpreted as \emph{formal grammars}, with the values being the
\emph{parses} of the grammar. The terms are a kind of \emph{intrinsically
verified} parse transformer: that is a term $x: A \vdash M : B$ implements a
function that takes any $A$-parse of a string $w$ and returns a $B$-parse of the
same string. As a special case, if $A$ is the trivial grammar of all strings,
then $M$ is an \emph{intrinsically verified total parser} a function that
returns a $B$ parse for every string. The benefit of this intrinsic approach to
constructing parsers is that the terms in our Lambek calculus look like ordinary
functional programs that obey an ordered linear typing discipline, yet this
discipline ensures by construction that the grammatical structure is respected.

To demonstrate that this is a \emph{practical} language for implementing grammar
formalisms and intrinsically verified parsers, we demonstrate several examples: ...

We have additionally implemented a proof of concept \emph{domain-specific proof
assistant} via a shallow embedding of our type theory in Agda.

intrinsically verified parsers for regular expressions and finite
automata, as well as some example LL(1) grammars.

Parsing data from a serialized format is one of the most common tasks
in all of computing. Accordingly, this is a quite well-studied problem
spanning theoretical computer science and linguistics with such
milestones as Chomsky's hierarchy of grammars
\cite{chomThreeModels1956}, practical algorithms for parsing of
regular and context-free grammars, and variants
\cite{KNUTH1965607,Earley1970}, as well as implementations of
practical tools for the generation of efficient parsers
\cite{Johnsonyacc}.

The theory of formal languages and parsing is one of the oldest and most
thoroughly developed area of theoretical computer science.
The central object of study is a \emph{formal language} $L$ over an alphabet
$\Sigma$, which is classically defined as a
\emph{subset} of strings $L \subseteq \String$. The expressivity of formal
languages is sufficient for defining a \emph{recognizer} of a language $L$:
a function $r : \String \to \Prop$ such that $r$ maps each string $w$ to the
proposition that $w$ is in the language $w \in L$. This technique of language
recognition can concisely specify lexical analysis, and even lead to verified
lexer implementations \cite{egolfVerbatim,Ouedraogo_2023}.

While there are ample thoery and tooling to handle the first stage of a language frontend, the final
frontier for parsing research is the development of
\emph{formally verified} parser implementations. Due to the
ubiquitous nature of parsing, these would useful components of many
formally verified software systems: compilers and servers most
especially. However, language recognition is insufficient for specifying a
parser, and for this task much of the structured foundations from the theory of
formal languages is rendered moot. Rather than returning a propositon-valued result, a parser produces a
structured \emph{parse tree} adherent to a \emph{formal grammar}.
The research on such grammars is often performed atop an ad-hoc choice of grammar
formalism, and results are proven with respect to that formalism. Critically,
the field is lacking a unifying theory across these different formalisms, even
though many of the same
compositional reason principles appear across distinct grammar formalisms.

In this work, we aim to elucidate these similarities and provide a
mathematical theory of formal grammars for constructing verified
parsers. Our approach can be summarized with the Curry-Howard-style
slogan: ``grammars as types''; and the addendum: ``parsers as
terms''. That is, we develop a type theory with a novel denotational
semantics in a category with abstract formal grammars as objects and
\emph{parse transformers} as morphisms. We say these are
\emph{abstract} formal grammars because they are not tied to any
particular syntactic formalism such as Chomskyan grammars, regular
expressions, Lambek calculus etc. The category we choose is a rich in
structure: it is a topos as well as being monoidal closed, and these
universal properties serve as the inspiration for our type theory.

The type theory itself is a form of dependent linear-non-linear type
theory: the linear types can be interpreted as grammars, whereas the
non-linear types can be interpreted as ordinary sets. As in prior work
on dependent linear-non-linear types, the linear types are allowed to
depend on the non-linear types but not vice-versa \cite{krishnaswami_integrating_2015}.

The substructural nature of \theoryabbv is well-aligned with the
requirements intrinsic to parsing and to the theory of formal
languages, where strings constitute a very clear notion of resource
that cannot be duplicated, reordered, or dropped. Moreover, the constructive
aspect of \theoryabbv ensures that verification of parsers written in the
calculus are \emph{correct-by-construction}. The type system is powerful enough
that derivations of a term type checking carry intrinsic proofs of correctness.

\paragraph{Contributions} This paper begins in \cref{sec:synindnotion} by
giving a universal semantic characterization of formal grammar.
From there, in \cref{sec:tt} we provide the syntax of \theoryname, a type theory
that axiomatizes the underlying structures needed to conduct formal grammar
theory. In \cref{sec:automata} we demonstrate the usage of
\theoryname for defining parser terms and internalizing results from formal language theory.
Then in \cref{sec:categorify}, we characterize the abstract categorical
models of our syntax. Afterwards in \cref{sec:othermodels}, we demonstrate several models,
including one to prove a metatheoretic canonicity theorem.
Our contributions are then:
%
\begin{itemize}
  \item A syntax-independent notion of grammar given by the category $\Grammar$
  \item \theoryname: A dependent
    linear-non-linear type theory meant to syntactically capture the latent structures
    in $\Grammar$ that enable parsing
  \item Categorical structures that capture the concepts integral to formal grammar theory, and that are notably modeled by $\Grammar$
  \item An implementation of the language, which is embedded shallowly in Agda
\end{itemize}

\section{A Syntax-Independent Notion of Syntax}
\label{sec:synindnotion}

The notion of formal language is central to the theory of parsing.
This definition is
especially useful as it gives a semantics to formal grammars that is completely
independent of any particular syntactic grammar formalism. Any new
notion of grammar can be given a language semantics and it provides a precise
mathematical specification for implementing a \emph{recognizer} of a language.
In this section we propose a \emph{proof-relevant} generalization of formal
languages that can accommodate richer aspects of formal language theory such as
parsing. We begin by proposing a presheaf approach to grammars that encompasses
other classes of grammars in a syntax-independent way.

\paragraph{From Formal Languages to Formal Grammars}

Our first contribution is a novel and syntax-independent characterization of
formal grammar:

\begin{definition}
  \label{def:grammar}
  A \emph{formal grammar} $G$ over a fixed alphabet $\Sigma$ is a function
  $G : \String \to \Set$.
\end{definition}

We say that a grammar $G$ associates to every string $w$ the set\footnote{We
  make little comment as to which foundations are used to encode $\Set$. Our
  construction is polymorphic in choice of a proper class of small sets, a
  Grothendieck universe, a type theoretic universe, or any similar foundation.} $G w$ of parse
trees showing that $w$ matches $G$. This definition of formal grammar serves as
a mathematical specification for a parser --- taking on the same role as
languages do for recognizers. In this sense, our new characterization of
grammars is just a proof-relevant generalization of the existing notion of
formal language.

By ranging over all input strings, notice that each grammar induces a language.
For each formal grammar $G$, define the language of accepted strings $L_{G}$ as,

\[ L_{G} = \{ w \in \String ~|~ G w~\text{is inhabited} \} \]

As they are simply sets, formal languages are naturally endowed with a partial
order via subset inclusion. Similarly, formal grammars naturally coalesce into a
\emph{category}. Define $\Grammar$ to be the category of formal grammars. Whose
objects are grammars, and whose morphisms are functions $f : G \to H$ between two
grammars presented by a family of functions

\[ f^{w} : G w \to H w \]

for every string $w \in \String$. Intuitively, we read a morphism of grammars as
a \emph{parse transformer} --- i.e. $f$ translates a $G$-parse of $w$ into an $H$-parse
of $w$. The induced notion of isomorphism in $\Grammar$ encodes that two grammars are equivalent if there is
a bijective translation of parses. In other words, two grammars are isomorphic
if they capture the same structural descriptions of strings, which is
precisely Chomsky's notion of \emph{strong equivalence} of
grammars \cite{chom1963}. In works like
\cite{yoshinaga2002formal}, this notion of strong equivalence is often treated
like the appropriate choice of ``sameness'' for two grammars. When abstracting over the latent
structure of formal grammars and describing it in the the language of
categories, it is encouraging to see the same definition of equivalence arise
naturally.

Upon inspection of the above definition, $\Grammar$ may equivalently
be described as the functor category $\Set^{\String}$ --- where the
set $\String$ is viewed as a discrete\footnote{Objects in $\String$
comprise the set of all strings, and the only morphisms are the
identity morphisms.}  category. In fact, in this paper we will take
$\Grammar$ to be defined as the functor category $\Set^{\String}$ ---
also called presheaf category. Such presheaf categories carry a
remarkable amount of structure.  Indeed they are a monoidal topos and,
as such, are a model to a substructural dependently typed theory.

\paragraph{From Grammars to Formal Grammars}
The category-theoretic framework can be further used to describe and compare
different notions of formal grammar. For instance, Chomskyan generative
grammars, semi-Thue systems, Montague grammars, and so on are all distinct syntactic
presentations of the same underlying idea of an abstract specification for
parsing. Each of these is an instance of a general \emph{notion of formal
  grammar} --- a category paired with a functor into
$\Grammar$. That is, they may all be interpreted in our category of grammars. In
this sense, $\Grammar$ provides an
all-encompassing semantic domain for parsing.

\max{this section kind of trails off at the end with a very vague aside about a $2$-category of grammar formalisms. Maybe this should go in future work/discussion? What is the message of this section? How does it tie in with the message of the paper? What even is the message of this paper?}

\pedro{Would building on this proposed structure make the message of the section clearer? In this case
  we would have to be a bit clearer in terms of how familiar notions of grammar can be seen as categories with functor into $\Grammar$.}

\section{\theoryname as a Syntax for Formal Grammars}
\label{sec:tt}

Omission of the structural rules of a deductive system, such as in
linear logic \cite{GIRARD19871}, offers precise control over how a value is used
in a derivation. Linear logic omits the weakening and contraction rules
to ensure that every value in context is used exactly once. This control enables
\emph{resource-sensitive} reasoning, where we may treat a resource as
\emph{consumed} after usage. This viewpoint is amenable to parsing applications,
as we may treat characters of a string as finite resources that are consumed at
parse-time. That is, when reading a string, the occurrence of any character is read once. Freely duplicating or dropping characters from a string changes the meaning of that string. \max{what about a backtracking parser?} \steven{Does the sentence sound better now? I've tried to avoid suggesting any particular parser design. Also, I think backtracking is a very interesting question to talk about in future work}. One may then envision a linear type system where the types comprise
formal grammars generated over some alphabet $\Sigma$, and the type constructors
correspond precisely to inductive constructions on grammars --- such as
conjunction, disjunction, concatenation, etc.

Unfortunately, programming in a linear term language is often
cumbersome and unintuitive for users who are unaccustomed to the usage
restrictions.\max{You are arguing against our own type theory here...} \steven{I
  want to argue against a \emph{purely linear} calculus to motivate the
  convenience of LNL, but perhaps this isn't appropriate here, and I don't want to
  make our own type theory sound bad.Should probably just cut the beginning of
  this paragraph and jump into LNL, but then there's not an immediate reason for
  why one might want/need it over a pure linear system}
\pedro{I think that it's reasonable to simply say something like ``programming in a purely linear term language
is limiting due to variable usage restrictions''. In this case we're not throwing our own TT under the bus}
Code in such a language can
become unnecessarily verbose when ensuring every function written is indeed linear. It would be more ergonomic to allow users using Girard encodings to translate
high-level reasoning into low-level linear terms.\pedro{What encodings are these?
We should add a citation here}
To alleviate this
concern, in 1995 Benton et al.\ proposed an alternative categorical
semantics of linear logic based on adjoint interactions between linear
and non-linear logics \cite{bentonMixedLinearNonlinear1995} ---
appropriately referred to as a \emph{linear-non-linear} (LNL)
system. This work is simply typed, so the boundary between linear and
non-linear subtheories is entirely characterized via a monoidal
adjunction $F \dashv G$ between linear terms and non-linear terms.

Inside of an LNL system, linearity may be thought of as an option that users can
choose to deliberately invoke at deliberate points in their developments in an
otherwise intuitionistic system. However, if we are wishing to treat parsers as
linear terms over input strings, the non-linear fragment of an LNL theory does
not really assist in the development of parsers. It is instead the case that
parsers may benefit from a \emph{dependence} on non-linear terms.
Through the approach described by Krishnaswami et al.\ in
\cite{krishnaswami_integrating_2015},
we may define a restricted form of dependent types. In particular, dependence
on linear terms
is disallowed; however, through dependence of a linear term on a non-linear
index, we recover the definition of Aho's indexed grammars \cite{AhoIndexed}
internal to \theoryabbv.

\steven{The above is meant to walk the reader into accepting LNL as a natural
  choice, but it doesn't really flow well to the beginning of the next subsection }

In this section we go on a tour of \theoryabbv. We start by presenting its syntax.
In order to avoid a notion of raw-term, we present it in an intrinsically-typed fashion,
where only well-typed terms are meaningful. Next, we show how it admits a model in $\Set^{\String}$
where the linear types have natural interpretation in terms of formal languages --- for instance,
the tensor corresponds to grammar concatenation. We conclude the section by showing how
classes of grammars correspond to certain sub-type-theories inside \theoryabbv.

\pedro{Hopefully this serves as a segue between this and the syntax subsection}
\subsection{Syntax}
\label{subsec:syntax}

Below we describe \theoryname~(\theoryabbv), an instance of an LNL theory with dependent types
to serve as a deductive
setting for formal grammar theory. \theoryabbv axiomatizes the necessary
structure underlying $\Grammar$ to
specify and construct parsers.

The structural judgments are shown in \cref{fig:structjdg}, the typing
well-formedness in \cref{fig:typewf}, and the intuitionistic typing
rules in \cref{fig:inttyping}. These are mostly just as they appear in
\cite{krishnaswami_integrating_2015}. It has two main differences. The
first one it the presence of two-distinct arrow types, one for adding
variables to the beginning of contexts and another one for adding
variables to the end of contexts. This is an adequate change in the
context of grammars, which are inherently non-symmetric. The second
change is the introduction of inductive types, which allows for the definition
of recursive grammars. The treatment of the non-linear
types is standard, so below we focus on the linear syntax.

\begin{figure}
  \begin{mathpar}
    \boxed{\ctxwff \Gamma}

    \inferrule{~}{\ctxwff \cdot}
    \and
    \inferrule{\ctxwff \Gamma \\ \ctxwffjdg \Gamma X}{\ctxwff {\Gamma, x : X}}

    \\

    \boxed{\linctxwff \Gamma \Delta}

    \inferrule{~}{\linctxwff \Gamma \cdot}
    \and
    \inferrule{\linctxwff \Gamma \Delta \\ \linctxwffjdg \gamma A}{\linctxwff \Gamma {\Delta, a : A}}

    \\

    \boxed{\ctxwffjdg \Gamma X}

    \inferrule{\Gamma \vdash X : U_i}{\ctxwffjdg \Gamma X}

    \boxed{\linctxwffjdg \Gamma A}

    \inferrule{\Gamma \vdash A : L_i}{\linctxwffjdg \Gamma A}

    \\

    \boxed{\ctxwffjdg \Gamma {X \equiv Y}}

    \inferrule{\Gamma \vdash X \equiv Y : U_i}{\ctxwffjdg \Gamma {X\equiv Y}}

    \boxed{\linctxwffjdg \Gamma {A \equiv B}}

    \inferrule{\Gamma \vdash A \equiv B : L_i}{\linctxwffjdg \Gamma {A \equiv B}}

  \end{mathpar}
  \caption{Structural judgments}
  \label{fig:structjdg}
\end{figure}

\begin{figure}
  \begin{mathpar}
    \boxed{\Gamma \vdash X : U_{i}}

    \boxed{\Gamma \vdash A : L_{i}}

    \inferrule{~}{\Gamma \vdash U_i : U_{i+1}}
 %
    \and
%
    \inferrule{~}{\Gamma \vdash L_i : U_{i+1}}
%
    \and
%
    \inferrule{\Gamma \vdash X : U_i \\ \hspace{-0.1cm} \Gamma, x : X \vdash Y : U_i}{\Gamma \vdash \PiTy x X Y : U_i }%
%
    \and
%
    \inferrule{\Gamma\vdash X : U_i \\ \hspace{-0.1cm} \Gamma, x : X \vdash Y : U_i}{\Gamma \vdash \SigTy x X Y : U_i}
%
    \and
%
    \inferrule{~}{\Gamma \vdash 1 : U_i}
%
    \and
%
    \inferrule{\Gamma \vdash A : L_i}{\Gamma \vdash G A : U_i}

    \and
%
    \inferrule{~}{\Gamma \vdash I : L_i}
 %
    \and
%
    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash A \otimes B : L_i}
%
    \and
%
    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash A \lto B : L_i}
%
    \and
%
    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash B \tol A : L_i}
%
    \and
%
    \inferrule{\Gamma \vdash X : U_i \\ \Gamma, x : X \vdash A : L_i}{\Gamma \vdash \LinPiTy x X A : L_i}
%
    \and
%
    \inferrule{\Gamma \vdash X : U_i \\ \Gamma, x : X \vdash A : L_i}{\Gamma \vdash \LinSigTy x X A : L_i}
%
    \and
%
    \inferrule{\Gamma \vdash X : U_i \quad \{\Gamma \vdash e_i : X\}_i}{\Gamma \vdash e_1 =_X e_2 : U_i}
    %
    \and
    %
    \inferrule{~}{\Gamma \vdash \top : L_i}
%
    \and

    \inferrule{~}{\Gamma \vdash \bot : L_i}

    \and
    \inferrule{c \in \Sigma}{\Gamma \vdash c : L_0}
    %
    \and
    %
    \inferrule{\Gamma, x : L_i \vdash A : L_i \and A \textrm{ strictly positive}}{\Gamma \vdash \mu x.\, A : L_i}
  \end{mathpar}
  \caption{Type well-formedness}
  \label{fig:typewf}
\end{figure}

\begin{figure}
  \begin{mathpar}
  \boxed{\Gamma \vdash x : X}

  \inferrule{~}{\Gamma, x : X, \Gamma' \vdash x : X}
  %
  \and
  %
  \inferrule{\Gamma \vdash e : Y \quad \ctxwffjdg \Gamma {X \equiv Y}}{\Gamma \vdash e : X}
  %
  \and
  %
  \inferrule{~}{\Gamma \vdash () : 1}
  %
  \and
  %
  \inferrule{\Gamma \vdash e : X \\ \Gamma \vdash e : \subst Y e x}{\Gamma \vdash (e, e') : \SigTy x X Y}
  %
  \and
%
  \inferrule{\Gamma \vdash e : \SigTy x X Y}{\Gamma \vdash \pi_1\, e : X}
  %
  \and
  %
  \inferrule{\Gamma \vdash e : \SigTy x X Y}{\Gamma \vdash \pi_2\, e : \subst Y {\pi_1\, e} x}
  \and
  \inferrule{\Gamma, x : X \vdash e : Y}{\Gamma \vdash \lamb x e : \PiTy x X Y}
  %
  \and
  %
  \inferrule{\Gamma \vdash e : \PiTy x X Y \\ \Gamma \vdash e' : X}{\Gamma \vdash \app e {e'} : \subst Y {e'} x}
  %
  \and
  %
  \inferrule{\Gamma \vdash e \equiv e' : X}{\Gamma \vdash \mathsf{refl} : e =_X e'}
  \and
  \inferrule{\Gamma ; \cdot \vdash e : A}{\Gamma \vdash \mathsf G e : \mathsf G A}
  \end{mathpar}
  \caption{Intuitionistic typing}
  \label{fig:inttyping}
\end{figure}

\begin{figure}
  \begin{mathpar}
    \boxed{\Gamma ; \Delta \vdash a : A}

    \inferrule{~}{\Gamma ; a : A \vdash a : A}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : B \\ \linctxwffjdg \Gamma {A \equiv B}}{\Gamma ; \Delta \vdash e : A}
    %
    \\
    %
    \inferrule{~}{\Gamma ; \cdot \vdash () : I}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : I \\ \Gamma ; \Delta_1',\Delta_2' \vdash e' : C}{\Gamma ; \Delta_1',\Delta,\Delta_2' \vdash \letin {()} e {e'} : C}
    %
    \\
    %
    \inferrule{\Gamma ; \Delta \vdash e : A \\ \Gamma ; \Delta' \vdash e' : B}{\Gamma ; \Delta, \Delta' \vdash e \otimes e' : A \otimes B}
    %
    \and
    %
    \inferrule{\Gamma ; \Delta \vdash e : A \otimes B \\ \Gamma ; \Delta'_1, a : A, b : B, \Delta'_2 \vdash e'}{\Gamma ;  \Delta_1', \Delta, \Delta'_2 \vdash \letin {a \otimes b} e {e'}}
    \\
    %
    \inferrule{\Gamma ; a : A , \Delta \vdash e : B}{\Gamma ; \Delta \vdash \lamblto a e : A\lto B}
    \and
    \inferrule{\Gamma ; \Delta' \vdash e' : A \\ \Gamma ; \Delta \vdash e : A \lto B}{\Gamma ; \Delta', \Delta \vdash \applto {e'} {e} : B}
    \\
    %
    \inferrule{\Gamma ; \Delta , a : A \vdash e : B}{\Gamma ; \Delta \vdash \lambtol a e : B\tol A}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : B \tol A \\ \Gamma ; \Delta' \vdash e' : A}{\Gamma ; \Delta, \Delta' \vdash \apptol e {e'} : B}
    %
    \\
    %
    \inferrule{\Gamma, x : X ; \Delta  \vdash e : A}{\Gamma ; \Delta \vdash \dlamb x e : \LinPiTy x X A}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : \LinPiTy x X A \\ \Gamma \vdash e' : X}{\Gamma ; \Delta \vdash \app e {e'} : \subst A {e'} x}
    %
    \\
    %
    \inferrule{\Gamma \vdash e : X \quad \Gamma ; \Delta \vdash e' : \subst A e x}{\Gamma ; \Delta \vdash (e, e') : \LinSigTy x X A}
    %
    \and
    %
    \inferrule{\Gamma ; \Delta \vdash e : \LinSigTy x X A \quad \Gamma, x : X ; \Delta'_1, a : A, \Delta'_2 \vdash e' : C}{\Gamma; \Delta'_1, \Delta, \Delta'_2 \vdash \letin {(x, a)} e {e'}: C}
    %
    \\
    %
    \inferrule{~}{\Gamma ; \Delta \vdash () : \top}

    \and

    \inferrule{~}{\Gamma ; x : \bot \vdash \mathsf{absurd}_A : A}
    %
    \and
    %
    %
    \inferrule{\Gamma \vdash e : \mathsf{G} A}{\Gamma ; \cdot \vdash \mathsf{G}^{-1}\, e : A}
    %
    \\
    %
    \inferrule{\Gamma; \Delta \vdash e : \subst A {\mu x.\, A} x}{\Gamma; \Delta \vdash \mathsf{cons}\, e : \mu x.\, A}
    \and
    \inferrule{\Gamma;\Delta\vdash e' : \mu x.\,A \and \Gamma; a:\subst A B x \vdash e : B}{\Gamma;\Delta\vdash \mathsf{fold}(a.e)(e') : B}
  \end{mathpar}
  \caption{Linear typing}
  \label{fig:linsyntax}
\end{figure}

\begin{figure}
  \begin{mathpar}
    \boxed{\Gamma \vdash e \equiv e' : X}

    \inferrule{\Gamma \vdash p : e =_X e'}{\Gamma \vdash e \equiv e' : X}
%
    \and
%
    \inferrule{~}{\Gamma \vdash \app {(\lamb x e)} {e'} \equiv \subst x e {e'} : X}
%
    \and
%
    \inferrule{~}{\Gamma \vdash e \equiv \lamb x {\app e x} : \PiTy x X Y}
%
    \and
%
    \inferrule{~}{\Gamma \vdash \pi_1\, (e_1, e_2) \equiv e_1 : X}
%
    \and
%
    \inferrule{~}{\Gamma \vdash \pi_2\, (e_1, e_2) \equiv e_2 : \subst x {e_1} Y}
%
    \and
%
    \inferrule{~}{\Gamma \vdash e \equiv (\pi_1\, e, \pi_2\, e) : \SigTy x X Y}
%
    \and
%
    \inferrule{~}{}
%
    \inferrule{~}{\Gamma \vdash t \equiv t' : 1}
%
    \and
%
    \inferrule{~}{\Gamma \vdash G\, (G^{-1} \, t) \equiv t : G A}

    \\

    \boxed{\Gamma ; \Delta \vdash a \equiv a' : A}

    \inferrule{~}{\Gamma; \cdot \vdash G^{-1}\, (G \, t ) \equiv t: A}
%
    \\
%
    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lamblto a e)} {e'} \equiv \subst e x {e'} : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lamblto a {\app e a} : A \lto B}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lambtol a e)} {e'} \equiv \subst e x {e'} : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lambtol a {\app e a} : A \tol B}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \app {(\dlamb x e)} {e'} \equiv \subst x a {e'} : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \dlamb x {\app e x} : \LinPiTy x X A}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv e' : \top}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e_i \equiv \pi_i (e_1, e_2) : A_i}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv (\pi_1 e, \pi_2 e) : A\& B}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} {()} e \equiv e : C}
    %
    \and
    %
    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} e {\subst {e'} {()} a} \equiv \subst {e'} a e : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \letin {e \otimes e'} {a \otimes a'} e'' \equiv \subst {e''} {a, a'} {e, e'} : C}
%
    \and
    %
    \inferrule{~}{\Gamma; \Delta \vdash \letin {a \otimes b} e {\subst {e'} {a \otimes b} c} \equiv \subst {e'} c e : C}
%
    \and
%
    \inferrule{~}{\Gamma;\Delta \vdash \letin {(x, a)} {(e, e')} {e''} \equiv \subst {e''} {x, a} {e, e'} : C}
    %
    \and
    %
    \inferrule{~}{\Gamma; \Delta \vdash \letin {(x, a)} e {\subst {e'} {(x, a)} y} \equiv \subst {e'} e y : C}
\end{mathpar}
  \caption{Judgmental equality}
  \label{fig:jdgeq}
\end{figure}

\subsection{Formal Grammars as Linear Types}
\label{sec:formaltype}

The linear typing judgment in our syntax takes on the following schematic form
$\Gamma ; \Delta \vdash a : A$. First, $A$ represents a \emph{linear type} in
our syntax. The intended semantics of these linear types are formal grammars.
That is, the linear typing system is designed to syntactically reflect the
behavior of formal grammars. For this reason, we may often interchangeably use
the terms ``linear type'' and ``grammar''.

The term $a$ is an inhabitant of type $A$, which is thought of as a parse tree of
the grammar $A$. The core idea of this entire paper follows precisely from this
single correspondence: grammars are types, and the inhabitants of these types
are parse trees for the grammars.

$\Gamma$ represents a non-linear context, while $\Delta$ represents a
\emph{linear context} dependent on $\Gamma$. These linear contexts behave
substructurally. As stated earlier, they are linear --- they do not obey
weakening or contraction --- because a character is exhausted once read by a
parsing procedure. Moreover, the characters in strings appear in the order in
which we read them. We do not have the freedom to freely permute characters,
therefore any type theory that is used to reason about formal grammars ought to
omit the structural rule of exchange as well. This means that every variable
within $\Delta$ must be used \emph{exactly once} and \emph{in order of
  occurrence}. Thus, we can think of the linear contexts as an ordered list of
limited resources. Once a resource is consumed, we cannot make reference to it
again. Variables in a linear context then act like building
blocks for constructing patterns over strings.

We give the base types and type constructors for linear terms. As the
interpretation of types as grammars in $\Grammar$ serves as our intended
semantics, we simultaneously give the interpretation $\sem \cdot$ of the
semantics as grammars.

\pedro{The paragraphs above are a bit circular, what about the following alternative?}
\steven{I am in favor of cutting what's above, but we haven't introduced the
  syntax in the main body of the paper yet, even if the figures appeared
  earlier. So what you write needs to take that into account}
\pedro{In which case we should introduce it in the previous section. I suggest
  adding this right after ``and the intuitionistic typing rules in fig 3''}

Now that we have specified the syntactic aspect of \theoryabbv, we must justify
its connections to formal grammars in order to show that it is a sound methodology
for reasoning about parsers. We do this by showing that $\mathbf{Gr}$ is a model
to our type theory.

Most of the non-linear semantics is standard \pedro{We should either cite something or add the non-linear semantics in the appendix}, so let us investigate in more
depth the linear semantics.

\begin{enumerate}
  \item A non-linear context $\Gamma$ denotes a set $\sem \Gamma$
  \item A non-linear type $\Gamma \vdash X : U_{i}$ denotes a family of sets
        $\sem X : \sem \Gamma \to \Set_{i}$
  \item A non-linear term $\Gamma \vdash e : X$ denotes a section
        $\sem e : \Pi(\gamma : \sem \Gamma)\sem{X} \gamma$ \pedro{we have not defined what sections are}
  \item Linear contexts $\Gamma \vdash \Delta$ and types
        $\Gamma \vdash A : L_{i}$ both denote families of grammars
        $\sem \Gamma \to \Gr_{i}$
  \item A linear term $\Gamma ; \Delta \vdash M : A$ denotes a family of parse
        transformers
        $\sem M : \Pi(\gamma : \sem \Gamma)\Pi(w : \String) \sem \Delta \gamma w \Rightarrow \sem M \gamma w$
\end{enumerate}

\paragraph{Linear Unit}
First, the linear unit $I$ may be built in the empty linear
context\footnote{When appropriate, statements like ``the empty linear context''
  may be taken polymorphically over all non-linear contexts that they might
  depend on.} $\cdot \dashv I$. $I$ serves as the unit for the operation
$\otimes$ described below.

As a grammar,
%
\[
  \sem {I} \gamma w = \{ () ~|~ w = \varepsilon \}
\]
%
That is, $I$ maps strings to the proposition that they are the empty string. $I$
is clearly only inhabited by the empty string $\varepsilon$, for which the
outputted set contains a single parse tree.

\paragraph{Base Types}
For each character $c$ in the alphabet $\Sigma$, we include a base type at the
lowest universe level $c : L_{0}$.

The grammar interpretation for characters is quite similar to that of $I$.
%
\[
  \sem {c} \gamma w = \{ () ~|~ w = c \}
\]
%
The grammar for the character $c$ likewise maps strings $w$ to the proposition that
they are equal to $c$.

\paragraph{Tensor Product}
The tensor product of two linear types is the first place where the ordering on
contexts really takes effect. That is, the tensor product of two types is
non-commutative.
When context $\Delta$ forms a linear term of
type $A$, and $\Delta'$ forms a linear term of type $B$, then the context
extension $\Delta, \Delta'$ forms a linear term of type $A \otimes B$.

As stated above, the type $I$ serves as the unit for $\otimes$. That is, for all
linear types $A$, the following equalities hold:
%
\[ I \otimes A \equiv A \otimes I \equiv A \]
%
In the grammar semantics,
%
\[
  \sem {A \otimes B} = \Sigma_{w_{1}, w_{2} : \String} (w_{1}w_{2} = w) \land \left(  \sem {A} \gamma w_{1} \times \sem {B} \gamma w_{2} \right)
\]
%
The string $w$ matches $A \otimes B$
precisely when it may be split into two
pieces such that the left one matches $A$
and the right one matches $B$.

\paragraph{Linear Function Types}
The monoidal structure provided by $\otimes$ is both left and right closed and
this is denoted by the left and right linear function types. Further,
for linear types $A$ and $B$ the left linear function type $A \lto B$
enjoys an elimination principle similar to function application or modus ponens.
%
\[
\inferrule{\Delta \vdash A \otimes A \lto B}{\Delta \vdash B}
\]
%
The interpretation of linear function types at a string $w$ is a linear function
that takes in parses of $A$ on a string $w_{a}$ and outputs parses of $B$ on the
string $w_{a} w$.
%
\[
  \sem{A \lto B} \gamma w = \Pi(w_a:\String) \left( A \gamma w_a \Rightarrow B\gamma (w_a w) \right)
\]
%
That is, strings match $A \lto B$ if when prepended with a parse of $A$ they
complete to parses of $B$. In this manner, the linear function types generalize
Brzozowksi's notion of derivative
\cite{brzozowskiDerivativesRegularExpressions1964}.
Brzozowski initially only gave an accounting of this operation for
generalized regular expressions, but later work from Might et al.\ demonstrates that the same
construction can be generalized to context free grammars
\cite{mightParsingDerivativesFunctional2011}. Here, via the linear function
types, the same notion of derivative extends to the grammars of \theoryabbv.

Note, to ensure that the linear function types do indeed generalize Brzozowski
derivatives, we must include the equalities in \cref{fig:brzozowskideriv} as axioms\max{we must? why must we}.
\pedro{We should further explain why these equations are required, or at the very
least give some intuition}

Of course, all the above statements for the left function type also have
corresponding analogues for the
right-handed counterpart.

\begin{figure}
\begin{align*}
  c\lto c &\equiv I\\
  c\lto d &\equiv \bot\\
  c\lto I &\equiv \bot\\
  c\lto 0 &\equiv \bot\\
  c\lto (A \otimes B) & \equiv (c\lto A) \otimes B + (A \amp I) \otimes (c\lto B)\\
  c\lto A^* &\equiv (A \amp I)^{*} \otimes (c \lto A) \otimes A^* \\
  c\lto (A + B) &\equiv (c\lto A) + (c\lto B)
\end{align*}
\caption{Equality for Brzozowski Derivatives}
\label{fig:brzozowskideriv}
\end{figure}

\paragraph{LNL Dependent Types}
Given a non-linear type $X$, we may form both the dependent product type
$\LinPiTy x X A$ and the dependent pair type $\LinSigTy x X A$ as linear types
where $x$ is free in $A$.

$\sem{\LinPiTy x X A} \gamma w = \Pi(x:\sem{X}\gamma) \sem{A}(\gamma,x) w$

$\sem{\LinSigTy x X A} \gamma w = \Sigma(x:\sem{X}\gamma) \sem{A}(\gamma,x) w$

The grammar semantics of the linear product type is
indeed a dependent function out of the
semantics of $X$. Likewise, the grammar semantics of the linear dependent pair type is a dependent
pair in $\Set$.\pedro{The connection to grammar semantics needs to be better explained. For instance,
we should explain that disjunction corresponds to ``or'' of grammars}

Note that even though we do not take the binary additive conjunction and
disjunction as primitive, we may define them via these LNL dependent types. In
particular, via a dependence on $\Bool$.
%
\[
  A \amp B := \LinPiTy b \Bool {C(b)}
\]
\[
  A \oplus B := \LinSigTy b \Bool {C(b)},
\]
where $C(\true) = A, C(\false) = B$.

\paragraph{Universal Type}
The universal type $\top$ may be formed in any context.
%
\[ \sem{\top} \gamma w = \{ \ast \}\]
%
Its grammar semantics in set outputs the unit type in $\Set$ for all input strings in all contexts.

\paragraph{Empty Type}
The empty type $\bot$ has no inhabitants. The elimination for the empty type
witnesses the principle of explosion --- i.e.\ from a term of type $\bot$ we may
introduce a term $\mathsf{absurd}_{A} : A$ for any type $A$.
%
\[ \sem {\bot} \gamma w = \emptyset \]
%
\paragraph{The $G$ Modality}
The $G$ modality provides the embedding of linear types in the empty context
into non-linear types. The introduction and elimination forms for $G A$ obey the
same rules as given in \cite{bentonMixedLinearNonlinear1995} and \cite{krishnaswami_integrating_2015}.

The left adjoint $F$ from non-linear types back to linear types may be defined
as,
%
\[
  F X := \LinSigTy a A I
\]
%
\steven{Elaborate on $G$ more}

\pedro{Agreed :) maybe mention connections to persistent propositions from the
separation logic literature}

The semantics of $G A$ are exactly the semantics of $A$ at the empty word $\varepsilon$.
%
\[ \sem{G A} \gamma = \sem{A} \gamma \varepsilon \]
%
%
\paragraph{Recursive Types}
Recursive linear types may be defined via the least-fixed point operator $\mu$.
The grammar semantics of which are the fixed-point of sets induced by the
grammar semantics of $A$.
%
\[ \sem{\mu x. A} \gamma = \mu (x:\Gr_i). \sem{A}(\gamma,x) \]
%
\pedro{We should justify, even if briefly, the existence of this
  fixed-point}
\steven{As a consequence of Knaster-Tarski}

Observe that we need not take the Kleene star as a primitive
grammar constructor, as it is definable as a fixed point.
The Kleene star of a grammar $g$ is given as,

\[
  g^{*} := \mu X . I \oplus (g \otimes X)
\]

\begin{figure}[h!]
\begin{mathpar}
  \inferrule
  {\Gamma ; \Delta \vdash p : I}
  {\Gamma ; \Delta \vdash \mathsf{nil} : g^{*}}

  \and

  \inferrule
  {\Gamma ; \Delta \vdash p : g \\ \Gamma ; \Delta' \vdash q
  : g^{*}}
  {\Gamma ; \Delta \vdash \mathsf{cons}(p , q) : g^{*}}

  \\

  \inferrule
  {
    \Gamma ; \Delta \vdash p : g^{*} \\
    \Gamma ; \cdot \vdash p_{\varepsilon} : h \\
    \Gamma ; x : g , y : h \vdash p_{\ast} : h
  }
  {\Gamma ; \Delta \vdash \mathsf{foldr}(p_{\varepsilon} , p_{\ast}) : g^{*}}
\end{mathpar}
\caption{Kleene Star Rules}
\label{fig:star}
\end{figure}

Likewise, $g^{*}$ has admissible introduction and
elimination rules, shown in \cref{fig:star}. Note that this
definition of $g^{*}$ and these
rules arbitrarily assigns a handedness to the Kleene star.
We could have just as well took it to be a fixed point of
$I \oplus (X \otimes g)$. In fact, the definitions are
equivalent, as the existence of the $\mathsf{foldl}$ term below
shows that $g^{*}$ is also a fixed point of
$I \oplus (X \otimes g)$.
This is a marked point of difference from Kleene
algebra with recursion, where the fixed points for the left and right variants
of Kleene star need not agree \cite{leiss}.

\begin{equation}
  \label{eq:foldl}
  \inferrule
  {
    \Gamma ; \Delta \vdash p : g^{*} \\
    \Gamma ; \cdot \vdash p_{\varepsilon} : h \\
    \Gamma ; y : h , x : h \vdash p_{\ast} : h
  }
  {\Gamma ; \Delta \vdash \mathsf{foldl}(p_{\varepsilon} , p_{\ast}) : g^{*}}
\end{equation}

In fact, the $\mathsf{foldl}$ term is defined using
$\mathsf{foldr}$ --- much in the same way one
may define a left fold over lists in terms of a right fold
in a functional programming language.
The underlying trick is to fold over a list of linear functions
instead of the original string. We curry each character $c$
of the string into a function that concatenates $c$, and
right fold over this list of linear functions. Because function
application is left-associative, this results in a left
fold over the original string.

We only take fixed points of a single variable as a
primitive operation in the type theory, but we may apply
Beki\`c's theorem \cite{Bekić1984} to define an admissible
notion of multivariate fixed point. This is particularly
useful for defining grammars that encode the states of an
automaton, as we will see in \cref{sec:automata}. In \cref{fig:multifix} we provide the
introduction and elimination principles for such a fixed
point, where $\sigma$ is the substitution that unrolls the
mutually recursive definitions one level. That is,

\begin{align*}
  \sigma = \{ & \mu(X_{1} = A_{1} \dots, X_{n} = A_{n}).X_{1} / X_{1} , \dots, \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{n} / X_{n} \}
\end{align*}

\begin{figure}
\begin{mathpar}
  \inferrule
  {\Gamma ; \Delta \vdash e : \simulsubst {A_{k}} {\sigma}}
  {\Gamma ; \Delta \vdash \mathsf{cons}~e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{k}}

  \and

  \inferrule
  {\Gamma ; \Delta \vdash e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{k} \\
             \Gamma ; x_{j} : \simulsubst {A_{j}}{\sigma} \vdash e_{j} : B_{j}\quad \forall j
  }
  {\Gamma; \Delta \vdash \mathsf{mfold}(x_{1}.e_{1}, \dots, x_{n}.e_{n})(e) : B_{k}}
\end{mathpar}
\caption{Multi-fixed Points}
\label{fig:multifix}
\end{figure}

\subsection{Subtheories}
Each choice of connective has direct implications on the expressivity of
\theoryabbv. In particular, the set of connectives used will determine where the
grammar semantics is placed in
the Chomsky hierarchy.

\paragraph{Regular Expressions} We may realize regular expression as a subtheory
of our language by restricting the type constructors to the
linear unit, characters, $\otimes$, $\oplus$, and Kleene star. Classically, the
languages recognized by these are exactly the regular languages --- the lowest
on the Chomsky hierarchy.

\paragraph{$\mu$-Regular Expressions} Similarly, we may restrict the connectives
to be linear unit, characters, $\otimes$, $\oplus$, and arbitrary recursion via
left-fixed point rather than only Kleene star.
Instead of regular expressions these correspond to Lei{\ss}'s $\mu$-regular
expressions, which are known to be equivalent to context-free grammars
\cite{leiss,krishnaswami_typed_2019}.

\paragraph{Beyond Context-Free Grammars} The previous two subtheories induce as semantics
regular grammars and context-free grammars, respectively; however, by including
the LNL dependent types we may actually express the entirety of the Chomsky
hierarchy. Through use of the LNL dependent types, we may encode indexed
grammars, which are known to be properly between context-free and
context-sensitive grammars within the Chomsky hierarchy \cite{AhoIndexed}. We
will further see in \cref{subsubsec:tm} how to use dependence to encode Turing machines internal to our
calculus, which of course induces a semantics of unrestricted grammars.

\section{Automata as Grammars}
\label{sec:automata}

The canonicity theorem from the previous section gives a tight connection
between terms in our type theory and the theory of formal languages. In this section
we further explore these connections by using the fact that, classically
formal language theory is closely related to the study of automata.
In the Chomsky hierarchy, each language class is associated to a class of
automata that serve as recognizers. Internal to \theoryabbv,
we can characterize these language classes syntactically; moreover, we
demonstrate the equivalence of these language classes to its associated automata class as a
proof term within our logic.

\subsection{Non-deterministic Finite Automata}
\label{subsec:finiteaut}
\begin{figure}
  \begin{tikzpicture}[node distance = 25mm ]
    \node[state, initial, accepting] (1) {$1$};
    \node[state, below left of=1] (2) {$2$};
    \node[state, right of=2] (3) {$3$};

    \path[->] (1) edge[above] node{$b$} (2)
              (1) edge[below, bend right, left=0.3] node{$\epsilon$} (3)
              (2) edge[loop left] node{$a$} (2)
              (2) edge[below] node{$a, b$} (3)
              (3) edge[above, bend right, right=0.3] node{$a$} (1);
  \end{tikzpicture}
  \caption{An example NFA}
  \label{fig:NFA}
\end{figure}

Classically, a \emph{nondeterministic finite automaton} (NFA) is a finite state machine where
transitions are labeled with characters from a fixed alphabet $\Sigma$. These
are often represented formally as a 5-tuple $(Q, \Sigma, \delta, q_{0}, F)$,

\begin{itemize}
  \item $Q$ a finite set of states
  \item $\Sigma$ a fixed, finite alphabet
  \item $\delta : Q \times (\Sigma \cup \{ \varepsilon\}) \to \mathcal{P}(Q)$ the labeled transition function
  \item $q_{0} \in Q$ the start state
  \item $F \subset Q$ a set of accepting states
\end{itemize}

Intuitively, this can be thought of like a directed graph with nodes in $Q$ with
an edge $q \overset{c}{\to} q'$ whenever $q' \in \delta(q, c)$. Note that
transitions in an NFA may be labeled with the empty string $\varepsilon$ --- such
transitions are referred to as \emph{$\varepsilon$-transitions}. We may see an
example of an NFA in \cref{fig:NFA}.

From an NFA, we may construct a grammar of traces as follows:

First, we define a mutual fixed point grammar that describes the traces through the NFA.\ Then, we have another
grammar that tells us if we're currently in an accepting state of the automaton.
A parse of the NFA grammar is then a pair of a trace and the data that we're in
an accepting state.

Consider the NFA $N$ pictured in \cref{fig:NFA}. There are three states, $1$, $2$,
and $3$. We introduce a non-linear type $Q$ with three inhabitants $q_{1}$,
$q_{2}$, and $q_{3}$ to represent each of these states, respectively. Given a
$q, q'$, we can
then define the type of traces from $q$ to $q'$. For instance, let's construct
the traces starting at $q_{1}$ and ending at $q_{2}$ as an example.

\begin{equation}
  \label{eq:nfatrace}
  \mathsf{Trace}_{N}(q_{1}, q_{2}) = \mu
  \begin{pmatrix}
    g_{q_{1}} := g_{q_{3}} \oplus ( b \otimes g_{q_{2}} ) \\
    g_{q_{2}} := ( a \otimes g_{q_{2}} ) \oplus ( a \otimes g_{q_{3}} ) \oplus ( b \otimes g_{q_{3}} ) \oplus I \\
    g_{q_{3}} := a \otimes g_{q_{1}} \\
  \end{pmatrix}. g_{q_{1}}
\end{equation}

We should read this as defining three mutually recursive grammars, one for each
state. The definitions of these mutually recursive grammars capture the
transitions of the automaton. To ensure that we only encode traces that end in
state 2, we only include the unit $I$ as a summand in
$g_{q_{2}}$. That is, by only including $I$ at this location, it makes it the
only place where derivations of $\mathsf{Trace}(q_{1}, q_{2})$ can terminate. We
can think of all of these definitions underneath
of the $\mu$ binder as bringing some local grammars into scope. With these local
grammars in scope we are ultimately constructing a term of type $g_{q_{1}}$ to denote
only traces starting in state 1.

To encode the acceptance criteria of $N$, we want to internalize a proposition
over each state of the NFA.\ That is, for each $q \in Q$ we define a term

\[
  \mathsf{acc}(q) := q \text{ is accepting}
\]

In this example, $\mathsf{acc}(q_{1}) = \top$ and
$\mathsf{acc}(q_{2}) = \mathsf{acc}(q_{3}) = \bot$. An accepting trace of the NFA $N$ is then given by the following dependent grammar,
\[
 \mathsf{AccTrace}_{N} := \LinSigTy q Q {\left( \mathsf{Trace}_{N}(q_{0} , q) \pair \mathsf{acc}(q) \right)}
\]

where $q_{0}$ is the initial state. That is, a trace is accepted by the NFA if
we can construct the trace, and the trace ends at an accepting state. This idea
is simple enough and aligns with how we intuitively treat these automata.

Generalizing over the above example, we want to define the type of traces as,

\[
  \mathsf{Trace}_{N}(q_{0}, q_{1}) = \mu
  \begin{pmatrix}
    g_{q} := \mathsf{Trans}(q), & q \neq q_{1} \\
    g_{q} := \mathsf{Trans}(q) \oplus I , & q = q_{1}
  \end{pmatrix}. g_{q_{0}}
\]

where $\mathsf{Trans}(q)$ is an iterated disjunction that describes which
transition you should take. For an NFA, $\mathsf{Trans}(q)$ must take on the
following syntactic form,

\begin{gather*}
 \mathsf{State} \in \{ g_{q} : q \in Q \} \\
 \mathsf{Char} \in \Sigma \\
 \mathsf{Trans}_{N}(q) ::= \mathsf{Char} \otimes \mathsf{State}~|~\mathsf{State}~|~\mathsf{Trans}_{N}(q) \oplus \mathsf{Trans}_{N}(q)
\end{gather*}

That is, $\mathsf{Trans}(q)$ is a disjunction of literals followed by grammars
that encode states.

When conducting proofs involving NFA grammars, we often to
need either construct terms of type $\mathsf{Trace}_{N}(q , q')$. To
this end, we give three admissible rules for constructing
traces,

\begin{figure}[h!]
  \label{fig:admissibleintro}
  \begin{mathpar}
    \inferrule
    {~}
    {\Gamma ; \cdot \vdash \mathsf{nil} : \mathsf{Trace}_N
      (q , q)}

    \and

    \inferrule
    {\Gamma ; \Delta \vdash M : \mathsf{Trace}_N
      (dst , q') \\
    \exists \text{ transition } src \overset{c}{\to} dst
    }
    {\Gamma ; x : c , \Delta \vdash \mathsf{cons}(M) : \mathsf{Trace}_N
    (src , q')}

    \and

    \inferrule
    {\Gamma ; \Delta \vdash M : \mathsf{Trace}_N (dst , q')
      \\
    \exists~\varepsilon\text{-transition } src
    \overset{\varepsilon}{\to} dst}
    {\Gamma ; \Delta \vdash \mathsf{\varepsilon cons}(M) : \mathsf{Trace}_N
    (src , q')}
  \end{mathpar}
  \caption{Admissible Trace Constructors}
\end{figure}

That is, we may use $\mathsf{nil}$ to terminate a trace that
begins and ends at state $q$. The rules $\mathsf{cons}$ and
$\mathsf{\varepsilon cons}$ are then used to inductively
stitch traces together when sound. For instance, we can read
the $\mathsf{cons}$ rule as saying that we can
create a trace coming from a state $src$ provided that we may first
transition via the character $c$ to state $dst$ and
inductively build a trace starting from $dst$. The
$\mathsf{\varepsilon cons}$ rule says something similar, but
with an $\varepsilon$-transition instead of a transition
labelled by a character. Recall that
these rules are \emph{admissible}. They are not strictly
necessary as primitives to conduct our proofs; however, they
do provide convenient shorthand notation for building terms
of type $\mathsf{Trace}_{N}(q , q')$.

Dual to constructing traces, we often want to construct
other terms in a context containing values of type
$\mathsf{Trace}_{N}(q , q')$. For this purpose, we make use
of the elimination principle for multiple-fixed points ---
which we write as $\mathsf{mfold}$ --- given
in \cref{fig:multifix}.

Note that this general construction will readily generalize to other types of
automata. If we wanted to define say deterministic finite automata\footnote{DFAs
\emph{could} just be defined as NFAs that happen to be deterministic. This is
one way to do so, or you may choose to present the transition relation for the
automaton as a transition \emph{function} instead. Concerns like these don't
matter so much when defining things on paper, but at formalization time these
are important and can make some proofs much easier.}, pushdown automata, Turing
machines, etc, we just swap out the type of traces for a different, but very
similar, trace construction.

With this setting for finite automata, we can now internalize some classical theorems inside of our formal system.

\subsection{A DFA Parser}
\label{subsec:regexparser}

Just as we encoded traces of NFAs as grammars, we likewise
encode the traces of a DFA as grammars. The key difference
between NFAs and DFAs is \emph{determinism} --- meaning,
that in a state $q$ inside of DFA, given a character $c$ there
will be exactly one transition that we may take leaving $q$
with label $c$. For us, this changes the definition of valid
transitions for a DFA, instead of the definition of
$\mathsf{Trans}$ provided in \cref{subsec:finiteaut} DFAs
obey

\begin{gather*}
 \mathsf{State} \in \{ g_{q} : q \in Q \} \\
 \mathsf{Trans}(q) ::= \bigoplus_{c \in \Sigma} (c \otimes \mathsf{State})
\end{gather*}

Meaning, each state has a transition for every character. \emph{Note:} the above use
  of a disjunction over $\Sigma$ is a bit of abuse of notation.
  As $\Sigma$ is a finite alphabet, we wish to think of this as an iterated
  binary sum, or perhaps as a finitely indexed sum defined via linear dependent
  sums in the same manner by which we defined binary sums.

We now wish to define a parser term for DFA grammars. In
particular, for a DFA $D$ we want to build a term,

\[
  w : \String \vdash \mathsf{parse}_{D} : \mathsf{AccTrace}_{D} \oplus \top
\]

where left injection into the output type denotes acceptance
by the parser, and right injection denotes rejection. To
build such a parser, it will suffice to construct a term

\[
  w : \String \vdash \mathsf{parse}_{D} : \LinSigTy q Q {\mathsf{Trace}_{D}(q_{0} , q)}
\]
This is because given a trace of a DFA, we may easily check
if we should accept or reject by simply testing
if the final state is accepting.

Because $w$ is a Kleene star of characters, we may construct
our desired $\mathsf{parse}_{D}$ term as a $\mathsf{foldl}$
over $w$. In the empty case, we just have the trace that
ends at the accepting state. In the recursive case, we
effectively add to our trace by transitioning one character
at a time, as we read them moving across $w$.

Perhaps this derivation is not too surprising. All it says
is that a DFA may be ran by transitioning a single character
at a time, and then accepting or rejecting based on the
final state. This is exactly what DFAs did initially, so
what did we gain? Well, this has the benefit of our type
system to ensure its correctness. Moreover, this construction exports to an
intrinsically verified and executable decision procedure for DFAs in \theoryabbv
embedded in Agda.

We should note that the
construction of a DFA parser requires the addition of an additional but seemingly innocuous
axiom.

\[
  \top \cong \left( \bigoplus_{c : \Sigma} c \right)^{*}
\]

That is, the characters in the alphabet $\Sigma$ really are all of the
characters. The need for this axiom arises when we go to take a transition
within the DFA,
as we must ensure that the next character of the string has a corresponding
transition from the current state
--- which would be guaranteed by determinism provided there are no surprise characters.

\subsection{Regular Expressions and DFAs}
\label{subsec:deriv}

In order to extend the DFA
parser from \cref{subsec:regexparser} to the construction of
a verified parser
generator for regular expression we need to perform some
plumbing establishing an equivalence between regular
expressions and DFAs.

There are several routes we may hope to take in establishing
this equivalence. First, we could prove an equivalence
between NFAs and regular expressions, and separately prove
an equivalence between NFAs and DFAs.
In \cref{subsec:eqproofs}, we include a version of
Thompson's construction --- which established the
equivalence between regular grammars and DFAs. We may additionally
hope to internalize a variant of the powerset construction \cite{rabinFiniteAutomataTheir1959}
--- which takes as input an NFA and constructs a DFA that
recognizes the same language --- and combine the results of
Thompson's construction and the powerset constructions to give an equivalence
between regular expressions and DFAs.  This route is alluring, as it
internalizes several classic grammar-theoretic constructions. However, it may
necessitate extensions to the LNL theory, like
a propositional truncation, and we have not yet investigated
how this would interact with the existing types in the
theory. The addition of a propositional truncation may seem
harmless, but it is not always immediately clear how
distinct constructions will interact. For instance, when
exploring LNL models, Benton discovered that the synthesis
of linear and dependent types require a new presentation of
the $!$ modality from linear logic
\cite{bentonMixedLinearNonlinear1995}. That is all to say,
this is a work in progress and
it is not immediate that the addition of a propositional
truncation is adequate for establishing the weak equivalence
between NFAs and DFAs.

We may instead hope to internalize an equivalence between
regular grammars and DFAs by using Brzozowski derivatives to
directly create a DFA that is weakly equivalent to a given
regular expression, as described by Owens et al.
\cite{owensRegularexpressionDerivativesReexamined2009}.
One characterization of regular grammars is that they are
precisely those grammars which have finitely many inequivalent Brzozowski
derivatives
\cite{brzozowskiDerivativesRegularExpressions1964}.
The algorithm used by Owens takes in a
regular grammar and generates a DFA that recognizes the same
language, and the states in this DFA are the finitely many
derivative equivalence classes. We initially had a version of
this theorem very roughly internalized in the LNL theory.
To our taste, too much of this presentation relied on
meta-arguments that lived outside of
our formalism, and thus this particular phrasing of the
theorem did not translate well into formalization.

In any case, we believe
that revisiting these lines of thought will lead to a
satisfactory internalization of the equivalence between
regular grammars and DFAs, and thus would bridge the gap
between our DFA parser and a full regular expression parser.

\subsection{Equivalence Between Regular Expressions and Finite Automata}
\label{subsec:eqproofs}
In this section, we describe a version of Thompson's
construction \cite{thompsonProgrammingTechniquesRegular1968}
   where we construct an NFA that recognizes a given regular
expression. Moreover, we will show that this NFA is strongly equivalent to the
original grammar. Witnessing this construction in our syntax has two benefits
\begin{enumerate}
  \item It reinforces this high-level view that the syntax is a natural and
        general setting for formal grammar reasoning, as we demonstrate that
        this formal system subsume results from existing systems, and
  \item Following the
        development of Thompson's construction, we then need
        only establish the equivalence of NFAs and DFAs to
        complete the full regular expression parser
\end{enumerate}

\begin{theorem}[Thompson]
  \label{thm:thompson}
  For $g$ a regular grammar $g$, there is an NFA $N$ that recognizes the same
  language as $g$.
\end{theorem}
We make a pretty straightforward adaptation of Thompson's theorem to our setting,

\begin{theorem}[Typed Thompson]
  \label{thm:typthompson}
  For $g$ a regular expression $g$, there is an NFA $N$ such that $g$ is isomorphic
  to $\mathsf{AccTrace}_{N}$.
\end{theorem}

\begin{proof}[Proof Sketch]
  Recall that regular expressions are inductively defined via
  disjunction, concatenation, and Kleene star over literals
  and the empty grammar. By induction over regular expressions,
  we will construct an NFA that is equivalent to $g$.

  First, define the recognizing NFA for the empty grammar
  $I$.

  \begin{figure}[h!]
  \begin{tikzpicture}[node distance = 25mm ]
    \node[state, initial] (1) {$1$};
    \node[state, right of=1, accepting] (2) {$2$};

    \path[->] (1) edge[below] node{$\varepsilon$} (2);
  \end{tikzpicture}
  \caption{$NFA(I)$}
  \label{fig:emptyNFA}
  \end{figure}

  The type of traces from the initial state of $NFA(I)$ to the single
  accepting state is given by,

  \[
    \mathsf{Trace}_{NFA(I)}(q_{1}, q_{2}) = \mu
      \begin{pmatrix}
         g_{q_{1}} := g_{q_{2}} \\
         g_{q_{2}} := I
      \end{pmatrix}. g_{q_{1}}
  \]

  The accepting traces through $NFA(I)$ are then described
  as,

  \[
    \mathsf{AccTrace}_{NFA(I)} = \LinSigTy q {\{1 , 2\}} {\left( \mathsf{Trace}_{NFA(I)}(q_{1} , q) \pair \mathsf{acc}(q) \right)}
  \]

  A quick inspection of \cref{fig:emptyNFA} reveals that the
  only reasonable choice for $q$ is $q_{2}$ --- because
  state 2 is accepting while state 1 is not. Therefore,

  \begin{align*}
    \mathsf{AccTrace}_{NFA(I)}
    & \cong \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2}) \pair \mathsf{acc}(q_{2}) \\
    & \cong \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2}) \pair \top \\
    & \cong \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2})
  \end{align*}

  From here, to prove
  $I \cong \mathsf{AccTrace}_{NFA(I)}$ it suffices to show
  $I \cong \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2})$. Below we
  give two parse transformers, one from $I$ to
  $\mathsf{Trace}_{NFA(I)}(q_{1} , q_{2})$ and vice versa.

  \[
    \inferrule
    {p : I \vdash \mathsf{nil} : \mathsf{Trace}(q_{2} , q_{2}) \\
     \exists \text{~transition~} q_{1} \overset{\varepsilon}{\to} q_{2}
    }
    {p : I \vdash \mathsf{\varepsilon cons}(\mathsf{nil}) : \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2})}
  \]

  Let $\gamma$ be the substitution $\{ g_{q_{2}} / g_{q_{1}}, I / g_{q_{2}} \}$,

  \[
    \inferrule
    {
      p : \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2}) \vdash p :
        \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2}) \\
      x_{1} : \simulsubst {g_{q_{2}}} {\gamma} = I \vdash x_{1} : I \\
      x_{2} : \simulsubst {I} {\gamma} = I \vdash x_{2} : I
    }
    {
      p : \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2}) \vdash \mathsf{mfold}(x_{1}.x_{1} , x_{2}.x_{2})(p) : I
    }
  \]

  This concludes the proof for the case of the empty
  grammar. Let's now walk through the construction for
  literal grammars. Given a character $c$, we construct an
  NFA that recognizes only the string containing the single
  character $c$ as,

  \begin{figure}[h!]
  \begin{tikzpicture}[node distance = 25mm ]
    \node[state, initial] (1) {$1$};
    \node[state, right of=1, accepting] (2) {$2$};

    \path[->] (1) edge[below] node{$c$} (2);
  \end{tikzpicture}
  \caption{$NFA(c)$}
  \label{fig:literalNFA}
  \end{figure}

  The automaton in \cref{fig:literalNFA} induces the
  following type of traces from $q_{1}$ to $q_{2}$.

  \[
    \mathsf{Trace}_{NFA(c)}(q_{1}, q_{2}) = \mu
      \begin{pmatrix}
         g_{q_{1}} := c \otimes g_{q_{2}} \\
         g_{q_{2}} := I
      \end{pmatrix}. g_{q_{1}}
  \]

  Through the same argument as the empty grammar, the only
  state that is accepting is $q_{2}$ and thus,

  \[
    \mathsf{AccTrace}_{NFA(c)} \cong \mathsf{Trace}_{NFA(c)}(q_{1} , q_{2})
  \]

  To show the desired isomorphism of $c \cong NFA(c)$ we
  make a similar argument as we did for the empty grammar
  $I$, except we leverage the $\mathsf{cons}$ rule instead
  of $\mathsf{\varepsilon cons}$. That is, the parse
  transformers in either direction are given as,

  \[
    \inferrule
    {\cdot \vdash \mathsf{nil} : \mathsf{Trace}(q_{2} , q_{2}) \\
     \exists \text{~transition~} q_{1} \overset{c}{\to} q_{2}
    }
    {p : c \vdash \mathsf{cons}(\mathsf{nil}) :
      \mathsf{Trace}_{NFA(c)}(q_{1} , q_{2})}
  \]

  \[
    \inferrule
    {
      p : \mathsf{Trace}_{NFA(c)}(q_{1} , q_{2}) \vdash p :
        \mathsf{Trace}_{NFA(c)}(q_{1} , q_{2}) \\
      x_{1} : \simulsubst {(c \otimes g_{q_{2}})} {\gamma} = c \otimes I \vdash \mathsf{unitR}(x_{1}) : c \\
      x_{2} : \simulsubst {I} {\gamma} = I \vdash x_{2} : I
    }
    {
      p : \mathsf{Trace}_{NFA(c)}(q_{1} , q_{2}) \vdash \mathsf{mfold}(x_{1}.\mathsf{unitR}(x_{1}) , x_{2}.x_{2})(p) : c
    }
  \]

  Where $\gamma$ is the substitution
  $\{ c \otimes g_{q_{2}} / g_{q_{1}}, I / g_{q_{2}} \}$ and
  $\mathsf{unitR}$ is a witness to the isomorphism
  $c \otimes I \cong c$. Again, we may see that these do
  indeed mutually invert each other in the Agda code.

  It remains to show that the desired isomorphisms are
  preserved by $\otimes$, $\oplus$, and Kleene star. Here,
  we will give the argument for the disjunction case, the
  others are defined quite similarly.

  Given two NFAs $N$ and $M$, we may define a new NFA that
  encodes the disjunction of $N$ and $M$. Denote the
  internal states of $N$ by $q_{j}$'s and the internal states
  of $M$ by $r_{k}$'s,

  \begin{figure}[h!]
  \begin{tikzpicture}[node distance = 20mm ]
    \node[state] (2) {$q_{init}$};
    \node[state, initial, below left of=2] (1) {$init$};
    \node[state, below right of=1] (3) {$r_{init}$};
    \node[right of=2] (4) {$\cdots$};
    \node[right of=3] (5) {$\cdots$};
    % \node[state, below right of=1] (3) {$3$};
    % \node[state, below of=2] (4) {$4$};

    \path[->] (1) edge[below] node{$\varepsilon$} (2);
    \path[->] (1) edge[below] node{$\varepsilon$} (3);
    \path[->] (2) edge[below] node{} (4);
    \path[->] (3) edge[below] node{} (5);

    \node[label={[name=l] $N$}, draw,line width=2pt,rounded corners=5pt, fit=(2)(4)] {};
    \node[label={[name=l] $M$}, draw,line width=2pt,rounded corners=5pt, fit=(3)(5)] {};
  \end{tikzpicture}
  \caption{$N \oplus_{NFA} M$}
  \label{fig:disjunctionNFA}
  \end{figure}

  \Cref{fig:disjunctionNFA} shows the process of
  disjunctively combining NFAs. Precisely, we add a single
  new state and we included copies of the states from each of $N$
  and $M$. The new state acts as the initial state and has
  $\varepsilon$-transitions to the initial states of $N$ and
  $M$. We include all of the internal transitions from $N$
  and $M$, and the accepting states of $N \oplus_{NFA} M$
  are exactly the accepting states in each subautomaton.

  Let $g$ and $g'$ be two regular grammars such that
  $g \cong NFA(g)$ and $g' \cong NFA(g')$. As a matter of
  notation\footnote{We shall similarly abuse notation for
    $\otimes$ and Kleene. That is, for a regular grammar
    $g$, when we write $NFA(g)$ we mean the NFA inductively
    built up with the NFA-analogues to the constructors that
    built up $g$.}, we will write $NFA(g \oplus g')$ for
  $NFA(g) \oplus_{NFA} NFA(g')$. The traces of
  $NFA(g \oplus g')$ are then given by,

  \[
    \mathsf{Trace}_{NFA(g \oplus g')}(src , dst) = \mu
      \begin{pmatrix}
        g_{init} := g_{q_{0}} \oplus g_{r_{0}} \\
        g_{q_{j}} := \mathsf{Trans}_{NFA(g)}(q_{j}) \oplus \mathsf{isDst}(q_{j}) \\
        g_{r_{k}} := \mathsf{Trans}_{NFA(g')}(r_{k}) \oplus \mathsf{isDst}(r_{k})
      \end{pmatrix}.g_{src}
    \]

  where $\mathsf{Trans}$ is used to echo the same syntactic
  definitions that appear in the $NFA(g)$ and $NFA(g')$.
  Also, $src$ and $dst$ may take on any value in
  $Q := \{init\} \cup \{q_{j}\} \cup \{r_{k}\}$, and
  $\mathsf{isDst}(q)$ checks if $q$ is equal to $dst$. Which
  is all to say, the traces of $NFA(g \oplus g')$ comprise
  either a trace of $NFA(g)$, or a trace of $NFA(g')$, and
  the transition coming out of $g_{init}$ determines which
  subautomaton we step into.

  The parse transformer from $g \oplus g'$ checks which side
  of the sum type we are on, then takes the appropriate step
  from $g_{init}$ in the automaton.

  \[
    \inferrule
    {
      u : g \vdash \iota (\phi (u)) : \mathsf{AccTrace}_{NFA(g \oplus g')} \\
      v : g' \vdash \iota' (\psi (v)) : \mathsf{AccTrace}_{NFA(g \oplus g')}
    }
    {p : g \oplus g' \vdash \mathsf{case}~p \{ \mathsf{inl}(u) \mapsto s , \mathsf{inr}(v) \mapsto r \} : \mathsf{AccTrace}_{NFA(g \oplus g')}}
  \]

  with $\iota$ and $\iota'$ as embeddings from $NFA(g)$ and
  $NFA(g')$, respectively, into $NFA(g \oplus g')$,
  $\phi: g \cong \mathsf{AccTrace}_{NFA(g)}$, and $\psi: g' \cong \mathsf{AccTrace}_{NFA(g')}$. On a
  high level, all this construction does is turn a parse of
  $g$ into a parse of $NFA(g)$ and then embeds that inside
  of the larger automaton $NFA(g \oplus g')$. Likewise for $g'$.

  In the other direction, recall that the data of an
  accepting trace for $NFA(g \oplus g')$ is a pair of a
  trace and a proof that
  the end state $q'$ of that trace is accepting. By
  multifolding over the first part of that pair, we turn the
  term of type
  $\mathsf{Trace}_{NFA(g \oplus g')}(init , q')$ into a
  trace of either of the subautomata,

  \[
    p : \mathsf{Trace}_{NFA(g \oplus g')}(init , q') \vdash \mathsf{mfold}_{NFA(g \oplus g')} : \mathsf{Trace}_{NFA(g)}(q_{0} , q') \oplus \mathsf{Trace}_{NFA(g')}(r_{0} , q')
  \]

  Additionally, we leverage the fact that the only accepting
  states for $NFA(g \oplus g')$ are those from the
  subautomata to extract that $q'$ must be an accepting
  state from a subautomaton.

  \[
    x : \mathsf{acc}_{NFA(g \oplus g')}(q') \vdash M : \mathsf{acc}_{NFA(g)}(q') \oplus \mathsf{acc}_{NFA(g')}(q')
  \]

  We then combine the trace and proof of acceptance into an
  accepting trace of one of the subautomata,

  \[
    p : \mathsf{AccTrace}_{NFA(g \oplus g')} \vdash N : \mathsf{AccTrace}_{NFA(g)} \oplus \mathsf{AccTrace}_{NFA(g')}
  \]

  Lastly, we then inductively use the isomorphisms $\phi$
  and $\psi$ to turn the accepting traces into a parse of
  $g$ or $g'$,

  \[
     N : \mathsf{AccTrace}_{NFA(g)} \oplus \mathsf{AccTrace}_{NFA(g')} \vdash \mathsf{case}~N~\{\mathsf{inl}(n) \mapsto \phi^{-1}(n), \mathsf{inr}(n') \mapsto \psi^{-1}(n')\} : g \oplus g'
  \]

  When you wish to show that these actually maps invert each other to form an isomorphism\max{jumbled sentence}, we invoke the
  universal properties of the $\mathsf{Trace}$ types. That is,
  $\mathsf{Trace}_{\oplus NFA}$, $\mathsf{Trace}_{N}$, and $\mathsf{Trace}_{M}$
  are \emph{initial} algebras to the the functors that define them. Thus, the
  embedding of traces through $N$ into $\oplus NFA$ and then mapped back down to $N$
  compose to a map from $\mathsf{Trace}_{N}$ to itself. Such a map is unique and
  thus must be the identity. The same reasoning shows that the desired
  composite do indeed invert each other to form an isomorphism.

  As discussed above, the other cases and
  follow through a similar argument. This
  concludes the proof of our variant of Thompson's construction.
\end{proof}

\subsection{Other Automata}
\subsubsection{Pushdown Automata}
\label{subsubsec:pda}
A (nondeterministic) \emph{pushdown automaton} is an automaton that employs a
stack. Just like NFAs, they have transitions labeled with characters from a
fixed string alphabet $\Sigma$. Additionally, they maintain a stack of
characters drawn from a stack alphabet $\Gamma$. They are often represented
formally as a 7-tuple $(Q, \Sigma, \Gamma, \delta, q_{0}, Z, F)$,

\begin{itemize}
  \item $Q$ a finite set of states
  \item $\Sigma$ a fixed, finite string alphabet
  \item $S$ a fixed, finite stack alphabet
  \item
        $\delta \subset Q \times (\Sigma \cup \{ \varepsilon \}) \times S \to \mathcal{P}(Q \times S^{*})$
        the transition function
  \item $q_{0} \in Q$ the start state
  \item $Z \in S$ the initial stack symbol
  \item $F \subset Q$ the accepting states
\end{itemize}

We encode the traces of a pushdown automaton very similarly to those of an NFA,
except the transitions of a PDA are instead encoded via the LNL product type
$\bigamp$. This is because when simply transitioning via character, a PDA must
also pop and push characters onto a stack, which is used as the argument to
these dependent functions. Thus, the appropriate formation of traces through a
PDA is dependent on a stack.

Let $S$ be a non-linear type encoding the stack
alphabet, and build lists over $S$ as the (non-linear) least fixed-point
$\mathsf{List}(S) := \mu X . 1 + S \times X$. Then, the type of states for a
PDA $P$ with stack alphabet $S$ are given as a functions that takes in lists $L$,
and then makes a case distinction between possible transitions based on what was witnessed as $\mathsf{head}(L)$. The choice of transition
will then determine which character to transition by and what word $w$ should be
pushed onto the stack. The word that is added to the top of the stack is
appended to $\mathsf{tail}(L)$ and then we recursively step into another state
called on argument $w \cdot \mathsf{tail}(L)$.

% \begin{gather*}
%   \mathsf{State} \in {g _{q} : q \in Q} \\
%   \mathsf{Word} \in \String \\
%   \mathsf{Char} \in \Sigma   \\
%   \mathsf{StackChar} \in \Gamma \\
% \end{gather*}
% \begin{align*}
%   \mathsf{Trans}_{P}(q) ::=~&
%                            \LinPiTy {(hd :: tl)} {\mathsf{List}(S)} {\left( \mathsf{Char} \otimes \mathsf{State}(\mathsf{Word} + tl) \right)}~| \\
%   & \LinPiTy {(hd :: tl)} {\mathsf{List}(S)} {\left(  \mathsf{State}(\mathsf{Word} + tl) \right)}~|~\mathsf{Trans}_{P}(q) \oplus \mathsf{Trans}_{P}(q)
% \end{align*}

% That is to say, when transitioning a PDA pops off the head $hd$ of the stack

\subsubsection{Turing Machines}
\label{subsubsec:tm}

Above we gave a grammar presentation of
traces through a PDA by using a non-linear type $S$
to encode the stack. We may similarly use pairs $S \times S$
to encode the tape of a Turing machine. With two stacks we can simulate the behavior of the
infinite tape of a Turing machine. The intuition behind this correspondence is
that the left half of the tape is on one stack, the right
half of the tape the other, and we treat the tops of stacks
like the head of the tape.

\pedro{I like the overall structure of this section! Showing how the type theory
can encode and reason about different kinds of languages is very instructive!}

\section{Categorical Semantics of \theoryabbv}
\label{sec:categorify}

As described in previous sections, \theoryabbv has a ``standard''
interpretation where non-linear types denote sets, linear types denote
formal grammars, non-linear terms denote functions and linear terms
denote parse transformers.
%
Further, as is typical for type theory, every type constructor in the
calculus is interpreted by using a \emph{universal construction} in
either the category of sets or the category of grammars\footnote{the
only exception is that the axioms for Brzozowski derivatives we
considered do not follow solely from universal properties. In this
section we focus on models that do not necessarily satisfy these
axioms.}.
%
This leads to an immediate opportunity for generalization: the type
theory has in addition to the standard grammar-theoretic
interpretation, an interpretation in any categorical structure that
exhibits the same universal constructions.
%
On the one hand this gives us a structured way to formalize the
standard semantics precisely, but additionally it enables us to
consider \emph{non-standard} models in the next section that point to
further applications as well as meta-theoretic results.

In developing this categorical semantics, we will start with the
closest analogue from formal language theory: Kleene algebra.  Kleene
algebras are an important tool in the theory of regular languages
serving as a bridge between algebraic reasoning and equivalence of
regular expressions. More broadly, through various extensions, they
serve as a theoretical substrate to studying different kinds of formal
languages.
%
We can then see that the ``regular fragment'' of our type theory
(i.e., just characters, $\otimes$, $\oplus$, $0,1$ and a Kleene star)
has models in what we call a \emph{Kleene category} a categorification
of Kleene algebra from posets to categories.
%
We then further develop this into our final notion of model, which we
call a \emph{Chomsky category} as it can model not just regular
languages, but the full Chomsky hierarchy.

\subsection{Kleene Algebra and Kleene Categories}
A Kleene algebra is a tuple $(A, +, \cdot, (-)^*, 1, 0)$, where $A$ is
a set, $+$ and $\cdot$ are binary operations over $A$, $(-)^*$ is a
function over $A$, and $1$ and $0$ are constants. These structures
satisfy the axioms depicted in Figure~\ref{fig:axioms}.

\begin{figure}
  \begin{align*}
    x + (y + z) &= (x + y) + z & x + y &= y + x\\
    x + 0 &= x & x + x &= x\\
    x(yz) &= (xy)z & x1 &= 1x = x\\
    x(y + z) &= xy + xz & (x + y)z &= xz + yz\\
    x0 &= 0x = x & & \\
    1 + aa^* &\leq a^* & 1 + a^*a &\leq a^*\\
     b + ax \leq x &\implies a^*b \leq x &  b + xa \leq x &\implies ba^* \leq x
  \end{align*}
  \caption{Kleene algebra axioms}
  \label{fig:axioms}
\end{figure}

The addition operation can be used to define the partial order
structure $a \leq b$ if, and only if, $a + b = b$. In the theory of
formal languages, this order structure can be used to model language
containment. In this section, we categorify the concept of Kleene
algebra and build on top of it in order to define an abstract theory
of parsing. We start by defining \emph{Kleene categories}.

\begin{definition}
  A Kleene category is a distributive monoidal category $\cat{K}$
  such that for every objects $A$ and $B$, the endofunctors $F_{A, B}
  = B + A \otimes X$ and $G_{A, B} = B + X \otimes A$ have initial
  algebras (denoted $\mu X.\, F_{A, B}(X)$) such that $B \otimes (\mu
  X.\, F_{A, 1}) \cong \mu X.\, F_{A, B}(X)$ and the analogous isomorphism
  for $G_{A,B}$ also holds.
\end{definition}

As a sanity check, note that Kleene algebras are indeed examples of
Kleene categories.

\begin{example}
  Every Kleene algebra, seen a posetal category, is a Kleene category.
  The product $\cdot$ is a monoidal product and the addition is a
  least-upper bound, which corresponds to a coproduct. Lastly, the
  axioms of the Kleene star have a direct correspondence to the
  coherence conditions postulated by the initial algebras of Kleene
  categories.
\end{example}

This example provides a neat categorical justification to how
restrictive Kleene algebras are in terms of reasoning about
languages. By only having at most one morphism between objects, there
is not a lot of information they can convey. In this case, the only
information you get is language containment. As demonstrated by
$\mathbf{Gr}$, the extra degrees of freedom granted by having more
morphisms gives you more algebraic structure for reasoning about
languages.

For the next example, we see an unexpected connection with the theory
of substructural logics.

\begin{example}
  The opposite category of every Kleene category is a model of a variant of
  conjunctive ordered logic, where the Kleene star plays the role of the ``of
  course'' modality from substructural logics which allows hypotheses to
  be discarded or duplicated.
\end{example}

As we have seen, the proposed axioms are a direct translation of the
Kleene algebra axioms to a categorical setting. Its most unusual aspect is the
axiomatization of the Kleene star as a family of initial algebras
satisfying certain isomorphisms. If the Kleene category $\cat{K}$ has
more structure, then these isomorphisms hold ``for free''.

\begin{theorem}
  \label{th:kleeneclosed}
  Let $\cat{K}$ be a Kleene category such that it is also monoidal
  closed.  Then, the initial algebras isomorphisms hold automatically.
\end{theorem}
\begin{proof}
  We prove this by the unicity (up-to isomorphism) of initial
  algebras. Let $[hd, tl]: I + (\mu X.\, F_{A, I}(X)) \otimes A \to
  (\mu X.\, F_{A, I}(X))$ be the initial algebra structure of $(\mu
  X.\, F_{A, I}(X))$ and consider the map $[hd, tl] : B + B \otimes
  (\mu X.\, F_{A, I}(X)) \otimes A \to B\otimes (\mu X.\, F_{A,
    I}(X))$.

  Now, let $[f,g] : B + A \otimes Y \to Y$ be an $F_{A,B}$-algebra and
  we want to show that there is a unique algebra morphism $h : \mu X.\, F_{A,I} \to B \lto Y$. We can show existence and
  uniqueness by showing that the diagram on top commutes if, and
  only if, the diagram on the bottom commutes:

% https://q.uiver.app/#q=WzAsOCxbMCwwLCJCICsgQiBcXG90aW1lcyAoXFxtdSBYLlxcLCAxICsgWCBcXG90aW1lcyBBKSJdLFsyLDAsIkIgKyBZIFxcb3RpbWVzIEEiXSxbMiwyLCJZIl0sWzAsMiwiQiBcXG90aW1lcyAoXFxtdSBYLlxcLCAxICsgWCBcXG90aW1lcyBBKSJdLFswLDMsIjEgKyAoXFxtdSBYLlxcLCAxICsgWCBcXG90aW1lcyBBKSJdLFswLDUsIlxcbXUgWC5cXCwgMSArIFggXFxvdGltZXMgQSJdLFsyLDMsIjEgKyAoIEIgXFxsdG8gWSkgXFxvdGltZXMgQSJdLFsyLDUsIkIgXFxsdG8gWSJdLFswLDEsImlkICsgKGlkIFxcb3RpbWVzIGg7IGV2KSBcXG90aW1lcyBpZF9BIl0sWzEsMiwiW2YsZ10iXSxbMywyLCJpZCBcXG90aW1lcyBoOyBldiIsMl0sWzAsMywiW2lkIFxcb3RpbWVzIGgsIGlkIFxcb3RpbWVzIHRsXSIsMl0sWzQsNiwiaWQgKyAoaCBcXG90aW1lcyBYKVxcb3RpbWVzIGlkIl0sWzUsNywiaCIsMl0sWzYsNywiW2YnLCBnJ10iXSxbNCw1LCJbaGQsIHRsXSIsMl1d
\[\begin{tikzcd}
	{B + B \otimes (\mu X.\, I + X \otimes A)} && {B + Y \otimes A} \\
	\\
	{B \otimes (\mu X.\, I + X \otimes A)} && Y \\
	{I + (\mu X.\, I + X \otimes A)} && {I + ( B \lto Y) \otimes A} \\
	\\
	{\mu X.\, I + X \otimes A} && {B \lto Y}
	\arrow["{id + (id \otimes h; ev) \otimes id_A}", from=1-1, to=1-3]
	\arrow["{[f,g]}", from=1-3, to=3-3]
	\arrow["{id \otimes h; ev}"', from=3-1, to=3-3]
	\arrow["{[id \otimes h, id \otimes tl]}"', from=1-1, to=3-1]
	\arrow["{id + (h \otimes X)\otimes id}", from=4-1, to=4-3]
	\arrow["h"', from=6-1, to=6-3]
	\arrow["{[f', g']}", from=4-3, to=6-3]
	\arrow["{[hd, tl]}"', from=4-1, to=6-1]
\end{tikzcd}\]
  This equivalence follows by using the adjunction structure given
  by the monoidal closed structure of $\cat{K}$. A completely analogous
  argument for $G_{A,B}$ also holds. Furthermore, by generalizing the
  construction of \Cref{sec:formaltype}, we can also show that from
  the monoidal closed assumption it follows that $\mu X.\, F_{A, I}(X) \cong \mu X.\, G_{A, I}(X)$
\end{proof}

Something surprising about this lemma is that it provides an alternative
perspective on the observation that if a Kleene algebra has an
residuation operation, also called action algebra \cite{kozen1994action},
then the Kleene star admits a simpler axiomatization.

Since we want Kleene categories to generalize our notion of formal
grammars as presheaves $\String \to \Set$, we prove that they do
indeed form a Kleene category. We start by presenting a well-known
construction from presheaf categories.

\begin{definition}
  Let $\cat{C}$ be a locally small monoidal category and $F$, $G$ be
  two functors $\cat{C} \to \Set$. Their Day convolution tensor
  product is defined as the following coend formula:
  \[
  (F \otimes_{Day} G)(x) = \int^{(y,z) \in \cat{C}\times\cat{C}}\cat{C}(y\otimes z, x) \times F(y) \times G(z)
  \]
  Dually, its internal hom is given by the following end formula:
  \[
  (F \lto_{Day} G)(x) = \int_{y} \Set(F(y), G(x \otimes y))
  \]
\end{definition}

\begin{lemma}[Day \cite{day1970construction}]
  Under the assumptions above, the presheaf category $\Set^{\cat{C}}$ is
  monoidal closed.
\end{lemma}

%% \begin{theorem}
%%   Let $\cat{K}$ be a Kleene category and $A$ a discrete category.
%%   The functor category $[A, \cat{K}]$.
%%   (HOW GENERAL SHOULD THIS THEOREM BE? BY ASSUMING ENOUGH STRUCTURE,
%%   E.G. K = Set, THIS THEOREM BECOMES SIMPLE TO PROVE)
%% \end{theorem}
\begin{theorem}
  If $\cat{C}$ is a locally small monoidal category, then
  $\Set^{\cat{C}}$ is a Kleene category.
\end{theorem}
\begin{proof}

  By the lemma above, $\Set^{\cat{C}}$ is monoidal closed, and since it
  is a presheaf category, it has coproducts. Furthermore, the tensor
  is a left adjoint, i.e. it preserves colimits and, therefore, it is
  a distributive category.

  As for the Kleene star, since presheaf categories admit small colimits,
  the initial algebra of the functors $F_{A,B}$ and $G_{A,B}$ can be
  defined as the filtered colimit of the diagrams:

  From Theorem~\ref{th:kleeneclosed} it follows that these initial
  algebras satisfy the required isomorphisms and this concludes the
  proof.
\end{proof}

\begin{corollary}
  For every alphabet $\Sigma$, the presheaf category $\Set^{\cat{\Sigma^*}}$
  is a Kleene category.
\end{corollary}
\begin{proof}
  Note that string concatenation and the empty string make the
  discrete category $\Sigma^*$ a strict monoidal category.
\end{proof}

Much like in the posetal case, the abstract structure of a Kleene
category is expressive enough to synthetically reason about regular
languages. A significant difference between them is that while Kleene
algebras can reason about language containment, Kleene categories can
reason about \emph{ambiguity}, \emph{strong equivalence} of grammars.

\subsection{Chomsky Categories}

Though Kleene categories are expressive enough to reason about
concepts that are outside of reach of Kleene algebras, their
simply-typed nature makes them not so expressive from a type theoretic
point of view. This is limiting because type theories are successful
syntactic frameworks for manipulating complicated categorical
structures while avoiding some issues common in category theory, such
as coherence issues.

With this in mind, we want to design a categorical semantics that
builds on top of Kleene categories with the goal of extending them
with dependent types and making them capable reasoning about languages
and their parsers. This leads us to the abstract notion of model we
are interested in capturing with \theoryabbv: a \emph{Chomsky
category}.

\pedro{We should probably define some of the words in this definition}
\begin{definition}
  A Chomsky category is a locally Cartesian category with two hierarchies of
  universes $\{L_i\}_{i\in \nat}$ and $\{U_i\}_{i\in \nat}$ such that
  every $L_i$ and $U_i$ are $U_{i+1}$-small. Furthermore, we require
  $U_i$ to be closed under dependent products and sum,
  $L_i$ to be closed under the Kleene category connectives,
  dependent products, left and right closed structures, with
  a type constructor $G : L_i \to U_i$ and a linear dependent sum
  going the other direction.
\end{definition}

\steven{Max suggests augmenting the definition of a Chomsky category to
  something like two categories $L$ and $U$, and $U$ is Cartesian closed I don't
  fully recall the rest.

  My best guess is that you take $L$ a Kleene category with a hierarchy of universes
  two, and you further require that $L$ is $Psh(U)$-enriched, except he
  suggested a further adjective on these presheaves. Perhaps representability?
}

\begin{theorem}
  The presheaf category $\Grammar = \Set^{\cat{\Sigma^*}}$ is a Chomsky category.
\end{theorem}

Further, the syntactic category of \theoryname is manifestly a Chomsky category.

\pedro{This is likely true, but if we explicitly say so, this warrants a proof. I think that
if we don't say anything about the syntactic category, reviewers won't mind.}

\steven{I agree that we don't want to say anything that opens unnecessary
  questions for proofs we haven't written. However, it seems hard to make the
  case that we have the right categorical model of the syntax if this statement
  isn't true. By restating our definition of Chomsky category, this should be
  obvious or a quick proof}

\pedro{I agree, then we should add the quick proof :) }

\section{Concrete Models of \theoryabbv}
\label{sec:othermodels}

One of the powers of type theories is that it can be profitable to interpret them
in various models. In this section, by using our just-defined Chomsky categories,
we show how other useful concepts from formal language theory can also be organized
as models of \theoryabbv. We illustrate this point by providing two examples that
are closely related to the theory of formal languages: language equivalence and
semantic actions. Furthermore, in order to justify how $\mathbf{Gr}$ relates to
more traditional notions of parsing, we define a glued model that proves a
canonicity property of grammar terms.

\subsection{Language Semantics}
Every grammar induces a language semantics. Also languages can be taken as a
propositionally truncated view of the syntax. Logical equivalence should induce
weak equivalence, and thus even give a syntactic way to reason about language equivalence.
\steven{TODO language semantics}

\subsection{Semantic Actions}
Returning to the problem of parsing, the output of a parse usually is not the
literal parse tree. Rather, the output is the result of some \emph{semantic
  actions} ran on the parse tree, which usually serve to remove some syntactic
details that are unnecessary for later processing.

Given a grammar $G : \String \to \Set$, we define a semantic action to be a set
$X$ with a function $f$ that produces a semantic element from any parse of $G$.

\[
  f : \PiTy w \String {G w \to X}
\]

Further, semantic actions can be arranged into a structured category.
Define $\SemAct$ as the comma
category $\Grammar / \Delta$, where $\Delta : \Set \to \Grammar$ defines a
discrete presheaf. That is, for a set $X$, $\Delta (X)(w) = X$ for all
$w \in \String$. As $\SemAct$ is defined as a comma category, it has a forgetful
functor into $\Grammar$. That is, $\SemAct$ serves as a notion of formal
grammar. Moreover, $\SemAct$ is a model of \theoryabbv.

\steven{It being a notion of formal grammar is distinct from being a model. This
probably warrants a proof}

\steven{TODO semantic actions}

\pedro{This is a very nice opportunity of showing off the supremacy of denotational
  reasoning ;) We should probably prove the gluing lemma in the previous section
  and apply it here and in the canonicity section. The actual proof might have to be
  moved to the appendix, though}

\subsection{Parse Canonicity}
Canonicity is an important metatheoretic theorem in the type theory
literature.  It provides insight on the normal forms of terms and,
therefore, on its computational aspects. Frequently, proving
canonicity for boolean types, i.e. every closed term of type bool
reduces to either true or false, is enough to justify that the type
theory being studied is well-behaved. In our case, however, since we
want to connect \theoryabbv to parsers, we must provide a more
detailed account of canonicity. In particular, we give a nonstandard semantics
of \theoryabbv that carries a proof of canonicity along with it.

If $\cdot \vdash A$ is a closed linear type then there are
two obvious notions of what constitutes a ``parse'' of a string w
according to the grammar $A$:
\begin{enumerate}
\item On the one hand we have the set-theoretic semantics just
  defined, $\llbracket A \rrbracket \cdot w$
\item On the other hand, we can view the string $w = c_1c_2\cdots$ as
  a linear context $\lceil w \rceil = x_1:c_1,x_2:c_2,\ldots$ and
  define a parse to be a $\beta\eta$-equivalence class of linear terms $\cdot;
  \lceil w \rceil \vdash e : G$.
\end{enumerate}
It is not difficult to see that at least for the ``purely positive''
formulae (those featuring only the positive connectives
$0,+,I,\otimes,\mu, \overline\Sigma,c$) that
every element $t \in \llbracket A \rrbracket w$ is a kind of tree and
that the nodes of the tree correspond precisely to the introduction
forms of the type. However it is far less obvious that \emph{every}
linear term $\lceil w \rceil \vdash p : \phi$ is equal to some
sequence of introduction forms since proofs can include elimination
forms as well. To show that this is indeed the case we give a
\emph{canonicity} result for the calculus: that the parses for .

\begin{definition}
  A non-linear type $X$ is purely positive if it is built up using
  only finite sums, finite products and least fixed points.

  A linear type is purely positive if it is built up using only finite
  sums, tensor products, generators $c$, least fixed points and linear
  sigma types over purely positive non-linear types.
\end{definition}

\begin{definition}
  %% Let $X$ be a closed non-linear type. The closed elements $\textrm{Cl}(X)$ of $X$ are the definitional equivalence classes of terms $\cdot \vdash e : X$.

  Let $A$ be a closed linear type. The nerve $N(A)$ is a presheaf on
  strings that takes a string $w$ to the definitional equivalence
  classes of terms $\cdot; \lceil w\rceil \vdash e: N(A)$.
\end{definition}

\begin{theorem}[Canonicity]
  Let $A$ be a closed, purely positive linear type. Then there is an
  isomorphism between $\llbracket A\rrbracket$ and $N(A)$.
\end{theorem}
\begin{proof}
  We outline the proof here, more details are in the appendix. The
  proof proceeds first by a standard logical families construction
  that combines canonicity arguments for dependent type theory
  TODO cite coquand
  % \cite{coquand,etc}
  with logical relations constructions for linear
  types
  TODO cite hylandschalk
  % \cite{hylandschalk}
  . It is easy to see by induction that the
  logical family for $A$, $\hat A$ is isomorphic to $\llbracket A
  \rrbracket$ and the fundamental lemma proves that the projection
  morphism $p : \hat A \to N(A)$ has a section, the canonicalization
  procedure. Then we establish again by induction that
  canonicalization is also a retraction by observing that introduction
  forms are taken to constructors.
\end{proof}

\begin{enumerate}
\item Every term $\lceil w \rceil \vdash p : G + H$ is equal to $\sigma_1q$ or $\sigma_2 r$ (but not both)
\item There are no terms $\lceil w \rceil \vdash p : 0$
\item If there is a term $\lceil w \rceil \vdash p : c$ then $w = c$ and $p = x$.
\item Every term $\lceil w \rceil \vdash p : G \otimes H$ is equal to $(q,r)$ for some $q,r$
\item Every term $\lceil w \rceil \vdash p : \epsilon$ is equal to $()$
\item Every term $\lceil w \rceil \vdash p : c$ is equal to $x:c$
\item Every term $\lceil w \rceil \vdash p : \mu X. G$ is equal to $\textrm{roll}(q)$ where $q : G(\mu X.G/X)$
\item Every term $\lceil w \rceil \vdash p : (x:A) \times G$ is equal
  to $(M,q)$ where $\cdot \vdash M : A$
\end{enumerate}

To prove this result we will use a logical families model. We give a
brief overview of this model concretely:
\begin{enumerate}
\item A context $\Gamma$ denotes a family of sets indexed by closing substitutions $\hat\Gamma : (\cdot \vdash \Gamma) \Rightarrow \Set_i$
\item A type $\Gamma \vdash X : U_i$ denotes a family of sets $\hat X : \Pi(\gamma:\cdot \vdash \Gamma) \hat\Gamma \Rightarrow (\cdot \vdash \simulsubst X \gamma) \Rightarrow \Set_i$
\item A term $\Gamma \vdash e : X$ denotes a section $\hat e : \Pi(\gamma)\Pi(\hat\gamma)\hat X \gamma \hat\gamma (\simulsubst e \gamma)$
\item A linear type $\Gamma \vdash A : L_i$ denotes a family of grammars $\hat A : \Pi(\gamma:\cdot\vdash\Gamma)\,\hat\Gamma \Rightarrow \Pi(w:\Sigma^*) (\cdot;\lceil w\rceil \vdash A[\gamma])\Rightarrow \Set_i$, and the denotation of a linear context $\Delta$ is similar.
\item A linear term $\Gamma;\Delta \vdash e : A$ denotes a function \[\hat e : \Pi(\gamma)\Pi(\hat\gamma)\Pi(w)\Pi(\delta : \lceil w \rceil \vdash \simulsubst \Delta \gamma) \hat\Delta \gamma \hat\gamma \delta \Rightarrow \hat A \gamma \hat\gamma w {(\simulsubst {\simulsubst e \gamma} \delta)}\]
\end{enumerate}
And some of the constructions:
\begin{enumerate}
\item $\widehat {(G A)} \gamma \hat\gamma e = \hat A \gamma \hat\gamma \varepsilon (G^{-1}e)$
\item $\widehat {(A \otimes B)} \gamma \hat\gamma w e = \Sigma(w_Aw_B = w)\Sigma(e_A)\Sigma(e_B) (e_A,e_B) = e \wedge \hat A \gamma \hat \gamma w_A e_A \times \hat B \gamma \hat \gamma w_B e_B$
\item $\widehat {(A \lto B)} \gamma \hat\gamma w e = \Pi(w_A)\Pi(e_A) \hat A \gamma\hat\gamma w_A e_A \Rightarrow \hat B \gamma\hat\gamma (ww_A) (\applto {e_A} e)$
\end{enumerate}

First, the category with families will be
the category of logical families over set contexts/types
$\Delta$/$A$. Then the propositional portion will be defined by
mapping a logical family $\hat \Gamma \to \Gamma$

First, let $L$ be the category of BI formulae and proofs (quotiented
by $\beta\eta$ equality). Define a functor $N : L \to \Set^{\Sigma^*}$ by
\[ N(\phi)(w) = L(w,\phi) \]

Then define the gluing category $\mathcal G$ as the comma category
$\Set^{\Sigma^*}/N$. That is, an object of this category is a pair of
a formula $\phi \in L$ and an object $S \in \mathcal
P(\Sigma^*)/N(\phi)$. We can then use the equivalence $\mathcal
P(\Sigma^*)/N(\phi) \cong \mathcal P(\int N(\phi))$ to get a simple
description of such an $S$: it is simply a family of sets indexed by
proofs $L(w,\phi)$:
\[ \prod_{w\in\Sigma^*} L(w,\phi) \to \Set \]
This category clearly comes with a projection functor $\pi : \mathcal
G \to \mathcal L$ and then our goal is to define a section by using
the universal property of $\mathcal L$.

To this end we define
\begin{enumerate}
\item $(\phi, S) \otimes (\psi, T) = (\phi \otimes \psi, S\otimes T)$ where
  \[ (S \otimes T)(w, p) = (w_1w_2 = w) \times (q_1,q_2 = p) \times S\,w_1\,q_1 \times T\,w_2\,q_2\]
\item $(\phi, S) \multimap (\psi, T) = (\phi \multimap \psi, S \multimap T)$ where
  \[ (S \multimap T)(w,p) = w' \to q \to S\,w'\,q \to T (ww') (p\,q) \]
\item $\mu X. ??$ ??
\end{enumerate}

\pedro{We should conclude this section by explaining the relevance of the canonicity
theorem. Could also be done before stating the theorem.}

\section{Discussion and Future Work}
\label{sec:discussion}

The core objects of study in the theory of formal grammars have stood stalwart
for decades. Study of interesting extensions or restrictions of particular
grammar formalisms has remained consistent line of research over the years.
Occasionally, researchers will either explicitly create a type system for their
formalism, or researchers will leverage the same underlying structure of their
systems. In this paper, we have not
done much \emph{per se}, other than gather these ideas, abstract over their
commonalities, and give the common language between formalisms.

\paragraph{Typed Approaches to Grammars}
The usage of a simple type system to reason about regular expressions was
introduced by Frisch and Cardelli in 2004 \cite{frischCardelli}, and later expanded
by Hengelin and Nielsen in 2011 \cite{henglein_regular_2011} to handle proofs of regular expression
containment.

In 1992, while investigating the addition of arbitrary recursion to Kleene
algebra \cite{leiss},
Lei{\ss} used a least fixed-point operator instead of the Kleene star. Although
he did not explicitly make use of a type system, or have access to the view of Frisch and
Cardelli, Lei{\ss} had added the full power of inductive types to a type system of regular expressions.

Pratt's work on the residuation of action algebras in \cite{prattActionLogicPure1991} closely mirrors
our observation that Brzozowski derivatives mirror linear function types.
Moreover, he shows that his residuals form a Galois connection with sequential
composition. This is precisely a posetal --- or \emph{thin} --- version of the
adjunction between
linear function types and tensor product in \theoryabbv.

In \cite{firsovCertifiedNormalizationContextFree2015}, Firsov and Uustalu
constructed a verified normalization procedure to turn a
context-free grammar into an equivalent grammar in Chomsky normal form.
Combining their work with a CYK parser, they have built a verified context-free
grammar parser, \emph{up to weak equivalence}. To upgrade from a weak
equivalence to a strong equivalence, they suggest updating their grammar
formalism to treat parse trees as first-class proofs of language membership.
That is, without explicitly stating so, they suggest that they next natural step
is to reason about grammars in a system where parse trees are terms, as we have
given in this paper.

\steven{Make the above paragraph shorter. Also idk if its worth including, but I
like that they practically ask for a type theory}

\steven{Mention Neel's typed algebraic approach paper here. Say it is similar at
  first glance, as its parsing with types, but that he actually is representing
  something quite different. Or is it so different that its not worth mentioning?}

\paragraph{Kleene Algebra}
Since the early works in the theory of formal languages, Kleene
algebras have played an important role in its development. They
generalize the operations known from regular languages by introducing
operations generalizing language composition, language union and the
Kleene star.  More generally, they are defined as inequational theory
where the inequality is meant to capture language containment. This
theory is extremely successful, having found applications in algebraic
path problems, theory of programming languages, compiler optimizations
and more.

A frequently fruitful research direction is exploring varying
extensions of Kleene algebras, Kleene algebra with tests (KAT) being
one of the most notable ones. Our approach is radically different from
most extensions, which usually aim at modifying or adding new
operations to Kleene algebras, but still keeping it as an inequational
theory. By adopting a category-theoretic treatment and allowing the
``order structure'' to encode more information than merely
inequalities, we were able to extend Kleene algebra to reason about
parsing as well.

\steven{Can we sketch here what an apt categorification of KATs might be? As a
  poor first pass: perhaps a Kleene category with a full involutive
  subcategory that corresponds to the Boolean algebra?.}
\steven{What can we say about internalizing Kleene algebra proofs or extensions?}
\steven{any KA citations needed above?}

\paragraph{Linearity with Dependent Types}
\steven{I don't know if it worth discussing the different influences of
  categorical model here? We take Neel's model and add to it as necessary}

Vakar might be worth citing to discuss different characterizations of the
notion of model \cite{vakarSyntaxSemanticsLinear2015}

\cite{fu2023twolevellineardependenttype} give a compiler for a two-level
linear dependent type theory. This could serve as a template for a
compiler implementation of \theoryname that is not just inference rules
embedded in Agda.

\paragraph{Implementations in Parsing}
Talk about how we have a DFA parser in Agda and want to to get out a regex
parser. Then context free, etc

future work related to building a usable correct-by-construction parser
combinator/semantic action combinator library in this language.

Contrast this
potential library with existing verified parsers. Citations for some in prelim presentation

\paragraph{Applications to Type Checking}
While parsing typically refers to the generation of semantic
objects from string input, many tasks in programming can be
viewed as parsing of objects with more structure, such as
trees with binding structure or graphs. Fundamental to the
frontend of many
programming language implementations are type systems. In
particular, \emph{type checking}
--- analogous to language recognition --- and \emph{typed
  elaboration} --- analogous to parsing --- arise when
producing a semantic object subject to some analysis. Just
as our string grammars were given as functors from $\String$
to $\Set$, we envision adapting the same philosophy
to functors from $\String$ to some \emph{category of trees} to craft a syntax
that natively captures typed elaboration. This suggests an
unusual sort of bunched type theory, where context extension
no longer resembles concatenation of strings but instead
takes on the form of tree constructors.

\steven{condense the future work section?}
\steven{Put the 2 category stuff here?}

\paragraph{Other Future Work}
Backtracking parsing: via (delimited) continuations? effect handlers?

\bibliographystyle{plain}
\bibliography{refs.bib}

\end{document}

### 2024-10-03 paper/paper.tex
some rules for strictly positive functors

@@
 
+\newcommand{\SPF}{\mathsf{SPF}}
+\newcommand{\Var}{\mathsf{Var}}
+\newcommand{\K}{\mathsf{K}}
+\newcommand{\map}{\mathsf{map}}
 \newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
@@
 \newcommand{\Set}{\mathbf{Set}}
+\newcommand{\Syn}{\mathbf{Synx}}
 \newcommand{\SemAct}{\mathbf{SemAct}}
@@
 \newcommand{\nat}{\mathbb{N}}
-\newcommand{\theoryname}{Grammar Type Theory\xspace}
-\newcommand{\theoryabbv}{GramTT\xspace}
+\newcommand{\theoryname}{Dependent Lambek Calculus\xspace}
+\newcommand{\theoryabbv}{$\textrm{Lambek}^D$}
 
@@
 \DeclareMathOperator*{\bigamp}{\scalerel*{\&}{\bigoplus}}
+\DeclareMathOperator*{\bigwith}{\scalerel*{\&}{\bigoplus}}
+
+\newcommand{\oplusinj}[2]{\sigma\,#1\,#2}
+\newcommand{\withprj}[2]{\pi\,#1\,#2}
 
@@
 \newcommand{\subst}[3]{\simulsubst {#1} {#2/#3}}
+\newcommand{\el}{\mathsf{el}}
 \newcommand{\letin}[3]{\mathsf{let}\, #1 = #2 \, \mathsf{in}\, #3}
@@
 \newcommand{\lambtol}[2]{\lambda^{{\tol}} #1.\, #2}
-\newcommand{\dlamb}[2]{\overline{\lambda} #1.\, #2}
+\newcommand{\dlamb}[2]{{\lambda}^{{\&}} #1.\, #2}
+\newcommand{\withlamb}[2]{{\lambda}^{{\&}} #1.\, #2}
 \newcommand{\app}[2]{#1 \, #2}
@@
 \newcommand{\LinSigTy}[3]{\textstyle\bigoplus (#1 : #2). #3}
+%% \newcommand{\DepWith}[2]{{\textstyle\bigamp}\limits_{#1}{#2}
+%% \newcommand{\DepPlus}[2]{{\textstyle\bigoplus}\limits_{#1}{#2}
 \newcommand{\GrTy}{\mathsf{Gr}}
@@
 
-\title{A Dependent Lambek Calculus for Intrinsic Verification of Parsers and Formal Grammar Theory}
+\title{Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus}
 
@@
 \begin{abstract}
-  We propose \theoryname~(\theoryabbv) a new type-theoretic calculus
-  for the specification of formal grammars and intrinsic verification
-  of parsers. \theoryabbv is a dependent non-commutative
-  linear/non-linear type theory where the linear types can be viewed
-  as formal grammars, and linear terms can be viewed as parse
-  transformers, of which parsers are a special case. We formalize this
-  interpretation by giving a denotational semantics of the theory in a
-  novel categorical semantics of formal grammars in a category of
-  families indexed by strings.
-
-  To demonstrate the usefulness of the calculus for formal grammar
-  theory, we show that existing grammar formalisms --- e.g.\ regular
-  expressions, context free expressions, and their associated automata \max{which ones do
-    we actually show?} \steven{regular expressions, $\mu$-regular expressions
-    (or context free expressions), and different automata classes (finite,
-    pushdown, Turing)} --- can be expressed as linear types in
-  \theoryabbv. We give simple characterizations of the regular and
-  context-free grammars as expressible by restricting which
-  connectives of \theoryabbv are allowed, a form of type theoretic
-  characterization of these grammar classes. Further, we demonstrate
-  some central proofs of formal grammar theory can be intrinsically
-  verified: \emph{strong} equivalence between regular expressions and
-  NFAs is expressed by a syntactic \emph{isomorphism} definable in
-  \theoryabbv.
-
-  Finally, since \theoryabbv is a syntactic calculus, in addition to
-  the ``standard'' semantics of linear types as grammars, we
-  demonstrate two non-standard semantic interpretations. One is a
-  semantics where linear types are instead interpreted as
-  \emph{semantic actions}, rather than grammars, which are more common
-  in application than formal grammars proper. Lastly we use a
-  non-standard semantics to prove a \emph{canonicity} metatheorem that
-  shows that the interpretation of types as grammars is canonical: the
-  type-theoretic notion of a parse agrees with the expected semantic
-  notion.
+  We propose \theoryname~(\theoryabbv), a domain-specific dependent type theory
+  for the definition of intrinsically verified parsers and verification of weak
+  and strong equivalences of formal grammars. The linear types in \theoryabbv
+  have a simple interpretation as formal grammars, with values of the types
+  interpreted as parse trees and general terms as a form of intrinsically
+  verified parse transformers, of which parsers being a special case. We
+  demonstrate the expressivity of this system by showing that the indexed
+  inductive linear types in this calculus can be used to encode many different
+  familiar grammar formalisms, including regular and context-free grammars but
+  also traces of various types of automata. We can then construct strong
+  equivalences between grammar and automata formalisms and construct verified
+  parsers as simple linear functional programs. All of our examples have been
+  implemented in a prototype implementation of \theoryabbv as a shallow
+  embedding in Agda.
+
+  Our shallow embedding is based on a denotational semantics of \theoryabbv
+  where non-linear types denote ordinary sets and linear types denote formal
+  grammars. To develop this semantics compositionally, we define a notion of
+  general notion of model we call a \emph{Chomsky hyperdoctrine}, a
+  generalization of the Kleene algebra approach to formal language theory. In
+  addition to our ``standard model'' where linear types denote formal grammars,
+  we demonstrate two non-standard models. The first interprets linear types as
+  semantic actions over formal grammars, suggesting future extensions of our
+  calculus. The second is used to structure a logical relations argument to
+  prove a canonicity metatheorem for the type theory, proving that all closed
+  terms of context-free linear types normalize to a parse tree.
 \end{abstract}
@@
 
-Parsing is central to computing. Verification of parsers is a crucial
-task in many verified applications....There is a great deal of
-literature on specific parsing formalisms and proving them correct,
-but of course each formalism is restrictive and must be given a new
-formal treatement of its correctness.
-
-% TODO: more about why a *framework* is useful
-
-To address these issues with verified parsing, we have designed a
-\emph{domain-specific type theory} which supports the definition of grammar
-formalisms and \emph{intrinsic verification} of parser implementations. The type
-theory is based on a \emph{dependent non-commutative linear type theory}, which
-we call the \emph{dependent Lambek calculus}. In this type theory, the linear
-types are interpreted as \emph{formal grammars}, with the values being the
-\emph{parses} of the grammar. The terms are a kind of \emph{intrinsically
-verified} parse transformer: that is a term $x: A \vdash M : B$ implements a
-function that takes any $A$-parse of a string $w$ and returns a $B$-parse of the
-same string. As a special case, if $A$ is the trivial grammar of all strings,
-then $M$ is an \emph{intrinsically verified total parser} a function that
-returns a $B$ parse for every string. The benefit of this intrinsic approach to
-constructing parsers is that the terms in our Lambek calculus look like ordinary
-functional programs that obey an ordered linear typing discipline, yet this
-discipline ensures by construction that the grammatical structure is respected.
-
-To demonstrate that this is a \emph{practical} language for implementing grammar
-formalisms and intrinsically verified parsers, we demonstrate several examples: ...
-
-We have additionally implemented a proof of concept \emph{domain-specific proof
-assistant} via a shallow embedding of our type theory in Agda.
-
-intrinsically verified parsers for regular expressions and finite
-automata, as well as some example LL(1) grammars.
+%% Outline
+%% - Why a framework for parser verification
+%% - what intrinsic verification gets us
+%% - the secret sauce: indexed linear inductive types
+%% - prototype implementation
+%% - overview of "abstract" formal grammars?
 
@@
 
-In this work, we aim to elucidate these similarities and provide a
-mathematical theory of formal grammars for constructing verified
-parsers. Our approach can be summarized with the Curry-Howard-style
-slogan: ``grammars as types''; and the addendum: ``parsers as
-terms''. That is, we develop a type theory with a novel denotational
-semantics in a category with abstract formal grammars as objects and
-\emph{parse transformers} as morphisms. We say these are
+In this work, we aim to elucidate these similarities and
+provide a mathematical theory of formal grammars for constructing verified
+parsers. Our approach can be summarized with the
+Curry-Howard-style slogan: ``grammars as types''; and the addendum: ``parsers as
+terms''. That is, we develop a type theory with a novel
+denotational semantics in a category with abstract formal grammars as
+objects and \emph{parse transformers} as morphisms. We say these are
 \emph{abstract} formal grammars because they are not tied to any
 particular syntactic formalism such as Chomskyan grammars, regular
-expressions, Lambek calculus etc. The category we choose is a rich in
-structure: it is a topos as well as being monoidal closed, and these
+expressions, Lambek calculus etc. The category we choose is a rich
+in structure: it is a topos as well as being monoidal closed, and these
 universal properties serve as the inspiration for our type theory.
@@
 
-\section{A Syntax-Independent Notion of Syntax}
+\subsection{A Syntax-Independent Notion of Syntax}
 \label{sec:synindnotion}
@@
 
-\section{\theoryname as a Syntax for Formal Grammars}
+\section{Dependent Lambek Calculus by Example}
+\label{sec:type-theory-examples}
+
+%% 1. First example: finite grammars (regex minus Kleene star)
+%%    - why ordered linear:
+%%      + no exchange
+%%      + no contraction
+%%      + no weakening
+%%      + ambiguity
+%% 2. Second example: Kleene star
+%% 3. Third example: NFA traces to show the indexed inductive types
+
+\section{Syntax and Typing for \theoryname}
 \label{sec:tt}
+
+Next we present the full details of the syntax and typing of \theoryname.
+\theoryname has two kinds of types, linear types $A$ which represent formal
+grammars, and non-linear types $X$, which act like ordinary types or sets. Both
+linear and non-linear types are checked for well-formedness in a context
+$\Gamma$ containing \emph{only} non-linear variables $\Gamma = x:X,\ldots$.
+
+\begin{figure}
+  \begin{mathpar}
+    \boxed{\Gamma ; \Delta \vdash a : A}
+
+    \inferrule{~}{\Gamma ; a : A \vdash a : A}
+    \and
+    \inferrule{\Gamma ; \Delta \vdash e : B \\ \linctxwffjdg \Gamma {A \equiv B}}{\Gamma ; \Delta \vdash e : A}
+    %
+    \\
+    %
+    \inferrule{~}{\Gamma ; \cdot \vdash () : I}
+    \and
+    \inferrule{\Gamma ; \Delta \vdash e : I \\ \Gamma ; \Delta_1',\Delta_2' \vdash e' : C}{\Gamma ; \Delta_1',\Delta,\Delta_2' \vdash \letin {()} e {e'} : C}
+    %
+    \\
+    %
+    \inferrule{\Gamma ; \Delta \vdash e : A \\ \Gamma ; \Delta' \vdash e' : B}{\Gamma ; \Delta, \Delta' \vdash e \otimes e' : A \otimes B}
+    %
+    \and
+    %
+    \inferrule{\Gamma ; \Delta \vdash e : A \otimes B \\ \Gamma ; \Delta'_1, a : A, b : B, \Delta'_2 \vdash e'}{\Gamma ;  \Delta_1', \Delta, \Delta'_2 \vdash \letin {a \otimes b} e {e'}}
+    \\
+    %
+    \inferrule{\Gamma ; a : A , \Delta \vdash e : B}{\Gamma ; \Delta \vdash \lamblto a e : A\lto B}
+    \and
+    \inferrule{\Gamma ; \Delta' \vdash e' : A \\ \Gamma ; \Delta \vdash e : A \lto B}{\Gamma ; \Delta', \Delta \vdash \applto {e'} {e} : B}
+    \\
+    %
+    \inferrule{\Gamma ; \Delta , a : A \vdash e : B}{\Gamma ; \Delta \vdash \lambtol a e : B\tol A}
+    \and
+    \inferrule{\Gamma ; \Delta \vdash e : B \tol A \\ \Gamma ; \Delta' \vdash e' : A}{\Gamma ; \Delta, \Delta' \vdash \apptol e {e'} : B}
+    %
+    \\
+    %
+    \inferrule{\Gamma, x : X ; \Delta  \vdash e : A}
+              {\Gamma ; \Delta \vdash \dlamb x e : \LinPiTy x X A}
+    \and
+    \inferrule{\Gamma ; \Delta \vdash e : \LinPiTy x X A \\ \Gamma \vdash e' : X}{\Gamma ; \Delta \vdash e\,.\pi\,M : \subst A {e'} x}
+    %
+    \\
+    %
+    \inferrule{\Gamma \vdash M : X \quad \Gamma ; \Delta \vdash e : \subst A e x}{\Gamma ; \Delta \vdash \sigma\,M\,e : \bigoplus\limits_{x:X} A}
+    %
+    \and
+    %
+    \inferrule{\Gamma ; \Delta \vdash e : \bigoplus\limits_{x:X} A \quad \Gamma, x : X ; \Delta'_1, a : A, \Delta'_2 \vdash e' : C}{\Gamma; \Delta'_1, \Delta, \Delta'_2 \vdash \letin {\sigma\,x\,a} e {e'}: C}
+    %% %
+    %% \\
+    %% %
+    %% \inferrule{~}{\Gamma ; \Delta \vdash () : \top}
+
+    %% \and
+
+    %% \inferrule{~}{\Gamma ; x : \bot \vdash \mathsf{absurd}_A : A}
+    %% %
+    %% \and
+    %% %
+    %% %
+    %% \inferrule{\Gamma \vdash e : \mathsf{G} A}{\Gamma ; \cdot \vdash \mathsf{G}^{-1}\, e : A}
+    %% %
+  \end{mathpar}
+  \caption{Linear typing}
+  \label{fig:linsyntax}
+\end{figure}
+
+\begin{figure}
+  \begin{mathpar}
+    \inferrule{~}{\Gamma \vdash \Set_0 : \Set_1}
+ %
+    \and
+%
+    \inferrule{~}{\Gamma \vdash L_i : U_{i+1}}
+    {}{}
+  \end{mathpar}
+  \caption{Universe Types}
+\end{figure}
+
+\begin{figure}
+  \textrm{TODO: var, konstant, bigoplus, bigamp and tensor}\\
+  \textrm{TODO: behavior of map}
+  \begin{mathpar}
+    \inferrule
+    {\Gamma \vdash M : X}
+    {\Gamma \vdash \mathsf{Var}\,M : \SPF\,X}
+
+    \inferrule
+    {\Gamma \vdash A : L}
+    {\Gamma \vdash \mathsf{K}\,A : \SPF\,X}
+
+    \inferrule
+    {\Gamma,y:Y \vdash A : \SPF\,X}
+    {\Gamma \vdash \bigoplus_{y:Y}\,A : \SPF\,X}
+
+    \inferrule
+    {\Gamma,y:Y \vdash A : \SPF\,X}
+    {\Gamma \vdash \bigamp_{y:Y}\,A : \SPF\,X}
+
+    \inferrule
+    {\Gamma \vdash A : \SPF\,X\and \Gamma \vdash B : \SPF\,X}
+    {\Gamma \vdash A \otimes B : \SPF\,X}
+
+    \inferrule
+    {A : \SPF\,X}
+    {\el(A) : (X \to L) \to L}
+
+    \el(\Var\,M) B = B\,M\and
+    \el(\mathsf{K}\,A) B = A\and
+    \el(\bigoplus_{y:Y} A) B = \bigoplus_{y:Y}\el(A)B\and
+    \el(\bigamp_{y:Y} A) B = \bigamp_{y:Y}\el(A)B\and
+    \el(A \otimes A') B = \el(A)B \otimes \el(A')B
+
+  \inferrule
+  {A : \textrm{SPF} X\and
+  \Gamma,x:X;b: B x \vdash f : C x\and
+  \Delta \vdash e : \el(A)(B)
+  }
+  {\Gamma;\Delta \vdash \mathsf{map}\,A\,(x.b.f)\,e : \textrm{el}(A)\, C}
+
+  \map(\Var\,M)\,(x.b.f)\,e = f[M/x][e/b]\and
+  \map(\K\,A)\,(x.b.f)\,e = e\and
+  \map(\bigoplus\limits_{y:Y} A)\,(x.b.f)\,e =
+  \letin{\oplusinj y a}{e}{\oplusinj y {\map(A)\,(x.b.f)\,e}}\\
+  \map(\bigwith\limits_{y:Y} A)\,(x.b.f)\,e =
+  \withlamb{y}{\map(A)\,(x.b.f)\,(\withprj a e)}\\
+  \map(A \otimes A')\,(x.b.f)\,e =
+  \letin{(a,a')}{e}{(\map(A)\,(x.b.f)\,a,\map(A')\,(x.b.f)\,a')}
+  \end{mathpar}
+  \caption{Strictly Positive Functors}
+\end{figure}
+
+\begin{figure}
+  \begin{mathpar}
+    \inferrule*[right=IndIntro]
+    {\Gamma \vdash A : X \to \textrm{SPF}\,X\and
+     \Gamma \vdash M : X\and
+     \Gamma; \Delta \vdash e : \textrm{el}(A\,M) (\mu\,A)}
+    {\Gamma; \Delta \vdash \mathsf{roll}\, e : \mu\,A\, M}
+
+    \inferrule*[right=IndElim]
+    {}
+    {\Gamma; \Delta \vdash \mathsf{fold}(x.b.f)(M, e) : B\,M}
+  \end{mathpar}
+  \caption{Indexed Inductive Linear Types}
+\end{figure}
+
+\begin{figure}
+  \begin{mathpar}
+    \begin{array}{rcl}
+      \textrm{non-linear types} X,Y,Z & ::= & \textrm{Empty}\pipe\textrm{Unit}\pipe\textrm{Bool}\pipe\textrm{Nat}
+      \pipe \sum_{x:X} Y \pipe \prod_{x:X}Y\pipe \textrm{SPF}\,X \pipe \cdots \max{also eliminators...}\\
+      \textrm{linear types} A,B,C & ::= & c \pipe I \pipe A \otimes B \pipe A \lto B \pipe B \tol A
+      \pipe \bigoplus_{x:X} A \pipe \bigamp_{x:X} A \pipe \mu e e'\\
+      \textrm{non-linear terms} M,N,O & ::= & \cdots\\
+      \textrm{linear terms} 
+    \end{array}
+  \end{mathpar}
+  \caption{Syntax of Types and Terms}
+\end{figure}
 
@@
   \begin{mathpar}
-    \boxed{\Gamma ; \Delta \vdash a : A}
-
-    \inferrule{~}{\Gamma ; a : A \vdash a : A}
-    \and
-    \inferrule{\Gamma ; \Delta \vdash e : B \\ \linctxwffjdg \Gamma {A \equiv B}}{\Gamma ; \Delta \vdash e : A}
-    %
-    \\
-    %
-    \inferrule{~}{\Gamma ; \cdot \vdash () : I}
-    \and
-    \inferrule{\Gamma ; \Delta \vdash e : I \\ \Gamma ; \Delta_1',\Delta_2' \vdash e' : C}{\Gamma ; \Delta_1',\Delta,\Delta_2' \vdash \letin {()} e {e'} : C}
-    %
-    \\
-    %
-    \inferrule{\Gamma ; \Delta \vdash e : A \\ \Gamma ; \Delta' \vdash e' : B}{\Gamma ; \Delta, \Delta' \vdash e \otimes e' : A \otimes B}
-    %
-    \and
-    %
-    \inferrule{\Gamma ; \Delta \vdash e : A \otimes B \\ \Gamma ; \Delta'_1, a : A, b : B, \Delta'_2 \vdash e'}{\Gamma ;  \Delta_1', \Delta, \Delta'_2 \vdash \letin {a \otimes b} e {e'}}
-    \\
-    %
-    \inferrule{\Gamma ; a : A , \Delta \vdash e : B}{\Gamma ; \Delta \vdash \lamblto a e : A\lto B}
-    \and
-    \inferrule{\Gamma ; \Delta' \vdash e' : A \\ \Gamma ; \Delta \vdash e : A \lto B}{\Gamma ; \Delta', \Delta \vdash \applto {e'} {e} : B}
-    \\
-    %
-    \inferrule{\Gamma ; \Delta , a : A \vdash e : B}{\Gamma ; \Delta \vdash \lambtol a e : B\tol A}
-    \and
-    \inferrule{\Gamma ; \Delta \vdash e : B \tol A \\ \Gamma ; \Delta' \vdash e' : A}{\Gamma ; \Delta, \Delta' \vdash \apptol e {e'} : B}
-    %
-    \\
-    %
-    \inferrule{\Gamma, x : X ; \Delta  \vdash e : A}{\Gamma ; \Delta \vdash \dlamb x e : \LinPiTy x X A}
-    \and
-    \inferrule{\Gamma ; \Delta \vdash e : \LinPiTy x X A \\ \Gamma \vdash e' : X}{\Gamma ; \Delta \vdash \app e {e'} : \subst A {e'} x}
-    %
-    \\
-    %
-    \inferrule{\Gamma \vdash e : X \quad \Gamma ; \Delta \vdash e' : \subst A e x}{\Gamma ; \Delta \vdash (e, e') : \LinSigTy x X A}
-    %
-    \and
-    %
-    \inferrule{\Gamma ; \Delta \vdash e : \LinSigTy x X A \quad \Gamma, x : X ; \Delta'_1, a : A, \Delta'_2 \vdash e' : C}{\Gamma; \Delta'_1, \Delta, \Delta'_2 \vdash \letin {(x, a)} e {e'}: C}
-    %
-    \\
-    %
-    \inferrule{~}{\Gamma ; \Delta \vdash () : \top}
-
-    \and
-
-    \inferrule{~}{\Gamma ; x : \bot \vdash \mathsf{absurd}_A : A}
-    %
-    \and
-    %
-    %
-    \inferrule{\Gamma \vdash e : \mathsf{G} A}{\Gamma ; \cdot \vdash \mathsf{G}^{-1}\, e : A}
-    %
-    \\
-    %
-    \inferrule{\Gamma; \Delta \vdash e : \subst A {\mu x.\, A} x}{\Gamma; \Delta \vdash \mathsf{cons}\, e : \mu x.\, A}
-    \and
-    \inferrule{\Gamma;\Delta\vdash e' : \mu x.\,A \and \Gamma; a:\subst A B x \vdash e : B}{\Gamma;\Delta\vdash \mathsf{fold}(a.e)(e') : B}
-  \end{mathpar}
-  \caption{Linear typing}
-  \label{fig:linsyntax}
-\end{figure}
-
-\begin{figure}
-  \begin{mathpar}
     \boxed{\Gamma \vdash e \equiv e' : X}
@@
 
-\section{Automata as Grammars}
-\label{sec:automata}
+\section{Formal Grammar Theory in Lambek}
+\label{sec:applications}
+
+%% Make sure to say which results are mechanized
+%% 1. Basic concepts: unambiguity, parseability, weak and strong equivalence
+%% 2. Regex, NFA, DFA
+%% 3. Dyck language
+%% 4. LL(1) parser
+%% 5. Expressiveness: Turing Machines
 
@@
 
-\section{Categorical Semantics of \theoryabbv}
-\label{sec:categorify}
+\section{Semantics and Metatheory}
+\label{sec:semantics-and-metatheory}
+%% \section{Categorical Semantics of \theoryabbv}
+%% \label{sec:categorify}
 
@@
 
-\subsection{Chomsky Categories}
+\subsection{Lambek Hyperdoctines and Chomsky Hyperdoctrines}
 
@@
 
+We do this in two stages: first we define a \emph{Lambek hyperdoctrine} to be a
+notion of model for the judgmental structure of \theoryabbv: that is we have
+notions of linear and non-linear type, contexts, substitution and terms, but do
+not assume any particular type constructions exist. Then we will define when a
+Lambek hyperdoctrine is a Chomsky hyperdoctrine, meaning it can interpret all
+the type formers of \theoryabbv and therefore arbitrary strength grammars.
+
+\begin{definition}[Lambek Hyperdoctrine]
+  A \emph{Lambek hyperdoctrine} consists of
+  \begin{enumerate}
+  \item A category $\mathcal C$ with a terminal object.
+  \item A category with families structure over $\mathcal C$.
+  \item A contravariant functor $L : \mathcal C^{o} \to \textrm{MultiCat}$ from
+    $\mathcal C$ to the category of Multicategories.
+  \end{enumerate}
+\end{definition}
+Here the objects of $\mathcal C$ model the non-linear contexts, and morphisms
+model non-linear substitutions. The category with families structure over
+$\mathcal C$ models the dependent non-linear types. Finally the
+``hyperdoctrine'' of multicategories $L$ models the linear types and terms. The
+fact that this is functorial in $\mathcal C$ corresponds to the fact that linear
+types are all relative to a non-linear context and that linear variables and
+composition commute with non-linear substitution.
+
+We will have three main Lambek hyperdoctrines of interest in this paper: the
+``standard model'' of sets and grammars, the ``syntactic model'' given by our
+type theory itself and lastly a gluing model introduced in
+Section~\ref{sec:canonicity} to prove the canonicity theorem.
+\begin{example}
+  \begin{enumerate}
+  \item Let $\mathcal C$ be the category of sets, equipped with its usual
+    category with families structures of families. Define $L : \mathcal C^{o}
+    \to \textrm{MultiCat}$ to map $L(X)$ to the representable multicategory
+    $(\Set^{\Sigma^*})^X$ of $X$-indexed families of grammars where the monoidal
+    structure is given pointwise by the Day convolution monoidal structure.
+  \item We can build a ``syntactic'' Lambek hyperdoctrine $\Syn(\Sigma)$ from the syntax
+    itself: the category $\mathcal C$ is given by non-linear contexts and
+    substitutions, the category with families by the non-linear types and terms
+    and the hyperdoctrine of multicategories by the linear types and terms.
+  \end{enumerate}
+\end{example}
+
+Next each type linear and non-linear type constructor corresponds to extra data
+on a Lambek hyperdoctrine.
+\begin{definition}
+  \label{def:chomsky-data}
+  \begin{enumerate}
+  \item Dependent type structure: standard (say extensional type theory, give a reference)
+  \item Universes
+  \item Inductive grammars?
+  \item Non-inductive Grammar constructors: standard monoidal category stuff
+  \end{enumerate}
+\end{definition}
+\begin{definition}[Chomsky Hyperdoctrine]
+  A Chomsky Hyperdoctrine is a Lambek hyperdoctrine equipped with all of the data in Definition~\ref{def:chomsky-data}.
+\end{definition}
 \pedro{We should probably define some of the words in this definition}
@@
 
-\section{Concrete Models of \theoryabbv}
+\subsection{Concrete Models of \theoryabbv}
 \label{sec:othermodels}
@@
 
+%% 1. More practical: integrating it into a larger verified development
+%%    - verified imperative implementation: Lambek logic ala separation
+%%      logic?
+%% 2. Semantic actions
+%% 3. Type systems as tree grammars
+
 The core objects of study in the theory of formal grammars have stood stalwart

### 2024-10-10 paper/paper.tex
intro/elim/beta/eta for indexed inductives on paper, and start agda formalization of on-paper rules
remove unused fig
some syntax

@@
 \newcommand{\map}{\mathsf{map}}
+\newcommand{\roll}{\mathsf{roll}}
+\newcommand{\fold}{\mathsf{fold}}
 \newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
@@
 
+\newcommand{\equalizer}[4]{\{#1\,|\,\apptol {#2}{#1} = \apptol{#3}{#1 \}}
+
 \newcommand{\ctxwff}[1]{#1 \,\, \mathsf{ok}}
@@
 \begin{figure}
+  \begin{array}{rrcl}
+    \textrm{non-linear contexts} & \Gamma & ::= & \cdot \pipe \Gamma,x:X \\
+    % 
+    \textrm{non-linear types} & X , Y , Z & ::= & \mathsf{Nat} \pipe \mathsf{Int} \pipe \mathsf{Empty} \pipe \mathsf{Unit} \pipe \mathsf{SPF}\,X\pipe U \pipe L \pipe !A \pipe \sum_{x:X} Y \pipe \prod_{x:X} Y\pipe M =_A N\\
+    \textrm{linear contexts} & \Delta & ::= & \cdot \pipe \Delta , a:A\\
+    \textrm{linear types} & A , B , C & ::= &
+    c \pipe I \pipe A \otimes B \pipe \bigoplus\limits_{x:X} A \pipe \mu A_f M
+    A \lto B \pipe B \tol A \pipe \bigwith\limits_{x:X} A \pipe \equalizer {a}{e}{f}\\
+  \end{array}
+  \caption{Overview of Types}
+\end{figure}
+
+\begin{figure}
   \begin{mathpar}
@@
     \inferrule*[right=IndElim]
+    {\Gamma;\Delta \vdash e : \mu A M\and
+     \Gamma,x:X;b:\el(A\,x)\,B \vdash f : B\,x
+    }
+    {\Gamma; \Delta \vdash \mathsf{fold}(A)(x.b.f)(M, e) : B\,M}
+
+    \inferrule*[right=Ind$\beta$]
     {}
-    {\Gamma; \Delta \vdash \mathsf{fold}(x.b.f)(M, e) : B\,M}
+    {\fold(A)(x.b.f)(M, \roll(e)) =
+      f[M/x][\map(A\,M)\,(y.a.\fold(A)(x.b.f)(y,a))\,e/b]}
+
+    \inferrule*[right=Ind$\eta$]
+    {\Gamma,x:X; a:\mu\,A\,x \vdash e' : B\,x
+      \and \Gamma;\Delta \vdash e : \mu A M\\\\
+      \Gamma,x:X; a':\el(A x)(\mu A) \vdash
+      e'[x/x][\roll(a')/a] = f[x/x][\map(A\,x)\,(y.a.e'[y/x][a/a])\,a/b] : B\,x
+    }
+    {\Gamma; \Delta \vdash \fold(A)(x.b.f)(M,e) = e'[M/x][e/a] : B\,M}
   \end{mathpar}
   \caption{Indexed Inductive Linear Types}
-\end{figure}
-
-\begin{figure}
-  \begin{mathpar}
-    \begin{array}{rcl}
-      \textrm{non-linear types} X,Y,Z & ::= & \textrm{Empty}\pipe\textrm{Unit}\pipe\textrm{Bool}\pipe\textrm{Nat}
-      \pipe \sum_{x:X} Y \pipe \prod_{x:X}Y\pipe \textrm{SPF}\,X \pipe \cdots \max{also eliminators...}\\
-      \textrm{linear types} A,B,C & ::= & c \pipe I \pipe A \otimes B \pipe A \lto B \pipe B \tol A
-      \pipe \bigoplus_{x:X} A \pipe \bigamp_{x:X} A \pipe \mu e e'\\
-      \textrm{non-linear terms} M,N,O & ::= & \cdots\\
-      \textrm{linear terms} 
-    \end{array}
-  \end{mathpar}
-  \caption{Syntax of Types and Terms}
 \end{figure}
@@
   \inferrule{\Gamma ; \cdot \vdash e : A}{\Gamma \vdash \mathsf G e : \mathsf G A}
+
+  \textrm{TODO: inductive types (boolean, nat) with their eliminators allowed in }
   \end{mathpar}

### 2024-10-11 paper/paper.tex
fixed typo

--- 2024-10-03 paper/paper.tex
@@
 
-While there are ample thoery and tooling to handle the first stage of a language frontend, the final
+While there are ample theory and tooling to handle the first stage of a language frontend, the final
 frontier for parsing research is the development of
@@
     \inferrule*[right=IndElim]
+    {\Gamma;\Delta \vdash e : \mu A M\and
+     \Gamma,x:X;b:\el(A\,x)\,B \vdash f : B\,x
+    }
+    {\Gamma; \Delta \vdash \mathsf{fold}(A)(x.b.f)(M, e) : B\,M}
+
+    \inferrule*[right=Ind$\beta$]
     {}
-    {\Gamma; \Delta \vdash \mathsf{fold}(x.b.f)(M, e) : B\,M}
+    {\fold(A)(x.b.f)(M, \roll(e)) =
+      f[M/x][\map(A\,M)\,(y.a.\fold(A)(x.b.f)(y,a))\,e/b]}
+
+    \inferrule*[right=Ind$\eta$]
+    {\Gamma,x:X; a:\mu\,A\,x \vdash e' : B\,x
+      \and \Gamma;\Delta \vdash e : \mu A M\\\\
+      \Gamma,x:X; a':\el(A x)(\mu A) \vdash
+      e'[x/x][\roll(a')/a] = f[x/x][\map(A\,x)\,(y.a.e'[y/x][a/a])\,a/b] : B\,x
+    }
+    {\Gamma; \Delta \vdash \fold(A)(x.b.f)(M,e) = e'[M/x][e/a] : B\,M}
   \end{mathpar}

### 2024-10-24 paper/paper.tex
fixed equalizer command
equalizer intro, beta, eta
Some examples

--- 2024-10-11 paper/paper.tex
@@
 \usepackage{xspace}
+\usepackage{listings}
 \usepackage{hyperref}
@@
 \newcommand{\fold}{\mathsf{fold}}
+\newcommand{\inl}{\mathsf{inl}}
+\newcommand{\inr}{\mathsf{inr}}
 \newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
@@
 \DeclareMathOperator*{\bigwith}{\scalerel*{\&}{\bigoplus}}
+
+\newcommand{\bang}{~\textbf{!}~}
+\newcommand{\nil}{\mathsf{nil}}
+\newcommand{\cons}{\mathsf{cons}}
+\newcommand{\data}{\mathsf{data}}
+\newcommand{\stringquote}[1]{\texttt{\textquotesingle#1\textquotesingle}}
 
@@
 
+\newcommand{\equalizer}[3]{\{#1\,|\,\apptol {#2}{#1} = \apptol{#3}{#1} \}}
+\newcommand{\equalizerpair}[2]{\langle #1 , #2 \rangle}
+\newcommand{\equalizerpi}[1]{#1.\pi}
+
 \newcommand{\ctxwff}[1]{#1 \,\, \mathsf{ok}}
@@
 
-\section{Dependent Lambek Calculus by Example}
+\section{\theoryname by Example}
 \label{sec:type-theory-examples}
+
+To gain intuiton for working in \theoryname, let's walk through some examples of
+grammars and parses internal to it.
+
+\paragraph{Finite Grammars}
+First, consider finite grammars over the alphabet $\{ a , b , c \}$. Each letter of the
+alphabet corresponds to a base type in our calculus. For instance, the grammar
+for the single character $a$ has a single parse tree for the string \stringquote{a}.
+and no parse trees at any other strings.
+
+A finite grammar is built from base types via dijunctions $\oplus$ and
+concatenations $\otimes$ --- for instance, $(a \oplus b) \otimes c$.
+In \cref{fig:fingram} we find a proof tree showing that the input string \stringquote{ac} matches
+the grammar $(a \oplus b) \otimes c$.
+
+\begin{figure}
+\begin{mathpar}
+  \inferrule
+  {
+    \inferrule
+    {~}
+    {x : a \vdash \inl(x) : a \oplus b}
+    \\
+    \inferrule
+    {~}
+    {y : c \vdash y : c}
+  }
+  {x : a , y : c \vdash \inl(x) \otimes y : (a \oplus b) \otimes c}
+\end{mathpar}
+\caption{\stringquote{ac} matches $(a \oplus b) \otimes c$}
+\label{fig:fingram}
+\end{figure}
+
+To parse the specific string \stringquote{ac}, we populate the context with an
+internalized encoding of that string. In contexts of this form, linear terms
+describe the structure of a parse tree. That is, $\inl(x) \otimes y$ details the
+manner by which \stringquote{ac} matches $(a \oplus b) \otimes c$.
+
+Because derivations in a linear context correspond to parse trees in this
+manner, we cannot affirm all of the usual structural rules of a type system.
+Weakening and contraction each allow for modification of the string in context ---
+weakening allows for characters to be dropped at will, while contraction allows
+for duplication of characters. Omission of these rules offers a
+resource-sensitive view that ensures only the inputted characters are parsed.
+Moreover, strings are not simply sets of characters, as they carry an inherent
+ordering. Thus, the rule of exchange is also omitted in this calculus so that
+the constituent characters of a string are not freely permuted.
+
+Without the above structural rules we
+arrive at an ordered, linear type system, and to summarize this
+particular choice of substructural restrictions: when parsing, characters must
+be processed
+without modification and in the order in which they were written.
+
+\paragraph{Regular Expressions}
+Next, we demonstrate how to parse a regular expression in \theoryabbv. Again,
+consider grammars over $\{ a , b , c \}$. Similar to finite grammars, regular
+expresssions are built inductively using base types, $\oplus$, and $\otimes$ ;
+however, unlike finite grammars, regular expressions additionally have a
+restricted form of recursion via Kleene star $(\cdot)^{\ast}$.
+
+% TODO this should be beter formatted but can't use listings with \otimes ^{\ast}
+\begin{figure}
+\begin{align*}
+\data &~A^{*}:\\
+      & \nil : I \to A* \\
+      & \cons : A \otimes A* \to A*
+\end{align*}
+\caption{Kleene Star as an inductive type}
+\label{fig:kleenestarinductive}
+\end{figure}
+
+For a grammar $A$, the recursive nature of $A^{*}$ is captured by defining
+$A^{*}$ to be an inductive linear
+type, as shown in \cref{fig:kleenestarinductive}. Just as in \cref{fig:fingram}, we can show that particular strings match a Kleene
+star. In \cref{fig:kleenestarderivation}, we demonstrate that \stringquote{ab}
+matches $a^{*} \otimes b$ through repeated application of the Kleene star constructors.
+
+\begin{figure}
+\begin{mathpar}
+  \inferrule
+  {
+    \inferrule
+    {
+      \inferrule
+      {~}
+      {x : a \vdash x : a}
+      \\
+      \inferrule
+      {
+        \inferrule
+        {~}
+        {I \vdash \nil : a^{*}}
+      }
+      {\cdot \vdash \nil : a^{*}}
+    }
+    {x : a \vdash \cons (x , \nil) : a^{*}}
+    \\
+    \inferrule
+    {~}
+    {y : b \vdash y : b}
+  }
+  {x : a , y : b \vdash \cons(x ) \otimes y : a^{*} \otimes b}
+\end{mathpar}
+\caption{\stringquote{ab} matches $a^{*} \otimes b$}
+\label{fig:kleenestarderivation}
+\end{figure}
+
+However, we can have derivations where the term in context is not simply a
+string of base types. For instance, by using the elimination principle for
+Kleene star, \cref{fig:kleeneabstractproof} builds a term of type
+$A^{*}$ in context comprising abstract\footnote{A linear type that is not a
+  finite grammar.} linear types. In these contexts, it is not quite appropriate
+to treat linear terms as parse \textit{trees}, because the information in
+context is not a string. Instead, it is more
+appropriate to think of such terms as parse \textit{transformers}. That is, in
+\cref{fig:kleeneabstractproof} the
+inhabitant $x : (A \otimes A)^{*}$ in the context is not itself data to be parsed; rather,
+$x$ is a parse tree and the term from context $x : (A \otimes A)^{*}$ to the
+term $A^{*}$ is a transformation of parse trees of the input grammar into parse
+trees of the output grammar.
+
+\steven{TODO : describe elimination principle somewhere. Maybe just in prose?}
+\steven{I probably can't get away with the inline definition of $f$ below}
+\begin{figure}
+\begin{mathpar}
+  \inferrule
+  {
+    \inferrule
+    {~}
+    {x_I : I \vdash \nil : A^*}
+    \\
+    \inferrule
+    {
+      \inferrule
+      {~}
+      {x_1 : A , x_2 : A , y : A* \vdash \cons (x_1 , \cons(x_2 , y)) : A^*}
+    }
+    {x : A \otimes A , y : A^*  \vdash f := \letin {x_1 \otimes x_2} {x} {\cons(x_1 ,
+        \cons(x_2 , y))} : A^*}
+  }
+  {x : (A \otimes A)^* \vdash \fold(\nil , f) : A^* }
+\end{mathpar}
+\caption{An example mapping out of abstract grammars}
+\label{fig:kleeneabstractproof}
+\end{figure}
+
+% \begin{tikzpicture}[%opacity=0.5,
+%   treenode/.style = {shape=rectangle, rounded corners,
+%     draw, align=center,
+%     top color=white, bottom color=white!20},
+%     line/.style={draw, -latex'},
+%     edge from parent/.style={draw,-latex'},
+%     level 1/.style={sibling distance=20mm, level distance=0.5cm},
+%     level 2/.style={sibling distance=1cm, level distance=1cm},
+%     visible on=<3->
+%     ]
+%   \node[treenode] (tree) {$\otimes$}
+%     child { node[treenode] (a) {${\color{blue} a}$} }
+%     child { node[treenode] (b){${\color{magenta} \mathsf{inl}}$}
+%         child { node[treenode] (c) {${\color{magenta} b}$} }
+%       };
+%       \node[draw, fit=(tree)(a)(b)(c),
+%       label={[name=l] $g(ab)$}, visible on=<3->] {};
+%     \node[below=2.5cm of tree] (mt) {};
+%     \node[below=2.5cm of a] (mta) {};
+%     \node[below=2.5cm of b] (mtb) {};
+%     \node[below=-0.2cm of mt, visible on=<4->] (mtset) {};
+%       \node[draw, fit=(mt)(mtset)(mta)(mtb),
+%       label={[name=l, visible on=<4->] $g(aa)$}, visible on=<4->] {};
+% \end{tikzpicture}
 
@@
 
+\steven{This figure is broken}
+\begin{figure}
+  % \begin{array}{rrcl}
+  %   \textrm{non-linear contexts} & \Gamma & ::= & \cdot \pipe \Gamma,x:X \\
+  %   \textrm{non-linear types} & X , Y , Z & ::= & \mathsf{Nat} \pipe
+  %                                                 \mathsf{Int} \pipe
+  %                                                 \mathsf{Empty} \pipe
+  %                                                 \mathsf{Unit} \pipe
+  %                                                 \mathsf{SPF}\,X\pipe U \pipe L
+  %                                                 \pipe !A \pipe \sum_{x:X} Y
+  %                                                 \pipe \prod_{x:X} Y\pipe M =_A
+  %                                                 N \\
+  %   \textrm{linear contexts} & \Delta & ::= & \cdot \pipe \Delta , a:A\\
+  %   \textrm{linear types} & A , B , C & ::= &
+  %   c \pipe I \pipe A \otimes B \pipe \bigoplus\limits_{x:X} A \pipe \mu A_f M
+  %   A \lto B \pipe B \tol A \pipe \bigwith\limits_{x:X} A \pipe \equalizer {a}{f}{g}\\
+  % \end{array}
+  \caption{Overview of Types}
+\end{figure}
+
 \begin{figure}
@@
     \inferrule{\Gamma ; \Delta \vdash e : \bigoplus\limits_{x:X} A \quad \Gamma, x : X ; \Delta'_1, a : A, \Delta'_2 \vdash e' : C}{\Gamma; \Delta'_1, \Delta, \Delta'_2 \vdash \letin {\sigma\,x\,a} e {e'}: C}
-    %% %
+    %
+    \and
+    %
+    \inferrule
+    {~}
+    {\Gamma , x : \square A ; \cdot \vdash \bang x : A}
+    %
+    \and
+    %
+    \inferrule
+    {\Gamma \vdash f : \square (A \lto B) \\
+      \Gamma \vdash g : \square (A \lto B) \\
+      \Gamma ; \Delta \vdash a : A \\
+      \Gamma ; \Delta \vdash p : \apptol {\bang f}{a} = \apptol {\bang g}{a}}
+    {\Gamma ; \Delta \vdash \equalizerpair{a}{p} : \equalizer {a}{\bang f}{\bang
+      g}}
+    %
+    \and
+    %
+    \inferrule
+    {\Gamma \vdash f : \square (A \lto B) \\
+      \Gamma \vdash g : \square (A \lto B) \\
+      \Gamma ; \Delta \vdash e : \equalizer{a}{\bang f}{\bang g}}
+    {\Gamma ; \Delta \vdash \equalizerpi {e} : A}
     %% \\
@@
   \caption{Indexed Inductive Linear Types}
-\end{figure}
-
-\begin{figure}
-  \begin{mathpar}
-    \begin{array}{rcl}
-      \textrm{non-linear types} X,Y,Z & ::= & \textrm{Empty}\pipe\textrm{Unit}\pipe\textrm{Bool}\pipe\textrm{Nat}
-      \pipe \sum_{x:X} Y \pipe \prod_{x:X}Y\pipe \textrm{SPF}\,X \pipe \cdots \max{also eliminators...}\\
-      \textrm{linear types} A,B,C & ::= & c \pipe I \pipe A \otimes B \pipe A \lto B \pipe B \tol A
-      \pipe \bigoplus_{x:X} A \pipe \bigamp_{x:X} A \pipe \mu e e'\\
-      \textrm{non-linear terms} M,N,O & ::= & \cdots\\
-      \textrm{linear terms} 
-    \end{array}
-  \end{mathpar}
-  \caption{Syntax of Types and Terms}
 \end{figure}
@@
     \inferrule{\Gamma, x : L_i \vdash A : L_i \and A \textrm{ strictly positive}}{\Gamma \vdash \mu x.\, A : L_i}
+    %
+    \and
+    %
+    \inferrule
+    {\Gamma \vdash A : L_i}
+    {\Gamma \vdash \square A : U_i}
+    %
+    \and
+    \inferrule
+    {\Gamma ; \Delta \vdash f : A \\ \Gamma \vdash g : \square (A \lto B)}
+    {\Gamma ; \Delta \vdash \equalizer {a}{\bang f}{\bang g} : L_i}
   \end{mathpar}
@@
     \inferrule{~}{\Gamma; \Delta \vdash \letin {(x, a)} e {\subst {e'} {(x, a)} y} \equiv \subst {e'} e y : C}
+    \and
+    %
+    \inferrule
+    {\Gamma \vdash f : \square (A \lto B) \\
+      \Gamma \vdash g : \square (A \lto B) \\
+      \Gamma ; \Delta \vdash a : A \\
+      \Gamma ; \Delta \vdash p : \apptol {\bang f}{a} \equiv \apptol {\bang g}{a}}
+    {\Gamma ; \Delta \vdash \equalizerpi {\equalizerpair {a}{p}} \equiv a : A}
+    %
+    \and
+    %
+    \inferrule
+    {\Gamma \vdash f : \square (A \lto B) \\
+      \Gamma \vdash g : \square (A \lto B) \\
+      \Gamma ; \Delta \vdash p : \apptol {\bang f}{a} = \apptol {\bang g}{a} \\
+      \Gamma ; \Delta \vdash e : \equalizer{a}{\bang f}{\bang g}}
+    {\Gamma ; \Delta \vdash \equalizerpair {\equalizerpi {e}} {p} \equiv e : A}
 \end{mathpar}

### 2024-10-25 paper/paper.tex
example rough draft

@@
 % -*- fill-column: 80; -*-
-\documentclass[sigconf,review,anonymous,screen,acmsmall]{acmart}
+\documentclass[review,anonymous,screen,acmsmall,nonacm]{acmart}
 \usepackage{mathpartir}
@@
 \newcommand{\data}{\mathsf{data}}
+\newcommand{\where}{\mathsf{where}}
+\newcommand{\Trace}{\mathsf{Trace}}
 \newcommand{\stringquote}[1]{\texttt{\textquotesingle#1\textquotesingle}}
@@
 
+\steven{To save space, each of these example figures can be
+  minified/potentially inlined}
+
 To gain intuiton for working in \theoryname, let's walk through some examples of
-grammars and parses internal to it.
-
-\paragraph{Finite Grammars}
-First, consider finite grammars over the alphabet $\{ a , b , c \}$. Each letter of the
+grammars and parses internal to it. In each of our examples, we will look at
+grammars over the three character alphabet $\{ a , b , c\}$.
+
+Each letter of the
 alphabet corresponds to a base type in our calculus. For instance, the grammar
@@
 
-A finite grammar is built from base types via dijunctions $\oplus$ and
-concatenations $\otimes$ --- for instance, $(a \oplus b) \otimes c$.
+\paragraph{Finite Grammars}
+First consider finite grammars --- those built from base types via dijunctions $\oplus$ and
+concatenations $\otimes$. For instance, $(a \oplus b) \otimes c$.
 In \cref{fig:fingram} we find a proof tree showing that the input string \stringquote{ac} matches
@@
 \paragraph{Regular Expressions}
-Next, we demonstrate how to parse a regular expression in \theoryabbv. Again,
-consider grammars over $\{ a , b , c \}$. Similar to finite grammars, regular
+Next, we demonstrate how to parse a regular expression in \theoryabbv. Similar
+to finite grammars, regular
 expresssions are built inductively using base types, $\oplus$, and $\otimes$ ;
@@
 \begin{align*}
-\data &~A^{*}:\\
-      & \nil : I \to A* \\
-      & \cons : A \otimes A* \to A*
+\data &~A^{*}: L_{i}~\where\\
+      & \nil : I \lto A^{*} \\
+      & \cons : A \otimes A^{*} \lto A^{*}
 \end{align*}
@@
 Kleene star, \cref{fig:kleeneabstractproof} builds a term of type
-$A^{*}$ in context comprising abstract\footnote{A linear type that is not a
-  finite grammar.} linear types. In these contexts, it is not quite appropriate
+$A^{*}$ in context comprising abstract linear types. In these contexts, it is
+not quite appropriate
 to treat linear terms as parse \textit{trees}, because the information in
@@
 $x$ is a parse tree and the term from context $x : (A \otimes A)^{*}$ to the
-term $A^{*}$ is a transformation of parse trees of the input grammar into parse
+type $A^{*}$ is a transformation of parse trees of the input grammar into parse
 trees of the output grammar.
@@
 
-% \begin{tikzpicture}[%opacity=0.5,
-%   treenode/.style = {shape=rectangle, rounded corners,
-%     draw, align=center,
-%     top color=white, bottom color=white!20},
-%     line/.style={draw, -latex'},
-%     edge from parent/.style={draw,-latex'},
-%     level 1/.style={sibling distance=20mm, level distance=0.5cm},
-%     level 2/.style={sibling distance=1cm, level distance=1cm},
-%     visible on=<3->
-%     ]
-%   \node[treenode] (tree) {$\otimes$}
-%     child { node[treenode] (a) {${\color{blue} a}$} }
-%     child { node[treenode] (b){${\color{magenta} \mathsf{inl}}$}
-%         child { node[treenode] (c) {${\color{magenta} b}$} }
-%       };
-%       \node[draw, fit=(tree)(a)(b)(c),
-%       label={[name=l] $g(ab)$}, visible on=<3->] {};
-%     \node[below=2.5cm of tree] (mt) {};
-%     \node[below=2.5cm of a] (mta) {};
-%     \node[below=2.5cm of b] (mtb) {};
-%     \node[below=-0.2cm of mt, visible on=<4->] (mtset) {};
-%       \node[draw, fit=(mt)(mtset)(mta)(mtb),
-%       label={[name=l, visible on=<4->] $g(aa)$}, visible on=<4->] {};
-% \end{tikzpicture}
-
-%% 1. First example: finite grammars (regex minus Kleene star)
-%%    - why ordered linear:
-%%      + no exchange
-%%      + no contraction
-%%      + no weakening
-%%      + ambiguity
-%% 2. Second example: Kleene star
+\paragraph{Non-deterministic Finite Automata}
+As a final example, consider the non-deterministic finite
+automaton (NFA) $N$ from \cref{fig:exampleNFA}. Such automata play a critical
+role, as they serve as operationalized and
+executable presentation of a formal grammar. Therefore, it is crucial that we
+can represent them internal to \theoryabbv.
+
+Our method of representing these automata internally is to provide an \emph{indexed}
+inductive type of traces through the machine. For $N$, this indexed inductive
+type is given in \cref{fig:nfainductive}.
+
+The type of traces in $N$ is indexed by two pieces of data: a state in $N$ and a
+Boolean. For each state $s : N.\mathsf{state}$ and $acc : \Bool$, we define the
+type of traces from state $s$, $\Trace_{N}~s~acc$. The Boolean flag $acc$ marks if the trace
+ends in either an accepting or a rejecting state. Let us dissect the structure
+of this type in more detail for state 1 of $N$.
+
+$\Trace_{N}~1~acc$ has three constructors. The first constructor
+$\mathsf{stop}_{1}$ allows a trace to end at state one, and because state
+1 is accepting $acc$ is true when terminating at state 1.
+
+The constructor
+$\mathsf{1to2}$ denotes the $b$-labeled transition in $N$ from state 1 to state
+2. The character $b$ is consumed when taking this transition and then a trace
+beginning at state 2 must be produced, which is denoted by the tensor product of
+the grammar $b$ and $Trace_{N}~2~acc$. Finally, the constructor $\mathsf{1to2}$
+abstracts over the acceptance Boolean because it is not yet determined.
+Because the Boolean is set by the acceptance criteria of the \emph{final} state,
+it is provided inductively by the trace out of state 2.
+
+The constructor $\mathsf{1to3}$ is built quite similarly as $\mathsf{1to2}$;
+however, $\mathsf{1to3}$ does not involve a tensor product. This is because the
+transition from state 1 to state 3 is labeled by the empty string $\epsilon$, so
+the transition consumes no input. In this case, when building a trace out of
+state 1, it suffices to build a trace out of state 3 \emph{over the same string}.
+
+\begin{figure}
+  \begin{tikzpicture}[node distance = 25mm ]
+    \node[state, initial, accepting] (1) {$1$};
+    \node[state, below left of=1] (2) {$2$};
+    \node[state, right of=2] (3) {$3$};
+
+    \path[->] (1) edge[above] node{$b$} (2)
+              (1) edge[below, bend right, left=0.3] node{$\epsilon$} (3)
+              (2) edge[loop left] node{$a$} (2)
+              (2) edge[below] node{$b$} (3)
+              (3) edge[above, bend right, right=0.3] node{$a$} (1);
+  \end{tikzpicture}
+  \caption{NFA $N$}
+  \label{fig:exampleNFA}
+\end{figure}
+
+\begin{figure}
+\begin{align*}
+\data &~\Trace_{N}~1~(acc : \Bool) : L_{i}~\where\\
+      & \mathsf{stop}_{1} : I \lto \Trace_{N}~1~\true \\
+      & \mathsf{1to2} : \forall (acc : \Bool) \to b \otimes \Trace_{N}~2~acc \lto \Trace_{N}~1~acc \\
+      & \mathsf{1to3} : \forall (acc : \Bool) \to \Trace_{N}~3~acc \lto \Trace_{N}~1~acc \\
+\\
+\data &~\Trace_{N}~2~(acc : \Bool) : L_{i}~\where\\
+      & \mathsf{stop}_{2} : I \lto \Trace_{N}~2~\false \\
+      & \mathsf{2to2} : \forall (acc : \Bool) \to a \otimes \Trace_{N}~2~acc \lto \Trace_{N}~2~acc \\
+      & \mathsf{2to3} : \forall (acc : \Bool) \to b \otimes \Trace_{N}~3~acc \lto \Trace_{N}~2~acc \\
+\\
+\data &~\Trace_{N}~3~(acc : \Bool) : L_{i}~\where\\
+      & \mathsf{stop}_{3} : I \lto \Trace_{N}~3~\true \\
+      & \mathsf{3to1} : \forall (acc : \Bool) \to a \otimes \Trace_{N}~1~acc \lto \Trace_{N}~3~acc \\
+\end{align*}
+\caption{$N$ as an indexed inductive type}
+\label{fig:nfainductive}
+\end{figure}
+
 %% 3. Third example: NFA traces to show the indexed inductive types
@@
 \label{sec:applications}
+
+\steven{this entire section should become much more compact with the examples
+  section from the beginning in mind}
 

### 2024-10-25 paper/paper.tex
fix broken figure, remove some outdated comments

@@
 
+Overall, leveraging \emph{indexed} inductive types provides a concise interface
+for defineing a family
+of mutually recursive types over the indexing data.
+
 \begin{figure}
@@
 
-\steven{This figure is broken}
 \begin{figure}
-  % \begin{array}{rrcl}
-  %   \textrm{non-linear contexts} & \Gamma & ::= & \cdot \pipe \Gamma,x:X \\
-  %   \textrm{non-linear types} & X , Y , Z & ::= & \mathsf{Nat} \pipe
-  %                                                 \mathsf{Int} \pipe
-  %                                                 \mathsf{Empty} \pipe
-  %                                                 \mathsf{Unit} \pipe
-  %                                                 \mathsf{SPF}\,X\pipe U \pipe L
-  %                                                 \pipe !A \pipe \sum_{x:X} Y
-  %                                                 \pipe \prod_{x:X} Y\pipe M =_A
-  %                                                 N \\
-  %   \textrm{linear contexts} & \Delta & ::= & \cdot \pipe \Delta , a:A\\
-  %   \textrm{linear types} & A , B , C & ::= &
-  %   c \pipe I \pipe A \otimes B \pipe \bigoplus\limits_{x:X} A \pipe \mu A_f M
-  %   A \lto B \pipe B \tol A \pipe \bigwith\limits_{x:X} A \pipe \equalizer {a}{f}{g}\\
-  % \end{array}
+  \[
+  \begin{array}{rrcl}
+    \textrm{non-linear contexts} & \Gamma & ::= & \cdot \pipe \Gamma,x:X \\
+    \textrm{non-linear types} & X , Y , Z & ::= & \mathsf{Nat} \pipe
+                                                  \mathsf{Int} \pipe
+                                                  \mathsf{Empty} \pipe
+                                                  \mathsf{Unit} \pipe
+                                                  \mathsf{SPF}\,X\pipe U \pipe L
+                                                  \pipe !A \pipe \sum_{x:X} Y
+                                                  \pipe \prod_{x:X} Y\pipe M =_A
+                                                  N \\
+    \textrm{linear contexts} & \Delta & ::= & \cdot \pipe \Delta , a:A\\
+    \textrm{linear types} & A , B , C & ::= &
+    c \pipe I \pipe A \otimes B \pipe \bigoplus\limits_{x:X} A \pipe \mu A_f M
+    A \lto B \pipe B \tol A \pipe \bigwith\limits_{x:X} A \pipe \equalizer {a}{f}{g}\\
+  \end{array}
+  \]
   \caption{Overview of Types}
@@
 Freely duplicating or dropping characters from a string changes the meaning of that string.
-\max{what about a backtracking parser?} \steven{Does the sentence sound better now?
-I've tried to avoid suggesting any particular parser design.
-Also, I think backtracking is a very interesting question to talk about in future work}.
+\max{what about a backtracking parser?}.
 One may then envision a linear type system where the types comprise
@@
 
-Unfortunately, programming in a linear term language is often
-cumbersome and unintuitive for users who are unaccustomed to the usage
-restrictions.\max{You are arguing against our own type theory here...} \steven{I
-  want to argue against a \emph{purely linear} calculus to motivate the
-  convenience of LNL, but perhaps this isn't appropriate here, and I don't want to
-  make our own type theory sound bad.Should probably just cut the beginning of
-  this paragraph and jump into LNL, but then there's not an immediate reason for
-  why one might want/need it over a pure linear system}
-\pedro{I think that it's reasonable to simply say something like ``programming in a purely linear term language
-is limiting due to variable usage restrictions''. In this case we're not throwing our own TT under the bus}
-Code in such a language can
-become unnecessarily verbose when ensuring every function written is indeed linear. It would be more ergonomic to allow users using Girard encodings to translate
-high-level reasoning into low-level linear terms.\pedro{What encodings are these?
-We should add a citation here}
+Programming in a purely linear term language is limiting due to the variable
+usage restrictions. Code in such a language can
+become unnecessarily verbose when ensuring the linearity invariant.
 To alleviate this
@@
 internal to \theoryabbv.
-
-\steven{The above is meant to walk the reader into accepting LNL as a natural
-  choice, but it doesn't really flow well to the beginning of the next subsection }
 

### 2024-10-25 paper/paper.tex
scaffolding for properties/semantic actions

@@
 \newcommand{\tol}{\mathrel{\rotatebox[origin=c]{180}{$\lto$}}}
-\newcommand{\String}{\Sigma^{*}}
+\newcommand{\String}{\textbf{String}}
 \newcommand{\Set}{\mathbf{Set}}
@@
 $\Sigma$, which is classically defined as a
-\emph{subset} of strings $L \subseteq \String$. The expressivity of formal
+\emph{subset} of strings $L \subseteq \String = \Sigma^{*}$. The expressivity of formal
 languages is sufficient for defining a \emph{recognizer} of a language $L$:
@@
 
-\pedro{Hopefully this serves as a segue between this and the syntax subsection}
 \subsection{Syntax}
@@
 
-\section{Formal Grammar Theory in Lambek}
+\section{Formal Grammar Theory in \theoryname}
 \label{sec:applications}
 
-\steven{this entire section should become much more compact with the examples
+\steven{below this is unedited word vomit}
+\steven{this entire ``Formal Grammar Theory in Lambek'' section should become much more compact with the examples
   section from the beginning in mind}
 
+\subsection{Grammar Properties}
+\steven{Mention that $\top$ and $\String$ are iso. Highlight this as a needed axiom}
+
+%% 1. Basic concepts: unambiguity, parseability, weak and strong equivalence
+\paragraph{Unambiguity}
+In the grammars model, we say that a grammar is unambiguous if there is at most
+one parse tree for any input string. There are several ways to capture this
+concept internally.
+
+In the presence of distributive coproducts, the following are equivalent:
+\begin{itemize}
+  \item $A$ is an unambiguous grammar
+  \item The morphism $! : A \to \top$ is a monomorphism
+  \item Given a grammar $B$, all morphisms $B \to A$ are equal
+  \item $A \& A \cong A$
+\end{itemize}
+
+\paragraph{Parseability}
+We say $A$ is \emph{weakly parseable} if there exists a type $B$ and a map
+$\top \to A \oplus B$. Because $\top \cong \String$, this says that any string
+can be parsed into either an $A$ or a $B$.
+
+If $\top \cong A \oplus B$, call $A$ \emph{parseable}. This notion of
+parseability implies that $A$ and $B$ are each unambiguous.
+
+\paragraph{Decidability}
+In the case that $A$ is parseable with respect to $\neg A$ --- i.e.
+$\top \cong A \oplus \neg A$ --- then call $A$ \emph{decidable}.
+
+\paragraph{Equivalence}
+Grammars $A$ and $B$ are \emph{weakly equivalent} if there exists maps
+$f : A \to B$ and $g : B \to A$.
+
+If $f$ and $g$ form an isomorphsim, then $A$ and $B$ are \emph{strongly
+  equivalent}.
+\steven{TODO cite that these are the same as Chomsky's notions of equivalence}
+
+\subsection{Regular Expressions and Finite Automata}
+%% 2. Regex, NFA, DFA
+
+\subsection{The Dyck Grammar}
+%% 3. Dyck language
+The Dyck grammar $D$ for balanced parens (summarize what it is + citation)
+
+\steven{this is a VERY loose sketch of the things I want to say about $D$}
+\Cref{fig:DyckAutomaton} presents an automaton $M$ that recognizes $D$. We show this
+precisely by constructing an isomorphism between $D$ and $\Trace_{M}~0~\true$.
+Unlike the finite automata seen thus far, $M$ has infinitely many states --- one
+for each $n : \mathbb{N}$ with an additional fail state. This diverges a bit
+from classical language theory, where this machine would be encoded as a
+pushdown automaton with finitely many
+states and stack of memory. Such PDAs are encodable in our theory, although they
+are a little cumbersome to use. Moreover, they
+seem overfit to classical reasoning as they don't preserve the gramamr up to
+strong equivalence, only weak equivalnce in general. \steven{cite lambek
+  automata and state that is is a secondary goal to preserve as much structure as possible}
+By moving the stack of memory (the nats) into the
+state, we obtain infinitely many states, which is fine bc the states need only
+be a type for us to interface with it.
+
+\steven{TODO sketch proof of equivalence. Mention that it is mechanized}
+
+\begin{figure}
+  \begin{tikzpicture}[node distance = 17mm ]
+    \node[state, initial, accepting] (0) {$0$};
+    \node[state, right of=0] (1) {$1$};
+    \node[state, right of=1] (2) {$2$};
+    \node[right of=2] (3) {$\dots$};
+    \node[state, below of=0] (fail) {$\mathsf{fail}$};
+
+    \path[->] (0) edge[above, bend left] node{\stringquote{(}} (1)
+              (1) edge[below, bend left] node{\stringquote{)}} (0)
+              (1) edge[above, bend left] node{\stringquote{(}} (2)
+              (2) edge[below, bend left] node{\stringquote{)}} (1)
+              (2) edge[above, bend left] node{\stringquote{(}} (3)
+              (3) edge[below, bend left] node{\stringquote{)}} (2)
+              (fail) edge[loop right] node{\stringquote{(}, \stringquote{)}} (fail)
+              (0) edge[left] node{\stringquote{)}} (fail);
+  \end{tikzpicture}
+  \caption{Automaton $M$ for the Dyck grammar}
+  \label{fig:DyckAutomaton}
+\end{figure}
+
+\subsection{LL(1)}
+%% 4. LL(1) parser
+
+\steven{TODO impute LL(1) example, mention that its weak equivalence is mechanized}
+
+\subsection{UnReStRiCtEd}
+%% 5. Expressiveness: Turing Machines
+
 %% Make sure to say which results are mechanized
-%% 1. Basic concepts: unambiguity, parseability, weak and strong equivalence
-%% 2. Regex, NFA, DFA
-%% 3. Dyck language
-%% 4. LL(1) parser
-%% 5. Expressiveness: Turing Machines
-
+
+\subsection{Semantic Actions}
+\steven{TODO edit this. synthesize it with the later presentation of sematnic
+  actions as models, and then remove the model section}
+All constructions discussed until now detail how a string may be parsed into a
+\emph{concrete} syntax tree; however, this is often not the representation that later
+compiler passes would take in. Instead, they often take in an \emph{abstract}
+syntax tree that forgets syntactic details that are superfluous to the later
+passes. For example, after using parentheses to disambiguate a parse of a string
+into a concrete syntax tree, we may remove these parentheses to have a cleaner
+structured representation of the underlying data. The procedure that takes in a
+concrete syntax tree and returns an abstract one is referred to as a
+\emph{semantic action}.
+
+Given a linear type $A$, the inhaitants of $A$ serve as concrete syntax. We may
+then chose a non-linear type $X$ that serves as abstract syntax. A semantic
+action can then be defined as $\square (A \lto \Delta X)$ where
+$\Delta X = \LinSigTy x X \top$.
+\steven{Note here that even though $\Delta$ is a map from nonlinear types into
+  linear ones, it is very different from $F$/whatever we call that half of the adjunction}
+
+\steven{Everything in this section below this marker is outdated and needs to be
+folded into the above or cut}
+
+\steven{Need a different segue because canonicity is no longer before this}
 The canonicity theorem from the previous section gives a tight connection
@@
 \end{itemize}
-
-Intuitively, this can be thought of like a directed graph with nodes in $Q$ with
-an edge $q \overset{c}{\to} q'$ whenever $q' \in \delta(q, c)$. Note that
-transitions in an NFA may be labeled with the empty string $\varepsilon$ --- such
-transitions are referred to as \emph{$\varepsilon$-transitions}. We may see an
-example of an NFA in \cref{fig:NFA}.
-
-From an NFA, we may construct a grammar of traces as follows:
-
-First, we define a mutual fixed point grammar that describes the traces through the NFA.\ Then, we have another
-grammar that tells us if we're currently in an accepting state of the automaton.
-A parse of the NFA grammar is then a pair of a trace and the data that we're in
-an accepting state.
-
-Consider the NFA $N$ pictured in \cref{fig:NFA}. There are three states, $1$, $2$,
-and $3$. We introduce a non-linear type $Q$ with three inhabitants $q_{1}$,
-$q_{2}$, and $q_{3}$ to represent each of these states, respectively. Given a
-$q, q'$, we can
-then define the type of traces from $q$ to $q'$. For instance, let's construct
-the traces starting at $q_{1}$ and ending at $q_{2}$ as an example.
-
-\begin{equation}
-  \label{eq:nfatrace}
-  \mathsf{Trace}_{N}(q_{1}, q_{2}) = \mu
-  \begin{pmatrix}
-    g_{q_{1}} := g_{q_{3}} \oplus ( b \otimes g_{q_{2}} ) \\
-    g_{q_{2}} := ( a \otimes g_{q_{2}} ) \oplus ( a \otimes g_{q_{3}} ) \oplus ( b \otimes g_{q_{3}} ) \oplus I \\
-    g_{q_{3}} := a \otimes g_{q_{1}} \\
-  \end{pmatrix}. g_{q_{1}}
-\end{equation}
-
-We should read this as defining three mutually recursive grammars, one for each
-state. The definitions of these mutually recursive grammars capture the
-transitions of the automaton. To ensure that we only encode traces that end in
-state 2, we only include the unit $I$ as a summand in
-$g_{q_{2}}$. That is, by only including $I$ at this location, it makes it the
-only place where derivations of $\mathsf{Trace}(q_{1}, q_{2})$ can terminate. We
-can think of all of these definitions underneath
-of the $\mu$ binder as bringing some local grammars into scope. With these local
-grammars in scope we are ultimately constructing a term of type $g_{q_{1}}$ to denote
-only traces starting in state 1.
-
-To encode the acceptance criteria of $N$, we want to internalize a proposition
-over each state of the NFA.\ That is, for each $q \in Q$ we define a term
-
-\[
-  \mathsf{acc}(q) := q \text{ is accepting}
-\]
-
-In this example, $\mathsf{acc}(q_{1}) = \top$ and
-$\mathsf{acc}(q_{2}) = \mathsf{acc}(q_{3}) = \bot$. An accepting trace of the NFA $N$ is then given by the following dependent grammar,
-\[
- \mathsf{AccTrace}_{N} := \LinSigTy q Q {\left( \mathsf{Trace}_{N}(q_{0} , q) \pair \mathsf{acc}(q) \right)}
-\]
-
-where $q_{0}$ is the initial state. That is, a trace is accepted by the NFA if
-we can construct the trace, and the trace ends at an accepting state. This idea
-is simple enough and aligns with how we intuitively treat these automata.
-
-Generalizing over the above example, we want to define the type of traces as,
-
-\[
-  \mathsf{Trace}_{N}(q_{0}, q_{1}) = \mu
-  \begin{pmatrix}
-    g_{q} := \mathsf{Trans}(q), & q \neq q_{1} \\
-    g_{q} := \mathsf{Trans}(q) \oplus I , & q = q_{1}
-  \end{pmatrix}. g_{q_{0}}
-\]
-
-where $\mathsf{Trans}(q)$ is an iterated disjunction that describes which
-transition you should take. For an NFA, $\mathsf{Trans}(q)$ must take on the
-following syntactic form,
-
-\begin{gather*}
- \mathsf{State} \in \{ g_{q} : q \in Q \} \\
- \mathsf{Char} \in \Sigma \\
- \mathsf{Trans}_{N}(q) ::= \mathsf{Char} \otimes \mathsf{State}~|~\mathsf{State}~|~\mathsf{Trans}_{N}(q) \oplus \mathsf{Trans}_{N}(q)
-\end{gather*}
-
-That is, $\mathsf{Trans}(q)$ is a disjunction of literals followed by grammars
-that encode states.
-
-When conducting proofs involving NFA grammars, we often to
-need either construct terms of type $\mathsf{Trace}_{N}(q , q')$. To
-this end, we give three admissible rules for constructing
-traces,
-
-\begin{figure}[h!]
-  \label{fig:admissibleintro}
-  \begin{mathpar}
-    \inferrule
-    {~}
-    {\Gamma ; \cdot \vdash \mathsf{nil} : \mathsf{Trace}_N
-      (q , q)}
-
-    \and
-
-    \inferrule
-    {\Gamma ; \Delta \vdash M : \mathsf{Trace}_N
-      (dst , q') \\
-    \exists \text{ transition } src \overset{c}{\to} dst
-    }
-    {\Gamma ; x : c , \Delta \vdash \mathsf{cons}(M) : \mathsf{Trace}_N
-    (src , q')}
-
-    \and
-
-    \inferrule
-    {\Gamma ; \Delta \vdash M : \mathsf{Trace}_N (dst , q')
-      \\
-    \exists~\varepsilon\text{-transition } src
-    \overset{\varepsilon}{\to} dst}
-    {\Gamma ; \Delta \vdash \mathsf{\varepsilon cons}(M) : \mathsf{Trace}_N
-    (src , q')}
-  \end{mathpar}
-  \caption{Admissible Trace Constructors}
-\end{figure}
-
-That is, we may use $\mathsf{nil}$ to terminate a trace that
-begins and ends at state $q$. The rules $\mathsf{cons}$ and
-$\mathsf{\varepsilon cons}$ are then used to inductively
-stitch traces together when sound. For instance, we can read
-the $\mathsf{cons}$ rule as saying that we can
-create a trace coming from a state $src$ provided that we may first
-transition via the character $c$ to state $dst$ and
-inductively build a trace starting from $dst$. The
-$\mathsf{\varepsilon cons}$ rule says something similar, but
-with an $\varepsilon$-transition instead of a transition
-labelled by a character. Recall that
-these rules are \emph{admissible}. They are not strictly
-necessary as primitives to conduct our proofs; however, they
-do provide convenient shorthand notation for building terms
-of type $\mathsf{Trace}_{N}(q , q')$.
-
-Dual to constructing traces, we often want to construct
-other terms in a context containing values of type
-$\mathsf{Trace}_{N}(q , q')$. For this purpose, we make use
-of the elimination principle for multiple-fixed points ---
-which we write as $\mathsf{mfold}$ --- given
-in \cref{fig:multifix}.
-
-Note that this general construction will readily generalize to other types of
-automata. If we wanted to define say deterministic finite automata\footnote{DFAs
-\emph{could} just be defined as NFAs that happen to be deterministic. This is
-one way to do so, or you may choose to present the transition relation for the
-automaton as a transition \emph{function} instead. Concerns like these don't
-matter so much when defining things on paper, but at formalization time these
-are important and can make some proofs much easier.}, pushdown automata, Turing
-machines, etc, we just swap out the type of traces for a different, but very
-similar, trace construction.
-
-With this setting for finite automata, we can now internalize some classical theorems inside of our formal system.
 
@@
 \subsection{Semantic Actions}
+\steven{Tentatively planning to cut this subsection for an internal representation
+  of semantic actions}
 Returning to the problem of parsing, the output of a parse usually is not the
@@
 
+\steven{TODO add lambek calculus/categorial grammars here}
+
 \paragraph{Kleene Algebra}

### 2024-10-25 paper/paper.tex
change syntax for the linear to the nonlinear functor
add todos

@@
 \newcommand{\bang}{~\textbf{!}~}
+% Tentatively using uparrow for linear to non-linear to be in line with
+% Pfennings adjoint functional programming
+\newcommand{\ltonl}[1]{~\uparrow #1}
 \newcommand{\nil}{\mathsf{nil}}
@@
 
-However, we can have derivations where the term in context is not simply a
+We may also have derivations where the term in context is not simply a
 string of base types. For instance, by using the elimination principle for
 Kleene star, \cref{fig:kleeneabstractproof} builds a term of type
-$A^{*}$ in context comprising abstract linear types. In these contexts, it is
+$A^{*}$ in a context comprising abstract linear types. In these contexts, it is
 not quite appropriate
@@
 \steven{TODO : describe elimination principle somewhere. Maybe just in prose?}
-\steven{I probably can't get away with the inline definition of $f$ below}
+\steven{I probably can't get away with the inline definition of $f$ in \cref{fig:kleeneabstractproof}}
 \begin{figure}
@@
                                                   \mathsf{SPF}\,X\pipe U \pipe L
-                                                  \pipe !A \pipe \sum_{x:X} Y
+                                                  \pipe \ltonl A \pipe \sum_{x:X} Y
                                                   \pipe \prod_{x:X} Y\pipe M =_A
@@
     {~}
-    {\Gamma , x : \square A ; \cdot \vdash \bang x : A}
-    %
-    \and
-    %
-    \inferrule
-    {\Gamma \vdash f : \square (A \lto B) \\
-      \Gamma \vdash g : \square (A \lto B) \\
+    {\Gamma , x : \ltonl A ; \cdot \vdash x : A}
+    %
+    \and
+    %
+    \inferrule
+    {\Gamma \vdash f : \ltonl {(A \lto B)} \\
+      \Gamma \vdash g : \ltonl { (A \lto B) } \\
       \Gamma ; \Delta \vdash a : A \\
-      \Gamma ; \Delta \vdash p : \apptol {\bang f}{a} = \apptol {\bang g}{a}}
-    {\Gamma ; \Delta \vdash \equalizerpair{a}{p} : \equalizer {a}{\bang f}{\bang
-      g}}
-    %
-    \and
-    %
-    \inferrule
-    {\Gamma \vdash f : \square (A \lto B) \\
-      \Gamma \vdash g : \square (A \lto B) \\
-      \Gamma ; \Delta \vdash e : \equalizer{a}{\bang f}{\bang g}}
+      \Gamma ; \Delta \vdash p : \apptol {f}{a} = \apptol {g}{a}}
+    {\Gamma ; \Delta \vdash \equalizerpair{a}{p} : \equalizer {a}{f}{g}}
+    %
+    \and
+    %
+    \inferrule
+    {\Gamma \vdash f : \ltonl { (A \lto B) } \\
+      \Gamma \vdash g : \ltonl { (A \lto B) } \\
+      \Gamma ; \Delta \vdash e : \equalizer{a}{f}{g}}
     {\Gamma ; \Delta \vdash \equalizerpi {e} : A}
@@
 non-linear subtheories is entirely characterized via a monoidal
-adjunction $F \dashv G$ between linear terms and non-linear terms.
+adjunction between linear terms and non-linear terms.
+
+\steven{Describe this monoidal adjunction with $\ltonl A$ and the derivable
+  version of $F$}
 
@@
 is disallowed; however, through dependence of a linear term on a non-linear
-index, we recover the definition of Aho's indexed grammars \cite{AhoIndexed}
-internal to \theoryabbv.
+index, we can recover the definition of Aho's indexed grammars \cite{AhoIndexed}
+internal to \theoryabbv. Although, we can define strictly more things, as Aho's
+indexed grammars provide restricted access to the index whereas we can do
+whatever with it.
+
+\steven{elaborate on indexed grammars in a more articulate way in the last sentence}
+\steven{Put the aho stuff in the subtheory section}
 
@@
 where only well-typed terms are meaningful. Next, we show how it admits a model in $\Set^{\String}$
-where the linear types have natural interpretation in terms of formal languages --- for instance,
+where the linear types have natural interpretation in terms of formal grammars --- for instance,
 the tensor corresponds to grammar concatenation. We conclude the section by showing how
@@
 %
-    \and
-%
-    \inferrule{\Gamma \vdash A : L_i}{\Gamma \vdash G A : U_i}
+    % \and
+%
+    % \inferrule{\Gamma \vdash A : L_i}{\Gamma \vdash G A : U_i}
 
@@
     {\Gamma \vdash A : L_i}
-    {\Gamma \vdash \square A : U_i}
-    %
-    \and
-    \inferrule
-    {\Gamma ; \Delta \vdash f : A \\ \Gamma \vdash g : \square (A \lto B)}
-    {\Gamma ; \Delta \vdash \equalizer {a}{\bang f}{\bang g} : L_i}
+    {\Gamma \vdash \ltonl { A } : U_i}
+    %
+    \and
+    \inferrule
+    {\Gamma ; \Delta \vdash f : A \\ \Gamma \vdash g : \ltonl { (A \lto B) }}
+    {\Gamma ; \Delta \vdash \equalizer {a}{f}{g} : L_i}
   \end{mathpar}
@@
   \and
-  \inferrule{\Gamma ; \cdot \vdash e : A}{\Gamma \vdash \mathsf G e : \mathsf G A}
+  \inferrule{\Gamma ; \cdot \vdash e : A}{\Gamma \vdash e : \ltonl A}
 
@@
 %
-    \and
-%
-    \inferrule{~}{\Gamma \vdash G\, (G^{-1} \, t) \equiv t : G A}
+    % \and
+%
+    % \inferrule{~}{\Gamma \vdash G\, (G^{-1} \, t) \equiv t : G A}
 
@@
 
-    \inferrule{~}{\Gamma; \cdot \vdash G^{-1}\, (G \, t ) \equiv t: A}
-%
-    \\
+    % \inferrule{~}{\Gamma; \cdot \vdash G^{-1}\, (G \, t ) \equiv t: A}
+%
 %
@@
     \inferrule
-    {\Gamma \vdash f : \square (A \lto B) \\
-      \Gamma \vdash g : \square (A \lto B) \\
+    {\Gamma \vdash f : \ltonl { (A \lto B) } \\
+      \Gamma \vdash g : \ltonl { (A \lto B) } \\
       \Gamma ; \Delta \vdash a : A \\
-      \Gamma ; \Delta \vdash p : \apptol {\bang f}{a} \equiv \apptol {\bang g}{a}}
+      \Gamma ; \Delta \vdash p : \apptol {f}{a} \equiv \apptol {g}{a}}
     {\Gamma ; \Delta \vdash \equalizerpi {\equalizerpair {a}{p}} \equiv a : A}
@@
     \inferrule
-    {\Gamma \vdash f : \square (A \lto B) \\
-      \Gamma \vdash g : \square (A \lto B) \\
-      \Gamma ; \Delta \vdash p : \apptol {\bang f}{a} = \apptol {\bang g}{a} \\
-      \Gamma ; \Delta \vdash e : \equalizer{a}{\bang f}{\bang g}}
+    {\Gamma \vdash f : \ltonl { (A \lto B) } \\
+      \Gamma \vdash g : \ltonl { (A \lto B) } \\
+      \Gamma ; \Delta \vdash p : \apptol {f}{a} = \apptol {g}{a} \\
+      \Gamma ; \Delta \vdash e : \equalizer{a}{f}{g}}
     {\Gamma ; \Delta \vdash \equalizerpair {\equalizerpi {e}} {p} \equiv e : A}
@@
 \end{figure}
+\steven{Do we need inverse for $\ltonl A$ in judgemental equality figure?}
+\steven{Do we need to say anything about the equalizer proofs being unique in
+  judgemental equality figure?}
 
@@
 \paragraph{LNL Dependent Types}
+\steven{talk about how binary sum and binary with are definable over $\Bool$}
 Given a non-linear type $X$, we may form both the dependent product type
@@
 \paragraph{Universal Type}
+\steven{Definable as nullary product}
 The universal type $\top$ may be formed in any context.
@@
 \paragraph{Empty Type}
+\steven{Definable as nullary sum}
 The empty type $\bot$ has no inhabitants. The elimination for the empty type
@@
 \paragraph{The $G$ Modality}
+\steven{Rename this}
+\steven{Compare to persistance modalities from separation logic}
 The $G$ modality provides the embedding of linear types in the empty context
@@
 %
-\steven{Elaborate on $G$ more}
+\steven{Elaborate on $G$ more and rename $G A$to $\ltonl A$ in this section}
 
@@
 \paragraph{Recursive Types}
+\steven{reflavor this section as ``indexed inductive'' instead of ``recursive'',
+and entirely rework with the new inductive definition}
+\steven{The non-indexed inductive are definable as being trivially indexed over unit}
 Recursive linear types may be defined via the least-fixed point operator $\mu$.
@@
   fixed-point}
-\steven{As a consequence of Knaster-Tarski}
 
@@
 
+\steven{Aho's indexed grammars to be moved into here to reflect on yet another
+  grammar formalism that is internally definable}
+
 \section{Formal Grammar Theory in \theoryname}
@@
 %% 2. Regex, NFA, DFA
+\steven{How much is worth explicitly walking through? The examples from earlier
+  give motivation, and we are pressed for space}
+
+\paragraph{Regex-NFA/Thompson}
+\steven{TODO put regex-NFA iso here. Call it Thompson's constructon?. Mention
+  its mechanized}
+
+\paragraph{Determinization of an NFA}
+\steven{TODO put determinzation here. Mention its mechanized}
 
@@
 The Dyck grammar $D$ for balanced parens (summarize what it is + citation)
+
+\steven{somewhere in here, sneak in the benchmark where the dyck parser could
+  handle 10's of thousands of characters in about 30s}
 
@@
 
+\steven{Because we can define a version of semantic actions internally, we
+  shouldn't put it as a separate model}
+
 One of the powers of type theories is that it can be profitable to interpret them
@@
 
-\subsection{Semantic Actions}
-\steven{Tentatively planning to cut this subsection for an internal representation
-  of semantic actions}
-Returning to the problem of parsing, the output of a parse usually is not the
-literal parse tree. Rather, the output is the result of some \emph{semantic
-  actions} ran on the parse tree, which usually serve to remove some syntactic
-details that are unnecessary for later processing.
-
-Given a grammar $G : \String \to \Set$, we define a semantic action to be a set
-$X$ with a function $f$ that produces a semantic element from any parse of $G$.
-
-\[
-  f : \PiTy w \String {G w \to X}
-\]
-
-Further, semantic actions can be arranged into a structured category.
-Define $\SemAct$ as the comma
-category $\Grammar / \Delta$, where $\Delta : \Set \to \Grammar$ defines a
-discrete presheaf. That is, for a set $X$, $\Delta (X)(w) = X$ for all
-$w \in \String$. As $\SemAct$ is defined as a comma category, it has a forgetful
-functor into $\Grammar$. That is, $\SemAct$ serves as a notion of formal
-grammar. Moreover, $\SemAct$ is a model of \theoryabbv.
-
-\steven{It being a notion of formal grammar is distinct from being a model. This
-probably warrants a proof}
-
-\steven{TODO semantic actions}
-
-\pedro{This is a very nice opportunity of showing off the supremacy of denotational
-  reasoning ;) We should probably prove the gluing lemma in the previous section
-  and apply it here and in the canonicity section. The actual proof might have to be
-  moved to the appendix, though}
+% \subsection{Semantic Actions}
+% \steven{Tentatively planning to cut this subsection for an internal representation
+%   of semantic actions}
+% Returning to the problem of parsing, the output of a parse usually is not the
+% literal parse tree. Rather, the output is the result of some \emph{semantic
+%   actions} ran on the parse tree, which usually serve to remove some syntactic
+% details that are unnecessary for later processing.
+
+% Given a grammar $G : \String \to \Set$, we define a semantic action to be a set
+% $X$ with a function $f$ that produces a semantic element from any parse of $G$.
+
+% \[
+%   f : \PiTy w \String {G w \to X}
+% \]
+
+% Further, semantic actions can be arranged into a structured category.
+% Define $\SemAct$ as the comma
+% category $\Grammar / \Delta$, where $\Delta : \Set \to \Grammar$ defines a
+% discrete presheaf. That is, for a set $X$, $\Delta (X)(w) = X$ for all
+% $w \in \String$. As $\SemAct$ is defined as a comma category, it has a forgetful
+% functor into $\Grammar$. That is, $\SemAct$ serves as a notion of formal
+% grammar. Moreover, $\SemAct$ is a model of \theoryabbv.
+
+% \steven{It being a notion of formal grammar is distinct from being a model. This
+% probably warrants a proof}
+
+% \steven{TODO semantic actions}
+
+% \pedro{This is a very nice opportunity of showing off the supremacy of denotational
+%   reasoning ;) We should probably prove the gluing lemma in the previous section
+%   and apply it here and in the canonicity section. The actual proof might have to be
+%   moved to the appendix, though}
 

### 2024-10-28 paper/paper.tex
flip linear arrows

@@
 \data &~A^{*}: L_{i}~\where\\
-      & \nil : I \lto A^{*} \\
-      & \cons : A \otimes A^{*} \lto A^{*}
+      & \nil : \ltonl {A^{*}} \\
+      & \cons : \ltonl {(A \lto A^{*} \lto A^{*})}
 \end{align*}
@@
 \end{figure}
+
+\steven{TODO need to update the proof tree example derivations (\cref{fig:kleenestarderivation,fig:kleeneabstractproof}) to use the presentation of cons that is now in \cref{fig:kleenestarinductive}}
 
@@
 \data &~\Trace_{N}~1~(acc : \Bool) : L_{i}~\where\\
-      & \mathsf{stop}_{1} : I \lto \Trace_{N}~1~\true \\
-      & \mathsf{1to2} : \forall (acc : \Bool) \to b \otimes \Trace_{N}~2~acc \lto \Trace_{N}~1~acc \\
-      & \mathsf{1to3} : \forall (acc : \Bool) \to \Trace_{N}~3~acc \lto \Trace_{N}~1~acc \\
+      & \mathsf{stop}_{1} : \ltonl {\Trace_{N}~1~\true}  \\
+      & \mathsf{1to2} : \forall (acc : \Bool) \to \ltonl {(b \lto \Trace_{N}~2~acc \lto \Trace_{N}~1~acc)} \\
+      & \mathsf{1to3} : \forall (acc : \Bool) \to \ltonl {(\Trace_{N}~3~acc \lto \Trace_{N}~1~acc)} \\
 \\
 \data &~\Trace_{N}~2~(acc : \Bool) : L_{i}~\where\\
-      & \mathsf{stop}_{2} : I \lto \Trace_{N}~2~\false \\
-      & \mathsf{2to2} : \forall (acc : \Bool) \to a \otimes \Trace_{N}~2~acc \lto \Trace_{N}~2~acc \\
-      & \mathsf{2to3} : \forall (acc : \Bool) \to b \otimes \Trace_{N}~3~acc \lto \Trace_{N}~2~acc \\
+      & \mathsf{stop}_{2} : \ltonol {\Trace_{N}~2~\false} \\
+      & \mathsf{2to2} : \forall (acc : \Bool) \to \ltonl {(a \lto \Trace_{N}~2~acc \lto \Trace_{N}~2~acc)} \\
+      & \mathsf{2to3} : \forall (acc : \Bool) \to \ltonl {(b \lto \Trace_{N}~3~acc \lto \Trace_{N}~2~acc)}  \\
 \\
 \data &~\Trace_{N}~3~(acc : \Bool) : L_{i}~\where\\
-      & \mathsf{stop}_{3} : I \lto \Trace_{N}~3~\true \\
-      & \mathsf{3to1} : \forall (acc : \Bool) \to a \otimes \Trace_{N}~1~acc \lto \Trace_{N}~3~acc \\
+      & \mathsf{stop}_{3} : \ltonl {\Trace_{N}~3~\true} \\
+      & \mathsf{3to1} : \forall (acc : \Bool) \to \ltonl {(a \lto \Trace_{N}~1~acc \lto \Trace_{N}~3~acc)} \\
 \end{align*}
@@
     \textrm{linear types} & A , B , C & ::= &
-    c \pipe I \pipe A \otimes B \pipe \bigoplus\limits_{x:X} A \pipe \mu A_f M
+    c \pipe I \pipe A \otimes B \pipe \bigoplus\limits_{x:X} A \pipe \mu A_f M \pipe
     A \lto B \pipe B \tol A \pipe \bigwith\limits_{x:X} A \pipe \equalizer {a}{f}{g}\\
@@
     %
-    \inferrule{\Gamma ; a : A , \Delta \vdash e : B}{\Gamma ; \Delta \vdash \lamblto a e : A\lto B}
-    \and
-    \inferrule{\Gamma ; \Delta' \vdash e' : A \\ \Gamma ; \Delta \vdash e : A \lto B}{\Gamma ; \Delta', \Delta \vdash \applto {e'} {e} : B}
+    \inferrule{\Gamma ; \Delta , a : A \vdash e : B}{\Gamma ; \Delta \vdash \lamblto a e : A\lto B}
+    \and
+    \inferrule{\Gamma ; \Delta' \vdash e' : A \\ \Gamma ; \Delta \vdash e : A \lto B}{\Gamma ; \Delta, \Delta' \vdash \applto {e'} {e} : B}
     \\
     %
-    \inferrule{\Gamma ; \Delta , a : A \vdash e : B}{\Gamma ; \Delta \vdash \lambtol a e : B\tol A}
-    \and
-    \inferrule{\Gamma ; \Delta \vdash e : B \tol A \\ \Gamma ; \Delta' \vdash e' : A}{\Gamma ; \Delta, \Delta' \vdash \apptol e {e'} : B}
+    \inferrule{\Gamma ; a : A , \Delta \vdash e : B}{\Gamma ; \Delta \vdash \lambtol a e : B\tol A}
+    \and
+    \inferrule{\Gamma ; \Delta \vdash e : B \tol A \\ \Gamma ; \Delta' \vdash e' : A}{\Gamma ; \Delta', \Delta \vdash \apptol e {e'} : B}
     %
@@
 %
-    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lambtol a {\app e a} : A \tol B}
+    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lambtol a {\app e a} : B \tol A}
 %

### 2024-10-28 paper/paper.tex
comment out old bits in sec 4. edit the examples some

@@
 \newcommand{\stringquote}[1]{\texttt{\textquotesingle#1\textquotesingle}}
+\newcommand{\internalize}[1]{\lceil#1\rceil}
 
@@
 
-\newcommand{\equalizer}[3]{\{#1\,|\,\apptol {#2}{#1} = \apptol{#3}{#1} \}}
+\newcommand{\equalizer}[3]{\{#1\,|\,\applto {#2}{#1} = \applto{#3}{#1} \}}
 \newcommand{\equalizerpair}[2]{\langle #1 , #2 \rangle}
@@
 
+\steven{TODO need to go through and ensure that $\Sigma^{*}$ and $\String$ are each used appropriately}
+
 We say that a grammar $G$ associates to every string $w$ the set\footnote{We
@@
 As they are simply sets, formal languages are naturally endowed with a partial
-order via subset inclusion. Similarly, formal grammars naturally coalesce into a
-\emph{category}. Define $\Grammar$ to be the category of formal grammars. Whose
+order via subset inclusion. Just, formal grammars naturally coalesce into a
+\emph{category}.
+
+Define $\Grammar$ to be the category of formal grammars. Whose
 objects are grammars, and whose morphisms are functions $f : G \to H$ between two
-grammars presented by a family of functions
-
-\[ f^{w} : G w \to H w \]
-
-for every string $w \in \String$. Intuitively, we read a morphism of grammars as
-a \emph{parse transformer} --- i.e. $f$ translates a $G$-parse of $w$ into an $H$-parse
-of $w$. The induced notion of isomorphism in $\Grammar$ encodes that two grammars are equivalent if there is
-a bijective translation of parses. In other words, two grammars are isomorphic
-if they capture the same structural descriptions of strings, which is
-precisely Chomsky's notion of \emph{strong equivalence} of
-grammars \cite{chom1963}. In works like
-\cite{yoshinaga2002formal}, this notion of strong equivalence is often treated
-like the appropriate choice of ``sameness'' for two grammars. When abstracting over the latent
-structure of formal grammars and describing it in the the language of
-categories, it is encouraging to see the same definition of equivalence arise
-naturally.
-
-Upon inspection of the above definition, $\Grammar$ may equivalently
-be described as the functor category $\Set^{\String}$ --- where the
-set $\String$ is viewed as a discrete\footnote{Objects in $\String$
-comprise the set of all strings, and the only morphisms are the
-identity morphisms.}  category. In fact, in this paper we will take
-$\Grammar$ to be defined as the functor category $\Set^{\String}$ ---
-also called presheaf category. Such presheaf categories carry a
-remarkable amount of structure.  Indeed they are a monoidal topos and,
-as such, are a model to a substructural dependently typed theory.
+grammars presented by a family of functions $f^{w} : G w \to H w$ for
+$w : \String$. Intuitively, we interpret these morphisms as parse
+\emph{transformers} --- as constructive translations of $G$-parses into $H$-parses. Throughout this paper, we will make frequent reference to this categorical
+structure on $\Grammar$.
+
+\steven{Above sounds awkward. I want to additionally say something here that signposts that we will give more
+  information about that structure without just listing it all here. }
 
@@
 
-\max{this section kind of trails off at the end with a very vague aside about a $2$-category of grammar formalisms.
+Upon further inspection, $\Grammar$ may equivalently
+be described as the functor category $\Set^{\String}$ --- where the
+set $\String$ is viewed as a discrete\footnote{Objects in $\String$
+comprise the set of all strings, and the only morphisms are the
+identity morphisms.} category. In fact, in this paper we will take
+$\Grammar$ to be defined as the functor category $\Set^{\String}$. Such functor
+categories in $\Set$, called presheaf categories, carry a
+remarkable amount of structure.  Indeed they are a monoidal topos and,
+as such, are a model to a substructural dependently typed theory.
+
+Thus we
+present \theoryname (\theoryabbv) as a domain-specific type theory that
+abstracts over the structure of $\Grammar$. To begin, let us look at what some
+programs in \theoryabbv look like.
+
+\steven{TODO Reword this transition into the examples}
+
+% \max{this section kind of trails off at the end with a very vague aside about a $2$-category of grammar formalisms.
 Maybe this should go in future work/discussion?
@@
 
-\pedro{Would building on this proposed structure make the message of the section clearer? In this case
-  we would have to be a bit clearer in terms of how familiar notions of grammar can be seen as categories with functor into $\Grammar$.}
+% \pedro{Would building on this proposed structure make the message of the section clearer? In this case
+%   we would have to be a bit clearer in terms of how familiar notions of grammar can be seen as categories with functor into $\Grammar$.}
 
@@
 
-To gain intuiton for working in \theoryname, let's walk through some examples of
-grammars and parses internal to it. In each of our examples, we will look at
+To gain intuiton for working in \theoryabbv, we study some examples of
+grammars and parses of them internal to \theoryabbc. In each of our examples, we will look at
 grammars over the three character alphabet $\{ a , b , c\}$.
@@
 \paragraph{Finite Grammars}
-First consider finite grammars --- those built from base types via dijunctions $\oplus$ and
+First consider finite grammars --- those built from base types via disjunctions $\oplus$ and
 concatenations $\otimes$. For instance, $(a \oplus b) \otimes c$.
-In \cref{fig:fingram} we find a proof tree showing that the input string \stringquote{ac} matches
-the grammar $(a \oplus b) \otimes c$.
+In \cref{fig:fingram} we find a proof tree that constructs a term of type
+$(a \oplus b) \otimes c$ in the context $x : a, y : c$. We think of this term as
+a parse tree internal to \theoryabbv.
+
+Parses of a fixed string $w$ into a grammar $A$ are internally represented as
+terms of type $A$ in the context $\internalize w$, where $\internalize w$ is a
+comma-separated list of the constituent characters of $w$. In particular, to
+parse \stringquote{ac}, we build parse trees in the context
+$\internalize{\stringquote{ac}} = x : A , y : C$. Thus the term
+$\inl(x) \otimes y$ from \cref{fig:fingram} details the
+manner by which \stringquote{ac} matches the grammar $(a \oplus b) \otimes c$.
 
@@
 
-To parse the specific string \stringquote{ac}, we populate the context with an
-internalized encoding of that string. In contexts of this form, linear terms
-describe the structure of a parse tree. That is, $\inl(x) \otimes y$ details the
-manner by which \stringquote{ac} matches $(a \oplus b) \otimes c$.
-
 Because derivations in a linear context correspond to parse trees in this
@@
 For a grammar $A$, the recursive nature of $A^{*}$ is captured by defining
-$A^{*}$ to be an inductive linear
+$A^{*}$ to be an inductive\footnote{As we will soon see, our theory defines
+  \emph{indexed} inductive types. We can recover non-indexed inductive types by
+  indexing over the unit type.} linear
 type, as shown in \cref{fig:kleenestarinductive}. Just as in \cref{fig:fingram}, we can show that particular strings match a Kleene
@@
 automaton (NFA) $N$ from \cref{fig:exampleNFA}. Such automata play a critical
-role, as they serve as operationalized and
+role in defining parsers, as they often serve as an operationalized and
 executable presentation of a formal grammar. Therefore, it is crucial that we
@@
 Our method of representing these automata internally is to provide an \emph{indexed}
-inductive type of traces through the machine. For $N$, this indexed inductive
-type is given in \cref{fig:nfainductive}.
+inductive type of traces through the machine. An ordered, linear theory is
+integral to our design of \theoryabbv ; however, there are some aspects of parsing that
+cannot be captured through linear types alone. Often one wants to define a
+\emph{family} of linear types varying over a set, which is a primary motivator
+for us to make \theoryabbv a linear-non-linear theory. For the NFA $N$, an indexed inductive
+type of traces through $N$ is given in \cref{fig:nfainductive}.
 
 The type of traces in $N$ is indexed by two pieces of data: a state in $N$ and a
-Boolean. For each state $s : N.\mathsf{state}$ and $acc : \Bool$, we define the
-type of traces from state $s$, $\Trace_{N}~s~acc$. The Boolean flag $acc$ marks if the trace
+Boolean. For each state $s : N.\mathsf{state}$ and $acc : \Bool$, we define
+$\Trace_{N}~s~acc$ which denotes the
+type of traces starting at state $s$ and terminating at an arbitrary state. The Boolean flag $acc$ marks if the trace
 ends in either an accepting or a rejecting state. Let us dissect the structure
@@
 
-Overall, leveraging \emph{indexed} inductive types provides a concise interface
-for defineing a family
-of mutually recursive types over the indexing data.
+Overall, leveraging the \emph{indexed} inductive types provides a concise interface
+for defining the traces of $N$ as a family
+of mutually recursive types varying coherently over indexing data drawn from
+$N.\mathsf{state} \times \Bool$.
 
@@
       \Gamma ; \Delta \vdash a : A \\
-      \Gamma ; \Delta \vdash p : \apptol {f}{a} = \apptol {g}{a}}
+      \Gamma ; \Delta \vdash p : \applto {f}{a} = \applto {g}{a}}
     {\Gamma ; \Delta \vdash \equalizerpair{a}{p} : \equalizer {a}{f}{g}}
@@
 
-Omission of the structural rules of a deductive system, such as in
-linear logic \cite{GIRARD19871}, offers precise control over how a value is used
-in a derivation. Linear logic omits the weakening and contraction rules
-to ensure that every value in context is used exactly once. This control enables
-\emph{resource-sensitive} reasoning, where we may treat a resource as
-\emph{consumed} after usage. This viewpoint is amenable to parsing applications,
-as we may treat characters of a string as finite resources that are consumed at
-parse-time.
+% \steven{TODO the intro to this section is to be condensed based on what is
+%   included in the examples section as motivation and intuition building}
+
+% Omission of the structural rules of a deductive system, such as in
+% linear logic \cite{GIRARD19871}, offers precise control over how a value is used
+% in a derivation. Linear logic omits the weakening and contraction rules
+% to ensure that every value in context is used exactly once. This control enables
+% \emph{resource-sensitive} reasoning, where we may treat a resource as
+% \emph{consumed} after usage. This viewpoint is amenable to parsing applications,
+% as we may treat characters of a string as finite resources that are consumed at
+% parse-time.
 That is, when reading a string, the occurrence of any character is read once.
@@
 One may then envision a linear type system where the types comprise
-formal grammars generated over some alphabet $\Sigma$, and the type constructors
-correspond precisely to inductive constructions on grammars --- such as
-conjunction, disjunction, concatenation, etc.
-
-Programming in a purely linear term language is limiting due to the variable
-usage restrictions. Code in such a language can
-become unnecessarily verbose when ensuring the linearity invariant.
-To alleviate this
-concern, in 1995 Benton et al.\ proposed an alternative categorical
-semantics of linear logic based on adjoint interactions between linear
-and non-linear logics \cite{bentonMixedLinearNonlinear1995} ---
-appropriately referred to as a \emph{linear-non-linear} (LNL)
-system. This work is simply typed, so the boundary between linear and
-non-linear subtheories is entirely characterized via a monoidal
-adjunction between linear terms and non-linear terms.
-
-\steven{Describe this monoidal adjunction with $\ltonl A$ and the derivable
-  version of $F$}
-
-Inside of an LNL system, linearity may be thought of as an option that users can
-choose to deliberately invoke at deliberate points in their developments in an
-otherwise intuitionistic system. However, if we are wishing to treat parsers as
-linear terms over input strings, the non-linear fragment of an LNL theory does
-not really assist in the development of parsers. It is instead the case that
-parsers may benefit from a \emph{dependence} on non-linear terms.
-Through the approach described by Krishnaswami et al.\ in
-\cite{krishnaswami_integrating_2015},
-we may define a restricted form of dependent types. In particular, dependence
-on linear terms
-is disallowed; however, through dependence of a linear term on a non-linear
-index, we can recover the definition of Aho's indexed grammars \cite{AhoIndexed}
-internal to \theoryabbv. Although, we can define strictly more things, as Aho's
-indexed grammars provide restricted access to the index whereas we can do
-whatever with it.
-
-\steven{elaborate on indexed grammars in a more articulate way in the last sentence}
-\steven{Put the aho stuff in the subtheory section}
+% formal grammars generated over some alphabet $\Sigma$, and the type constructors
+% correspond precisely to inductive constructions on grammars --- such as
+% conjunction, disjunction, concatenation, etc.
+
+% Programming in a purely linear term language is limiting due to the variable
+% usage restrictions. Code in such a language can
+% become unnecessarily verbose when ensuring the linearity invariant.
+% To alleviate this
+% concern, in 1995 Benton et al.\ proposed an alternative categorical
+% semantics of linear logic based on adjoint interactions between linear
+% and non-linear logics \cite{bentonMixedLinearNonlinear1995} ---
+% appropriately referred to as a \emph{linear-non-linear} (LNL)
+% system. This work is simply typed, so the boundary between linear and
+% non-linear subtheories is entirely characterized via a monoidal
+% adjunction between linear terms and non-linear terms.
+
+% \steven{Describe this monoidal adjunction with $\ltonl A$ and the derivable
+%   version of $F$}
+
+% Inside of an LNL system, linearity may be thought of as an option that users can
+% choose to deliberately invoke at deliberate points in their developments in an
+% otherwise intuitionistic system. However, if we are wishing to treat parsers as
+% linear terms over input strings, the non-linear fragment of an LNL theory does
+% not really assist in the development of parsers. It is instead the case that
+% parsers may benefit from a \emph{dependence} on non-linear terms.
+% Through the approach described by Krishnaswami et al.\ in
+% \cite{krishnaswami_integrating_2015},
+% we may define a restricted form of dependent types.
+
+% In particular, dependence
+% on linear terms
+% is disallowed; however, through dependence of a linear term on a non-linear
+% index, we can recover the definition of Aho's indexed grammars \cite{AhoIndexed}
+% internal to \theoryabbv. Although, we can define strictly more things, as Aho's
+% indexed grammars provide restricted access to the index whereas we can do
+% whatever with it.
+
+% \steven{elaborate on indexed grammars in a more articulate way in the last sentence}
+% \steven{Put the aho stuff in the subtheory section}
+%
 
@@
 
-The linear typing judgment in our syntax takes on the following schematic form
-$\Gamma ; \Delta \vdash a : A$. First, $A$ represents a \emph{linear type} in
-our syntax. The intended semantics of these linear types are formal grammars.
-That is, the linear typing system is designed to syntactically reflect the
-behavior of formal grammars. For this reason, we may often interchangeably use
-the terms ``linear type'' and ``grammar''.
-
-The term $a$ is an inhabitant of type $A$, which is thought of as a parse tree of
-the grammar $A$. The core idea of this entire paper follows precisely from this
-single correspondence: grammars are types, and the inhabitants of these types
-are parse trees for the grammars.
-
-$\Gamma$ represents a non-linear context, while $\Delta$ represents a
-\emph{linear context} dependent on $\Gamma$. These linear contexts behave
-substructurally. As stated earlier, they are linear --- they do not obey
-weakening or contraction --- because a character is exhausted once read by a
-parsing procedure. Moreover, the characters in strings appear in the order in
-which we read them. We do not have the freedom to freely permute characters,
-therefore any type theory that is used to reason about formal grammars ought to
-omit the structural rule of exchange as well. This means that every variable
-within $\Delta$ must be used \emph{exactly once} and \emph{in order of
-  occurrence}. Thus, we can think of the linear contexts as an ordered list of
-limited resources. Once a resource is consumed, we cannot make reference to it
-again. Variables in a linear context then act like building
-blocks for constructing patterns over strings.
-
-We give the base types and type constructors for linear terms. As the
-interpretation of types as grammars in $\Grammar$ serves as our intended
-semantics, we simultaneously give the interpretation $\sem \cdot$ of the
-semantics as grammars.
-
-\pedro{The paragraphs above are a bit circular, what about the following alternative?}
-\steven{I am in favor of cutting what's above, but we haven't introduced the
-  syntax in the main body of the paper yet, even if the figures appeared
-  earlier. So what you write needs to take that into account}
-\pedro{In which case we should introduce it in the previous section. I suggest
-  adding this right after ``and the intuitionistic typing rules in fig 3''}
+% The linear typing judgment in our syntax takes on the following schematic form
+% $\Gamma ; \Delta \vdash a : A$. First, $A$ represents a \emph{linear type} in
+% our syntax. The intended semantics of these linear types are formal grammars.
+% That is, the linear typing system is designed to syntactically reflect the
+% behavior of formal grammars. For this reason, we may often interchangeably use
+% the terms ``linear type'' and ``grammar''.
+
+% The term $a$ is an inhabitant of type $A$, which is thought of as a parse tree of
+% the grammar $A$. The core idea of this entire paper follows precisely from this
+% single correspondence: grammars are types, and the inhabitants of these types
+% are parse trees for the grammars.
+
+% $\Gamma$ represents a non-linear context, while $\Delta$ represents a
+% \emph{linear context} dependent on $\Gamma$. These linear contexts behave
+% substructurally. As stated earlier, they are linear --- they do not obey
+% weakening or contraction --- because a character is exhausted once read by a
+% parsing procedure. Moreover, the characters in strings appear in the order in
+% which we read them. We do not have the freedom to freely permute characters,
+% therefore any type theory that is used to reason about formal grammars ought to
+% omit the structural rule of exchange as well. This means that every variable
+% within $\Delta$ must be used \emph{exactly once} and \emph{in order of
+%   occurrence}. Thus, we can think of the linear contexts as an ordered list of
+% limited resources. Once a resource is consumed, we cannot make reference to it
+% again. Variables in a linear context then act like building
+% blocks for constructing patterns over strings.
+
+% We give the base types and type constructors for linear terms. As the
+% interpretation of types as grammars in $\Grammar$ serves as our intended
+% semantics, we simultaneously give the interpretation $\sem \cdot$ of the
+% semantics as grammars.
+
+% \pedro{The paragraphs above are a bit circular, what about the following alternative?}
+% \steven{I am in favor of cutting what's above, but we haven't introduced the
+%   syntax in the main body of the paper yet, even if the figures appeared
+%   earlier. So what you write needs to take that into account}
+% \pedro{In which case we should introduce it in the previous section. I suggest
+%   adding this right after ``and the intuitionistic typing rules in fig 3''}
 
@@
 Recursive linear types may be defined via the least-fixed point operator $\mu$.
-The grammar semantics of which are the fixed-point of sets induced by the
-grammar semantics of $A$.
-%
-\[ \sem{\mu x. A} \gamma = \mu (x:\Gr_i). \sem{A}(\gamma,x) \]
-%
-\pedro{We should justify, even if briefly, the existence of this
-  fixed-point}
-
-Observe that we need not take the Kleene star as a primitive
-grammar constructor, as it is definable as a fixed point.
-The Kleene star of a grammar $g$ is given as,
-
-\[
-  g^{*} := \mu X . I \oplus (g \otimes X)
-\]
-
-\begin{figure}[h!]
-\begin{mathpar}
-  \inferrule
-  {\Gamma ; \Delta \vdash p : I}
-  {\Gamma ; \Delta \vdash \mathsf{nil} : g^{*}}
-
-  \and
-
-  \inferrule
-  {\Gamma ; \Delta \vdash p : g \\ \Gamma ; \Delta' \vdash q
-  : g^{*}}
-  {\Gamma ; \Delta \vdash \mathsf{cons}(p , q) : g^{*}}
-
-  \\
-
-  \inferrule
-  {
-    \Gamma ; \Delta \vdash p : g^{*} \\
-    \Gamma ; \cdot \vdash p_{\varepsilon} : h \\
-    \Gamma ; x : g , y : h \vdash p_{\ast} : h
-  }
-  {\Gamma ; \Delta \vdash \mathsf{foldr}(p_{\varepsilon} , p_{\ast}) : g^{*}}
-\end{mathpar}
-\caption{Kleene Star Rules}
-\label{fig:star}
-\end{figure}
-
-Likewise, $g^{*}$ has admissible introduction and
-elimination rules, shown in \cref{fig:star}. Note that this
-definition of $g^{*}$ and these
-rules arbitrarily assigns a handedness to the Kleene star.
-We could have just as well took it to be a fixed point of
-$I \oplus (X \otimes g)$. In fact, the definitions are
-equivalent, as the existence of the $\mathsf{foldl}$ term below
-shows that $g^{*}$ is also a fixed point of
-$I \oplus (X \otimes g)$.
-This is a marked point of difference from Kleene
-algebra with recursion, where the fixed points for the left and right variants
-of Kleene star need not agree \cite{leiss}.
-
-\begin{equation}
-  \label{eq:foldl}
-  \inferrule
-  {
-    \Gamma ; \Delta \vdash p : g^{*} \\
-    \Gamma ; \cdot \vdash p_{\varepsilon} : h \\
-    \Gamma ; y : h , x : h \vdash p_{\ast} : h
-  }
-  {\Gamma ; \Delta \vdash \mathsf{foldl}(p_{\varepsilon} , p_{\ast}) : g^{*}}
-\end{equation}
-
-In fact, the $\mathsf{foldl}$ term is defined using
-$\mathsf{foldr}$ --- much in the same way one
-may define a left fold over lists in terms of a right fold
-in a functional programming language.
-The underlying trick is to fold over a list of linear functions
-instead of the original string. We curry each character $c$
-of the string into a function that concatenates $c$, and
-right fold over this list of linear functions. Because function
-application is left-associative, this results in a left
-fold over the original string.
-
-We only take fixed points of a single variable as a
-primitive operation in the type theory, but we may apply
-Beki\`c's theorem \cite{Bekić1984} to define an admissible
-notion of multivariate fixed point. This is particularly
-useful for defining grammars that encode the states of an
-automaton, as we will see in \cref{sec:automata}. In \cref{fig:multifix} we provide the
-introduction and elimination principles for such a fixed
-point, where $\sigma$ is the substitution that unrolls the
-mutually recursive definitions one level. That is,
-
-\begin{align*}
-  \sigma = \{ & \mu(X_{1} = A_{1} \dots, X_{n} = A_{n}).X_{1} / X_{1} , \dots, \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{n} / X_{n} \}
-\end{align*}
-
-\begin{figure}
-\begin{mathpar}
-  \inferrule
-  {\Gamma ; \Delta \vdash e : \simulsubst {A_{k}} {\sigma}}
-  {\Gamma ; \Delta \vdash \mathsf{cons}~e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{k}}
-
-  \and
-
-  \inferrule
-  {\Gamma ; \Delta \vdash e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{k} \\
-             \Gamma ; x_{j} : \simulsubst {A_{j}}{\sigma} \vdash e_{j} : B_{j}\quad \forall j
-  }
-  {\Gamma; \Delta \vdash \mathsf{mfold}(x_{1}.e_{1}, \dots, x_{n}.e_{n})(e) : B_{k}}
-\end{mathpar}
-\caption{Multi-fixed Points}
-\label{fig:multifix}
-\end{figure}
+% The grammar semantics of which are the fixed-point of sets induced by the
+% grammar semantics of $A$.
+% %
+% \[ \sem{\mu x. A} \gamma = \mu (x:\Gr_i). \sem{A}(\gamma,x) \]
+% %
+% \pedro{We should justify, even if briefly, the existence of this
+%   fixed-point}
+
+% Observe that we need not take the Kleene star as a primitive
+% grammar constructor, as it is definable as a fixed point.
+% The Kleene star of a grammar $g$ is given as,
+
+% \[
+%   g^{*} := \mu X . I \oplus (g \otimes X)
+% \]
+
+% \begin{figure}[h!]
+% \begin{mathpar}
+%   \inferrule
+%   {\Gamma ; \Delta \vdash p : I}
+%   {\Gamma ; \Delta \vdash \mathsf{nil} : g^{*}}
+
+%   \and
+
+%   \inferrule
+%   {\Gamma ; \Delta \vdash p : g \\ \Gamma ; \Delta' \vdash q
+%   : g^{*}}
+%   {\Gamma ; \Delta \vdash \mathsf{cons}(p , q) : g^{*}}
+
+%   \\
+
+%   \inferrule
+%   {
+%     \Gamma ; \Delta \vdash p : g^{*} \\
+%     \Gamma ; \cdot \vdash p_{\varepsilon} : h \\
+%     \Gamma ; x : g , y : h \vdash p_{\ast} : h
+%   }
+%   {\Gamma ; \Delta \vdash \mathsf{foldr}(p_{\varepsilon} , p_{\ast}) : g^{*}}
+% \end{mathpar}
+% \caption{Kleene Star Rules}
+% \label{fig:star}
+% \end{figure}
+
+% Likewise, $g^{*}$ has admissible introduction and
+% elimination rules, shown in \cref{fig:star}. Note that this
+% definition of $g^{*}$ and these
+% rules arbitrarily assigns a handedness to the Kleene star.
+% We could have just as well took it to be a fixed point of
+% $I \oplus (X \otimes g)$. In fact, the definitions are
+% equivalent, as the existence of the $\mathsf{foldl}$ term below
+% shows that $g^{*}$ is also a fixed point of
+% $I \oplus (X \otimes g)$.
+% This is a marked point of difference from Kleene
+% algebra with recursion, where the fixed points for the left and right variants
+% of Kleene star need not agree \cite{leiss}.
+
+% \begin{equation}
+%   \label{eq:foldl}
+%   \inferrule
+%   {
+%     \Gamma ; \Delta \vdash p : g^{*} \\
+%     \Gamma ; \cdot \vdash p_{\varepsilon} : h \\
+%     \Gamma ; y : h , x : h \vdash p_{\ast} : h
+%   }
+%   {\Gamma ; \Delta \vdash \mathsf{foldl}(p_{\varepsilon} , p_{\ast}) : g^{*}}
+% \end{equation}
+
+% In fact, the $\mathsf{foldl}$ term is defined using
+% $\mathsf{foldr}$ --- much in the same way one
+% may define a left fold over lists in terms of a right fold
+% in a functional programming language.
+% The underlying trick is to fold over a list of linear functions
+% instead of the original string. We curry each character $c$
+% of the string into a function that concatenates $c$, and
+% right fold over this list of linear functions. Because function
+% application is left-associative, this results in a left
+% fold over the original string.
+
+% We only take fixed points of a single variable as a
+% primitive operation in the type theory, but we may apply
+% Beki\`c's theorem \cite{Bekić1984} to define an admissible
+% notion of multivariate fixed point. This is particularly
+% useful for defining grammars that encode the states of an
+% automaton, as we will see in \cref{sec:automata}. In \cref{fig:multifix} we provide the
+% introduction and elimination principles for such a fixed
+% point, where $\sigma$ is the substitution that unrolls the
+% mutually recursive definitions one level. That is,
+
+% \begin{align*}
+%   \sigma = \{ & \mu(X_{1} = A_{1} \dots, X_{n} = A_{n}).X_{1} / X_{1} , \dots, \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{n} / X_{n} \}
+% \end{align*}
+
+% \begin{figure}
+% \begin{mathpar}
+%   \inferrule
+%   {\Gamma ; \Delta \vdash e : \simulsubst {A_{k}} {\sigma}}
+%   {\Gamma ; \Delta \vdash \mathsf{cons}~e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{k}}
+
+%   \and
+
+%   \inferrule
+%   {\Gamma ; \Delta \vdash e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{k} \\
+%              \Gamma ; x_{j} : \simulsubst {A_{j}}{\sigma} \vdash e_{j} : B_{j}\quad \forall j
+%   }
+%   {\Gamma; \Delta \vdash \mathsf{mfold}(x_{1}.e_{1}, \dots, x_{n}.e_{n})(e) : B_{k}}
+% \end{mathpar}
+% \caption{Multi-fixed Points}
+% \label{fig:multifix}
+% \end{figure}
 
@@
 
-\steven{Aho's indexed grammars to be moved into here to reflect on yet another
-  grammar formalism that is internally definable}
+\paragraph{Indexed Grammars}
+The LNL dependent types can be used to encode Aho's indexed grammars
+\cite{AhoIndexed} internal to \theoryabbv. Although, we can define strictly more
+things with this non-linear index, as we have unstricted access to the index
+whereas Aho's grammars can only manipulate it in a particular way.
+
+\steven{Elaborate on these, and note that they are bw context free and context
+  sensitive on the chomsky hierarchy}
 
@@
 
-\steven{below this is unedited word vomit}
-\steven{this entire ``Formal Grammar Theory in Lambek'' section should become much more compact with the examples
-  section from the beginning in mind}
+\steven{this section is mostly word vomit, need to edit}
 
@@
 one parse tree for any input string. There are several ways to capture this
-concept internally.
-
-In the presence of distributive coproducts, the following are equivalent:
+concept inside of the theory of \theoryabbv.
+
+In the presence of distributive coproducts (which we probably add a) the following are equivalent:
 \begin{itemize}
@@
 \paragraph{Parseability}
-We say $A$ is \emph{weakly parseable} if there exists a type $B$ and a map
+We say $A$ is \emph{parseable} if there exists a type $B$ and a map
 $\top \to A \oplus B$. Because $\top \cong \String$, this says that any string
@@
 
-If $\top \cong A \oplus B$, call $A$ \emph{parseable}. This notion of
-parseability implies that $A$ and $B$ are each unambiguous.
+If $\top \cong A \oplus B$, call $A$ \emph{unambiguously parseable}.
 
@@
 In the case that $A$ is parseable with respect to $\neg A$ --- i.e.
-$\top \cong A \oplus \neg A$ --- then call $A$ \emph{decidable}.
+there's a map $\top \to A \oplus \neg A$ --- then call $A$ \emph{decidable}.
 
@@
 \steven{TODO cite that these are the same as Chomsky's notions of equivalence}
+
+The induced notion of isomorphism in $\Grammar$ encodes that two grammars are equivalent if there is
+a bijective translation of parses. In other words, two grammars are isomorphic
+if they capture the same structural descriptions of strings, which is
+precisely Chomsky's notion of \emph{strong equivalence} of
+grammars \cite{chom1963}. In works like
+\cite{yoshinaga2002formal}, this notion of strong equivalence is often treated
+like the appropriate choice of ``sameness'' for two grammars. When abstracting over the latent
+structure of formal grammars and describing it in the the language of
+categories, it is encouraging to see the same definition of equivalence arise
+naturally.
 
@@
 for each $n : \mathbb{N}$ with an additional fail state. This diverges a bit
-from classical language theory, where this machine would be encoded as a
+from classical language theory, where the recognizer for $D$ would be encoded as a
 pushdown automaton with finitely many
@@
 are a little cumbersome to use. Moreover, they
-seem overfit to classical reasoning as they don't preserve the gramamr up to
+seem overfit to classical reasoning as they don't preserve the grammar up to
 strong equivalence, only weak equivalnce in general. \steven{cite lambek
@@
 
-\steven{Everything in this section below this marker is outdated and needs to be
-folded into the above or cut}
-
-\steven{Need a different segue because canonicity is no longer before this}
-The canonicity theorem from the previous section gives a tight connection
-between terms in our type theory and the theory of formal languages. In this section
-we further explore these connections by using the fact that, classically
-formal language theory is closely related to the study of automata.
-In the Chomsky hierarchy, each language class is associated to a class of
-automata that serve as recognizers. Internal to \theoryabbv,
-we can characterize these language classes syntactically; moreover, we
-demonstrate the equivalence of these language classes to its associated automata class as a
-proof term within our logic.
-
-\subsection{Non-deterministic Finite Automata}
-\label{subsec:finiteaut}
-\begin{figure}
-  \begin{tikzpicture}[node distance = 25mm ]
-    \node[state, initial, accepting] (1) {$1$};
-    \node[state, below left of=1] (2) {$2$};
-    \node[state, right of=2] (3) {$3$};
-
-    \path[->] (1) edge[above] node{$b$} (2)
-              (1) edge[below, bend right, left=0.3] node{$\epsilon$} (3)
-              (2) edge[loop left] node{$a$} (2)
-              (2) edge[below] node{$a, b$} (3)
-              (3) edge[above, bend right, right=0.3] node{$a$} (1);
-  \end{tikzpicture}
-  \caption{An example NFA}
-  \label{fig:NFA}
-\end{figure}
-
-Classically, a \emph{nondeterministic finite automaton} (NFA) is a finite state machine where
-transitions are labeled with characters from a fixed alphabet $\Sigma$. These
-are often represented formally as a 5-tuple $(Q, \Sigma, \delta, q_{0}, F)$,
-
-\begin{itemize}
-  \item $Q$ a finite set of states
-  \item $\Sigma$ a fixed, finite alphabet
-  \item $\delta : Q \times (\Sigma \cup \{ \varepsilon\}) \to \mathcal{P}(Q)$ the labeled transition function
-  \item $q_{0} \in Q$ the start state
-  \item $F \subset Q$ a set of accepting states
-\end{itemize}
-
-\subsection{A DFA Parser}
-\label{subsec:regexparser}
-
-Just as we encoded traces of NFAs as grammars, we likewise
-encode the traces of a DFA as grammars. The key difference
-between NFAs and DFAs is \emph{determinism} --- meaning,
-that in a state $q$ inside of DFA, given a character $c$ there
-will be exactly one transition that we may take leaving $q$
-with label $c$. For us, this changes the definition of valid
-transitions for a DFA, instead of the definition of
-$\mathsf{Trans}$ provided in \cref{subsec:finiteaut} DFAs
-obey
-
-\begin{gather*}
- \mathsf{State} \in \{ g_{q} : q \in Q \} \\
- \mathsf{Trans}(q) ::= \bigoplus_{c \in \Sigma} (c \otimes \mathsf{State})
-\end{gather*}
-
-Meaning, each state has a transition for every character. \emph{Note:} the above use
-  of a disjunction over $\Sigma$ is a bit of abuse of notation.
-  As $\Sigma$ is a finite alphabet, we wish to think of this as an iterated
-  binary sum, or perhaps as a finitely indexed sum defined via linear dependent
-  sums in the same manner by which we defined binary sums.
-
-We now wish to define a parser term for DFA grammars. In
-particular, for a DFA $D$ we want to build a term,
-
-\[
-  w : \String \vdash \mathsf{parse}_{D} : \mathsf{AccTrace}_{D} \oplus \top
-\]
-
-where left injection into the output type denotes acceptance
-by the parser, and right injection denotes rejection. To
-build such a parser, it will suffice to construct a term
-
-\[
-  w : \String \vdash \mathsf{parse}_{D} : \LinSigTy q Q {\mathsf{Trace}_{D}(q_{0} , q)}
-\]
-This is because given a trace of a DFA, we may easily check
-if we should accept or reject by simply testing
-if the final state is accepting.
-
-Because $w$ is a Kleene star of characters, we may construct
-our desired $\mathsf{parse}_{D}$ term as a $\mathsf{foldl}$
-over $w$. In the empty case, we just have the trace that
-ends at the accepting state. In the recursive case, we
-effectively add to our trace by transitioning one character
-at a time, as we read them moving across $w$.
-
-Perhaps this derivation is not too surprising. All it says
-is that a DFA may be ran by transitioning a single character
-at a time, and then accepting or rejecting based on the
-final state. This is exactly what DFAs did initially, so
-what did we gain? Well, this has the benefit of our type
-system to ensure its correctness. Moreover, this construction exports to an
-intrinsically verified and executable decision procedure for DFAs in \theoryabbv
-embedded in Agda.
-
-We should note that the
-construction of a DFA parser requires the addition of an additional but seemingly innocuous
-axiom.
-
-\[
-  \top \cong \left( \bigoplus_{c : \Sigma} c \right)^{*}
-\]
-
-That is, the characters in the alphabet $\Sigma$ really are all of the
-characters. The need for this axiom arises when we go to take a transition
-within the DFA,
-as we must ensure that the next character of the string has a corresponding
-transition from the current state
---- which would be guaranteed by determinism provided there are no surprise characters.
-
-\subsection{Regular Expressions and DFAs}
-\label{subsec:deriv}
-
-In order to extend the DFA
-parser from \cref{subsec:regexparser} to the construction of
-a verified parser
-generator for regular expression we need to perform some
-plumbing establishing an equivalence between regular
-expressions and DFAs.
-
-There are several routes we may hope to take in establishing
-this equivalence. First, we could prove an equivalence
-between NFAs and regular expressions, and separately prove
-an equivalence between NFAs and DFAs.
-In \cref{subsec:eqproofs}, we include a version of
-Thompson's construction --- which established the
-equivalence between regular grammars and DFAs. We may additionally
-hope to internalize a variant of the powerset construction \cite{rabinFiniteAutomataTheir1959}
---- which takes as input an NFA and constructs a DFA that
-recognizes the same language --- and combine the results of
-Thompson's construction and the powerset constructions to give an equivalence
-between regular expressions and DFAs.  This route is alluring, as it
-internalizes several classic grammar-theoretic constructions. However, it may
-necessitate extensions to the LNL theory, like
-a propositional truncation, and we have not yet investigated
-how this would interact with the existing types in the
-theory. The addition of a propositional truncation may seem
-harmless, but it is not always immediately clear how
-distinct constructions will interact. For instance, when
-exploring LNL models, Benton discovered that the synthesis
-of linear and dependent types require a new presentation of
-the $!$ modality from linear logic
-\cite{bentonMixedLinearNonlinear1995}. That is all to say,
-this is a work in progress and
-it is not immediate that the addition of a propositional
-truncation is adequate for establishing the weak equivalence
-between NFAs and DFAs.
-
-We may instead hope to internalize an equivalence between
-regular grammars and DFAs by using Brzozowski derivatives to
-directly create a DFA that is weakly equivalent to a given
-regular expression, as described by Owens et al.
-\cite{owensRegularexpressionDerivativesReexamined2009}.
-One characterization of regular grammars is that they are
-precisely those grammars which have finitely many inequivalent Brzozowski
-derivatives
-\cite{brzozowskiDerivativesRegularExpressions1964}.
-The algorithm used by Owens takes in a
-regular grammar and generates a DFA that recognizes the same
-language, and the states in this DFA are the finitely many
-derivative equivalence classes. We initially had a version of
-this theorem very roughly internalized in the LNL theory.
-To our taste, too much of this presentation relied on
-meta-arguments that lived outside of
-our formalism, and thus this particular phrasing of the
-theorem did not translate well into formalization.
-
-In any case, we believe
-that revisiting these lines of thought will lead to a
-satisfactory internalization of the equivalence between
-regular grammars and DFAs, and thus would bridge the gap
-between our DFA parser and a full regular expression parser.
-
-\subsection{Equivalence Between Regular Expressions and Finite Automata}
-\label{subsec:eqproofs}
-In this section, we describe a version of Thompson's
-construction \cite{thompsonProgrammingTechniquesRegular1968}
-   where we construct an NFA that recognizes a given regular
-expression. Moreover, we will show that this NFA is strongly equivalent to the
-original grammar. Witnessing this construction in our syntax has two benefits
-\begin{enumerate}
-  \item It reinforces this high-level view that the syntax is a natural and
-        general setting for formal grammar reasoning, as we demonstrate that
-        this formal system subsume results from existing systems, and
-  \item Following the
-        development of Thompson's construction, we then need
-        only establish the equivalence of NFAs and DFAs to
-        complete the full regular expression parser
-\end{enumerate}
-
-\begin{theorem}[Thompson]
-  \label{thm:thompson}
-  For $g$ a regular grammar $g$, there is an NFA $N$ that recognizes the same
-  language as $g$.
-\end{theorem}
-We make a pretty straightforward adaptation of Thompson's theorem to our setting,
-
-\begin{theorem}[Typed Thompson]
-  \label{thm:typthompson}
-  For $g$ a regular expression $g$, there is an NFA $N$ such that $g$ is isomorphic
-  to $\mathsf{AccTrace}_{N}$.
-\end{theorem}
-
-\begin{proof}[Proof Sketch]
-  Recall that regular expressions are inductively defined via
-  disjunction, concatenation, and Kleene star over literals
-  and the empty grammar. By induction over regular expressions,
-  we will construct an NFA that is equivalent to $g$.
-
-  First, define the recognizing NFA for the empty grammar
-  $I$.
-
-  \begin{figure}[h!]
-  \begin{tikzpicture}[node distance = 25mm ]
-    \node[state, initial] (1) {$1$};
-    \node[state, right of=1, accepting] (2) {$2$};
-
-    \path[->] (1) edge[below] node{$\varepsilon$} (2);
-  \end{tikzpicture}
-  \caption{$NFA(I)$}
-  \label{fig:emptyNFA}
-  \end{figure}
-
-  The type of traces from the initial state of $NFA(I)$ to the single
-  accepting state is given by,
-
-  \[
-    \mathsf{Trace}_{NFA(I)}(q_{1}, q_{2}) = \mu
-      \begin{pmatrix}
-         g_{q_{1}} := g_{q_{2}} \\
-         g_{q_{2}} := I
-      \end{pmatrix}. g_{q_{1}}
-  \]
-
-  The accepting traces through $NFA(I)$ are then described
-  as,
-
-  \[
-    \mathsf{AccTrace}_{NFA(I)} = \LinSigTy q {\{1 , 2\}} {\left( \mathsf{Trace}_{NFA(I)}(q_{1} , q) \pair \mathsf{acc}(q) \right)}
-  \]
-
-  A quick inspection of \cref{fig:emptyNFA} reveals that the
-  only reasonable choice for $q$ is $q_{2}$ --- because
-  state 2 is accepting while state 1 is not. Therefore,
-
-  \begin{align*}
-    \mathsf{AccTrace}_{NFA(I)}
-    & \cong \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2}) \pair \mathsf{acc}(q_{2}) \\
-    & \cong \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2}) \pair \top \\
-    & \cong \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2})
-  \end{align*}
-
-  From here, to prove
-  $I \cong \mathsf{AccTrace}_{NFA(I)}$ it suffices to show
-  $I \cong \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2})$. Below we
-  give two parse transformers, one from $I$ to
-  $\mathsf{Trace}_{NFA(I)}(q_{1} , q_{2})$ and vice versa.
-
-  \[
-    \inferrule
-    {p : I \vdash \mathsf{nil} : \mathsf{Trace}(q_{2} , q_{2}) \\
-     \exists \text{~transition~} q_{1} \overset{\varepsilon}{\to} q_{2}
-    }
-    {p : I \vdash \mathsf{\varepsilon cons}(\mathsf{nil}) : \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2})}
-  \]
-
-  Let $\gamma$ be the substitution $\{ g_{q_{2}} / g_{q_{1}}, I / g_{q_{2}} \}$,
-
-  \[
-    \inferrule
-    {
-      p : \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2}) \vdash p :
-        \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2}) \\
-      x_{1} : \simulsubst {g_{q_{2}}} {\gamma} = I \vdash x_{1} : I \\
-      x_{2} : \simulsubst {I} {\gamma} = I \vdash x_{2} : I
-    }
-    {
-      p : \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2}) \vdash \mathsf{mfold}(x_{1}.x_{1} , x_{2}.x_{2})(p) : I
-    }
-  \]
-
-  This concludes the proof for the case of the empty
-  grammar. Let's now walk through the construction for
-  literal grammars. Given a character $c$, we construct an
-  NFA that recognizes only the string containing the single
-  character $c$ as,
-
-  \begin{figure}[h!]
-  \begin{tikzpicture}[node distance = 25mm ]
-    \node[state, initial] (1) {$1$};
-    \node[state, right of=1, accepting] (2) {$2$};
-
-    \path[->] (1) edge[below] node{$c$} (2);
-  \end{tikzpicture}
-  \caption{$NFA(c)$}
-  \label{fig:literalNFA}
-  \end{figure}
-
-  The automaton in \cref{fig:literalNFA} induces the
-  following type of traces from $q_{1}$ to $q_{2}$.
-
-  \[
-    \mathsf{Trace}_{NFA(c)}(q_{1}, q_{2}) = \mu
-      \begin{pmatrix}
-         g_{q_{1}} := c \otimes g_{q_{2}} \\
-         g_{q_{2}} := I
-      \end{pmatrix}. g_{q_{1}}
-  \]
-
-  Through the same argument as the empty grammar, the only
-  state that is accepting is $q_{2}$ and thus,
-
-  \[
-    \mathsf{AccTrace}_{NFA(c)} \cong \mathsf{Trace}_{NFA(c)}(q_{1} , q_{2})
-  \]
-
-  To show the desired isomorphism of $c \cong NFA(c)$ we
-  make a similar argument as we did for the empty grammar
-  $I$, except we leverage the $\mathsf{cons}$ rule instead
-  of $\mathsf{\varepsilon cons}$. That is, the parse
-  transformers in either direction are given as,
-
-  \[
-    \inferrule
-    {\cdot \vdash \mathsf{nil} : \mathsf{Trace}(q_{2} , q_{2}) \\
-     \exists \text{~transition~} q_{1} \overset{c}{\to} q_{2}
-    }
-    {p : c \vdash \mathsf{cons}(\mathsf{nil}) :
-      \mathsf{Trace}_{NFA(c)}(q_{1} , q_{2})}
-  \]
-
-  \[
-    \inferrule
-    {
-      p : \mathsf{Trace}_{NFA(c)}(q_{1} , q_{2}) \vdash p :
-        \mathsf{Trace}_{NFA(c)}(q_{1} , q_{2}) \\
-      x_{1} : \simulsubst {(c \otimes g_{q_{2}})} {\gamma} = c \otimes I \vdash \mathsf{unitR}(x_{1}) : c \\
-      x_{2} : \simulsubst {I} {\gamma} = I \vdash x_{2} : I
-    }
-    {
-      p : \mathsf{Trace}_{NFA(c)}(q_{1} , q_{2}) \vdash \mathsf{mfold}(x_{1}.\mathsf{unitR}(x_{1}) , x_{2}.x_{2})(p) : c
-    }
-  \]
-
-  Where $\gamma$ is the substitution
-  $\{ c \otimes g_{q_{2}} / g_{q_{1}}, I / g_{q_{2}} \}$ and
-  $\mathsf{unitR}$ is a witness to the isomorphism
-  $c \otimes I \cong c$. Again, we may see that these do
-  indeed mutually invert each other in the Agda code.
-
-  It remains to show that the desired isomorphisms are
-  preserved by $\otimes$, $\oplus$, and Kleene star. Here,
-  we will give the argument for the disjunction case, the
-  others are defined quite similarly.
-
-  Given two NFAs $N$ and $M$, we may define a new NFA that
-  encodes the disjunction of $N$ and $M$. Denote the
-  internal states of $N$ by $q_{j}$'s and the internal states
-  of $M$ by $r_{k}$'s,
-
-  \begin{figure}[h!]
-  \begin{tikzpicture}[node distance = 20mm ]
-    \node[state] (2) {$q_{init}$};
-    \node[state, initial, below left of=2] (1) {$init$};
-    \node[state, below right of=1] (3) {$r_{init}$};
-    \node[right of=2] (4) {$\cdots$};
-    \node[right of=3] (5) {$\cdots$};
-    % \node[state, below right of=1] (3) {$3$};
-    % \node[state, below of=2] (4) {$4$};
-
-    \path[->] (1) edge[below] node{$\varepsilon$} (2);
-    \path[->] (1) edge[below] node{$\varepsilon$} (3);
-    \path[->] (2) edge[below] node{} (4);
-    \path[->] (3) edge[below] node{} (5);
-
-    \node[label={[name=l] $N$}, draw,line width=2pt,rounded corners=5pt, fit=(2)(4)] {};
-    \node[label={[name=l] $M$}, draw,line width=2pt,rounded corners=5pt, fit=(3)(5)] {};
-  \end{tikzpicture}
-  \caption{$N \oplus_{NFA} M$}
-  \label{fig:disjunctionNFA}
-  \end{figure}
-
-  \Cref{fig:disjunctionNFA} shows the process of
-  disjunctively combining NFAs. Precisely, we add a single
-  new state and we included copies of the states from each of $N$
-  and $M$. The new state acts as the initial state and has
-  $\varepsilon$-transitions to the initial states of $N$ and
-  $M$. We include all of the internal transitions from $N$
-  and $M$, and the accepting states of $N \oplus_{NFA} M$
-  are exactly the accepting states in each subautomaton.
-
-  Let $g$ and $g'$ be two regular grammars such that
-  $g \cong NFA(g)$ and $g' \cong NFA(g')$. As a matter of
-  notation\footnote{We shall similarly abuse notation for
-    $\otimes$ and Kleene. That is, for a regular grammar
-    $g$, when we write $NFA(g)$ we mean the NFA inductively
-    built up with the NFA-analogues to the constructors that
-    built up $g$.}, we will write $NFA(g \oplus g')$ for
-  $NFA(g) \oplus_{NFA} NFA(g')$. The traces of
-  $NFA(g \oplus g')$ are then given by,
-
-  \[
-    \mathsf{Trace}_{NFA(g \oplus g')}(src , dst) = \mu
-      \begin{pmatrix}
-        g_{init} := g_{q_{0}} \oplus g_{r_{0}} \\
-        g_{q_{j}} := \mathsf{Trans}_{NFA(g)}(q_{j}) \oplus \mathsf{isDst}(q_{j}) \\
-        g_{r_{k}} := \mathsf{Trans}_{NFA(g')}(r_{k}) \oplus \mathsf{isDst}(r_{k})
-      \end{pmatrix}.g_{src}
-    \]
-
-  where $\mathsf{Trans}$ is used to echo the same syntactic
-  definitions that appear in the $NFA(g)$ and $NFA(g')$.
-  Also, $src$ and $dst$ may take on any value in
-  $Q := \{init\} \cup \{q_{j}\} \cup \{r_{k}\}$, and
-  $\mathsf{isDst}(q)$ checks if $q$ is equal to $dst$. Which
-  is all to say, the traces of $NFA(g \oplus g')$ comprise
-  either a trace of $NFA(g)$, or a trace of $NFA(g')$, and
-  the transition coming out of $g_{init}$ determines which
-  subautomaton we step into.
-
-  The parse transformer from $g \oplus g'$ checks which side
-  of the sum type we are on, then takes the appropriate step
-  from $g_{init}$ in the automaton.
-
-  \[
-    \inferrule
-    {
-      u : g \vdash \iota (\phi (u)) : \mathsf{AccTrace}_{NFA(g \oplus g')} \\
-      v : g' \vdash \iota' (\psi (v)) : \mathsf{AccTrace}_{NFA(g \oplus g')}
-    }
-    {p : g \oplus g' \vdash \mathsf{case}~p \{ \mathsf{inl}(u) \mapsto s , \mathsf{inr}(v) \mapsto r \} : \mathsf{AccTrace}_{NFA(g \oplus g')}}
-  \]
-
-  with $\iota$ and $\iota'$ as embeddings from $NFA(g)$ and
-  $NFA(g')$, respectively, into $NFA(g \oplus g')$,
-  $\phi: g \cong \mathsf{AccTrace}_{NFA(g)}$, and $\psi: g' \cong \mathsf{AccTrace}_{NFA(g')}$. On a
-  high level, all this construction does is turn a parse of
-  $g$ into a parse of $NFA(g)$ and then embeds that inside
-  of the larger automaton $NFA(g \oplus g')$. Likewise for $g'$.
-
-  In the other direction, recall that the data of an
-  accepting trace for $NFA(g \oplus g')$ is a pair of a
-  trace and a proof that
-  the end state $q'$ of that trace is accepting. By
-  multifolding over the first part of that pair, we turn the
-  term of type
-  $\mathsf{Trace}_{NFA(g \oplus g')}(init , q')$ into a
-  trace of either of the subautomata,
-
-  \[
-    p : \mathsf{Trace}_{NFA(g \oplus g')}(init , q') \vdash \mathsf{mfold}_{NFA(g \oplus g')} : \mathsf{Trace}_{NFA(g)}(q_{0} , q') \oplus \mathsf{Trace}_{NFA(g')}(r_{0} , q')
-  \]
-
-  Additionally, we leverage the fact that the only accepting
-  states for $NFA(g \oplus g')$ are those from the
-  subautomata to extract that $q'$ must be an accepting
-  state from a subautomaton.
-
-  \[
-    x : \mathsf{acc}_{NFA(g \oplus g')}(q') \vdash M : \mathsf{acc}_{NFA(g)}(q') \oplus \mathsf{acc}_{NFA(g')}(q')
-  \]
-
-  We then combine the trace and proof of acceptance into an
-  accepting trace of one of the subautomata,
-
-  \[
-    p : \mathsf{AccTrace}_{NFA(g \oplus g')} \vdash N : \mathsf{AccTrace}_{NFA(g)} \oplus \mathsf{AccTrace}_{NFA(g')}
-  \]
-
-  Lastly, we then inductively use the isomorphisms $\phi$
-  and $\psi$ to turn the accepting traces into a parse of
-  $g$ or $g'$,
-
-  \[
-     N : \mathsf{AccTrace}_{NFA(g)} \oplus \mathsf{AccTrace}_{NFA(g')} \vdash \mathsf{case}~N~\{\mathsf{inl}(n) \mapsto \phi^{-1}(n), \mathsf{inr}(n') \mapsto \psi^{-1}(n')\} : g \oplus g'
-  \]
-
-  When you wish to show that these actually maps invert each other to form an isomorphism\max{jumbled sentence}, we invoke the
-  universal properties of the $\mathsf{Trace}$ types. That is,
-  $\mathsf{Trace}_{\oplus NFA}$, $\mathsf{Trace}_{N}$, and $\mathsf{Trace}_{M}$
-  are \emph{initial} algebras to the the functors that define them. Thus, the
-  embedding of traces through $N$ into $\oplus NFA$ and then mapped back down to $N$
-  compose to a map from $\mathsf{Trace}_{N}$ to itself. Such a map is unique and
-  thus must be the identity. The same reasoning shows that the desired
-  composite do indeed invert each other to form an isomorphism.
-
-  As discussed above, the other cases and
-  follow through a similar argument. This
-  concludes the proof of our variant of Thompson's construction.
-\end{proof}
-
-\subsection{Other Automata}
-\subsubsection{Pushdown Automata}
-\label{subsubsec:pda}
-A (nondeterministic) \emph{pushdown automaton} is an automaton that employs a
-stack. Just like NFAs, they have transitions labeled with characters from a
-fixed string alphabet $\Sigma$. Additionally, they maintain a stack of
-characters drawn from a stack alphabet $\Gamma$. They are often represented
-formally as a 7-tuple $(Q, \Sigma, \Gamma, \delta, q_{0}, Z, F)$,
-
-\begin{itemize}
-  \item $Q$ a finite set of states
-  \item $\Sigma$ a fixed, finite string alphabet
-  \item $S$ a fixed, finite stack alphabet
-  \item
-        $\delta \subset Q \times (\Sigma \cup \{ \varepsilon \}) \times S \to \mathcal{P}(Q \times S^{*})$
-        the transition function
-  \item $q_{0} \in Q$ the start state
-  \item $Z \in S$ the initial stack symbol
-  \item $F \subset Q$ the accepting states
-\end{itemize}
-
-We encode the traces of a pushdown automaton very similarly to those of an NFA,
-except the transitions of a PDA are instead encoded via the LNL product type
-$\bigamp$. This is because when simply transitioning via character, a PDA must
-also pop and push characters onto a stack, which is used as the argument to
-these dependent functions. Thus, the appropriate formation of traces through a
-PDA is dependent on a stack.
-
-Let $S$ be a non-linear type encoding the stack
-alphabet, and build lists over $S$ as the (non-linear) least fixed-point
-$\mathsf{List}(S) := \mu X . 1 + S \times X$. Then, the type of states for a
-PDA $P$ with stack alphabet $S$ are given as a functions that takes in lists $L$,
-and then makes a case distinction between possible transitions based on what was witnessed as $\mathsf{head}(L)$. The choice of transition
-will then determine which character to transition by and what word $w$ should be
-pushed onto the stack. The word that is added to the top of the stack is
-appended to $\mathsf{tail}(L)$ and then we recursively step into another state
-called on argument $w \cdot \mathsf{tail}(L)$.
+% \steven{Everything in this section below this marker is outdated and needs to be
+% folded into the above or cut}
+
+% \steven{Need a different segue because canonicity is no longer before this}
+% The canonicity theorem from the previous section gives a tight connection
+% between terms in our type theory and the theory of formal languages. In this section
+% we further explore these connections by using the fact that, classically
+% formal language theory is closely related to the study of automata.
+% In the Chomsky hierarchy, each language class is associated to a class of
+% automata that serve as recognizers. Internal to \theoryabbv,
+% we can characterize these language classes syntactically; moreover, we
+% demonstrate the equivalence of these language classes to its associated automata class as a
+% proof term within our logic.
+
+% \subsection{Non-deterministic Finite Automata}
+% \label{subsec:finiteaut}
+% \begin{figure}
+%   \begin{tikzpicture}[node distance = 25mm ]
+%     \node[state, initial, accepting] (1) {$1$};
+%     \node[state, below left of=1] (2) {$2$};
+%     \node[state, right of=2] (3) {$3$};
+
+%     \path[->] (1) edge[above] node{$b$} (2)
+%               (1) edge[below, bend right, left=0.3] node{$\epsilon$} (3)
+%               (2) edge[loop left] node{$a$} (2)
+%               (2) edge[below] node{$a, b$} (3)
+%               (3) edge[above, bend right, right=0.3] node{$a$} (1);
+%   \end{tikzpicture}
+%   \caption{An example NFA}
+%   \label{fig:NFA}
+% \end{figure}
+
+% Classically, a \emph{nondeterministic finite automaton} (NFA) is a finite state machine where
+% transitions are labeled with characters from a fixed alphabet $\Sigma$. These
+% are often represented formally as a 5-tuple $(Q, \Sigma, \delta, q_{0}, F)$,
+
+% \begin{itemize}
+%   \item $Q$ a finite set of states
+%   \item $\Sigma$ a fixed, finite alphabet
+%   \item $\delta : Q \times (\Sigma \cup \{ \varepsilon\}) \to \mathcal{P}(Q)$ the labeled transition function
+%   \item $q_{0} \in Q$ the start state
+%   \item $F \subset Q$ a set of accepting states
+% \end{itemize}
+
+% \subsection{A DFA Parser}
+% \label{subsec:regexparser}
+
+% Just as we encoded traces of NFAs as grammars, we likewise
+% encode the traces of a DFA as grammars. The key difference
+% between NFAs and DFAs is \emph{determinism} --- meaning,
+% that in a state $q$ inside of DFA, given a character $c$ there
+% will be exactly one transition that we may take leaving $q$
+% with label $c$. For us, this changes the definition of valid
+% transitions for a DFA, instead of the definition of
+% $\mathsf{Trans}$ provided in \cref{subsec:finiteaut} DFAs
+% obey
+
+% \begin{gather*}
+%  \mathsf{State} \in \{ g_{q} : q \in Q \} \\
+%  \mathsf{Trans}(q) ::= \bigoplus_{c \in \Sigma} (c \otimes \mathsf{State})
+% \end{gather*}
+
+% Meaning, each state has a transition for every character. \emph{Note:} the above use
+%   of a disjunction over $\Sigma$ is a bit of abuse of notation.
+%   As $\Sigma$ is a finite alphabet, we wish to think of this as an iterated
+%   binary sum, or perhaps as a finitely indexed sum defined via linear dependent
+%   sums in the same manner by which we defined binary sums.
+
+% We now wish to define a parser term for DFA grammars. In
+% particular, for a DFA $D$ we want to build a term,
+
+% \[
+%   w : \String \vdash \mathsf{parse}_{D} : \mathsf{AccTrace}_{D} \oplus \top
+% \]
+
+% where left injection into the output type denotes acceptance
+% by the parser, and right injection denotes rejection. To
+% build such a parser, it will suffice to construct a term
+
+% \[
+%   w : \String \vdash \mathsf{parse}_{D} : \LinSigTy q Q {\mathsf{Trace}_{D}(q_{0} , q)}
+% \]
+% This is because given a trace of a DFA, we may easily check
+% if we should accept or reject by simply testing
+% if the final state is accepting.
+
+% Because $w$ is a Kleene star of characters, we may construct
+% our desired $\mathsf{parse}_{D}$ term as a $\mathsf{foldl}$
+% over $w$. In the empty case, we just have the trace that
+% ends at the accepting state. In the recursive case, we
+% effectively add to our trace by transitioning one character
+% at a time, as we read them moving across $w$.
+
+% Perhaps this derivation is not too surprising. All it says
+% is that a DFA may be ran by transitioning a single character
+% at a time, and then accepting or rejecting based on the
+% final state. This is exactly what DFAs did initially, so
+% what did we gain? Well, this has the benefit of our type
+% system to ensure its correctness. Moreover, this construction exports to an
+% intrinsically verified and executable decision procedure for DFAs in \theoryabbv
+% embedded in Agda.
+
+% We should note that the
+% construction of a DFA parser requires the addition of an additional but seemingly innocuous
+% axiom.
+
+% \[
+%   \top \cong \left( \bigoplus_{c : \Sigma} c \right)^{*}
+% \]
+
+% That is, the characters in the alphabet $\Sigma$ really are all of the
+% characters. The need for this axiom arises when we go to take a transition
+% within the DFA,
+% as we must ensure that the next character of the string has a corresponding
+% transition from the current state
+% --- which would be guaranteed by determinism provided there are no surprise characters.
+
+% \subsection{Regular Expressions and DFAs}
+% \label{subsec:deriv}
+
+% In order to extend the DFA
+% parser from \cref{subsec:regexparser} to the construction of
+% a verified parser
+% generator for regular expression we need to perform some
+% plumbing establishing an equivalence between regular
+% expressions and DFAs.
+
+% There are several routes we may hope to take in establishing
+% this equivalence. First, we could prove an equivalence
+% between NFAs and regular expressions, and separately prove
+% an equivalence between NFAs and DFAs.
+% In \cref{subsec:eqproofs}, we include a version of
+% Thompson's construction --- which established the
+% equivalence between regular grammars and DFAs. We may additionally
+% hope to internalize a variant of the powerset construction \cite{rabinFiniteAutomataTheir1959}
+% --- which takes as input an NFA and constructs a DFA that
+% recognizes the same language --- and combine the results of
+% Thompson's construction and the powerset constructions to give an equivalence
+% between regular expressions and DFAs.  This route is alluring, as it
+% internalizes several classic grammar-theoretic constructions. However, it may
+% necessitate extensions to the LNL theory, like
+% a propositional truncation, and we have not yet investigated
+% how this would interact with the existing types in the
+% theory. The addition of a propositional truncation may seem
+% harmless, but it is not always immediately clear how
+% distinct constructions will interact. For instance, when
+% exploring LNL models, Benton discovered that the synthesis
+% of linear and dependent types require a new presentation of
+% the $!$ modality from linear logic
+% \cite{bentonMixedLinearNonlinear1995}. That is all to say,
+% this is a work in progress and
+% it is not immediate that the addition of a propositional
+% truncation is adequate for establishing the weak equivalence
+% between NFAs and DFAs.
+
+% We may instead hope to internalize an equivalence between
+% regular grammars and DFAs by using Brzozowski derivatives to
+% directly create a DFA that is weakly equivalent to a given
+% regular expression, as described by Owens et al.
+% \cite{owensRegularexpressionDerivativesReexamined2009}.
+% One characterization of regular grammars is that they are
+% precisely those grammars which have finitely many inequivalent Brzozowski
+% derivatives
+% \cite{brzozowskiDerivativesRegularExpressions1964}.
+% The algorithm used by Owens takes in a
+% regular grammar and generates a DFA that recognizes the same
+% language, and the states in this DFA are the finitely many
+% derivative equivalence classes. We initially had a version of
+% this theorem very roughly internalized in the LNL theory.
+% To our taste, too much of this presentation relied on
+% meta-arguments that lived outside of
+% our formalism, and thus this particular phrasing of the
+% theorem did not translate well into formalization.
+
+% In any case, we believe
+% that revisiting these lines of thought will lead to a
+% satisfactory internalization of the equivalence between
+% regular grammars and DFAs, and thus would bridge the gap
+% between our DFA parser and a full regular expression parser.
+
+% \subsection{Equivalence Between Regular Expressions and Finite Automata}
+% \label{subsec:eqproofs}
+% In this section, we describe a version of Thompson's
+% construction \cite{thompsonProgrammingTechniquesRegular1968}
+%    where we construct an NFA that recognizes a given regular
+% expression. Moreover, we will show that this NFA is strongly equivalent to the
+% original grammar. Witnessing this construction in our syntax has two benefits
+% \begin{enumerate}
+%   \item It reinforces this high-level view that the syntax is a natural and
+%         general setting for formal grammar reasoning, as we demonstrate that
+%         this formal system subsume results from existing systems, and
+%   \item Following the
+%         development of Thompson's construction, we then need
+%         only establish the equivalence of NFAs and DFAs to
+%         complete the full regular expression parser
+% \end{enumerate}
+
+% \begin{theorem}[Thompson]
+%   \label{thm:thompson}
+%   For $g$ a regular grammar $g$, there is an NFA $N$ that recognizes the same
+%   language as $g$.
+% \end{theorem}
+% We make a pretty straightforward adaptation of Thompson's theorem to our setting,
+
+% \begin{theorem}[Typed Thompson]
+%   \label{thm:typthompson}
+%   For $g$ a regular expression $g$, there is an NFA $N$ such that $g$ is isomorphic
+%   to $\mathsf{AccTrace}_{N}$.
+% \end{theorem}
+
+% \begin{proof}[Proof Sketch]
+%   Recall that regular expressions are inductively defined via
+%   disjunction, concatenation, and Kleene star over literals
+%   and the empty grammar. By induction over regular expressions,
+%   we will construct an NFA that is equivalent to $g$.
+
+%   First, define the recognizing NFA for the empty grammar
+%   $I$.
+
+%   \begin{figure}[h!]
+%   \begin{tikzpicture}[node distance = 25mm ]
+%     \node[state, initial] (1) {$1$};
+%     \node[state, right of=1, accepting] (2) {$2$};
+
+%     \path[->] (1) edge[below] node{$\varepsilon$} (2);
+%   \end{tikzpicture}
+%   \caption{$NFA(I)$}
+%   \label{fig:emptyNFA}
+%   \end{figure}
+
+%   The type of traces from the initial state of $NFA(I)$ to the single
+%   accepting state is given by,
+
+%   \[
+%     \mathsf{Trace}_{NFA(I)}(q_{1}, q_{2}) = \mu
+%       \begin{pmatrix}
+%          g_{q_{1}} := g_{q_{2}} \\
+%          g_{q_{2}} := I
+%       \end{pmatrix}. g_{q_{1}}
+%   \]
+
+%   The accepting traces through $NFA(I)$ are then described
+%   as,
+
+%   \[
+%     \mathsf{AccTrace}_{NFA(I)} = \LinSigTy q {\{1 , 2\}} {\left( \mathsf{Trace}_{NFA(I)}(q_{1} , q) \pair \mathsf{acc}(q) \right)}
+%   \]
+
+%   A quick inspection of \cref{fig:emptyNFA} reveals that the
+%   only reasonable choice for $q$ is $q_{2}$ --- because
+%   state 2 is accepting while state 1 is not. Therefore,
+
+%   \begin{align*}
+%     \mathsf{AccTrace}_{NFA(I)}
+%     & \cong \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2}) \pair \mathsf{acc}(q_{2}) \\
+%     & \cong \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2}) \pair \top \\
+%     & \cong \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2})
+%   \end{align*}
+
+%   From here, to prove
+%   $I \cong \mathsf{AccTrace}_{NFA(I)}$ it suffices to show
+%   $I \cong \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2})$. Below we
+%   give two parse transformers, one from $I$ to
+%   $\mathsf{Trace}_{NFA(I)}(q_{1} , q_{2})$ and vice versa.
+
+%   \[
+%     \inferrule
+%     {p : I \vdash \mathsf{nil} : \mathsf{Trace}(q_{2} , q_{2}) \\
+%      \exists \text{~transition~} q_{1} \overset{\varepsilon}{\to} q_{2}
+%     }
+%     {p : I \vdash \mathsf{\varepsilon cons}(\mathsf{nil}) : \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2})}
+%   \]
+
+%   Let $\gamma$ be the substitution $\{ g_{q_{2}} / g_{q_{1}}, I / g_{q_{2}} \}$,
+
+%   \[
+%     \inferrule
+%     {
+%       p : \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2}) \vdash p :
+%         \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2}) \\
+%       x_{1} : \simulsubst {g_{q_{2}}} {\gamma} = I \vdash x_{1} : I \\
+%       x_{2} : \simulsubst {I} {\gamma} = I \vdash x_{2} : I
+%     }
+%     {
+%       p : \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2}) \vdash \mathsf{mfold}(x_{1}.x_{1} , x_{2}.x_{2})(p) : I
+%     }
+%   \]
+
+%   This concludes the proof for the case of the empty
+%   grammar. Let's now walk through the construction for
+%   literal grammars. Given a character $c$, we construct an
+%   NFA that recognizes only the string containing the single
+%   character $c$ as,
+
+%   \begin{figure}[h!]
+%   \begin{tikzpicture}[node distance = 25mm ]
+%     \node[state, initial] (1) {$1$};
+%     \node[state, right of=1, accepting] (2) {$2$};
+
+%     \path[->] (1) edge[below] node{$c$} (2);
+%   \end{tikzpicture}
+%   \caption{$NFA(c)$}
+%   \label{fig:literalNFA}
+%   \end{figure}
+
+%   The automaton in \cref{fig:literalNFA} induces the
+%   following type of traces from $q_{1}$ to $q_{2}$.
+
+%   \[
+%     \mathsf{Trace}_{NFA(c)}(q_{1}, q_{2}) = \mu
+%       \begin{pmatrix}
+%          g_{q_{1}} := c \otimes g_{q_{2}} \\
+%          g_{q_{2}} := I
+%       \end{pmatrix}. g_{q_{1}}
+%   \]
+
+%   Through the same argument as the empty grammar, the only
+%   state that is accepting is $q_{2}$ and thus,
+
+%   \[
+%     \mathsf{AccTrace}_{NFA(c)} \cong \mathsf{Trace}_{NFA(c)}(q_{1} , q_{2})
+%   \]
+
+%   To show the desired isomorphism of $c \cong NFA(c)$ we
+%   make a similar argument as we did for the empty grammar
+%   $I$, except we leverage the $\mathsf{cons}$ rule instead
+%   of $\mathsf{\varepsilon cons}$. That is, the parse
+%   transformers in either direction are given as,
+
+%   \[
+%     \inferrule
+%     {\cdot \vdash \mathsf{nil} : \mathsf{Trace}(q_{2} , q_{2}) \\
+%      \exists \text{~transition~} q_{1} \overset{c}{\to} q_{2}
+%     }
+%     {p : c \vdash \mathsf{cons}(\mathsf{nil}) :
+%       \mathsf{Trace}_{NFA(c)}(q_{1} , q_{2})}
+%   \]
+
+%   \[
+%     \inferrule
+%     {
+%       p : \mathsf{Trace}_{NFA(c)}(q_{1} , q_{2}) \vdash p :
+%         \mathsf{Trace}_{NFA(c)}(q_{1} , q_{2}) \\
+%       x_{1} : \simulsubst {(c \otimes g_{q_{2}})} {\gamma} = c \otimes I \vdash \mathsf{unitR}(x_{1}) : c \\
+%       x_{2} : \simulsubst {I} {\gamma} = I \vdash x_{2} : I
+%     }
+%     {
+%       p : \mathsf{Trace}_{NFA(c)}(q_{1} , q_{2}) \vdash \mathsf{mfold}(x_{1}.\mathsf{unitR}(x_{1}) , x_{2}.x_{2})(p) : c
+%     }
+%   \]
+
+%   Where $\gamma$ is the substitution
+%   $\{ c \otimes g_{q_{2}} / g_{q_{1}}, I / g_{q_{2}} \}$ and
+%   $\mathsf{unitR}$ is a witness to the isomorphism
+%   $c \otimes I \cong c$. Again, we may see that these do
+%   indeed mutually invert each other in the Agda code.
+
+%   It remains to show that the desired isomorphisms are
+%   preserved by $\otimes$, $\oplus$, and Kleene star. Here,
+%   we will give the argument for the disjunction case, the
+%   others are defined quite similarly.
+
+%   Given two NFAs $N$ and $M$, we may define a new NFA that
+%   encodes the disjunction of $N$ and $M$. Denote the
+%   internal states of $N$ by $q_{j}$'s and the internal states
+%   of $M$ by $r_{k}$'s,
+
+%   \begin{figure}[h!]
+%   \begin{tikzpicture}[node distance = 20mm ]
+%     \node[state] (2) {$q_{init}$};
+%     \node[state, initial, below left of=2] (1) {$init$};
+%     \node[state, below right of=1] (3) {$r_{init}$};
+%     \node[right of=2] (4) {$\cdots$};
+%     \node[right of=3] (5) {$\cdots$};
+%     % \node[state, below right of=1] (3) {$3$};
+%     % \node[state, below of=2] (4) {$4$};
+
+%     \path[->] (1) edge[below] node{$\varepsilon$} (2);
+%     \path[->] (1) edge[below] node{$\varepsilon$} (3);
+%     \path[->] (2) edge[below] node{} (4);
+%     \path[->] (3) edge[below] node{} (5);
+
+%     \node[label={[name=l] $N$}, draw,line width=2pt,rounded corners=5pt, fit=(2)(4)] {};
+%     \node[label={[name=l] $M$}, draw,line width=2pt,rounded corners=5pt, fit=(3)(5)] {};
+%   \end{tikzpicture}
+%   \caption{$N \oplus_{NFA} M$}
+%   \label{fig:disjunctionNFA}
+%   \end{figure}
+
+%   \Cref{fig:disjunctionNFA} shows the process of
+%   disjunctively combining NFAs. Precisely, we add a single
+%   new state and we included copies of the states from each of $N$
+%   and $M$. The new state acts as the initial state and has
+%   $\varepsilon$-transitions to the initial states of $N$ and
+%   $M$. We include all of the internal transitions from $N$
+%   and $M$, and the accepting states of $N \oplus_{NFA} M$
+%   are exactly the accepting states in each subautomaton.
+
+%   Let $g$ and $g'$ be two regular grammars such that
+%   $g \cong NFA(g)$ and $g' \cong NFA(g')$. As a matter of
+%   notation\footnote{We shall similarly abuse notation for
+%     $\otimes$ and Kleene. That is, for a regular grammar
+%     $g$, when we write $NFA(g)$ we mean the NFA inductively
+%     built up with the NFA-analogues to the constructors that
+%     built up $g$.}, we will write $NFA(g \oplus g')$ for
+%   $NFA(g) \oplus_{NFA} NFA(g')$. The traces of
+%   $NFA(g \oplus g')$ are then given by,
+
+%   \[
+%     \mathsf{Trace}_{NFA(g \oplus g')}(src , dst) = \mu
+%       \begin{pmatrix}
+%         g_{init} := g_{q_{0}} \oplus g_{r_{0}} \\
+%         g_{q_{j}} := \mathsf{Trans}_{NFA(g)}(q_{j}) \oplus \mathsf{isDst}(q_{j}) \\
+%         g_{r_{k}} := \mathsf{Trans}_{NFA(g')}(r_{k}) \oplus \mathsf{isDst}(r_{k})
+%       \end{pmatrix}.g_{src}
+%     \]
+
+%   where $\mathsf{Trans}$ is used to echo the same syntactic
+%   definitions that appear in the $NFA(g)$ and $NFA(g')$.
+%   Also, $src$ and $dst$ may take on any value in
+%   $Q := \{init\} \cup \{q_{j}\} \cup \{r_{k}\}$, and
+%   $\mathsf{isDst}(q)$ checks if $q$ is equal to $dst$. Which
+%   is all to say, the traces of $NFA(g \oplus g')$ comprise
+%   either a trace of $NFA(g)$, or a trace of $NFA(g')$, and
+%   the transition coming out of $g_{init}$ determines which
+%   subautomaton we step into.
+
+%   The parse transformer from $g \oplus g'$ checks which side
+%   of the sum type we are on, then takes the appropriate step
+%   from $g_{init}$ in the automaton.
+
+%   \[
+%     \inferrule
+%     {
+%       u : g \vdash \iota (\phi (u)) : \mathsf{AccTrace}_{NFA(g \oplus g')} \\
+%       v : g' \vdash \iota' (\psi (v)) : \mathsf{AccTrace}_{NFA(g \oplus g')}
+%     }
+%     {p : g \oplus g' \vdash \mathsf{case}~p \{ \mathsf{inl}(u) \mapsto s , \mathsf{inr}(v) \mapsto r \} : \mathsf{AccTrace}_{NFA(g \oplus g')}}
+%   \]
+
+%   with $\iota$ and $\iota'$ as embeddings from $NFA(g)$ and
+%   $NFA(g')$, respectively, into $NFA(g \oplus g')$,
+%   $\phi: g \cong \mathsf{AccTrace}_{NFA(g)}$, and $\psi: g' \cong \mathsf{AccTrace}_{NFA(g')}$. On a
+%   high level, all this construction does is turn a parse of
+%   $g$ into a parse of $NFA(g)$ and then embeds that inside
+%   of the larger automaton $NFA(g \oplus g')$. Likewise for $g'$.
+
+%   In the other direction, recall that the data of an
+%   accepting trace for $NFA(g \oplus g')$ is a pair of a
+%   trace and a proof that
+%   the end state $q'$ of that trace is accepting. By
+%   multifolding over the first part of that pair, we turn the
+%   term of type
+%   $\mathsf{Trace}_{NFA(g \oplus g')}(init , q')$ into a
+%   trace of either of the subautomata,
+
+%   \[
+%     p : \mathsf{Trace}_{NFA(g \oplus g')}(init , q') \vdash \mathsf{mfold}_{NFA(g \oplus g')} : \mathsf{Trace}_{NFA(g)}(q_{0} , q') \oplus \mathsf{Trace}_{NFA(g')}(r_{0} , q')
+%   \]
+
+%   Additionally, we leverage the fact that the only accepting
+%   states for $NFA(g \oplus g')$ are those from the
+%   subautomata to extract that $q'$ must be an accepting
+%   state from a subautomaton.
+
+%   \[
+%     x : \mathsf{acc}_{NFA(g \oplus g')}(q') \vdash M : \mathsf{acc}_{NFA(g)}(q') \oplus \mathsf{acc}_{NFA(g')}(q')
+%   \]
+
+%   We then combine the trace and proof of acceptance into an
+%   accepting trace of one of the subautomata,
+
+%   \[
+%     p : \mathsf{AccTrace}_{NFA(g \oplus g')} \vdash N : \mathsf{AccTrace}_{NFA(g)} \oplus \mathsf{AccTrace}_{NFA(g')}
+%   \]
+
+%   Lastly, we then inductively use the isomorphisms $\phi$
+%   and $\psi$ to turn the accepting traces into a parse of
+%   $g$ or $g'$,
+
+%   \[
+%      N : \mathsf{AccTrace}_{NFA(g)} \oplus \mathsf{AccTrace}_{NFA(g')} \vdash \mathsf{case}~N~\{\mathsf{inl}(n) \mapsto \phi^{-1}(n), \mathsf{inr}(n') \mapsto \psi^{-1}(n')\} : g \oplus g'
+%   \]
+
+%   When you wish to show that these actually maps invert each other to form an isomorphism\max{jumbled sentence}, we invoke the
+%   universal properties of the $\mathsf{Trace}$ types. That is,
+%   $\mathsf{Trace}_{\oplus NFA}$, $\mathsf{Trace}_{N}$, and $\mathsf{Trace}_{M}$
+%   are \emph{initial} algebras to the the functors that define them. Thus, the
+%   embedding of traces through $N$ into $\oplus NFA$ and then mapped back down to $N$
+%   compose to a map from $\mathsf{Trace}_{N}$ to itself. Such a map is unique and
+%   thus must be the identity. The same reasoning shows that the desired
+%   composite do indeed invert each other to form an isomorphism.
+
+%   As discussed above, the other cases and
+%   follow through a similar argument. This
+%   concludes the proof of our variant of Thompson's construction.
+% \end{proof}
+
+% \subsection{Other Automata}
+% \subsubsection{Pushdown Automata}
+% \label{subsubsec:pda}
+% A (nondeterministic) \emph{pushdown automaton} is an automaton that employs a
+% stack. Just like NFAs, they have transitions labeled with characters from a
+% fixed string alphabet $\Sigma$. Additionally, they maintain a stack of
+% characters drawn from a stack alphabet $\Gamma$. They are often represented
+% formally as a 7-tuple $(Q, \Sigma, \Gamma, \delta, q_{0}, Z, F)$,
+
+% \begin{itemize}
+%   \item $Q$ a finite set of states
+%   \item $\Sigma$ a fixed, finite string alphabet
+%   \item $S$ a fixed, finite stack alphabet
+%   \item
+%         $\delta \subset Q \times (\Sigma \cup \{ \varepsilon \}) \times S \to \mathcal{P}(Q \times S^{*})$
+%         the transition function
+%   \item $q_{0} \in Q$ the start state
+%   \item $Z \in S$ the initial stack symbol
+%   \item $F \subset Q$ the accepting states
+% \end{itemize}
+
+% We encode the traces of a pushdown automaton very similarly to those of an NFA,
+% except the transitions of a PDA are instead encoded via the LNL product type
+% $\bigamp$. This is because when simply transitioning via character, a PDA must
+% also pop and push characters onto a stack, which is used as the argument to
+% these dependent functions. Thus, the appropriate formation of traces through a
+% PDA is dependent on a stack.
+
+% Let $S$ be a non-linear type encoding the stack
+% alphabet, and build lists over $S$ as the (non-linear) least fixed-point
+% $\mathsf{List}(S) := \mu X . 1 + S \times X$. Then, the type of states for a
+% PDA $P$ with stack alphabet $S$ are given as a functions that takes in lists $L$,
+% and then makes a case distinction between possible transitions based on what was witnessed as $\mathsf{head}(L)$. The choice of transition
+% will then determine which character to transition by and what word $w$ should be
+% pushed onto the stack. The word that is added to the top of the stack is
+% appended to $\mathsf{tail}(L)$ and then we recursively step into another state
+% called on argument $w \cdot \mathsf{tail}(L)$.
 
@@
 
-\subsubsection{Turing Machines}
-\label{subsubsec:tm}
-
-Above we gave a grammar presentation of
-traces through a PDA by using a non-linear type $S$
-to encode the stack. We may similarly use pairs $S \times S$
-to encode the tape of a Turing machine. With two stacks we can simulate the behavior of the
-infinite tape of a Turing machine. The intuition behind this correspondence is
-that the left half of the tape is on one stack, the right
-half of the tape the other, and we treat the tops of stacks
-like the head of the tape.
-
-\pedro{I like the overall structure of this section! Showing how the type theory
-can encode and reason about different kinds of languages is very instructive!}
+% \subsubsection{Turing Machines}
+% \label{subsubsec:tm}
+
+% Above we gave a grammar presentation of
+% traces through a PDA by using a non-linear type $S$
+% to encode the stack. We may similarly use pairs $S \times S$
+% to encode the tape of a Turing machine. With two stacks we can simulate the behavior of the
+% infinite tape of a Turing machine. The intuition behind this correspondence is
+% that the left half of the tape is on one stack, the right
+% half of the tape the other, and we treat the tops of stacks
+% like the head of the tape.
+
+% \pedro{I like the overall structure of this section! Showing how the type theory
+% can encode and reason about different kinds of languages is very instructive!}
 

### 2024-10-28 paper/paper.tex
categorical grammar related work sketch

@@
 \label{sec:discussion}
+\paragraph{Lambek Calculus and Categorial Grammar}
+In 1958 \cite{lambek1958mathematics}, Joachim Lambek introduced his syntactic
+calculus as a logical system for linguistic derivations. In fact, one may view
+the subtheory of our calculus generated over $\otimes, \lto, \tol$ to be a proof
+relevant presentation of Lambek's original system.
+
+\steven{Should discussion of this connection to lambek calculus be spread
+  throughout the paper?}
+
+Lambek later explicitly connects his syntactic calculus to the structure of a
+non-commutative biclosed monoidal category \cite{lambek1988categorial}. There is
+a rich line of subsequent research after these revelations by Lambek, including
+the study of pregroup grammars
+\cite{coeckeMathematicalFoundationsCompositional2010}, the presentation of
+abstract categorial grammars \cite{degrooteAbstractCategorialParsing2015}, and
+even Lambek calculus with dependent types \cite{luoLambekCalculusDependent}. As
+far as the authors can tell, these works seem to be primarily interested in natural
+language semantics in the style of Montague grammars, rather than parsing formal
+languages. That is, they seem to try to give a functorial semantic
+interpretation of natrual language.
+
+There seem to be some implementations from the abstract categorial grammar
+toolkit, but again they seem natural language focused and not up to the task of
+verified formal language parsing.
+
+Also mention here that \cite{luoLambekCalculusDependent} seems similar at first
+glance, but isn't as similar as it might look because the judgements and
+dependent types seem to
+take on a different interpretation of ours. Say something here like, ``as we
+understand it, it differs in x,y,z ways''
+
+\steven{TODO Edit the above}
+\steven{Even though the above sounds bad, are there lambek/categorial things that are missed?}
 
@@
 
-The core objects of study in the theory of formal grammars have stood stalwart
-for decades. Study of interesting extensions or restrictions of particular
-grammar formalisms has remained consistent line of research over the years.
-Occasionally, researchers will either explicitly create a type system for their
-formalism, or researchers will leverage the same underlying structure of their
-systems. In this paper, we have not
-done much \emph{per se}, other than gather these ideas, abstract over their
-commonalities, and give the common language between formalisms.
-
-\paragraph{Typed Approaches to Grammars}
+\paragraph{Typed Approaches to Regular Expressions}
 The usage of a simple type system to reason about regular expressions was
@@
 by Hengelin and Nielsen in 2011 \cite{henglein_regular_2011} to handle proofs of regular expression
-containment.
-
-In 1992, while investigating the addition of arbitrary recursion to Kleene
+containment. In 1992, while investigating the addition of arbitrary recursion to Kleene
 algebra \cite{leiss},
@@
 he did not explicitly make use of a type system, or have access to the view of Frisch and
-Cardelli, Lei{\ss} had added the full power of inductive types to a type system of regular expressions.
+Cardelli, Lei{\ss} had added the full power of inductive types to a type system
+of regular expressions.
 
@@
 
-In \cite{firsovCertifiedNormalizationContextFree2015}, Firsov and Uustalu
-constructed a verified normalization procedure to turn a
-context-free grammar into an equivalent grammar in Chomsky normal form.
-Combining their work with a CYK parser, they have built a verified context-free
-grammar parser, \emph{up to weak equivalence}. To upgrade from a weak
-equivalence to a strong equivalence, they suggest updating their grammar
-formalism to treat parse trees as first-class proofs of language membership.
-That is, without explicitly stating so, they suggest that they next natural step
-is to reason about grammars in a system where parse trees are terms, as we have
-given in this paper.
-
-\steven{Make the above paragraph shorter. Also idk if its worth including, but I
-like that they practically ask for a type theory}
+Our primary observation in this paper can be summarized as saying that all of
+these above works can embed naturally into a rich, unifying theory.
+
+\steven{I'd like to synthesize the discussion of these works with the categorial
+grammar works, but I am not sure how. It seems as though there are these
+parallel tracks of research that have missed each other for decades}
+
+% In \cite{firsovCertifiedNormalizationContextFree2015}, Firsov and Uustalu
+% constructed a verified normalization procedure to turn a
+% context-free grammar into an equivalent grammar in Chomsky normal form.
+% Combining their work with a CYK parser, they have built a verified context-free
+% grammar parser, \emph{up to weak equivalence}. To upgrade from a weak
+% equivalence to a strong equivalence, they suggest updating their grammar
+% formalism to treat parse trees as first-class proofs of language membership.
+% That is, without explicitly stating so, they suggest that they next natural step
+% is to reason about grammars in a system where parse trees are terms, as we have
+% given in this paper.
+
+% \steven{Make the above paragraph shorter. Also idk if its worth including, but I
+% like that they practically ask for a type theory}
 
@@
 
-Vakar might be worth citing to discuss different characterizations of the
-notion of model \cite{vakarSyntaxSemanticsLinear2015}
+% Vakar might be worth citing to discuss different characterizations of the
+% notion of model \cite{vakarSyntaxSemanticsLinear2015}
+
+% \cite{fu2023twolevellineardependenttype} give a compiler for a two-level
+% linear dependent type theory. This could serve as a template for a
+% compiler implementation of \theoryname that is not just inference rules
+% embedded in Agda.
+
+\paragraph{Implementations in Parsing}
+\steven{Talk about future implementation-side work}
 
@@
 
-\paragraph{Implementations in Parsing}
-Talk about how we have a DFA parser in Agda and want to to get out a regex
-parser. Then context free, etc
-
-future work related to building a usable correct-by-construction parser
-combinator/semantic action combinator library in this language.
-
-Contrast this
-potential library with existing verified parsers. Citations for some in prelim presentation
+%%    - verified imperative implementation: Lambek logic ala separation
+%%      logic?
+\paragraph{Verified Imperative Implementations}
+Lambek logic ala separation logic
 
@@
 to $\Set$, we envision adapting the same philosophy
-to functors from $\String$ to some \emph{category of trees} to craft a syntax
+to functors from a \emph{category of trees} to $\Set$ to craft a syntax
 that natively captures typed elaboration. This suggests an
@@
 
-\steven{condense the future work section?}
-\steven{Put the 2 category stuff here?}
-
-\paragraph{Other Future Work}
-Backtracking parsing: via (delimited) continuations? effect handlers?
-
 \bibliographystyle{plain}

### 2024-10-28 paper/paper.tex
fix build

@@
 To gain intuiton for working in \theoryabbv, we study some examples of
-grammars and parses of them internal to \theoryabbc. In each of our examples, we will look at
+grammars and parses of them internal to \theoryabbv. In each of our examples, we will look at
 grammars over the three character alphabet $\{ a , b , c\}$.
@@
 \data &~\Trace_{N}~2~(acc : \Bool) : L_{i}~\where\\
-      & \mathsf{stop}_{2} : \ltonol {\Trace_{N}~2~\false} \\
+      & \mathsf{stop}_{2} : \ltonl {\Trace_{N}~2~\false} \\
       & \mathsf{2to2} : \forall (acc : \Bool) \to \ltonl {(a \lto \Trace_{N}~2~acc \lto \Trace_{N}~2~acc)} \\

### 2024-10-28 paper/paper.tex
some comments

@@
 
+\max{Is this really true?
+I don't think it's accurate to say that the ordering restriction on variable usage actually dictates an order in processing variables.
+And I don't know what you mean by processed without modification.}
 Without the above structural rules we
@@
 For a grammar $A$, the recursive nature of $A^{*}$ is captured by defining
-$A^{*}$ to be an inductive\footnote{As we will soon see, our theory defines
+$A^{*}$ to be an \max{this footnote seems unnecessary}inductive\footnote{As we will soon see, our theory defines
   \emph{indexed} inductive types. We can recover non-indexed inductive types by
@@
 \steven{I probably can't get away with the inline definition of $f$ in \cref{fig:kleeneabstractproof}}
+\max{shouldn't the fold be applied to $x$ and shouldn't $f$ be $\lambda x.\lambda y.\cdots$?}
 \begin{figure}
@@
 integral to our design of \theoryabbv ; however, there are some aspects of parsing that
-cannot be captured through linear types alone. Often one wants to define a
+cannot be captured through linear types\max{not exactly clear what you mean, maybe you mean ``simple'' linear types?} alone. Often one wants to define a
 \emph{family} of linear types varying over a set, which is a primary motivator
-for us to make \theoryabbv a linear-non-linear theory. For the NFA $N$, an indexed inductive
+for us to make \theoryabbv a linear-non-linear theory\max{jargon-y. Be more direct}. For the NFA $N$, an indexed inductive
 type of traces through $N$ is given in \cref{fig:nfainductive}.
@@
 type of traces starting at state $s$ and terminating at an arbitrary state. The Boolean flag $acc$ marks if the trace
-ends in either an accepting or a rejecting state. Let us dissect the structure
+ends in either an accepting or a rejecting state\max{confusing. One sentence you say arbitrary state and then the next sentence you say it has to match $acc$}. Let us dissect the structure
 of this type in more detail for state 1 of $N$.
@@
 
+\max{this notation is inconsistent with Agda syntax. You write $acc : \Bool$ but then in the $\mathsf{stop}$ cases you instantiate $\acc$ with a specific boolean}
+\max{is the boolean just overcomplicating stuff here?}
 \begin{figure}
@@
 
+\max{was this paragraph supposed to be deleted?}
 Next we present the full details of the syntax and typing of \theoryname.
@@
 In order to avoid a notion of raw-term, we present it in an intrinsically-typed fashion,
-where only well-typed terms are meaningful. Next, we show how it admits a model in $\Set^{\String}$
+where only well-typed terms are meaningful\max{this sentence says the same thing multiple times}. Next, we show how it admits a model in $\Set^{\String}$
 where the linear types have natural interpretation in terms of formal grammars --- for instance,
 the tensor corresponds to grammar concatenation. We conclude the section by showing how
-classes of grammars correspond to certain sub-type-theories inside \theoryabbv.
+classes of grammars correspond to certain sub-type-theories inside \theoryabbv.\max{why are you putting the entire paper into this section?}
 
@@
 rules in \cref{fig:inttyping}. These are mostly just as they appear in
-\cite{krishnaswami_integrating_2015}. It has two main differences. The
+\cite{krishnaswami_integrating_2015}.
+It has two main differences\max{you absolutely cannot just describe the differences between our calculus and one from prior work.
+This paper needs to be self-contained}.
+The
 first one it the presence of two-distinct arrow types, one for adding
 variables to the beginning of contexts and another one for adding
-variables to the end of contexts. This is an adequate change in the
+variables to the end of contexts. This is an adequate\max{adequate?} change in the
 context of grammars, which are inherently non-symmetric. The second
@@
 \label{sec:formaltype}
+\max{this should be in its own section on semantics}
 
@@
 %   adding this right after ``and the intuitionistic typing rules in fig 3''}
-
 Now that we have specified the syntactic aspect of \theoryabbv, we must justify
@@
 
-\paragraph{Universal Type}
+\paragraph{Universal Type}\max{this is absolutely not the right terminology. Universal type means multiple other completely different things}
 \steven{Definable as nullary product}

### 2024-10-30 paper/paper.tex
move semantic stuff to semantics section (unedited for now)
fix inference rules, modify examples

@@
 \newcommand{\equalizer}[3]{\{#1\,|\,\applto {#2}{#1} = \applto{#3}{#1} \}}
-\newcommand{\equalizerpair}[2]{\langle #1 , #2 \rangle}
+\newcommand{\equalizerin}[1]{\langle #1 \rangle}
 \newcommand{\equalizerpi}[1]{#1.\pi}
@@
 \paragraph{Finite Grammars}
+\newcommand{\fingram}{$(a \otimes b) \oplus c$}
+
 First consider finite grammars --- those built from base types via disjunctions $\oplus$ and
-concatenations $\otimes$. For instance, $(a \oplus b) \otimes c$.
+concatenations $\otimes$. For instance, \fingram.
 In \cref{fig:fingram} we find a proof tree that constructs a term of type
-$(a \oplus b) \otimes c$ in the context $x : a, y : c$. We think of this term as
+\fingram in the context $x : a, y : b$. We think of this term as
 a parse tree internal to \theoryabbv.
@@
 parse \stringquote{ac}, we build parse trees in the context
-$\internalize{\stringquote{ac}} = x : A , y : C$. Thus the term
-$\inl(x) \otimes y$ from \cref{fig:fingram} details the
-manner by which \stringquote{ac} matches the grammar $(a \oplus b) \otimes c$.
-
-\begin{figure}
-\begin{mathpar}
-  \inferrule
-  {
-    \inferrule
-    {~}
-    {x : a \vdash \inl(x) : a \oplus b}
-    \\
-    \inferrule
-    {~}
-    {y : c \vdash y : c}
-  }
-  {x : a , y : c \vdash \inl(x) \otimes y : (a \oplus b) \otimes c}
-\end{mathpar}
-\caption{\stringquote{ac} matches $(a \oplus b) \otimes c$}
-\label{fig:fingram}
-\end{figure}
-
-Because derivations in a linear context correspond to parse trees in this
-manner, we cannot affirm all of the usual structural rules of a type system.
-Weakening and contraction each allow for modification of the string in context ---
-weakening allows for characters to be dropped at will, while contraction allows
-for duplication of characters. Omission of these rules offers a
-resource-sensitive view that ensures only the inputted characters are parsed.
-Moreover, strings are not simply sets of characters, as they carry an inherent
-ordering. Thus, the rule of exchange is also omitted in this calculus so that
-the constituent characters of a string are not freely permuted.
-
-\max{Is this really true?
-I don't think it's accurate to say that the ordering restriction on variable usage actually dictates an order in processing variables.
-And I don't know what you mean by processed without modification.}
-Without the above structural rules we
-arrive at an ordered, linear type system, and to summarize this
-particular choice of substructural restrictions: when parsing, characters must
-be processed
-without modification and in the order in which they were written.
-
-\paragraph{Regular Expressions}
-Next, we demonstrate how to parse a regular expression in \theoryabbv. Similar
-to finite grammars, regular
-expresssions are built inductively using base types, $\oplus$, and $\otimes$ ;
-however, unlike finite grammars, regular expressions additionally have a
-restricted form of recursion via Kleene star $(\cdot)^{\ast}$.
-
-% TODO this should be beter formatted but can't use listings with \otimes ^{\ast}
-\begin{figure}
-\begin{align*}
-\data &~A^{*}: L_{i}~\where\\
-      & \nil : \ltonl {A^{*}} \\
-      & \cons : \ltonl {(A \lto A^{*} \lto A^{*})}
-\end{align*}
-\caption{Kleene Star as an inductive type}
-\label{fig:kleenestarinductive}
-\end{figure}
-
-For a grammar $A$, the recursive nature of $A^{*}$ is captured by defining
-$A^{*}$ to be an \max{this footnote seems unnecessary}inductive\footnote{As we will soon see, our theory defines
-  \emph{indexed} inductive types. We can recover non-indexed inductive types by
-  indexing over the unit type.} linear
-type, as shown in \cref{fig:kleenestarinductive}. Just as in \cref{fig:fingram}, we can show that particular strings match a Kleene
-star. In \cref{fig:kleenestarderivation}, we demonstrate that \stringquote{ab}
-matches $a^{*} \otimes b$ through repeated application of the Kleene star constructors.
+$\internalize{\stringquote{ab}} = x : a , y : b$. Thus the term
+$\inl(x \otimes y)$ from \cref{fig:fingram} details the
+manner by which \stringquote{ac} matches the grammar \fingram.
 
@@
       \inferrule
+      {~}
+      {y : b \vdash y : b}
+    }
+    {x : a , y : b \vdash x \otimes y : a \otimes b}
+  }
+  {x : a , y : b \vdash \inl(x \otimes y) : \fingram}
+\end{mathpar}
+\caption{\stringquote{ac} matches $(a \oplus b) \otimes c$}
+\label{fig:fingram}
+\end{figure}
+
+Because derivations in a linear context correspond to parse trees in this
+manner, we cannot affirm all of the usual structural rules of a type system.
+Weakening and contraction each allow for modification of the string in context ---
+weakening allows for characters to be dropped at will, while contraction allows
+for duplication of characters. Omission of these rules offers a
+resource-sensitive view that ensures only the inputted characters are parsed.
+Moreover, strings are not simply sets of characters, as they carry an inherent
+ordering. Thus, the rule of exchange is also omitted in this calculus so that
+the constituent characters of a string are not freely permuted. Without the above structural rules we
+arrive at an ordered, linear type system.
+
+\paragraph{Regular Expressions}
+Next, we demonstrate how to parse a regular expression in \theoryabbv. Similar
+to finite grammars, regular
+expresssions are built inductively using base types, $\oplus$, and $\otimes$ ;
+however, unlike finite grammars, regular expressions additionally have a
+restricted form of recursion via Kleene star $(\cdot)^{\ast}$.
+
+% TODO this should be beter formatted but can't use listings with \otimes ^{\ast}
+\begin{figure}
+\begin{align*}
+\data &~A^{*}~\where\\
+      & \nil : \ltonl {A^{*}} \\
+      & \cons : \ltonl {(A \lto A^{*} \lto A^{*})}
+\end{align*}
+\caption{Kleene Star as an inductive type}
+\label{fig:kleenestarinductive}
+\end{figure}
+
+For a grammar $A$, the recursive nature of $A^{*}$ is captured by defining
+$A^{*}$ to be an inductive linear
+type, as shown in \cref{fig:kleenestarinductive}. $A^{*}$ has two constructors:
+$\nil$, which builds a parse of type $A^{*}$ in the empty context; and $\cons$,
+which takes in of $A$ and a parse of $A^{*}$ and linearly combines them to build
+a parse of $A^{*}$.
+
+The arrow, $\uparrow$, wrapping these constructors denotes
+an interaction between the linear and nonlinear parts of the theory.
+In short, a nonlinear term
+$M : \ntonl A^{*}$ denotes a linear term in the empty context
+$\cdot \vdash M : A^{*}$. Later in
+this paper, we will further discuss the behavior of $\uparrow$.
+At this moment,
+it suffices to read these arrows as allow us to use the reuse the names
+``$\nil$'' and ``$\cons$''
+as symbols for contructors. That is, the $\nil$ constructor is not linearly
+consumed at usage time
+
+Just as in \cref{fig:fingram}, we can show that particular strings match a Kleene
+star. In \cref{fig:kleenestarderivation}, we demonstrate that \stringquote{ab}
+matches $(a^{*} \otimes b) \oplus c$ through repeated application of the Kleene star constructors.
+
+\begin{figure}
+\begin{mathpar}
+\inferrule
 {
   \inferrule
-        {~}
-        {I \vdash \nil : a^{*}}
-      }
-      {\cdot \vdash \nil : a^{*}}
-    }
-    {x : a \vdash \cons (x , \nil) : a^{*}}
+  {
+    \inferrule
+    {
+      \inferrule
+      {
+        \inferrule
+        {\cdot \vdash \cons : \ltonl {(a \lto a^* \lto a^*)}}
+        {\cdot ; \cdot \vdash \cons : a \lto a^* \lto a^*}
         \\
@@
         {~}
-    {y : b \vdash y : b}
+        {\cdot ; x : a \vdash x : a}
       }
-  {x : a , y : b \vdash \cons(x ) \otimes y : a^{*} \otimes b}
+      {\cdot ; x : a \vdash \cons~x : a^* \lto a^*}
+      \\
+      \inferrule
+      {~}
+      {\cdot ; \cdot \vdash \nil : a^*}
+    }
+    {\cdot ; x : a \vdash \cons (x , \nil) : a^{*}}
+    \\
+    \inferrule
+    {~}
+    {\cdot ; y : b \vdash y : b}
+  }
+  {\cdot ; x : a , y : b \vdash \cons(x, \nil) \otimes y : a^{*} \otimes b}
+}
+{\cdot ; x : a , y : b \vdash \inl (\cons(x) \otimes y) : (a^{*} \otimes b) \oplus c}
 \end{mathpar}
-\caption{\stringquote{ab} matches $a^{*} \otimes b$}
+\caption{\stringquote{ab} matches $(a^{*} \otimes b) \oplus c$}
 \label{fig:kleenestarderivation}
@@
 \max{shouldn't the fold be applied to $x$ and shouldn't $f$ be $\lambda x.\lambda y.\cdots$?}
+
 \begin{figure}
@@
     {~}
-    {x_I : I \vdash \nil : A^*}
+    {\cdot \vdash \nil : A^*}
     \\
@@
       \inferrule
-      {~}
-      {x_1 : A , x_2 : A , y : A* \vdash \cons (x_1 , \cons(x_2 , y)) : A^*}
+      {x_1 : A, x_2 : A , y : A^* \vdash \cons (x_1 , \cons(x_2 , y)) : A^*}
+      {x' : A \otimes A , y : A^* \vdash g := \letin {x_1 \otimes x_2} {x'} {\cons (x_1 , \cons(x_2 , y)) : A^*}}
     }
-    {x : A \otimes A , y : A^*  \vdash f := \letin {x_1 \otimes x_2} {x} {\cons(x_1 ,
-        \cons(x_2 , y))} : A^*}
+    {\cdot \vdash f := \lamblto x {\lamblto y {\app {\app g x} {y} }} : (A \otimes A) \lto A^* \lto A^*}
   }
-  {x : (A \otimes A)^* \vdash \fold(\nil , f) : A^* }
+  {x : (A \otimes A)^* \vdash \fold(\nil , f)(x) : A^* }
 \end{mathpar}
@@
   \begin{tikzpicture}[node distance = 25mm ]
-    \node[state, initial, accepting] (1) {$1$};
+    \node[state, initial] (1) {$1$};
     \node[state, below left of=1] (2) {$2$};
-    \node[state, right of=2] (3) {$3$};
-
-    \path[->] (1) edge[above] node{$b$} (2)
-              (1) edge[below, bend right, left=0.3] node{$\epsilon$} (3)
+    \node[state, right of=2, accepting] (3) {$3$};
+
+    \path[->] (1) edge[above] node{$\epsilon$} (2)
+              (1) edge[right] node{$c$} (3)
               (2) edge[loop left] node{$a$} (2)
-              (2) edge[below] node{$b$} (3)
-              (3) edge[above, bend right, right=0.3] node{$a$} (1);
+              (2) edge[below] node{$b$} (3);
   \end{tikzpicture}
-  \caption{NFA $N$}
+  \caption{NFA for $(a^{*} \otimes b) \oplus c$}
   \label{fig:exampleNFA}
@@
 
-\max{this notation is inconsistent with Agda syntax. You write $acc : \Bool$ but then in the $\mathsf{stop}$ cases you instantiate $\acc$ with a specific boolean}
+\max{this notation is inconsistent with Agda syntax. You write $acc : \Bool$ but then in the $\mathsf{stop}$ cases you instantiate $acc$ with a specific boolean}
 \max{is the boolean just overcomplicating stuff here?}
+
 \begin{figure}
 \begin{align*}
-\data &~\Trace_{N}~1~(acc : \Bool) : L_{i}~\where\\
-      & \mathsf{stop}_{1} : \ltonl {\Trace_{N}~1~\true}  \\
-      & \mathsf{1to2} : \forall (acc : \Bool) \to \ltonl {(b \lto \Trace_{N}~2~acc \lto \Trace_{N}~1~acc)} \\
-      & \mathsf{1to3} : \forall (acc : \Bool) \to \ltonl {(\Trace_{N}~3~acc \lto \Trace_{N}~1~acc)} \\
-\\
-\data &~\Trace_{N}~2~(acc : \Bool) : L_{i}~\where\\
-      & \mathsf{stop}_{2} : \ltonl {\Trace_{N}~2~\false} \\
-      & \mathsf{2to2} : \forall (acc : \Bool) \to \ltonl {(a \lto \Trace_{N}~2~acc \lto \Trace_{N}~2~acc)} \\
-      & \mathsf{2to3} : \forall (acc : \Bool) \to \ltonl {(b \lto \Trace_{N}~3~acc \lto \Trace_{N}~2~acc)}  \\
-\\
-\data &~\Trace_{N}~3~(acc : \Bool) : L_{i}~\where\\
-      & \mathsf{stop}_{3} : \ltonl {\Trace_{N}~3~\true} \\
-      & \mathsf{3to1} : \forall (acc : \Bool) \to \ltonl {(a \lto \Trace_{N}~1~acc \lto \Trace_{N}~3~acc)} \\
+\data &~\Trace : (s : N.\mathsf{state}) \to L~\where\\
+      & \mathsf{stop} : \ltonl {(\Trace~3)}  \\
+      & \mathsf{1to2} : \ltonl {(\Trace~2 \lto \Trace~1)} \\
+      & \mathsf{2to2} : \ltonl {(a \lto \Trace~2 \lto \Trace~2)} \\
+      & \mathsf{2to3} : \ltonl {(b \lto \Trace~3 \lto \Trace~2)}  \\
+      & \mathsf{1to3} : \ltonl {(c \lto \Trace~3\lto \Trace~1)} \\
 \end{align*}
@@
   \begin{mathpar}
-    \boxed{\Gamma ; \Delta \vdash a : A}
+    \boxed{\Gamma ; \Delta \vdash e : A}
 
@@
     \inferrule
-    {~}
-    {\Gamma , x : \ltonl A ; \cdot \vdash x : A}
+    {\Gamma \vdash M : \ltonl A}
+    {\Gamma ; \cdot \vdash M : A}
     %
@@
     \inferrule
-    {\Gamma \vdash f : \ltonl {(A \lto B)} \\
-      \Gamma \vdash g : \ltonl { (A \lto B) } \\
-      \Gamma ; \Delta \vdash a : A \\
-      \Gamma ; \Delta \vdash p : \applto {f}{a} = \applto {g}{a}}
-    {\Gamma ; \Delta \vdash \equalizerpair{a}{p} : \equalizer {a}{f}{g}}
+    { \Gamma ; \Delta \vdash e : A \\
+      \Gamma ; \Delta \vdash \applto {f}{e} \equiv \applto {g}{e}}
+    {\Gamma ; \Delta \vdash \equalizerin{e} : \equalizer {e}{f}{g}}
     %
@@
     \inferrule
-    {\Gamma \vdash f : \ltonl { (A \lto B) } \\
-      \Gamma \vdash g : \ltonl { (A \lto B) } \\
-      \Gamma ; \Delta \vdash e : \equalizer{a}{f}{g}}
+    {\Gamma ; \Delta \vdash e : \equalizer{a}{f}{g}}
     {\Gamma ; \Delta \vdash \equalizerpi {e} : A}
-    %% \\
-    %% %
-    %% \inferrule{~}{\Gamma ; \Delta \vdash () : \top}
-
-    %% \and
-
-    %% \inferrule{~}{\Gamma ; x : \bot \vdash \mathsf{absurd}_A : A}
-    %% %
-    %% \and
-    %% %
-    %% %
-    %% \inferrule{\Gamma \vdash e : \mathsf{G} A}{\Gamma ; \cdot \vdash \mathsf{G}^{-1}\, e : A}
-    %% %
   \end{mathpar}
@@
     \inferrule
-    {\Gamma ; \Delta \vdash f : A \\ \Gamma \vdash g : \ltonl { (A \lto B) }}
+    {\Gamma ; \Delta \vdash f : \ltonl {(A \lto B)} \\ \Gamma \vdash g : \ltonl { (A \lto B) }}
     {\Gamma ; \Delta \vdash \equalizer {a}{f}{g} : L_i}
@@
   \begin{mathpar}
-  \boxed{\Gamma \vdash x : X}
+  \boxed{\Gamma \vdash M : X}
 
@@
   %
-  \inferrule{\Gamma \vdash e : Y \quad \ctxwffjdg \Gamma {X \equiv Y}}{\Gamma \vdash e : X}
+  \inferrule{\Gamma \vdash M : Y \quad \ctxwffjdg \Gamma {X \equiv Y}}{\Gamma \vdash M : X}
   %
@@
   %
-  \inferrule{\Gamma \vdash e : X \\ \Gamma \vdash e : \subst Y e x}{\Gamma \vdash (e, e') : \SigTy x X Y}
-  %
-  \and
-%
-  \inferrule{\Gamma \vdash e : \SigTy x X Y}{\Gamma \vdash \pi_1\, e : X}
-  %
-  \and
-  %
-  \inferrule{\Gamma \vdash e : \SigTy x X Y}{\Gamma \vdash \pi_2\, e : \subst Y {\pi_1\, e} x}
-  \and
-  \inferrule{\Gamma, x : X \vdash e : Y}{\Gamma \vdash \lamb x e : \PiTy x X Y}
-  %
-  \and
-  %
-  \inferrule{\Gamma \vdash e : \PiTy x X Y \\ \Gamma \vdash e' : X}{\Gamma \vdash \app e {e'} : \subst Y {e'} x}
-  %
-  \and
-  %
-  \inferrule{\Gamma \vdash e \equiv e' : X}{\Gamma \vdash \mathsf{refl} : e =_X e'}
+  \inferrule{\Gamma \vdash M : X \\ \Gamma \vdash N : \subst Y M x}{\Gamma \vdash (M, N) : \SigTy x X Y}
+  %
+  \and
+%
+  \inferrule{\Gamma \vdash M : \SigTy x X Y}{\Gamma \vdash \pi_1\, M : X}
+  %
+  \and
+  %
+  \inferrule{\Gamma \vdash M : \SigTy x X Y}{\Gamma \vdash \pi_2\, M : \subst Y {\pi_1\, M} x}
+  \and
+  \inferrule{\Gamma, x : X \vdash M : Y}{\Gamma \vdash \lamb x M : \PiTy x X Y}
+  %
+  \and
+  %
+  \inferrule{\Gamma \vdash M : \PiTy x X Y \\ \Gamma \vdash N : X}{\Gamma \vdash \app M {N} : \subst Y {N} x}
+  %
+  \and
+  %
+  \inferrule{\Gamma \vdash M \equiv N : X}{\Gamma \vdash \mathsf{refl} : M =_X N}
   \and
@@
   \begin{mathpar}
-    \boxed{\Gamma \vdash e \equiv e' : X}
-
-    \inferrule{\Gamma \vdash p : e =_X e'}{\Gamma \vdash e \equiv e' : X}
-%
-    \and
-%
-    \inferrule{~}{\Gamma \vdash \app {(\lamb x e)} {e'} \equiv \subst x e {e'} : X}
-%
-    \and
-%
-    \inferrule{~}{\Gamma \vdash e \equiv \lamb x {\app e x} : \PiTy x X Y}
-%
-    \and
-%
-    \inferrule{~}{\Gamma \vdash \pi_1\, (e_1, e_2) \equiv e_1 : X}
-%
-    \and
-%
-    \inferrule{~}{\Gamma \vdash \pi_2\, (e_1, e_2) \equiv e_2 : \subst x {e_1} Y}
-%
-    \and
-%
-    \inferrule{~}{\Gamma \vdash e \equiv (\pi_1\, e, \pi_2\, e) : \SigTy x X Y}
-%
-    \and
-%
-    \inferrule{~}{}
-%
-    \inferrule{~}{\Gamma \vdash t \equiv t' : 1}
+    \boxed{\Gamma \vdash M \equiv N : X}
+
+    \inferrule{\Gamma \vdash p : M =_X N}{\Gamma \vdash M \equiv N : X}
+%
+    \and
+%
+    \inferrule{~}{\Gamma \vdash \app {(\lamb x M)} {N} \equiv \subst x M {N} : X}
+%
+    \and
+%
+    \inferrule{~}{\Gamma \vdash m \equiv \lamb x {\app M x} : \PiTy x X Y}
+%
+    \and
+%
+    \inferrule{~}{\Gamma \vdash \pi_1\, (M, N) \equiv M : X}
+%
+    \and
+%
+    \inferrule{~}{\Gamma \vdash \pi_2\, (M , N) \equiv N : \subst x {M} Y}
+%
+    \and
+%
+    \inferrule{~}{\Gamma \vdash M \equiv (\pi_1\, M, \pi_2\, M) : \SigTy x X Y}
+%
+    \and
+%
+    % \inferrule{~}{}
+%
+    \inferrule{~}{\Gamma \vdash M \equiv N : 1}
 %
@@
 
-    \boxed{\Gamma ; \Delta \vdash a \equiv a' : A}
+    \boxed{\Gamma ; \Delta \vdash e \equiv e' : A}
 
@@
 %
-    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lamblto a e)} {e'} \equiv \subst e x {e'} : C}
+    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lamblto a e)} {e'} \equiv \subst e {e'} a : C}
 %
@@
 %
-    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lambtol a e)} {e'} \equiv \subst e x {e'} : C}
+    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lambtol a e)} {e'} \equiv \subst e {e'} a : C}
 %
@@
 %
-    \inferrule{~}{\Gamma; \Delta \vdash \app {(\dlamb x e)} {e'} \equiv \subst x a {e'} : C}
+    \inferrule{~}{\Gamma; \Delta \vdash \app {(\dlamb x e)} {M} \equiv \subst {e} M x : C}
 %
@@
 %
-    \inferrule{~}{\Gamma; \Delta \vdash e \equiv e' : \top}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash e_i \equiv \pi_i (e_1, e_2) : A_i}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash e \equiv (\pi_1 e, \pi_2 e) : A\& B}
+%     \inferrule{~}{\Gamma; \Delta \vdash e \equiv e' : \top}
+% %
+%     \and
+%
+%     \inferrule{~}{\Gamma; \Delta \vdash e_i \equiv \pi_i (e_1, e_2) : A_i}
+% %
+%     \and
+% %
+%     \inferrule{~}{\Gamma; \Delta \vdash e \equiv (\pi_1 e, \pi_2 e) : A\& B}
 %
@@
     %
-    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} e {\subst {e'} {()} a} \equiv \subst {e'} a e : C}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash \letin {e \otimes e'} {a \otimes a'} e'' \equiv \subst {e''} {a, a'} {e, e'} : C}
-%
-    \and
-    %
-    \inferrule{~}{\Gamma; \Delta \vdash \letin {a \otimes b} e {\subst {e'} {a \otimes b} c} \equiv \subst {e'} c e : C}
-%
-    \and
-%
-    \inferrule{~}{\Gamma;\Delta \vdash \letin {(x, a)} {(e, e')} {e''} \equiv \subst {e''} {x, a} {e, e'} : C}
+    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} e {\subst {e'} {()} a} \equiv \subst {e'} e a : C}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash \letin {a \otimes b} {e \otimes e'} e'' \equiv e'' \{ e/a , e'/b \} : C}
+%
+    \and
+    %
+    \inferrule{~}{\Gamma; \Delta \vdash \letin {a \otimes b} e {\subst {e'} {a \otimes b} c} \equiv \subst {e'} e c : C}
+%
+    \and
+%
+    \inferrule{~}{\Gamma;\Delta \vdash \letin {(x, a)} {(M, e)} {e'} \equiv e' \{ M/x , e/a \} : C}
     %
@@
     \inferrule
-    {\Gamma \vdash f : \ltonl { (A \lto B) } \\
-      \Gamma \vdash g : \ltonl { (A \lto B) } \\
-      \Gamma ; \Delta \vdash a : A \\
-      \Gamma ; \Delta \vdash p : \apptol {f}{a} \equiv \apptol {g}{a}}
-    {\Gamma ; \Delta \vdash \equalizerpi {\equalizerpair {a}{p}} \equiv a : A}
+    {~}
+    {\Gamma ; \Delta \vdash \equalizerpi {\equalizerin {e}} \equiv e : A}
     %
@@
     \inferrule
-    {\Gamma \vdash f : \ltonl { (A \lto B) } \\
-      \Gamma \vdash g : \ltonl { (A \lto B) } \\
-      \Gamma ; \Delta \vdash p : \apptol {f}{a} = \apptol {g}{a} \\
-      \Gamma ; \Delta \vdash e : \equalizer{a}{f}{g}}
-    {\Gamma ; \Delta \vdash \equalizerpair {\equalizerpi {e}} {p} \equiv e : A}
+    {~}
+    {\Gamma ; \Delta \vdash \equalizerin {\equalizerpi {e}} \equiv e : A}
 \end{mathpar}
@@
   judgemental equality figure?}
-
-\subsection{Formal Grammars as Linear Types}
-\label{sec:formaltype}
-\max{this should be in its own section on semantics}
-
-% The linear typing judgment in our syntax takes on the following schematic form
-% $\Gamma ; \Delta \vdash a : A$. First, $A$ represents a \emph{linear type} in
-% our syntax. The intended semantics of these linear types are formal grammars.
-% That is, the linear typing system is designed to syntactically reflect the
-% behavior of formal grammars. For this reason, we may often interchangeably use
-% the terms ``linear type'' and ``grammar''.
-
-% The term $a$ is an inhabitant of type $A$, which is thought of as a parse tree of
-% the grammar $A$. The core idea of this entire paper follows precisely from this
-% single correspondence: grammars are types, and the inhabitants of these types
-% are parse trees for the grammars.
-
-% $\Gamma$ represents a non-linear context, while $\Delta$ represents a
-% \emph{linear context} dependent on $\Gamma$. These linear contexts behave
-% substructurally. As stated earlier, they are linear --- they do not obey
-% weakening or contraction --- because a character is exhausted once read by a
-% parsing procedure. Moreover, the characters in strings appear in the order in
-% which we read them. We do not have the freedom to freely permute characters,
-% therefore any type theory that is used to reason about formal grammars ought to
-% omit the structural rule of exchange as well. This means that every variable
-% within $\Delta$ must be used \emph{exactly once} and \emph{in order of
-%   occurrence}. Thus, we can think of the linear contexts as an ordered list of
-% limited resources. Once a resource is consumed, we cannot make reference to it
-% again. Variables in a linear context then act like building
-% blocks for constructing patterns over strings.
-
-% We give the base types and type constructors for linear terms. As the
-% interpretation of types as grammars in $\Grammar$ serves as our intended
-% semantics, we simultaneously give the interpretation $\sem \cdot$ of the
-% semantics as grammars.
-
-% \pedro{The paragraphs above are a bit circular, what about the following alternative?}
-% \steven{I am in favor of cutting what's above, but we haven't introduced the
-%   syntax in the main body of the paper yet, even if the figures appeared
-%   earlier. So what you write needs to take that into account}
-% \pedro{In which case we should introduce it in the previous section. I suggest
-%   adding this right after ``and the intuitionistic typing rules in fig 3''}
-Now that we have specified the syntactic aspect of \theoryabbv, we must justify
-its connections to formal grammars in order to show that it is a sound methodology
-for reasoning about parsers. We do this by showing that $\mathbf{Gr}$ is a model
-to our type theory.
-
-Most of the non-linear semantics is standard \pedro{We should either cite something or add the non-linear semantics in the appendix}, so let us investigate in more
-depth the linear semantics.
-
-\begin{enumerate}
-  \item A non-linear context $\Gamma$ denotes a set $\sem \Gamma$
-  \item A non-linear type $\Gamma \vdash X : U_{i}$ denotes a family of sets
-        $\sem X : \sem \Gamma \to \Set_{i}$
-  \item A non-linear term $\Gamma \vdash e : X$ denotes a section
-        $\sem e : \Pi(\gamma : \sem \Gamma)\sem{X} \gamma$ \pedro{we have not defined what sections are}
-  \item Linear contexts $\Gamma \vdash \Delta$ and types
-        $\Gamma \vdash A : L_{i}$ both denote families of grammars
-        $\sem \Gamma \to \Gr_{i}$
-  \item A linear term $\Gamma ; \Delta \vdash M : A$ denotes a family of parse
-        transformers
-        $\sem M : \Pi(\gamma : \sem \Gamma)\Pi(w : \String) \sem \Delta \gamma w \Rightarrow \sem M \gamma w$
-\end{enumerate}
-
-\paragraph{Linear Unit}
-First, the linear unit $I$ may be built in the empty linear
-context\footnote{When appropriate, statements like ``the empty linear context''
-  may be taken polymorphically over all non-linear contexts that they might
-  depend on.} $\cdot \dashv I$. $I$ serves as the unit for the operation
-$\otimes$ described below.
-
-As a grammar,
-%
-\[
-  \sem {I} \gamma w = \{ () ~|~ w = \varepsilon \}
-\]
-%
-That is, $I$ maps strings to the proposition that they are the empty string. $I$
-is clearly only inhabited by the empty string $\varepsilon$, for which the
-outputted set contains a single parse tree.
-
-\paragraph{Base Types}
-For each character $c$ in the alphabet $\Sigma$, we include a base type at the
-lowest universe level $c : L_{0}$.
-
-The grammar interpretation for characters is quite similar to that of $I$.
-%
-\[
-  \sem {c} \gamma w = \{ () ~|~ w = c \}
-\]
-%
-The grammar for the character $c$ likewise maps strings $w$ to the proposition that
-they are equal to $c$.
-
-\paragraph{Tensor Product}
-The tensor product of two linear types is the first place where the ordering on
-contexts really takes effect. That is, the tensor product of two types is
-non-commutative.
-When context $\Delta$ forms a linear term of
-type $A$, and $\Delta'$ forms a linear term of type $B$, then the context
-extension $\Delta, \Delta'$ forms a linear term of type $A \otimes B$.
-
-As stated above, the type $I$ serves as the unit for $\otimes$. That is, for all
-linear types $A$, the following equalities hold:
-%
-\[ I \otimes A \equiv A \otimes I \equiv A \]
-%
-In the grammar semantics,
-%
-\[
-  \sem {A \otimes B} = \Sigma_{w_{1}, w_{2} : \String} (w_{1}w_{2} = w) \land \left(  \sem {A} \gamma w_{1} \times \sem {B} \gamma w_{2} \right)
-\]
-%
-The string $w$ matches $A \otimes B$
-precisely when it may be split into two
-pieces such that the left one matches $A$
-and the right one matches $B$.
-
-\paragraph{Linear Function Types}
-The monoidal structure provided by $\otimes$ is both left and right closed and
-this is denoted by the left and right linear function types. Further,
-for linear types $A$ and $B$ the left linear function type $A \lto B$
-enjoys an elimination principle similar to function application or modus ponens.
-%
-\[
-\inferrule{\Delta \vdash A \otimes A \lto B}{\Delta \vdash B}
-\]
-%
-The interpretation of linear function types at a string $w$ is a linear function
-that takes in parses of $A$ on a string $w_{a}$ and outputs parses of $B$ on the
-string $w_{a} w$.
-%
-\[
-  \sem{A \lto B} \gamma w = \Pi(w_a:\String) \left( A \gamma w_a \Rightarrow B\gamma (w_a w) \right)
-\]
-%
-That is, strings match $A \lto B$ if when prepended with a parse of $A$ they
-complete to parses of $B$. In this manner, the linear function types generalize
-Brzozowksi's notion of derivative
-\cite{brzozowskiDerivativesRegularExpressions1964}.
-Brzozowski initially only gave an accounting of this operation for
-generalized regular expressions, but later work from Might et al.\ demonstrates that the same
-construction can be generalized to context free grammars
-\cite{mightParsingDerivativesFunctional2011}. Here, via the linear function
-types, the same notion of derivative extends to the grammars of \theoryabbv.
-
-Note, to ensure that the linear function types do indeed generalize Brzozowski
-derivatives, we must include the equalities in \cref{fig:brzozowskideriv} as axioms\max{we must? why must we}.
-\pedro{We should further explain why these equations are required, or at the very
-least give some intuition}
-
-Of course, all the above statements for the left function type also have
-corresponding analogues for the
-right-handed counterpart.
-
-\begin{figure}
-\begin{align*}
-  c\lto c &\equiv I\\
-  c\lto d &\equiv \bot\\
-  c\lto I &\equiv \bot\\
-  c\lto 0 &\equiv \bot\\
-  c\lto (A \otimes B) & \equiv (c\lto A) \otimes B + (A \amp I) \otimes (c\lto B)\\
-  c\lto A^* &\equiv (A \amp I)^{*} \otimes (c \lto A) \otimes A^* \\
-  c\lto (A + B) &\equiv (c\lto A) + (c\lto B)
-\end{align*}
-\caption{Equality for Brzozowski Derivatives}
-\label{fig:brzozowskideriv}
-\end{figure}
-
-\paragraph{LNL Dependent Types}
-\steven{talk about how binary sum and binary with are definable over $\Bool$}
-Given a non-linear type $X$, we may form both the dependent product type
-$\LinPiTy x X A$ and the dependent pair type $\LinSigTy x X A$ as linear types
-where $x$ is free in $A$.
-
-$\sem{\LinPiTy x X A} \gamma w = \Pi(x:\sem{X}\gamma) \sem{A}(\gamma,x) w$
-
-$\sem{\LinSigTy x X A} \gamma w = \Sigma(x:\sem{X}\gamma) \sem{A}(\gamma,x) w$
-
-The grammar semantics of the linear product type is
-indeed a dependent function out of the
-semantics of $X$. Likewise, the grammar semantics of the linear dependent pair type is a dependent
-pair in $\Set$.\pedro{The connection to grammar semantics needs to be better explained. For instance,
-we should explain that disjunction corresponds to ``or'' of grammars}
-
-Note that even though we do not take the binary additive conjunction and
-disjunction as primitive, we may define them via these LNL dependent types. In
-particular, via a dependence on $\Bool$.
-%
-\[
-  A \amp B := \LinPiTy b \Bool {C(b)}
-\]
-\[
-  A \oplus B := \LinSigTy b \Bool {C(b)},
-\]
-where $C(\true) = A, C(\false) = B$.
-
-\paragraph{Universal Type}\max{this is absolutely not the right terminology. Universal type means multiple other completely different things}
-\steven{Definable as nullary product}
-The universal type $\top$ may be formed in any context.
-%
-\[ \sem{\top} \gamma w = \{ \ast \}\]
-%
-Its grammar semantics in set outputs the unit type in $\Set$ for all input strings in all contexts.
-
-\paragraph{Empty Type}
-\steven{Definable as nullary sum}
-The empty type $\bot$ has no inhabitants. The elimination for the empty type
-witnesses the principle of explosion --- i.e.\ from a term of type $\bot$ we may
-introduce a term $\mathsf{absurd}_{A} : A$ for any type $A$.
-%
-\[ \sem {\bot} \gamma w = \emptyset \]
-%
-\paragraph{The $G$ Modality}
-\steven{Rename this}
-\steven{Compare to persistance modalities from separation logic}
-The $G$ modality provides the embedding of linear types in the empty context
-into non-linear types. The introduction and elimination forms for $G A$ obey the
-same rules as given in \cite{bentonMixedLinearNonlinear1995} and \cite{krishnaswami_integrating_2015}.
-
-The left adjoint $F$ from non-linear types back to linear types may be defined
-as,
-%
-\[
-  F X := \LinSigTy a A I
-\]
-%
-\steven{Elaborate on $G$ more and rename $G A$to $\ltonl A$ in this section}
-
-\pedro{Agreed :) maybe mention connections to persistent propositions from the
-separation logic literature}
-
-The semantics of $G A$ are exactly the semantics of $A$ at the empty word $\varepsilon$.
-%
-\[ \sem{G A} \gamma = \sem{A} \gamma \varepsilon \]
-%
-%
-\paragraph{Recursive Types}
-\steven{reflavor this section as ``indexed inductive'' instead of ``recursive'',
-and entirely rework with the new inductive definition}
-\steven{The non-indexed inductive are definable as being trivially indexed over unit}
-Recursive linear types may be defined via the least-fixed point operator $\mu$.
-% The grammar semantics of which are the fixed-point of sets induced by the
-% grammar semantics of $A$.
-% %
-% \[ \sem{\mu x. A} \gamma = \mu (x:\Gr_i). \sem{A}(\gamma,x) \]
-% %
-% \pedro{We should justify, even if briefly, the existence of this
-%   fixed-point}
-
-% Observe that we need not take the Kleene star as a primitive
-% grammar constructor, as it is definable as a fixed point.
-% The Kleene star of a grammar $g$ is given as,
-
-% \[
-%   g^{*} := \mu X . I \oplus (g \otimes X)
-% \]
-
-% \begin{figure}[h!]
-% \begin{mathpar}
-%   \inferrule
-%   {\Gamma ; \Delta \vdash p : I}
-%   {\Gamma ; \Delta \vdash \mathsf{nil} : g^{*}}
-
-%   \and
-
-%   \inferrule
-%   {\Gamma ; \Delta \vdash p : g \\ \Gamma ; \Delta' \vdash q
-%   : g^{*}}
-%   {\Gamma ; \Delta \vdash \mathsf{cons}(p , q) : g^{*}}
-
-%   \\
-
-%   \inferrule
-%   {
-%     \Gamma ; \Delta \vdash p : g^{*} \\
-%     \Gamma ; \cdot \vdash p_{\varepsilon} : h \\
-%     \Gamma ; x : g , y : h \vdash p_{\ast} : h
-%   }
-%   {\Gamma ; \Delta \vdash \mathsf{foldr}(p_{\varepsilon} , p_{\ast}) : g^{*}}
-% \end{mathpar}
-% \caption{Kleene Star Rules}
-% \label{fig:star}
-% \end{figure}
-
-% Likewise, $g^{*}$ has admissible introduction and
-% elimination rules, shown in \cref{fig:star}. Note that this
-% definition of $g^{*}$ and these
-% rules arbitrarily assigns a handedness to the Kleene star.
-% We could have just as well took it to be a fixed point of
-% $I \oplus (X \otimes g)$. In fact, the definitions are
-% equivalent, as the existence of the $\mathsf{foldl}$ term below
-% shows that $g^{*}$ is also a fixed point of
-% $I \oplus (X \otimes g)$.
-% This is a marked point of difference from Kleene
-% algebra with recursion, where the fixed points for the left and right variants
-% of Kleene star need not agree \cite{leiss}.
-
-% \begin{equation}
-%   \label{eq:foldl}
-%   \inferrule
-%   {
-%     \Gamma ; \Delta \vdash p : g^{*} \\
-%     \Gamma ; \cdot \vdash p_{\varepsilon} : h \\
-%     \Gamma ; y : h , x : h \vdash p_{\ast} : h
-%   }
-%   {\Gamma ; \Delta \vdash \mathsf{foldl}(p_{\varepsilon} , p_{\ast}) : g^{*}}
-% \end{equation}
-
-% In fact, the $\mathsf{foldl}$ term is defined using
-% $\mathsf{foldr}$ --- much in the same way one
-% may define a left fold over lists in terms of a right fold
-% in a functional programming language.
-% The underlying trick is to fold over a list of linear functions
-% instead of the original string. We curry each character $c$
-% of the string into a function that concatenates $c$, and
-% right fold over this list of linear functions. Because function
-% application is left-associative, this results in a left
-% fold over the original string.
-
-% We only take fixed points of a single variable as a
-% primitive operation in the type theory, but we may apply
-% Beki\`c's theorem \cite{Bekić1984} to define an admissible
-% notion of multivariate fixed point. This is particularly
-% useful for defining grammars that encode the states of an
-% automaton, as we will see in \cref{sec:automata}. In \cref{fig:multifix} we provide the
-% introduction and elimination principles for such a fixed
-% point, where $\sigma$ is the substitution that unrolls the
-% mutually recursive definitions one level. That is,
-
-% \begin{align*}
-%   \sigma = \{ & \mu(X_{1} = A_{1} \dots, X_{n} = A_{n}).X_{1} / X_{1} , \dots, \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{n} / X_{n} \}
-% \end{align*}
-
-% \begin{figure}
-% \begin{mathpar}
-%   \inferrule
-%   {\Gamma ; \Delta \vdash e : \simulsubst {A_{k}} {\sigma}}
-%   {\Gamma ; \Delta \vdash \mathsf{cons}~e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{k}}
-
-%   \and
-
-%   \inferrule
-%   {\Gamma ; \Delta \vdash e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{k} \\
-%              \Gamma ; x_{j} : \simulsubst {A_{j}}{\sigma} \vdash e_{j} : B_{j}\quad \forall j
-%   }
-%   {\Gamma; \Delta \vdash \mathsf{mfold}(x_{1}.e_{1}, \dots, x_{n}.e_{n})(e) : B_{k}}
-% \end{mathpar}
-% \caption{Multi-fixed Points}
-% \label{fig:multifix}
-% \end{figure}
 
@@
 \label{sec:semantics-and-metatheory}
+
+\subsection{Formal Grammars as Linear Types}
+\label{sec:formaltype}
+\max{this should be in its own section on semantics}
+
+% The linear typing judgment in our syntax takes on the following schematic form
+% $\Gamma ; \Delta \vdash a : A$. First, $A$ represents a \emph{linear type} in
+% our syntax. The intended semantics of these linear types are formal grammars.
+% That is, the linear typing system is designed to syntactically reflect the
+% behavior of formal grammars. For this reason, we may often interchangeably use
+% the terms ``linear type'' and ``grammar''.
+
+% The term $a$ is an inhabitant of type $A$, which is thought of as a parse tree of
+% the grammar $A$. The core idea of this entire paper follows precisely from this
+% single correspondence: grammars are types, and the inhabitants of these types
+% are parse trees for the grammars.
+
+% $\Gamma$ represents a non-linear context, while $\Delta$ represents a
+% \emph{linear context} dependent on $\Gamma$. These linear contexts behave
+% substructurally. As stated earlier, they are linear --- they do not obey
+% weakening or contraction --- because a character is exhausted once read by a
+% parsing procedure. Moreover, the characters in strings appear in the order in
+% which we read them. We do not have the freedom to freely permute characters,
+% therefore any type theory that is used to reason about formal grammars ought to
+% omit the structural rule of exchange as well. This means that every variable
+% within $\Delta$ must be used \emph{exactly once} and \emph{in order of
+%   occurrence}. Thus, we can think of the linear contexts as an ordered list of
+% limited resources. Once a resource is consumed, we cannot make reference to it
+% again. Variables in a linear context then act like building
+% blocks for constructing patterns over strings.
+
+% We give the base types and type constructors for linear terms. As the
+% interpretation of types as grammars in $\Grammar$ serves as our intended
+% semantics, we simultaneously give the interpretation $\sem \cdot$ of the
+% semantics as grammars.
+
+% \pedro{The paragraphs above are a bit circular, what about the following alternative?}
+% \steven{I am in favor of cutting what's above, but we haven't introduced the
+%   syntax in the main body of the paper yet, even if the figures appeared
+%   earlier. So what you write needs to take that into account}
+% \pedro{In which case we should introduce it in the previous section. I suggest
+%   adding this right after ``and the intuitionistic typing rules in fig 3''}
+Now that we have specified the syntactic aspect of \theoryabbv, we must justify
+its connections to formal grammars in order to show that it is a sound methodology
+for reasoning about parsers. We do this by showing that $\mathbf{Gr}$ is a model
+to our type theory.
+
+Most of the non-linear semantics is standard \pedro{We should either cite something or add the non-linear semantics in the appendix}, so let us investigate in more
+depth the linear semantics.
+
+\begin{enumerate}
+  \item A non-linear context $\Gamma$ denotes a set $\sem \Gamma$
+  \item A non-linear type $\Gamma \vdash X : U_{i}$ denotes a family of sets
+        $\sem X : \sem \Gamma \to \Set_{i}$
+  \item A non-linear term $\Gamma \vdash e : X$ denotes a section
+        $\sem e : \Pi(\gamma : \sem \Gamma)\sem{X} \gamma$ \pedro{we have not defined what sections are}
+  \item Linear contexts $\Gamma \vdash \Delta$ and types
+        $\Gamma \vdash A : L_{i}$ both denote families of grammars
+        $\sem \Gamma \to \Gr_{i}$
+  \item A linear term $\Gamma ; \Delta \vdash M : A$ denotes a family of parse
+        transformers
+        $\sem M : \Pi(\gamma : \sem \Gamma)\Pi(w : \String) \sem \Delta \gamma w \Rightarrow \sem M \gamma w$
+\end{enumerate}
+
+\paragraph{Linear Unit}
+First, the linear unit $I$ may be built in the empty linear
+context\footnote{When appropriate, statements like ``the empty linear context''
+  may be taken polymorphically over all non-linear contexts that they might
+  depend on.} $\cdot \dashv I$. $I$ serves as the unit for the operation
+$\otimes$ described below.
+
+As a grammar,
+%
+\[
+  \sem {I} \gamma w = \{ () ~|~ w = \varepsilon \}
+\]
+%
+That is, $I$ maps strings to the proposition that they are the empty string. $I$
+is clearly only inhabited by the empty string $\varepsilon$, for which the
+outputted set contains a single parse tree.
+
+\paragraph{Base Types}
+For each character $c$ in the alphabet $\Sigma$, we include a base type at the
+lowest universe level $c : L_{0}$.
+
+The grammar interpretation for characters is quite similar to that of $I$.
+%
+\[
+  \sem {c} \gamma w = \{ () ~|~ w = c \}
+\]
+%
+The grammar for the character $c$ likewise maps strings $w$ to the proposition that
+they are equal to $c$.
+
+\paragraph{Tensor Product}
+The tensor product of two linear types is the first place where the ordering on
+contexts really takes effect. That is, the tensor product of two types is
+non-commutative.
+When context $\Delta$ forms a linear term of
+type $A$, and $\Delta'$ forms a linear term of type $B$, then the context
+extension $\Delta, \Delta'$ forms a linear term of type $A \otimes B$.
+
+As stated above, the type $I$ serves as the unit for $\otimes$. That is, for all
+linear types $A$, the following equalities hold:
+%
+\[ I \otimes A \equiv A \otimes I \equiv A \]
+%
+In the grammar semantics,
+%
+\[
+  \sem {A \otimes B} = \Sigma_{w_{1}, w_{2} : \String} (w_{1}w_{2} = w) \land \left(  \sem {A} \gamma w_{1} \times \sem {B} \gamma w_{2} \right)
+\]
+%
+The string $w$ matches $A \otimes B$
+precisely when it may be split into two
+pieces such that the left one matches $A$
+and the right one matches $B$.
+
+\paragraph{Linear Function Types}
+The monoidal structure provided by $\otimes$ is both left and right closed and
+this is denoted by the left and right linear function types. Further,
+for linear types $A$ and $B$ the left linear function type $A \lto B$
+enjoys an elimination principle similar to function application or modus ponens.
+%
+\[
+\inferrule{\Delta \vdash A \otimes A \lto B}{\Delta \vdash B}
+\]
+%
+The interpretation of linear function types at a string $w$ is a linear function
+that takes in parses of $A$ on a string $w_{a}$ and outputs parses of $B$ on the
+string $w_{a} w$.
+%
+\[
+  \sem{A \lto B} \gamma w = \Pi(w_a:\String) \left( A \gamma w_a \Rightarrow B\gamma (w_a w) \right)
+\]
+%
+That is, strings match $A \lto B$ if when prepended with a parse of $A$ they
+complete to parses of $B$. In this manner, the linear function types generalize
+Brzozowksi's notion of derivative
+\cite{brzozowskiDerivativesRegularExpressions1964}.
+Brzozowski initially only gave an accounting of this operation for
+generalized regular expressions, but later work from Might et al.\ demonstrates that the same
+construction can be generalized to context free grammars
+\cite{mightParsingDerivativesFunctional2011}. Here, via the linear function
+types, the same notion of derivative extends to the grammars of \theoryabbv.
+
+Note, to ensure that the linear function types do indeed generalize Brzozowski
+derivatives, we must include the equalities in \cref{fig:brzozowskideriv} as axioms\max{we must? why must we}.
+\pedro{We should further explain why these equations are required, or at the very
+least give some intuition}
+
+Of course, all the above statements for the left function type also have
+corresponding analogues for the
+right-handed counterpart.
+
+\begin{figure}
+\begin{align*}
+  c\lto c &\equiv I\\
+  c\lto d &\equiv \bot\\
+  c\lto I &\equiv \bot\\
+  c\lto 0 &\equiv \bot\\
+  c\lto (A \otimes B) & \equiv (c\lto A) \otimes B + (A \amp I) \otimes (c\lto B)\\
+  c\lto A^* &\equiv (A \amp I)^{*} \otimes (c \lto A) \otimes A^* \\
+  c\lto (A + B) &\equiv (c\lto A) + (c\lto B)
+\end{align*}
+\caption{Equality for Brzozowski Derivatives}
+\label{fig:brzozowskideriv}
+\end{figure}
+
+\paragraph{LNL Dependent Types}
+\steven{talk about how binary sum and binary with are definable over $\Bool$}
+Given a non-linear type $X$, we may form both the dependent product type
+$\LinPiTy x X A$ and the dependent pair type $\LinSigTy x X A$ as linear types
+where $x$ is free in $A$.
+
+$\sem{\LinPiTy x X A} \gamma w = \Pi(x:\sem{X}\gamma) \sem{A}(\gamma,x) w$
+
+$\sem{\LinSigTy x X A} \gamma w = \Sigma(x:\sem{X}\gamma) \sem{A}(\gamma,x) w$
+
+The grammar semantics of the linear product type is
+indeed a dependent function out of the
+semantics of $X$. Likewise, the grammar semantics of the linear dependent pair type is a dependent
+pair in $\Set$.\pedro{The connection to grammar semantics needs to be better explained. For instance,
+we should explain that disjunction corresponds to ``or'' of grammars}
+
+Note that even though we do not take the binary additive conjunction and
+disjunction as primitive, we may define them via these LNL dependent types. In
+particular, via a dependence on $\Bool$.
+%
+\[
+  A \amp B := \LinPiTy b \Bool {C(b)}
+\]
+\[
+  A \oplus B := \LinSigTy b \Bool {C(b)},
+\]
+where $C(\true) = A, C(\false) = B$.
+
+\paragraph{Universal Type}\max{this is absolutely not the right terminology. Universal type means multiple other completely different things}
+\steven{Definable as nullary product}
+The universal type $\top$ may be formed in any context.
+%
+\[ \sem{\top} \gamma w = \{ \ast \}\]
+%
+Its grammar semantics in set outputs the unit type in $\Set$ for all input strings in all contexts.
+
+\paragraph{Empty Type}
+\steven{Definable as nullary sum}
+The empty type $\bot$ has no inhabitants. The elimination for the empty type
+witnesses the principle of explosion --- i.e.\ from a term of type $\bot$ we may
+introduce a term $\mathsf{absurd}_{A} : A$ for any type $A$.
+%
+\[ \sem {\bot} \gamma w = \emptyset \]
+%
+\paragraph{The $G$ Modality}
+\steven{Rename this}
+\steven{Compare to persistance modalities from separation logic}
+The $G$ modality provides the embedding of linear types in the empty context
+into non-linear types. The introduction and elimination forms for $G A$ obey the
+same rules as given in \cite{bentonMixedLinearNonlinear1995} and \cite{krishnaswami_integrating_2015}.
+
+The left adjoint $F$ from non-linear types back to linear types may be defined
+as,
+%
+\[
+  F X := \LinSigTy a A I
+\]
+%
+\steven{Elaborate on $G$ more and rename $G A$to $\ltonl A$ in this section}
+
+\pedro{Agreed :) maybe mention connections to persistent propositions from the
+separation logic literature}
+
+The semantics of $G A$ are exactly the semantics of $A$ at the empty word $\varepsilon$.
+%
+\[ \sem{G A} \gamma = \sem{A} \gamma \varepsilon \]
+%
+%
+\paragraph{Recursive Types}
+\steven{reflavor this section as ``indexed inductive'' instead of ``recursive'',
+and entirely rework with the new inductive definition}
+\steven{The non-indexed inductive are definable as being trivially indexed over unit}
+Recursive linear types may be defined via the least-fixed point operator $\mu$.
+% The grammar semantics of which are the fixed-point of sets induced by the
+% grammar semantics of $A$.
+% %
+% \[ \sem{\mu x. A} \gamma = \mu (x:\Gr_i). \sem{A}(\gamma,x) \]
+% %
+% \pedro{We should justify, even if briefly, the existence of this
+%   fixed-point}
+
+% Observe that we need not take the Kleene star as a primitive
+% grammar constructor, as it is definable as a fixed point.
+% The Kleene star of a grammar $g$ is given as,
+
+% \[
+%   g^{*} := \mu X . I \oplus (g \otimes X)
+% \]
+
+% \begin{figure}[h!]
+% \begin{mathpar}
+%   \inferrule
+%   {\Gamma ; \Delta \vdash p : I}
+%   {\Gamma ; \Delta \vdash \mathsf{nil} : g^{*}}
+
+%   \and
+
+%   \inferrule
+%   {\Gamma ; \Delta \vdash p : g \\ \Gamma ; \Delta' \vdash q
+%   : g^{*}}
+%   {\Gamma ; \Delta \vdash \mathsf{cons}(p , q) : g^{*}}
+
+%   \\
+
+%   \inferrule
+%   {
+%     \Gamma ; \Delta \vdash p : g^{*} \\
+%     \Gamma ; \cdot \vdash p_{\varepsilon} : h \\
+%     \Gamma ; x : g , y : h \vdash p_{\ast} : h
+%   }
+%   {\Gamma ; \Delta \vdash \mathsf{foldr}(p_{\varepsilon} , p_{\ast}) : g^{*}}
+% \end{mathpar}
+% \caption{Kleene Star Rules}
+% \label{fig:star}
+% \end{figure}
+
+% Likewise, $g^{*}$ has admissible introduction and
+% elimination rules, shown in \cref{fig:star}. Note that this
+% definition of $g^{*}$ and these
+% rules arbitrarily assigns a handedness to the Kleene star.
+% We could have just as well took it to be a fixed point of
+% $I \oplus (X \otimes g)$. In fact, the definitions are
+% equivalent, as the existence of the $\mathsf{foldl}$ term below
+% shows that $g^{*}$ is also a fixed point of
+% $I \oplus (X \otimes g)$.
+% This is a marked point of difference from Kleene
+% algebra with recursion, where the fixed points for the left and right variants
+% of Kleene star need not agree \cite{leiss}.
+
+% \begin{equation}
+%   \label{eq:foldl}
+%   \inferrule
+%   {
+%     \Gamma ; \Delta \vdash p : g^{*} \\
+%     \Gamma ; \cdot \vdash p_{\varepsilon} : h \\
+%     \Gamma ; y : h , x : h \vdash p_{\ast} : h
+%   }
+%   {\Gamma ; \Delta \vdash \mathsf{foldl}(p_{\varepsilon} , p_{\ast}) : g^{*}}
+% \end{equation}
+
+% In fact, the $\mathsf{foldl}$ term is defined using
+% $\mathsf{foldr}$ --- much in the same way one
+% may define a left fold over lists in terms of a right fold
+% in a functional programming language.
+% The underlying trick is to fold over a list of linear functions
+% instead of the original string. We curry each character $c$
+% of the string into a function that concatenates $c$, and
+% right fold over this list of linear functions. Because function
+% application is left-associative, this results in a left
+% fold over the original string.
+
+% We only take fixed points of a single variable as a
+% primitive operation in the type theory, but we may apply
+% Beki\`c's theorem \cite{Bekić1984} to define an admissible
+% notion of multivariate fixed point. This is particularly
+% useful for defining grammars that encode the states of an
+% automaton, as we will see in \cref{sec:automata}. In \cref{fig:multifix} we provide the
+% introduction and elimination principles for such a fixed
+% point, where $\sigma$ is the substitution that unrolls the
+% mutually recursive definitions one level. That is,
+
+% \begin{align*}
+%   \sigma = \{ & \mu(X_{1} = A_{1} \dots, X_{n} = A_{n}).X_{1} / X_{1} , \dots, \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{n} / X_{n} \}
+% \end{align*}
+
+% \begin{figure}
+% \begin{mathpar}
+%   \inferrule
+%   {\Gamma ; \Delta \vdash e : \simulsubst {A_{k}} {\sigma}}
+%   {\Gamma ; \Delta \vdash \mathsf{cons}~e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{k}}
+
+%   \and
+
+%   \inferrule
+%   {\Gamma ; \Delta \vdash e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{k} \\
+%              \Gamma ; x_{j} : \simulsubst {A_{j}}{\sigma} \vdash e_{j} : B_{j}\quad \forall j
+%   }
+%   {\Gamma; \Delta \vdash \mathsf{mfold}(x_{1}.e_{1}, \dots, x_{n}.e_{n})(e) : B_{k}}
+% \end{mathpar}
+% \caption{Multi-fixed Points}
+% \label{fig:multifix}
+% \end{figure}
+
 %% \section{Categorical Semantics of \theoryabbv}

### 2024-10-30 paper/paper.tex
wrapping up examples for now

@@
 \paragraph{Finite Grammars}
-\newcommand{\fingram}{$(a \otimes b) \oplus c$}
-
 First consider finite grammars --- those built from base types via disjunctions $\oplus$ and
-concatenations $\otimes$. For instance, \fingram.
+concatenations $\otimes$. For instance, $(a \otimes b) \oplus c$.
 In \cref{fig:fingram} we find a proof tree that constructs a term of type
-\fingram in the context $x : a, y : b$. We think of this term as
+$(a \otimes b) \oplus c$ in the context $x : a, y : b$. We think of this term as
 a parse tree internal to \theoryabbv.
@@
 $\inl(x \otimes y)$ from \cref{fig:fingram} details the
-manner by which \stringquote{ac} matches the grammar \fingram.
+manner by which \stringquote{ac} matches the grammar $(a \otimes b) \oplus c$.
 
@@
   }
-  {x : a , y : b \vdash \inl(x \otimes y) : \fingram}
+  {x : a , y : b \vdash \inl(x \otimes y) : (a \otimes b) \oplus c}
 \end{mathpar}
@@
 \begin{align*}
-\data &~A^{*}~\where\\
+\data &~A^{*} : L~\where\\
       & \nil : \ltonl {A^{*}} \\
@@
 which takes in of $A$ and a parse of $A^{*}$ and linearly combines them to build
-a parse of $A^{*}$.
+a parse of $A^{*}$. The type defined by these constructors belong to the
+universe of linear types $L$.
 
@@
 star. In \cref{fig:kleenestarderivation}, we demonstrate that \stringquote{ab}
-matches $(a^{*} \otimes b) \oplus c$ through repeated application of the Kleene star constructors.
+matches $(a^{*} \otimes b) \oplus c$ through repeated application of the Kleene
+star constructors. In this proof tree we expand some syntactic details that are often
+left implicit. For instance, every linear context is parametrized by a nonlinear
+context, which is syntactically separate by a semicolon. If the nonlinear
+context is empty, we may often omit it; however, in this figure we make this
+distinction explicit. Further consider the leaves of the tree. First, the
+leaves involving the constructors are nonlinear terms, and thus they occur in
+the nonlinear empty context. Then we explicit cast from
+an nonlinear $uparrow$-wrapped type into a linear one. Such casts may often be
+made implicitly, such as in $\cref{fig:kleeneabstractproof}$.
 
@@
         \inferrule
+        {
+          \inferrule
+          {~}
           {\cdot \vdash \cons : \ltonl {(a \lto a^* \lto a^*)}}
+        }
         {\cdot ; \cdot \vdash \cons : a \lto a^* \lto a^*}
@@
       \inferrule
+      {
+        \inferrule
         {~}
+        {\cdot \vdash \nil : \ltonl {(a^*)}}
+      }
       {\cdot ; \cdot \vdash \nil : a^*}
@@
 
-\steven{TODO need to update the proof tree example derivations (\cref{fig:kleenestarderivation,fig:kleeneabstractproof}) to use the presentation of cons that is now in \cref{fig:kleenestarinductive}}
-
 We may also have derivations where the term in context is not simply a
-string of base types. For instance, by using the elimination principle for
-Kleene star, \cref{fig:kleeneabstractproof} builds a term of type
-$A^{*}$ in a context comprising abstract linear types. In these contexts, it is
+string of base types. For instance,
+in \cref{fig:kleeneabstractproof} we show that every parse of the grammar
+$(A \otimes A)^{*}$ induces a parse of $A^{*}$ for an arbitrary grammar $A$.
+
+The context $(A \otimes A)^{*}$ does not correspond directly to a string, so it is
 not quite appropriate
-to treat linear terms as parse \textit{trees}, because the information in
-context is not a string. Instead, it is more
+to treat linear terms here as parse \textit{trees}. Instead, it is more
 appropriate to think of such terms as parse \textit{transformers}. That is, in
@@
 \steven{TODO : describe elimination principle somewhere. Maybe just in prose?}
-\steven{I probably can't get away with the inline definition of $f$ in \cref{fig:kleeneabstractproof}}
-\max{shouldn't the fold be applied to $x$ and shouldn't $f$ be $\lambda x.\lambda y.\cdots$?}
 
@@
       \inferrule
+      {
+        \inferrule
+        {~}
         {x_1 : A, x_2 : A , y : A^* \vdash \cons (x_1 , \cons(x_2 , y)) : A^*}
+      }
       {x' : A \otimes A , y : A^* \vdash g := \letin {x_1 \otimes x_2} {x'} {\cons (x_1 , \cons(x_2 , y)) : A^*}}
@@
 \paragraph{Non-deterministic Finite Automata}
-As a final example, consider the non-deterministic finite
-automaton (NFA) $N$ from \cref{fig:exampleNFA}. Such automata play a critical
+Automata play a critical
 role in defining parsers, as they often serve as an operationalized and
 executable presentation of a formal grammar. Therefore, it is crucial that we
-can represent them internal to \theoryabbv.
+can represent them internal to \theoryabbv. As a final example, consider the
+representation of $(a^{*} \otimes b) \oplus c$ as a non-deterministic finite
+automaton (NFA), shown in \cref{fig:exampleNFA}.
 
@@
 integral to our design of \theoryabbv ; however, there are some aspects of parsing that
-cannot be captured through linear types\max{not exactly clear what you mean, maybe you mean ``simple'' linear types?} alone. Often one wants to define a
-\emph{family} of linear types varying over a set, which is a primary motivator
-for us to make \theoryabbv a linear-non-linear theory\max{jargon-y. Be more direct}. For the NFA $N$, an indexed inductive
-type of traces through $N$ is given in \cref{fig:nfainductive}.
-
-The type of traces in $N$ is indexed by two pieces of data: a state in $N$ and a
-Boolean. For each state $s : N.\mathsf{state}$ and $acc : \Bool$, we define
-$\Trace_{N}~s~acc$ which denotes the
-type of traces starting at state $s$ and terminating at an arbitrary state. The Boolean flag $acc$ marks if the trace
-ends in either an accepting or a rejecting state\max{confusing. One sentence you say arbitrary state and then the next sentence you say it has to match $acc$}. Let us dissect the structure
-of this type in more detail for state 1 of $N$.
-
-$\Trace_{N}~1~acc$ has three constructors. The first constructor
-$\mathsf{stop}_{1}$ allows a trace to end at state one, and because state
-1 is accepting $acc$ is true when terminating at state 1.
-
-The constructor
-$\mathsf{1to2}$ denotes the $b$-labeled transition in $N$ from state 1 to state
-2. The character $b$ is consumed when taking this transition and then a trace
-beginning at state 2 must be produced, which is denoted by the tensor product of
-the grammar $b$ and $Trace_{N}~2~acc$. Finally, the constructor $\mathsf{1to2}$
-abstracts over the acceptance Boolean because it is not yet determined.
-Because the Boolean is set by the acceptance criteria of the \emph{final} state,
-it is provided inductively by the trace out of state 2.
-
-The constructor $\mathsf{1to3}$ is built quite similarly as $\mathsf{1to2}$;
-however, $\mathsf{1to3}$ does not involve a tensor product. This is because the
-transition from state 1 to state 3 is labeled by the empty string $\epsilon$, so
-the transition consumes no input. In this case, when building a trace out of
-state 1, it suffices to build a trace out of state 3 \emph{over the same string}.
+cannot be captured through simple linear types alone. Often one wants to define a
+\emph{family} of linear types varying over a set, such as a type of traces through an
+NFA parametrized by the first state in the trace. For the example NFA, the data
+type $\Trace$ is provided in \cref{fig:nfainductive}.
+
+$\Trace~s$ denotes traces through the NFA beginning in state $s$ and ending in
+an accepting state. In this instance there is only a sinlge accepting state, and
+thus the traces necessarily end in state 3. Let us dissect the structure
+of $\Trace~s$ in more detail.
+
+There are two kinds of constructors for a trace: those those modelling the
+termination of a trace at a given state and those modelling transitions in the
+NFA.\ The constructor $\mathsf{stop}$ denotes the first kind of constructor and
+terminates the trace in state 3. The constructor
+$\mathsf{2to3}$ denotes the $b$-labeled transition in the NFA from state 2 to state
+3. In this case, we build a trace beginning at state 2
+by consuming a parse of the label, $b$, and consuming a trace beginning at state 3.
+Finally, the constructor $\mathsf{1to2}$ is built quite similarly as $\mathsf{2to3}$;
+however, $\mathsf{1to2}$  may transition states without reading a character. This is because the
+transition from state 1 to state 2 is labeled by the empty string $\epsilon$, so
+transition consumes no input. In this case, when building a trace out of
+state 1, it suffices to build a trace out of state 2 \emph{over the same string}.
 
 Overall, leveraging the \emph{indexed} inductive types provides a concise interface
-for defining the traces of $N$ as a family
+for defining the traces of the NFA as a family
 of mutually recursive types varying coherently over indexing data drawn from
-$N.\mathsf{state} \times \Bool$.
+the states of the NFA.
+
+Just as in \cref{fig:kleenestarderivation} shows the \stringquote{ab} matches
+$(a^{*} \otimes b) \oplus c$, we show in TODO CREF that \stingquote{ab} is
+accepted by the NFA representation of this regular expression. This is denote by
+showing that \stringquote{ab} matches the trace grammar where the index is the
+initial state of the NFA.\ Further, as we
+will see in LATER SECTION, we can derive an isomorphism between parses of the
+regular expression and accepting traces of the corresponding NFA.
 
@@
 
-\max{this notation is inconsistent with Agda syntax. You write $acc : \Bool$ but then in the $\mathsf{stop}$ cases you instantiate $acc$ with a specific boolean}
-\max{is the boolean just overcomplicating stuff here?}
-
 \begin{figure}
 \begin{align*}
-\data &~\Trace : (s : N.\mathsf{state}) \to L~\where\\
+\data &~\Trace : (s : \mathsf{Fin}~3) \to L~\where\\
       & \mathsf{stop} : \ltonl {(\Trace~3)}  \\
@@
 \label{fig:nfainductive}
+\end{figure}
+
+\begin{figure}
+\begin{mathpar}
+\inferrule
+{
+  \inferrule
+  {
+    \inferrule
+    {~}
+    {x : a \vdash x : a}
+    \\
+    \inferrule
+    {
+      \inferrule
+      {~}
+      {y : b \vdash y : b}
+      \\
+      \inferrule
+      {~}
+      {\cdot \vdash \mathsf{stop} : \Trace~3}
+    }
+    {y : b \vdash \mathsf{2to3}(y , \mathsf{stop}) : \Trace~2}
+  }
+  {x : a , y : b \vdash \mathsf{2to2}(x , \mathsf{2to3}(y , \mathsf{stop})) : \Trace~2}
+}
+{x : a , y : b \vdash \mathsf{1to2}(\mathsf{2to2}(x , \mathsf{2to3}(y ,
+  \mathsf{stop}))) : \Trace~1}
+\end{mathpar}
+\caption{\stringquote{ab} is accepted by the NFA for $(a^{*} \otimes b) \oplus c$}
+\label{fig:kleenestarderivation}
 \end{figure}

### 2024-10-31 paper/paper.tex
determinization

@@
 parser, and for this task much of the structured foundations from the theory of
-formal languages is rendered moot. Rather than returning a propositon-valued result, a parser produces a
+formal languages is rendered moot. Rather than returning a proposition-valued result, a parser produces a
 structured \emph{parse tree} adherent to a \emph{formal grammar}.
@@
 
-To gain intuiton for working in \theoryabbv, we study some examples of
+To gain intuition for working in \theoryabbv, we study some examples of
 grammars and parses of them internal to \theoryabbv. In each of our examples, we will look at
@@
 to finite grammars, regular
-expresssions are built inductively using base types, $\oplus$, and $\otimes$ ;
+expressions are built inductively using base types, $\oplus$, and $\otimes$ ;
 however, unlike finite grammars, regular expressions additionally have a
@@
 
-% TODO this should be beter formatted but can't use listings with \otimes ^{\ast}
+% TODO this should be better formatted but can't use listings with \otimes ^{\ast}
 \begin{figure}
@@
 ``$\nil$'' and ``$\cons$''
-as symbols for contructors. That is, the $\nil$ constructor is not linearly
+as symbols for constructors. That is, the $\nil$ constructor is not linearly
 consumed at usage time
@@
 star constructors. In this proof tree we expand some syntactic details that are often
-left implicit. For instance, every linear context is parametrized by a nonlinear
+left implicit. For instance, every linear context is parameterized by a nonlinear
 context, which is syntactically separate by a semicolon. If the nonlinear
@@
 \emph{family} of linear types varying over a set, such as a type of traces through an
-NFA parametrized by the first state in the trace. For the example NFA, the data
+NFA parameterized by the first state in the trace. For the example NFA, the data
 type $\Trace$ is provided in \cref{fig:nfainductive}.
@@
 $\Trace~s$ denotes traces through the NFA beginning in state $s$ and ending in
-an accepting state. In this instance there is only a sinlge accepting state, and
+an accepting state. In this instance there is only a single accepting state, and
 thus the traces necessarily end in state 3. Let us dissect the structure
@@
 
-There are two kinds of constructors for a trace: those those modelling the
-termination of a trace at a given state and those modelling transitions in the
+There are two kinds of constructors for a trace: those those modeling the
+termination of a trace at a given state and those modeling transitions in the
 NFA.\ The constructor $\mathsf{stop}$ denotes the first kind of constructor and
@@
 
-Just as in \cref{fig:kleenestarderivation} shows the \stringquote{ab} matches
-$(a^{*} \otimes b) \oplus c$, we show in TODO CREF that \stingquote{ab} is
-accepted by the NFA representation of this regular expression. This is denote by
-showing that \stringquote{ab} matches the trace grammar where the index is the
+Just as \cref{fig:kleenestarderivation} shows the \stringquote{ab} matches
+$(a^{*} \otimes b) \oplus c$, we show in \cref{fig:nfaderivation} that \stingquote{ab} is
+accepted by the NFA representation of this regular expression. This is denoted by
+showing that \stringquote{ab} matches the trace grammar when the index is the
 initial state of the NFA.\ Further, as we
 will see in LATER SECTION, we can derive an isomorphism between parses of the
-regular expression and accepting traces of the corresponding NFA.
+regular expression and accepting traces of their corresponding NFA.
 
@@
 \caption{\stringquote{ab} is accepted by the NFA for $(a^{*} \otimes b) \oplus c$}
-\label{fig:kleenestarderivation}
+\label{fig:nfaderivation}
 \end{figure}
@@
 
-\paragraph{Regular Expressions} We may realize regular expression as a subtheory
+\paragraph{Regular Expressions}
+We may realize regular expression as a subtheory
 of our language by restricting the type constructors to the
@@
 \cite{AhoIndexed} internal to \theoryabbv. Although, we can define strictly more
-things with this non-linear index, as we have unstricted access to the index
+things with this non-linear index, as we have unrestricted access to the index
 whereas Aho's grammars can only manipulate it in a particular way.
@@
 \steven{Elaborate on these, and note that they are bw context free and context
-  sensitive on the chomsky hierarchy}
+  sensitive on the Chomsky hierarchy}
 
@@
 
-If $f$ and $g$ form an isomorphsim, then $A$ and $B$ are \emph{strongly
+If $f$ and $g$ form an isomorphism, then $A$ and $B$ are \emph{strongly
   equivalent}.
@@
 %% 2. Regex, NFA, DFA
-\steven{How much is worth explicitly walking through? The examples from earlier
-  give motivation, and we are pressed for space}
 
 \paragraph{Regex-NFA/Thompson}
-\steven{TODO put regex-NFA iso here. Call it Thompson's constructon?. Mention
+
+\steven{TODO put regex-NFA iso here. Call it Thompson's construction?. Mention
   its mechanized}
@@
 \paragraph{Determinization of an NFA}
-\steven{TODO put determinzation here. Mention its mechanized}
+\newcommand{\Agda}{Formalized in Agda}
+\begin{theorem}[Determinization, \Agda]
+  Given an NFA $N$, there exists a DFA $D$ such that
+  $\Trace_{N}~N.\mathsf{init}$ is weakly equivalent to $\Trace_{D}~D.\mathsf{init}$.
+\end{theorem}
+
+\begin{proof}
+  We present a variant of the classic powerset construction CITE. First, we build
+  the DFA $D$. Define the
+  states of $D$ to be the $\mathbb{P}_{\epsilon}(N.\mathsf{states})$ --- the type of $\epsilon$-closed subsets of
+  $N.\mathsf{states}$. A subset is accepting in $D$ if it contains an accepting state
+  from $N$. Construct the initial state of $D$ as the
+  $\epsilon$-closure of $N.\mathsf{init}$. Lastly define the transtion function
+  of $D$ to send the subset $X$ under the character $c$ to the
+  $\epsilon$-closure of all the states reachable from $X$ via a transition
+  labelled with the character $c$.
+
+  We demonstrate the weak equivalence between $\Trace_{N}~(N.\mathsf{init})$ and
+  $\Trace_{D}~(D.\mathsf{init})$ by constructing morphisms between the two
+  grammars.
+
+  To build a term
+  $\Trace_{N}~(N.\mathsf{init}) \vdash \Trace_{D}~(D.\mathsf{init})$, it
+  suffices to construct the stronger term
+
+  \[
+    f : \ltonl {\left(\Trace_{N}~(s) \lto \bigwith\limits_{X : D.\mathsf{states}} \bigwith\limits_{s \in X} \Trace_{D}~X\right)}
+  \]
+  where $s$ is arbitary and then instantiate it at $s = N.\mathsf{init}$, $X = D.\mathsf{init}$.
+
+Because
+the indexing state $s$ is now arbitrary, we can define the desired term via
+induction the trace through the NFA.
+
+If the NFA trace is terminating at
+accepting state
+$q$ and $q \in X$,
+then we return a trace in $D$ that terminates at
+$X$, which is accepting because it contains the accepting state $q$.
+
+If the NFA trace is a labeled transition $s \overset{c}{\to} s'$ such that
+$s \in X$, then we first decompose the trace to a parse of the character $c$ and a
+trace starting from $s'$ --- that is, to a term of type
+$c \otimes \Trace_{N}(s')$. We then inductively call $f$ on traces out of $s'$
+to get traces in a parse of $c$ concatenated with a family of traces
+parametrized by $\epslion$-closed subsets containing $s'$.
+. We instantiate this family at $\delta_{c}(X)$, as
+$s'$ is reachable from $s$ by a single transition labelled by $c$, so
+$s' \in \overline{\epsilon}\overline{c}{X} = D.\delta_{c}(X)$. Therefore we can
+linearly combine the parse of character $c$ and the trace out of $D.\delta_{c}(X)$
+to get a trace in $D$ rooted at $X$.
+
+\[\begin{tikzcd}
+    {c \otimes \mathsf{Trace}_{N}(s')} &
+    {c \otimes \bigwith\limits_{Y : D.\mathsf{states}} \bigwith\limits_{s' \in Y} \Trace_{D}~Y} &
+    {c \otimes \mathsf{Trace}_{D}(D.\delta_{c}(X))} & {\mathsf{Trace}_{D}(X)}
+	\arrow["{id\otimes f}", from=1-1, to=1-2]
+	\arrow["{id\otimes\pi}", from=1-2, to=1-3]
+	\arrow["{D.\mathsf{cons}}", from=1-3, to=1-4]
+\end{tikzcd}\]
+
+Lastly, if the NFA trace was an $\epsilon$-transition
+$s \overset{\epsilon}{\to} s'$ with $s \in X$, then the construction is quite similar. In this
+case, we again inductively call $f$ on traces out of $s'$ except we instantiate
+the choice of subset containing $s'$ with $X$ itself. We may do this as all
+states in $D$ are taken to be $\epsilon$-closed, therefore $s' \in X$ because
+$s$ is in $X$ and $\epsilon$-transitions to $s'$.
+
+\[\begin{tikzcd}
+    {\mathsf{Trace}_{N}(s')} &
+    {\bigwith\limits_{Y : D.\mathsf{states}} \bigwith\limits_{s' \in Y} \Trace_{D}~Y} &
+    {\mathsf{Trace}_{D}(X)}
+	\arrow["{id\otimes f}", from=1-1, to=1-2]
+	\arrow["{id\otimes\pi}", from=1-2, to=1-3]
+\end{tikzcd}\]
+
+This concludes the construction of $f$.
+
+To construct a term from DFA traces to NFA traces, we similarly strengthen our
+induction hypothesis and build a term
+\[
+  g : \ltonl {\left( \Trace_{D}(X) \lto \bigoplus\limits_{s : N.\mathsf{states}}\bigoplus\limits_{s \in X} \Trace_{N}(s) \right)}
+\]
+
+Let's turn our heads for a moment to the content behind the traces in $D$.
+Traces from subset $X$ to subset $Y$ denote the \emph{existence} of some trace
+in the $N$ from some $x \in X$ to some $y \in Y$. In general it is not clear how
+to extract out a choice of trace in $N$, which is what $g$ is asking of us.
+However, we may further take the states and transitions of $N$ to be ordered.
+This allows us to
+induce an order on traces in $D$ from $X$ to $Y$. From these orderings we can then
+define the necessary choice functions on each type by taking the smallest with
+respect to the ordering.
+
+We define $g$ by induction on traces through $D$. First if the trace stops at
+some accepting $\epsilon$-closed subset $X$, then we may choose some state
+$s \in X$ such that $s$ is accepting and terminate the trace in $N$ at $s$.
+
+Next if the trace in $D$ is defined via transition
+$X \overset{c}{\to} D.\delta_{c}(X)$ then we may choose some state in
+$s \in D.\delta_{c}(X)$. Inductively, $g$ sends the trace rooted at
+$D.\delta_{c}(X)$ to one rooted at $s$. By definition of $D.\delta_{c}(X)$,
+there exists some $s' \in X$ such that $s'$ tranitions to $s$ by some number of
+$\epsilon$-transitions followed by a transition labeled by $c$.
+We order all
+such traces of this form and choose the smallest one, which is then linearly
+combined with the parse of $c$ to build an $N$ trace. While using
+$N.\mathsf{cons}$ to build this final trace, we appropriately shift the indexing
+data to $s'$, which is adherent to the defintion of $D.\delta_{c}(X)$. This concludes the
+definition of $g$.
+
+\[\begin{tikzcd}
+    {c \otimes \mathsf{Trace}_{D}(D.\delta_{c}(X))} &
+    {c \otimes \bigoplus\limits_{s : N.\mathsf{states}} \bigoplus\limits_{s \in D.\delta_{c}(X)} \Trace_{N}~s} \\
+    {\bigoplus\limits_{s' : N.\mathsf{states}} \bigoplus\limits_{s' \in X} \Trace_{N}~s'} &
+    {\bigoplus\limits_{s : N.\mathsf{states}} \bigoplus\limits_{s \in D.\delta_{c}(X)} \left( c \otimes \Trace_{N}~s \right)}
+	\arrow["{id\otimes g}", from=1-1, to=1-2]
+	\arrow["{\mathsf{dist}}", from=1-2, to=2-2]
+	\arrow["{}", from=2-2, to=2-1]
+\end{tikzcd}\]
+
+Therefore $\Trace_{N}(N.\mathsf{init})$ and $\Trace_{D}(D.\mathsf{init})$ are
+weakly equivalent, and thus $N$ and $D$ recognize the same language of strings.
+\end{proof}
 
@@
 
-\steven{somewhere in here, sneak in the benchmark where the dyck parser could
+\steven{somewhere in here, sneak in the benchmark where the Dyck parser could
   handle 10's of thousands of characters in about 30s}
@@
 seem overfit to classical reasoning as they don't preserve the grammar up to
-strong equivalence, only weak equivalnce in general. \steven{cite lambek
+strong equivalence, only weak equivalence in general. \steven{cite Lambek
   automata and state that is is a secondary goal to preserve as much structure as possible}
@@
 \subsection{Semantic Actions}
-\steven{TODO edit this. synthesize it with the later presentation of sematnic
+\steven{TODO edit this. synthesize it with the later presentation of semantic
   actions as models, and then remove the model section}
@@
 
-Given a linear type $A$, the inhaitants of $A$ serve as concrete syntax. We may
+Given a linear type $A$, the inhabitants of $A$ serve as concrete syntax. We may
 then chose a non-linear type $X$ that serves as abstract syntax. A semantic
@@
 \steven{Rename this}
-\steven{Compare to persistance modalities from separation logic}
+\steven{Compare to persistence modalities from separation logic}
 The $G$ modality provides the embedding of linear types in the empty context
@@
 
-\subsection{Lambek Hyperdoctines and Chomsky Hyperdoctrines}
+\subsection{Lambek Hyperdoctrines and Chomsky Hyperdoctrines}
 

### 2024-10-31 paper/paper.tex
turing machine wip

@@
 
-\subsection{UnReStRiCtEd}
+\subsection{Unrestricted Grammars}
 %% 5. Expressiveness: Turing Machines
+The expressive power of \theoryabbv spans the entire Chomsky hierachy, as we may
+encode the grammars accepted by Turing machines.
+
+We nonlinearly encode the tape of the Turing machine as
+function from $\mathbb{Z} \to \Gamma$, where $\Gamma$ is your tape alphabet. The
+grammar of traces for a Turing machine linear consumes the input string and
+writes it to the tape. From here, we nonlinear reason about machine transitions.
+We execute the machine and track its current state as part of the nonlinear
+indexing data.
 

### 2024-10-31 paper/paper.tex
semantics progress

--- 2024-10-31 paper/paper.tex
@@
 \newcommand{\String}{\textbf{String}}
+\newcommand{\stringg}{\mathsf{string}}
+\newcommand{\charg}{\mathsf{char}}
 \newcommand{\Set}{\mathbf{Set}}
@@
 
+\newcommand{\isTy}{\textrm{ type}}
+\newcommand{\isCtx}{\textrm{ ctx}}
+\newcommand{\isSmall}{\textrm{ small}}
+\newcommand{\isLinTy}{\textrm{ lin. type}}
+\newcommand{\isLinCtx}{\textrm{ lin. ctx.}}
+
 \newcommand{\gluedNL}{{\mathcal G}_S}
@@
 \newcommand{\cons}{\mathsf{cons}}
+\newcommand{\epscons}{\epsilon\mathsf{cons}}
 \newcommand{\data}{\mathsf{data}}
@@
 \newcommand{\pedro}[1]{\ifdraft{\color{red}[{\bf Pedro says}: #1]}\fi}
+\newcommand{\nathan}[1]{\ifdraft{\color{green}[{\bf Nathan says}: #1]}\fi}
 \newcommand{\pipe}{\,|\,}
@@
 \email{stschaef@umich.edu}
+
+\author{Nathan Varner}
+\affiliation{\department{Electrical Engineering and Computer Science}
+  \institution{University of Michigan}
+  \country{USA}
+}
+\email{nvarner@umich.edu}
 
@@
 
-% TODO this should be better formatted but can't use listings with \otimes ^{\ast}
 \begin{figure}
@@
 In short, a nonlinear term
-$M : \ntonl A^{*}$ denotes a linear term in the empty context
+$M : \ltonl A^{*}$ denotes a linear term in the empty context
 $\cdot \vdash M : A^{*}$. Later in
@@
 left implicit. For instance, every linear context is parameterized by a nonlinear
-context, which is syntactically separate by a semicolon. If the nonlinear
+context, which is syntactically separated by a semicolon. If the nonlinear
 context is empty, we may often omit it; however, in this figure we make this
@@
 the nonlinear empty context. Then we explicit cast from
-an nonlinear $uparrow$-wrapped type into a linear one. Such casts may often be
+an nonlinear $\uparrow$-wrapped type into a linear one. Such casts may often be
 made implicitly, such as in $\cref{fig:kleeneabstractproof}$.
@@
 Just as \cref{fig:kleenestarderivation} shows the \stringquote{ab} matches
-$(a^{*} \otimes b) \oplus c$, we show in \cref{fig:nfaderivation} that \stingquote{ab} is
+$(a^{*} \otimes b) \oplus c$, we show in \cref{fig:nfaderivation} that \stringquote{ab} is
 accepted by the NFA representation of this regular expression. This is denoted by
@@
 
+\steven{TODO the example figures need to changed to not use $x$'s for linear variables}
+
 %% 3. Third example: NFA traces to show the indexed inductive types
@@
                                                   \mathsf{SPF}\,X\pipe U \pipe L
-                                                  \pipe \ltonl A \pipe \sum_{x:X} Y
-                                                  \pipe \prod_{x:X} Y\pipe M =_A
+                                                  \pipe \ltonl A \pipe \sum\limits_{x:X} Y
+                                                  \pipe \prod\limits_{x:X} Y\pipe M =_A
                                                   N \\
@@
 
-\subsection{Subtheories}
-Each choice of connective has direct implications on the expressivity of
-\theoryabbv. In particular, the set of connectives used will determine where the
-grammar semantics is placed in
-the Chomsky hierarchy.
-
-\paragraph{Regular Expressions}
-We may realize regular expression as a subtheory
-of our language by restricting the type constructors to the
-linear unit, characters, $\otimes$, $\oplus$, and Kleene star. Classically, the
-languages recognized by these are exactly the regular languages --- the lowest
-on the Chomsky hierarchy.
-
-\paragraph{$\mu$-Regular Expressions} Similarly, we may restrict the connectives
-to be linear unit, characters, $\otimes$, $\oplus$, and arbitrary recursion via
-left-fixed point rather than only Kleene star.
-Instead of regular expressions these correspond to Lei{\ss}'s $\mu$-regular
-expressions, which are known to be equivalent to context-free grammars
-\cite{leiss,krishnaswami_typed_2019}.
-
-\paragraph{Beyond Context-Free Grammars} The previous two subtheories induce as semantics
-regular grammars and context-free grammars, respectively; however, by including
-the LNL dependent types we may actually express the entirety of the Chomsky
-hierarchy. Through use of the LNL dependent types, we may encode indexed
-grammars, which are known to be properly between context-free and
-context-sensitive grammars within the Chomsky hierarchy \cite{AhoIndexed}. We
-will further see in \cref{subsubsec:tm} how to use dependence to encode Turing machines internal to our
-calculus, which of course induces a semantics of unrestricted grammars.
-
-\paragraph{Indexed Grammars}
-The LNL dependent types can be used to encode Aho's indexed grammars
-\cite{AhoIndexed} internal to \theoryabbv. Although, we can define strictly more
-things with this non-linear index, as we have unrestricted access to the index
-whereas Aho's grammars can only manipulate it in a particular way.
-
-\steven{Elaborate on these, and note that they are bw context free and context
-  sensitive on the Chomsky hierarchy}
+% \paragraph{Regular Expressions}
+% We may realize regular expression as a subtheory
+% of our language by restricting the type constructors to the
+% linear unit, characters, $\otimes$, $\oplus$, and Kleene star. Classically, the
+% languages recognized by these are exactly the regular languages --- the lowest
+% on the Chomsky hierarchy.
+
+% \paragraph{$\mu$-Regular Expressions} Similarly, we may restrict the connectives
+% to be linear unit, characters, $\otimes$, $\oplus$, and arbitrary recursion via
+% left-fixed point rather than only Kleene star.
+% Instead of regular expressions these correspond to Lei{\ss}'s $\mu$-regular
+% expressions, which are known to be equivalent to context-free grammars
+% \cite{leiss,krishnaswami_typed_2019}.
+
+% \paragraph{Beyond Context-Free Grammars} The previous two subtheories induce as semantics
+% regular grammars and context-free grammars, respectively; however, by including
+% the LNL dependent types we may actually express the entirety of the Chomsky
+% hierarchy. Through use of the LNL dependent types, we may encode indexed
+% grammars, which are known to be properly between context-free and
+% context-sensitive grammars within the Chomsky hierarchy \cite{AhoIndexed}. We
+% will further see in \cref{subsubsec:tm} how to use dependence to encode Turing machines internal to our
+% calculus, which of course induces a semantics of unrestricted grammars.
+
+% \paragraph{Indexed Grammars}
+% The LNL dependent types can be used to encode Aho's indexed grammars
+% \cite{AhoIndexed} internal to \theoryabbv. Although, we can define strictly more
+% things with this non-linear index, as we have unrestricted access to the index
+% whereas Aho's grammars can only manipulate it in a particular way.
+
+% \steven{Elaborate on these, and note that they are bw context free and context
+%   sensitive on the Chomsky hierarchy}
 
@@
 \label{sec:applications}
-
-\steven{this section is mostly word vomit, need to edit}
+This section explores the applications of \theoryabbv to conducting formal
+gramamr theory. We demonstrate that several classical notions and constructions
+integral to the theory of formal languages are faithfully represented
+in \theoryname. By encoding well-established formal grammar concepts, we ensure
+that our framework remains grounded in the
+foundational principles of formal language theory while opening the door to
+compositional formal verification of parsers.
 
 \subsection{Grammar Properties}
-\steven{Mention that $\top$ and $\String$ are iso. Highlight this as a needed axiom}
+
+\steven{Need an intro for this subsection}
 
@@
 \paragraph{Unambiguity}
-In the grammars model, we say that a grammar is unambiguous if there is at most
-one parse tree for any input string. There are several ways to capture this
-concept inside of the theory of \theoryabbv.
-
-In the presence of distributive coproducts (which we probably add a) the following are equivalent:
+Ordinarily, we think of a formal grammar as unambiguous when there is at most
+one parse tree for any input string; however, this description of unambiguity is
+a semantic property that is not expressible in the syntax of \theoryabbv. Rather
+than capturing this uniqueness aspect directly, we provide several notions
+internal to \theoryabbv that imply unambiguity in the grammars model.
+
+We say that $A$ is \emph{unambiguous} if the unique map $! : A \to \top$ to the
+terminal object is a monomorphism. Further, in the presence of distributive coproducts, the following are equivalent:
 \begin{itemize}
   \item $A$ is an unambiguous grammar
-  \item The morphism $! : A \to \top$ is a monomorphism
   \item Given a grammar $B$, all morphisms $B \to A$ are equal
@@
 
+\paragraph{A Grammar of Strings}
+Define $\charg$ to be the disjunction of characters in the alphabet $\Sigma$,
+$\charg = \LinSigTy {c} {\Sigma} c$. Then define a linear type of strings,
+$\stringg = \charg^{*}$. In the grammars model, we may construct an
+isomorphism between $\top$ and $\stringg$. Thus when building parsers, we may wish to enforce this
+syntactically by including $\top \cong \stringg$ as an axiom.
+
 \paragraph{Parseability}
-We say $A$ is \emph{parseable} if there exists a type $B$ and a map
-$\top \to A \oplus B$. Because $\top \cong \String$, this says that any string
-can be parsed into either an $A$ or a $B$.
-
-If $\top \cong A \oplus B$, call $A$ \emph{unambiguously parseable}.
+We say $A$ is \emph{parseable} if there exists a grammar $B$ and a map
+$\top \to A \oplus B$. Taking $\top \cong \stringg$, a grammar $A$ is parseable
+exactly when we may read an arbitrary string into a parse of $A$ or a $B$-valued
+refutation.
 
 \paragraph{Decidability}
-In the case that $A$ is parseable with respect to $\neg A$ --- i.e.
+In the case that $A$ is parseable with respect to $\neg A$ --- i.e.\
 there's a map $\top \to A \oplus \neg A$ --- then call $A$ \emph{decidable}.
@@
 \paragraph{Equivalence}
-Grammars $A$ and $B$ are \emph{weakly equivalent} if there exists maps
-$f : A \to B$ and $g : B \to A$.
-
-If $f$ and $g$ form an isomorphism, then $A$ and $B$ are \emph{strongly
-  equivalent}.
-\steven{TODO cite that these are the same as Chomsky's notions of equivalence}
-
-The induced notion of isomorphism in $\Grammar$ encodes that two grammars are equivalent if there is
-a bijective translation of parses. In other words, two grammars are isomorphic
-if they capture the same structural descriptions of strings, which is
-precisely Chomsky's notion of \emph{strong equivalence} of
-grammars \cite{chom1963}. In works like
-\cite{yoshinaga2002formal}, this notion of strong equivalence is often treated
-like the appropriate choice of ``sameness'' for two grammars. When abstracting over the latent
+In \cite{chom1963}, Chomsky differentiates between the weak generative capacity
+of a formal grammar --- the set of strings accepted by it --- and its strong
+generative capacity --- the set of ``structural descriptions'', or parse trees,
+that it generates. Two grammars are said to be \emph{weakly equivalent} by
+Chomsky if they
+agree in their weak generative capacity, and likewise \emph{strongly equivalent}
+if they agree in their strong generative capacity. We aim to recover each of
+these notions of equivalence internally.
+
+Internal to \theoryabbv, we define grammars $A$ and $B$ to be \emph{weakly equivalent} if there exists maps
+$f : A \to B$ and $g : B \to A$. Note that in the grammars model, this implies the expected equality on
+the set of accepted strings.
+
+If $f$ and $g$ form an isomorphism, then $A$ and $B$ defined to be \emph{strongly
+  equivalent}. This implies Chomsky's notion of strong equivalence, as for any
+input string $w$
+$A \cong B$ induces a bijective translation of parses $A w \cong B w$.
+When abstracting over the latent
 structure of formal grammars and describing it in the the language of
-categories, it is encouraging to see the same definition of equivalence arise
-naturally.
+categories, it is encouraging to see the same definition of strong equivalence
+emerge naturally.
 
@@
 
-\paragraph{Regex-NFA/Thompson}
-
-\steven{TODO put regex-NFA iso here. Call it Thompson's construction?. Mention
-  its mechanized}
-
-\paragraph{Determinization of an NFA}
+Regular expressions
+are among one of the most studied and
+well-understood concepts in all of computer science. Pratically, they are
+essential tools for lexical analysis and many text
+processing tasks. So they are a natural concept to encoding into our
+domain-specific grammars language.
+
+Nearly every theoretical discussion of regular expressions is accompanied by a
+discussion of finite automata, as it is well known that regular expressions and finite automata are
+equivalent as grammar formalisms. In this section, we make the equivalences
+between regular expressions and finite automata precise by encoding two theorems
+from classical language theory: (1) every regular expression is strongly equivalent
+to the traces through some nondeterministic finite automaton (NFA); and, (2) for
+every NFA, there is some deterministic finite automaton (DFA) such that the type
+of traces through the NFA is weakly equivalent to the type of traces through the DFA.
+
+\newcommand{\states}{\mathsf{states}}
+\newcommand{\labelt}{\mathsf{label}}
+\newcommand{\transitions}{\mathsf{transitions}}
+\newcommand{\epstransitions}{\epsilon\mathsf{transitions}}
+\newcommand{\isAcc}{\mathsf{isAcc}}
+\newcommand{\init}{\mathsf{init}}
+\newcommand{\src}{\mathsf{src}}
+\newcommand{\dst}{\mathsf{dst}}
+\newcommand{\epssrc}{\epsilon\mathsf{src}}
+\newcommand{\epsdst}{\epsilon\mathsf{dst}}
+
+\paragraph{Regular Expressions} The theory of regular expressions emerges as the
+subtheory of \theoryabbv generated by
+$\otimes$, $\oplus$, and Kleene star over the types $\bot$, $I$, and characters drawn from $\Sigma$.
+
+\paragraph{Nondeterministic Finite Automata} In \cref{sec:type-theory-examples}, we saw the demonstration of
+one particular NFA corresponding to the regular expression
+$(a^{*} \otimes b) \oplus c$. The definition of the linear type of traces
+through this
+NFA, as
+presented in \cref{fig:nfainductive}, inspects the concrete structure of the
+NFA.\ When generalizing from this example to an arbitrary NFA, we do not have the
+luxury of inspecting a concrete defintion. Therefore, we abstract over the
+specification of an arbitrary NFA and define traces with respect to that
+specification.
+
+Let the data of an NFA $N$ be given by the following nonlinear data,
+\begin{itemize}
+\item $N.\states$ --- a finite set of states
+\item $N.\init$ --- the initial state
+\item $N.\isAcc$ --- a decidable proposition over $N.\states$ denoting which
+        states are accepting
+\item $N.\transitions$ --- a finite set of labeled transitions
+\item $N.\labelt$ --- a mapping from $N.\transitions$ to the character labeling
+        the transition
+\item $N.\src$ --- a mapping from $N.\transitions$ to source state of a transition
+\item $N.\dst$ --- a mapping from $N.\transitions$ to destination state of a transition
+\item $N.\epstransitions$ --- a finite set of $\epsilon$-transitions
+\item $N.\epssrc$ --- a mapping from $N.\epstransitions$ to source state of a $\epsilon$-transition
+\item $N.\epsdst$ --- a mapping from $N.\epstransitions$ to destination state of a $\epsilon$-transition
+\end{itemize}
+
+For the above definition of an NFA $N$, in \cref{fig:nfatrace} we define a linear type of traces
+through $N$. Traces in $N$ may be built through one of three constructors. We
+may terminate a trace at an accepting state with the constructor $\nil$. If we
+had a trace beginning at the destination state of a transition, then we may use
+the $\cons$ constructor to linearly combine that trace with a parse of the label
+of the transition to build a trace beginning at the source of the transition.
+Finally, if we had a trace beginning at the destination of an
+$\epsilon$-transition then we may use $\epscons$ to pull it back along the
+$\epsilon$-transition and construct a trace beginning at the source of the $\epsilon$-transition.
+
+\begin{figure}
+\begin{align*}
+\data &~\Trace_{N} : (s : N.\states) \to L~\where\\
+      & \nil : \ltonl {\left( \LinPiTy {s} {N.\states} {\LinPiTy {\mathsf{acc}} {N.\isAcc(s)} {\Trace_{N}~s}} \right)}  \\
+      & \cons : \ltonl {\left( \LinPiTy {t} {N.\transitions} {(N.\labelt~t) \lto \Trace_{N}~(N.\dst(t)) \lto \Trace_{N}(N.\src(t))} \right)} \\
+      & \epscons : \ltonl {\left( \LinPiTy {t} {N.\epstransitions} {\Trace_{N}~(N.\epsdst(t)) \lto \Trace_{N}(N.\epssrc(t))} \right)}
+\end{align*}
+\caption{Traces through the NFA $N$ as an indexed inductive type}
+\label{fig:nfatrace}
+\end{figure}
+
+\paragraph{Determinstic Finite Automata}
+Just as above, we provide a specification for an arbitrary DFA.\ A DFA $D$ is
+given by the following data,
+
+\begin{itemize}
+\item $D.\states$ --- a finite set of states
+\item $D.\init$ --- the initial state
+\item $D.\isAcc$ --- a decidable proposition over $D.\states$ denoting which
+      states are accepting
+\item $D.\delta$ --- a transition function. Given a character $c$ and a state
+        $s$, $D.\delta_{c}(s)$ is the state for which there is a transition $s \overset{c}{\to} D.\delta_{c}(s)$
+\end{itemize}
+
+The linear type of traces through $D$ is given in \cref{fig:dfatrace}. These
+traces are either terminated in an accepting state with the $\nil$ constructor,
+or the $\cons$ constructor builds a trace out of state $s$ by linearly combining
+a parse of some character $c$ with a
+trace out of the state $D.\delta_{c}(s)$.
+
+\begin{figure}
+\begin{align*}
+\data &~\Trace_{D} : (s : D.\states) \to L~\where\\
+      & \nil : \ltonl {\left( \LinPiTy {s} {D.\states} {\LinPiTy {\mathsf{acc}} {D.\isAcc(s)} {\Trace_{D}~s}} \right)}  \\
+      & \cons : \ltonl {\left( \LinPiTy {c} {\Sigma} {\LinPiTy {s} {D.\states} {c \lto \Trace_{D}~(D.\delta_{c}(s)) \lto \Trace_{D}(s)}} \right)}
+\end{align*}
+\caption{Traces through the DFA $D$ as an indexed inductive type}
+\label{fig:dfatrace}
+\end{figure}
+
 \newcommand{\Agda}{Formalized in Agda}
@@
 \begin{proof}
-  We present a variant of the classic powerset construction CITE. First, we build
+  We present a variant of the Rabin and Scotts's classic powerset construction \cite{rabinFiniteAutomataTheir1959}. First, we build
   the DFA $D$. Define the
@@
 to get traces in a parse of $c$ concatenated with a family of traces
-parametrized by $\epslion$-closed subsets containing $s'$.
+parametrized by $\epsilon$-closed subsets containing $s'$.
 . We instantiate this family at $\delta_{c}(X)$, as
@@
 
-Let's turn our heads for a moment to the content behind the traces in $D$.
+For a moment, let us turn our attention to the content behind the traces in $D$.
 Traces from subset $X$ to subset $Y$ denote the \emph{existence} of some trace
@@
 
+\begin{theorem}[Thompson's Construction, \Agda]
+  Given a regular expression $R$, there exists an NFA $N$ such that $R$ is
+  strongly equivalent to $\Trace_{N}(N.\init)$.
+\end{theorem}
+
+\begin{proof}
+In this proof, we present a variant of Thompson's construction
+\cite{thompsonProgrammingTechniquesRegular1968} in which we induct on the
+structure of $R$ to construct strongly equivalent NFAs. We will sketch this
+proof by walking through some of the inductive cases, the remaining cases follow
+similar reasoning and the details may be found in the formalized artifact.
+
+\begin{figure}
+  \begin{tikzpicture}[node distance = 17mm ]
+    \node[state, initial] (0) {$0$};
+    \node[state, right of=0, accepting] (1) {$1$};
+
+    \path[->] (0) edge[above] node{$c$} (1);
+  \end{tikzpicture}
+  \caption{NFA $N_{c}$ for a single character $c$}
+  \label{fig:litnfa}
+\end{figure}
+
+First handle the case where $R$ is a literal character $c$. The NFA $N_{c}$ pictured in
+\cref{fig:litnfa} is strongly equivalent to the grammar $c$. To demonstrate this
+strong equivalence, we build terms between $c$ and $\Trace_{N_{c}}(N_{c}.\init)$
+and show they are mutually inverse.
+
+Define $f : \ltonl {(c \lto \Trace_{N_{c}}(N_{c}.\init))}$ as follows,
+
+\[
+  a : c \vdash f := \cons(a, \nil) : \Trace_{N_{c}}(N_{c}.\init)
+\]
+
+Define $g : \ltonl {(\Trace_{N_{c}}(N_{c}.\init) \lto c)}$ via induction on the
+trace. Map traces beginning in state 1, which are necessarily built via $\nil$,
+to the linear unit $I$. Then we must handle traces beginning in state 0, which
+are necessarily built via $\cons$. We decompose such a trace into a parse of the
+character $c$ and a trace beginning in state 1. We inductively transform the
+trace beginning in state 1 into a term of type $I$. Finally, we use the right
+unitor to turn a term of type $c \otimes I$ into one of type $c$.
+
+\[\begin{tikzcd}
+    {c \otimes \mathsf{Trace}_{N_{c}}(1)} &
+    {c \otimes I} &
+    c
+	\arrow["{id\otimes g}", from=1-1, to=1-2]
+	\arrow["{}", from=1-2, to=1-3]
+\end{tikzcd}\]
+
+Note that the grammar $c$ is unambiguous, so $g \circ f$ is equal to the
+identity on $c$. Additionally, we can show that $f$ carries the structure of an
+indexed algebra homomorphism for the family of functors defining the trace type
+$\Trace_{N_{c}}$. Because $g$ is defined via induction, it carries the structure
+of an algebra homomorphism as well. Thus, $f \circ g$ is also an algebra
+homomorphism. $f \circ g$ and the identity map are both then algebra
+homomorphisms from $\Trace_{N_{c}}$ to itself. Because $\Trace_{N_{c}}$ is the
+initial algebra for the relevant family of functors, there is only one
+homomorphism from it to itself and thus $f \circ g = \mathsf{id}$. Therefore,
+$f$ and $g$ are mutually inverse and the regular expression is strongly
+equivalent to the traces through the NFA in this case.
+
+Now consider the case where $R$ is the disjunction of two regular expressions
+$R_{1}$ and $R_{2}$. Inductively, we may build NFAs $N_{1}$ and $N_{2}$ that are
+strongly equivalent to $R_{1}$ and $R_{2}$, respectively. We define $N_{\oplus}$
+as a new NFA built as the disjunction of $N_{1}$ and $N_{2}$.
+$N_{\oplus}.\states$ comprises the states from $N_{1}$, the states from
+$N_{2}$, and a new marked initial state. The accepting states are precisely
+those that were accepting in the subautomata. The initial state has an
+$\epsilon$-transition to each of the initial states of the subautomata. Once in
+$N_{1}.\init$, the downstream transitions are a copy of those in $N_{1}$,
+likewise for $N_{2}.\init$.  That is, $N_{\oplus}$ contains a copy of each
+$N_{1}$ and $N_{2}$, and we choose which copy to inhabit by deciding which
+$\epsilon$-transition out of the initial state to take. A schematic
+representation of this automaton is given in \cref{fig:disjunctionnfa}.
+
+\begin{figure}
+  \begin{tikzpicture}[node distance = 17mm ]
+    \node[state, initial] (init) {$\init$};
+    \node[state, above right of=init] (init1) {$N_{1}.\init$};
+    \node[state, below right of=init] (init2) {$N_{2}.\init$};
+    \node[right of=init1] (dots1) {$\dots$};
+    \node[right of=init2] (dots2) {$\dots$};
+    \node[state, accepting, right of=dots1] (acc1) {$s_{1}$};
+    \node[state, accepting, right of=dots2] (acc2) {$s_{2}$};
+
+    \path[->] (init) edge[below] node{$\epsilon$} (init1)
+              (init) edge[above] node{$\epsilon$} (init2)
+              (init1) edge[above] node{} (dots1)
+              (init2) edge[above] node{} (dots2)
+              (dots1) edge[above] node{} (acc1)
+              (dots2) edge[above] node{} (acc2);
+  \end{tikzpicture}
+  \caption{NFA $N_{\oplus}$ as a disjunction of $N_{1}$ and $N_{2}$}
+  \label{fig:disjunctionnfa}
+\end{figure}
+
+Because $N_{1}$ and $N_{2}$ are each strongly equivalent to $R_{1}$ and $R_{2}$,
+it suffices to show that
+$\Trace_{N_{1}}(N_{1}.\init) \oplus \Trace_{N_{2}}(N_{2}.\init)$ is strongly
+equivalent to $\Trace_{N_{\oplus}}(N_{\oplus}.\init)$.
+
+Just like the case of a single character, when building the maps in either
+direction, we strengthen the induction hypothesis by mapping out of traces
+beginning with an arbitrary state then proceed by induction on traces.
+
+First, define the
+mapping out of $\Trace_{N_{\oplus}}(s)$. If $s$ is a state from $N_{1}$ then
+send the trace to a trace in $N_{1}$ beginning at $s$. Likewise, if $s$ is from
+$N_{2}$, then we map to a trace in $N_{2}$ beginning at $s$. If $s$ is
+$N_{\oplus}.\init$, then we map the trace to a disjunction of traces
+$\Trace_{N_{1}}(N_{1}.\init) \oplus \Trace_{N_{2}}(N_{2}.\init)$.
+
+In the other direction, we need to map a disjunction of traces through $N_{1}$
+or $N_{2}$ to a trace in $N_{\oplus}$. Without loss of generality, assume that
+the trace came from $N_{1}$. That is, assume we currently have a term of type
+$\Trace_{{N_{1}}}(N_{1}.\init)$ and we aim to construct a term of type
+$\Trace_{N_{\oplus}}(N_{\oplus}.\init)$. We induct on the trace $N$ trace an map
+it over to a trace beginning at the copy of $N_{1}.\init$ in $N_{\oplus}$, and
+then we conclude by using the $\epscons$ constructor to turn this trace into one
+beginning at $N_{\oplus}.\init$.
+
+We show that these maps mutually invert each other by the same sort of argument
+we used with the single character NFA.\ That is, we prove that the each map is a
+homomorphism of algebras and then appeal to the initiality of each of
+$\Trace_{N_{1}}$, $\Trace_{N_{2}}$, and $\Trace_{N_{\oplus}}$ to prove that the
+relevant maps compose to the identity.
+
+In this sketch, we elide the proof that these maps do indeed form homomorphisms.
+Further, we are eliding the needed NFAs for the cases where the regular
+expression is $I$, $\bot$, a Kleene star, or a concatenation. All of these
+elided details may be found in the Agda formalization.
+\end{proof}
+
 \subsection{The Dyck Grammar}
 %% 3. Dyck language
-The Dyck grammar $D$ for balanced parens (summarize what it is + citation)
+\newcommand{\balanced}{\mathsf{balanced}}
+\newcommand{\Dyck}{\mathsf{Dyck}}
+Moving up the Chomsky hierarchy from regular grammars, we arrive at context-free
+grammars (CFGs). In many ways, the smallest nontrivial CFG is the Dyck grammar
+$\Dyck$ --- the grammar of balanced parentheses. Moreover, the Chomsky-Sch\"utzenberger
+representation theorem \cite{CHOMSKY1959118} demonstrates that $\Dyck$, and grammars like it,
+cut to the essence of what it means to be context free.
+
+$\Dyck$ is a grammar over the alphabet $\{ \stringquote{(}, \stringquote{)} \}$
+and is defined in \cref{fig:dyckinductive}. The $\nil$ constructor shows that
+the empty string is balanced, and the $\balanced$ constrcutors builds a balanced
+parse by wrapping an already balanced parse in an additional set of parentheses
+then following it with another balanced parse.
+
+\begin{figure}
+\begin{align*}
+\data &~\Dyck : L~\where\\
+      & \nil : \ltonl {\Dyck} \\
+      & \balanced : \ltonl {(\stringquote{(} \lto \Dyck \lto \stringquote{)} \lto \Dyck \lto \Dyck)}
+\end{align*}
+\caption{Kleene Star as an inductive type}
+\label{fig:dyckinductive}
+\end{figure}
 
@@
 
-\steven{this is a VERY loose sketch of the things I want to say about $D$}
-\Cref{fig:DyckAutomaton} presents an automaton $M$ that recognizes $D$. We show this
-precisely by constructing an isomorphism between $D$ and $\Trace_{M}~0~\true$.
-Unlike the finite automata seen thus far, $M$ has infinitely many states --- one
-for each $n : \mathbb{N}$ with an additional fail state. This diverges a bit
-from classical language theory, where the recognizer for $D$ would be encoded as a
-pushdown automaton with finitely many
-states and stack of memory. Such PDAs are encodable in our theory, although they
-are a little cumbersome to use. Moreover, they
-seem overfit to classical reasoning as they don't preserve the grammar up to
-strong equivalence, only weak equivalence in general. \steven{cite Lambek
-  automata and state that is is a secondary goal to preserve as much structure as possible}
-By moving the stack of memory (the nats) into the
-state, we obtain infinitely many states, which is fine bc the states need only
-be a type for us to interface with it.
+We construct a parser for $\Dyck$ by building a determinstic automaton $M$ such
+that $\Trace_{M}(M.\init) \cong \Dyck$.
+
+\Cref{fig:DyckAutomaton} presents a sketch of the automaton $M$. The traditional
+approach to parse grammars like $\Dyck$ is to employ a stack to count how many
+open parentheses have been seen so far. We may think of
+$M$ as performing the same task as a stack machine, except instead of storing
+the number of open parentheses in a distinguished stack of memory, $M$ operates
+by storing this data in the current state. This is made feasible by specifying
+$M$ almost identically to how we defined DFAs, except we omit the
+finiteness condition on the states of the automaton.
+
+\steven{Mention that $\Trace_{M}$ is iso to string. Mention the same for DFAs above }
+
+% Unlike the finite automata seen thus far, $M$ has infinitely many states --- one
+% for each $n : \mathbb{N}$ with an additional fail state. This diverges a bit
+% from classical language theory, where the recognizer for $D$ would be encoded as a
+% pushdown automaton with finitely many
+% states and stack of memory. Such PDAs are encodable in our theory, although they
+% are a little cumbersome to use. Moreover, they
+% seem overfit to classical reasoning as they don't preserve the grammar up to
+% strong equivalence, only weak equivalence in general. \steven{cite Lambek
+%   automata and state that is is a secondary goal to preserve as much structure as possible}
+% By moving the stack of memory (the nats) into the
+% state, we obtain infinitely many states, which is fine bc the states need only
+% be a type for us to interface with it.
 
@@
 
-\subsection{LL(1)}
-%% 4. LL(1) parser
-
-\steven{TODO impute LL(1) example, mention that its weak equivalence is mechanized}
-
-\subsection{UnReStRiCtEd}
+% \subsection{LL(1)}
+% %% 4. LL(1) parser
+
+% \steven{TODO impute LL(1) example, mention that its weak equivalence is mechanized}
+
+\subsection{Turing Machines}
 %% 5. Expressiveness: Turing Machines
-
-%% Make sure to say which results are mechanized
+The expressive power of \theoryabbv spans the entire Chomsky hierachy, as we may
+encode the acccepting traces of Turing machines as linear types.
+
+\newcommand{\Turing}{\mathsf{Turing}}
+\newcommand{\snoc}{\mathsf{snoc}}
+
+\begin{figure}
+\begin{align*}
+\data &~\Turing~(t : \mathbb{Z} \to \Gamma): L~\where\\
+      & \nil : \ltonl {\left(\LinPiTy {t} {\mathbb{N} \to \Gamma} {\Turing~t} \right)} \\
+      & \snoc : \ltonl {\left( \LinPiTy {c}{\Sigma}{\LinPiTy {t} {\mathbb{N} \to \Gamma} {\Turing(\mathsf{push}(c , t)) \lto c \lto \Turing(t)}} \right)}
+\end{align*}
+\caption{Kleene Star as an inductive non-linear type}
+\label{fig:turinginductive}
+\end{figure}
+
+We nonlinearly encode the tape of a (one-sided) Turing machine as
+function from $\mathbb{N} \to \Gamma$, where $\Gamma$ is the tape alphabet.
+
+We may then define a term of type
+$\mathsf{parseTape} : \ltonl {\left( \stringg \lto \LinSigTy {t} {\mathbb{N} \to \Gamma} {\Turing(t)} \right)}$.
+In this manner, all of the interesting computations that the Turing machine
+would do occur solely inside of the nonlinear index. That is,
+$\mathsf{parseTape}$ does not discern which strings are accepted by the Turing
+machine, nor does it do any computation. $\mathsf{parseTape}$ simply loads the
+input string into the initial spaces on the tape.
+
+Once the string has been loaded onto the tape, we may further define the
+following term to run the Turing machine.
+\[
+  \mathsf{run} : \ltonl {\left( \bigoplus\limits_{t : \mathbb{N} \to \Gamma} \Turing(t) \lto \bigwith\limits_{\mathsf{fuel} : \mathbb{N}} \mathsf{Maybe} \left( \bigoplus\limits_{t : \mathbb{N} \to \Gamma} \bigoplus\limits_{b : \Bool} \bigoplus\limits_{\mathsf{trace} : \mathsf{TuringTrace}(b , t)} \Turing(t) \right) \right)}
+\]
+
+where $\mathsf{TuringTrace}$ is a nonlinear type denoting terminating traces
+through the Turing machine. Note that $\mathsf{run}$ returns a $\mathsf{Maybe}$
+type in a fuel parameter. Turing machines may not terminate. If they do
+terminate within $\mathsf{fuel}$-many steps, then we return the trace along with
+the current tape and a boolean denoting if the trace is accepting or rejcting.
+If the trace does not terminate within the alloted amount of fuel, then we
+timeout and return nothing.
 
@@
 
-\section{Semantics and Metatheory}
+\section{Denotational Semantics}
 \label{sec:semantics-and-metatheory}
 
-\subsection{Formal Grammars as Linear Types}
-\label{sec:formaltype}
-\max{this should be in its own section on semantics}
+To justify our intuitive explanation of linear types as formal
+grammars, we will now define a \emph{denotational semantics} that
+makes this mathematically precise. This semantics is the basis for our
+prototype implementation in Agda.
+
+\subsection{What is a formal grammar?}
+
+\begin{definition}
+  A \emph{formal language} $L$ is a function from strings to propositions.
+  A \emph{formal grammar} $A$ is a function from strings to small sets.
+\end{definition}
+
+\max{todo: need to decide if we need this section here or if it's in the intro already}.
 
@@
 
+\begin{definition}
+
+\end{definition}
+
+To handle our universe, we assume a universe of small sets in our
+metatheory that is closed under all of our type formers.
+\begin{definition}[Grammar Semantics]
+  We define the following interpretations by mutual recursion on the
+  judgments of \theoryabbv:
+  \begin{enumerate}
+  \item For each non-linear context $\Gamma \isCtx$, we define a set $\sem \Gamma$.
+  \item For each non-linear type $\Gamma \vdash X \isTy$, and element
+    $\gamma \in \sem\Gamma$, we define a set $\sem X \gamma$.
+  \item For each linear type $\Gamma A \isLinTy$ and element $\gamma
+    \in \sem\Gamma$, we define a formal grammar $\sem{A}\gamma$. We
+    similarly define a formal grammar $\sem\Delta\gamma$ for each
+    linear context $\Gamma \Delta\isLinCtx$.
+  \item For each non-linear term $\Gamma \vdash M : X$ and $\gamma \in \sem{\Gamma}$, we define an element $\sem{M}\gamma \in \sem{X}\gamma$.
+  \item For each linear term $\Gamma; \Delta \vdash e : A$ and $\gamma \in \sem{\Gamma}$ we define a parse transformer from $\sem{\Delta}\gamma$ to $\sem{A}\gamma$.
+  \end{enumerate}
+  And we verify the following conditions:
+  \begin{enumerate}
+  \item If $\Gamma \vdash X \isSmall$, then $\sem X \gamma$ is a small set.
+  \item If $\Gamma \vdash X \cong X'$ then for every $\gamma$, $\sem{X}\gamma = \sem{X'}\gamma$
+  \item If $\Gamma \vdash A \cong A'$ then for every $\gamma$, $\sem{A}\gamma = \sem{A'}\gamma$
+  \item If $\Gamma \vdash M \cong M' : X$ then for every $\gamma$, $\sem{M}\gamma = \sem{M'}\gamma$
+  \item If $\Gamma;\Delta \vdash e \cong e' : A$ then for every $\gamma$, $\sem{e}\gamma = \sem{e'}\gamma$
+  \end{enumerate}
+\end{definition}
+
+For reasons of brevity, we present only a selection of cases here. The full details are in the appendix.\max{todo: put the full details in the appendix}.
+
 Most of the non-linear semantics is standard \pedro{We should either cite something or add the non-linear semantics in the appendix}, so let us investigate in more
 depth the linear semantics.
-
 \begin{enumerate}
@@
         $\sem X : \sem \Gamma \to \Set_{i}$
-  \item A non-linear term $\Gamma \vdash e : X$ denotes a section
-        $\sem e : \Pi(\gamma : \sem \Gamma)\sem{X} \gamma$ \pedro{we have not defined what sections are}
+  \item A non-linear term $\Gamma \vdash e : X$ denotes a dependently typed function
+        $\sem e : \prod_{\gamma : \sem \Gamma}\sem{X} \gamma$ \pedro{we have not defined what sections are}
   \item Linear contexts $\Gamma \vdash \Delta$ and types

### 2024-10-31 paper/paper.tex
nfa/dfa defs

--- 2024-10-31 paper/paper.tex
@@
 
+\author{Nathan Varner}
+\affiliation{\department{Electrical Engineering and Computer Science}
+  \institution{University of Michigan}
+  \country{USA}
+}
+\email{nvarner@umich.edu}
+
 \author{Max S. New}
@@
 \paragraph{Unambiguity}
-In the grammars model, we say that a grammar is unambiguous if there is at most
+In the grammars-as-presheaves model of \theoryabbv, we say that a grammar is unambiguous if there is at most
 one parse tree for any input string. There are several ways to capture this
@@
 
-In the presence of distributive coproducts (which we probably add a) the following are equivalent:
+In the presence of distributive coproducts, the following are equivalent:
 \begin{itemize}
@@
 
+\newcommand{\states}{\mathsf{states}}
+\newcommand{\labelt}{\mathsf{label}}
+\newcommand{\transitions}{\mathsf{transitions}}
+\newcommand{\epstransitions}{\epsilon\mathsf{transitions}}
+\newcommand{\isAcc}{\mathsf{isAcc}}
+\newcommand{\init}{\mathsf{init}}
+\newcommand{\src}{\mathsf{src}}
+\newcommand{\dst}{\mathsf{dst}}
+\newcommand{\epssrc}{\epsilon\mathsf{src}}
+\newcommand{\epsdst}{\epsilon\mathsf{dst}}
+
+\paragraph{Nondeterministic Finite Automata} In \cref{sec:type-theory-examples}, we saw the demonstration of
+one particular NFA corresponding to the regular expression
+$(a^{*} \otimes b) \oplus c$. The definition of the linear type of traces
+through this
+NFA, as
+presented in \cref{fig:nfainductive}, inspects the concrete structure of the
+NFA.\ When generalizing from this example to an arbitrary NFA, we do not have the
+luxury of inspecting a concrete defintion. Therefore, we abstract over the
+specification of an arbitrary NFA and define traces with respect to that
+specification.
+
+Let the data of an NFA $N$ be given by the following nonlinear data,
+\begin{itemize}
+\item $N.\states$ --- a finite set of states
+\item $N.\init$ --- the initial state
+\item $N.\isAcc$ --- a decidable proposition over $N.\states$ denoting which
+        states are accepting
+\item $N.\transitions$ --- a finite set of labeled transitions
+\item $N.\labelt$ --- a mapping from $N.\transitions$ to the character labeling
+        the transition
+\item $N.\src$ --- a mapping from $N.\transitions$ to source state of a transition
+\item $N.\dst$ --- a mapping from $N.\transitions$ to destination state of a transition
+\item $N.\epstransitions$ --- a finite set of $\epsilon$-transitions
+\item $N.\epssrc$ --- a mapping from $N.\epstransitions$ to source state of a $\epsilon$-transition
+\item $N.\epsdst$ --- a mapping from $N.\epstransitions$ to destination state of a $\epsilon$-transition
+\end{itemize}
+
+For the above definition of an NFA $N$, in \cref{fig:nfatrace} we define a linear type of traces
+through $N$. Traces in $N$ may be built through one of three constructors. We
+may terminate a trace at an accepting state with the constructor $\nil$. If we
+had a trace beginning at the destination state of a transition, then we may use
+the $\cons$ constructor to linearly combine that trace with a parse of the label
+of the transition to build a trace beginning at the source of the transition.
+Finally, if we had a trace beginning at the destination of an
+$\epsilon$-transition then we may use $\epscons$ to pull it back along the
+$\epsilon$-transition and construct a trace beginning at the source of the $\epsilon$-transition.
+
+\begin{figure}
+\begin{align*}
+\data &~\Trace_{N} : (s : N.\states) \to L~\where\\
+      & \nil : \ltonl {\left( \LinPiTy {s} {N.\states} {\LinPiTy {\mathsf{acc}} {N.\isAcc(s)} {\Trace_{N}~s}} \right)}  \\
+      & \cons : \ltonl {\left( \LinPiTy {t} {N.\transitions} {(N.\labelt~t) \lto \Trace_{N}~(N.\dst(t)) \lto \Trace_{N}(N.\src(t))} \right)} \\
+      & \epscons : \ltonl {\left( \LinPiTy {t} {N.\epstransitions} {\Trace_{N}~(N.\epsdst(t)) \lto \Trace_{N}(N.\epssrc(t))} \right)}
+\end{align*}
+\caption{Traces through the NFA $N$ as an indexed inductive type}
+\label{fig:nfatrace}
+\end{figure}
+
+\paragraph{Determinstic Finite Automata}
+Just as above, we provide a specification for an arbitrary DFA. A DFA $D$ is
+given by the following data,
+
+\begin{itemize}
+\item $D.\states$ --- a finite set of states
+\item $D.\init$ --- the initial state
+\item $D.\isAcc$ --- a decidable proposition over $D.\states$ denoting which
+      states are accepting
+\item $D.\delta$ --- a transition function. Given a character $c$ and a state
+        $s$, $D.\delta_{c}(s)$ is the state for which there is a transition $s \overset{c}{\to} D.\delta_{c}(s)$
+\end{itemize}
+
+The linear type of traces through $D$ is given in \cref{fig:dfatrace}. These
+traces are either terminated in an accepting state with the $\nil$ constructor,
+or the $\cons$ constructor builds a trace out of state $s$ by linearly combining
+a parse of some character $c$ with a
+trace out of the state $D.\delta_{c}(s)$.
+
+\begin{figure}
+\begin{align*}
+\data &~\Trace_{D} : (s : D.\states) \to L~\where\\
+      & \nil : \ltonl {\left( \LinPiTy {s} {D.\states} {\LinPiTy {\mathsf{acc}} {D.\isAcc(s)} {\Trace_{D}~s}} \right)}  \\
+      & \cons : \ltonl {\left( \LinPiTy {c} {\Sigma} {\LinPiTy {s} {D.\states} {c \lto \Trace_{D}~(D.\delta_{c}(s)) \lto \Trace_{D}(s)}} \right)}
+\end{align*}
+\caption{Traces through the DFA $D$ as an indexed inductive type}
+\label{fig:dfatrace}
+\end{figure}
+
 \paragraph{Regex-NFA/Thompson}
@@
 
-\paragraph{Determinization of an NFA}
 \newcommand{\Agda}{Formalized in Agda}
@@
 
-\subsection{Unrestricted Grammars}
+\subsection{Turing Machines}
 %% 5. Expressiveness: Turing Machines
 The expressive power of \theoryabbv spans the entire Chomsky hierachy, as we may
-encode the grammars accepted by Turing machines.
+encode the acccepting traces of Turing machines as linear types.
+
+\begin{figure}
+\begin{align*}
+\data &~\mathsf{Turing}~(t : \mathbb{Z} \to \Gamma): L~\where\\
+      & \mathsf{nil} : \ltonl {\left(\LinPiTy {t} {\mathbb{Z} \to \Gamma} {\mathsf{Turing}~t} \right)} \\
+      & \mathsf{snoc} : \ltonl {(\mathsf{Turing} \lto A^{*} \lto A^{*})}
+\end{align*}
+\caption{Kleene Star as an inductive non-linear type}
+\label{fig:turinginductive}
+\end{figure}
+
+However, unlike
+the traces through the automata we have seen until now, the encoding of Turing
+machines
 

### 2024-10-31 paper/paper.tex
thompson wip

@@
 
-\paragraph{Regex-NFA/Thompson}
-
-\steven{TODO put regex-NFA iso here. Call it Thompson's construction?. Mention
-  its mechanized}
-
 \newcommand{\Agda}{Formalized in Agda}
@@
 
-\subsection{The Dyck Grammar}
-%% 3. Dyck language
-The Dyck grammar $D$ for balanced parens (summarize what it is + citation)
-
-\steven{somewhere in here, sneak in the benchmark where the Dyck parser could
-  handle 10's of thousands of characters in about 30s}
-
-\steven{this is a VERY loose sketch of the things I want to say about $D$}
-\Cref{fig:DyckAutomaton} presents an automaton $M$ that recognizes $D$. We show this
-precisely by constructing an isomorphism between $D$ and $\Trace_{M}~0~\true$.
-Unlike the finite automata seen thus far, $M$ has infinitely many states --- one
-for each $n : \mathbb{N}$ with an additional fail state. This diverges a bit
-from classical language theory, where the recognizer for $D$ would be encoded as a
-pushdown automaton with finitely many
-states and stack of memory. Such PDAs are encodable in our theory, although they
-are a little cumbersome to use. Moreover, they
-seem overfit to classical reasoning as they don't preserve the grammar up to
-strong equivalence, only weak equivalence in general. \steven{cite Lambek
-  automata and state that is is a secondary goal to preserve as much structure as possible}
-By moving the stack of memory (the nats) into the
-state, we obtain infinitely many states, which is fine bc the states need only
-be a type for us to interface with it.
-
-\steven{TODO sketch proof of equivalence. Mention that it is mechanized}
+\begin{theorem}[Thompson's Construction, \Agda]
+  Given a regular expression $R$, there exists an NFA $N$ such that $R$ is
+  strongly equivalent to $\Trace_{N}(N.\init)$.
+\end{theorem}
+
+\begin{proof}
+In this proof, we present a variant of Thompson's construction
+\cite{thompsonProgrammingTechniquesRegular1968} in which we induct on the
+structure of $R$ to construct strongly equivalent NFAs. We will sketch this
+proof by walking through some of the inductive cases, the remaining cases follow
+similar reasoning and the details may be found in the formalized artifact.
 
@@
   \begin{tikzpicture}[node distance = 17mm ]
-    \node[state, initial, accepting] (0) {$0$};
-    \node[state, right of=0] (1) {$1$};
-    \node[state, right of=1] (2) {$2$};
-    \node[right of=2] (3) {$\dots$};
-    \node[state, below of=0] (fail) {$\mathsf{fail}$};
-
-    \path[->] (0) edge[above, bend left] node{\stringquote{(}} (1)
-              (1) edge[below, bend left] node{\stringquote{)}} (0)
-              (1) edge[above, bend left] node{\stringquote{(}} (2)
-              (2) edge[below, bend left] node{\stringquote{)}} (1)
-              (2) edge[above, bend left] node{\stringquote{(}} (3)
-              (3) edge[below, bend left] node{\stringquote{)}} (2)
-              (fail) edge[loop right] node{\stringquote{(}, \stringquote{)}} (fail)
-              (0) edge[left] node{\stringquote{)}} (fail);
+    \node[state, initial] (0) {$0$};
+    \node[state, right of=0, accepting] (1) {$1$};
+
+    \path[->] (0) edge[above] node{$c$} (1);
   \end{tikzpicture}
-  \caption{Automaton $M$ for the Dyck grammar}
-  \label{fig:DyckAutomaton}
+  \caption{NFA $N_{c}$ for a single character $c$}
+  \label{fig:litnfa}
 \end{figure}
+
+First handle the case where $R$ is a literal character $c$. The NFA $N_{c}$ pictured in
+\cref{fig:litnfa} is strongly equivalent to the grammar $c$. To demonstrate this
+strong equivalence, we build terms between $c$ and $\Trace_{N_{c}}(N_{c}.\init)$
+and show they are mutually inverse.
+
+Define $f : \ltonl {(c \lto \Trace_{N_{c}}(N_{c}.\init))}$ as follows,
+
+\[
+  a : c \vdash f := \cons(a, \nil) : \Trace_{N_{c}}(N_{c}.\init)
+\]
+
+Define $g : \ltonl {(\Trace_{N_{c}}(N_{c}.\init) \lto c)}$ via induction on the
+trace. Map traces beginning in state 1, which are necessarily built via $\nil$,
+to the linear unit $I$. Then we must handle traces beginning in state 0, which
+are necessarily built via $\cons$. We decompose such a trace into a parse of the
+character $c$ and a trace beginning in state 1. We inductively transform the
+trace beginning in state 1 into a term of type $I$. Finally, we use the right
+unitor to turn a term of type $c \otimes I$ into one of type $c$.
+
+\[\begin{tikzcd}
+    {c \otimes \mathsf{Trace}_{N_{c}}(1)} &
+    {c \otimes I} &
+    c
+	\arrow["{id\otimes g}", from=1-1, to=1-2]
+	\arrow["{}", from=1-2, to=1-3]
+\end{tikzcd}\]
+
+Note that the grammar $c$ is unambiguous, so $g \circ f$ is equal to the
+identity on $c$. Additionally, we can show that $f$ carries the structure of an
+indexed algebra homomorphism for the family of functors defining the trace type
+$\Trace_{N_{c}}$. Because $g$ is defined via induction, it carries the structure
+of an algebra homomorphism as well. Thus, $f \circ g$ is also an algebra
+homomorphism. $f \circ g$ and the identity map are both then algebra
+homomorphisms from $\Trace_{N_{c}}$ to itself. Because $\Trace_{N_{c}}$ is the
+initial algebra for the relevant family of functors, there is only one
+homomorphism from it to itself and thus $f \circ g = \mathsf{id}$. Therefore,
+$f$ and $g$ are mutually inverse and the regular expression is strongly
+equivalent to the traces through the NFA in this case.
+
+Now consider the case where $R$ is the disjunction of two regular expressions
+$R_{1}$ and $R_{2}$. Inductively, we may build NFAs $N_{1}$ and $N_{2}$ that are
+strongly equivalent to $R_{1}$ and $R_{2}$, respectively. We define $N_{\oplus}$
+as a new NFA built as the disjunction of $N_{1}$ and $N_{2}$.
+$N_{\oplus}.\states$ comprises the states from $N_{1}$, the states from
+$N_{2}$, and a new marked initial state. The accepting states are precisely
+those that were accepting in the subautomata. The initial state has an
+$\epsilon$-transition to each of the initial states of the subautomata. Once in
+$N_{1}.\init$, the downstream transitions are a copy of those in $N_{1}$,
+likewise for $N_{2}.\init$.  That is, $N_{\oplus}$ contains a copy of each
+$N_{1}$ and $N_{2}$, and we choose which copy to inhabit by deciding which
+$\epsilon$-transition out of the initial state to take. A schematic
+representation of this automaton is given in \cref{fig:disjunctionnfa}.
+
+\begin{figure}
+  \begin{tikzpicture}[node distance = 17mm ]
+    \node[state, initial] (init) {$\init$};
+    \node[state, above right of=init] (init1) {$N_{1}.\init$};
+    \node[state, below right of=init] (init2) {$N_{2}.\init$};
+    \node[right of=init1] (dots1) {$\dots$};
+    \node[right of=init2] (dots2) {$\dots$};
+    \node[state, accepting, right of=dots1] (acc1) {$s_{1}$};
+    \node[state, accepting, right of=dots2] (acc2) {$s_{2}$};
+
+    \path[->] (init) edge[below] node{$\epsilon$} (init1)
+              (init) edge[above] node{$\epsilon$} (init2)
+              (init1) edge[above] node{} (dots1)
+              (init2) edge[above] node{} (dots2)
+              (dots1) edge[above] node{} (acc1)
+              (dots2) edge[above] node{} (acc2);
+  \end{tikzpicture}
+  \caption{NFA $N_{\oplus}$ as a disjunction of $N_{1}$ and $N_{2}$}
+  \label{fig:disjunctionnfa}
+\end{figure}
+
+Because $N_{1}$ and $N_{2}$ are each strongly equivalent to $R_{1}$ and $R_{2}$,
+it suffices to show that
+$\Trace_{N_{1}}(N_{1}.\init) \oplus \Trace_{N_{2}}(N_{2}.\init)$ is strongly
+equivalent to $\Trace_{N_{\oplus}}(N_{\oplus}.\init)$.
+
+Just like the case of a single character, we induct on traces when defining this
+strong equivalence. First we map out of $\Trace_{N_{\oplus}}(s)$. If $s$ is a
+state from $N_{1}$, then we send the trace from $N_{\oplus}$
+
+\steven{TODO finish this, my brain isn't working rn}
+
+\end{proof}
 

### 2024-11-01 paper/paper.tex
finish thompson sketch

--- 2024-10-31 paper/paper.tex
@@
 
-\newcommand{\isTy}{\textrm{ type}}
-\newcommand{\isCtx}{\textrm{ ctx}}
-\newcommand{\isSmall}{\textrm{ small}}
-\newcommand{\isLinTy}{\textrm{ lin. type}}
-\newcommand{\isLinCtx}{\textrm{ lin. ctx.}}
-
 \newcommand{\gluedNL}{{\mathcal G}_S}
@@
 left implicit. For instance, every linear context is parameterized by a nonlinear
-context, which is syntactically separated by a semicolon. If the nonlinear
+context, which is syntactically separate by a semicolon. If the nonlinear
 context is empty, we may often omit it; however, in this figure we make this
@@
 the nonlinear empty context. Then we explicit cast from
-an nonlinear $\uparrow$-wrapped type into a linear one. Such casts may often be
+an nonlinear $uparrow$-wrapped type into a linear one. Such casts may often be
 made implicitly, such as in $\cref{fig:kleeneabstractproof}$.
@@
 generative capacity --- the set of ``structural descriptions'', or parse trees,
-that it generates. Two grammars are said to be \emph{weakly equivalent} by
-Chomsky if they
+that it generates. Two grammars are said to be \emph{weakly equivalent} if they
 agree in their weak generative capacity, and likewise \emph{strongly equivalent}
@@
 
-\paragraph{Regular Expressions} The theory of regular expressions emerges as the
-subtheory of \theoryabbv generated by
-$\otimes$, $\oplus$, and Kleene star over the types $\bot$, $I$, and characters drawn from $\Sigma$.
+\paragraph{Regular Expressions} The theory of regular expressions emerges as a
+subtheory of \theoryabbv by restricting the typing connectives to only use
+$\otimes$ and $\oplus$ generated over $\bot$, $I$, and characters drawn from $\Sigma$.
 
@@
 
-\subsection{The Dyck Grammar}
-%% 3. Dyck language
-\newcommand{\balanced}{\mathsf{balanced}}
-\newcommand{\Dyck}{\mathsf{Dyck}}
-Moving up the Chomsky hierarchy from regular grammars, we arrive at context-free
-grammars (CFGs). In many ways, the smallest nontrivial CFG is the Dyck grammar
-$\Dyck$ --- the grammar of balanced parentheses. Moreover, the Chomsky-Sch\"utzenberger
-representation theorem \cite{CHOMSKY1959118} demonstrates that $\Dyck$, and grammars like it,
-cut to the essence of what it means to be context free.
-
-$\Dyck$ is a grammar over the alphabet $\{ \stringquote{(}, \stringquote{)} \}$
-and is defined in \cref{fig:dyckinductive}. The $\nil$ constructor shows that
-the empty string is balanced, and the $\balanced$ constrcutors builds a balanced
-parse by wrapping an already balanced parse in an additional set of parentheses
-then following it with another balanced parse.
-
-\begin{figure}
-\begin{align*}
-\data &~\Dyck : L~\where\\
-      & \nil : \ltonl {\Dyck} \\
-      & \balanced : \ltonl {(\stringquote{(} \lto \Dyck \lto \stringquote{)} \lto \Dyck \lto \Dyck)}
-\end{align*}
-\caption{Kleene Star as an inductive type}
-\label{fig:dyckinductive}
-\end{figure}
-
-\steven{somewhere in here, sneak in the benchmark where the Dyck parser could
-  handle 10's of thousands of characters in about 30s}
-
-We construct a parser for $\Dyck$ by building a determinstic automaton $M$ such
-that $\Trace_{M}(M.\init) \cong \Dyck$.
-
-\Cref{fig:DyckAutomaton} presents a sketch of the automaton $M$. The traditional
-approach to parse grammars like $\Dyck$ is to employ a stack to count how many
-open parentheses have been seen so far. We may think of
-$M$ as performing the same task as a stack machine, except instead of storing
-the number of open parentheses in a distinguished stack of memory, $M$ operates
-by storing this data in the current state. This is made feasible by specifying
-$M$ almost identically to how we defined DFAs, except we omit the
-finiteness condition on the states of the automaton.
-
-\steven{Mention that $\Trace_{M}$ is iso to string. Mention the same for DFAs above }
-
-% Unlike the finite automata seen thus far, $M$ has infinitely many states --- one
-% for each $n : \mathbb{N}$ with an additional fail state. This diverges a bit
-% from classical language theory, where the recognizer for $D$ would be encoded as a
-% pushdown automaton with finitely many
-% states and stack of memory. Such PDAs are encodable in our theory, although they
-% are a little cumbersome to use. Moreover, they
-% seem overfit to classical reasoning as they don't preserve the grammar up to
-% strong equivalence, only weak equivalence in general. \steven{cite Lambek
-%   automata and state that is is a secondary goal to preserve as much structure as possible}
-% By moving the stack of memory (the nats) into the
-% state, we obtain infinitely many states, which is fine bc the states need only
-% be a type for us to interface with it.
-
-\steven{TODO sketch proof of equivalence. Mention that it is mechanized}
-
-\begin{figure}
-  \begin{tikzpicture}[node distance = 17mm ]
-    \node[state, initial, accepting] (0) {$0$};
-    \node[state, right of=0] (1) {$1$};
-    \node[state, right of=1] (2) {$2$};
-    \node[right of=2] (3) {$\dots$};
-    \node[state, below of=0] (fail) {$\mathsf{fail}$};
-
-    \path[->] (0) edge[above, bend left] node{\stringquote{(}} (1)
-              (1) edge[below, bend left] node{\stringquote{)}} (0)
-              (1) edge[above, bend left] node{\stringquote{(}} (2)
-              (2) edge[below, bend left] node{\stringquote{)}} (1)
-              (2) edge[above, bend left] node{\stringquote{(}} (3)
-              (3) edge[below, bend left] node{\stringquote{)}} (2)
-              (fail) edge[loop right] node{\stringquote{(}, \stringquote{)}} (fail)
-              (0) edge[left] node{\stringquote{)}} (fail);
-  \end{tikzpicture}
-  \caption{Automaton $M$ for the Dyck grammar}
-  \label{fig:DyckAutomaton}
-\end{figure}
-
-% \subsection{LL(1)}
-% %% 4. LL(1) parser
-
-% \steven{TODO impute LL(1) example, mention that its weak equivalence is mechanized}
+\subsection{LL(1)}
+%% 4. LL(1) parser
+
+\steven{TODO impute LL(1) example, mention that its weak equivalence is mechanized}
 
@@
 
-\newcommand{\Turing}{\mathsf{Turing}}
-\newcommand{\snoc}{\mathsf{snoc}}
-
 \begin{figure}
 \begin{align*}
-\data &~\Turing~(t : \mathbb{Z} \to \Gamma): L~\where\\
-      & \nil : \ltonl {\left(\LinPiTy {t} {\mathbb{N} \to \Gamma} {\Turing~t} \right)} \\
-      & \snoc : \ltonl {\left( \LinPiTy {c}{\Sigma}{\LinPiTy {t} {\mathbb{N} \to \Gamma} {\Turing(\mathsf{push}(c , t)) \lto c \lto \Turing(t)}} \right)}
+\data &~\mathsf{Turing}~(t : \mathbb{Z} \to \Gamma): L~\where\\
+      & \mathsf{nil} : \ltonl {\left(\LinPiTy {t} {\mathbb{Z} \to \Gamma} {\mathsf{Turing}~t} \right)} \\
+      & \mathsf{snoc} : \ltonl {(\mathsf{Turing} \lto A^{*} \lto A^{*})}
 \end{align*}
@@
 
-We nonlinearly encode the tape of a (one-sided) Turing machine as
-function from $\mathbb{N} \to \Gamma$, where $\Gamma$ is the tape alphabet.
-
-We may then define a term of type
-$\mathsf{parseTape} : \ltonl {\left( \stringg \lto \LinSigTy {t} {\mathbb{N} \to \Gamma} {\Turing(t)} \right)}$.
-In this manner, all of the interesting computations that the Turing machine
-would do occur solely inside of the nonlinear index. That is,
-$\mathsf{parseTape}$ does not discern which strings are accepted by the Turing
-machine, nor does it do any computation. $\mathsf{parseTape}$ simply loads the
-input string into the initial spaces on the tape.
-
-Once the string has been loaded onto the tape, we may further define the
-following term to run the Turing machine.
-\[
-  \mathsf{run} : \ltonl {\left( \bigoplus\limits_{t : \mathbb{N} \to \Gamma} \Turing(t) \lto \bigwith\limits_{\mathsf{fuel} : \mathbb{N}} \mathsf{Maybe} \left( \bigoplus\limits_{t : \mathbb{N} \to \Gamma} \bigoplus\limits_{b : \Bool} \bigoplus\limits_{\mathsf{trace} : \mathsf{TuringTrace}(b , t)} \Turing(t) \right) \right)}
-\]
-
-where $\mathsf{TuringTrace}$ is a nonlinear type denoting terminating traces
-through the Turing machine. Note that $\mathsf{run}$ returns a $\mathsf{Maybe}$
-type in a fuel parameter. Turing machines may not terminate. If they do
-terminate within $\mathsf{fuel}$-many steps, then we return the trace along with
-the current tape and a boolean denoting if the trace is accepting or rejcting.
-If the trace does not terminate within the alloted amount of fuel, then we
-timeout and return nothing.
+However, unlike
+the traces through the automata we have seen until now, the encoding of Turing
+machines
+
+We nonlinearly encode the tape of the Turing machine as
+function from $\mathbb{Z} \to \Gamma$, where $\Gamma$ is your tape alphabet. The
+grammar of traces for a Turing machine linear consumes the input string and
+writes it to the tape. From here, we nonlinear reason about machine transitions.
+We execute the machine and track its current state as part of the nonlinear
+indexing data.
+
+%% Make sure to say which results are mechanized
 
@@
 
-\section{Denotational Semantics}
+\section{Semantics and Metatheory}
 \label{sec:semantics-and-metatheory}
 
-To justify our intuitive explanation of linear types as formal
-grammars, we will now define a \emph{denotational semantics} that
-makes this mathematically precise. This semantics is the basis for our
-prototype implementation in Agda.
-
-\subsection{What is a formal grammar?}
-
-\begin{definition}
-  A \emph{formal language} $L$ is a function from strings to propositions.
-  A \emph{formal grammar} $A$ is a function from strings to small sets.
-\end{definition}
-
-\max{todo: need to decide if we need this section here or if it's in the intro already}.
+\subsection{Formal Grammars as Linear Types}
+\label{sec:formaltype}
+\max{this should be in its own section on semantics}
 
@@
 
-\begin{definition}
-
-\end{definition}
-
-To handle our universe, we assume a universe of small sets in our
-metatheory that is closed under all of our type formers.
-\begin{definition}[Grammar Semantics]
-  We define the following interpretations by mutual recursion on the
-  judgments of \theoryabbv:
-  \begin{enumerate}
-  \item For each non-linear context $\Gamma \isCtx$, we define a set $\sem \Gamma$.
-  \item For each non-linear type $\Gamma \vdash X \isTy$, and element
-    $\gamma \in \sem\Gamma$, we define a set $\sem X \gamma$.
-  \item For each linear type $\Gamma A \isLinTy$ and element $\gamma
-    \in \sem\Gamma$, we define a formal grammar $\sem{A}\gamma$. We
-    similarly define a formal grammar $\sem\Delta\gamma$ for each
-    linear context $\Gamma \Delta\isLinCtx$.
-  \item For each non-linear term $\Gamma \vdash M : X$ and $\gamma \in \sem{\Gamma}$, we define an element $\sem{M}\gamma \in \sem{X}\gamma$.
-  \item For each linear term $\Gamma; \Delta \vdash e : A$ and $\gamma \in \sem{\Gamma}$ we define a parse transformer from $\sem{\Delta}\gamma$ to $\sem{A}\gamma$.
-  \end{enumerate}
-  And we verify the following conditions:
-  \begin{enumerate}
-  \item If $\Gamma \vdash X \isSmall$, then $\sem X \gamma$ is a small set.
-  \item If $\Gamma \vdash X \cong X'$ then for every $\gamma$, $\sem{X}\gamma = \sem{X'}\gamma$
-  \item If $\Gamma \vdash A \cong A'$ then for every $\gamma$, $\sem{A}\gamma = \sem{A'}\gamma$
-  \item If $\Gamma \vdash M \cong M' : X$ then for every $\gamma$, $\sem{M}\gamma = \sem{M'}\gamma$
-  \item If $\Gamma;\Delta \vdash e \cong e' : A$ then for every $\gamma$, $\sem{e}\gamma = \sem{e'}\gamma$
-  \end{enumerate}
-\end{definition}
-
-For reasons of brevity, we present only a selection of cases here. The full details are in the appendix.\max{todo: put the full details in the appendix}.
-
 Most of the non-linear semantics is standard \pedro{We should either cite something or add the non-linear semantics in the appendix}, so let us investigate in more
 depth the linear semantics.
+
 \begin{enumerate}
@@
         $\sem X : \sem \Gamma \to \Set_{i}$
-  \item A non-linear term $\Gamma \vdash e : X$ denotes a dependently typed function
-        $\sem e : \prod_{\gamma : \sem \Gamma}\sem{X} \gamma$ \pedro{we have not defined what sections are}
+  \item A non-linear term $\Gamma \vdash e : X$ denotes a section
+        $\sem e : \Pi(\gamma : \sem \Gamma)\sem{X} \gamma$ \pedro{we have not defined what sections are}
   \item Linear contexts $\Gamma \vdash \Delta$ and types

### 2024-11-01 paper/paper.tex
Turing

@@
 
+\newcommand{\Turing}{\mathsf{Turing}}
+\newcommand{\snoc}{\mathsf{snoc}}
+
 \begin{figure}
 \begin{align*}
-\data &~\mathsf{Turing}~(t : \mathbb{Z} \to \Gamma): L~\where\\
-      & \mathsf{nil} : \ltonl {\left(\LinPiTy {t} {\mathbb{Z} \to \Gamma} {\mathsf{Turing}~t} \right)} \\
-      & \mathsf{snoc} : \ltonl {(\mathsf{Turing} \lto A^{*} \lto A^{*})}
+\data &~\Turing~(t : \mathbb{Z} \to \Gamma): L~\where\\
+      & \nil : \ltonl {\left(\LinPiTy {t} {\mathbb{N} \to \Gamma} {\Turing~t} \right)} \\
+      & \snoc : \ltonl {\left( \LinPiTy {c}{\Sigma}{\LinPiTy {t} {\mathbb{N} \to \Gamma} {\Turing(\mathsf{push}(c , t)) \lto c \lto \Turing(t)}} \right)}
 \end{align*}
@@
 
-However, unlike
-the traces through the automata we have seen until now, the encoding of Turing
-machines
-
-We nonlinearly encode the tape of the Turing machine as
-function from $\mathbb{Z} \to \Gamma$, where $\Gamma$ is your tape alphabet. The
-grammar of traces for a Turing machine linear consumes the input string and
-writes it to the tape. From here, we nonlinear reason about machine transitions.
-We execute the machine and track its current state as part of the nonlinear
-indexing data.
-
-%% Make sure to say which results are mechanized
+We nonlinearly encode the tape of a (one-sided) Turing machine as
+function from $\mathbb{N} \to \Gamma$, where $\Gamma$ is the tape alphabet.
+
+We may then define a term of type
+$\mathsf{parseTape} : \ltonl {\left( \stringg \lto \LinSigTy {t} {\mathbb{N} \to \Gamma} {\Turing(t)} \right)}$.
+In this manner, all of the interesting computations that the Turing machine
+would do occur solely inside of the nonlinear index. That is,
+$\mathsf{parseTape}$ does not discern which strings are accepted by the Turing
+machine, nor does it do any computation. $\mathsf{parseTape}$ simply loads the
+input string into the initial spaces on the tape.
+
+Once the string has been loaded onto the tape, we may further define the
+following term to run the Turing machine.
+\[
+  \mathsf{run} : \ltonl {\left( \bigoplus\limits_{t : \mathbb{N} \to \Gamma} \Turing(t) \lto \bigwith\limits_{\mathsf{fuel} : \mathbb{N}} \mathsf{Maybe} \left( \bigoplus\limits_{t : \mathbb{N} \to \Gamma} \bigoplus\limits_{b : \Bool} \bigoplus\limits_{\mathsf{trace} : \mathsf{TuringTrace}(b , t)} \Turing(t) \right) \right)}
+\]
+
+where $\mathsf{TuringTrace}$ is a nonlinear type denoting terminating traces
+through the Turing machine. Note that $\mathsf{run}$ returns a $\mathsf{Maybe}$
+type in a fuel parameter. Turing machines may not terminate. If they do
+terminate within $\mathsf{fuel}$-many steps, then we return the trace along with
+the current tape and a boolean denoting if the trace is accepting or rejcting.
+If the trace does not terminate within the alloted amount of fuel, then we
+timeout and return nothing.
 

### 2024-11-01 paper/paper.tex
dyck wip

@@
 generative capacity --- the set of ``structural descriptions'', or parse trees,
-that it generates. Two grammars are said to be \emph{weakly equivalent} if they
+that it generates. Two grammars are said to be \emph{weakly equivalent} by
+Chomsky if they
 agree in their weak generative capacity, and likewise \emph{strongly equivalent}
@@
 
-\paragraph{Regular Expressions} The theory of regular expressions emerges as a
-subtheory of \theoryabbv by restricting the typing connectives to only use
-$\otimes$ and $\oplus$ generated over $\bot$, $I$, and characters drawn from $\Sigma$.
+\paragraph{Regular Expressions} The theory of regular expressions emerges as the
+subtheory of \theoryabbv generated by
+$\otimes$, $\oplus$, and Kleene star over the types $\bot$, $I$, and characters drawn from $\Sigma$.
 
@@
 
-\subsection{LL(1)}
-%% 4. LL(1) parser
-
-\steven{TODO impute LL(1) example, mention that its weak equivalence is mechanized}
+\subsection{The Dyck Grammar}
+%% 3. Dyck language
+\newcommand{\balanced}{\mathsf{balanced}}
+\newcommand{\Dyck}{\mathsf{Dyck}}
+Moving up the Chomsky hierarchy from regular grammars, we arrive at context-free
+grammars (CFGs). In many ways, the smallest nontrivial CFG is the Dyck grammar
+$\Dyck$ --- the grammar of balanced parentheses. Moreover, the Chomsky-Sch\"utzenberger
+representation theorem \cite{CHOMSKY1959118} demonstrates that $\Dyck$, and grammars like it,
+cut to the essence of what it means to be context free.
+
+$\Dyck$ is a grammar over the alphabet $\{ \stringquote{(}, \stringquote{)} \}$
+and is defined in \cref{fig:dyckinductive}. The $\nil$ constructor shows that
+the empty string is balanced, and the $\balanced$ constrcutors builds a balanced
+parse by wrapping an already balanced parse in an additional set of parentheses
+then following it with another balanced parse.
+
+\begin{figure}
+\begin{align*}
+\data &~\Dyck : L~\where\\
+      & \nil : \ltonl {\Dyck} \\
+      & \balanced : \ltonl {(\stringquote{(} \lto \Dyck \lto \stringquote{)} \lto \Dyck \lto \Dyck)}
+\end{align*}
+\caption{Kleene Star as an inductive type}
+\label{fig:dyckinductive}
+\end{figure}
+
+\steven{somewhere in here, sneak in the benchmark where the Dyck parser could
+  handle 10's of thousands of characters in about 30s}
+
+We construct a parser for $\Dyck$ by building a determinstic automaton $M$ such
+that $\Trace_{M}(M.\init) \cong \Dyck$.
+
+\Cref{fig:DyckAutomaton} presents a sketch of the automaton $M$. The traditional
+approach to parse grammars like $\Dyck$ is to employ a stack to count how many
+open parentheses have been seen so far. We may think of
+$M$ as performing the same task as a stack machine, except instead of storing
+the number of open parentheses in a distinguished stack of memory, $M$ operates
+by storing this data in the current state. This is made feasible by specifying
+$M$ almost identically to how we defined DFAs, except we omit the
+finiteness condition on the states of the automaton.
+
+\steven{Mention that $\Trace_{M}$ is iso to string. Mention the same for DFAs above }
+
+% Unlike the finite automata seen thus far, $M$ has infinitely many states --- one
+% for each $n : \mathbb{N}$ with an additional fail state. This diverges a bit
+% from classical language theory, where the recognizer for $D$ would be encoded as a
+% pushdown automaton with finitely many
+% states and stack of memory. Such PDAs are encodable in our theory, although they
+% are a little cumbersome to use. Moreover, they
+% seem overfit to classical reasoning as they don't preserve the grammar up to
+% strong equivalence, only weak equivalence in general. \steven{cite Lambek
+%   automata and state that is is a secondary goal to preserve as much structure as possible}
+% By moving the stack of memory (the nats) into the
+% state, we obtain infinitely many states, which is fine bc the states need only
+% be a type for us to interface with it.
+
+\steven{TODO sketch proof of equivalence. Mention that it is mechanized}
+
+\begin{figure}
+  \begin{tikzpicture}[node distance = 17mm ]
+    \node[state, initial, accepting] (0) {$0$};
+    \node[state, right of=0] (1) {$1$};
+    \node[state, right of=1] (2) {$2$};
+    \node[right of=2] (3) {$\dots$};
+    \node[state, below of=0] (fail) {$\mathsf{fail}$};
+
+    \path[->] (0) edge[above, bend left] node{\stringquote{(}} (1)
+              (1) edge[below, bend left] node{\stringquote{)}} (0)
+              (1) edge[above, bend left] node{\stringquote{(}} (2)
+              (2) edge[below, bend left] node{\stringquote{)}} (1)
+              (2) edge[above, bend left] node{\stringquote{(}} (3)
+              (3) edge[below, bend left] node{\stringquote{)}} (2)
+              (fail) edge[loop right] node{\stringquote{(}, \stringquote{)}} (fail)
+              (0) edge[left] node{\stringquote{)}} (fail);
+  \end{tikzpicture}
+  \caption{Automaton $M$ for the Dyck grammar}
+  \label{fig:DyckAutomaton}
+\end{figure}
+
+% \subsection{LL(1)}
+% %% 4. LL(1) parser
+
+% \steven{TODO impute LL(1) example, mention that its weak equivalence is mechanized}
 

### 2024-11-01 paper/paper.tex
DFA parser wip

@@
 \newcommand{\LinPiTy}[3]{\textstyle\bigamp (#1 : #2). #3}
+\newcommand{\LinPiTyLimit}[3]{\bigwith\limits_{#1 : #2} #3}
 \newcommand{\LinSigTy}[3]{\textstyle\bigoplus (#1 : #2). #3}
+\newcommand{\LinSigTyLimit}[3]{\bigoplus\limits_{#1 : #2} #3}
 %% \newcommand{\DepWith}[2]{{\textstyle\bigamp}\limits_{#1}{#2}
@@
     \and
-    \inferrule{\Gamma ; \Delta \vdash e : \LinPiTy x X A \\ \Gamma \vdash e' : X}{\Gamma ; \Delta \vdash e\,.\pi\,M : \subst A {e'} x}
+    \inferrule{\Gamma ; \Delta \vdash e : \LinPiTy x X A \\ \Gamma \vdash M : X}{\Gamma ; \Delta \vdash e\,.\pi\,M : \subst A {e'} x}
     %
@@
 
+Through inclusion of this axiom, $! : \stringg \to \top$ is an isomorphism, and
+thus also a monomorphism. That is, $\stringg$ is unambiguous.
+
 \paragraph{Parseability}
@@
 \newcommand{\dst}{\mathsf{dst}}
+\newcommand{\parse}{\mathsf{parse}}
+\newcommand{\print}{\mathsf{print}}
 \newcommand{\epssrc}{\epsilon\mathsf{src}}
@@
       states are accepting
-\item $D.\delta$ --- a transition function. Given a character $c$ and a state
+\item $D.\delta$ --- a deterministic transition function. Given a character $c$ and a state
         $s$, $D.\delta_{c}(s)$ is the state for which there is a transition $s \overset{c}{\to} D.\delta_{c}(s)$
@@
 
-The linear type of traces through $D$ is given in \cref{fig:dfatrace}. These
-traces are either terminated in an accepting state with the $\nil$ constructor,
-or the $\cons$ constructor builds a trace out of state $s$ by linearly combining
-a parse of some character $c$ with a
-trace out of the state $D.\delta_{c}(s)$.
-
 \begin{figure}
 \begin{align*}
-\data &~\Trace_{D} : (s : D.\states) \to L~\where\\
-      & \nil : \ltonl {\left( \LinPiTy {s} {D.\states} {\LinPiTy {\mathsf{acc}} {D.\isAcc(s)} {\Trace_{D}~s}} \right)}  \\
-      & \cons : \ltonl {\left( \LinPiTy {c} {\Sigma} {\LinPiTy {s} {D.\states} {c \lto \Trace_{D}~(D.\delta_{c}(s)) \lto \Trace_{D}(s)}} \right)}
+\data &~\Trace_{D} : (s : D.\states) (b : \Bool) \to L~\where\\
+      & \nil : \ltonl {\left( \LinPiTyLimit {s} {D.\states} {\Trace_{D}(s , D.\isAcc(s))} \right)}  \\
+      & \cons : \ltonl {\left( \LinPiTyLimit {c} {\Sigma} {\LinPiTyLimit {s} {D.\states} {c \lto \LinSigTyLimit {b}{\Bool}{\Trace_{D}~(D.\delta_{c}(s), b)} \lto \Trace_{D}(s, b)}} \right)}
 \end{align*}
@@
 \label{fig:dfatrace}
+\end{figure}
+
+\steven{TODO make the DFA parser figure fit}
+
+The linear type of traces through $D$ is given in \cref{fig:dfatrace}. These
+traces are either terminated in a state $s$ with the $\nil$ constructor. When
+terminating a trace, we tag it with a Boolean that marks if the end state was
+accepting or not. The $\cons$ constructor builds a trace out of state $s$ by linearly combining
+a parse of some character $c$ with a
+trace out of the state $D.\delta_{c}(s)$. The trace built with $\cons$ is
+accepting if and only if the trace out of $D.\delta_{c}(s)$ is accepting.
+
+Because DFAs are deterministic, we are able to prove that their type of traces
+are unambiguous and define a parser for them. In particular we show that
+$\LinPiTy {s} {D.\states} {\LinSigTy {b} {\Bool} \Trace_{D}(s , b)}$ is isomorphic to
+$\stringg$, which simultaneously constructs a parser and demonstrates that
+traces through $D$ is unambiguous. This isomorphism intuitively says that we pay
+unambiguously build traces through $D$ beginning at any state $s$, and further
+that from any trace through $D$ we can uniquely recover the string that built
+the trace.
+
+The first half of the isomorphism,
+$\parse_{D} : \ltonl {\left( \stringg \lto \LinPiTy {s} {D.\states} {\Trace_{D}(s)} \right)}$,
+is defined by induction on strings in \cref{fig:dfaparser}. If this string is empty, then $\parse_{D}$
+defines a linear function that terminates a trace at the input state $s$. If the
+string is nonempty, then $\parse_{D}$ walks forward from the input state $s$ by
+the character at the head of the string.
+
+\begin{figure}
+\[
+\inferrule
+{
+  \cdot \vdash mt := \withlamb {s} {\sigma~(D.\isAcc(b))~(\nil.\pi~s)}   : \LinPiTyLimit {s} {D.\states} {\LinSigTyLimit {b} {\Bool} {\Trace_{D}(s, b)}}
+  \\
+  ch : \charg , tr : \LinPiTyLimit {s} {D.\states} {\LinSigTyLimit {b} {\Bool } {\Trace_{D}(s, b)}} \vdash
+  f := \withlamb {s} {\letin {\sigma~c~p_{c}} {ch} {\letin {\sigma~b~t} {tr(D.\delta_{c}(s))} {\sigma~b~(\cons(p_{c} , t))}}}
+  : \LinPiTyLimit {s} {D.\states} {\LinSigTyLimit {b} {\Bool } {\Trace_{D}(s, b)}}
+}
+{w : \stringg \vdash \parse_{D} := \fold(mt, f)(w) : \LinPiTyLimit {s} {D.\states} {\LinSigTyLimit {b} {\Bool } {\Trace_{D}(s, b)}}}
+\]
+\caption{Parser for DFA traces}
+\label{fig:dfaparser}
+\end{figure}
+
+The second half of the isomorphism,
+$\print_{D} : \ltonl {\left( \LinPiTy {s} {D.\states} {\Trace_{D}(s)} \lto \stringg \right)}$,
+is given in \cref{fig:printdfa} by induction on traces. If the trace is defined
+via $\nil$, then $\print_{D}$ returns the empty string. Otherwise, if the trace
+is defined by $\cons$ then $\parse_{D}$ appends the character $c$ to the output
+string and recurses.
+
+\begin{figure}
+\[
+\inferrule
+{
+}
+{tr : \LinPiTyLimit {s} {D.\states} {\LinSigTyLimit {b} {\Bool } {\Trace_{D}(s, b)}} \vdash asdf : \stringg}
+\]
+\caption{Parser for DFA traces}
+\label{fig:printdfa}
 \end{figure}

### 2024-11-04 paper/paper.tex
writing tweaks and comments

@@
 \newcommand{\cons}{\mathsf{cons}}
-\newcommand{\epscons}{\epsilon\mathsf{cons}}
+\newcommand{\epscons}{\varepsilon\mathsf{cons}}
 \newcommand{\data}{\mathsf{data}}
@@
 grammars and parses of them internal to \theoryabbv. In each of our examples, we will look at
-grammars over the three character alphabet $\{ a , b , c\}$.
-
-Each letter of the
-alphabet corresponds to a base type in our calculus. For instance, the grammar
-for the single character $a$ has a single parse tree for the string \stringquote{a}.
+grammars over the three character alphabet $\{ a , b , c\}$. Each letter of the
+alphabet is a base type in our calculus. For instance, the grammar
+for the single character $a$ has a single parse tree for the string \stringquote{a},
 and no parse trees at any other strings.
@@
 \paragraph{Finite Grammars}
+\max{epsilon? bottom?}
 First consider finite grammars --- those built from base types via disjunctions $\oplus$ and
@@
 
-Parses of a fixed string $w$ into a grammar $A$ are internally represented as
-terms of type $A$ in the context $\internalize w$, where $\internalize w$ is a
-comma-separated list of the constituent characters of $w$. In particular, to
-parse \stringquote{ac}, we build parse trees in the context
-$\internalize{\stringquote{ab}} = x : a , y : b$. Thus the term
-$\inl(x \otimes y)$ from \cref{fig:fingram} details the
-manner by which \stringquote{ac} matches the grammar $(a \otimes b) \oplus c$.
+For a grammar $A$, a parse tree of a string $w$ can be represented as
+a term of type $A$ in the context $\internalize w$, where
+$\internalize w$ is a context with one variable for each character of
+$w$. For example, to define a parse tree for \stringquote{ab}, we use
+the context $\internalize{\stringquote{ab}} = x : a , y : b$. Then
+\cref{fig:fingram} shows a typing derivation that shows that $\inl(x
+\otimes y)$ defines a valid parse tree for the grammar $(a \otimes b)
+\oplus c$.
 
@@
 
+\max{I think concrete examples here would be helpful}
+For this interpretation of terms as parse trees to make sense, our
+calculus cannot allow for \emph{any} of the usual structural rules of
+a type theory: weakening, contraction and exchange. Weakening allows
+for variables to go unused, while contraction allows for the same
+variable to be used twice, but in a parse tree, every character must
+be accounted for exactly once.
+%
+Finally, exchange allows for variables in the context to be reordered,
+but the ordering of characters in a string cannot be ignored in
+parsing: otherwise we would have that $\inl(x\otimes y)$ is a valid
+term in context $y:b,x:a$, but there should be no parses of $ba$ in
+this grammar.
+
 Because derivations in a linear context correspond to parse trees in this
@@
 \paragraph{Regular Expressions}
-Next, we demonstrate how to parse a regular expression in \theoryabbv. Similar
-to finite grammars, regular
-expressions are built inductively using base types, $\oplus$, and $\otimes$ ;
-however, unlike finite grammars, regular expressions additionally have a
-restricted form of recursion via Kleene star $(\cdot)^{\ast}$.
+Regular expressions can be encoded as types using base types,
+$\oplus$, and $\otimes$ and the Kleene star $(\cdot)^{\ast}$.  For a
+grammar $A$, we can define the Kleene star $A^{*}$ as a particular
+\emph{inductive linear type} of linear lists, as shown in
+\cref{fig:kleenestarinductive}. Here $A^{*} : L$ means we are defining
+a \emph{linear} type. $A^{*}$ has two constructors: $\nil$, which
+builds a parse of type $A^{*}$ from nothing; and $\cons$, which
+linearly consumes a parse of $A$ and a parse of $A^{*}$ and builds a
+parse of $A^{*}$. This linear consumption is defined by the linear
+function type $\ltonl$. The arrow, $\uparrow$, wrapping these
+constructors means that the constructors by themselves do not consume
+any resources, and so are \emph{non-linear} values themselves.
 
@@
 
-For a grammar $A$, the recursive nature of $A^{*}$ is captured by defining
-$A^{*}$ to be an inductive linear
-type, as shown in \cref{fig:kleenestarinductive}. $A^{*}$ has two constructors:
-$\nil$, which builds a parse of type $A^{*}$ in the empty context; and $\cons$,
-which takes in of $A$ and a parse of $A^{*}$ and linearly combines them to build
-a parse of $A^{*}$. The type defined by these constructors belong to the
-universe of linear types $L$.
-
-The arrow, $\uparrow$, wrapping these constructors denotes
-an interaction between the linear and nonlinear parts of the theory.
-In short, a nonlinear term
-$M : \ltonl A^{*}$ denotes a linear term in the empty context
-$\cdot \vdash M : A^{*}$. Later in
-this paper, we will further discuss the behavior of $\uparrow$.
-At this moment,
-it suffices to read these arrows as allow us to use the reuse the names
-``$\nil$'' and ``$\cons$''
-as symbols for constructors. That is, the $\nil$ constructor is not linearly
-consumed at usage time
-
-Just as in \cref{fig:fingram}, we can show that particular strings match a Kleene
-star. In \cref{fig:kleenestarderivation}, we demonstrate that \stringquote{ab}
-matches $(a^{*} \otimes b) \oplus c$ through repeated application of the Kleene
-star constructors. In this proof tree we expand some syntactic details that are often
-left implicit. For instance, every linear context is parameterized by a nonlinear
-context, which is syntactically separated by a semicolon. If the nonlinear
-context is empty, we may often omit it; however, in this figure we make this
-distinction explicit. Further consider the leaves of the tree. First, the
-leaves involving the constructors are nonlinear terms, and thus they occur in
-the nonlinear empty context. Then we explicit cast from
-an nonlinear $\uparrow$-wrapped type into a linear one. Such casts may often be
-made implicitly, such as in $\cref{fig:kleeneabstractproof}$.
+In \cref{fig:kleenestarderivation}, we demonstrate that
+\stringquote{ab} matches $(a^{*} \otimes b) \oplus c$ through repeated
+application of the Kleene star constructors. In this proof tree we
+expand some syntactic details that are often left implicit. For
+instance, every linear context is parameterized by a nonlinear
+context, which is syntactically separated by a semicolon. If the
+nonlinear context is empty, we may often omit it; however, in this
+figure we make this distinction explicit. Further consider the leaves
+of the tree. First, the leaves involving the constructors are
+nonlinear terms, and thus they occur in the nonlinear empty
+context. Then we explicit cast from an nonlinear $\uparrow$-wrapped
+type into a linear one. Such casts may often be made implicitly, such
+as in $\cref{fig:kleeneabstractproof}$.
 
@@
 trees of the output grammar.
+We also see here why $\cons$ and $\nil$ should themselves be
+non-linear values: we should be able to use them in constructing multiple parse trees.
 
@@
 \paragraph{Non-deterministic Finite Automata}
-Automata play a critical
-role in defining parsers, as they often serve as an operationalized and
-executable presentation of a formal grammar. Therefore, it is crucial that we
-can represent them internal to \theoryabbv. As a final example, consider the
-representation of $(a^{*} \otimes b) \oplus c$ as a non-deterministic finite
-automaton (NFA), shown in \cref{fig:exampleNFA}.
+\max{A lot of passive voice and indirect language in these paragraphs ``Often one wants to...'' ``Our method of representing...is''.}
+Automata play a critical role in defining parsers, as they often serve
+as an operationalized and executable presentation of a formal
+grammar. As a final example, consider the representation of $(a^{*}
+\otimes b) \oplus c$ as a non-deterministic finite automaton (NFA),
+shown in \cref{fig:exampleNFA}.
 
@@
 however, $\mathsf{1to2}$  may transition states without reading a character. This is because the
-transition from state 1 to state 2 is labeled by the empty string $\epsilon$, so
+transition from state 1 to state 2 is labeled by the empty string $\varepsilon$, so
 transition consumes no input. In this case, when building a trace out of
@@
 
-    \path[->] (1) edge[above] node{$\epsilon$} (2)
+    \path[->] (1) edge[above] node{$\varepsilon$} (2)
               (1) edge[right] node{$c$} (3)
@@
 \paragraph{Unambiguity}
+
 Ordinarily, we think of a formal grammar as unambiguous when there is at most
@@
 
+\max{what's a terminal object and what's a monomorphism and what's a distributive coproduct}
 We say that $A$ is \emph{unambiguous} if the unique map $! : A \to \top$ to the
@@
   \item $A$ is an unambiguous grammar
-  \item Given a grammar $B$, all morphisms $B \to A$ are equal
-  \item $A \& A \cong A$
+  \item Given a grammar $B$, all morphisms $B \to A$ are equal\max{is this internal statement? What is a ``morphism''}
+  \item $A \& A \cong A$\max{what is $\cong$?}
 \end{itemize}
@@
 thus also a monomorphism. That is, $\stringg$ is unambiguous.
+\max{we should have a section of the previous section which gives the axioms, this one and the distributivity axiom}
 
@@
 that it generates. Two grammars are said to be \emph{weakly equivalent} by
-Chomsky if they
+Chomsky\max{you can drop the ``by Chomsky'' at this point} if they
 agree in their weak generative capacity, and likewise \emph{strongly equivalent}
@@
 
+\max{indirect language}
 If $f$ and $g$ form an isomorphism, then $A$ and $B$ defined to be \emph{strongly
@@
 
+\max{this paragraph isn't really saying anything}
 Regular expressions
@@
 
+\max{who cares what every discussion of regular expressions is accompanied by}
 Nearly every theoretical discussion of regular expressions is accompanied by a
@@
 \newcommand{\transitions}{\mathsf{transitions}}
-\newcommand{\epstransitions}{\epsilon\mathsf{transitions}}
+\newcommand{\epstransitions}{\varepsilon\mathsf{transitions}}
 \newcommand{\isAcc}{\mathsf{isAcc}}
@@
 \newcommand{\print}{\mathsf{print}}
-\newcommand{\epssrc}{\epsilon\mathsf{src}}
-\newcommand{\epsdst}{\epsilon\mathsf{dst}}
+\newcommand{\epssrc}{\varepsilon\mathsf{src}}
+\newcommand{\epsdst}{\varepsilon\mathsf{dst}}
 
@@
 \item $N.\transitions$ --- a finite set of labeled transitions
-\item $N.\labelt$ --- a mapping from $N.\transitions$ to the character labeling
-        the transition
-\item $N.\src$ --- a mapping from $N.\transitions$ to source state of a transition
-\item $N.\dst$ --- a mapping from $N.\transitions$ to destination state of a transition
-\item $N.\epstransitions$ --- a finite set of $\epsilon$-transitions
-\item $N.\epssrc$ --- a mapping from $N.\epstransitions$ to source state of a $\epsilon$-transition
-\item $N.\epsdst$ --- a mapping from $N.\epstransitions$ to destination state of a $\epsilon$-transition
+\item $N.\labelt$ --- a mapping from $N.\transitions$ to the character labeling the transition
+\item $N.\src$ and $N.\dst$ --- mappings from $N.\transitions$ to source and target states of a labeled transition
+\item $N.\epstransitions$ --- a finite set of $\varepsilon$-transitions
+\item $N.\epssrc$ and $N.\epsdst$ --- mappings from $N.\epstransitions$ to source and target states of an $\varepsilon$-transition
 \end{itemize}
@@
 Finally, if we had a trace beginning at the destination of an
-$\epsilon$-transition then we may use $\epscons$ to pull it back along the
-$\epsilon$-transition and construct a trace beginning at the source of the $\epsilon$-transition.
+$\varepsilon$-transition then we may use $\epscons$ to pull it back along the
+$\varepsilon$-transition and construct a trace beginning at the source of the $\varepsilon$-transition.
 
@@
 \end{align*}
-\caption{Traces through the NFA $N$ as an indexed inductive type}
+\caption{Traces through an NFA $N$ as an indexed inductive type}
 \label{fig:nfatrace}
@@
 
+\max{I've seen people use a checkmark or a cute little agda logo for this}
 \newcommand{\Agda}{Formalized in Agda}
@@
   the DFA $D$. Define the
-  states of $D$ to be the $\mathbb{P}_{\epsilon}(N.\mathsf{states})$ --- the type of $\epsilon$-closed subsets of
+  states of $D$ to be the $\mathbb{P}_{\varepsilon}(N.\mathsf{states})$ --- the type of $\varepsilon$-closed subsets of
   $N.\mathsf{states}$. A subset is accepting in $D$ if it contains an accepting state
   from $N$. Construct the initial state of $D$ as the
-  $\epsilon$-closure of $N.\mathsf{init}$. Lastly define the transtion function
+  $\varepsilon$-closure of $N.\mathsf{init}$. Lastly define the transtion function
   of $D$ to send the subset $X$ under the character $c$ to the
-  $\epsilon$-closure of all the states reachable from $X$ via a transition
+  $\varepsilon$-closure of all the states reachable from $X$ via a transition
   labelled with the character $c$.
@@
 to get traces in a parse of $c$ concatenated with a family of traces
-parametrized by $\epsilon$-closed subsets containing $s'$.
+parametrized by $\varepsilon$-closed subsets containing $s'$.
 . We instantiate this family at $\delta_{c}(X)$, as
 $s'$ is reachable from $s$ by a single transition labelled by $c$, so
-$s' \in \overline{\epsilon}\overline{c}{X} = D.\delta_{c}(X)$. Therefore we can
+$s' \in \overline{\varepsilon}\overline{c}{X} = D.\delta_{c}(X)$. Therefore we can
 linearly combine the parse of character $c$ and the trace out of $D.\delta_{c}(X)$
@@
 
-Lastly, if the NFA trace was an $\epsilon$-transition
-$s \overset{\epsilon}{\to} s'$ with $s \in X$, then the construction is quite similar. In this
+Lastly, if the NFA trace was an $\varepsilon$-transition
+$s \overset{\varepsilon}{\to} s'$ with $s \in X$, then the construction is quite similar. In this
 case, we again inductively call $f$ on traces out of $s'$ except we instantiate
 the choice of subset containing $s'$ with $X$ itself. We may do this as all
-states in $D$ are taken to be $\epsilon$-closed, therefore $s' \in X$ because
-$s$ is in $X$ and $\epsilon$-transitions to $s'$.
+states in $D$ are taken to be $\varepsilon$-closed, therefore $s' \in X$ because
+$s$ is in $X$ and $\varepsilon$-transitions to $s'$.
 
@@
 We define $g$ by induction on traces through $D$. First if the trace stops at
-some accepting $\epsilon$-closed subset $X$, then we may choose some state
+some accepting $\varepsilon$-closed subset $X$, then we may choose some state
 $s \in X$ such that $s$ is accepting and terminate the trace in $N$ at $s$.
@@
 there exists some $s' \in X$ such that $s'$ tranitions to $s$ by some number of
-$\epsilon$-transitions followed by a transition labeled by $c$.
+$\varepsilon$-transitions followed by a transition labeled by $c$.
 We order all
@@
 In this proof, we present a variant of Thompson's construction
-\cite{thompsonProgrammingTechniquesRegular1968} in which we induct on the
+\cite{thompsonProgrammingTechniquesRegular1968} in which we \max{I induct this sentence into the hall of shame for misuse of the word ``induct''}induct on the
 structure of $R$ to construct strongly equivalent NFAs. We will sketch this
@@
 those that were accepting in the subautomata. The initial state has an
-$\epsilon$-transition to each of the initial states of the subautomata. Once in
+$\varepsilon$-transition to each of the initial states of the subautomata. Once in
 $N_{1}.\init$, the downstream transitions are a copy of those in $N_{1}$,
@@
 $N_{1}$ and $N_{2}$, and we choose which copy to inhabit by deciding which
-$\epsilon$-transition out of the initial state to take. A schematic
+$\varepsilon$-transition out of the initial state to take. A schematic
 representation of this automaton is given in \cref{fig:disjunctionnfa}.
@@
 
-    \path[->] (init) edge[below] node{$\epsilon$} (init1)
-              (init) edge[above] node{$\epsilon$} (init2)
+    \path[->] (init) edge[below] node{$\varepsilon$} (init1)
+              (init) edge[above] node{$\varepsilon$} (init2)
               (init1) edge[above] node{} (dots1)
@@
 Moving up the Chomsky hierarchy from regular grammars, we arrive at context-free
-grammars (CFGs). In many ways, the smallest nontrivial CFG is the Dyck grammar
+grammars (CFGs). \max{In many ways?}In many ways, the smallest nontrivial CFG is the Dyck grammar
 $\Dyck$ --- the grammar of balanced parentheses. Moreover, the Chomsky-Sch\"utzenberger
@@
 %     \path[->] (1) edge[above] node{$b$} (2)
-%               (1) edge[below, bend right, left=0.3] node{$\epsilon$} (3)
+%               (1) edge[below, bend right, left=0.3] node{$\varepsilon$} (3)
 %               (2) edge[loop left] node{$a$} (2)
@@
 \item Every term $\lceil w \rceil \vdash p : G \otimes H$ is equal to $(q,r)$ for some $q,r$
-\item Every term $\lceil w \rceil \vdash p : \epsilon$ is equal to $()$
+\item Every term $\lceil w \rceil \vdash p : \varepsilon$ is equal to $()$
 \item Every term $\lceil w \rceil \vdash p : c$ is equal to $x:c$
@@
 
-\paragraph{Linearity with Dependent Types}
-\steven{I don't know if it worth discussing the different influences of
-  categorical model here? We take Neel's model and add to it as necessary}
+%%    - verified imperative implementation: Lambek logic ala separation
+%%      logic?
+\paragraph{Relation to Separation Logic}
+
+\theoryabbv is similar in spirit to separation logic. Semantically,
+they are closely related: linear types in \theoryabbv denote families
+of sets indexed by a monoid of strings, whereas separation logic
+formulae typically denote families of predicates indexed by an ordered
+partial commutative monoid of worlds. The monoidal structure in both
+cases is are instances of the category-theoretic notion of Day
+convolution monoidal structure. From a separation-logic perspective,
+our notion of memory is very primitive: a memory shape is just a
+string of characters and the state of the memory is never allowed to
+evolve.
+
+Syntactically, our type theory can be viewed as a kind of ``separation
+type theory'' in that we care not just about provability of formulae
+but the constructive content of our linear terms, as these act as
+parsers. The design of \theoryabbv is based on the $\textrm{LNL}_D$
+calculus which explicitly was designed to be a dependent
+type-theoretic version of separation logic. Many concepts of
+separation logic then have direct analogues in our system. The
+characters $c$ are analogous to the ``points-to'' formulae $l \mapsto
+x$ in separation logic. Our $\otimes,\lto,\tol$ are analogous to the
+separating conjunction and magic-wand. The analogue of the persistence
+modality $\square$ of separation logic is $\bigoplus_{\_:\uparrow A} I$,
+which ``zeros out'' all but the parses from the empty string. If we
+think of the portions of the string we are parsing as our notion of
+resource, then these make sense as the ``resource-free'' elements of
+the type $A$.
+
+This semantic connection to separation logic also suggests an avenue
+of future work: rather than working with a dependent type theory, we
+could instead make a program logic based on non-commutative separation
+logic for verifying imperative implementations of parsers. This could
+be accomplished by modifying an existing separation logic
+implementation or embedding the logic within \theoryabbv.
+
+%% \paragraph{Linearity with Dependent Types}
+%% \steven{I don't know if it worth discussing the different influences of
+%%   categorical model here? We take Neel's model and add to it as necessary}
 
@@
 
-%%    - verified imperative implementation: Lambek logic ala separation
-%%      logic?
-\paragraph{Verified Imperative Implementations}
-Lambek logic ala separation logic
-
-\paragraph{Applications to Type Checking}
-While parsing typically refers to the generation of semantic
-objects from string input, many tasks in programming can be
-viewed as parsing of objects with more structure, such as
-trees with binding structure or graphs. Fundamental to the
-frontend of many
-programming language implementations are type systems. In
-particular, \emph{type checking}
---- analogous to language recognition --- and \emph{typed
-  elaboration} --- analogous to parsing --- arise when
-producing a semantic object subject to some analysis. Just
-as our string grammars were given as functors from $\String$
-to $\Set$, we envision adapting the same philosophy
-to functors from a \emph{category of trees} to $\Set$ to craft a syntax
-that natively captures typed elaboration. This suggests an
-unusual sort of bunched type theory, where context extension
-no longer resembles concatenation of strings but instead
-takes on the form of tree constructors.
+\paragraph{Type Checking and Elaboration}
+Our focus in this work has been on the verification of parsers for
+grammars over strings, but because \theoryabbv allows for the
+definition of arbitrarily powerful grammars, the system could also be
+used for more sophisticated semantic analysis such as scope checking
+or type checking. Alternatively, we could more directly encode type
+type systems as linear types in a modified version of \theoryabbv
+whose semantics is not in families of sets indexed by \emph{strings}
+but families of sets indexed by \emph{abstract syntax trees}.
+
+%% While parsing typically refers to the generation of semantic
+%% objects from string input, many tasks in programming can be
+%% viewed as parsing of objects with more structure, such as
+%% trees with binding structure or graphs. Fundamental to the
+%% frontend of many
+%% programming language implementations are type systems. In
+%% particular, \emph{type checking}
+%% --- analogous to language recognition --- and \emph{typed
+%%   elaboration} --- analogous to parsing --- arise when
+%% producing a semantic object subject to some analysis. Just
+%% as our string grammars were given as functors from $\String$
+%% to $\Set$, we envision adapting the same philosophy
+%% to functors from a \emph{category of trees} to $\Set$ to craft a syntax
+%% that natively captures typed elaboration. This suggests an
+%% unusual sort of bunched type theory, where context extension
+%% no longer resembles concatenation of strings but instead
+%% takes on the form of tree constructors.
+
+% type system              ~~ Formal Grammar
+% typing derivation        ~~ parse tree
+% algorithmic type system  ~~ LL(1), LR(1) grammar
+% Uniqueness of derivations ~~ unambiguous grammar
+% type elaborator           ~~ semantic actions
 

### 2024-11-06 paper/paper.tex
rough draft of syntax section

@@
 \newcommand{\theoryabbv}{$\textrm{Lambek}^D$}
+\newcommand{\lnld}{$\textrm{LNL}_D$}
 
@@
 \newcommand{\isLinCtx}{\textrm{ lin. ctx.}}
+
+\newcommand{\quoteTy}[1]{\lceil{#1}\rceil}
+\newcommand{\unquoteTy}[1]{\lfloor{#1}\rfloor}
 
@@
 \newcommand{\app}[2]{#1 \, #2}
-\newcommand{\applto}[2]{#1 \mathop{{}^{\lto}} #2}
+\newcommand{\applto}[2]{#1 \, #2}
 \newcommand{\apptol}[2]{#1 \mathop{{}^{\tol}} #2}
@@
 
-\newcommand{\ctxwff}[1]{#1 \,\, \mathsf{ok}}
-\newcommand{\ctxwffjdg}[2]{#1 \vdash #2 \,\, \mathsf{type}}
-\newcommand{\linctxwff}[2]{#1 \vdash #2 \,\, \mathsf{ok}}
-\newcommand{\linctxwffjdg}[2]{#1 \vdash #2 \,\, \mathsf{linear}}
+\newcommand{\ctxwff}[1]{#1 \isCtx}
+\newcommand{\ctxwffjdg}[2]{#1 \vdash #2 \isTy}
+\newcommand{\linctxwff}[2]{#1 \vdash #2 \isLinCtx}
+\newcommand{\linctxwffjdg}[2]{#1 \vdash #2 \isLinTy}
 
@@
 
-\max{was this paragraph supposed to be deleted?}
-Next we present the full details of the syntax and typing of \theoryname.
-\theoryname has two kinds of types, linear types $A$ which represent formal
-grammars, and non-linear types $X$, which act like ordinary types or sets. Both
-linear and non-linear types are checked for well-formedness in a context
-$\Gamma$ containing \emph{only} non-linear variables $\Gamma = x:X,\ldots$.
-
-\begin{figure}
-  \[
-  \begin{array}{rrcl}
-    \textrm{non-linear contexts} & \Gamma & ::= & \cdot \pipe \Gamma,x:X \\
-    \textrm{non-linear types} & X , Y , Z & ::= & \mathsf{Nat} \pipe
-                                                  \mathsf{Int} \pipe
-                                                  \mathsf{Empty} \pipe
-                                                  \mathsf{Unit} \pipe
-                                                  \mathsf{SPF}\,X\pipe U \pipe L
-                                                  \pipe \ltonl A \pipe \sum\limits_{x:X} Y
-                                                  \pipe \prod\limits_{x:X} Y\pipe M =_A
-                                                  N \\
-    \textrm{linear contexts} & \Delta & ::= & \cdot \pipe \Delta , a:A\\
-    \textrm{linear types} & A , B , C & ::= &
-    c \pipe I \pipe A \otimes B \pipe \bigoplus\limits_{x:X} A \pipe \mu A_f M \pipe
-    A \lto B \pipe B \tol A \pipe \bigwith\limits_{x:X} A \pipe \equalizer {a}{f}{g}\\
-  \end{array}
-  \]
-  \caption{Overview of Types}
-\end{figure}
+%% OUTLINE
+%%
+%% 1. Overview of syntax: what are the judgments
+%% 2. Cover the linear types except for inductives
+%% 3. Cover non-linear types, mainly focus on the universes
+%% 4. Cover indexed inductive linear types
+%% 5. Cover the "axioms" we add
+
+\max{was this paragraph supposed to be deleted?}  Next we present the
+full details of the syntax and typing of \theoryabbv. The syntax of
+\theoryabbv is based on the dependent linear-non-linear (\lnld)
+calculus from \cite{dlnl}. As in \lnld, \theoryabbv includes both
+non-linear dependent types, as well as linear types, which are allowed
+to depend on the non-linear types, but not on other linear types.
+%
+The main point of departure from \lnld's design in that the linear
+typing is \emph{non-commutative}, i.e., that exchange is not an
+admissible structural rule. Furthermore, we add a general-purpose
+indexed inductive linear type connective, as well as an
+\emph{equalizer} type, which we will show allows us to perform
+inductive proofs of equalities between linear terms.
+%
+Finally, we add base types and axioms to the system specifically to
+model formal grammars and parsing.
+
+The formation rules for the judgments of \theoryabbv are shown in
+\Cref{fig:formation}. $\Gamma$ stand for non-linear contexts, $X,Y,Z$
+stand for non-linear types, $M,N$ stand for non-linear terms, these
+act as in an ordinary dependent type theory. $\Delta$ stands for
+linear contexts, $A,B,C$ for linear types and $e,f,g$ for linear
+terms. These contexts, types and terms are allowed to depend on an
+ambient non-linear context $\Gamma$, but note that linear types $A$
+cannot depend on any \emph{linear} variables in $\Delta$. We include
+definitional equality judgments for both kinds of type and term
+judgments as well. Additionally, we have a judgment $\Gamma\vdash X
+\isSmall$ which is used in the definition of universe types.
 
@@
   \begin{mathpar}
-    \boxed{\Gamma ; \Delta \vdash e : A}
-
+    \inferrule{~}{\ctxwff \Gamma}
+
+    \inferrule{\ctxwff \Gamma}{\ctxwffjdg \Gamma X}
+
+    \inferrule{\ctxwffjdg \Gamma X}{\Gamma \vdash X \isSmall}
+
+    \inferrule{\ctxwffjdg \Gamma X \and \ctxwffjdg \Gamma Y}{\ctxwffjdg \Gamma {X \equiv Y}}
+
+    \inferrule{\ctxwffjdg \Gamma X}{\Gamma \vdash M : X}
+
+    \inferrule{\Gamma \vdash M : X \and \Gamma \vdash N : X}{\Gamma \vdash M \equiv N : X}
+
+    \inferrule{\ctxwff \Gamma}{\linctxwff \Gamma \Delta}
+
+    \inferrule{\ctxwff \Gamma}{\Gamma \vdash A \isLinTy}
+
+    \inferrule{\Gamma \vdash A \isLinTy \and \Gamma \vdash B \isLinTy}{\Gamma \vdash A \equiv B}
+
+    \inferrule{\Gamma \vdash \Delta \isLinCtx \and \Gamma \vdash A \isLinTy}{\Gamma;\Delta\vdash e : A}
+
+    \inferrule{\Gamma; \Delta \vdash e : A \and \Gamma;\Delta \vdash f : A}{\Gamma;\Delta\vdash e \equiv f : A}
+  \end{mathpar}
+  \caption{Formation Rules}
+  \label{fig:formation}
+\end{figure}
+
+\subsection{Non-linear Typing}
+
+The non-linear types of \theoryabbv include the standard dependent
+types for $\Pi,\Sigma$, extensional equality, natural numbers,
+booleans, unit and empty types. We include these rules in the
+appendix. We present the other non-linear type constructions in
+\Cref{fig:non-linear-stuff}. First, we include universe types $U$ of
+small non-linear types and $L$ of linear types. These are defined as
+``universes ala Coquand'' in that we define a judgment saying when a
+non-linear type is small and define the universe to internalize
+precisely this judgment. These universe types are needed so that we
+can define types by recursion on natural numbers. Next, we include a
+non-linear type $\ltonl A$ where $A$ is a linear type. The intuition
+for this type is that its elements are the linear terms that are
+``resource free'': its introduction rule says we can construct an
+$\ltonl A$ when we have a linear term of type $A$ with no free linear
+variables. This type is used extensively in our examples, playing a
+similar role to the $!$ modality of ordinary linear logic.
+
+%% \begin{figure}
+%%   \[
+%%   \begin{array}{rrcl}
+%%     \textrm{non-linear contexts} & \Gamma & ::= & \cdot \pipe \Gamma,x:X \\
+%%     \textrm{non-linear types} & X , Y , Z & ::= & \mathsf{Nat} \pipe
+%%                                                   \mathsf{Bool} \pipe
+%%                                                   \mathsf{Empty} \pipe
+%%                                                   \mathsf{Unit} \pipe
+%%                                                   \mathsf{SPF}\,X\pipe U \pipe L
+%%                                                   \pipe \ltonl A \pipe \sum\limits_{x:X} Y
+%%                                                   \pipe \prod\limits_{x:X} Y\pipe M =_A
+%%                                                   N
+%%   \end{array}
+%%   \]
+%%   \caption{Non-linear Types}
+%% \end{figure}
+
+\begin{figure}
+  \begin{mathpar}
+    \inferrule{}{\Gamma \vdash U \isTy}\and
+    \inferrule
+    {\Gamma \vdash M : U}
+    {\Gamma \vdash \unquoteTy M \isTy}\and
+    \inferrule
+    {\Gamma \vdash X \isTy}
+    {\Gamma \vdash \quoteTy X : U}\and
+    \unquoteTy{\quoteTy{X}} \equiv X \and \inferrule{\Gamma \vdash M : U}{\Gamma\vdash \quoteTy{\unquoteTy{M}}\equiv M : U}
+
+    \inferrule{}{\Gamma \vdash L \isTy}\and
+    \inferrule
+    {\Gamma \vdash M : L}
+    {\Gamma \vdash \unquoteTy M \isLinTy}\and
+    \inferrule
+    {\Gamma \vdash A \isLinTy}
+    {\Gamma \vdash \quoteTy A : L}\and
+    \unquoteTy{\quoteTy{A}} \equiv A \and \inferrule{\Gamma \vdash M : U}{\Gamma\vdash \quoteTy{\unquoteTy{M}}\equiv M : U}
+
+    \inferrule
+    {\Gamma \vdash A \isLinTy}
+    {\Gamma \vdash \ltonl { A } \isTy}
+    \and
+    \inferrule{\Gamma ; \cdot \vdash e : A}{\Gamma \vdash e : \ltonl A}\and
+    \inferrule{\Gamma \vdash M : \ltonl A}{\Gamma ; \cdot \vdash M : A} \and
+  \end{mathpar}
+  \caption{Non-linear types (selection)}
+  \label{fig:non-linear-types}
+\end{figure}
+
+\subsection{Linear Typing}
+
+We give an overview of the linear types in \Cref{fig:linear-types} and
+terms in \Cref{fig:linear-terms}. The equational theory for these
+types is straightforward and included in the appendix. First, the
+linear variable rule says that a linear variable can be used if it is
+the \emph{only} variable in the context.
+
+First, we cover the ``multiplicative'' connectives of non-commutative
+linear logic. The linear unit and tensor product are standard for a
+non-commutative linear logic: when we construct a linear unit we
+cannot use any variables and when we construct a tensor product, the
+two sides must use disjoint variables, and the variables the left side
+of the product uses must be to the left in the context of the
+variables used by the right side of the tensor product. Because we are
+non-commutative, there are two function types: $A \lto B$ and $B \tol
+A$. The difference between these is that the introduction rule for $A
+\lto B$ adds a variable to the right side of the context, whereas the
+introduction rule for $B \tol A$ adds a variable to the left side of
+the context. In our experience, because by convention parsing
+algorithms parse from left-to-right, we rarely need to use the $B \tol
+A$ connective. One of the main uses of linear types is that, in
+combination with the $\uparrow$ connective, we can abstract over
+linear terms, in that the non-linear type $\ltonl{A \lto B}$ is a
+non-linear type of linear terms with one linear input $A$ and output
+$B$.
+
+Next, we have the ``additive'' connectives. For these, we use the
+non-linear types to define \emph{indexed} versions of the additive
+disjunction $\oplus$ and additive conjunction $\&$ of linear
+logic. These are like a linear version of the $\Sigma$ and $\Pi$
+connectives of ordinary dependent type theory. The indexed $\&$ is
+defined by a $\lambda$ and eliminated using projection. The indexed
+$\oplus$ is like a \emph{weak} $\Sigma$ type: it has an injection
+introduction rule $\sigma$, but its elimination rule is given by
+\emph{pattern matching} rather than projections. Note that we can
+define the more typical nullary and binary versions of these
+connectives by using indexing over the empty and boolean type
+respectively. We will freely use $\bot$ to refer to this empty
+discunction and $\top$ to refer to the empty conjunction. Lastly, we
+include a type $\equalizer {a}{f}{g}$ that we call the
+\emph{equalizer} of $f$ and $g$. We think of this type as the
+``subset'' of $A$ that satisfy the equation $f\,a\equiv g\,a$. If
+these were non-linear types, such a type could be constructed using
+the equality type as $\sum_{x:X} f\,x=_Y g\,x$, but this construction
+can't be used because it uses a \emph{dependent} equality type. We
+note that the equalizer type is not used in defining any of our
+parsers or formal grammars, but instead will be used shortly to derive
+an induction principle for proving equalities of functions whose
+domain is an inductive type.
+
+\begin{figure}
+  \begin{mathpar}
+    \inferrule{}{\Gamma \vdash I \isLinTy}\and
+    \inferrule{\Gamma \vdash A \isLinTy \and \Gamma \vdash B \isLinTy}{\Gamma \vdash A \otimes B \isLinTy}\and
+    \inferrule{\Gamma \vdash A \isLinTy \and \Gamma \vdash B \isLinTy}{\Gamma \vdash A \lto B \isLinTy}\and
+    \inferrule{\Gamma \vdash A \isLinTy \and \Gamma \vdash B \isLinTy}{\Gamma \vdash A \tol B \isLinTy}\and
+    \inferrule{\Gamma,x:X \vdash A \isLinTy}{\Gamma \vdash \bigoplus\limits_{x:X} A}\and
+    \inferrule{\Gamma,x:X \vdash A \isLinTy}{\Gamma \vdash \bigamp\limits_{x:X} A}\and
+    \inferrule
+    {\Gamma \vdash f : \ltonl {(A \lto B)} \and \Gamma \vdash g : \ltonl { (A \lto B) }}
+    {\Gamma \vdash \equalizer {a}{f}{g} \isLinTy}
+  \end{mathpar}
+  \caption{Linear Types (selection)}
+\end{figure}
+
+\begin{figure}
+  \begin{mathpar}
     \inferrule{~}{\Gamma ; a : A \vdash a : A}
@@
     %
-    \inferrule{\Gamma ; a : A , \Delta \vdash e : B}{\Gamma ; \Delta \vdash \lambtol a e : B\tol A}
-    \and
-    \inferrule{\Gamma ; \Delta \vdash e : B \tol A \\ \Gamma ; \Delta' \vdash e' : A}{\Gamma ; \Delta', \Delta \vdash \apptol e {e'} : B}
-    %
-    \\
-    %
     \inferrule{\Gamma, x : X ; \Delta  \vdash e : A}
@@
     \inferrule{\Gamma ; \Delta \vdash e : \bigoplus\limits_{x:X} A \quad \Gamma, x : X ; \Delta'_1, a : A, \Delta'_2 \vdash e' : C}{\Gamma; \Delta'_1, \Delta, \Delta'_2 \vdash \letin {\sigma\,x\,a} e {e'}: C}
-    %
-    \and
-    %
-    \inferrule
-    {\Gamma \vdash M : \ltonl A}
-    {\Gamma ; \cdot \vdash M : A}
     %
@@
   \end{mathpar}
-  \caption{Linear typing}
+  \caption{Linear terms (selection)}
   \label{fig:linsyntax}
@@
 
+\subsection{Indexed Inductive Linear Types}
+
+Next, we introduce one of the most complex and important type linear
+type constructor of our development, \emph{indexed inductive linear
+types} $\mu F$. To encode these, we first add a non-linear type
+$\SPF\,X$ of \emph{strictly-positive functorial} linear type
+expressions indexed by a non-linear type $X$. We give the rules for
+strictly-positive functors in \Cref{fig:spf}. First, we add operations
+for a recursive reference $\Var{x}$ the constant functor, $K$, tensor
+products and additive conjunction and disjunction. As an example,
+\max{TODO: make an example}.  Next we add an operation $\el(F)$ which
+interprets our syntactic codes as an actual type constructor, and
+define it for each of the operations given above. Lastly, we give a
+\emph{functorial action} $\map(F)$ for each of the codes.\max{TODO:
+  find a citation for this kind of approach. Container papers work
+  this way for instance. I think there is a linear container paper
+  that uses this appraoch}
+
+With this notion of strictly-positive functor so defined, in
+\Cref{fig:iilt} we next add a linear type constructor $\mu F$, which
+defines a family of mutually-recursive inductive types for the family
+of type expressions $F$. The introduction rule for this is
+$\mathsf{roll}$, which constructs an element of $\mu F\,x$ from its
+one-level unfolding. The elimination principle is defined by a
+$\mathsf{fold}$ operation. The equations ensure that this makes the
+family an \emph{initial algebra} for the type expressions. That is the
+$\beta$ rule says that a $\mathsf{fold}$ applied to a $\mathsf{roll}$
+is equivalent to $\mathsf{map}$ping the $\mathsf{fold}$ over all the
+sub-expressions, which means that $\mathsf{fold}$ is an algebra
+homomorphism from $\mathsf{roll}$ to $\mathsf{f}$. Then the $\eta$
+rule says that $\mathsf{fold}$ is the \emph{unique} such homomorphism.
+
+This definition as an initial algebra is well-understood semantically
+but the $\eta$ principle in particular is somewhat cumbersome to use
+directly in proofs. In dependent type theory, we would have a
+dependent \emph{elimination} principle, which justifies
+\emph{inductive proofs}. However since linear types do not support
+dependency on linear types, we cannot directly adapt this
+approach. However, if we are trying to prove that two morphisms out of
+a mutually recursive type are equal, we can use the \emph{equalizer}
+type to prove their equality by induction.\max{TODO: show how this is
+  done.}
+
+\begin{figure}
+  \begin{align*}
+    \mathsf{Var} &: X \to \SPF\,X\\
+    \mathsf{K} &: L \to \SPF\,X\\
+    \mathsf{\bigoplus} &: (Y \to \SPF\,X) \to \SPF\,X\\
+    \mathsf{\bigamp} &: (Y \to \SPF\,X) \to \SPF\,X\\
+    \mathsf{\otimes} &: \SPF\,X \to \SPF\,X \to \SPF\,X\\
+    \el &: \prod_{X:U}\SPF\,X \to (X \to L) \to L\\
+    \map &: \prod_{X:U}\prod_{F : \SPF\,X}\prod_{A,B:X\to L}(\prod_{x:X}\uparrow(A\,x\lto B\,x)) \to \uparrow(\el(F)(A) \lto \el(F)(B))\\\\
+    \el(\Var\,M) B &= B\,M\\
+    \el(\mathsf{K}\,A) B &= A\\
+    \el(\bigoplus A) B &= \bigoplus_{y:Y}\el(A y)B\\
+    \el(\bigamp A) B &= \bigamp_{y:Y}\el(A y)B\\
+    \el(A \otimes A') B &= \el(A)B \otimes \el(A')B\\\\
+    \map(\Var\,M)\,f &= f\,M\\
+    \map(\K\,A)\,f &= \lambda a. a\\
+    \map(\bigoplus A)\,f &= \lambda a. \letin{\oplusinj y {a_y}}{a}{\oplusinj y {\map(A\,y)\,f\,a_y}}\\
+    \map(\bigwith\,A)\,f &= \lambda a. \withlamb{y}{\map(A\,y)\,f\,(\withprj y a)}\\
+    \map(A \otimes A')\,f &= \lambda b. \letin{(a,a')}{b}{(\map(A)\,f\,a,\map(A')\,f\,a')}
+  \end{align*}
+  \begin{mathpar}
+  \end{mathpar}
+  \caption{Strictly Positive Functors}
+  \label{fig:spf}
+\end{figure}
+
 \begin{figure}
   \begin{mathpar}
-    \inferrule{~}{\Gamma \vdash \Set_0 : \Set_1}
- %
-    \and
-%
-    \inferrule{~}{\Gamma \vdash L_i : U_{i+1}}
-    {}{}
-  \end{mathpar}
-  \caption{Universe Types}
-\end{figure}
-
-\begin{figure}
-  \textrm{TODO: var, konstant, bigoplus, bigamp and tensor}\\
-  \textrm{TODO: behavior of map}
-  \begin{mathpar}
-    \inferrule
-    {\Gamma \vdash M : X}
-    {\Gamma \vdash \mathsf{Var}\,M : \SPF\,X}
-
-    \inferrule
-    {\Gamma \vdash A : L}
-    {\Gamma \vdash \mathsf{K}\,A : \SPF\,X}
-
-    \inferrule
-    {\Gamma,y:Y \vdash A : \SPF\,X}
-    {\Gamma \vdash \bigoplus_{y:Y}\,A : \SPF\,X}
-
-    \inferrule
-    {\Gamma,y:Y \vdash A : \SPF\,X}
-    {\Gamma \vdash \bigamp_{y:Y}\,A : \SPF\,X}
-
-    \inferrule
-    {\Gamma \vdash A : \SPF\,X\and \Gamma \vdash B : \SPF\,X}
-    {\Gamma \vdash A \otimes B : \SPF\,X}
-
-    \inferrule
-    {A : \SPF\,X}
-    {\el(A) : (X \to L) \to L}
-
-    \el(\Var\,M) B = B\,M\and
-    \el(\mathsf{K}\,A) B = A\and
-    \el(\bigoplus_{y:Y} A) B = \bigoplus_{y:Y}\el(A)B\and
-    \el(\bigamp_{y:Y} A) B = \bigamp_{y:Y}\el(A)B\and
-    \el(A \otimes A') B = \el(A)B \otimes \el(A')B
-
-  \inferrule
-  {A : \textrm{SPF} X\and
-  \Gamma,x:X;b: B x \vdash f : C x\and
-  \Delta \vdash e : \el(A)(B)
-  }
-  {\Gamma;\Delta \vdash \mathsf{map}\,A\,(x.b.f)\,e : \textrm{el}(A)\, C}
-
-  \map(\Var\,M)\,(x.b.f)\,e = f[M/x][e/b]\and
-  \map(\K\,A)\,(x.b.f)\,e = e\and
-  \map(\bigoplus\limits_{y:Y} A)\,(x.b.f)\,e =
-  \letin{\oplusinj y a}{e}{\oplusinj y {\map(A)\,(x.b.f)\,e}}\\
-  \map(\bigwith\limits_{y:Y} A)\,(x.b.f)\,e =
-  \withlamb{y}{\map(A)\,(x.b.f)\,(\withprj a e)}\\
-  \map(A \otimes A')\,(x.b.f)\,e =
-  \letin{(a,a')}{e}{(\map(A)\,(x.b.f)\,a,\map(A')\,(x.b.f)\,a')}
-  \end{mathpar}
-  \caption{Strictly Positive Functors}
-\end{figure}
-
-\begin{figure}
-  \begin{mathpar}
-    \inferrule*[right=IndIntro]
-    {\Gamma \vdash A : X \to \textrm{SPF}\,X\and
-     \Gamma \vdash M : X\and
-     \Gamma; \Delta \vdash e : \textrm{el}(A\,M) (\mu\,A)}
-    {\Gamma; \Delta \vdash \mathsf{roll}\, e : \mu\,A\, M}
-
-    \inferrule*[right=IndElim]
-    {\Gamma;\Delta \vdash e : \mu A M\and
-     \Gamma,x:X;b:\el(A\,x)\,B \vdash f : B\,x
-    }
-    {\Gamma; \Delta \vdash \mathsf{fold}(A)(x.b.f)(M, e) : B\,M}
+    \mathsf{roll} : \prod_{X:U}\prod_{F:X \to \SPF\,X}\prod_{x:X}\ltonl{(\el(F\,x)(\mu\,F))}
+
+    %% \inferrule*[right=IndIntro]
+    %% {\Gamma \vdash A : X \to \textrm{SPF}\,X\and
+    %%  \Gamma \vdash M : X\and
+    %%  \Gamma; \Delta \vdash e : \textrm{el}(A\,M) (\mu\,A)}
+    %% {\Gamma; \Delta \vdash \mathsf{roll}\, e : \mu\,A\, M}
+
+    \mathsf{fold} : \prod_{X:U}\prod_{F:X\to\SPF\,X}\prod_{A:X \to L}
+    (\prod_{x:X}\ltonl{(\el(F\,x)(A) \lto A\,x)})
+    \to \prod_{x:X}\ltonl{(\mu F\,x \lto A\,x)}
 
@@
     {}
-    {\fold(A)(x.b.f)(M, \roll(e)) =
-      f[M/x][\map(A\,M)\,(y.a.\fold(A)(x.b.f)(y,a))\,e/b]}
+    {\fold\,F\,f\,x\,(\roll e) \equiv f\,x\,(\map(F\,x)\,(\fold\,F\,f))}
 
     \inferrule*[right=Ind$\eta$]
-    {\Gamma,x:X; a:\mu\,A\,x \vdash e' : B\,x
-      \and \Gamma;\Delta \vdash e : \mu A M\\\\
-      \Gamma,x:X; a':\el(A x)(\mu A) \vdash
-      e'[x/x][\roll(a')/a] = f[x/x][\map(A\,x)\,(y.a.e'[y/x][a/a])\,a/b] : B\,x
+    {x:X;a:\el(F\,x)(\mu F) \vdash e'\,x\,(\roll a) \equiv f\,x\,(\map(F\,x)\,e')
     }
-    {\Gamma; \Delta \vdash \fold(A)(x.b.f)(M,e) = e'[M/x][e/a] : B\,M}
+    {\fold\,F\,f\equiv e'}
+    %% {\Gamma,x:X; a:\mu\,A\,x \vdash e' : B\,x
+    %%   \and \Gamma;\Delta \vdash e : \mu A M\\\\
+    %%   \Gamma,x:X; a':\el(A x)(\mu A) \vdash
+    %%   e'[x/x][\roll(a')/a] = f[x/x][\map(A\,x)\,(y.a.e'[y/x][a/a])\,a/b] : B\,x
+    %% }
+    %% {\Gamma; \Delta \vdash \fold(A)(x.b.f)(M,e) = e'[M/x][e/a] : B\,M}
   \end{mathpar}
   \caption{Indexed Inductive Linear Types}
+  \label{fig:iilt}
 \end{figure}
+
+\subsection{Grammar-specific Additions}
+
+So far, our calculus is a generic combination of dependent types with
+non-commutative linear types. In order to carry out formal grammar
+theory and define parsers, however, we add a few carefully chosen base
+types, terms and axioms that allow us to reason about formal grammars.
+\max{TODO: more prose, also are these all of our axioms?}
+\begin{enumerate}
+\item (Characters) For each character of our alphabet $c \in \Sigma$, we add a linear type $c \isLinTy$.
+\item (Strings are total) The function $\ltonl{(\String \lto \top)}$ is an isomorphism.\max{for this need to define $\String$, and isomorphism, and possibly we should define $\top$ in more detail earlier?
+idk}
+\item (Distributivity of $\oplus,\&$) The function  $\ltonl{(\bigoplus\limits_{f : \prod_{x:X} Y(x)}\bigamp\limits_{x:X}A\,x\,(f\,x) \lto \bigamp\limits_{x:X}\bigoplus\limits_{y:Y(x)}A\,x\,y)}$ is an isomorphism.
+\item ($\ltonl$ preserves $\oplus$) The function $\ltonl{(\sum_{x:X}\ltonl{(A\,x)} \lto \ltonl{(\bigoplus\limits_{x:X}A\,x)})}$ is an isomorphism
+\end{enumerate}
 
@@
 %
-
-In this section we go on a tour of \theoryabbv. We start by presenting its syntax.
-In order to avoid a notion of raw-term, we present it in an intrinsically-typed fashion,
-where only well-typed terms are meaningful\max{this sentence says the same thing multiple times}. Next, we show how it admits a model in $\Set^{\String}$
-where the linear types have natural interpretation in terms of formal grammars --- for instance,
-the tensor corresponds to grammar concatenation. We conclude the section by showing how
-classes of grammars correspond to certain sub-type-theories inside \theoryabbv.\max{why are you putting the entire paper into this section?}
-
-\subsection{Syntax}
-\label{subsec:syntax}
-
-Below we describe \theoryname~(\theoryabbv), an instance of an LNL theory with dependent types
-to serve as a deductive
-setting for formal grammar theory. \theoryabbv axiomatizes the necessary
-structure underlying $\Grammar$ to
-specify and construct parsers.
-
-The structural judgments are shown in \cref{fig:structjdg}, the typing
-well-formedness in \cref{fig:typewf}, and the intuitionistic typing
-rules in \cref{fig:inttyping}. These are mostly just as they appear in
-\cite{krishnaswami_integrating_2015}.
-It has two main differences\max{you absolutely cannot just describe the differences between our calculus and one from prior work.
-This paper needs to be self-contained}.
-The
-first one it the presence of two-distinct arrow types, one for adding
-variables to the beginning of contexts and another one for adding
-variables to the end of contexts. This is an adequate\max{adequate?} change in the
-context of grammars, which are inherently non-symmetric. The second
-change is the introduction of inductive types, which allows for the definition
-of recursive grammars. The treatment of the non-linear
-types is standard, so below we focus on the linear syntax.
-
-\begin{figure}
-  \begin{mathpar}
-    \boxed{\ctxwff \Gamma}
-
-    \inferrule{~}{\ctxwff \cdot}
-    \and
-    \inferrule{\ctxwff \Gamma \\ \ctxwffjdg \Gamma X}{\ctxwff {\Gamma, x : X}}
-
-    \\
-
-    \boxed{\linctxwff \Gamma \Delta}
-
-    \inferrule{~}{\linctxwff \Gamma \cdot}
-    \and
-    \inferrule{\linctxwff \Gamma \Delta \\ \linctxwffjdg \gamma A}{\linctxwff \Gamma {\Delta, a : A}}
-
-    \\
-
-    \boxed{\ctxwffjdg \Gamma X}
-
-    \inferrule{\Gamma \vdash X : U_i}{\ctxwffjdg \Gamma X}
-
-    \boxed{\linctxwffjdg \Gamma A}
-
-    \inferrule{\Gamma \vdash A : L_i}{\linctxwffjdg \Gamma A}
-
-    \\
-
-    \boxed{\ctxwffjdg \Gamma {X \equiv Y}}
-
-    \inferrule{\Gamma \vdash X \equiv Y : U_i}{\ctxwffjdg \Gamma {X\equiv Y}}
-
-    \boxed{\linctxwffjdg \Gamma {A \equiv B}}
-
-    \inferrule{\Gamma \vdash A \equiv B : L_i}{\linctxwffjdg \Gamma {A \equiv B}}
-
-  \end{mathpar}
-  \caption{Structural judgments}
-  \label{fig:structjdg}
-\end{figure}
-
-\begin{figure}
-  \begin{mathpar}
-    \boxed{\Gamma \vdash X : U_{i}}
-
-    \boxed{\Gamma \vdash A : L_{i}}
-
-    \inferrule{~}{\Gamma \vdash U_i : U_{i+1}}
- %
-    \and
-%
-    \inferrule{~}{\Gamma \vdash L_i : U_{i+1}}
-%
-    \and
-%
-    \inferrule{\Gamma \vdash X : U_i \\ \hspace{-0.1cm} \Gamma, x : X \vdash Y : U_i}{\Gamma \vdash \PiTy x X Y : U_i }%
-%
-    \and
-%
-    \inferrule{\Gamma\vdash X : U_i \\ \hspace{-0.1cm} \Gamma, x : X \vdash Y : U_i}{\Gamma \vdash \SigTy x X Y : U_i}
-%
-    \and
-%
-    \inferrule{~}{\Gamma \vdash 1 : U_i}
-%
-    % \and
-%
-    % \inferrule{\Gamma \vdash A : L_i}{\Gamma \vdash G A : U_i}
-
-    \and
-%
-    \inferrule{~}{\Gamma \vdash I : L_i}
- %
-    \and
-%
-    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash A \otimes B : L_i}
-%
-    \and
-%
-    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash A \lto B : L_i}
-%
-    \and
-%
-    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash B \tol A : L_i}
-%
-    \and
-%
-    \inferrule{\Gamma \vdash X : U_i \\ \Gamma, x : X \vdash A : L_i}{\Gamma \vdash \LinPiTy x X A : L_i}
-%
-    \and
-%
-    \inferrule{\Gamma \vdash X : U_i \\ \Gamma, x : X \vdash A : L_i}{\Gamma \vdash \LinSigTy x X A : L_i}
-%
-    \and
-%
-    \inferrule{\Gamma \vdash X : U_i \quad \{\Gamma \vdash e_i : X\}_i}{\Gamma \vdash e_1 =_X e_2 : U_i}
-    %
-    \and
-    %
-    \inferrule{~}{\Gamma \vdash \top : L_i}
-%
-    \and
-
-    \inferrule{~}{\Gamma \vdash \bot : L_i}
-
-    \and
-    \inferrule{c \in \Sigma}{\Gamma \vdash c : L_0}
-    %
-    \and
-    %
-    \inferrule{\Gamma, x : L_i \vdash A : L_i \and A \textrm{ strictly positive}}{\Gamma \vdash \mu x.\, A : L_i}
-    %
-    \and
-    %
-    \inferrule
-    {\Gamma \vdash A : L_i}
-    {\Gamma \vdash \ltonl { A } : U_i}
-    %
-    \and
-    \inferrule
-    {\Gamma ; \Delta \vdash f : \ltonl {(A \lto B)} \\ \Gamma \vdash g : \ltonl { (A \lto B) }}
-    {\Gamma ; \Delta \vdash \equalizer {a}{f}{g} : L_i}
-  \end{mathpar}
-  \caption{Type well-formedness}
-  \label{fig:typewf}
-\end{figure}
-
-\begin{figure}
-  \begin{mathpar}
-  \boxed{\Gamma \vdash M : X}
-
-  \inferrule{~}{\Gamma, x : X, \Gamma' \vdash x : X}
-  %
-  \and
-  %
-  \inferrule{\Gamma \vdash M : Y \quad \ctxwffjdg \Gamma {X \equiv Y}}{\Gamma \vdash M : X}
-  %
-  \and
-  %
-  \inferrule{~}{\Gamma \vdash () : 1}
-  %
-  \and
-  %
-  \inferrule{\Gamma \vdash M : X \\ \Gamma \vdash N : \subst Y M x}{\Gamma \vdash (M, N) : \SigTy x X Y}
-  %
-  \and
-%
-  \inferrule{\Gamma \vdash M : \SigTy x X Y}{\Gamma \vdash \pi_1\, M : X}
-  %
-  \and
-  %
-  \inferrule{\Gamma \vdash M : \SigTy x X Y}{\Gamma \vdash \pi_2\, M : \subst Y {\pi_1\, M} x}
-  \and
-  \inferrule{\Gamma, x : X \vdash M : Y}{\Gamma \vdash \lamb x M : \PiTy x X Y}
-  %
-  \and
-  %
-  \inferrule{\Gamma \vdash M : \PiTy x X Y \\ \Gamma \vdash N : X}{\Gamma \vdash \app M {N} : \subst Y {N} x}
-  %
-  \and
-  %
-  \inferrule{\Gamma \vdash M \equiv N : X}{\Gamma \vdash \mathsf{refl} : M =_X N}
-  \and
-  \inferrule{\Gamma ; \cdot \vdash e : A}{\Gamma \vdash e : \ltonl A}
-
-  \textrm{TODO: inductive types (boolean, nat) with their eliminators allowed in }
-  \end{mathpar}
-  \caption{Intuitionistic typing}
-  \label{fig:inttyping}
-\end{figure}
-
-\begin{figure}
-  \begin{mathpar}
-    \boxed{\Gamma \vdash M \equiv N : X}
-
-    \inferrule{\Gamma \vdash p : M =_X N}{\Gamma \vdash M \equiv N : X}
-%
-    \and
-%
-    \inferrule{~}{\Gamma \vdash \app {(\lamb x M)} {N} \equiv \subst x M {N} : X}
-%
-    \and
-%
-    \inferrule{~}{\Gamma \vdash m \equiv \lamb x {\app M x} : \PiTy x X Y}
-%
-    \and
-%
-    \inferrule{~}{\Gamma \vdash \pi_1\, (M, N) \equiv M : X}
-%
-    \and
-%
-    \inferrule{~}{\Gamma \vdash \pi_2\, (M , N) \equiv N : \subst x {M} Y}
-%
-    \and
-%
-    \inferrule{~}{\Gamma \vdash M \equiv (\pi_1\, M, \pi_2\, M) : \SigTy x X Y}
-%
-    \and
-%
-    % \inferrule{~}{}
-%
-    \inferrule{~}{\Gamma \vdash M \equiv N : 1}
-%
-    % \and
-%
-    % \inferrule{~}{\Gamma \vdash G\, (G^{-1} \, t) \equiv t : G A}
-
-    \\
-
-    \boxed{\Gamma ; \Delta \vdash e \equiv e' : A}
-
-    % \inferrule{~}{\Gamma; \cdot \vdash G^{-1}\, (G \, t ) \equiv t: A}
-%
-%
-    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lamblto a e)} {e'} \equiv \subst e {e'} a : C}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lamblto a {\app e a} : A \lto B}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lambtol a e)} {e'} \equiv \subst e {e'} a : C}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lambtol a {\app e a} : B \tol A}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash \app {(\dlamb x e)} {M} \equiv \subst {e} M x : C}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \dlamb x {\app e x} : \LinPiTy x X A}
-%
-    \and
-%
-%     \inferrule{~}{\Gamma; \Delta \vdash e \equiv e' : \top}
-% %
-%     \and
-%
-%     \inferrule{~}{\Gamma; \Delta \vdash e_i \equiv \pi_i (e_1, e_2) : A_i}
-% %
-%     \and
-% %
-%     \inferrule{~}{\Gamma; \Delta \vdash e \equiv (\pi_1 e, \pi_2 e) : A\& B}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} {()} e \equiv e : C}
-    %
-    \and
-    %
-    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} e {\subst {e'} {()} a} \equiv \subst {e'} e a : C}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash \letin {a \otimes b} {e \otimes e'} e'' \equiv e'' \{ e/a , e'/b \} : C}
-%
-    \and
-    %
-    \inferrule{~}{\Gamma; \Delta \vdash \letin {a \otimes b} e {\subst {e'} {a \otimes b} c} \equiv \subst {e'} e c : C}
-%
-    \and
-%
-    \inferrule{~}{\Gamma;\Delta \vdash \letin {(x, a)} {(M, e)} {e'} \equiv e' \{ M/x , e/a \} : C}
-    %
-    \and
-    %
-    \inferrule{~}{\Gamma; \Delta \vdash \letin {(x, a)} e {\subst {e'} {(x, a)} y} \equiv \subst {e'} e y : C}
-    \and
-    %
-    \inferrule
-    {~}
-    {\Gamma ; \Delta \vdash \equalizerpi {\equalizerin {e}} \equiv e : A}
-    %
-    \and
-    %
-    \inferrule
-    {~}
-    {\Gamma ; \Delta \vdash \equalizerin {\equalizerpi {e}} \equiv e : A}
-\end{mathpar}
-  \caption{Judgmental equality}
-  \label{fig:jdgeq}
-\end{figure}
-\steven{Do we need inverse for $\ltonl A$ in judgemental equality figure?}
-\steven{Do we need to say anything about the equalizer proofs being unique in
-  judgemental equality figure?}
 
@@
 
+First, we need to establish a precise definition of a formal grammar,
+that accounts for the possibility of ambiguity. The most common
+definition of a formal grammar is to use the Chomskyan formalism of
+generative grammars where a grammar is given by a set of
+non-terminals, with a specified start symbol and set of production
+rules. We instead use a more abstract formulation that is closer in
+spirit to the standard definition of a formal language:
+
 \begin{definition}
@@
 \end{definition}
+
+We think of the grammar $A$ as taking a string to the set of all parse
+trees for that string. However since $A$ could be any function
+whatsoever there is no requirement that an element of $A(w)$ be a
+``tree'' in the usual sense. This definition provides a simple,
+syntax-independent definition of a grammar that can be used for any
+formalism: generative grammars, categorial grammars, or our own
+type-theoretic grammars. Note that the definition of a formal grammar
+is a generalization of the usual notion of formal language since a
+proposition can be equivalently defined as a subset of a one-element
+set. Then the difference between a formal grammar and a formal
+language is that formal grammars can be \emph{ambiguous} in that there
+can be more than one parse of the same string. Even for unambiguous
+grammars, we care not just about \emph{whether} a string has a parse
+tree, but \emph{which} parse tree it has, i.e., what the structure of
+the element of $A(w)$ is.
+
+The linear types of \theoryabbv can all be interpreted as formal
+grammars, but what about the linear terms? These can be interpreted as
+\emph{parse transformers}
+\begin{definition}
+  Let $L_1,L_2$ be formal languages. Then we say $L_1\subseteq L_2$ if
+  for every string $w$, $L_1(w)\Rightarrow L_2(w)$.
+
+  Let $A_1,A_2$ be formal grammars. Then a parse transformer $f$ from
+  $A_1$ to $A_2$ is a function assigning to each string $w$ a function
+  $f_w : A_1(w) \to A_2(w)$.
+\end{definition}
+First, note that parse transformers generalize language inclusion: if
+$A_1(w), A_2(w)$ are all subsets of a one-element set, then a parse
+transformer is equivalent to showing that $A_1(w) \subseteq A_2(w)$.
+In our denotational semantics, linear terms will be interpreted as
+such parse transformers, and the notions of unambiguous grammar,
+parseability, etc, introduced in \Cref{sec:concepts} can be verified
+to correspond to their intended meanings under this interpretation.
 
@@
 
-For reasons of brevity, we present only a selection of cases here. The full details are in the appendix.\max{todo: put the full details in the appendix}.
-
-Most of the non-linear semantics is standard \pedro{We should either cite something or add the non-linear semantics in the appendix}, so let us investigate in more
-depth the linear semantics.
-\begin{enumerate}
-  \item A non-linear context $\Gamma$ denotes a set $\sem \Gamma$
-  \item A non-linear type $\Gamma \vdash X : U_{i}$ denotes a family of sets
-        $\sem X : \sem \Gamma \to \Set_{i}$
-  \item A non-linear term $\Gamma \vdash e : X$ denotes a dependently typed function
-        $\sem e : \prod_{\gamma : \sem \Gamma}\sem{X} \gamma$ \pedro{we have not defined what sections are}
-  \item Linear contexts $\Gamma \vdash \Delta$ and types
-        $\Gamma \vdash A : L_{i}$ both denote families of grammars
-        $\sem \Gamma \to \Gr_{i}$
-  \item A linear term $\Gamma ; \Delta \vdash M : A$ denotes a family of parse
-        transformers
-        $\sem M : \Pi(\gamma : \sem \Gamma)\Pi(w : \String) \sem \Delta \gamma w \Rightarrow \sem M \gamma w$
-\end{enumerate}
-
-\paragraph{Linear Unit}
-First, the linear unit $I$ may be built in the empty linear
-context\footnote{When appropriate, statements like ``the empty linear context''
-  may be taken polymorphically over all non-linear contexts that they might
-  depend on.} $\cdot \dashv I$. $I$ serves as the unit for the operation
-$\otimes$ described below.
-
-As a grammar,
-%
-\[
-  \sem {I} \gamma w = \{ () ~|~ w = \varepsilon \}
-\]
-%
-That is, $I$ maps strings to the proposition that they are the empty string. $I$
-is clearly only inhabited by the empty string $\varepsilon$, for which the
-outputted set contains a single parse tree.
-
-\paragraph{Base Types}
-For each character $c$ in the alphabet $\Sigma$, we include a base type at the
-lowest universe level $c : L_{0}$.
-
-The grammar interpretation for characters is quite similar to that of $I$.
-%
-\[
-  \sem {c} \gamma w = \{ () ~|~ w = c \}
-\]
-%
-The grammar for the character $c$ likewise maps strings $w$ to the proposition that
-they are equal to $c$.
-
-\paragraph{Tensor Product}
-The tensor product of two linear types is the first place where the ordering on
-contexts really takes effect. That is, the tensor product of two types is
-non-commutative.
-When context $\Delta$ forms a linear term of
-type $A$, and $\Delta'$ forms a linear term of type $B$, then the context
-extension $\Delta, \Delta'$ forms a linear term of type $A \otimes B$.
-
-As stated above, the type $I$ serves as the unit for $\otimes$. That is, for all
-linear types $A$, the following equalities hold:
-%
-\[ I \otimes A \equiv A \otimes I \equiv A \]
-%
-In the grammar semantics,
-%
-\[
-  \sem {A \otimes B} = \Sigma_{w_{1}, w_{2} : \String} (w_{1}w_{2} = w) \land \left(  \sem {A} \gamma w_{1} \times \sem {B} \gamma w_{2} \right)
-\]
-%
-The string $w$ matches $A \otimes B$
-precisely when it may be split into two
-pieces such that the left one matches $A$
-and the right one matches $B$.
-
-\paragraph{Linear Function Types}
-The monoidal structure provided by $\otimes$ is both left and right closed and
-this is denoted by the left and right linear function types. Further,
-for linear types $A$ and $B$ the left linear function type $A \lto B$
-enjoys an elimination principle similar to function application or modus ponens.
-%
-\[
-\inferrule{\Delta \vdash A \otimes A \lto B}{\Delta \vdash B}
-\]
-%
-The interpretation of linear function types at a string $w$ is a linear function
-that takes in parses of $A$ on a string $w_{a}$ and outputs parses of $B$ on the
-string $w_{a} w$.
-%
-\[
-  \sem{A \lto B} \gamma w = \Pi(w_a:\String) \left( A \gamma w_a \Rightarrow B\gamma (w_a w) \right)
-\]
-%
-That is, strings match $A \lto B$ if when prepended with a parse of $A$ they
-complete to parses of $B$. In this manner, the linear function types generalize
-Brzozowksi's notion of derivative
-\cite{brzozowskiDerivativesRegularExpressions1964}.
-Brzozowski initially only gave an accounting of this operation for
-generalized regular expressions, but later work from Might et al.\ demonstrates that the same
-construction can be generalized to context free grammars
-\cite{mightParsingDerivativesFunctional2011}. Here, via the linear function
-types, the same notion of derivative extends to the grammars of \theoryabbv.
-
-Note, to ensure that the linear function types do indeed generalize Brzozowski
-derivatives, we must include the equalities in \cref{fig:brzozowskideriv} as axioms\max{we must? why must we}.
-\pedro{We should further explain why these equations are required, or at the very
-least give some intuition}
-
-Of course, all the above statements for the left function type also have
-corresponding analogues for the
-right-handed counterpart.
+For reasons of brevity, we present only a selection of cases in
+\Cref{fig:semantics}. The full details are in the appendix.\max{todo:
+  put the full details in the appendix}.
+\max{TODO: how to describe $\mu$ succinctly?}\max{make sure $\lto$ and $\tol$ are right}
 
@@
   \begin{align*}
-  c\lto c &\equiv I\\
-  c\lto d &\equiv \bot\\
-  c\lto I &\equiv \bot\\
-  c\lto 0 &\equiv \bot\\
-  c\lto (A \otimes B) & \equiv (c\lto A) \otimes B + (A \amp I) \otimes (c\lto B)\\
-  c\lto A^* &\equiv (A \amp I)^{*} \otimes (c \lto A) \otimes A^* \\
-  c\lto (A + B) &\equiv (c\lto A) + (c\lto B)
+    \sem{c}\gamma\,w &= \{ c | w = c\}\\
+    \sem{I}\gamma\,w &= \{ () \pipe w = \varepsilon \}\\
+    \sem{A \otimes B}\gamma\,w &= \{ (w_1,w_2,a,b) \pipe w_1w_2 = w \wedge a \in \sem{A}\gamma\,w_1 \wedge b \in \sem{B}\gamma\,w_2 \}\\
+    \sem{\bigoplus_{x:X} A}\gamma\,w &= \{ (x, a) \pipe x \in \sem{X}\gamma \wedge a \in \sem{A}(\gamma,x)w \}\\
+    \sem{\bigwith_{x:X} A}\gamma\,w &= \prod_{x\in \sem{X}\gamma} \sem{A}(\gamma, x)\,w\\
+    \sem{A \lto B}\gamma\,w &= \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,ww'\\
+    \sem{B \tol A}\gamma\,w &= \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,w'w\\
+    \sem{\uparrow A}\gamma &= \sem{A}\gamma\,\varepsilon\\
+    \sem{\SPF\,X}\gamma &= \{ F : \textrm{Functor}(\textrm{Grammar}^{\sem{X}\gamma}, \textrm{Grammar}) \pipe F \textrm{ preserves $\omega$-colimits.} \}\\
+    \sem{\el(A)}\gamma\,G &= \sem{A}\gamma\,G\\
+    \sem{\mu A}\gamma
   \end{align*}
-\caption{Equality for Brzozowski Derivatives}
-\label{fig:brzozowskideriv}
+  \caption{Grammar Semantics}
 \end{figure}
+
+%% That is, strings match $A \lto B$ if when prepended with a parse of $A$ they
+%% complete to parses of $B$. In this manner, the linear function types generalize
+%% Brzozowksi's notion of derivative
+%% \cite{brzozowskiDerivativesRegularExpressions1964}.
+%% Brzozowski initially only gave an accounting of this operation for
+%% generalized regular expressions, but later work from Might et al.\ demonstrates that the same
+%% construction can be generalized to context free grammars
+%% \cite{mightParsingDerivativesFunctional2011}. Here, via the linear function
+%% types, the same notion of derivative extends to the grammars of \theoryabbv.
+
+%% Note, to ensure that the linear function types do indeed generalize Brzozowski
+%% derivatives, we must include the equalities in \cref{fig:brzozowskideriv} as axioms\max{we must? why must we}.
+%% \pedro{We should further explain why these equations are required, or at the very
+%% least give some intuition}
+
+%% Of course, all the above statements for the left function type also have
+%% corresponding analogues for the
+%% right-handed counterpart.
+
+%% \begin{figure}
+%% \begin{align*}
+%%   c\lto c &\equiv I\\
+%%   c\lto d &\equiv \bot\\
+%%   c\lto I &\equiv \bot\\
+%%   c\lto 0 &\equiv \bot\\
+%%   c\lto (A \otimes B) & \equiv (c\lto A) \otimes B + (A \amp I) \otimes (c\lto B)\\
+%%   c\lto A^* &\equiv (A \amp I)^{*} \otimes (c \lto A) \otimes A^* \\
+%%   c\lto (A + B) &\equiv (c\lto A) + (c\lto B)
+%% \end{align*}
+%% \caption{Equality for Brzozowski Derivatives}
+%% \label{fig:brzozowskideriv}
+%% \end{figure}
 
@@
 
+\appendix
+
+\section{Elided Syntax}
+
+\max{TODO: add prose}
+
+\begin{figure}
+  \begin{mathpar}
+    \inferrule{~}{\ctxwff \cdot}
+    \and
+    \inferrule{\ctxwff \Gamma \\ \ctxwffjdg \Gamma X}{\ctxwff {\Gamma, x : X}}
+    \inferrule{\Gamma \vdash X : U_i}{\ctxwffjdg \Gamma X}
+
+    \boxed{\linctxwffjdg \Gamma A}
+
+    \inferrule{~}{\linctxwff \Gamma \cdot}
+    \and
+    \inferrule{\linctxwff \Gamma \Delta \\ \linctxwffjdg \gamma A}{\linctxwff \Gamma {\Delta, a : A}}
+  \end{mathpar}
+  \caption{Context Rules}
+\end{figure}
+
+\max{TODO: add rules for non-linear types $\Pi,\Sigma$, equality etc}
+
+\begin{figure}
+  \begin{mathpar}
+  \boxed{\Gamma \vdash M : X}
+
+  \inferrule{~}{\Gamma, x : X, \Gamma' \vdash x : X}
+  %
+  \and
+  %
+  \inferrule{\Gamma \vdash M : Y \quad \ctxwffjdg \Gamma {X \equiv Y}}{\Gamma \vdash M : X}
+  %
+  \and
+  %
+  \inferrule{~}{\Gamma \vdash () : 1}
+  %
+  \and
+  %
+  \inferrule{\Gamma \vdash M : X \\ \Gamma \vdash N : \subst Y M x}{\Gamma \vdash (M, N) : \SigTy x X Y}
+  %
+  \and
+%
+  \inferrule{\Gamma \vdash M : \SigTy x X Y}{\Gamma \vdash \pi_1\, M : X}
+  %
+  \and
+  %
+  \inferrule{\Gamma \vdash M : \SigTy x X Y}{\Gamma \vdash \pi_2\, M : \subst Y {\pi_1\, M} x}
+  \and
+  \inferrule{\Gamma, x : X \vdash M : Y}{\Gamma \vdash \lamb x M : \PiTy x X Y}
+  %
+  \and
+  %
+  \inferrule{\Gamma \vdash M : \PiTy x X Y \\ \Gamma \vdash N : X}{\Gamma \vdash \app M {N} : \subst Y {N} x}
+  %
+  \and
+  %
+  \inferrule{\Gamma \vdash M \equiv N : X}{\Gamma \vdash \mathsf{refl} : M =_X N}
+  \and
+  \textrm{TODO: inductive types (boolean, nat) with their eliminators allowed in }
+  \end{mathpar}
+  \caption{Intuitionistic typing}
+  \label{fig:inttyping}
+\end{figure}
+
+\begin{figure}
+  \begin{mathpar}
+    %
+    \inferrule{\Gamma ; a : A , \Delta \vdash e : B}{\Gamma ; \Delta \vdash \lambtol a e : B\tol A}
+    \and
+    \inferrule{\Gamma ; \Delta \vdash e : B \tol A \\ \Gamma ; \Delta' \vdash e' : A}{\Gamma ; \Delta', \Delta \vdash \apptol e {e'} : B}
+    %
+    \\
+
+  \end{mathpar}
+  \caption{Linear Terms, extended}
+\end{figure}
+
+\begin{figure}
+  \begin{mathpar}
+    \boxed{\Gamma \vdash M \equiv N : X}
+
+    \inferrule{\Gamma \vdash p : M =_X N}{\Gamma \vdash M \equiv N : X}
+%
+    \and
+%
+    \inferrule{~}{\Gamma \vdash \app {(\lamb x M)} {N} \equiv \subst x M {N} : X}
+%
+    \and
+%
+    \inferrule{~}{\Gamma \vdash m \equiv \lamb x {\app M x} : \PiTy x X Y}
+%
+    \and
+%
+    \inferrule{~}{\Gamma \vdash \pi_1\, (M, N) \equiv M : X}
+%
+    \and
+%
+    \inferrule{~}{\Gamma \vdash \pi_2\, (M , N) \equiv N : \subst x {M} Y}
+%
+    \and
+%
+    \inferrule{~}{\Gamma \vdash M \equiv (\pi_1\, M, \pi_2\, M) : \SigTy x X Y}
+%
+    \and
+%
+    % \inferrule{~}{}
+%
+    \inferrule{~}{\Gamma \vdash M \equiv N : 1}
+%
+    % \and
+%
+    % \inferrule{~}{\Gamma \vdash G\, (G^{-1} \, t) \equiv t : G A}
+
+    \\
+
+    \boxed{\Gamma ; \Delta \vdash e \equiv e' : A}
+
+    % \inferrule{~}{\Gamma; \cdot \vdash G^{-1}\, (G \, t ) \equiv t: A}
+%
+%
+    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lamblto a e)} {e'} \equiv \subst e {e'} a : C}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lamblto a {\app e a} : A \lto B}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lambtol a e)} {e'} \equiv \subst e {e'} a : C}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lambtol a {\app e a} : B \tol A}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash \app {(\dlamb x e)} {M} \equiv \subst {e} M x : C}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \dlamb x {\app e x} : \LinPiTy x X A}
+%
+    \and
+%
+%     \inferrule{~}{\Gamma; \Delta \vdash e \equiv e' : \top}
+% %
+%     \and
+%
+%     \inferrule{~}{\Gamma; \Delta \vdash e_i \equiv \pi_i (e_1, e_2) : A_i}
+% %
+%     \and
+% %
+%     \inferrule{~}{\Gamma; \Delta \vdash e \equiv (\pi_1 e, \pi_2 e) : A\& B}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} {()} e \equiv e : C}
+    %
+    \and
+    %
+    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} e {\subst {e'} {()} a} \equiv \subst {e'} e a : C}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash \letin {a \otimes b} {e \otimes e'} e'' \equiv e'' \{ e/a , e'/b \} : C}
+%
+    \and
+    %
+    \inferrule{~}{\Gamma; \Delta \vdash \letin {a \otimes b} e {\subst {e'} {a \otimes b} c} \equiv \subst {e'} e c : C}
+%
+    \and
+%
+    \inferrule{~}{\Gamma;\Delta \vdash \letin {(x, a)} {(M, e)} {e'} \equiv e' \{ M/x , e/a \} : C}
+    %
+    \and
+    %
+    \inferrule{~}{\Gamma; \Delta \vdash \letin {(x, a)} e {\subst {e'} {(x, a)} y} \equiv \subst {e'} e y : C}
+    \and
+    %
+    \inferrule
+    {~}
+    {\Gamma ; \Delta \vdash \equalizerpi {\equalizerin {e}} \equiv e : A}
+    %
+    \and
+    %
+    \inferrule
+    {~}
+    {\Gamma ; \Delta \vdash \equalizerin {\equalizerpi {e}} \equiv e : A}
+\end{mathpar}
+  \caption{Judgmental equality for linear terms}
+  \label{fig:jdgeq}
+\end{figure}
+\steven{Do we need inverse for $\ltonl A$ in judgemental equality figure?}
+\steven{Do we need to say anything about the equalizer proofs being unique in
+  judgemental equality figure?}
+
 \end{document}

### 2024-11-07 paper/paper.tex
tweak
checkin semantics

@@
 \item (Distributivity of $\oplus,\&$) The function  $\ltonl{(\bigoplus\limits_{f : \prod_{x:X} Y(x)}\bigamp\limits_{x:X}A\,x\,(f\,x) \lto \bigamp\limits_{x:X}\bigoplus\limits_{y:Y(x)}A\,x\,y)}$ is an isomorphism.
-\item ($\ltonl$ preserves $\oplus$) The function $\ltonl{(\sum_{x:X}\ltonl{(A\,x)} \lto \ltonl{(\bigoplus\limits_{x:X}A\,x)})}$ is an isomorphism
+\item ($\ltonl$ preserves $\oplus$) The function $\sum_{x:X}\ltonl{(A\,x)} \to \ltonl{(\bigoplus\limits_{x:X}A\,x)}$ is an isomorphism.
 \end{enumerate}
@@
 
-To justify our intuitive explanation of linear types as formal
-grammars, we will now define a \emph{denotational semantics} that
-makes this mathematically precise. This semantics is the basis for our
-prototype implementation in Agda.
-
-\subsection{What is a formal grammar?}
-
-First, we need to establish a precise definition of a formal grammar,
+To justify our assertion that \theoryabbv is a syntax for formal
+grammars and parse transformers, we will now define a
+\emph{denotational semantics} that makes this mathematically
+precise. To interpret our types of codes we assume a universe of
+\emph{small} sets.
+
+\subsection{Formal Grammars and Parse Transformers}
+
+First, we need to establish a precise definition of a formal grammar
 that accounts for the possibility of ambiguity. The most common
-definition of a formal grammar is to use the Chomskyan formalism of
-generative grammars where a grammar is given by a set of
-non-terminals, with a specified start symbol and set of production
-rules. We instead use a more abstract formulation that is closer in
-spirit to the standard definition of a formal language:
+definition of a formal grammar is as a generative grammars, defined by
+a set of non-terminals, with a specified start symbol and set of
+production rules. We instead use a more abstract formulation that is
+closer in spirit to the standard definition of a formal language:
 
@@
 
-\max{todo: need to decide if we need this section here or if it's in the intro already}.
-
+Parse transformers have an obvious notion of composition and identity
+transformer. This defines a \emph{category}.
+\begin{definition}
+  Define $\Grammar$ to be the category whose objects are formal
+  grammars and morphisms are parse transformers. 
+\end{definition}
+This category is equivalent to the slice category $\Set/\Sigma^*$ and
+as such is very well-behaved. It is complete, co-complete, cartesian
+closed and carries a monoidal biclosed structure. We will use these
+structures to model the linear types, terms and equalities in
+\theoryabbv.
+%% \begin{theorem}
+%%   The category $\Grammar$ is complete, co-complete and cartesian
+%%   closed. Furthermore, it is monoidal biclosed with the monoidal
+%%   product given by
+%%   \[ (A \otimes B) w = \{ (w_1,w_2,a,b) \pipe w_1w_2 = w \wedge a \in A w_1 \wedge b \in B w_2\} \]
+%%   and unit by
+%%   \[ I w = \{ () \pipe w = \varepsilon \} \]
+%% \end{theorem}
+
+%% \max{todo: need to decide if we need this section here or if it's in the intro already}.
 % The linear typing judgment in our syntax takes on the following schematic form
@@
 %   adding this right after ``and the intuitionistic typing rules in fig 3''}
-Now that we have specified the syntactic aspect of \theoryabbv, we must justify
-its connections to formal grammars in order to show that it is a sound methodology
-for reasoning about parsers. We do this by showing that $\mathbf{Gr}$ is a model
-to our type theory.
-
-\begin{definition}
-
-\end{definition}
-
-To handle our universe, we assume a universe of small sets in our
-metatheory that is closed under all of our type formers.
+\subsection{Semantics}
+
+We can now define our denotational semantics.
 \begin{definition}[Grammar Semantics]
@@
 \max{TODO: how to describe $\mu$ succinctly?}\max{make sure $\lto$ and $\tol$ are right}
-
 \begin{figure}
@@
     \sem{\SPF\,X}\gamma &= \{ F : \textrm{Functor}(\textrm{Grammar}^{\sem{X}\gamma}, \textrm{Grammar}) \pipe F \textrm{ preserves $\omega$-colimits.} \}\\
-    \sem{\el(A)}\gamma\,G &= \sem{A}\gamma\,G\\
-    \sem{\mu A}\gamma
+    \sem{\el(F)}\gamma\,G &= \sem{A}\gamma\,G\\
+    \sem{\mu A}\gamma &= \mu (\sem{A}\gamma) \tag{\Cref{thm:adamek}}
   \end{align*}
@@
 \end{figure}
+
+\max{TODO: prose}
+
+\max{todo: spelling}
+\begin{theorem}[Adamek]
+  If a functor $F : \mathcal C \to \mathcal C$ preserves
+  $\omega$-colimits and $\mathcal C$ has an initial object and
+  $\omega$-colimits, then $F$ has an initial algebra given by the
+  colimit of the chain
+  \[ 0 \to F(0) \to F(F(0)) \to \cdots \]
+\end{theorem}
 
@@
 
-\paragraph{LNL Dependent Types}
-\steven{talk about how binary sum and binary with are definable over $\Bool$}
-Given a non-linear type $X$, we may form both the dependent product type
-$\LinPiTy x X A$ and the dependent pair type $\LinSigTy x X A$ as linear types
-where $x$ is free in $A$.
-
-$\sem{\LinPiTy x X A} \gamma w = \Pi(x:\sem{X}\gamma) \sem{A}(\gamma,x) w$
-
-$\sem{\LinSigTy x X A} \gamma w = \Sigma(x:\sem{X}\gamma) \sem{A}(\gamma,x) w$
-
-The grammar semantics of the linear product type is
-indeed a dependent function out of the
-semantics of $X$. Likewise, the grammar semantics of the linear dependent pair type is a dependent
-pair in $\Set$.\pedro{The connection to grammar semantics needs to be better explained. For instance,
-we should explain that disjunction corresponds to ``or'' of grammars}
-
-Note that even though we do not take the binary additive conjunction and
-disjunction as primitive, we may define them via these LNL dependent types. In
-particular, via a dependence on $\Bool$.
-%
-\[
-  A \amp B := \LinPiTy b \Bool {C(b)}
-\]
-\[
-  A \oplus B := \LinSigTy b \Bool {C(b)},
-\]
-where $C(\true) = A, C(\false) = B$.
-
-\paragraph{Universal Type}\max{this is absolutely not the right terminology. Universal type means multiple other completely different things}
-\steven{Definable as nullary product}
-The universal type $\top$ may be formed in any context.
-%
-\[ \sem{\top} \gamma w = \{ \ast \}\]
-%
-Its grammar semantics in set outputs the unit type in $\Set$ for all input strings in all contexts.
-
-\paragraph{Empty Type}
-\steven{Definable as nullary sum}
-The empty type $\bot$ has no inhabitants. The elimination for the empty type
-witnesses the principle of explosion --- i.e.\ from a term of type $\bot$ we may
-introduce a term $\mathsf{absurd}_{A} : A$ for any type $A$.
-%
-\[ \sem {\bot} \gamma w = \emptyset \]
-%
-\paragraph{The $G$ Modality}
-\steven{Rename this}
-\steven{Compare to persistence modalities from separation logic}
-The $G$ modality provides the embedding of linear types in the empty context
-into non-linear types. The introduction and elimination forms for $G A$ obey the
-same rules as given in \cite{bentonMixedLinearNonlinear1995} and \cite{krishnaswami_integrating_2015}.
-
-The left adjoint $F$ from non-linear types back to linear types may be defined
-as,
-%
-\[
-  F X := \LinSigTy a A I
-\]
-%
-\steven{Elaborate on $G$ more and rename $G A$to $\ltonl A$ in this section}
-
-\pedro{Agreed :) maybe mention connections to persistent propositions from the
-separation logic literature}
-
-The semantics of $G A$ are exactly the semantics of $A$ at the empty word $\varepsilon$.
-%
-\[ \sem{G A} \gamma = \sem{A} \gamma \varepsilon \]
-%
-%
-\paragraph{Recursive Types}
-\steven{reflavor this section as ``indexed inductive'' instead of ``recursive'',
-and entirely rework with the new inductive definition}
-\steven{The non-indexed inductive are definable as being trivially indexed over unit}
-Recursive linear types may be defined via the least-fixed point operator $\mu$.
+%% \paragraph{LNL Dependent Types}
+%% \steven{talk about how binary sum and binary with are definable over $\Bool$}
+%% Given a non-linear type $X$, we may form both the dependent product type
+%% $\LinPiTy x X A$ and the dependent pair type $\LinSigTy x X A$ as linear types
+%% where $x$ is free in $A$.
+
+%% $\sem{\LinPiTy x X A} \gamma w = \Pi(x:\sem{X}\gamma) \sem{A}(\gamma,x) w$
+
+%% $\sem{\LinSigTy x X A} \gamma w = \Sigma(x:\sem{X}\gamma) \sem{A}(\gamma,x) w$
+
+%% The grammar semantics of the linear product type is
+%% indeed a dependent function out of the
+%% semantics of $X$. Likewise, the grammar semantics of the linear dependent pair type is a dependent
+%% pair in $\Set$.\pedro{The connection to grammar semantics needs to be better explained. For instance,
+%% we should explain that disjunction corresponds to ``or'' of grammars}
+
+%% Note that even though we do not take the binary additive conjunction and
+%% disjunction as primitive, we may define them via these LNL dependent types. In
+%% particular, via a dependence on $\Bool$.
+%% %
+%% \[
+%%   A \amp B := \LinPiTy b \Bool {C(b)}
+%% \]
+%% \[
+%%   A \oplus B := \LinSigTy b \Bool {C(b)},
+%% \]
+%% where $C(\true) = A, C(\false) = B$.
+
+%% \paragraph{Universal Type}\max{this is absolutely not the right terminology. Universal type means multiple other completely different things}
+%% \steven{Definable as nullary product}
+%% The universal type $\top$ may be formed in any context.
+%% %
+%% \[ \sem{\top} \gamma w = \{ \ast \}\]
+%% %
+%% Its grammar semantics in set outputs the unit type in $\Set$ for all input strings in all contexts.
+
+%% \paragraph{Empty Type}
+%% \steven{Definable as nullary sum}
+%% The empty type $\bot$ has no inhabitants. The elimination for the empty type
+%% witnesses the principle of explosion --- i.e.\ from a term of type $\bot$ we may
+%% introduce a term $\mathsf{absurd}_{A} : A$ for any type $A$.
+%% %
+%% \[ \sem {\bot} \gamma w = \emptyset \]
+%% %
+%% \paragraph{The $G$ Modality}
+%% \steven{Rename this}
+%% \steven{Compare to persistence modalities from separation logic}
+%% The $G$ modality provides the embedding of linear types in the empty context
+%% into non-linear types. The introduction and elimination forms for $G A$ obey the
+%% same rules as given in \cite{bentonMixedLinearNonlinear1995} and \cite{krishnaswami_integrating_2015}.
+
+%% The left adjoint $F$ from non-linear types back to linear types may be defined
+%% as,
+%% %
+%% \[
+%%   F X := \LinSigTy a A I
+%% \]
+%% %
+%% \steven{Elaborate on $G$ more and rename $G A$to $\ltonl A$ in this section}
+
+%% \pedro{Agreed :) maybe mention connections to persistent propositions from the
+%% separation logic literature}
+
+%% The semantics of $G A$ are exactly the semantics of $A$ at the empty word $\varepsilon$.
+%% %
+%% \[ \sem{G A} \gamma = \sem{A} \gamma \varepsilon \]
+%% %
+%% %
+%% \paragraph{Recursive Types}
+%% \steven{reflavor this section as ``indexed inductive'' instead of ``recursive'',
+%% and entirely rework with the new inductive definition}
+%% \steven{The non-indexed inductive are definable as being trivially indexed over unit}
+%% Recursive linear types may be defined via the least-fixed point operator $\mu$.
 % The grammar semantics of which are the fixed-point of sets induced by the
@@
 %     \Gamma ; \cdot \vdash p_{\varepsilon} : h \\
-%     \Gamma ; y : h , x : h \vdash p_{\ast} : h
-%   }
-%   {\Gamma ; \Delta \vdash \mathsf{foldl}(p_{\varepsilon} , p_{\ast}) : g^{*}}
-% \end{equation}
-
-% In fact, the $\mathsf{foldl}$ term is defined using
-% $\mathsf{foldr}$ --- much in the same way one
-% may define a left fold over lists in terms of a right fold
-% in a functional programming language.
-% The underlying trick is to fold over a list of linear functions
-% instead of the original string. We curry each character $c$
-% of the string into a function that concatenates $c$, and
-% right fold over this list of linear functions. Because function
-% application is left-associative, this results in a left
-% fold over the original string.
-
-% We only take fixed points of a single variable as a
-% primitive operation in the type theory, but we may apply
-% Beki\`c's theorem \cite{Bekić1984} to define an admissible
-% notion of multivariate fixed point. This is particularly
-% useful for defining grammars that encode the states of an
-% automaton, as we will see in \cref{sec:automata}. In \cref{fig:multifix} we provide the
-% introduction and elimination principles for such a fixed
-% point, where $\sigma$ is the substitution that unrolls the
-% mutually recursive definitions one level. That is,
-
-% \begin{align*}
-%   \sigma = \{ & \mu(X_{1} = A_{1} \dots, X_{n} = A_{n}).X_{1} / X_{1} , \dots, \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{n} / X_{n} \}
-% \end{align*}
-
-% \begin{figure}
-% \begin{mathpar}
-%   \inferrule
-%   {\Gamma ; \Delta \vdash e : \simulsubst {A_{k}} {\sigma}}
-%   {\Gamma ; \Delta \vdash \mathsf{cons}~e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{k}}
-
-%   \and
-
-%   \inferrule
-%   {\Gamma ; \Delta \vdash e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{k} \\
-%              \Gamma ; x_{j} : \simulsubst {A_{j}}{\sigma} \vdash e_{j} : B_{j}\quad \forall j
-%   }
-%   {\Gamma; \Delta \vdash \mathsf{mfold}(x_{1}.e_{1}, \dots, x_{n}.e_{n})(e) : B_{k}}
-% \end{mathpar}
-% \caption{Multi-fixed Points}
-% \label{fig:multifix}
-% \end{figure}
-
-%% \section{Categorical Semantics of \theoryabbv}
-%% \label{sec:categorify}
-
-As described in previous sections, \theoryabbv has a ``standard''
-interpretation where non-linear types denote sets, linear types denote
-formal grammars, non-linear terms denote functions and linear terms
-denote parse transformers.
-%
-Further, as is typical for type theory, every type constructor in the
-calculus is interpreted by using a \emph{universal construction} in
-either the category of sets or the category of grammars\footnote{the
-only exception is that the axioms for Brzozowski derivatives we
-considered do not follow solely from universal properties. In this
-section we focus on models that do not necessarily satisfy these
-axioms.}.
-%
-This leads to an immediate opportunity for generalization: the type
-theory has in addition to the standard grammar-theoretic
-interpretation, an interpretation in any categorical structure that
-exhibits the same universal constructions.
-%
-On the one hand this gives us a structured way to formalize the
-standard semantics precisely, but additionally it enables us to
-consider \emph{non-standard} models in the next section that point to
-further applications as well as meta-theoretic results.
-
-In developing this categorical semantics, we will start with the
-closest analogue from formal language theory: Kleene algebra.  Kleene
-algebras are an important tool in the theory of regular languages
-serving as a bridge between algebraic reasoning and equivalence of
-regular expressions. More broadly, through various extensions, they
-serve as a theoretical substrate to studying different kinds of formal
-languages.
-%
-We can then see that the ``regular fragment'' of our type theory
-(i.e., just characters, $\otimes$, $\oplus$, $0,1$ and a Kleene star)
-has models in what we call a \emph{Kleene category} a categorification
-of Kleene algebra from posets to categories.
-%
-We then further develop this into our final notion of model, which we
-call a \emph{Chomsky category} as it can model not just regular
-languages, but the full Chomsky hierarchy.
-
-\subsection{Kleene Algebra and Kleene Categories}
-A Kleene algebra is a tuple $(A, +, \cdot, (-)^*, 1, 0)$, where $A$ is
-a set, $+$ and $\cdot$ are binary operations over $A$, $(-)^*$ is a
-function over $A$, and $1$ and $0$ are constants. These structures
-satisfy the axioms depicted in Figure~\ref{fig:axioms}.
-
-\begin{figure}
-  \begin{align*}
-    x + (y + z) &= (x + y) + z & x + y &= y + x\\
-    x + 0 &= x & x + x &= x\\
-    x(yz) &= (xy)z & x1 &= 1x = x\\
-    x(y + z) &= xy + xz & (x + y)z &= xz + yz\\
-    x0 &= 0x = x & & \\
-    1 + aa^* &\leq a^* & 1 + a^*a &\leq a^*\\
-     b + ax \leq x &\implies a^*b \leq x &  b + xa \leq x &\implies ba^* \leq x
-  \end{align*}
-  \caption{Kleene algebra axioms}
-  \label{fig:axioms}
-\end{figure}
-
-The addition operation can be used to define the partial order
-structure $a \leq b$ if, and only if, $a + b = b$. In the theory of
-formal languages, this order structure can be used to model language
-containment. In this section, we categorify the concept of Kleene
-algebra and build on top of it in order to define an abstract theory
-of parsing. We start by defining \emph{Kleene categories}.
-
-\begin{definition}
-  A Kleene category is a distributive monoidal category $\cat{K}$
-  such that for every objects $A$ and $B$, the endofunctors $F_{A, B}
-  = B + A \otimes X$ and $G_{A, B} = B + X \otimes A$ have initial
-  algebras (denoted $\mu X.\, F_{A, B}(X)$) such that $B \otimes (\mu
-  X.\, F_{A, 1}) \cong \mu X.\, F_{A, B}(X)$ and the analogous isomorphism
-  for $G_{A,B}$ also holds.
-\end{definition}
-
-As a sanity check, note that Kleene algebras are indeed examples of
-Kleene categories.
-
-\begin{example}
-  Every Kleene algebra, seen a posetal category, is a Kleene category.
-  The product $\cdot$ is a monoidal product and the addition is a
-  least-upper bound, which corresponds to a coproduct. Lastly, the
-  axioms of the Kleene star have a direct correspondence to the
-  coherence conditions postulated by the initial algebras of Kleene
-  categories.
-\end{example}
-
-This example provides a neat categorical justification to how
-restrictive Kleene algebras are in terms of reasoning about
-languages. By only having at most one morphism between objects, there
-is not a lot of information they can convey. In this case, the only
-information you get is language containment. As demonstrated by
-$\mathbf{Gr}$, the extra degrees of freedom granted by having more
-morphisms gives you more algebraic structure for reasoning about
-languages.
-
-For the next example, we see an unexpected connection with the theory
-of substructural logics.
-
-\begin{example}
-  The opposite category of every Kleene category is a model of a variant of
-  conjunctive ordered logic, where the Kleene star plays the role of the ``of
-  course'' modality from substructural logics which allows hypotheses to
-  be discarded or duplicated.
-\end{example}
-
-As we have seen, the proposed axioms are a direct translation of the
-Kleene algebra axioms to a categorical setting. Its most unusual aspect is the
-axiomatization of the Kleene star as a family of initial algebras
-satisfying certain isomorphisms. If the Kleene category $\cat{K}$ has
-more structure, then these isomorphisms hold ``for free''.
-
-\begin{theorem}
-  \label{th:kleeneclosed}
-  Let $\cat{K}$ be a Kleene category such that it is also monoidal
-  closed.  Then, the initial algebras isomorphisms hold automatically.
-\end{theorem}
-\begin{proof}
-  We prove this by the unicity (up-to isomorphism) of initial
-  algebras. Let $[hd, tl]: I + (\mu X.\, F_{A, I}(X)) \otimes A \to
-  (\mu X.\, F_{A, I}(X))$ be the initial algebra structure of $(\mu
-  X.\, F_{A, I}(X))$ and consider the map $[hd, tl] : B + B \otimes
-  (\mu X.\, F_{A, I}(X)) \otimes A \to B\otimes (\mu X.\, F_{A,
-    I}(X))$.
-
-  Now, let $[f,g] : B + A \otimes Y \to Y$ be an $F_{A,B}$-algebra and
-  we want to show that there is a unique algebra morphism $h : \mu X.\, F_{A,I} \to B \lto Y$. We can show existence and
-  uniqueness by showing that the diagram on top commutes if, and
-  only if, the diagram on the bottom commutes:
-
-% https://q.uiver.app/#q=WzAsOCxbMCwwLCJCICsgQiBcXG90aW1lcyAoXFxtdSBYLlxcLCAxICsgWCBcXG90aW1lcyBBKSJdLFsyLDAsIkIgKyBZIFxcb3RpbWVzIEEiXSxbMiwyLCJZIl0sWzAsMiwiQiBcXG90aW1lcyAoXFxtdSBYLlxcLCAxICsgWCBcXG90aW1lcyBBKSJdLFswLDMsIjEgKyAoXFxtdSBYLlxcLCAxICsgWCBcXG90aW1lcyBBKSJdLFswLDUsIlxcbXUgWC5cXCwgMSArIFggXFxvdGltZXMgQSJdLFsyLDMsIjEgKyAoIEIgXFxsdG8gWSkgXFxvdGltZXMgQSJdLFsyLDUsIkIgXFxsdG8gWSJdLFswLDEsImlkICsgKGlkIFxcb3RpbWVzIGg7IGV2KSBcXG90aW1lcyBpZF9BIl0sWzEsMiwiW2YsZ10iXSxbMywyLCJpZCBcXG90aW1lcyBoOyBldiIsMl0sWzAsMywiW2lkIFxcb3RpbWVzIGgsIGlkIFxcb3RpbWVzIHRsXSIsMl0sWzQsNiwiaWQgKyAoaCBcXG90aW1lcyBYKVxcb3RpbWVzIGlkIl0sWzUsNywiaCIsMl0sWzYsNywiW2YnLCBnJ10iXSxbNCw1LCJbaGQsIHRsXSIsMl1d
-\[\begin{tikzcd}
-	{B + B \otimes (\mu X.\, I + X \otimes A)} && {B + Y \otimes A} \\
-	\\
-	{B \otimes (\mu X.\, I + X \otimes A)} && Y \\
-	{I + (\mu X.\, I + X \otimes A)} && {I + ( B \lto Y) \otimes A} \\
-	\\
-	{\mu X.\, I + X \otimes A} && {B \lto Y}
-	\arrow["{id + (id \otimes h; ev) \otimes id_A}", from=1-1, to=1-3]
-	\arrow["{[f,g]}", from=1-3, to=3-3]
-	\arrow["{id \otimes h; ev}"', from=3-1, to=3-3]
-	\arrow["{[id \otimes h, id \otimes tl]}"', from=1-1, to=3-1]
-	\arrow["{id + (h \otimes X)\otimes id}", from=4-1, to=4-3]
-	\arrow["h"', from=6-1, to=6-3]
-	\arrow["{[f', g']}", from=4-3, to=6-3]
-	\arrow["{[hd, tl]}"', from=4-1, to=6-1]
-\end{tikzcd}\]
-  This equivalence follows by using the adjunction structure given
-  by the monoidal closed structure of $\cat{K}$. A completely analogous
-  argument for $G_{A,B}$ also holds. Furthermore, by generalizing the
-  construction of \Cref{sec:formaltype}, we can also show that from
-  the monoidal closed assumption it follows that $\mu X.\, F_{A, I}(X) \cong \mu X.\, G_{A, I}(X)$
-\end{proof}
-
-Something surprising about this lemma is that it provides an alternative
-perspective on the observation that if a Kleene algebra has an
-residuation operation, also called action algebra \cite{kozen1994action},
-then the Kleene star admits a simpler axiomatization.
-
-Since we want Kleene categories to generalize our notion of formal
-grammars as presheaves $\String \to \Set$, we prove that they do
-indeed form a Kleene category. We start by presenting a well-known
-construction from presheaf categories.
-
-\begin{definition}
-  Let $\cat{C}$ be a locally small monoidal category and $F$, $G$ be
-  two functors $\cat{C} \to \Set$. Their Day convolution tensor
-  product is defined as the following coend formula:
-  \[
-  (F \otimes_{Day} G)(x) = \int^{(y,z) \in \cat{C}\times\cat{C}}\cat{C}(y\otimes z, x) \times F(y) \times G(z)
-  \]
-  Dually, its internal hom is given by the following end formula:
-  \[
-  (F \lto_{Day} G)(x) = \int_{y} \Set(F(y), G(x \otimes y))
-  \]
-\end{definition}
-
-\begin{lemma}[Day \cite{day1970construction}]
-  Under the assumptions above, the presheaf category $\Set^{\cat{C}}$ is
-  monoidal closed.
-\end{lemma}
+%% %     \Gamma ; y : h , x : h \vdash p_{\ast} : h
+%% %   }
+%% %   {\Gamma ; \Delta \vdash \mathsf{foldl}(p_{\varepsilon} , p_{\ast}) : g^{*}}
+%% % \end{equation}
+
+%% % In fact, the $\mathsf{foldl}$ term is defined using
+%% % $\mathsf{foldr}$ --- much in the same way one
+%% % may define a left fold over lists in terms of a right fold
+%% % in a functional programming language.
+%% % The underlying trick is to fold over a list of linear functions
+%% % instead of the original string. We curry each character $c$
+%% % of the string into a function that concatenates $c$, and
+%% % right fold over this list of linear functions. Because function
+%% % application is left-associative, this results in a left
+%% % fold over the original string.
+
+%% % We only take fixed points of a single variable as a
+%% % primitive operation in the type theory, but we may apply
+%% % Beki\`c's theorem \cite{Bekić1984} to define an admissible
+%% % notion of multivariate fixed point. This is particularly
+%% % useful for defining grammars that encode the states of an
+%% % automaton, as we will see in \cref{sec:automata}. In \cref{fig:multifix} we provide the
+%% % introduction and elimination principles for such a fixed
+%% % point, where $\sigma$ is the substitution that unrolls the
+%% % mutually recursive definitions one level. That is,
+
+%% % \begin{align*}
+%% %   \sigma = \{ & \mu(X_{1} = A_{1} \dots, X_{n} = A_{n}).X_{1} / X_{1} , \dots, \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{n} / X_{n} \}
+%% % \end{align*}
+
+%% % \begin{figure}
+%% % \begin{mathpar}
+%% %   \inferrule
+%% %   {\Gamma ; \Delta \vdash e : \simulsubst {A_{k}} {\sigma}}
+%% %   {\Gamma ; \Delta \vdash \mathsf{cons}~e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{k}}
+
+%% %   \and
+
+%% %   \inferrule
+%% %   {\Gamma ; \Delta \vdash e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{k} \\
+%% %              \Gamma ; x_{j} : \simulsubst {A_{j}}{\sigma} \vdash e_{j} : B_{j}\quad \forall j
+%% %   }
+%% %   {\Gamma; \Delta \vdash \mathsf{mfold}(x_{1}.e_{1}, \dots, x_{n}.e_{n})(e) : B_{k}}
+%% % \end{mathpar}
+%% % \caption{Multi-fixed Points}
+%% % \label{fig:multifix}
+%% % \end{figure}
+
+%% %% \section{Categorical Semantics of \theoryabbv}
+%% %% \label{sec:categorify}
+
+%% As described in previous sections, \theoryabbv has a ``standard''
+%% interpretation where non-linear types denote sets, linear types denote
+%% formal grammars, non-linear terms denote functions and linear terms
+%% denote parse transformers.
+%% %
+%% Further, as is typical for type theory, every type constructor in the
+%% calculus is interpreted by using a \emph{universal construction} in
+%% either the category of sets or the category of grammars\footnote{the
+%% only exception is that the axioms for Brzozowski derivatives we
+%% considered do not follow solely from universal properties. In this
+%% section we focus on models that do not necessarily satisfy these
+%% axioms.}.
+%% %
+%% This leads to an immediate opportunity for generalization: the type
+%% theory has in addition to the standard grammar-theoretic
+%% interpretation, an interpretation in any categorical structure that
+%% exhibits the same universal constructions.
+%% %
+%% On the one hand this gives us a structured way to formalize the
+%% standard semantics precisely, but additionally it enables us to
+%% consider \emph{non-standard} models in the next section that point to
+%% further applications as well as meta-theoretic results.
+
+%% In developing this categorical semantics, we will start with the
+%% closest analogue from formal language theory: Kleene algebra.  Kleene
+%% algebras are an important tool in the theory of regular languages
+%% serving as a bridge between algebraic reasoning and equivalence of
+%% regular expressions. More broadly, through various extensions, they
+%% serve as a theoretical substrate to studying different kinds of formal
+%% languages.
+%% %
+%% We can then see that the ``regular fragment'' of our type theory
+%% (i.e., just characters, $\otimes$, $\oplus$, $0,1$ and a Kleene star)
+%% has models in what we call a \emph{Kleene category} a categorification
+%% of Kleene algebra from posets to categories.
+%% %
+%% We then further develop this into our final notion of model, which we
+%% call a \emph{Chomsky category} as it can model not just regular
+%% languages, but the full Chomsky hierarchy.
+
+%% \subsection{Comparison to Kleene Algebra}
+%% \subsection{Kleene Algebra and Kleene Categories}
+%% A Kleene algebra is a tuple $(A, +, \cdot, (-)^*, 1, 0)$, where $A$ is
+%% a set, $+$ and $\cdot$ are binary operations over $A$, $(-)^*$ is a
+%% function over $A$, and $1$ and $0$ are constants. These structures
+%% satisfy the axioms depicted in Figure~\ref{fig:axioms}.
+
+%% \begin{figure}
+%%   \begin{align*}
+%%     x + (y + z) &= (x + y) + z & x + y &= y + x\\
+%%     x + 0 &= x & x + x &= x\\
+%%     x(yz) &= (xy)z & x1 &= 1x = x\\
+%%     x(y + z) &= xy + xz & (x + y)z &= xz + yz\\
+%%     x0 &= 0x = x & & \\
+%%     1 + aa^* &\leq a^* & 1 + a^*a &\leq a^*\\
+%%      b + ax \leq x &\implies a^*b \leq x &  b + xa \leq x &\implies ba^* \leq x
+%%   \end{align*}
+%%   \caption{Kleene algebra axioms}
+%%   \label{fig:axioms}
+%% \end{figure}
+
+%% The addition operation can be used to define the partial order
+%% structure $a \leq b$ if, and only if, $a + b = b$. In the theory of
+%% formal languages, this order structure can be used to model language
+%% containment. In this section, we categorify the concept of Kleene
+%% algebra and build on top of it in order to define an abstract theory
+%% of parsing. We start by defining \emph{Kleene categories}.
+
+%% \begin{definition}
+%%   A Kleene category is a distributive monoidal category $\cat{K}$
+%%   such that for every objects $A$ and $B$, the endofunctors $F_{A, B}
+%%   = B + A \otimes X$ and $G_{A, B} = B + X \otimes A$ have initial
+%%   algebras (denoted $\mu X.\, F_{A, B}(X)$) such that $B \otimes (\mu
+%%   X.\, F_{A, 1}) \cong \mu X.\, F_{A, B}(X)$ and the analogous isomorphism
+%%   for $G_{A,B}$ also holds.
+%% \end{definition}
+
+%% As a sanity check, note that Kleene algebras are indeed examples of
+%% Kleene categories.
+
+%% \begin{example}
+%%   Every Kleene algebra, seen a posetal category, is a Kleene category.
+%%   The product $\cdot$ is a monoidal product and the addition is a
+%%   least-upper bound, which corresponds to a coproduct. Lastly, the
+%%   axioms of the Kleene star have a direct correspondence to the
+%%   coherence conditions postulated by the initial algebras of Kleene
+%%   categories.
+%% \end{example}
+
+%% This example provides a neat categorical justification to how
+%% restrictive Kleene algebras are in terms of reasoning about
+%% languages. By only having at most one morphism between objects, there
+%% is not a lot of information they can convey. In this case, the only
+%% information you get is language containment. As demonstrated by
+%% $\mathbf{Gr}$, the extra degrees of freedom granted by having more
+%% morphisms gives you more algebraic structure for reasoning about
+%% languages.
+
+%% For the next example, we see an unexpected connection with the theory
+%% of substructural logics.
+
+%% \begin{example}
+%%   The opposite category of every Kleene category is a model of a variant of
+%%   conjunctive ordered logic, where the Kleene star plays the role of the ``of
+%%   course'' modality from substructural logics which allows hypotheses to
+%%   be discarded or duplicated.
+%% \end{example}
+
+%% As we have seen, the proposed axioms are a direct translation of the
+%% Kleene algebra axioms to a categorical setting. Its most unusual aspect is the
+%% axiomatization of the Kleene star as a family of initial algebras
+%% satisfying certain isomorphisms. If the Kleene category $\cat{K}$ has
+%% more structure, then these isomorphisms hold ``for free''.
 
 %% \begin{theorem}
-%%   Let $\cat{K}$ be a Kleene category and $A$ a discrete category.
-%%   The functor category $[A, \cat{K}]$.
-%%   (HOW GENERAL SHOULD THIS THEOREM BE? BY ASSUMING ENOUGH STRUCTURE,
-%%   E.G. K = Set, THIS THEOREM BECOMES SIMPLE TO PROVE)
+%%   \label{th:kleeneclosed}
+%%   Let $\cat{K}$ be a Kleene category such that it is also monoidal
+%%   closed.  Then, the initial algebras isomorphisms hold automatically.
 %% \end{theorem}
-\begin{theorem}
-  If $\cat{C}$ is a locally small monoidal category, then
-  $\Set^{\cat{C}}$ is a Kleene category.
-\end{theorem}
-\begin{proof}
-
-  By the lemma above, $\Set^{\cat{C}}$ is monoidal closed, and since it
-  is a presheaf category, it has coproducts. Furthermore, the tensor
-  is a left adjoint, i.e. it preserves colimits and, therefore, it is
-  a distributive category.
-
-  As for the Kleene star, since presheaf categories admit small colimits,
-  the initial algebra of the functors $F_{A,B}$ and $G_{A,B}$ can be
-  defined as the filtered colimit of the diagrams:
-
-  From Theorem~\ref{th:kleeneclosed} it follows that these initial
-  algebras satisfy the required isomorphisms and this concludes the
-  proof.
-\end{proof}
-
-\begin{corollary}
-  For every alphabet $\Sigma$, the presheaf category $\Set^{\cat{\Sigma^*}}$
-  is a Kleene category.
-\end{corollary}
-\begin{proof}
-  Note that string concatenation and the empty string make the
-  discrete category $\Sigma^*$ a strict monoidal category.
-\end{proof}
-
-Much like in the posetal case, the abstract structure of a Kleene
-category is expressive enough to synthetically reason about regular
-languages. A significant difference between them is that while Kleene
-algebras can reason about language containment, Kleene categories can
-reason about \emph{ambiguity}, \emph{strong equivalence} of grammars.
-
-\subsection{Lambek Hyperdoctrines and Chomsky Hyperdoctrines}
-
-Though Kleene categories are expressive enough to reason about
-concepts that are outside of reach of Kleene algebras, their
-simply-typed nature makes them not so expressive from a type theoretic
-point of view. This is limiting because type theories are successful
-syntactic frameworks for manipulating complicated categorical
-structures while avoiding some issues common in category theory, such
-as coherence issues.
-
-With this in mind, we want to design a categorical semantics that
-builds on top of Kleene categories with the goal of extending them
-with dependent types and making them capable reasoning about languages
-and their parsers. This leads us to the abstract notion of model we
-are interested in capturing with \theoryabbv: a \emph{Chomsky
-category}.
-
-We do this in two stages: first we define a \emph{Lambek hyperdoctrine} to be a
-notion of model for the judgmental structure of \theoryabbv: that is we have
-notions of linear and non-linear type, contexts, substitution and terms, but do
-not assume any particular type constructions exist. Then we will define when a
-Lambek hyperdoctrine is a Chomsky hyperdoctrine, meaning it can interpret all
-the type formers of \theoryabbv and therefore arbitrary strength grammars.
-
-\begin{definition}[Lambek Hyperdoctrine]
-  A \emph{Lambek hyperdoctrine} consists of
-  \begin{enumerate}
-  \item A category $\mathcal C$ with a terminal object.
-  \item A category with families structure over $\mathcal C$.
-  \item A contravariant functor $L : \mathcal C^{o} \to \textrm{MultiCat}$ from
-    $\mathcal C$ to the category of Multicategories.
-  \end{enumerate}
-\end{definition}
-Here the objects of $\mathcal C$ model the non-linear contexts, and morphisms
-model non-linear substitutions. The category with families structure over
-$\mathcal C$ models the dependent non-linear types. Finally the
-``hyperdoctrine'' of multicategories $L$ models the linear types and terms. The
-fact that this is functorial in $\mathcal C$ corresponds to the fact that linear
-types are all relative to a non-linear context and that linear variables and
-composition commute with non-linear substitution.
-
-We will have three main Lambek hyperdoctrines of interest in this paper: the
-``standard model'' of sets and grammars, the ``syntactic model'' given by our
-type theory itself and lastly a gluing model introduced in
-Section~\ref{sec:canonicity} to prove the canonicity theorem.
-\begin{example}
-  \begin{enumerate}
-  \item Let $\mathcal C$ be the category of sets, equipped with its usual
-    category with families structures of families. Define $L : \mathcal C^{o}
-    \to \textrm{MultiCat}$ to map $L(X)$ to the representable multicategory
-    $(\Set^{\Sigma^*})^X$ of $X$-indexed families of grammars where the monoidal
-    structure is given pointwise by the Day convolution monoidal structure.
-  \item We can build a ``syntactic'' Lambek hyperdoctrine $\Syn(\Sigma)$ from the syntax
-    itself: the category $\mathcal C$ is given by non-linear contexts and
-    substitutions, the category with families by the non-linear types and terms
-    and the hyperdoctrine of multicategories by the linear types and terms.
-  \end{enumerate}
-\end{example}
-
-Next each type linear and non-linear type constructor corresponds to extra data
-on a Lambek hyperdoctrine.
-\begin{definition}
-  \label{def:chomsky-data}
-  \begin{enumerate}
-  \item Dependent type structure: standard (say extensional type theory, give a reference)
-  \item Universes
-  \item Inductive grammars?
-  \item Non-inductive Grammar constructors: standard monoidal category stuff
-  \end{enumerate}
-\end{definition}
-\begin{definition}[Chomsky Hyperdoctrine]
-  A Chomsky Hyperdoctrine is a Lambek hyperdoctrine equipped with all of the data in Definition~\ref{def:chomsky-data}.
-\end{definition}
-\pedro{We should probably define some of the words in this definition}
-\begin{definition}
-  A Chomsky category is a locally Cartesian category with two hierarchies of
-  universes $\{L_i\}_{i\in \nat}$ and $\{U_i\}_{i\in \nat}$ such that
-  every $L_i$ and $U_i$ are $U_{i+1}$-small. Furthermore, we require
-  $U_i$ to be closed under dependent products and sum,
-  $L_i$ to be closed under the Kleene category connectives,
-  dependent products, left and right closed structures, with
-  a type constructor $G : L_i \to U_i$ and a linear dependent sum
-  going the other direction.
-\end{definition}
-
-\steven{Max suggests augmenting the definition of a Chomsky category to
-  something like two categories $L$ and $U$, and $U$ is Cartesian closed I don't
-  fully recall the rest.
-
-  My best guess is that you take $L$ a Kleene category with a hierarchy of universes
-  two, and you further require that $L$ is $Psh(U)$-enriched, except he
-  suggested a further adjective on these presheaves. Perhaps representability?
-}
-
-\begin{theorem}
-  The presheaf category $\Grammar = \Set^{\cat{\Sigma^*}}$ is a Chomsky category.
-\end{theorem}
-
-Further, the syntactic category of \theoryname is manifestly a Chomsky category.
-
-\pedro{This is likely true, but if we explicitly say so, this warrants a proof. I think that
-if we don't say anything about the syntactic category, reviewers won't mind.}
-
-\steven{I agree that we don't want to say anything that opens unnecessary
-  questions for proofs we haven't written. However, it seems hard to make the
-  case that we have the right categorical model of the syntax if this statement
-  isn't true. By restating our definition of Chomsky category, this should be
-  obvious or a quick proof}
-
-\pedro{I agree, then we should add the quick proof :) }
-
-\subsection{Concrete Models of \theoryabbv}
-\label{sec:othermodels}
-
-\steven{Because we can define a version of semantic actions internally, we
-  shouldn't put it as a separate model}
-
-One of the powers of type theories is that it can be profitable to interpret them
-in various models. In this section, by using our just-defined Chomsky categories,
-we show how other useful concepts from formal language theory can also be organized
-as models of \theoryabbv. We illustrate this point by providing two examples that
-are closely related to the theory of formal languages: language equivalence and
-semantic actions. Furthermore, in order to justify how $\mathbf{Gr}$ relates to
-more traditional notions of parsing, we define a glued model that proves a
-canonicity property of grammar terms.
-
-\subsection{Language Semantics}
-Every grammar induces a language semantics. Also languages can be taken as a
-propositionally truncated view of the syntax. Logical equivalence should induce
-weak equivalence, and thus even give a syntactic way to reason about language equivalence.
-\steven{TODO language semantics}
-
-% \subsection{Semantic Actions}
-% \steven{Tentatively planning to cut this subsection for an internal representation
-%   of semantic actions}
-% Returning to the problem of parsing, the output of a parse usually is not the
-% literal parse tree. Rather, the output is the result of some \emph{semantic
-%   actions} ran on the parse tree, which usually serve to remove some syntactic
-% details that are unnecessary for later processing.
-
-% Given a grammar $G : \String \to \Set$, we define a semantic action to be a set
-% $X$ with a function $f$ that produces a semantic element from any parse of $G$.
-
-% \[
-%   f : \PiTy w \String {G w \to X}
-% \]
-
-% Further, semantic actions can be arranged into a structured category.
-% Define $\SemAct$ as the comma
-% category $\Grammar / \Delta$, where $\Delta : \Set \to \Grammar$ defines a
-% discrete presheaf. That is, for a set $X$, $\Delta (X)(w) = X$ for all
-% $w \in \String$. As $\SemAct$ is defined as a comma category, it has a forgetful
-% functor into $\Grammar$. That is, $\SemAct$ serves as a notion of formal
-% grammar. Moreover, $\SemAct$ is a model of \theoryabbv.
-
-% \steven{It being a notion of formal grammar is distinct from being a model. This
-% probably warrants a proof}
-
-% \steven{TODO semantic actions}
-
-% \pedro{This is a very nice opportunity of showing off the supremacy of denotational
-%   reasoning ;) We should probably prove the gluing lemma in the previous section
-%   and apply it here and in the canonicity section. The actual proof might have to be
-%   moved to the appendix, though}
-
-\subsection{Parse Canonicity}
-Canonicity is an important metatheoretic theorem in the type theory
-literature.  It provides insight on the normal forms of terms and,
-therefore, on its computational aspects. Frequently, proving
-canonicity for boolean types, i.e. every closed term of type bool
-reduces to either true or false, is enough to justify that the type
-theory being studied is well-behaved. In our case, however, since we
-want to connect \theoryabbv to parsers, we must provide a more
-detailed account of canonicity. In particular, we give a nonstandard semantics
-of \theoryabbv that carries a proof of canonicity along with it.
-
-If $\cdot \vdash A$ is a closed linear type then there are
-two obvious notions of what constitutes a ``parse'' of a string w
-according to the grammar $A$:
-\begin{enumerate}
-\item On the one hand we have the set-theoretic semantics just
-  defined, $\llbracket A \rrbracket \cdot w$
-\item On the other hand, we can view the string $w = c_1c_2\cdots$ as
-  a linear context $\lceil w \rceil = x_1:c_1,x_2:c_2,\ldots$ and
-  define a parse to be a $\beta\eta$-equivalence class of linear terms $\cdot;
-  \lceil w \rceil \vdash e : G$.
-\end{enumerate}
-It is not difficult to see that at least for the ``purely positive''
-formulae (those featuring only the positive connectives
-$0,+,I,\otimes,\mu, \overline\Sigma,c$) that
-every element $t \in \llbracket A \rrbracket w$ is a kind of tree and
-that the nodes of the tree correspond precisely to the introduction
-forms of the type. However it is far less obvious that \emph{every}
-linear term $\lceil w \rceil \vdash p : \phi$ is equal to some
-sequence of introduction forms since proofs can include elimination
-forms as well. To show that this is indeed the case we give a
-\emph{canonicity} result for the calculus: that the parses for .
-
-\begin{definition}
-  A non-linear type $X$ is purely positive if it is built up using
-  only finite sums, finite products and least fixed points.
-
-  A linear type is purely positive if it is built up using only finite
-  sums, tensor products, generators $c$, least fixed points and linear
-  sigma types over purely positive non-linear types.
-\end{definition}
-
-\begin{definition}
-  %% Let $X$ be a closed non-linear type. The closed elements $\textrm{Cl}(X)$ of $X$ are the definitional equivalence classes of terms $\cdot \vdash e : X$.
-
-  Let $A$ be a closed linear type. The nerve $N(A)$ is a presheaf on
-  strings that takes a string $w$ to the definitional equivalence
-  classes of terms $\cdot; \lceil w\rceil \vdash e: N(A)$.
-\end{definition}
-
-\begin{theorem}[Canonicity]
-  Let $A$ be a closed, purely positive linear type. Then there is an
-  isomorphism between $\llbracket A\rrbracket$ and $N(A)$.
-\end{theorem}
-\begin{proof}
-  We outline the proof here, more details are in the appendix. The
-  proof proceeds first by a standard logical families construction
-  that combines canonicity arguments for dependent type theory
-  TODO cite coquand
-  % \cite{coquand,etc}
-  with logical relations constructions for linear
-  types
-  TODO cite hylandschalk
-  % \cite{hylandschalk}
-  . It is easy to see by induction that the
-  logical family for $A$, $\hat A$ is isomorphic to $\llbracket A
-  \rrbracket$ and the fundamental lemma proves that the projection
-  morphism $p : \hat A \to N(A)$ has a section, the canonicalization
-  procedure. Then we establish again by induction that
-  canonicalization is also a retraction by observing that introduction
-  forms are taken to constructors.
-\end{proof}
-
-\begin{enumerate}
-\item Every term $\lceil w \rceil \vdash p : G + H$ is equal to $\sigma_1q$ or $\sigma_2 r$ (but not both)
-\item There are no terms $\lceil w \rceil \vdash p : 0$
-\item If there is a term $\lceil w \rceil \vdash p : c$ then $w = c$ and $p = x$.
-\item Every term $\lceil w \rceil \vdash p : G \otimes H$ is equal to $(q,r)$ for some $q,r$
-\item Every term $\lceil w \rceil \vdash p : \varepsilon$ is equal to $()$
-\item Every term $\lceil w \rceil \vdash p : c$ is equal to $x:c$
-\item Every term $\lceil w \rceil \vdash p : \mu X. G$ is equal to $\textrm{roll}(q)$ where $q : G(\mu X.G/X)$
-\item Every term $\lceil w \rceil \vdash p : (x:A) \times G$ is equal
-  to $(M,q)$ where $\cdot \vdash M : A$
-\end{enumerate}
-
-To prove this result we will use a logical families model. We give a
-brief overview of this model concretely:
-\begin{enumerate}
-\item A context $\Gamma$ denotes a family of sets indexed by closing substitutions $\hat\Gamma : (\cdot \vdash \Gamma) \Rightarrow \Set_i$
-\item A type $\Gamma \vdash X : U_i$ denotes a family of sets $\hat X : \Pi(\gamma:\cdot \vdash \Gamma) \hat\Gamma \Rightarrow (\cdot \vdash \simulsubst X \gamma) \Rightarrow \Set_i$
-\item A term $\Gamma \vdash e : X$ denotes a section $\hat e : \Pi(\gamma)\Pi(\hat\gamma)\hat X \gamma \hat\gamma (\simulsubst e \gamma)$
-\item A linear type $\Gamma \vdash A : L_i$ denotes a family of grammars $\hat A : \Pi(\gamma:\cdot\vdash\Gamma)\,\hat\Gamma \Rightarrow \Pi(w:\Sigma^*) (\cdot;\lceil w\rceil \vdash A[\gamma])\Rightarrow \Set_i$, and the denotation of a linear context $\Delta$ is similar.
-\item A linear term $\Gamma;\Delta \vdash e : A$ denotes a function \[\hat e : \Pi(\gamma)\Pi(\hat\gamma)\Pi(w)\Pi(\delta : \lceil w \rceil \vdash \simulsubst \Delta \gamma) \hat\Delta \gamma \hat\gamma \delta \Rightarrow \hat A \gamma \hat\gamma w {(\simulsubst {\simulsubst e \gamma} \delta)}\]
-\end{enumerate}
-And some of the constructions:
-\begin{enumerate}
-\item $\widehat {(G A)} \gamma \hat\gamma e = \hat A \gamma \hat\gamma \varepsilon (G^{-1}e)$
-\item $\widehat {(A \otimes B)} \gamma \hat\gamma w e = \Sigma(w_Aw_B = w)\Sigma(e_A)\Sigma(e_B) (e_A,e_B) = e \wedge \hat A \gamma \hat \gamma w_A e_A \times \hat B \gamma \hat \gamma w_B e_B$
-\item $\widehat {(A \lto B)} \gamma \hat\gamma w e = \Pi(w_A)\Pi(e_A) \hat A \gamma\hat\gamma w_A e_A \Rightarrow \hat B \gamma\hat\gamma (ww_A) (\applto {e_A} e)$
-\end{enumerate}
-
-First, the category with families will be
-the category of logical families over set contexts/types
-$\Delta$/$A$. Then the propositional portion will be defined by
-mapping a logical family $\hat \Gamma \to \Gamma$
-
-First, let $L$ be the category of BI formulae and proofs (quotiented
-by $\beta\eta$ equality). Define a functor $N : L \to \Set^{\Sigma^*}$ by
-\[ N(\phi)(w) = L(w,\phi) \]
-
-Then define the gluing category $\mathcal G$ as the comma category
-$\Set^{\Sigma^*}/N$. That is, an object of this category is a pair of
-a formula $\phi \in L$ and an object $S \in \mathcal
-P(\Sigma^*)/N(\phi)$. We can then use the equivalence $\mathcal
-P(\Sigma^*)/N(\phi) \cong \mathcal P(\int N(\phi))$ to get a simple
-description of such an $S$: it is simply a family of sets indexed by
-proofs $L(w,\phi)$:
-\[ \prod_{w\in\Sigma^*} L(w,\phi) \to \Set \]
-This category clearly comes with a projection functor $\pi : \mathcal
-G \to \mathcal L$ and then our goal is to define a section by using
-the universal property of $\mathcal L$.
-
-To this end we define
-\begin{enumerate}
-\item $(\phi, S) \otimes (\psi, T) = (\phi \otimes \psi, S\otimes T)$ where
-  \[ (S \otimes T)(w, p) = (w_1w_2 = w) \times (q_1,q_2 = p) \times S\,w_1\,q_1 \times T\,w_2\,q_2\]
-\item $(\phi, S) \multimap (\psi, T) = (\phi \multimap \psi, S \multimap T)$ where
-  \[ (S \multimap T)(w,p) = w' \to q \to S\,w'\,q \to T (ww') (p\,q) \]
-\item $\mu X. ??$ ??
-\end{enumerate}
-
-\pedro{We should conclude this section by explaining the relevance of the canonicity
-theorem. Could also be done before stating the theorem.}
+%% \begin{proof}
+%%   We prove this by the unicity (up-to isomorphism) of initial
+%%   algebras. Let $[hd, tl]: I + (\mu X.\, F_{A, I}(X)) \otimes A \to
+%%   (\mu X.\, F_{A, I}(X))$ be the initial algebra structure of $(\mu
+%%   X.\, F_{A, I}(X))$ and consider the map $[hd, tl] : B + B \otimes
+%%   (\mu X.\, F_{A, I}(X)) \otimes A \to B\otimes (\mu X.\, F_{A,
+%%     I}(X))$.
+
+%%   Now, let $[f,g] : B + A \otimes Y \to Y$ be an $F_{A,B}$-algebra and
+%%   we want to show that there is a unique algebra morphism $h : \mu X.\, F_{A,I} \to B \lto Y$. We can show existence and
+%%   uniqueness by showing that the diagram on top commutes if, and
+%%   only if, the diagram on the bottom commutes:
+
+%% % https://q.uiver.app/#q=WzAsOCxbMCwwLCJCICsgQiBcXG90aW1lcyAoXFxtdSBYLlxcLCAxICsgWCBcXG90aW1lcyBBKSJdLFsyLDAsIkIgKyBZIFxcb3RpbWVzIEEiXSxbMiwyLCJZIl0sWzAsMiwiQiBcXG90aW1lcyAoXFxtdSBYLlxcLCAxICsgWCBcXG90aW1lcyBBKSJdLFswLDMsIjEgKyAoXFxtdSBYLlxcLCAxICsgWCBcXG90aW1lcyBBKSJdLFswLDUsIlxcbXUgWC5cXCwgMSArIFggXFxvdGltZXMgQSJdLFsyLDMsIjEgKyAoIEIgXFxsdG8gWSkgXFxvdGltZXMgQSJdLFsyLDUsIkIgXFxsdG8gWSJdLFswLDEsImlkICsgKGlkIFxcb3RpbWVzIGg7IGV2KSBcXG90aW1lcyBpZF9BIl0sWzEsMiwiW2YsZ10iXSxbMywyLCJpZCBcXG90aW1lcyBoOyBldiIsMl0sWzAsMywiW2lkIFxcb3RpbWVzIGgsIGlkIFxcb3RpbWVzIHRsXSIsMl0sWzQsNiwiaWQgKyAoaCBcXG90aW1lcyBYKVxcb3RpbWVzIGlkIl0sWzUsNywiaCIsMl0sWzYsNywiW2YnLCBnJ10iXSxbNCw1LCJbaGQsIHRsXSIsMl1d
+%% \[\begin{tikzcd}
+%% 	{B + B \otimes (\mu X.\, I + X \otimes A)} && {B + Y \otimes A} \\
+%% 	\\
+%% 	{B \otimes (\mu X.\, I + X \otimes A)} && Y \\
+%% 	{I + (\mu X.\, I + X \otimes A)} && {I + ( B \lto Y) \otimes A} \\
+%% 	\\
+%% 	{\mu X.\, I + X \otimes A} && {B \lto Y}
+%% 	\arrow["{id + (id \otimes h; ev) \otimes id_A}", from=1-1, to=1-3]
+%% 	\arrow["{[f,g]}", from=1-3, to=3-3]
+%% 	\arrow["{id \otimes h; ev}"', from=3-1, to=3-3]
+%% 	\arrow["{[id \otimes h, id \otimes tl]}"', from=1-1, to=3-1]
+%% 	\arrow["{id + (h \otimes X)\otimes id}", from=4-1, to=4-3]
+%% 	\arrow["h"', from=6-1, to=6-3]
+%% 	\arrow["{[f', g']}", from=4-3, to=6-3]
+%% 	\arrow["{[hd, tl]}"', from=4-1, to=6-1]
+%% \end{tikzcd}\]
+%%   This equivalence follows by using the adjunction structure given
+%%   by the monoidal closed structure of $\cat{K}$. A completely analogous
+%%   argument for $G_{A,B}$ also holds. Furthermore, by generalizing the
+%%   construction of \Cref{sec:formaltype}, we can also show that from
+%%   the monoidal closed assumption it follows that $\mu X.\, F_{A, I}(X) \cong \mu X.\, G_{A, I}(X)$
+%% \end{proof}
+
+%% Something surprising about this lemma is that it provides an alternative
+%% perspective on the observation that if a Kleene algebra has an
+%% residuation operation, also called action algebra \cite{kozen1994action},
+%% then the Kleene star admits a simpler axiomatization.
+
+%% Since we want Kleene categories to generalize our notion of formal
+%% grammars as presheaves $\String \to \Set$, we prove that they do
+%% indeed form a Kleene category. We start by presenting a well-known
+%% construction from presheaf categories.
+
+%% \begin{definition}
+%%   Let $\cat{C}$ be a locally small monoidal category and $F$, $G$ be
+%%   two functors $\cat{C} \to \Set$. Their Day convolution tensor
+%%   product is defined as the following coend formula:
+%%   \[
+%%   (F \otimes_{Day} G)(x) = \int^{(y,z) \in \cat{C}\times\cat{C}}\cat{C}(y\otimes z, x) \times F(y) \times G(z)
+%%   \]
+%%   Dually, its internal hom is given by the following end formula:
+%%   \[
+%%   (F \lto_{Day} G)(x) = \int_{y} \Set(F(y), G(x \otimes y))
+%%   \]
+%% \end{definition}
+
+%% \begin{lemma}[Day \cite{day1970construction}]
+%%   Under the assumptions above, the presheaf category $\Set^{\cat{C}}$ is
+%%   monoidal closed.
+%% \end{lemma}
+
+%% %% \begin{theorem}
+%% %%   Let $\cat{K}$ be a Kleene category and $A$ a discrete category.
+%% %%   The functor category $[A, \cat{K}]$.
+%% %%   (HOW GENERAL SHOULD THIS THEOREM BE? BY ASSUMING ENOUGH STRUCTURE,
+%% %%   E.G. K = Set, THIS THEOREM BECOMES SIMPLE TO PROVE)
+%% %% \end{theorem}
+%% \begin{theorem}
+%%   If $\cat{C}$ is a locally small monoidal category, then
+%%   $\Set^{\cat{C}}$ is a Kleene category.
+%% \end{theorem}
+%% \begin{proof}
+
+%%   By the lemma above, $\Set^{\cat{C}}$ is monoidal closed, and since it
+%%   is a presheaf category, it has coproducts. Furthermore, the tensor
+%%   is a left adjoint, i.e. it preserves colimits and, therefore, it is
+%%   a distributive category.
+
+%%   As for the Kleene star, since presheaf categories admit small colimits,
+%%   the initial algebra of the functors $F_{A,B}$ and $G_{A,B}$ can be
+%%   defined as the filtered colimit of the diagrams:
+
+%%   From Theorem~\ref{th:kleeneclosed} it follows that these initial
+%%   algebras satisfy the required isomorphisms and this concludes the
+%%   proof.
+%% \end{proof}
+
+%% \begin{corollary}
+%%   For every alphabet $\Sigma$, the presheaf category $\Set^{\cat{\Sigma^*}}$
+%%   is a Kleene category.
+%% \end{corollary}
+%% \begin{proof}
+%%   Note that string concatenation and the empty string make the
+%%   discrete category $\Sigma^*$ a strict monoidal category.
+%% \end{proof}
+
+%% Much like in the posetal case, the abstract structure of a Kleene
+%% category is expressive enough to synthetically reason about regular
+%% languages. A significant difference between them is that while Kleene
+%% algebras can reason about language containment, Kleene categories can
+%% reason about \emph{ambiguity}, \emph{strong equivalence} of grammars.
+
+%% \subsection{Lambek Hyperdoctrines and Chomsky Hyperdoctrines}
+
+%% Though Kleene categories are expressive enough to reason about
+%% concepts that are outside of reach of Kleene algebras, their
+%% simply-typed nature makes them not so expressive from a type theoretic
+%% point of view. This is limiting because type theories are successful
+%% syntactic frameworks for manipulating complicated categorical
+%% structures while avoiding some issues common in category theory, such
+%% as coherence issues.
+
+%% With this in mind, we want to design a categorical semantics that
+%% builds on top of Kleene categories with the goal of extending them
+%% with dependent types and making them capable reasoning about languages
+%% and their parsers. This leads us to the abstract notion of model we
+%% are interested in capturing with \theoryabbv: a \emph{Chomsky
+%% category}.
+
+%% We do this in two stages: first we define a \emph{Lambek hyperdoctrine} to be a
+%% notion of model for the judgmental structure of \theoryabbv: that is we have
+%% notions of linear and non-linear type, contexts, substitution and terms, but do
+%% not assume any particular type constructions exist. Then we will define when a
+%% Lambek hyperdoctrine is a Chomsky hyperdoctrine, meaning it can interpret all
+%% the type formers of \theoryabbv and therefore arbitrary strength grammars.
+
+%% \begin{definition}[Lambek Hyperdoctrine]
+%%   A \emph{Lambek hyperdoctrine} consists of
+%%   \begin{enumerate}
+%%   \item A category $\mathcal C$ with a terminal object.
+%%   \item A category with families structure over $\mathcal C$.
+%%   \item A contravariant functor $L : \mathcal C^{o} \to \textrm{MultiCat}$ from
+%%     $\mathcal C$ to the category of Multicategories.
+%%   \end{enumerate}
+%% \end{definition}
+%% Here the objects of $\mathcal C$ model the non-linear contexts, and morphisms
+%% model non-linear substitutions. The category with families structure over
+%% $\mathcal C$ models the dependent non-linear types. Finally the
+%% ``hyperdoctrine'' of multicategories $L$ models the linear types and terms. The
+%% fact that this is functorial in $\mathcal C$ corresponds to the fact that linear
+%% types are all relative to a non-linear context and that linear variables and
+%% composition commute with non-linear substitution.
+
+%% We will have three main Lambek hyperdoctrines of interest in this paper: the
+%% ``standard model'' of sets and grammars, the ``syntactic model'' given by our
+%% type theory itself and lastly a gluing model introduced in
+%% Section~\ref{sec:canonicity} to prove the canonicity theorem.
+%% \begin{example}
+%%   \begin{enumerate}
+%%   \item Let $\mathcal C$ be the category of sets, equipped with its usual
+%%     category with families structures of families. Define $L : \mathcal C^{o}
+%%     \to \textrm{MultiCat}$ to map $L(X)$ to the representable multicategory
+%%     $(\Set^{\Sigma^*})^X$ of $X$-indexed families of grammars where the monoidal
+%%     structure is given pointwise by the Day convolution monoidal structure.
+%%   \item We can build a ``syntactic'' Lambek hyperdoctrine $\Syn(\Sigma)$ from the syntax
+%%     itself: the category $\mathcal C$ is given by non-linear contexts and
+%%     substitutions, the category with families by the non-linear types and terms
+%%     and the hyperdoctrine of multicategories by the linear types and terms.
+%%   \end{enumerate}
+%% \end{example}
+
+%% Next each type linear and non-linear type constructor corresponds to extra data
+%% on a Lambek hyperdoctrine.
+%% \begin{definition}
+%%   \label{def:chomsky-data}
+%%   \begin{enumerate}
+%%   \item Dependent type structure: standard (say extensional type theory, give a reference)
+%%   \item Universes
+%%   \item Inductive grammars?
+%%   \item Non-inductive Grammar constructors: standard monoidal category stuff
+%%   \end{enumerate}
+%% \end{definition}
+%% \begin{definition}[Chomsky Hyperdoctrine]
+%%   A Chomsky Hyperdoctrine is a Lambek hyperdoctrine equipped with all of the data in Definition~\ref{def:chomsky-data}.
+%% \end{definition}
+%% \pedro{We should probably define some of the words in this definition}
+%% \begin{definition}
+%%   A Chomsky category is a locally Cartesian category with two hierarchies of
+%%   universes $\{L_i\}_{i\in \nat}$ and $\{U_i\}_{i\in \nat}$ such that
+%%   every $L_i$ and $U_i$ are $U_{i+1}$-small. Furthermore, we require
+%%   $U_i$ to be closed under dependent products and sum,
+%%   $L_i$ to be closed under the Kleene category connectives,
+%%   dependent products, left and right closed structures, with
+%%   a type constructor $G : L_i \to U_i$ and a linear dependent sum
+%%   going the other direction.
+%% \end{definition}
+
+%% \steven{Max suggests augmenting the definition of a Chomsky category to
+%%   something like two categories $L$ and $U$, and $U$ is Cartesian closed I don't
+%%   fully recall the rest.
+
+%%   My best guess is that you take $L$ a Kleene category with a hierarchy of universes
+%%   two, and you further require that $L$ is $Psh(U)$-enriched, except he
+%%   suggested a further adjective on these presheaves. Perhaps representability?
+%% }
+
+%% \begin{theorem}
+%%   The presheaf category $\Grammar = \Set^{\cat{\Sigma^*}}$ is a Chomsky category.
+%% \end{theorem}
+
+%% Further, the syntactic category of \theoryname is manifestly a Chomsky category.
+
+%% \pedro{This is likely true, but if we explicitly say so, this warrants a proof. I think that
+%% if we don't say anything about the syntactic category, reviewers won't mind.}
+
+%% \steven{I agree that we don't want to say anything that opens unnecessary
+%%   questions for proofs we haven't written. However, it seems hard to make the
+%%   case that we have the right categorical model of the syntax if this statement
+%%   isn't true. By restating our definition of Chomsky category, this should be
+%%   obvious or a quick proof}
+
+%% \pedro{I agree, then we should add the quick proof :) }
+
+%% \subsection{Concrete Models of \theoryabbv}
+%% \label{sec:othermodels}
+
+%% \steven{Because we can define a version of semantic actions internally, we
+%%   shouldn't put it as a separate model}
+
+%% One of the powers of type theories is that it can be profitable to interpret them
+%% in various models. In this section, by using our just-defined Chomsky categories,
+%% we show how other useful concepts from formal language theory can also be organized
+%% as models of \theoryabbv. We illustrate this point by providing two examples that
+%% are closely related to the theory of formal languages: language equivalence and
+%% semantic actions. Furthermore, in order to justify how $\mathbf{Gr}$ relates to
+%% more traditional notions of parsing, we define a glued model that proves a
+%% canonicity property of grammar terms.
+
+%% \subsection{Language Semantics}
+%% Every grammar induces a language semantics. Also languages can be taken as a
+%% propositionally truncated view of the syntax. Logical equivalence should induce
+%% weak equivalence, and thus even give a syntactic way to reason about language equivalence.
+%% \steven{TODO language semantics}
+
+%% % \subsection{Semantic Actions}
+%% % \steven{Tentatively planning to cut this subsection for an internal representation
+%% %   of semantic actions}
+%% % Returning to the problem of parsing, the output of a parse usually is not the
+%% % literal parse tree. Rather, the output is the result of some \emph{semantic
+%% %   actions} ran on the parse tree, which usually serve to remove some syntactic
+%% % details that are unnecessary for later processing.
+
+%% % Given a grammar $G : \String \to \Set$, we define a semantic action to be a set
+%% % $X$ with a function $f$ that produces a semantic element from any parse of $G$.
+
+%% % \[
+%% %   f : \PiTy w \String {G w \to X}
+%% % \]
+
+%% % Further, semantic actions can be arranged into a structured category.
+%% % Define $\SemAct$ as the comma
+%% % category $\Grammar / \Delta$, where $\Delta : \Set \to \Grammar$ defines a
+%% % discrete presheaf. That is, for a set $X$, $\Delta (X)(w) = X$ for all
+%% % $w \in \String$. As $\SemAct$ is defined as a comma category, it has a forgetful
+%% % functor into $\Grammar$. That is, $\SemAct$ serves as a notion of formal
+%% % grammar. Moreover, $\SemAct$ is a model of \theoryabbv.
+
+%% % \steven{It being a notion of formal grammar is distinct from being a model. This
+%% % probably warrants a proof}
+
+%% % \steven{TODO semantic actions}
+
+%% % \pedro{This is a very nice opportunity of showing off the supremacy of denotational
+%% %   reasoning ;) We should probably prove the gluing lemma in the previous section
+%% %   and apply it here and in the canonicity section. The actual proof might have to be
+%% %   moved to the appendix, though}
+
+%% \subsection{Parse Canonicity}
+%% Canonicity is an important metatheoretic theorem in the type theory
+%% literature.  It provides insight on the normal forms of terms and,
+%% therefore, on its computational aspects. Frequently, proving
+%% canonicity for boolean types, i.e. every closed term of type bool
+%% reduces to either true or false, is enough to justify that the type
+%% theory being studied is well-behaved. In our case, however, since we
+%% want to connect \theoryabbv to parsers, we must provide a more
+%% detailed account of canonicity. In particular, we give a nonstandard semantics
+%% of \theoryabbv that carries a proof of canonicity along with it.
+
+%% If $\cdot \vdash A$ is a closed linear type then there are
+%% two obvious notions of what constitutes a ``parse'' of a string w
+%% according to the grammar $A$:
+%% \begin{enumerate}
+%% \item On the one hand we have the set-theoretic semantics just
+%%   defined, $\llbracket A \rrbracket \cdot w$
+%% \item On the other hand, we can view the string $w = c_1c_2\cdots$ as
+%%   a linear context $\lceil w \rceil = x_1:c_1,x_2:c_2,\ldots$ and
+%%   define a parse to be a $\beta\eta$-equivalence class of linear terms $\cdot;
+%%   \lceil w \rceil \vdash e : G$.
+%% \end{enumerate}
+%% It is not difficult to see that at least for the ``purely positive''
+%% formulae (those featuring only the positive connectives
+%% $0,+,I,\otimes,\mu, \overline\Sigma,c$) that
+%% every element $t \in \llbracket A \rrbracket w$ is a kind of tree and
+%% that the nodes of the tree correspond precisely to the introduction
+%% forms of the type. However it is far less obvious that \emph{every}
+%% linear term $\lceil w \rceil \vdash p : \phi$ is equal to some
+%% sequence of introduction forms since proofs can include elimination
+%% forms as well. To show that this is indeed the case we give a
+%% \emph{canonicity} result for the calculus: that the parses for .
+
+%% \begin{definition}
+%%   A non-linear type $X$ is purely positive if it is built up using
+%%   only finite sums, finite products and least fixed points.
+
+%%   A linear type is purely positive if it is built up using only finite
+%%   sums, tensor products, generators $c$, least fixed points and linear
+%%   sigma types over purely positive non-linear types.
+%% \end{definition}
+
+%% \begin{definition}
+%%   %% Let $X$ be a closed non-linear type. The closed elements $\textrm{Cl}(X)$ of $X$ are the definitional equivalence classes of terms $\cdot \vdash e : X$.
+
+%%   Let $A$ be a closed linear type. The nerve $N(A)$ is a presheaf on
+%%   strings that takes a string $w$ to the definitional equivalence
+%%   classes of terms $\cdot; \lceil w\rceil \vdash e: N(A)$.
+%% \end{definition}
+
+%% \begin{theorem}[Canonicity]
+%%   Let $A$ be a closed, purely positive linear type. Then there is an
+%%   isomorphism between $\llbracket A\rrbracket$ and $N(A)$.
+%% \end{theorem}
+%% \begin{proof}
+%%   We outline the proof here, more details are in the appendix. The
+%%   proof proceeds first by a standard logical families construction
+%%   that combines canonicity arguments for dependent type theory
+%%   TODO cite coquand
+%%   % \cite{coquand,etc}
+%%   with logical relations constructions for linear
+%%   types
+%%   TODO cite hylandschalk
+%%   % \cite{hylandschalk}
+%%   . It is easy to see by induction that the
+%%   logical family for $A$, $\hat A$ is isomorphic to $\llbracket A
+%%   \rrbracket$ and the fundamental lemma proves that the projection
+%%   morphism $p : \hat A \to N(A)$ has a section, the canonicalization
+%%   procedure. Then we establish again by induction that
+%%   canonicalization is also a retraction by observing that introduction
+%%   forms are taken to constructors.
+%% \end{proof}
+
+%% \begin{enumerate}
+%% \item Every term $\lceil w \rceil \vdash p : G + H$ is equal to $\sigma_1q$ or $\sigma_2 r$ (but not both)
+%% \item There are no terms $\lceil w \rceil \vdash p : 0$
+%% \item If there is a term $\lceil w \rceil \vdash p : c$ then $w = c$ and $p = x$.
+%% \item Every term $\lceil w \rceil \vdash p : G \otimes H$ is equal to $(q,r)$ for some $q,r$
+%% \item Every term $\lceil w \rceil \vdash p : \varepsilon$ is equal to $()$
+%% \item Every term $\lceil w \rceil \vdash p : c$ is equal to $x:c$
+%% \item Every term $\lceil w \rceil \vdash p : \mu X. G$ is equal to $\textrm{roll}(q)$ where $q : G(\mu X.G/X)$
+%% \item Every term $\lceil w \rceil \vdash p : (x:A) \times G$ is equal
+%%   to $(M,q)$ where $\cdot \vdash M : A$
+%% \end{enumerate}
+
+%% To prove this result we will use a logical families model. We give a
+%% brief overview of this model concretely:
+%% \begin{enumerate}
+%% \item A context $\Gamma$ denotes a family of sets indexed by closing substitutions $\hat\Gamma : (\cdot \vdash \Gamma) \Rightarrow \Set_i$
+%% \item A type $\Gamma \vdash X : U_i$ denotes a family of sets $\hat X : \Pi(\gamma:\cdot \vdash \Gamma) \hat\Gamma \Rightarrow (\cdot \vdash \simulsubst X \gamma) \Rightarrow \Set_i$
+%% \item A term $\Gamma \vdash e : X$ denotes a section $\hat e : \Pi(\gamma)\Pi(\hat\gamma)\hat X \gamma \hat\gamma (\simulsubst e \gamma)$
+%% \item A linear type $\Gamma \vdash A : L_i$ denotes a family of grammars $\hat A : \Pi(\gamma:\cdot\vdash\Gamma)\,\hat\Gamma \Rightarrow \Pi(w:\Sigma^*) (\cdot;\lceil w\rceil \vdash A[\gamma])\Rightarrow \Set_i$, and the denotation of a linear context $\Delta$ is similar.
+%% \item A linear term $\Gamma;\Delta \vdash e : A$ denotes a function \[\hat e : \Pi(\gamma)\Pi(\hat\gamma)\Pi(w)\Pi(\delta : \lceil w \rceil \vdash \simulsubst \Delta \gamma) \hat\Delta \gamma \hat\gamma \delta \Rightarrow \hat A \gamma \hat\gamma w {(\simulsubst {\simulsubst e \gamma} \delta)}\]
+%% \end{enumerate}
+%% And some of the constructions:
+%% \begin{enumerate}
+%% \item $\widehat {(G A)} \gamma \hat\gamma e = \hat A \gamma \hat\gamma \varepsilon (G^{-1}e)$
+%% \item $\widehat {(A \otimes B)} \gamma \hat\gamma w e = \Sigma(w_Aw_B = w)\Sigma(e_A)\Sigma(e_B) (e_A,e_B) = e \wedge \hat A \gamma \hat \gamma w_A e_A \times \hat B \gamma \hat \gamma w_B e_B$
+%% \item $\widehat {(A \lto B)} \gamma \hat\gamma w e = \Pi(w_A)\Pi(e_A) \hat A \gamma\hat\gamma w_A e_A \Rightarrow \hat B \gamma\hat\gamma (ww_A) (\applto {e_A} e)$
+%% \end{enumerate}
+
+%% First, the category with families will be
+%% the category of logical families over set contexts/types
+%% $\Delta$/$A$. Then the propositional portion will be defined by
+%% mapping a logical family $\hat \Gamma \to \Gamma$
+
+%% First, let $L$ be the category of BI formulae and proofs (quotiented
+%% by $\beta\eta$ equality). Define a functor $N : L \to \Set^{\Sigma^*}$ by
+%% \[ N(\phi)(w) = L(w,\phi) \]
+
+%% Then define the gluing category $\mathcal G$ as the comma category
+%% $\Set^{\Sigma^*}/N$. That is, an object of this category is a pair of
+%% a formula $\phi \in L$ and an object $S \in \mathcal
+%% P(\Sigma^*)/N(\phi)$. We can then use the equivalence $\mathcal
+%% P(\Sigma^*)/N(\phi) \cong \mathcal P(\int N(\phi))$ to get a simple
+%% description of such an $S$: it is simply a family of sets indexed by
+%% proofs $L(w,\phi)$:
+%% \[ \prod_{w\in\Sigma^*} L(w,\phi) \to \Set \]
+%% This category clearly comes with a projection functor $\pi : \mathcal
+%% G \to \mathcal L$ and then our goal is to define a section by using
+%% the universal property of $\mathcal L$.
+
+%% To this end we define
+%% \begin{enumerate}
+%% \item $(\phi, S) \otimes (\psi, T) = (\phi \otimes \psi, S\otimes T)$ where
+%%   \[ (S \otimes T)(w, p) = (w_1w_2 = w) \times (q_1,q_2 = p) \times S\,w_1\,q_1 \times T\,w_2\,q_2\]
+%% \item $(\phi, S) \multimap (\psi, T) = (\phi \multimap \psi, S \multimap T)$ where
+%%   \[ (S \multimap T)(w,p) = w' \to q \to S\,w'\,q \to T (ww') (p\,q) \]
+%% \item $\mu X. ??$ ??
+%% \end{enumerate}
+
+%% \pedro{We should conclude this section by explaining the relevance of the canonicity
+%% theorem. Could also be done before stating the theorem.}
 

### 2024-11-13 paper/paper.tex
added some comments and references

@@
 \usepackage{xspace}
+
+\usepackage[LGR,T1]{fontenc}
+\DeclareMathAlphabet{\mathgtt}{LGR}{cmtt}{m}{n}
+
+\renewcommand{\Gamma}{\mathgtt{G}}
+\renewcommand{\Delta}{\mathgtt{D}}
+\renewcommand{\Sigma}{\mathgtt{S}}
+\renewcommand{\mu}{\mathgtt{m}}
+
+\renewcommand{\familydefault}{\ttdefault}
+\usepackage{mathastext}
+\renewcommand{\familydefault}{\rmdefault}
+
+\definecolor{codegreen}{rgb}{0,0.6,0}
+\definecolor{codegray}{rgb}{0.5,0.5,0.5}
+\definecolor{codepurple}{rgb}{0.58,0,0.82}
+\definecolor{backcolour}{rgb}{0.95,0.95,0.92}
+
 \usepackage{listings}
+\lstdefinelanguage{Agda}{
+keywords={data, where, module, import, open, public,
+record, field, let, in, if, then, else, case, of, with,
+do, postulate, primitive, mutual, abstract, private,
+forall, exists, cong, set, prop, sort, Level, Data, Type,
+Renamer},
+morekeywords=[2]{Set, Type, Prop, moduleFuncs, instance, Named},
+sensitive=true,
+comment=[l]{--},
+morecomment=[s]{\{-}{-\}},
+morestring=[b]",
+mathescape=true,
+escapeinside={(*@}{@*)}
+}
+\lstset{
+language=Agda,
+basicstyle=\ttfamily\small,
+keywordstyle=\color{blue},
+keywordstyle=[2]\color{teal},
+identifierstyle=\color{black},
+commentstyle=\color{gray}\textit,
+stringstyle=\color{orange},
+numbers=none,
+numberstyle=\tiny\color{gray},
+stepnumber=1,
+numbersep=5pt,
+showstringspaces=false,
+tabsize=4,
+captionpos=b,
+breaklines=true,
+breakatwhitespace=false,
+rulecolor=\color{black},
+texcl=true,
+backgroundcolor=\color{backcolour},
+frame=single
+}
+
 \usepackage{hyperref}
-\usepackage{cleveref}
+\usepackage[capitalise]{cleveref}
 \usepackage{stmaryrd}
@@
 
+\newcommand{\Subst}{\textrm{Subst}}
 \newcommand{\SPF}{\mathsf{SPF}}
@@
 \newcommand{\String}{\textbf{String}}
-\newcommand{\stringg}{\mathsf{string}}
-\newcommand{\charg}{\mathsf{char}}
+\newcommand{\stringg}{\texttt{String}}
+\newcommand{\charg}{\mathtt{Char}}
 \newcommand{\Set}{\mathbf{Set}}
@@
 \newcommand{\Prop}{\mathbf{Prop}}
-\newcommand{\Bool}{\mathbf{Bool}}
-\newcommand{\true}{\mathbf{true}}
-\newcommand{\false}{\mathbf{false}}
+\newcommand{\Bool}{\mathtt{Bool}}
+\newcommand{\true}{\mathtt{true}}
+\newcommand{\false}{\mathtt{false}}
 \newcommand{\nat}{\mathbb{N}}
 \newcommand{\theoryname}{Dependent Lambek Calculus\xspace}
-\newcommand{\theoryabbv}{$\textrm{Lambek}^D$}
+\newcommand{\theoryabbv}{$\textrm{Lambek}^D$~}
 \newcommand{\lnld}{$\textrm{LNL}_D$}
@@
 \newcommand{\ltonl}[1]{~\uparrow #1}
-\newcommand{\nil}{\mathsf{nil}}
-\newcommand{\cons}{\mathsf{cons}}
-\newcommand{\epscons}{\varepsilon\mathsf{cons}}
+\newcommand{\nil}{\texttt{nil}}
+\newcommand{\ident}{\texttt{id}}
+\newcommand{\cons}{\texttt{cons}}
+\newcommand{\epscons}{\varepsilon\texttt{cons}}
 \newcommand{\data}{\mathsf{data}}
 \newcommand{\where}{\mathsf{where}}
-\newcommand{\Trace}{\mathsf{Trace}}
-\newcommand{\stringquote}[1]{\texttt{\textquotesingle#1\textquotesingle}}
+\newcommand{\Trace}{\texttt{Trace}}
+\newcommand{\literal}[1]{\texttt{\textquotesingle#1\textquotesingle}}
+\newcommand{\stringquote}[1]{\texttt{\textquotedbl#1\textquotedbl}}
 \newcommand{\internalize}[1]{\lceil#1\rceil}
@@
 
+\newsavebox{\logoagdabox}
+\sbox{\logoagdabox}{%
+  %
+  \raisebox{-2pt}{\includegraphics[height=1em]{logo-agda.pdf}}%
+}
+\newcommand{\agdalogo}{%
+  \usebox{\logoagdabox}}%
+
 \newif\ifdraft
@@
 
+\author{Pedro H. Azevedo de Amorim}
+\affiliation{
+  \department{Department of Computer Science}
+  \institution{University of Oxford}
+  \country{UK}
+}
+\email{pedro.azevedo.de.amorim@cs.ox.ac.uk}
+
 \author{Max S. New}
@@
 \email{maxsnew@umich.edu}
-
-\author{Pedro H. Azevedo de Amorim}
-\affiliation{
-  \department{Department of Computer Science}
-  \institution{University of Oxford}
-  \country{UK}
-}
-\email{pedro.azevedo.de.amorim@cs.ox.ac.uk}
 
@@
   embedding in Agda.
-
-  Our shallow embedding is based on a denotational semantics of \theoryabbv
-  where non-linear types denote ordinary sets and linear types denote formal
-  grammars. To develop this semantics compositionally, we define a notion of
-  general notion of model we call a \emph{Chomsky hyperdoctrine}, a
-  generalization of the Kleene algebra approach to formal language theory. In
-  addition to our ``standard model'' where linear types denote formal grammars,
-  we demonstrate two non-standard models. The first interprets linear types as
-  semantic actions over formal grammars, suggesting future extensions of our
-  calculus. The second is used to structure a logical relations argument to
-  prove a canonicity metatheorem for the type theory, proving that all closed
-  terms of context-free linear types normalize to a parse tree.
 \end{abstract}
@@
 \label{sec:intro}
-
-%% Outline
-%% - Why a framework for parser verification
-%% - what intrinsic verification gets us
-%% - the secret sauce: indexed linear inductive types
-%% - prototype implementation
-%% - overview of "abstract" formal grammars?
-
-Parsing data from a serialized format is one of the most common tasks
-in all of computing. Accordingly, this is a quite well-studied problem
-spanning theoretical computer science and linguistics with such
-milestones as Chomsky's hierarchy of grammars
-\cite{chomThreeModels1956}, practical algorithms for parsing of
-regular and context-free grammars, and variants
-\cite{KNUTH1965607,Earley1970}, as well as implementations of
-practical tools for the generation of efficient parsers
-\cite{Johnsonyacc}.
-
-The theory of formal languages and parsing is one of the oldest and most
-thoroughly developed area of theoretical computer science.
-The central object of study is a \emph{formal language} $L$ over an alphabet
-$\Sigma$, which is classically defined as a
-\emph{subset} of strings $L \subseteq \String = \Sigma^{*}$. The expressivity of formal
-languages is sufficient for defining a \emph{recognizer} of a language $L$:
-a function $r : \String \to \Prop$ such that $r$ maps each string $w$ to the
-proposition that $w$ is in the language $w \in L$. This technique of language
-recognition can concisely specify lexical analysis, and even lead to verified
-lexer implementations \cite{egolfVerbatim,Ouedraogo_2023}.
-
-While there are ample theory and tooling to handle the first stage of a language frontend, the final
-frontier for parsing research is the development of
-\emph{formally verified} parser implementations. Due to the
-ubiquitous nature of parsing, these would useful components of many
-formally verified software systems: compilers and servers most
-especially. However, language recognition is insufficient for specifying a
-parser, and for this task much of the structured foundations from the theory of
-formal languages is rendered moot. Rather than returning a proposition-valued result, a parser produces a
-structured \emph{parse tree} adherent to a \emph{formal grammar}.
-The research on such grammars is often performed atop an ad-hoc choice of grammar
-formalism, and results are proven with respect to that formalism. Critically,
-the field is lacking a unifying theory across these different formalisms, even
-though many of the same
-compositional reason principles appear across distinct grammar formalisms.
-
-In this work, we aim to elucidate these similarities and
-provide a mathematical theory of formal grammars for constructing verified
-parsers. Our approach can be summarized with the
-Curry-Howard-style slogan: ``grammars as types''; and the addendum: ``parsers as
-terms''. That is, we develop a type theory with a novel
-denotational semantics in a category with abstract formal grammars as
-objects and \emph{parse transformers} as morphisms. We say these are
-\emph{abstract} formal grammars because they are not tied to any
-particular syntactic formalism such as Chomskyan grammars, regular
-expressions, Lambek calculus etc. The category we choose is a rich
-in structure: it is a topos as well as being monoidal closed, and these
-universal properties serve as the inspiration for our type theory.
-
-The type theory itself is a form of dependent linear-non-linear type
+Formal verification encompasses a broad suite of techniques one may use to
+mathematically guarantee the correctness of software systems. We can only
+reap the benefits of
+any proof effort if its underlying assumptions hold. For instance, verifying the
+functional correctness of an algorithm is no use if it is run on buggy hardware.
+Likewise, the promised guarantees of a program may be jeopardized if it is
+parsed incorrectly.
+
+Large verified developments, like CompCert \cite{leroy_formal_2009},
+invest a significant number of person-hours into proof development. In early
+versions of CompCert, the apparent
+assurances associated with their effort were at risk of invalidation by
+parser bugs \cite{yangFindingUnderstandingBugs}. Parsing is a ubiquitous task
+throughout computing and yet parsers are frequently unverified, even in
+otherwise verified systems. The CompCert team later
+developed a method for validating the correctness of LR parsers
+\cite{jourdanValidatingLRParsers2012} and then replaced their parser with one
+that is certified to be correct. Despite their success in  validating their C parser and other work in the
+generation of  verified parsers
+\cite{lasserCoStarVerifiedALL2021,firsovCertifiedNormalizationContextFree2015,danielssonTotalParserCombinators2010},
+it is still arduous and highly nontrivial to write a verified
+parser. Further, approaches are sensitive to a choice of grammar and automata formalism to serve
+as a
+foundation, which can make it difficult to relate and interoperate different
+styles of verified parsers.
+
+In \cite{elliottSymbolicAutomaticDifferentiation2021}, Elliott describes formal
+grammar as type-level predicates on strings that prove
+language membership. That is, a \emph{formal grammar} $A$ is a function
+$\String \to \Set$ such that for a string $w$, $A~w$ is the set of proofs
+showing that $w$ belongs to the language recognized by $A$. This typed approach
+to parsing serves a general setting that can simultaneuously interpret many distinct
+grammar formalisms. In this paper, we adopt Elliott's view on formal grammars and present a
+syntax that axiomatizes this semantic notion of grammar.
+
+We present \theoryname (\theoryabbv) as a domain specific
+type theory for building verified parsers. The type theory itself is a form of dependent linear-non-linear type
 theory: the linear types can be interpreted as grammars, whereas the
@@
 
-\paragraph{Contributions} This paper begins in \cref{sec:synindnotion} by
-giving a universal semantic characterization of formal grammar.
+Parsers written in \theoryabbv take on a linear functional style, which makes
+them familiar to write and amenable to compositional verification techniques. We
+implement \theoryabbv as a shallowly embedded domain-specific langauge in the
+Cubical Agda proof assistant. Throughout this paper, we will use \agdalogo~ to
+mark results that are mechanized in our Agda development.
+
+\paragraph{Contributions} This paper begins in \cref{sec:type-theory-examples}
+by studying small example programs from \theoryabbv to build intuition.
 From there, in \cref{sec:tt} we provide the syntax of \theoryname, a type theory
 that axiomatizes the underlying structures needed to conduct formal grammar
-theory. In \cref{sec:automata} we demonstrate the usage of
-\theoryname for defining parser terms and internalizing results from formal language theory.
-Then in \cref{sec:categorify}, we characterize the abstract categorical
-models of our syntax. Afterwards in \cref{sec:othermodels}, we demonstrate several models,
-including one to prove a metatheoretic canonicity theorem.
+theory. Later in \cref{sec:applications} we demonstrate the
+applicability of \theoryabbv for relating familiar grammar and automata
+formalisms as well as building concrete parsers.
+Then in \cref{sec:semantics-and-metatheory}, we give a denotational semantics
+that makes precise the connection between \theoryabbv syntax and formal grammars.
+
 Our contributions are then:
@@
 \begin{itemize}
-  \item A syntax-independent notion of grammar given by the category $\Grammar$
-  \item \theoryname: A dependent
-    linear-non-linear type theory meant to syntactically capture the latent structures
-    in $\Grammar$ that enable parsing
-  \item Categorical structures that capture the concepts integral to formal grammar theory, and that are notably modeled by $\Grammar$
-  \item An implementation of the language, which is embedded shallowly in Agda
+  \item \theoryname (\theoryabbv): A dependent
+        linear-non-linear type theory for building verified parsers
+  \item A denotational grammars semantics for \theoryabbv
+  \item An implementation of \theoryabbv in Agda
 \end{itemize}
-
-\subsection{A Syntax-Independent Notion of Syntax}
-\label{sec:synindnotion}
-
-The notion of formal language is central to the theory of parsing.
-This definition is
-especially useful as it gives a semantics to formal grammars that is completely
-independent of any particular syntactic grammar formalism. Any new
-notion of grammar can be given a language semantics and it provides a precise
-mathematical specification for implementing a \emph{recognizer} of a language.
-In this section we propose a \emph{proof-relevant} generalization of formal
-languages that can accommodate richer aspects of formal language theory such as
-parsing. We begin by proposing a presheaf approach to grammars that encompasses
-other classes of grammars in a syntax-independent way.
-
-\paragraph{From Formal Languages to Formal Grammars}
-
-Our first contribution is a novel and syntax-independent characterization of
-formal grammar:
-
-\begin{definition}
-  \label{def:grammar}
-  A \emph{formal grammar} $G$ over a fixed alphabet $\Sigma$ is a function
-  $G : \String \to \Set$.
-\end{definition}
-
-\steven{TODO need to go through and ensure that $\Sigma^{*}$ and $\String$ are each used appropriately}
-
-We say that a grammar $G$ associates to every string $w$ the set\footnote{We
-  make little comment as to which foundations are used to encode $\Set$. Our
-  construction is polymorphic in choice of a proper class of small sets, a
-  Grothendieck universe, a type theoretic universe, or any similar foundation.} $G w$ of parse
-trees showing that $w$ matches $G$. This definition of formal grammar serves as
-a mathematical specification for a parser --- taking on the same role as
-languages do for recognizers. In this sense, our new characterization of
-grammars is just a proof-relevant generalization of the existing notion of
-formal language.
-
-By ranging over all input strings, notice that each grammar induces a language.
-For each formal grammar $G$, define the language of accepted strings $L_{G}$ as,
-
-\[ L_{G} = \{ w \in \String ~|~ G w~\text{is inhabited} \} \]
-
-As they are simply sets, formal languages are naturally endowed with a partial
-order via subset inclusion. Just, formal grammars naturally coalesce into a
-\emph{category}.
-
-Define $\Grammar$ to be the category of formal grammars. Whose
-objects are grammars, and whose morphisms are functions $f : G \to H$ between two
-grammars presented by a family of functions $f^{w} : G w \to H w$ for
-$w : \String$. Intuitively, we interpret these morphisms as parse
-\emph{transformers} --- as constructive translations of $G$-parses into $H$-parses. Throughout this paper, we will make frequent reference to this categorical
-structure on $\Grammar$.
-
-\steven{Above sounds awkward. I want to additionally say something here that signposts that we will give more
-  information about that structure without just listing it all here. }
-
-\paragraph{From Grammars to Formal Grammars}
-The category-theoretic framework can be further used to describe and compare
-different notions of formal grammar. For instance, Chomskyan generative
-grammars, semi-Thue systems, Montague grammars, and so on are all distinct syntactic
-presentations of the same underlying idea of an abstract specification for
-parsing. Each of these is an instance of a general \emph{notion of formal
-  grammar} --- a category paired with a functor into
-$\Grammar$. That is, they may all be interpreted in our category of grammars. In
-this sense, $\Grammar$ provides an
-all-encompassing semantic domain for parsing.
-
-Upon further inspection, $\Grammar$ may equivalently
-be described as the functor category $\Set^{\String}$ --- where the
-set $\String$ is viewed as a discrete\footnote{Objects in $\String$
-comprise the set of all strings, and the only morphisms are the
-identity morphisms.} category. In fact, in this paper we will take
-$\Grammar$ to be defined as the functor category $\Set^{\String}$. Such functor
-categories in $\Set$, called presheaf categories, carry a
-remarkable amount of structure.  Indeed they are a monoidal topos and,
-as such, are a model to a substructural dependently typed theory.
-
-Thus we
-present \theoryname (\theoryabbv) as a domain-specific type theory that
-abstracts over the structure of $\Grammar$. To begin, let us look at what some
-programs in \theoryabbv look like.
-
-\steven{TODO Reword this transition into the examples}
-
-% \max{this section kind of trails off at the end with a very vague aside about a $2$-category of grammar formalisms.
-Maybe this should go in future work/discussion?
-What is the message of this section?
-How does it tie in with the message of the paper?
-What even is the message of this paper?}
-
-% \pedro{Would building on this proposed structure make the message of the section clearer? In this case
-%   we would have to be a bit clearer in terms of how familiar notions of grammar can be seen as categories with functor into $\Grammar$.}
 
@@
 \label{sec:type-theory-examples}
-
-\steven{To save space, each of these example figures can be
-  minified/potentially inlined}
-
-To gain intuition for working in \theoryabbv, we study some examples of
-grammars and parses of them internal to \theoryabbv. In each of our examples, we will look at
-grammars over the three character alphabet $\{ a , b , c\}$. Each letter of the
-alphabet is a base type in our calculus. For instance, the grammar
-for the single character $a$ has a single parse tree for the string \stringquote{a},
-and no parse trees at any other strings.
+To gain intuition for working in \theoryabbv, we begin with some illustrative
+examples drawn from the theory of formal languages. In particular, we
+demonstrate how to encode finite grammars, regular expressions, and finite
+automata in \theoryabbv. Each of our examples will be defined for strings over the three character alphabet
+$\Sigma = \{ \texttt{a} , \texttt{b}, \texttt{c} \}$.
+
+\newcommand{\A}{\texttt{A}}
+\newcommand{\B}{\texttt{B}}
+\newcommand{\I}{\texttt{I}}
+\newcommand{\f}{\texttt{f}}
+\newcommand{\g}{\texttt{g}}
+\renewcommand{\L}{\texttt{L}}
+\renewcommand{\a}{\texttt{a}}
+\renewcommand{\b}{\texttt{b}}
+\renewcommand{\c}{\texttt{c}}
+\newcommand{\w}{\texttt{w}}
 
 \paragraph{Finite Grammars}
-\max{epsilon? bottom?}
-First consider finite grammars --- those built from base types via disjunctions $\oplus$ and
-concatenations $\otimes$. For instance, $(a \otimes b) \oplus c$.
-In \cref{fig:fingram} we find a proof tree that constructs a term of type
-$(a \otimes b) \oplus c$ in the context $x : a, y : b$. We think of this term as
-a parse tree internal to \theoryabbv.
-
-For a grammar $A$, a parse tree of a string $w$ can be represented as
-a term of type $A$ in the context $\internalize w$, where
-$\internalize w$ is a context with one variable for each character of
-$w$. For example, to define a parse tree for \stringquote{ab}, we use
-the context $\internalize{\stringquote{ab}} = x : a , y : b$. Then
-\cref{fig:fingram} shows a typing derivation that shows that $\inl(x
-\otimes y)$ defines a valid parse tree for the grammar $(a \otimes b)
-\oplus c$.
+First consider finite grammars --- those built from base types via disjunctions and
+concatenations. The base types comprise characters drawn from the alphabet, the
+empty string, and the empty grammar.
+For instance, the grammar corresponding to the character $a$ is given by $\literal a$ which
+has a single parse tree for the string
+\stringquote{a} and no parse trees at any other strings. The grammar $\I$ recognizes has a single
+parse tree for the empty string $\epsilon = \stringquote{}$ and no parses for any other strings.
+The final base type, the empty grammar $\bot$, has no parses for any string. We
+use type-theoretic syntax to represent disjunction $\oplus$ and concatenation
+$\otimes$ of
+grammars. Over an input string $\w$, a parse of the disjunction $\A \oplus \B$ is either
+a parse of $\A$ over the string $\w$ or a
+parse of $\B$ over the string $\w$. Similarly, $\w$ matches $\A \otimes \B$ if
+$\w$ can be split into two strings $\w_{\A}$ and $\w_{\B}$ that match $\A$
+and $\B$, respectively.
+
+For a grammar $\A$, a parse tree of a string $\w$ can be represented as
+a term of type $\A$ in the context $\internalize \w$, where
+$\internalize \w$ is a context with one variable for each character of
+$\w$. For example, to define a parse tree for \stringquote{ab}, we use
+the context $\internalize{\stringquote{ab}} = a : \literal a , b : \literal b$. Then the left of
+\cref{fig:fingram} gives a typing derivation that shows that $\inl(a
+\otimes b)$ defines a valid parse tree for the finite grammar
+$(\literal a \otimes \literal b)
+\oplus \literal c$. Underneath the derivation in \cref{fig:fingram}, we give a linear $\lambda$-term
+that implements the program given in the derivation.
 
@@
       {~}
-      {x : a \vdash x : a}
+      {a : \literal a \vdash a : \literal a}
       \\
@@
       {~}
-      {y : b \vdash y : b}
+      {b : \literal b \vdash b : \literal b}
     }
-    {x : a , y : b \vdash x \otimes y : a \otimes b}
+    {a : \literal a , b : \literal b \vdash a \otimes b : \literal a \otimes
+      \literal b}
   }
-  {x : a , y : b \vdash \inl(x \otimes y) : (a \otimes b) \oplus c}
+  {a : \literal a , b : \literal b \vdash \texttt{f} := \inl(a \otimes b) :
+    (\literal a \otimes \literal b) \oplus \literal c}
 \end{mathpar}
-\caption{\stringquote{ac} matches $(a \oplus b) \otimes c$}
+\begin{lstlisting}
+f : (*@$\uparrow$@*)('a' (*@$\otimes$@*) 'b' (*@$\lto$@*) ('a' (*@$\otimes$@*) 'b') (*@$\oplus$@*) 'c')
+f (a , b) = inl (a (*@$\otimes$@*) b)
+\end{lstlisting}
+\caption{\stringquote{ac} matches
+  $(\literal a \oplus \literal b) \otimes \literal c$}
 \label{fig:fingram}
 \end{figure}
-
-\max{I think concrete examples here would be helpful}
 For this interpretation of terms as parse trees to make sense, our
@@
 variable to be used twice, but in a parse tree, every character must
-be accounted for exactly once.
-%
-Finally, exchange allows for variables in the context to be reordered,
-but the ordering of characters in a string cannot be ignored in
-parsing: otherwise we would have that $\inl(x\otimes y)$ is a valid
-term in context $y:b,x:a$, but there should be no parses of $ba$ in
-this grammar.
-
-Because derivations in a linear context correspond to parse trees in this
-manner, we cannot affirm all of the usual structural rules of a type system.
-Weakening and contraction each allow for modification of the string in context ---
-weakening allows for characters to be dropped at will, while contraction allows
-for duplication of characters. Omission of these rules offers a
-resource-sensitive view that ensures only the inputted characters are parsed.
-Moreover, strings are not simply sets of characters, as they carry an inherent
-ordering. Thus, the rule of exchange is also omitted in this calculus so that
-the constituent characters of a string are not freely permuted. Without the above structural rules we
-arrive at an ordered, linear type system.
+be accounted for exactly once. That is, we want to prevent the following
+erroneous derivations,
+\[
+  a : \literal a , b : \literal b \not \vdash a : \literal a \qquad a : \literal a , b : \literal b \not \vdash (a, a) : \literal a \otimes \literal a
+\]
+
+%
+Finally, the ordering of characters in a string cannot be ignored while
+parsing, so we omit the exchange rule because it would allow
+for variables in the context to be reordered,
+\[
+  a : \literal a , b : \literal b \not \vdash (b , a) : \literal b \otimes \literal a
+\]
+% otherwise we would have that $\inl(x\otimes y)$ is a valid
+% term in context $y:b,x:a$, but there should be no parses of $ba$ in
+% this grammar.
 
 \paragraph{Regular Expressions}
-Regular expressions can be encoded as types using base types,
-$\oplus$, and $\otimes$ and the Kleene star $(\cdot)^{\ast}$.  For a
-grammar $A$, we can define the Kleene star $A^{*}$ as a particular
+Regular expressions can be encoded as types generated by base types,
+$\oplus$, and $\otimes$, and the Kleene star $(\cdot)^{\ast}$.  For a
+grammar $\A$, we can define the Kleene star $\A^{*}$ as a particular
 \emph{inductive linear type} of linear lists, as shown in
-\cref{fig:kleenestarinductive}. Here $A^{*} : L$ means we are defining
-a \emph{linear} type. $A^{*}$ has two constructors: $\nil$, which
-builds a parse of type $A^{*}$ from nothing; and $\cons$, which
-linearly consumes a parse of $A$ and a parse of $A^{*}$ and builds a
-parse of $A^{*}$. This linear consumption is defined by the linear
-function type $\ltonl$. The arrow, $\uparrow$, wrapping these
-constructors means that the constructors by themselves do not consume
-any resources, and so are \emph{non-linear} values themselves.
+\cref{fig:kleenestarinductive}. Here $\A^{*} : \L$ means we are defining
+a \emph{linear} type. $\A^{*}$ has two constructors: $\nil$, which
+builds a parse of type $\A^{*}$ from nothing; and $\cons$, which
+linearly consumes a parse of $\A$ and a parse of $\A^{*}$ and builds a
+parse of $\A^{*}$. This linear consumption is defined by the linear
+function type $\lto$. The linear function type $\A \lto \B$ defines functions that
+take in parses of $\A$ as input, \emph{consume} the input, and return a parse of
+$\B$ as output. The arrow, $\uparrow$, wrapping these
+constructors means that the constructors by themselves are not consumed
+upon usage, and so are \emph{non-linear} values themselves. That is, the
+names $\nil$ and $\cons$ are function symbols that may be reused as many times
+as we wish.
 
 \begin{figure}
-\begin{align*}
-\data &~A^{*} : L~\where\\
-      & \nil : \ltonl {A^{*}} \\
-      & \cons : \ltonl {(A \lto A^{*} \lto A^{*})}
-\end{align*}
+\begin{lstlisting}
+data A(*@$^*$@*) : L where
+  nil : (*@$\uparrow$@*)(A(*@$^*$@*))
+  cons : (*@$\uparrow$@*)(A (*@$\lto$@*) A(*@$^*$@*) (*@$\lto$@*) A(*@$^*$@*))
+\end{lstlisting}
+% \begin{align*}
+% \data &~A^{*} : L~\where\\
+%       & \nil : \ltonl {A^{*}} \\
+%       & \cons : \ltonl {(A \lto A^{*} \lto A^{*})}
+% \end{align*}
 \caption{Kleene Star as an inductive type}
@@
 
-In \cref{fig:kleenestarderivation}, we demonstrate that
-\stringquote{ab} matches $(a^{*} \otimes b) \oplus c$ through repeated
-application of the Kleene star constructors. In this proof tree we
-expand some syntactic details that are often left implicit. For
-instance, every linear context is parameterized by a nonlinear
-context, which is syntactically separated by a semicolon. If the
-nonlinear context is empty, we may often omit it; however, in this
-figure we make this distinction explicit. Further consider the leaves
-of the tree. First, the leaves involving the constructors are
-nonlinear terms, and thus they occur in the nonlinear empty
-context. Then we explicit cast from an nonlinear $\uparrow$-wrapped
-type into a linear one. Such casts may often be made implicitly, such
-as in $\cref{fig:kleeneabstractproof}$.
-
+Through repeated application of the Kleene star constructors, \cref{fig:kleenestarderivation} gives a derivation that shows \stringquote{ab} matches the
+regular expression $({\literal a}^{*} \otimes \literal b) \oplus \literal c$. The leaves of the proof tree
+that mention the arrow $\uparrow$ describe a cast from a non-linear type to a linear type.
+For instance, the premise of the leaf involving $\nil$
+views $\nil : \ltonl {({\literal a}^{*})}$ as the name of a
+constructor, and a constructor should be nonlinearly valued because we may call
+it several time. However, the conclusion of this leaf views $\nil : {\literal a}^{*}$ as a
+linear value. After we call the constructor it returns a value that may only be
+used a single time. In this paper such casts are often made implicitly, but this
+does not introduce any ambiguity.
+Even though we use $\nil$ to refer to both a constructor and the
+linear value produced by it
+, we differentiate between the two views by inspecting
+the type.
+
+\steven{TODO fig 3 sizing}
 \begin{figure}
 \begin{mathpar}
+\footnotesize
 \inferrule
@@
           {~}
-          {\cdot \vdash \cons : \ltonl {(a \lto a^* \lto a^*)}}
+          {\cdot \vdash \cons : \ltonl {({\literal a} \lto {\literal a}^* \lto {\literal a}^*)}}
         }
-        {\cdot ; \cdot \vdash \cons : a \lto a^* \lto a^*}
+        {\cdot \vdash \cons : {\literal a} \lto {\literal a}^* \lto {\literal a}^*}
         \\
@@
         {~}
-        {\cdot ; x : a \vdash x : a}
+        {a : {\literal a} \vdash a : {\literal a}}
       }
-      {\cdot ; x : a \vdash \cons~x : a^* \lto a^*}
+      {a : a \vdash \cons~a : {\literal a}^* \lto {\literal a}^*}
       \\
@@
         {~}
-        {\cdot \vdash \nil : \ltonl {(a^*)}}
+        {\cdot \vdash \nil : \ltonl {({\literal a}^*)}}
       }
-      {\cdot ; \cdot \vdash \nil : a^*}
+      {\cdot \vdash \nil : {\literal a}^*}
     }
-    {\cdot ; x : a \vdash \cons (x , \nil) : a^{*}}
+    {a : {\literal a} \vdash \cons~a~\nil : {\literal a}^{*}}
     \\
@@
     {~}
-    {\cdot ; y : b \vdash y : b}
+    {b : {\literal b} \vdash b : {\literal b}}
   }
-  {\cdot ; x : a , y : b \vdash \cons(x, \nil) \otimes y : a^{*} \otimes b}
+  {a : {\literal a} , b : {\literal b} \vdash (\cons~a~\nil) \otimes b : {\literal a}^{*} \otimes {\literal b}}
 }
-{\cdot ; x : a , y : b \vdash \inl (\cons(x) \otimes y) : (a^{*} \otimes b) \oplus c}
+{a : \literal a , b : \literal b \vdash \texttt{g} := \inl ((\cons~a~\nil)
+  \otimes b) : ({\literal a}^{*} \otimes \literal b) \oplus \literal c}
 \end{mathpar}
-\caption{\stringquote{ab} matches $(a^{*} \otimes b) \oplus c$}
+\begin{lstlisting}
+g : (*@$\uparrow$@*)(('a' (*@$\otimes$@*) 'b') (*@$\lto$@*) ('a'(*@$^*$@*) (*@$\otimes$@*) 'b') (*@$\oplus$@*) 'c')
+g (a , b) = inl (cons a' nil (*@$\otimes$@*) b')
+\end{lstlisting}
+\caption{\stringquote{ab} matches
+  $({\literal a}^{*} \otimes \literal b) \oplus \literal c$}
 \label{fig:kleenestarderivation}
@@
 We may also have derivations where the term in context is not simply a
-string of base types. For instance,
-in \cref{fig:kleeneabstractproof} we show that every parse of the grammar
-$(A \otimes A)^{*}$ induces a parse of $A^{*}$ for an arbitrary grammar $A$.
-
-The context $(A \otimes A)^{*}$ does not correspond directly to a string, so it is
+string of base types. In \cref{fig:kleeneabstractproof} we show that every parse
+of the grammar $(A \otimes A)^{*}$ induces a parse of $A^{*}$ for an
+arbitrary grammar $A$. The context $(A \otimes A)^{*}$ does not correspond directly to a string, so it is
 not quite appropriate
-to treat linear terms here as parse \textit{trees}. Instead, it is more
-appropriate to think of such terms as parse \textit{transformers}. That is, in
-\cref{fig:kleeneabstractproof} the
-inhabitant $x : (A \otimes A)^{*}$ in the context is not itself data to be parsed; rather,
-$x$ is a parse tree and the term from context $x : (A \otimes A)^{*}$ to the
-type $A^{*}$ is a transformation of parse trees of the input grammar into parse
-trees of the output grammar.
-We also see here why $\cons$ and $\nil$ should themselves be
-non-linear values: we should be able to use them in constructing multiple parse trees.
-
-\steven{TODO : describe elimination principle somewhere. Maybe just in prose?}
+to treat linear terms here as parse \textit{trees}.
+The
+context $a : (A \otimes A)^{*}$ does not contain concrete data to be parsed; rather, there may be many choices of string underlying the parse tree captured
+by the variable $a$. Thus, the term $\texttt{h}$ from
+\cref{fig:kleeneabstractproof} is not a parse of a string, and
+it is more
+appropriate to think of it as a parse \textit{transformer} --- a function from
+parses of $(\A \otimes \A)^{*}$ to parse of $\A^{*}$.
+
+We define $\texttt{h}$ by recursion on terms of type $(\A \otimes \A)^{*}$.
+This recursion is expressed in the derivation tree by invoking the
+elimination principle for Kleene star, written
+as $\fold$. The parse transformer $\texttt{h}$ is more compactly presented in the pseudocode
+of \cref{fig:kleeneabstractproof} by pattern matching on the input and making an
+explicit recursive call in the body of its definition. When appropriate we will favor this code-like
+presentation of terms in our calculus, as it is more readable and more
+faithful to the implementation of \theoryabbv in Agda.
 
@@
     {~}
-    {\cdot \vdash \nil : A^*}
+    {\cdot \vdash \nil : \A^*}
     \\
@@
         {~}
-        {x_1 : A, x_2 : A , y : A^* \vdash \cons (x_1 , \cons(x_2 , y)) : A^*}
+        {a_1 : \A, a_2 : \A , a_{*} : \A^* \vdash \cons (a_1 , \cons(a_2 , a_{*})) : \A^*}
       }
-      {x' : A \otimes A , y : A^* \vdash g := \letin {x_1 \otimes x_2} {x'} {\cons (x_1 , \cons(x_2 , y)) : A^*}}
+      {a' : \A \otimes \A , a_{*} : \A^* \vdash g := \letin {a_1 \otimes a_2} {a'} {\cons (a_1 , \cons(a_2 , a_{*})) : \A^*}}
     }
-    {\cdot \vdash f := \lamblto x {\lamblto y {\app {\app g x} {y} }} : (A \otimes A) \lto A^* \lto A^*}
+    {\cdot \vdash f := \lamblto a {\lamblto {a_{*}} {\app {\app g a} {a_{*}} }} : (\A \otimes \A) \lto \A^* \lto \A^*}
   }
-  {x : (A \otimes A)^* \vdash \fold(\nil , f)(x) : A^* }
+  {a : (\A \otimes \A)^* \vdash \texttt{h} := \fold(\nil , f)(a) : \A^* }
 \end{mathpar}
-\caption{An example mapping out of abstract grammars}
+\begin{lstlisting}
+h : (*@$\uparrow$@*)((A (*@$\otimes$@*) A)(*@$^*$@*) (*@$\lto$@*) A(*@$^*$@*))
+h nil = nil
+h (cons (a1 (*@$\otimes$@*) a2) as) = cons a1 (cons a2 (h as))
+\end{lstlisting}
+\caption{A parse transformer for abstract grammars}
 \label{fig:kleeneabstractproof}
@@
 \paragraph{Non-deterministic Finite Automata}
-\max{A lot of passive voice and indirect language in these paragraphs ``Often one wants to...'' ``Our method of representing...is''.}
-Automata play a critical role in defining parsers, as they often serve
-as an operationalized and executable presentation of a formal
-grammar. As a final example, consider the representation of $(a^{*}
-\otimes b) \oplus c$ as a non-deterministic finite automaton (NFA),
-shown in \cref{fig:exampleNFA}.
-
-Our method of representing these automata internally is to provide an \emph{indexed}
-inductive type of traces through the machine. An ordered, linear theory is
-integral to our design of \theoryabbv ; however, there are some aspects of parsing that
-cannot be captured through simple linear types alone. Often one wants to define a
-\emph{family} of linear types varying over a set, such as a type of traces through an
-NFA parameterized by the first state in the trace. For the example NFA, the data
-type $\Trace$ is provided in \cref{fig:nfainductive}.
-
-$\Trace~s$ denotes traces through the NFA beginning in state $s$ and ending in
-an accepting state. In this instance there is only a single accepting state, and
-thus the traces necessarily end in state 3. Let us dissect the structure
-of $\Trace~s$ in more detail.
-
-There are two kinds of constructors for a trace: those those modeling the
-termination of a trace at a given state and those modeling transitions in the
-NFA.\ The constructor $\mathsf{stop}$ denotes the first kind of constructor and
-terminates the trace in state 3. The constructor
-$\mathsf{2to3}$ denotes the $b$-labeled transition in the NFA from state 2 to state
-3. In this case, we build a trace beginning at state 2
-by consuming a parse of the label, $b$, and consuming a trace beginning at state 3.
-Finally, the constructor $\mathsf{1to2}$ is built quite similarly as $\mathsf{2to3}$;
-however, $\mathsf{1to2}$  may transition states without reading a character. This is because the
-transition from state 1 to state 2 is labeled by the empty string $\varepsilon$, so
-transition consumes no input. In this case, when building a trace out of
-state 1, it suffices to build a trace out of state 2 \emph{over the same string}.
-
-Overall, leveraging the \emph{indexed} inductive types provides a concise interface
-for defining the traces of the NFA as a family
-of mutually recursive types varying coherently over indexing data drawn from
-the states of the NFA.
+\newcommand{\s}{\texttt{s}}
+\newcommand{\0}{\texttt{0}}
+\newcommand{\1}{\texttt{1}}
+\newcommand{\2}{\texttt{2}}
+Regular expressions are a compact formalism for defining a formal grammar,
+however simply defining a regular expression --- such as
+$({\literal a}^{*} \otimes \literal b) \oplus \literal c$ --- does not give any insight into how to parse
+it. Many parsing algorithms are based on a notion of automaton, and to verify
+these algorithms we need to represent their associated automata in
+\theoryabbv.
+
+Finite automata are precisely the class of machines that recognize regular
+expressions. \cref{fig:exampleNFA} shows a non-deterministic
+finite automaton (NFA) for the regular expression
+$({\literal a}^{*} \otimes \literal b) \oplus \literal c$.
+\cref{fig:nfainductive} defines $\Trace$ an \emph{indexed} inductive linear type of
+traces through this automaton. Defining an indexed inductive type can be thought
+of like defining a family of mutually inductive types varying coherently over a
+type of indexing data. That is, \cref{fig:nfainductive} defines the linear types
+$\Trace~\0$,
+$\Trace~\1$, and $\Trace~\2$ mutually inductively. The index $\s : \texttt{Fin 3}$ picks out which
+state in the automaton a trace begins at --- where $\texttt{Fin 3}$ is the
+finite type containing inhabitants $\{\0 , \1 , \2\}$.
+There are three kinds of constructors for $\Trace$:
+(1) those that terminate traces, (2) those that correspond to transitions
+labeled by a character, and (3) those that correspond to transitions labeled by
+the empty string $\epsilon$. The constructor $\texttt{stop}$ terminates a trace
+in the accepting state $\2$. The constructors $\texttt{1to1}$, $\texttt{1to2}$,
+$\texttt{0to2}$ each define a labeled transition through the NFA, and each of
+these consumes a parse of the label's character and a trace beginning at the
+destination of a transition to produce a trace beginning at the source of a
+transition. The constructor $\texttt{0to1}$ behaves similarly, except its
+transition is labeled with the empty string $\epsilon$. Therefore,
+$\texttt{0to1}$ takes in a trace beginning a state $\1$ and returns a trace
+beginning at state $\0$ corresponding to the same underlying string.
 
 Just as \cref{fig:kleenestarderivation} shows the \stringquote{ab} matches
-$(a^{*} \otimes b) \oplus c$, we show in \cref{fig:nfaderivation} that \stringquote{ab} is
-accepted by the NFA representation of this regular expression. This is denoted by
-showing that \stringquote{ab} matches the trace grammar when the index is the
-initial state of the NFA.\ Further, as we
-will see in LATER SECTION, we can derive an isomorphism between parses of the
-regular expression and accepting traces of their corresponding NFA.
+$({\literal a}^{*} \otimes \literal b) \oplus \literal c$, \cref{fig:nfaderivation} demonstrates that \stringquote{ab} is
+accepted by the NFA representation of this regular expression. In particular, we
+show that \stringquote{ab} matches the grammar $\Trace~0$ because $\0$ is the
+initial state of the NFA.\ We may imagine defining parse transformers from
+$({\literal a}^{*} \otimes \literal b) \oplus \literal c$ to $\Trace~\1$, and vice versa. The existence of
+such maps between the two types would demonstrate that the regular expression
+and the automaton capture the same language --- i.e.\ they recognize the same
+set of strings. As we will see in LATER SECTION, we can define parse
+transformers between regular expressions and their associated automaton.
+Moreover, we prove that these parse transformers define a bijective
+correspondence of parses over every input string. That is, in a strong sense the
+parse of a regular expression and the traces of their associated automata
+capture precisely the same data.
 
@@
   \begin{tikzpicture}[node distance = 25mm ]
-    \node[state, initial] (1) {$1$};
-    \node[state, below left of=1] (2) {$2$};
-    \node[state, right of=2, accepting] (3) {$3$};
-
-    \path[->] (1) edge[above] node{$\varepsilon$} (2)
-              (1) edge[right] node{$c$} (3)
-              (2) edge[loop left] node{$a$} (2)
-              (2) edge[below] node{$b$} (3);
+    \node[state, initial] (0) {$\0$};
+    \node[state, below left of=0] (1) {$\1$};
+    \node[state, right of=1, accepting] (2) {$\2$};
+
+    \path[->] (0) edge[above] node{$\epsilon$} (1)
+              (0) edge[right] node{$\stringquote{c}$} (2)
+              (1) edge[loop left] node{$\stringquote{a}$} (1)
+              (1) edge[below] node{$\stringquote{b}$} (2);
   \end{tikzpicture}
-  \caption{NFA for $(a^{*} \otimes b) \oplus c$}
+  \caption{NFA for $(\a^{*} \otimes \b) \oplus \c$}
   \label{fig:exampleNFA}
@@
 \begin{figure}
-\begin{align*}
-\data &~\Trace : (s : \mathsf{Fin}~3) \to L~\where\\
-      & \mathsf{stop} : \ltonl {(\Trace~3)}  \\
-      & \mathsf{1to2} : \ltonl {(\Trace~2 \lto \Trace~1)} \\
-      & \mathsf{2to2} : \ltonl {(a \lto \Trace~2 \lto \Trace~2)} \\
-      & \mathsf{2to3} : \ltonl {(b \lto \Trace~3 \lto \Trace~2)}  \\
-      & \mathsf{1to3} : \ltonl {(c \lto \Trace~3\lto \Trace~1)} \\
-\end{align*}
+\begin{lstlisting}
+data Trace : (s : Fin 3) (*@$\to$@*) L where
+  stop : (*@$\uparrow$@*)(Trace 2)
+  1to1 : (*@$\uparrow$@*)('a' (*@$\lto$@*) Trace 1 (*@$\lto$@*) Trace 1)
+  1to2 : (*@$\uparrow$@*)('b' (*@$\lto$@*) Trace 2 (*@$\lto$@*) Trace 1)
+  0to2 : (*@$\uparrow$@*)('c' (*@$\lto$@*) Trace 2 (*@$\lto$@*) Trace 0)
+  0to1 : (*@$\uparrow$@*)(Trace 1 (*@$\lto$@*) Trace 0)
+\end{lstlisting}
 \caption{$N$ as an indexed inductive type}
@@
 \begin{figure}
-\begin{mathpar}
-\inferrule
-{
-  \inferrule
-  {
-    \inferrule
-    {~}
-    {x : a \vdash x : a}
-    \\
-    \inferrule
-    {
-      \inferrule
-      {~}
-      {y : b \vdash y : b}
-      \\
-      \inferrule
-      {~}
-      {\cdot \vdash \mathsf{stop} : \Trace~3}
-    }
-    {y : b \vdash \mathsf{2to3}(y , \mathsf{stop}) : \Trace~2}
-  }
-  {x : a , y : b \vdash \mathsf{2to2}(x , \mathsf{2to3}(y , \mathsf{stop})) : \Trace~2}
-}
-{x : a , y : b \vdash \mathsf{1to2}(\mathsf{2to2}(x , \mathsf{2to3}(y ,
-  \mathsf{stop}))) : \Trace~1}
-\end{mathpar}
-\caption{\stringquote{ab} is accepted by the NFA for $(a^{*} \otimes b) \oplus c$}
+\begin{lstlisting}
+k : (*@$\uparrow$@*)(('a' (*@$\otimes$@*) 'b') (*@$\lto$@*)  Trace 0)
+k (a , b) = 0to1 (1to1 a (1to2 b stop))
+\end{lstlisting}
+\caption{\stringquote{ab} is accepted by the NFA for
+  $({\literal a}^{*} \otimes {\literal b }) \oplus \literal c$}
 \label{fig:nfaderivation}
 \end{figure}
-
-\steven{TODO the example figures need to changed to not use $x$'s for linear variables}
 
@@
 
-\max{was this paragraph supposed to be deleted?}  Next we present the
-full details of the syntax and typing of \theoryabbv. The syntax of
+The syntax of
 \theoryabbv is based on the dependent linear-non-linear (\lnld)
-calculus from \cite{dlnl}. As in \lnld, \theoryabbv includes both
+calculus from \cite{krishnaswami_integrating_2015}. As in \lnld, \theoryabbv includes both
 non-linear dependent types, as well as linear types, which are allowed
@@
 variables. This type is used extensively in our examples, playing a
-similar role to the $!$ modality of ordinary linear logic.
+similar role to the $!$ modality of ordinary linear logic or the
+persitent modality $\square$ from separation logic.
 
@@
 %% \end{figure}
-
 \begin{figure}
@@
 $B$.
+\pedro{Isn't this is a slightly incorrect desciption of $\ltonl{A \lto B}$?
+How is the description above distinct form $A \lto B$?}
 
@@
 respectively. We will freely use $\bot$ to refer to this empty
-discunction and $\top$ to refer to the empty conjunction. Lastly, we
+disjunction and $\top$ to refer to the empty conjunction. Lastly, we
 include a type $\equalizer {a}{f}{g}$ that we call the
@@
     %
-    \inferrule{\Gamma ; \Delta \vdash e : A \otimes B \\ \Gamma ; \Delta'_1, a : A, b : B, \Delta'_2 \vdash e'}{\Gamma ;  \Delta_1', \Delta, \Delta'_2 \vdash \letin {a \otimes b} e {e'}}
+    \inferrule{\Gamma ; \Delta \vdash e : A \otimes B \\ \Gamma ; \Delta'_1, a : A, b : B, \Delta'_2 \vdash e' : C}{\Gamma ;  \Delta_1', \Delta, \Delta'_2 \vdash \letin {a \otimes b} e {e'} : C}
     \\
@@
 \end{figure}
+\pedro{Fixed tensor elimination rule (added output type)}
+\pedro{Perhaps we should modify the equalizer introduction rule.
+  As it stands, readers might be confused by the same symbol $e$
+  being used everywhere.}
 
@@
 
-Next, we introduce one of the most complex and important type linear
-type constructor of our development, \emph{indexed inductive linear
+Next, we introduce one of the most complex and important linear
+type constructors of our development, \emph{indexed inductive linear
 types} $\mu F$. To encode these, we first add a non-linear type
@@
 strictly-positive functors in \Cref{fig:spf}. First, we add operations
-for a recursive reference $\Var{x}$ the constant functor, $K$, tensor
+for a recursive reference $\Var~x$ the constant functor, $K$, tensor
 products and additive conjunction and disjunction. As an example,
@@
   this way for instance. I think there is a linear container paper
-  that uses this appraoch}
+  that uses this approach}
 
@@
 of type expressions $F$. The introduction rule for this is
-$\mathsf{roll}$, which constructs an element of $\mu F\,x$ from its
+$\roll$, which constructs an element of $\mu F\,x$ from its
 one-level unfolding. The elimination principle is defined by a
-$\mathsf{fold}$ operation. The equations ensure that this makes the
+$\fold$ operation. The equations ensure that this makes the
 family an \emph{initial algebra} for the type expressions. That is the
@@
 This section explores the applications of \theoryabbv to conducting formal
-gramamr theory. We demonstrate that several classical notions and constructions
+grammar theory. We demonstrate that several classical notions and constructions
 integral to the theory of formal languages are faithfully represented
@@
 
-\subsection{Grammar Properties}
-
-\steven{Need an intro for this subsection}
-
-%% 1. Basic concepts: unambiguity, parseability, weak and strong equivalence
-\paragraph{Unambiguity}
-
-Ordinarily, we think of a formal grammar as unambiguous when there is at most
-one parse tree for any input string; however, this description of unambiguity is
-a semantic property that is not expressible in the syntax of \theoryabbv. Rather
-than capturing this uniqueness aspect directly, we provide several notions
-internal to \theoryabbv that imply unambiguity in the grammars model.
-
-\max{what's a terminal object and what's a monomorphism and what's a distributive coproduct}
-We say that $A$ is \emph{unambiguous} if the unique map $! : A \to \top$ to the
-terminal object is a monomorphism. Further, in the presence of distributive coproducts, the following are equivalent:
-\begin{itemize}
-  \item $A$ is an unambiguous grammar
-  \item Given a grammar $B$, all morphisms $B \to A$ are equal\max{is this internal statement? What is a ``morphism''}
-  \item $A \& A \cong A$\max{what is $\cong$?}
-\end{itemize}
-
-\paragraph{A Grammar of Strings}
-Define $\charg$ to be the disjunction of characters in the alphabet $\Sigma$,
-$\charg = \LinSigTy {c} {\Sigma} c$. Then define a linear type of strings,
-$\stringg = \charg^{*}$. In the grammars model, we may construct an
-isomorphism between $\top$ and $\stringg$. Thus when building parsers, we may wish to enforce this
-syntactically by including $\top \cong \stringg$ as an axiom.
-
-Through inclusion of this axiom, $! : \stringg \to \top$ is an isomorphism, and
-thus also a monomorphism. That is, $\stringg$ is unambiguous.
-\max{we should have a section of the previous section which gives the axioms, this one and the distributivity axiom}
-
-\paragraph{Parseability}
-We say $A$ is \emph{parseable} if there exists a grammar $B$ and a map
-$\top \to A \oplus B$. Taking $\top \cong \stringg$, a grammar $A$ is parseable
-exactly when we may read an arbitrary string into a parse of $A$ or a $B$-valued
-refutation.
-
-\paragraph{Decidability}
-In the case that $A$ is parseable with respect to $\neg A$ --- i.e.\
-there's a map $\top \to A \oplus \neg A$ --- then call $A$ \emph{decidable}.
-
 \paragraph{Equivalence}
@@
 generative capacity --- the set of ``structural descriptions'', or parse trees,
-that it generates. Two grammars are said to be \emph{weakly equivalent} by
-Chomsky\max{you can drop the ``by Chomsky'' at this point} if they
+that it generates. Two grammars are said to be \emph{weakly equivalent} if they
 agree in their weak generative capacity, and likewise \emph{strongly equivalent}
 if they agree in their strong generative capacity. We aim to recover each of
-these notions of equivalence internally.
-
-Internal to \theoryabbv, we define grammars $A$ and $B$ to be \emph{weakly equivalent} if there exists maps
-$f : A \to B$ and $g : B \to A$. Note that in the grammars model, this implies the expected equality on
-the set of accepted strings.
-
-\max{indirect language}
-If $f$ and $g$ form an isomorphism, then $A$ and $B$ defined to be \emph{strongly
-  equivalent}. This implies Chomsky's notion of strong equivalence, as for any
-input string $w$
-$A \cong B$ induces a bijective translation of parses $A w \cong B w$.
-When abstracting over the latent
-structure of formal grammars and describing it in the the language of
-categories, it is encouraging to see the same definition of strong equivalence
-emerge naturally.
+these notions of equivalence in \theoryabbv.
+
+\begin{definition}
+  \label{def:weakequiv}
+  Grammars $\A$ and $\B$ are \emph{weakly equivalent} if there exists parse
+  transformers $\f : \ltonl{(\A \lto \B)}$ and $\g : \ltonl {(\B \lto \A)}$.
+\end{definition}
+\pedro{Up to this point, we have mentioned a couple of times parse transformers,
+  but we have never defined what they are. In particular, why does the definition
+  require the $\uparrow$ modality. I'm assuming that it's to properly deal with
+  linear variables, but these things are not obvious at first.}
+
+\newcommand{\lang}{\texttt{Lang}}
+A parse transformer $\f : \ltonl{(\A \lto \B)}$ takes as input parses of $\A$ for string
+$\w$ and returns a parse of $\B$ for $\w$. Therefore, $\f$ demonstrates an
+inclusion of languages $\lang(\A) \subset \lang(\B)$. Similarly, the existence of
+$\g$ shows that $\lang(\B) \subset \lang(\A)$, and thus $\lang(\A) = \lang(\B)$.
+
+\steven{Need better notation for language that is cohesive throughout the paper}
+
+\begin{definition}
+  \label{def:strongequiv}
+  Grammars $\A$ and $\B$ are \emph{strongly equivalent} if there exists parse
+  transformers $\f : \ltonl{(\A \lto \B)}$ and $\g : \ltonl {(\B \lto \A)}$ such
+  that $\f \circ \g \equiv \ident$ and $\g \circ \f \equiv \ident$.
+\end{definition}
+
+For a string $\w$, \cref{def:strongequiv} induces a bijection between the
+$\A$-parses of $\w$ and the $\B$-parses of $\w$, thus demonstrating that $\A$
+and $\B$ agree in their strong generative capacity.
+
+\paragraph{Ambiguity}
+A formal grammar $\A$ is ambiguous if there is a string $\w$ for which $\A$ has
+several parse trees of $\w$.
+For example, $\a \oplus \a$ is ambiguous because
+$\stringquote{a}$ has two parse trees --- on either side of the disjunction. On
+the other hand, a formal grammar is unambiguous when there is at most
+one parse tree for any input string. These notions of ambiguity are entirely
+semantic, so we seek to find a property expressible in the syntax of \theoryabbv
+that reifies the semantic description of ambiguity.
+
+\begin{definition}
+  \label{def:unambig}
+  A grammar $\A$ is \emph{unambiguous} if for every linear type $\B$,
+  $\f : \ltonl{(\B \lto \A)}$, and $\g : \ltonl{(\B \lto \A)}$
+  then $\f \equiv \g$
+\end{definition}
+\cref{def:unambig} can be read more intuitively as stating that $\A$ is
+unambiguous if there is at most one way to transform parses of any other grammar
+$\B$ into parse of $\A$.
+
+Note that this definition of unambiguity is preserved by strong equivalence, so a common proof technique for showing $\A$ is unambiguous is by proving $\A$
+is strongly equivalent to an unambiguous grammar $\B$. In this style of proof, we
+build the functions $\f : \ltonl{(\A \lto \B)}$ and
+$\g : \ltonl{(\B \lto \A)}$, then we have the obligation to show
+$\f \circ \g \equiv \ident$ and $\g \circ \f \equiv \ident$. However, because $\B$ is
+unambiguous there is only one function of type $\ltonl{(\B \lto \B)}$, so
+$\f \circ \g : \ltonl{(\B \lto \B)}$ and $\ident : \ltonl{(\B \lto \B)}$ are equal.
+Thus, in this type of proof, we really only have the obligation of building each of $\f$
+and $\g$ and demonstrating that $\g \circ \f \equiv \ident$.
+
+% \paragraph{A Grammar of Strings}
+% Define the character grammar $\charg$ to be the disjunction over characters in the alphabet $\Sigma$,
+% $\charg = \LinSigTy {c} {\Sigma} c$. Then define a linear type of strings as a
+% linear list of characters,
+% $\stringg = \charg^{*}$.
+
+% In the grammars model, we can build a
+% strong equivalence between $\top$ and $\stringg$. hen building parsers, we may wish to enforce this
+% syntactically by including $\top \cong \stringg$ as an axiom.
+
+% Through inclusion of this axiom, $! : \stringg \to \top$ is an isomorphism, and
+% thus also a monomorphism. That is, $\stringg$ is unambiguous.
+% \max{we should have a section of the previous section which gives the axioms, this one and the distributivity axiom}
+
+% \paragraph{Parseability}
+% We say $A$ is \emph{parseable} if there exists a grammar $B$ and a map
+% $\top \to A \oplus B$. Taking $\top \cong \stringg$, a grammar $A$ is parseable
+% exactly when we may read an arbitrary string into a parse of $A$ or a $B$-valued
+% refutation.
+
+% \paragraph{Decidability}
+% In the case that $A$ is parseable with respect to $\neg A$ --- i.e.\
+% there's a map $\top \to A \oplus \neg A$ --- then call $A$ \emph{decidable}.
 
@@
 %% 2. Regex, NFA, DFA
-
-\max{this paragraph isn't really saying anything}
-Regular expressions
-are among one of the most studied and
-well-understood concepts in all of computer science. Pratically, they are
-essential tools for lexical analysis and many text
-processing tasks. So they are a natural concept to encoding into our
-domain-specific grammars language.
-
-\max{who cares what every discussion of regular expressions is accompanied by}
-Nearly every theoretical discussion of regular expressions is accompanied by a
-discussion of finite automata, as it is well known that regular expressions and finite automata are
-equivalent as grammar formalisms. In this section, we make the equivalences
+In this section, we make the equivalences
 between regular expressions and finite automata precise by encoding two theorems
 from classical language theory: (1) every regular expression is strongly equivalent
-to the traces through some nondeterministic finite automaton (NFA); and, (2) for
+to the traces through some non-deterministic finite automaton (NFA); and, (2) for
 every NFA, there is some deterministic finite automaton (DFA) such that the type
@@
 
-\newcommand{\states}{\mathsf{states}}
-\newcommand{\labelt}{\mathsf{label}}
-\newcommand{\transitions}{\mathsf{transitions}}
-\newcommand{\epstransitions}{\varepsilon\mathsf{transitions}}
-\newcommand{\isAcc}{\mathsf{isAcc}}
-\newcommand{\init}{\mathsf{init}}
-\newcommand{\src}{\mathsf{src}}
-\newcommand{\dst}{\mathsf{dst}}
-\newcommand{\parse}{\mathsf{parse}}
-\newcommand{\print}{\mathsf{print}}
-\newcommand{\epssrc}{\varepsilon\mathsf{src}}
-\newcommand{\epsdst}{\varepsilon\mathsf{dst}}
+\newcommand{\states}{\mathtt{states}}
+\newcommand{\labelt}{label}
+\newcommand{\transitions}{\texttt{transitions}}
+\newcommand{\epstransitions}{\epsilon\texttt{transitions}}
+\newcommand{\isAcc}{\mathtt{isAcc}}
+\newcommand{\init}{\mathtt{init}}
+\newcommand{\src}{\mathtt{src}}
+\newcommand{\dst}{\mathtt{dst}}
+\newcommand{\parse}{\mathtt{parse}}
+\newcommand{\print}{\mathtt{print}}
+\newcommand{\epssrc}{\epsilon\mathtt{src}}
+\newcommand{\epsdst}{\epsilon\mathtt{dst}}
+
+\newcommand{\N}{\texttt{N}}
 
@@
 
-\paragraph{Nondeterministic Finite Automata} In \cref{sec:type-theory-examples}, we saw the demonstration of
+\paragraph{Nondeterministic Finite Automata} In \cref{sec:type-theory-examples}, we saw
 one particular NFA corresponding to the regular expression
-$(a^{*} \otimes b) \oplus c$. The definition of the linear type of traces
+$({\literal a}^{*} \otimes \literal b) \oplus \literal c$. The definition of the linear type of traces
 through this
@@
 NFA.\ When generalizing from this example to an arbitrary NFA, we do not have the
-luxury of inspecting a concrete defintion. Therefore, we abstract over the
+luxury of inspecting a concrete definition. Therefore, we abstract over the
 specification of an arbitrary NFA and define traces with respect to that
@@
 
-Let the data of an NFA $N$ be given by the following nonlinear data,
+Let the data of an NFA $\N$ be given by the following nonlinear data,
 \begin{itemize}
-\item $N.\states$ --- a finite set of states
-\item $N.\init$ --- the initial state
-\item $N.\isAcc$ --- a decidable proposition over $N.\states$ denoting which
-        states are accepting
-\item $N.\transitions$ --- a finite set of labeled transitions
-\item $N.\labelt$ --- a mapping from $N.\transitions$ to the character labeling the transition
-\item $N.\src$ and $N.\dst$ --- mappings from $N.\transitions$ to source and target states of a labeled transition
-\item $N.\epstransitions$ --- a finite set of $\varepsilon$-transitions
-\item $N.\epssrc$ and $N.\epsdst$ --- mappings from $N.\epstransitions$ to source and target states of an $\varepsilon$-transition
+\item $\N.\states$ --- a finite set of states
+\item $\N.\init : \N.\states$ --- the initial state
+\item $\N.\isAcc : \N.\states \to \Bool$ --- a mapping marking which states are
+        accepting
+\item $\N.\transitions$ --- a finite set of labeled transitions
+\item $\N.\labelt : \N.\transitions \to \Sigma$ --- a mapping from $N.\transitions$ to the character labeling the transition
+\item $\N.\src , \N.\dst : \N.\transitions \to \N.\states$ --- mappings from
+        $\N.\transitions$ to the source and target states, respectively, of a labeled transition
+\item $\N.\epstransitions$ --- a finite set of $\varepsilon$-transitions
+\item $\N.\epssrc, \N.\epsdst : \N.\epstransitions \to \N.\states$ --- mappings
+        from $\N.\epstransitions$ to the source and target states of an $\epsilon$-transition
 \end{itemize}
 
-For the above definition of an NFA $N$, in \cref{fig:nfatrace} we define a linear type of traces
-through $N$. Traces in $N$ may be built through one of three constructors. We
+In \cref{fig:nfatrace} we define a linear type of traces
+through an NFA $\N$. $\Trace_{\N}$ is an inductive type indexed by the first
+state $\s : \N.\states$ of a trace and a Boolean $b : \Bool$ that marks if the
+trace ends in an accepting state. Traces in $\N$ may be built through one of three constructors. We
 may terminate a trace at an accepting state with the constructor $\nil$. If we
@@
 Finally, if we had a trace beginning at the destination of an
-$\varepsilon$-transition then we may use $\epscons$ to pull it back along the
-$\varepsilon$-transition and construct a trace beginning at the source of the $\varepsilon$-transition.
+$\epsilon$-transition then we may use $\epscons$ to pull it back along the
+$\epsilon$-transition and construct a trace beginning at the source of the
+$\epsilon$-transition. As a shorthand, write $Parse_{N}$ for the accepting
+traces out of $N.init$.
 
 \begin{figure}
-\begin{align*}
-\data &~\Trace_{N} : (s : N.\states) \to L~\where\\
-      & \nil : \ltonl {\left( \LinPiTy {s} {N.\states} {\LinPiTy {\mathsf{acc}} {N.\isAcc(s)} {\Trace_{N}~s}} \right)}  \\
-      & \cons : \ltonl {\left( \LinPiTy {t} {N.\transitions} {(N.\labelt~t) \lto \Trace_{N}~(N.\dst(t)) \lto \Trace_{N}(N.\src(t))} \right)} \\
-      & \epscons : \ltonl {\left( \LinPiTy {t} {N.\epstransitions} {\Trace_{N}~(N.\epsdst(t)) \lto \Trace_{N}(N.\epssrc(t))} \right)}
-\end{align*}
+\begin{lstlisting}
+data Trace(*@$_{\N}$@*) : (s : N.states) (b : Bool) (*@$\to$@*) L where
+  nil : (*@$\uparrow$@*)(&[ s (*@$\in$@*) N.states ] Trace(*@$_{\N}$@*) s (N.isAcc s))
+  cons : (*@$\uparrow$@*)(&[ t (*@$\in$@*) N.transitions ] &[ b (*@$\in$@*) Bool ]
+           ('N.label t' (*@$\lto$@*) Trace(*@$_{\N}$@*) (N.dst t) b (*@$\lto$@*) Trace(*@$_{\N}$@*) (N.src t) b))
+  (*@$\epsilon$@*)cons : (*@$\uparrow$@*)(&[ t (*@$\in$@*) N.(*@$\epsilon$@*)transitions ] &[ b (*@$\in$@*) Bool ]
+           (Trace(*@$_{\N}$@*) (N.(*@$\epsilon$@*)dst t) b (*@$\lto$@*) Trace(*@$_{\N}$@*)(N.(*@$\epsilon$@*)src t) b))
+\end{lstlisting}
 \caption{Traces through an NFA $N$ as an indexed inductive type}
@@
 
-\paragraph{Determinstic Finite Automata}
-Just as above, we provide a specification for an arbitrary DFA.\ A DFA $D$ is
+\paragraph{Deterministic Finite Automata}
+\newcommand{\D}{\texttt{D}}
+Just as the above defines an NFA, we provide a specification for an arbitrary
+deterministic finite automaton.\ A DFA $D$ is
 given by the following data,
@@
 \begin{itemize}
-\item $D.\states$ --- a finite set of states
-\item $D.\init$ --- the initial state
-\item $D.\isAcc$ --- a decidable proposition over $D.\states$ denoting which
-      states are accepting
-\item $D.\delta$ --- a deterministic transition function. Given a character $c$ and a state
-        $s$, $D.\delta_{c}(s)$ is the state for which there is a transition $s \overset{c}{\to} D.\delta_{c}(s)$
+\item $\D.\states$ --- a finite set of states
+\item $\D.\init : \D.\states$ --- the initial state
+  \item $\D.\isAcc : \D.\states \to \Bool$ --- a mapping marking which states
+        are accepting
+\item $\D.\delta : \Sigma \to \D.\states \to \D.\states$ --- a deterministic transition function. Given a character $\c$ and a state
+        $\s$, $\D.\delta~\c~\s$ is the state for which there is a transition $\s \overset{\stringquote{c}}{\to} (\D.\delta~\c~\s)$
 \end{itemize}
@@
 \begin{figure}
-\begin{align*}
-\data &~\Trace_{D} : (s : D.\states) (b : \Bool) \to L~\where\\
-      & \nil : \ltonl {\left( \LinPiTyLimit {s} {D.\states} {\Trace_{D}(s , D.\isAcc(s))} \right)}  \\
-      & \cons : \ltonl {\left( \LinPiTyLimit {c} {\Sigma} {\LinPiTyLimit {s} {D.\states} {c \lto \LinSigTyLimit {b}{\Bool}{\Trace_{D}~(D.\delta_{c}(s), b)} \lto \Trace_{D}(s, b)}} \right)}
-\end{align*}
+\begin{lstlisting}
+data Trace(*@$_{\D}$@*) : (s : D.states) (b : Bool) (*@$\to$@*) L where
+  nil : (*@$\uparrow$@*)(&[ s (*@$\in$@*) D.states ] Trace(*@$_{\D}$@*) s D.isAcc s)
+  cons : (*@$\uparrow$@*)(&[ c (*@$\in$@*) (*@$\Sigma$@*) ] &[ s (*@$\in$@*) D.states ] &[ b (*@$\in$@*) Bool ]
+             ('c' (*@$\lto$@*) Trace(*@$_{\D}$@*) (D.(*@$\delta$@*) c s) b (*@$\lto$@*) Trace(*@$_{\D}$@*) s b))
+\end{lstlisting}
 \caption{Traces through the DFA $D$ as an indexed inductive type}
@@
 
-\steven{TODO make the DFA parser figure fit}
-
-The linear type of traces through $D$ is given in \cref{fig:dfatrace}. These
-traces are either terminated in a state $s$ with the $\nil$ constructor. When
-terminating a trace, we tag it with a Boolean that marks if the end state was
-accepting or not. The $\cons$ constructor builds a trace out of state $s$ by linearly combining
-a parse of some character $c$ with a
-trace out of the state $D.\delta_{c}(s)$. The trace built with $\cons$ is
-accepting if and only if the trace out of $D.\delta_{c}(s)$ is accepting.
+$\Trace_{\D}$, the linear type of traces through $\D$, is given in
+\cref{fig:dfatrace}. These
+traces may be terminated in an accepting state $\s$ with the $\nil$ constructor.
+The $\cons$ constructor builds a trace out of state $\s$ by linearly combining
+a parse of some character $\c$ with a
+trace out of the state $\D.\delta~\c~\s$. The trace built with $\cons$ is
+accepting if and only if the trace out of $D.\delta~c~s$ is accepting.
 
@@
 are unambiguous and define a parser for them. In particular we show that
-$\LinPiTy {s} {D.\states} {\LinSigTy {b} {\Bool} \Trace_{D}(s , b)}$ is isomorphic to
+$\LinPiTy {s} {D.\states} {\LinSigTy {b} {\Bool} \Trace_{D}~s~b}$ is strongly
+equivalent to
 $\stringg$, which simultaneously constructs a parser and demonstrates that
-traces through $D$ is unambiguous. This isomorphism intuitively says that we pay
+traces through $D$ is unambiguous. Intuitively this strong equivalence says that we may
 unambiguously build traces through $D$ beginning at any state $s$, and further
@@
 
-The first half of the isomorphism,
-$\parse_{D} : \ltonl {\left( \stringg \lto \LinPiTy {s} {D.\states} {\Trace_{D}(s)} \right)}$,
+The first half of the equivalence,
+$\parse_{D}$,
 is defined by induction on strings in \cref{fig:dfaparser}. If this string is empty, then $\parse_{D}$
 defines a linear function that terminates a trace at the input state $s$. If the
-string is nonempty, then $\parse_{D}$ walks forward from the input state $s$ by
+string is nonempty, then $\parse_{D}$ walks forward in $D$ from the input state $s$ by
 the character at the head of the string.
@@
 \begin{figure}
-\[
-\inferrule
-{
-  \cdot \vdash mt := \withlamb {s} {\sigma~(D.\isAcc(b))~(\nil.\pi~s)}   : \LinPiTyLimit {s} {D.\states} {\LinSigTyLimit {b} {\Bool} {\Trace_{D}(s, b)}}
-  \\
-  ch : \charg , tr : \LinPiTyLimit {s} {D.\states} {\LinSigTyLimit {b} {\Bool } {\Trace_{D}(s, b)}} \vdash
-  f := \withlamb {s} {\letin {\sigma~c~p_{c}} {ch} {\letin {\sigma~b~t} {tr(D.\delta_{c}(s))} {\sigma~b~(\cons(p_{c} , t))}}}
-  : \LinPiTyLimit {s} {D.\states} {\LinSigTyLimit {b} {\Bool } {\Trace_{D}(s, b)}}
-}
-{w : \stringg \vdash \parse_{D} := \fold(mt, f)(w) : \LinPiTyLimit {s} {D.\states} {\LinSigTyLimit {b} {\Bool } {\Trace_{D}(s, b)}}}
-\]
+\begin{lstlisting}
+parse(*@$_D$@*) : (*@$\uparrow$@*)(String (*@$\lto$@*) &[ s (*@$\in$@*) D.states ] (*@$\oplus$@*)[ b (*@$\in$@*) Bool ] Trace(*@$_{\D}$@*) s b)
+parse(*@$_D$@*) String.nil s = (*@$\sigma$@*) (D.isAcc s) (Trace(*@$_{\D}$@*).nil s)
+parse(*@$_D$@*) (String.cons ((*@$\sigma$@*) c a) w) s =
+   let (*@$\sigma$@*) b t = parse w (D.(*@$\delta$@*) c s) in
+   (*@$\sigma$@*) b (Trace(*@$_{\D}$@*).cons c s b a t)
+\end{lstlisting}
 \caption{Parser for DFA traces}
@@
 
-The second half of the isomorphism,
-$\print_{D} : \ltonl {\left( \LinPiTy {s} {D.\states} {\Trace_{D}(s)} \lto \stringg \right)}$,
+The second half of the equivalence,
+$\print_{D}$,
 is given in \cref{fig:printdfa} by induction on traces. If the trace is defined
 via $\nil$, then $\print_{D}$ returns the empty string. Otherwise, if the trace
-is defined by $\cons$ then $\parse_{D}$ appends the character $c$ to the output
+is defined by $\cons$ then $\parse_{D}$ appends the character from the most
+recent transition to the output
 string and recurses.
@@
 \begin{figure}
-\[
-\inferrule
-{
-}
-{tr : \LinPiTyLimit {s} {D.\states} {\LinSigTyLimit {b} {\Bool } {\Trace_{D}(s, b)}} \vdash asdf : \stringg}
-\]
-\caption{Parser for DFA traces}
+\begin{lstlisting}
+print(*@$_D$@*) : (s : D.states) (*@$\to$@*) (*@$\uparrow$@*)(((*@$\oplus$@*)[ b (*@$\in$@*) Bool ] Trace(*@$_{\D}$@*) s b) (*@$\lto$@*) String)
+print(*@$_D$@*) s ((*@$\sigma$@*) b (Trace(*@$_{\D}$@*).nil .s)) = String.nil
+print(*@$_D$@*) s ((*@$\sigma$@*) b (Trace(*@$_{\D}$@*).cons c (D.(*@$\delta$@*) c .s) b a trace)) =
+  String.cons ((*@$\sigma$@*) c a) (print(*@$_D$@*) (D.(*@$\delta$@*) c s) ((*@$\sigma$@*) b trace))
+\end{lstlisting}
+\caption{Printer for DFA traces}
 \label{fig:printdfa}
@@
 
-\max{I've seen people use a checkmark or a cute little agda logo for this}
-\newcommand{\Agda}{Formalized in Agda}
+\steven{TODO make the Agda logo clickable, and mention its meaning somewhere}
+\newcommand{\Agda}{\agdalogo}
 \begin{theorem}[Determinization, \Agda]
+  \label{thm:determinization}
   Given an NFA $N$, there exists a DFA $D$ such that
-  $\Trace_{N}~N.\mathsf{init}$ is weakly equivalent to $\Trace_{D}~D.\mathsf{init}$.
+  $Parse_{N}$ is weakly equivalent to $Parse_{D}$.
 \end{theorem}
-
+\newcommand{\X}{\texttt{X}}
+\newcommand{\Y}{\texttt{Y}}
+\newcommand{\x}{\texttt{x}}
+\newcommand{\y}{\texttt{y}}
 \begin{proof}
-  We present a variant of the Rabin and Scotts's classic powerset construction \cite{rabinFiniteAutomataTheir1959}. First, we build
-  the DFA $D$. Define the
-  states of $D$ to be the $\mathbb{P}_{\varepsilon}(N.\mathsf{states})$ --- the type of $\varepsilon$-closed subsets of
-  $N.\mathsf{states}$. A subset is accepting in $D$ if it contains an accepting state
-  from $N$. Construct the initial state of $D$ as the
-  $\varepsilon$-closure of $N.\mathsf{init}$. Lastly define the transtion function
-  of $D$ to send the subset $X$ under the character $c$ to the
-  $\varepsilon$-closure of all the states reachable from $X$ via a transition
-  labelled with the character $c$.
-
-  We demonstrate the weak equivalence between $\Trace_{N}~(N.\mathsf{init})$ and
-  $\Trace_{D}~(D.\mathsf{init})$ by constructing morphisms between the two
-  grammars.
-
-  To build a term
-  $\Trace_{N}~(N.\mathsf{init}) \vdash \Trace_{D}~(D.\mathsf{init})$, it
-  suffices to construct the stronger term
-
+  We present a variant of Rabin and Scott's classic powerset construction \cite{rabinFiniteAutomataTheir1959}. First, we build
+  the DFA $\D$. Define the
+  states of $\D$ to be the $\mathbb{P}_{\varepsilon}(\N.\states)$ --- the type
+  of $\epsilon$-closed\footnote{A subset of states $\X$ is $\epsilon$-closed if
+    for every $\s \in \X$ and $\epsilon$-transition $\s \overset{\epsilon}{\to} \s'$ we have $\s' \in \X$.} subsets of
+  $\N.\states$. A subset is accepting in $\D$ if it contains an accepting state
+  from $\N$. Construct the initial state of $\D$ as the
+  $\epsilon$-closure of $\N.\init$. Lastly define the transition function
+  of $\D$ to send the subset $X$ under the character $\c$ to the
+  $\epsilon$-closure of all the states reachable from $X$ via a transition
+  labelled with the character $\c$.
+
+  We demonstrate the weak equivalence between $Parse_{N}$ and
+  $Parse_{D}$ by constructing parse transformers between the two
+  grammars. To build the parse transformer
+  $\ltonl{(Parse_{N} \lto Parse_{D})}$, we strengthen our inductive hypothesis
+  and build the term
+\newcommand{\NtoD}{\texttt{NtoD}}
+\newcommand{\DtoN}{\texttt{DtoN}}
   \[
-    f : \ltonl {\left(\Trace_{N}~(s) \lto \bigwith\limits_{X : D.\mathsf{states}} \bigwith\limits_{s \in X} \Trace_{D}~X\right)}
+    \NtoD : \ltonl {\left(\Trace_{\N}~\s~\true \lto \bigwith\limits_{\X : \D.\states} \bigwith\limits_{\texttt{sInX} : \X \ni \s} \Trace_{\D}~\X~\true \right)}
   \]
-  where $s$ is arbitary and then instantiate it at $s = N.\mathsf{init}$, $X = D.\mathsf{init}$.
-
-Because
-the indexing state $s$ is now arbitrary, we can define the desired term via
-induction the trace through the NFA.
-
-If the NFA trace is terminating at
-accepting state
-$q$ and $q \in X$,
-then we return a trace in $D$ that terminates at
-$X$, which is accepting because it contains the accepting state $q$.
-
-If the NFA trace is a labeled transition $s \overset{c}{\to} s'$ such that
-$s \in X$, then we first decompose the trace to a parse of the character $c$ and a
-trace starting from $s'$ --- that is, to a term of type
-$c \otimes \Trace_{N}(s')$. We then inductively call $f$ on traces out of $s'$
-to get traces in a parse of $c$ concatenated with a family of traces
-parametrized by $\varepsilon$-closed subsets containing $s'$.
-. We instantiate this family at $\delta_{c}(X)$, as
-$s'$ is reachable from $s$ by a single transition labelled by $c$, so
-$s' \in \overline{\varepsilon}\overline{c}{X} = D.\delta_{c}(X)$. Therefore we can
-linearly combine the parse of character $c$ and the trace out of $D.\delta_{c}(X)$
-to get a trace in $D$ rooted at $X$.
-
-\[\begin{tikzcd}
-    {c \otimes \mathsf{Trace}_{N}(s')} &
-    {c \otimes \bigwith\limits_{Y : D.\mathsf{states}} \bigwith\limits_{s' \in Y} \Trace_{D}~Y} &
-    {c \otimes \mathsf{Trace}_{D}(D.\delta_{c}(X))} & {\mathsf{Trace}_{D}(X)}
-	\arrow["{id\otimes f}", from=1-1, to=1-2]
-	\arrow["{id\otimes\pi}", from=1-2, to=1-3]
-	\arrow["{D.\mathsf{cons}}", from=1-3, to=1-4]
-\end{tikzcd}\]
-
-Lastly, if the NFA trace was an $\varepsilon$-transition
-$s \overset{\varepsilon}{\to} s'$ with $s \in X$, then the construction is quite similar. In this
-case, we again inductively call $f$ on traces out of $s'$ except we instantiate
-the choice of subset containing $s'$ with $X$ itself. We may do this as all
-states in $D$ are taken to be $\varepsilon$-closed, therefore $s' \in X$ because
-$s$ is in $X$ and $\varepsilon$-transitions to $s'$.
-
-\[\begin{tikzcd}
-    {\mathsf{Trace}_{N}(s')} &
-    {\bigwith\limits_{Y : D.\mathsf{states}} \bigwith\limits_{s' \in Y} \Trace_{D}~Y} &
-    {\mathsf{Trace}_{D}(X)}
-	\arrow["{id\otimes f}", from=1-1, to=1-2]
-	\arrow["{id\otimes\pi}", from=1-2, to=1-3]
-\end{tikzcd}\]
-
-This concludes the construction of $f$.
+  that maps a trace in $\N$ from an arbitrary state $\s$ to a trace in $\D$
+  that may begin at any subset of states $\X$ that contains $\s$. $\NtoD$ may then be instantiated at
+  $\s = \N.\init$ and $\X = D.\mathsf{init}$ to get the desired
+  parse transformer.
+
+\begin{figure}
+\begin{lstlisting}
+NtoD : (s : N.states) (*@$\to$@*)
+       (*@$\uparrow$@*)(Trace(*@$_{\N}$@*) s true (*@$\lto$@*)
+           &[ X (*@$\in$@*) D.states ] &[ sInX (*@$\in$@*) (X (*@$\ni$@*) s) ] Trace(*@$_{\D}$@*) X true)
+NtoD .s (nil s) X sInX = nil s
+NtoD .(N.src t) (N.cons trans .true a dstTrace) X sInX =
+  D.cons (N.label t) X true a
+    (((NtoD s dstTrace).(*@$\pi$@*) (D.(*@$\delta$@*) (N.label t) X)).(*@$\pi$@*) (N.dst t))
+NtoD .(N.(*@$\epsilon$@*)src t) ((*@$\epsilon$@*)cons (*@$\epsilon$@*)trans .true (*@$\epsilon$@*)DstTrace) X sInX =
+  ((NtoD s (*@$\epsilon$@*)DstTrace).(*@$\pi$@*) X).(*@$\pi$@*) (N.(*@$\epsilon$@*)dst t)
+\end{lstlisting}
+\caption{Parse transformer from NFA traces to Powerset DFA traces}
+\label{fig:NtoD}
+\end{figure}
+
+Because the state $\s$ is now arbitrary, we can define $\NtoD$ via
+recursion on the trace through the NFA. Code for this is given in \cref{fig:NtoD}.
+If the NFA trace is terminating at the accepting state $\s$ and $\s \in \X$,
+then we return a trace in $\D$ that terminates at
+$\X$, which is accepting because it contains the accepting state $\s$.
+
+If the NFA trace is defined via $\cons$ for a labeled transition
+$t : N.transitions$ such that $N.src~t \in \X$, then
+we first decompose the trace into a parse of $N.label~t$ and a
+trace starting from $N.dst~t$. We then recursively call $\NtoD$
+on this trace which returns a family of traces parameterized by $\epsilon$-closed
+subsets containing $N.\dst~t$. We instantiate this family at the
+set of states
+$\D.\delta~(N.label~t)~X$. Because $\N.\dst~t$ is
+reachable from $\N.\src~\texttt{trans}$ by a single transition labeled by
+$N.label~t$ and $N.src~t \in X$ we have that
+$N.dst~t \in \D.\delta~(N.label~t)~X$. Therefore
+we can linearly combine the parse of the label and the trace built from the
+recursive call to get a trace in $\D$ rooted at $X$.
+
+Lastly, if the NFA trace is defined via $\epsilon\cons$ for an $\epsilon$-transition
+$t : \N.\epstransitions$ such that
+$\N.\epssrc~t \in \X$, then the construction is quite
+similar. In this case we again recursively call $\NtoD$, except afterwards we
+instantiate the family of grammars at the set $X$ itself. We may do this because
+all states in $\D$ are taken to be $\epsilon$-closed, therefore
+$\N.\epsdst~t \in \X$ because
+$\N.\epssrc~t$ is in $X$. This concludes the construction of $\NtoD$.
 
 To construct a term from DFA traces to NFA traces, we similarly strengthen our
-induction hypothesis and build a term
+induction hypothesis and build a parse transformer
 \[
-  g : \ltonl {\left( \Trace_{D}(X) \lto \bigoplus\limits_{s : N.\mathsf{states}}\bigoplus\limits_{s \in X} \Trace_{N}(s) \right)}
+  \DtoN : \ltonl {\left( \Trace_{\D}~\X~\true \lto \bigoplus\limits_{s : \N.\mathsf{states}}\bigoplus\limits_{\texttt{sInX} : \X \ni \s} \Trace_{\N}~s~\true \right)}
 \]
 
-For a moment, let us turn our attention to the content behind the traces in $D$.
-Traces from subset $X$ to subset $Y$ denote the \emph{existence} of some trace
-in the $N$ from some $x \in X$ to some $y \in Y$. In general it is not clear how
-to extract out a choice of trace in $N$, which is what $g$ is asking of us.
-However, we may further take the states and transitions of $N$ to be ordered.
+For a moment, let us turn our attention to the content behind the traces in $\D$.
+Traces from subset $\X$ to subset $\Y$ describes the \emph{existence} of some trace
+in the $\N$ from some $\x \in \X$ to some $\y \in \Y$. In general it is not clear how
+to extract out a choice of trace in $\N$, which is what $\DtoN$ is asking of us.
+However, we may further take the states and transitions of $\N$ to be \emph{ordered}.
 This allows us to
-induce an order on traces in $D$ from $X$ to $Y$. From these orderings we can then
+induce an order on traces in $\D$ from $\X$ to $\Y$. From these orderings we can then
 define the necessary choice functions on each type by taking the smallest with
@@
 
-We define $g$ by induction on traces through $D$. First if the trace stops at
-some accepting $\varepsilon$-closed subset $X$, then we may choose some state
-$s \in X$ such that $s$ is accepting and terminate the trace in $N$ at $s$.
-
-Next if the trace in $D$ is defined via transition
-$X \overset{c}{\to} D.\delta_{c}(X)$ then we may choose some state in
-$s \in D.\delta_{c}(X)$. Inductively, $g$ sends the trace rooted at
-$D.\delta_{c}(X)$ to one rooted at $s$. By definition of $D.\delta_{c}(X)$,
-there exists some $s' \in X$ such that $s'$ tranitions to $s$ by some number of
-$\varepsilon$-transitions followed by a transition labeled by $c$.
+We define $\DtoN$ by recursion on traces through $\D$. First if the trace stops at
+some accepting $\epsilon$-closed subset $\X$, then we may choose some state
+$\s \in X$ such that $\s$ is accepting and terminate the trace in $\N$ at $\s$.
+
+Next if the trace in $\D$ is defined via transition
+$\X \overset{\stringquote{c}}{\to} (\D.\delta~\c~\X)$ then we may choose some state in
+$\s \in \D.\delta~\c~\X$. Recursively, $\DtoN$ sends the trace rooted at
+$\D.\delta~\c~\X$ to one rooted at $\s$. By definition of $\D.\delta~\c~\X$,
+there exists some $\s' \in \X$ such that $\s'$ transitions to $\s$ by some number of
+$\epsilon$-transitions followed by a transition labeled by $\c$.
 We order all
 such traces of this form and choose the smallest one, which is then linearly
-combined with the parse of $c$ to build an $N$ trace. While using
-$N.\mathsf{cons}$ to build this final trace, we appropriately shift the indexing
-data to $s'$, which is adherent to the defintion of $D.\delta_{c}(X)$. This concludes the
-definition of $g$.
-
-\[\begin{tikzcd}
-    {c \otimes \mathsf{Trace}_{D}(D.\delta_{c}(X))} &
-    {c \otimes \bigoplus\limits_{s : N.\mathsf{states}} \bigoplus\limits_{s \in D.\delta_{c}(X)} \Trace_{N}~s} \\
-    {\bigoplus\limits_{s' : N.\mathsf{states}} \bigoplus\limits_{s' \in X} \Trace_{N}~s'} &
-    {\bigoplus\limits_{s : N.\mathsf{states}} \bigoplus\limits_{s \in D.\delta_{c}(X)} \left( c \otimes \Trace_{N}~s \right)}
-	\arrow["{id\otimes g}", from=1-1, to=1-2]
-	\arrow["{\mathsf{dist}}", from=1-2, to=2-2]
-	\arrow["{}", from=2-2, to=2-1]
-\end{tikzcd}\]
-
-Therefore $\Trace_{N}(N.\mathsf{init})$ and $\Trace_{D}(D.\mathsf{init})$ are
-weakly equivalent, and thus $N$ and $D$ recognize the same language of strings.
+combined with the parse of $\c$ to build an $\N$ trace. This concludes the
+definition of $\DtoN$, thus showing that $Parse_{N}$ and $Parse_{D}$ are
+weakly equivalent.
 \end{proof}
 
+\newcommand{\R}{\texttt{R}}
 \begin{theorem}[Thompson's Construction, \Agda]
-  Given a regular expression $R$, there exists an NFA $N$ such that $R$ is
-  strongly equivalent to $\Trace_{N}(N.\init)$.
+  Given a regular expression $\R$, there exists an NFA $\N$ such that $\R$ is
+  strongly equivalent to $\Trace_{N}(\N.\init)$.
 \end{theorem}
@@
 \begin{proof}
-In this proof, we present a variant of Thompson's construction
-\cite{thompsonProgrammingTechniquesRegular1968} in which we \max{I induct this sentence into the hall of shame for misuse of the word ``induct''}induct on the
-structure of $R$ to construct strongly equivalent NFAs. We will sketch this
-proof by walking through some of the inductive cases, the remaining cases follow
+  \steven{TODO this proof needs to be rewritten. Likely through use of equalizers}
+  \pedro{Since this proof and the determinization one are well-known and we have
+    proved it in Agda, it's probably a good idea to include a much shorter proof sketch,
+    or move it to the appendix, though this would feel a bit superfluous.}
+We present a variant of Thompson's construction
+\cite{thompsonProgrammingTechniquesRegular1968} in which we recurse on the
+structure of $\R$ to construct a strongly equivalent NFAs. In this paper we will walk through
+some of the cases of this proof. The remaining cases
+follow
 similar reasoning and the details may be found in the formalized artifact.
@@
   \begin{tikzpicture}[node distance = 17mm ]
-    \node[state, initial] (0) {$0$};
-    \node[state, right of=0, accepting] (1) {$1$};
-
-    \path[->] (0) edge[above] node{$c$} (1);
+    \node[state, initial] (0) {$\0$};
+    \node[state, right of=0, accepting] (1) {$\1$};
+
+    \path[->] (0) edge[above] node{$\stringquote{c}$} (1);
   \end{tikzpicture}
-  \caption{NFA $N_{c}$ for a single character $c$}
+  \caption{NFA $\N_{\c}$ for a single character $\c$}
   \label{fig:litnfa}
@@
 
-First handle the case where $R$ is a literal character $c$. The NFA $N_{c}$ pictured in
+First handle the case where $R$ is a literal character $\literal c$. The NFA $N_{c}$ pictured in
 \cref{fig:litnfa} is strongly equivalent to the grammar $c$. To demonstrate this
-strong equivalence, we build terms between $c$ and $\Trace_{N_{c}}(N_{c}.\init)$
+strong equivalence, we build functions between $\literal c$ and $Parse_{N_{c}}$
 and show they are mutually inverse.
 
-Define $f : \ltonl {(c \lto \Trace_{N_{c}}(N_{c}.\init))}$ as follows,
-
-\[
-  a : c \vdash f := \cons(a, \nil) : \Trace_{N_{c}}(N_{c}.\init)
-\]
-
-Define $g : \ltonl {(\Trace_{N_{c}}(N_{c}.\init) \lto c)}$ via induction on the
-trace. Map traces beginning in state 1, which are necessarily built via $\nil$,
-to the linear unit $I$. Then we must handle traces beginning in state 0, which
-are necessarily built via $\cons$. We decompose such a trace into a parse of the
-character $c$ and a trace beginning in state 1. We inductively transform the
-trace beginning in state 1 into a term of type $I$. Finally, we use the right
-unitor to turn a term of type $c \otimes I$ into one of type $c$.
-
-\[\begin{tikzcd}
-    {c \otimes \mathsf{Trace}_{N_{c}}(1)} &
-    {c \otimes I} &
-    c
-	\arrow["{id\otimes g}", from=1-1, to=1-2]
-	\arrow["{}", from=1-2, to=1-3]
-\end{tikzcd}\]
-
-Note that the grammar $c$ is unambiguous, so $g \circ f$ is equal to the
-identity on $c$. Additionally, we can show that $f$ carries the structure of an
-indexed algebra homomorphism for the family of functors defining the trace type
-$\Trace_{N_{c}}$. Because $g$ is defined via induction, it carries the structure
-of an algebra homomorphism as well. Thus, $f \circ g$ is also an algebra
-homomorphism. $f \circ g$ and the identity map are both then algebra
-homomorphisms from $\Trace_{N_{c}}$ to itself. Because $\Trace_{N_{c}}$ is the
-initial algebra for the relevant family of functors, there is only one
-homomorphism from it to itself and thus $f \circ g = \mathsf{id}$. Therefore,
-$f$ and $g$ are mutually inverse and the regular expression is strongly
-equivalent to the traces through the NFA in this case.
-
+\begin{figure}
+\begin{lstlisting}
+A : (s : Fin 2) (*@$\to$@*) L
+A 0 = 'c'
+A 1 = I
+
+toN(*@$_c$@*) : (s : Fin 2) (*@$\to$@*) (*@$\uparrow$@*)(A s (*@$\lto$@*) Trace(*@$_{N_c}$@*) s true)
+toN(*@$_c$@*) 0 c = cons t true c (nil 1)
+toN(*@$_c$@*) 1 () = nil 1
+
+fromN(*@$_c$@*) : (s : Fin 2) (*@$\to$@*) (*@$\uparrow$@*)(Trace(*@$_{N_c}$@*) s true (*@$\lto$@*) A s)
+fromN(*@$_c$@*) 0 (cons t .true c tr) = let a (*@$\otimes$@*) () = c (*@$\otimes$@*) fromN(*@$_c$@*) 1 tr in a
+fromN(*@$_c$@*) 1 (nil .1) = ()
+\end{lstlisting}
+\caption{Parse transformer out of $Trace_{N_{c}}~s~true$}
+\label{fig:litNFAterms}
+\end{figure}
+
+In \cref{fig:litNFAterms} we define a family of grammars $A$ that serve as
+interpretations for each state in $N_{c}$. We then prove that each $A~s$ is
+equivalent to the accepting traces through $N_{c}$ out of state $s$.
+
+When
+defining $fromN_{c}$ we pattern match on the accepting traces beginning in each
+state. Out of state $0$ the only accepting traces are built via $cons$ over the
+single transition $t$, and out
+of state $1$ the only accepting trace is built via $nil$.
+
+Each $A~s$ is unambiguous, so $(fromN_{c}~s) \circ (toN_{c}~s)$ is equal to the
+identity on each $A~s$. To complete the proof of strong equivalence it remains
+to show that $(toN_{c}~s) \circ (fromN_{c}~s)$ is equal to the identity on
+$Trace~s~true$, which we show by using the equalizer type.
+
+\begin{figure}
+\begin{lstlisting}
+eqHom(*@$_{N_c}$@*) : (s : N(*@$_c$@*).state) (*@$\to$@*)
+      (*@$\uparrow$@*)(Trace(*@$_{N_c}$@*) s true (*@$\lto$@*) {tr | (toN(*@$_c$@*) s (*@$\circ$@*) fromN(*@$_{c}$@*) s) tr = id tr})
+\end{lstlisting}
+\caption{Signature for the equalizer term}
+\label{fig:litequalizer}
+\end{figure}
+
+The type signature for the family of functions, $eqHom_{N_{c}}$, into the equalizer type is given in
+\cref{fig:litequalizer}. By defining an $Trace_{N_{c}}$-algebra homomorphism
+into the equalizer, we enable a
+proof strategy for equality by induction on traces through $N_{c}$. First consider the traces
+out of state $1$. The only case to handle here is $nil 1$, and by definition
+$toN_{c}~1$ and $fromN_{c}~1$ invert each other.
+When defining $eqHom_{N_{c}}$ for traces out of state $0$, we must leverage an
+inductive hypothesis,
+
+\begin{align*}
+  (toN_{c}~0 \circ fromN_{c}~0) (cons~t~true~c~tr)
+  & \equiv (toN_{c}~0) (let~a~\otimes~()~=~c~\otimes~fromN_{c}~1~tr~in~a) \\
+  & \equiv cons~t~true~(let~a~\otimes~()~=~c~\otimes~fromN_{c}~1~tr~in~a)~(nil~1) \\
+  & \equiv cons~t~true~c~(nil~1) \\
+  & \equiv cons~t~true~c~tr \\
+\end{align*}
+
+The first two equalities are by defintion. The thrid equality holds because
+$\literal c$ is unambiguous and
+$let~a~\otimes~()~=~c~\otimes~fromN_{c}~1~tr~in~a$ and $c$ have type
+$\literal c$. The final equality holds because our inductive hypothesis states
+that $nil~1 \equiv tr$. Thus, each $A s$ and $Trace_{N_{c}}~s~true$ are strongly
+equivalent --- in particular, $\literal c$ and $Parse_{N_{c}}$.
+
+\steven{Rewrite the disjunction case with equalizers}
 Now consider the case where $R$ is the disjunction of two regular expressions
@@
 it suffices to show that
-$\Trace_{N_{1}}(N_{1}.\init) \oplus \Trace_{N_{2}}(N_{2}.\init)$ is strongly
-equivalent to $\Trace_{N_{\oplus}}(N_{\oplus}.\init)$.
+$Parse_{N_{1}} \oplus Parse_{N_{2}}$ is strongly
+equivalent to $Parse_{N_{\oplus}}$.
 
@@
 %% 3. Dyck language
-\newcommand{\balanced}{\mathsf{balanced}}
-\newcommand{\Dyck}{\mathsf{Dyck}}
+\newcommand{\balanced}{\mathtt{balanced}}
+\newcommand{\Dyck}{\mathtt{Dyck}}
 Moving up the Chomsky hierarchy from regular grammars, we arrive at context-free
-grammars (CFGs). \max{In many ways?}In many ways, the smallest nontrivial CFG is the Dyck grammar
-$\Dyck$ --- the grammar of balanced parentheses. Moreover, the Chomsky-Sch\"utzenberger
-representation theorem \cite{CHOMSKY1959118} demonstrates that $\Dyck$, and grammars like it,
-cut to the essence of what it means to be context free.
+grammars (CFGs). An interesting nontrivial CFG is the Dyck grammar
+$\Dyck$ --- the grammar of balanced parentheses.
 
@@
 and is defined in \cref{fig:dyckinductive}. The $\nil$ constructor shows that
-the empty string is balanced, and the $\balanced$ constrcutors builds a balanced
+the empty string is balanced, and the $bal$ constructor builds a balanced
 parse by wrapping an already balanced parse in an additional set of parentheses
-then following it with another balanced parse.
-
+then following it with another balanced parse. We construct a parser for $\Dyck$ by building a deterministic automaton $M$ such
+that $Parse_{M}$ is strongly equivalent to $\Dyck$.
 \begin{figure}
-\begin{align*}
-\data &~\Dyck : L~\where\\
-      & \nil : \ltonl {\Dyck} \\
-      & \balanced : \ltonl {(\stringquote{(} \lto \Dyck \lto \stringquote{)} \lto \Dyck \lto \Dyck)}
-\end{align*}
-\caption{Kleene Star as an inductive type}
+\begin{lstlisting}
+data Dyck : L where
+  nil : (*@$\uparrow$@*) Dyck
+  bal : (*@$\uparrow$@*)('(' (*@$\lto$@*) Dyck (*@$\lto$@*) ')' (*@$\lto$@*) Dyck (*@$\lto$@*) Dyck)
+\end{lstlisting}
+\caption{The Dyck grammar as an inductive linear type}
 \label{fig:dyckinductive}
 \end{figure}
-
-\steven{somewhere in here, sneak in the benchmark where the Dyck parser could
-  handle 10's of thousands of characters in about 30s}
-
-We construct a parser for $\Dyck$ by building a determinstic automaton $M$ such
-that $\Trace_{M}(M.\init) \cong \Dyck$.
 
@@
 open parentheses have been seen so far. We may think of
-$M$ as performing the same task as a stack machine, except instead of storing
+$M$ as performing the same task as a stack machine through slightly different means. Instead of storing
 the number of open parentheses in a distinguished stack of memory, $M$ operates
-by storing this data in the current state. This is made feasible by specifying
-$M$ almost identically to how we defined DFAs, except we omit the
-finiteness condition on the states of the automaton.
-
-\steven{Mention that $\Trace_{M}$ is iso to string. Mention the same for DFAs above }
-
-% Unlike the finite automata seen thus far, $M$ has infinitely many states --- one
-% for each $n : \mathbb{N}$ with an additional fail state. This diverges a bit
-% from classical language theory, where the recognizer for $D$ would be encoded as a
-% pushdown automaton with finitely many
-% states and stack of memory. Such PDAs are encodable in our theory, although they
-% are a little cumbersome to use. Moreover, they
-% seem overfit to classical reasoning as they don't preserve the grammar up to
-% strong equivalence, only weak equivalence in general. \steven{cite Lambek
-%   automata and state that is is a secondary goal to preserve as much structure as possible}
-% By moving the stack of memory (the nats) into the
-% state, we obtain infinitely many states, which is fine bc the states need only
-% be a type for us to interface with it.
-
-\steven{TODO sketch proof of equivalence. Mention that it is mechanized}
+by storing this data in the current state. $M$ is defined almost identically to how we defined DFAs, except we omit the
+finiteness condition on the states of the automaton. This similarity allows
+functions $parse_{M}$ and $print_{M}$ to be defined analogously to
+\cref{fig:dfaparser,fig:printdfa}, and so
+$\LinSigTy {s} {M.states} {\LinSigTy {b}{Bool} {Trace_{M}~s~b}}$ is likewise unambiguous.
+
+\begin{theorem}[\Agda]
+  \label{thm:dyck}
+  $Dyck$ and $Parse_{M}$ are strongly equivalent.
+\end{theorem}
+
+We omit the details of \cref{thm:dyck} in this paper, and the ful proof is
+mechanized in our Agda development. We can build a $Dyck$ parser by utilizing
+this artifact, and the
+runtime of the $Dyck$ parser is about 30 seconds for input that is 25,000
+characters long.
+\pedro{It might be a good idea to mention explicitly in the introduction
+  that we are not claiming that our verified parsers are competitive with
+  the state of the art.}
 
@@
     \node[right of=2] (3) {$\dots$};
-    \node[state, below of=0] (fail) {$\mathsf{fail}$};
+    \node[state, below of=0] (fail) {$fail$};
 
@@
 
-\subsection{Turing Machines}
-%% 5. Expressiveness: Turing Machines
-The expressive power of \theoryabbv spans the entire Chomsky hierachy, as we may
-encode the acccepting traces of Turing machines as linear types.
-
-\newcommand{\Turing}{\mathsf{Turing}}
-\newcommand{\snoc}{\mathsf{snoc}}
-
-\begin{figure}
-\begin{align*}
-\data &~\Turing~(t : \mathbb{Z} \to \Gamma): L~\where\\
-      & \nil : \ltonl {\left(\LinPiTy {t} {\mathbb{N} \to \Gamma} {\Turing~t} \right)} \\
-      & \snoc : \ltonl {\left( \LinPiTy {c}{\Sigma}{\LinPiTy {t} {\mathbb{N} \to \Gamma} {\Turing(\mathsf{push}(c , t)) \lto c \lto \Turing(t)}} \right)}
-\end{align*}
-\caption{Kleene Star as an inductive non-linear type}
-\label{fig:turinginductive}
-\end{figure}
-
-We nonlinearly encode the tape of a (one-sided) Turing machine as
-function from $\mathbb{N} \to \Gamma$, where $\Gamma$ is the tape alphabet.
-
-We may then define a term of type
-$\mathsf{parseTape} : \ltonl {\left( \stringg \lto \LinSigTy {t} {\mathbb{N} \to \Gamma} {\Turing(t)} \right)}$.
-In this manner, all of the interesting computations that the Turing machine
-would do occur solely inside of the nonlinear index. That is,
-$\mathsf{parseTape}$ does not discern which strings are accepted by the Turing
-machine, nor does it do any computation. $\mathsf{parseTape}$ simply loads the
-input string into the initial spaces on the tape.
-
-Once the string has been loaded onto the tape, we may further define the
-following term to run the Turing machine.
+\subsection{Unrestricted Grammars}
+\newcommand{\stringNL}{String'}
+\steven{Need to better introduce nonlinear type of strings. We haven't talked
+  about semantics nor have we included a nonlinear type of strings in the syntax}
+Define $\stringNL$ to be the non-linear list type over the alphabet $\Sigma$ and
+let $P : \stringNL \to U$ be a mapping from the set of strings to the universe of
+nonlinear types. The function $P$ describes a family of nonlinear types
+dependent on $\stringNL$.
+We view the image of $P$ as a set of parse trees for each input string, and thus
+$P$ carries the structure of a formal grammar. However, even though $P$ describes this grammatical data, it is not
+a first-class linear type in \theoryabbv. To address this gap, below we define a linear type that captures the same
+semantic grammar as $P$,
+
 \[
-  \mathsf{run} : \ltonl {\left( \bigoplus\limits_{t : \mathbb{N} \to \Gamma} \Turing(t) \lto \bigwith\limits_{\mathsf{fuel} : \mathbb{N}} \mathsf{Maybe} \left( \bigoplus\limits_{t : \mathbb{N} \to \Gamma} \bigoplus\limits_{b : \Bool} \bigoplus\limits_{\mathsf{trace} : \mathsf{TuringTrace}(b , t)} \Turing(t) \right) \right)}
+Reify~P = \LinSigTyLimit {w} {\stringNL} {\LinSigTyLimit {x} {P~w} {~\lceil w \rceil}}
 \]
 
-where $\mathsf{TuringTrace}$ is a nonlinear type denoting terminating traces
-through the Turing machine. Note that $\mathsf{run}$ returns a $\mathsf{Maybe}$
-type in a fuel parameter. Turing machines may not terminate. If they do
-terminate within $\mathsf{fuel}$-many steps, then we return the trace along with
-the current tape and a boolean denoting if the trace is accepting or rejcting.
-If the trace does not terminate within the alloted amount of fuel, then we
-timeout and return nothing.
+\steven{Probably put this ceiling business around the discussion of the linear
+  type of strings. Is this in section 3?}
+
+where $\lceil \stringquote{} \rceil = I$ and
+$\lceil c :: w \rceil = \literal c \otimes \lceil w \rceil$.
+
+This reification operation on functions $\stringNL \to U$ is incredibly expressive,
+as it allows to sidestep our linear typing connectives and utilize the whole of nonlinear
+dependent type theory to define a grammar. This power allows us to encode
+arbitrary grammars. In particular, given a
+Turing machine $Tur$ one may
+define a predicate $accepts : \stringNL \to U$ such that $accept~w$ is equal to $\top$ if
+$T$ accepts $w$ and equal to $\bot$ otherwise. Then, $Reify~accepts$ is a linear
+type that captures precisely the strings accepted by $Tur$. That is,
+$Lang(Reify~accepts)$ is recursively enumerable --- the most general class of
+languages in the Chomsky hierarchy.
 
 \subsection{Semantic Actions}
-\steven{TODO edit this. synthesize it with the later presentation of semantic
-  actions as models, and then remove the model section}
-All constructions discussed until now detail how a string may be parsed into a
-\emph{concrete} syntax tree; however, this is often not the representation that later
-compiler passes would take in. Instead, they often take in an \emph{abstract}
-syntax tree that forgets syntactic details that are superfluous to the later
-passes. For example, after using parentheses to disambiguate a parse of a string
-into a concrete syntax tree, we may remove these parentheses to have a cleaner
-structured representation of the underlying data. The procedure that takes in a
+So far, each construction has been primarily concerned with the parsing of a
+string into a \emph{concrete} syntax tree. However, in a larger verified
+developement --- such as a compiler --- a parser will often emit an
+\emph{abstract} syntax tree that omits superflous syntactic details that aren't
+needed for later passes.  The procedure that takes in a
 concrete syntax tree and returns an abstract one is referred to as a
@@
 
-Given a linear type $A$, the inhabitants of $A$ serve as concrete syntax. We may
-then chose a non-linear type $X$ that serves as abstract syntax. A semantic
-action can then be defined as $\square (A \lto \Delta X)$ where
-$\Delta X = \LinSigTy x X \top$.
-\steven{Note here that even though $\Delta$ is a map from nonlinear types into
-  linear ones, it is very different from $F$/whatever we call that half of the adjunction}
-
-% \steven{Everything in this section below this marker is outdated and needs to be
-% folded into the above or cut}
-
-% \steven{Need a different segue because canonicity is no longer before this}
-% The canonicity theorem from the previous section gives a tight connection
-% between terms in our type theory and the theory of formal languages. In this section
-% we further explore these connections by using the fact that, classically
-% formal language theory is closely related to the study of automata.
-% In the Chomsky hierarchy, each language class is associated to a class of
-% automata that serve as recognizers. Internal to \theoryabbv,
-% we can characterize these language classes syntactically; moreover, we
-% demonstrate the equivalence of these language classes to its associated automata class as a
-% proof term within our logic.
-
-% \subsection{Non-deterministic Finite Automata}
-% \label{subsec:finiteaut}
-% \begin{figure}
-%   \begin{tikzpicture}[node distance = 25mm ]
-%     \node[state, initial, accepting] (1) {$1$};
-%     \node[state, below left of=1] (2) {$2$};
-%     \node[state, right of=2] (3) {$3$};
-
-%     \path[->] (1) edge[above] node{$b$} (2)
-%               (1) edge[below, bend right, left=0.3] node{$\varepsilon$} (3)
-%               (2) edge[loop left] node{$a$} (2)
-%               (2) edge[below] node{$a, b$} (3)
-%               (3) edge[above, bend right, right=0.3] node{$a$} (1);
-%   \end{tikzpicture}
-%   \caption{An example NFA}
-%   \label{fig:NFA}
-% \end{figure}
-
-% Classically, a \emph{nondeterministic finite automaton} (NFA) is a finite state machine where
-% transitions are labeled with characters from a fixed alphabet $\Sigma$. These
-% are often represented formally as a 5-tuple $(Q, \Sigma, \delta, q_{0}, F)$,
-
-% \begin{itemize}
-%   \item $Q$ a finite set of states
-%   \item $\Sigma$ a fixed, finite alphabet
-%   \item $\delta : Q \times (\Sigma \cup \{ \varepsilon\}) \to \mathcal{P}(Q)$ the labeled transition function
-%   \item $q_{0} \in Q$ the start state
-%   \item $F \subset Q$ a set of accepting states
-% \end{itemize}
-
-% \subsection{A DFA Parser}
-% \label{subsec:regexparser}
-
-% Just as we encoded traces of NFAs as grammars, we likewise
-% encode the traces of a DFA as grammars. The key difference
-% between NFAs and DFAs is \emph{determinism} --- meaning,
-% that in a state $q$ inside of DFA, given a character $c$ there
-% will be exactly one transition that we may take leaving $q$
-% with label $c$. For us, this changes the definition of valid
-% transitions for a DFA, instead of the definition of
-% $\mathsf{Trans}$ provided in \cref{subsec:finiteaut} DFAs
-% obey
-
-% \begin{gather*}
-%  \mathsf{State} \in \{ g_{q} : q \in Q \} \\
-%  \mathsf{Trans}(q) ::= \bigoplus_{c \in \Sigma} (c \otimes \mathsf{State})
-% \end{gather*}
-
-% Meaning, each state has a transition for every character. \emph{Note:} the above use
-%   of a disjunction over $\Sigma$ is a bit of abuse of notation.
-%   As $\Sigma$ is a finite alphabet, we wish to think of this as an iterated
-%   binary sum, or perhaps as a finitely indexed sum defined via linear dependent
-%   sums in the same manner by which we defined binary sums.
-
-% We now wish to define a parser term for DFA grammars. In
-% particular, for a DFA $D$ we want to build a term,
-
-% \[
-%   w : \String \vdash \mathsf{parse}_{D} : \mathsf{AccTrace}_{D} \oplus \top
-% \]
-
-% where left injection into the output type denotes acceptance
-% by the parser, and right injection denotes rejection. To
-% build such a parser, it will suffice to construct a term
-
-% \[
-%   w : \String \vdash \mathsf{parse}_{D} : \LinSigTy q Q {\mathsf{Trace}_{D}(q_{0} , q)}
-% \]
-% This is because given a trace of a DFA, we may easily check
-% if we should accept or reject by simply testing
-% if the final state is accepting.
-
-% Because $w$ is a Kleene star of characters, we may construct
-% our desired $\mathsf{parse}_{D}$ term as a $\mathsf{foldl}$
-% over $w$. In the empty case, we just have the trace that
-% ends at the accepting state. In the recursive case, we
-% effectively add to our trace by transitioning one character
-% at a time, as we read them moving across $w$.
-
-% Perhaps this derivation is not too surprising. All it says
-% is that a DFA may be ran by transitioning a single character
-% at a time, and then accepting or rejecting based on the
-% final state. This is exactly what DFAs did initially, so
-% what did we gain? Well, this has the benefit of our type
-% system to ensure its correctness. Moreover, this construction exports to an
-% intrinsically verified and executable decision procedure for DFAs in \theoryabbv
-% embedded in Agda.
-
-% We should note that the
-% construction of a DFA parser requires the addition of an additional but seemingly innocuous
-% axiom.
-
-% \[
-%   \top \cong \left( \bigoplus_{c : \Sigma} c \right)^{*}
-% \]
-
-% That is, the characters in the alphabet $\Sigma$ really are all of the
-% characters. The need for this axiom arises when we go to take a transition
-% within the DFA,
-% as we must ensure that the next character of the string has a corresponding
-% transition from the current state
-% --- which would be guaranteed by determinism provided there are no surprise characters.
-
-% \subsection{Regular Expressions and DFAs}
-% \label{subsec:deriv}
-
-% In order to extend the DFA
-% parser from \cref{subsec:regexparser} to the construction of
-% a verified parser
-% generator for regular expression we need to perform some
-% plumbing establishing an equivalence between regular
-% expressions and DFAs.
-
-% There are several routes we may hope to take in establishing
-% this equivalence. First, we could prove an equivalence
-% between NFAs and regular expressions, and separately prove
-% an equivalence between NFAs and DFAs.
-% In \cref{subsec:eqproofs}, we include a version of
-% Thompson's construction --- which established the
-% equivalence between regular grammars and DFAs. We may additionally
-% hope to internalize a variant of the powerset construction \cite{rabinFiniteAutomataTheir1959}
-% --- which takes as input an NFA and constructs a DFA that
-% recognizes the same language --- and combine the results of
-% Thompson's construction and the powerset constructions to give an equivalence
-% between regular expressions and DFAs.  This route is alluring, as it
-% internalizes several classic grammar-theoretic constructions. However, it may
-% necessitate extensions to the LNL theory, like
-% a propositional truncation, and we have not yet investigated
-% how this would interact with the existing types in the
-% theory. The addition of a propositional truncation may seem
-% harmless, but it is not always immediately clear how
-% distinct constructions will interact. For instance, when
-% exploring LNL models, Benton discovered that the synthesis
-% of linear and dependent types require a new presentation of
-% the $!$ modality from linear logic
-% \cite{bentonMixedLinearNonlinear1995}. That is all to say,
-% this is a work in progress and
-% it is not immediate that the addition of a propositional
-% truncation is adequate for establishing the weak equivalence
-% between NFAs and DFAs.
-
-% We may instead hope to internalize an equivalence between
-% regular grammars and DFAs by using Brzozowski derivatives to
-% directly create a DFA that is weakly equivalent to a given
-% regular expression, as described by Owens et al.
-% \cite{owensRegularexpressionDerivativesReexamined2009}.
-% One characterization of regular grammars is that they are
-% precisely those grammars which have finitely many inequivalent Brzozowski
-% derivatives
-% \cite{brzozowskiDerivativesRegularExpressions1964}.
-% The algorithm used by Owens takes in a
-% regular grammar and generates a DFA that recognizes the same
-% language, and the states in this DFA are the finitely many
-% derivative equivalence classes. We initially had a version of
-% this theorem very roughly internalized in the LNL theory.
-% To our taste, too much of this presentation relied on
-% meta-arguments that lived outside of
-% our formalism, and thus this particular phrasing of the
-% theorem did not translate well into formalization.
-
-% In any case, we believe
-% that revisiting these lines of thought will lead to a
-% satisfactory internalization of the equivalence between
-% regular grammars and DFAs, and thus would bridge the gap
-% between our DFA parser and a full regular expression parser.
-
-% \subsection{Equivalence Between Regular Expressions and Finite Automata}
-% \label{subsec:eqproofs}
-% In this section, we describe a version of Thompson's
-% construction \cite{thompsonProgrammingTechniquesRegular1968}
-%    where we construct an NFA that recognizes a given regular
-% expression. Moreover, we will show that this NFA is strongly equivalent to the
-% original grammar. Witnessing this construction in our syntax has two benefits
-% \begin{enumerate}
-%   \item It reinforces this high-level view that the syntax is a natural and
-%         general setting for formal grammar reasoning, as we demonstrate that
-%         this formal system subsume results from existing systems, and
-%   \item Following the
-%         development of Thompson's construction, we then need
-%         only establish the equivalence of NFAs and DFAs to
-%         complete the full regular expression parser
-% \end{enumerate}
-
-% \begin{theorem}[Thompson]
-%   \label{thm:thompson}
-%   For $g$ a regular grammar $g$, there is an NFA $N$ that recognizes the same
-%   language as $g$.
-% \end{theorem}
-% We make a pretty straightforward adaptation of Thompson's theorem to our setting,
-
-% \begin{theorem}[Typed Thompson]
-%   \label{thm:typthompson}
-%   For $g$ a regular expression $g$, there is an NFA $N$ such that $g$ is isomorphic
-%   to $\mathsf{AccTrace}_{N}$.
-% \end{theorem}
-
-% \begin{proof}[Proof Sketch]
-%   Recall that regular expressions are inductively defined via
-%   disjunction, concatenation, and Kleene star over literals
-%   and the empty grammar. By induction over regular expressions,
-%   we will construct an NFA that is equivalent to $g$.
-
-%   First, define the recognizing NFA for the empty grammar
-%   $I$.
-
-%   \begin{figure}[h!]
-%   \begin{tikzpicture}[node distance = 25mm ]
-%     \node[state, initial] (1) {$1$};
-%     \node[state, right of=1, accepting] (2) {$2$};
-
-%     \path[->] (1) edge[below] node{$\varepsilon$} (2);
-%   \end{tikzpicture}
-%   \caption{$NFA(I)$}
-%   \label{fig:emptyNFA}
-%   \end{figure}
-
-%   The type of traces from the initial state of $NFA(I)$ to the single
-%   accepting state is given by,
-
-%   \[
-%     \mathsf{Trace}_{NFA(I)}(q_{1}, q_{2}) = \mu
-%       \begin{pmatrix}
-%          g_{q_{1}} := g_{q_{2}} \\
-%          g_{q_{2}} := I
-%       \end{pmatrix}. g_{q_{1}}
-%   \]
-
-%   The accepting traces through $NFA(I)$ are then described
-%   as,
-
-%   \[
-%     \mathsf{AccTrace}_{NFA(I)} = \LinSigTy q {\{1 , 2\}} {\left( \mathsf{Trace}_{NFA(I)}(q_{1} , q) \pair \mathsf{acc}(q) \right)}
-%   \]
-
-%   A quick inspection of \cref{fig:emptyNFA} reveals that the
-%   only reasonable choice for $q$ is $q_{2}$ --- because
-%   state 2 is accepting while state 1 is not. Therefore,
-
-%   \begin{align*}
-%     \mathsf{AccTrace}_{NFA(I)}
-%     & \cong \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2}) \pair \mathsf{acc}(q_{2}) \\
-%     & \cong \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2}) \pair \top \\
-%     & \cong \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2})
-%   \end{align*}
-
-%   From here, to prove
-%   $I \cong \mathsf{AccTrace}_{NFA(I)}$ it suffices to show
-%   $I \cong \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2})$. Below we
-%   give two parse transformers, one from $I$ to
-%   $\mathsf{Trace}_{NFA(I)}(q_{1} , q_{2})$ and vice versa.
-
-%   \[
-%     \inferrule
-%     {p : I \vdash \mathsf{nil} : \mathsf{Trace}(q_{2} , q_{2}) \\
-%      \exists \text{~transition~} q_{1} \overset{\varepsilon}{\to} q_{2}
-%     }
-%     {p : I \vdash \mathsf{\varepsilon cons}(\mathsf{nil}) : \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2})}
-%   \]
-
-%   Let $\gamma$ be the substitution $\{ g_{q_{2}} / g_{q_{1}}, I / g_{q_{2}} \}$,
-
-%   \[
-%     \inferrule
-%     {
-%       p : \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2}) \vdash p :
-%         \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2}) \\
-%       x_{1} : \simulsubst {g_{q_{2}}} {\gamma} = I \vdash x_{1} : I \\
-%       x_{2} : \simulsubst {I} {\gamma} = I \vdash x_{2} : I
-%     }
-%     {
-%       p : \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2}) \vdash \mathsf{mfold}(x_{1}.x_{1} , x_{2}.x_{2})(p) : I
-%     }
-%   \]
-
-%   This concludes the proof for the case of the empty
-%   grammar. Let's now walk through the construction for
-%   literal grammars. Given a character $c$, we construct an
-%   NFA that recognizes only the string containing the single
-%   character $c$ as,
-
-%   \begin{figure}[h!]
-%   \begin{tikzpicture}[node distance = 25mm ]
-%     \node[state, initial] (1) {$1$};
-%     \node[state, right of=1, accepting] (2) {$2$};
-
-%     \path[->] (1) edge[below] node{$c$} (2);
-%   \end{tikzpicture}
-%   \caption{$NFA(c)$}
-%   \label{fig:literalNFA}
-%   \end{figure}
-
-%   The automaton in \cref{fig:literalNFA} induces the
-%   following type of traces from $q_{1}$ to $q_{2}$.
-
-%   \[
-%     \mathsf{Trace}_{NFA(c)}(q_{1}, q_{2}) = \mu
-%       \begin{pmatrix}
-%          g_{q_{1}} := c \otimes g_{q_{2}} \\
-%          g_{q_{2}} := I
-%       \end{pmatrix}. g_{q_{1}}
-%   \]
-
-%   Through the same argument as the empty grammar, the only
-%   state that is accepting is $q_{2}$ and thus,
-
-%   \[
-%     \mathsf{AccTrace}_{NFA(c)} \cong \mathsf{Trace}_{NFA(c)}(q_{1} , q_{2})
-%   \]
-
-%   To show the desired isomorphism of $c \cong NFA(c)$ we
-%   make a similar argument as we did for the empty grammar
-%   $I$, except we leverage the $\mathsf{cons}$ rule instead
-%   of $\mathsf{\varepsilon cons}$. That is, the parse
-%   transformers in either direction are given as,
-
-%   \[
-%     \inferrule
-%     {\cdot \vdash \mathsf{nil} : \mathsf{Trace}(q_{2} , q_{2}) \\
-%      \exists \text{~transition~} q_{1} \overset{c}{\to} q_{2}
-%     }
-%     {p : c \vdash \mathsf{cons}(\mathsf{nil}) :
-%       \mathsf{Trace}_{NFA(c)}(q_{1} , q_{2})}
-%   \]
-
-%   \[
-%     \inferrule
-%     {
-%       p : \mathsf{Trace}_{NFA(c)}(q_{1} , q_{2}) \vdash p :
-%         \mathsf{Trace}_{NFA(c)}(q_{1} , q_{2}) \\
-%       x_{1} : \simulsubst {(c \otimes g_{q_{2}})} {\gamma} = c \otimes I \vdash \mathsf{unitR}(x_{1}) : c \\
-%       x_{2} : \simulsubst {I} {\gamma} = I \vdash x_{2} : I
-%     }
-%     {
-%       p : \mathsf{Trace}_{NFA(c)}(q_{1} , q_{2}) \vdash \mathsf{mfold}(x_{1}.\mathsf{unitR}(x_{1}) , x_{2}.x_{2})(p) : c
-%     }
-%   \]
-
-%   Where $\gamma$ is the substitution
-%   $\{ c \otimes g_{q_{2}} / g_{q_{1}}, I / g_{q_{2}} \}$ and
-%   $\mathsf{unitR}$ is a witness to the isomorphism
-%   $c \otimes I \cong c$. Again, we may see that these do
-%   indeed mutually invert each other in the Agda code.
-
-%   It remains to show that the desired isomorphisms are
-%   preserved by $\otimes$, $\oplus$, and Kleene star. Here,
-%   we will give the argument for the disjunction case, the
-%   others are defined quite similarly.
-
-%   Given two NFAs $N$ and $M$, we may define a new NFA that
-%   encodes the disjunction of $N$ and $M$. Denote the
-%   internal states of $N$ by $q_{j}$'s and the internal states
-%   of $M$ by $r_{k}$'s,
-
-%   \begin{figure}[h!]
-%   \begin{tikzpicture}[node distance = 20mm ]
-%     \node[state] (2) {$q_{init}$};
-%     \node[state, initial, below left of=2] (1) {$init$};
-%     \node[state, below right of=1] (3) {$r_{init}$};
-%     \node[right of=2] (4) {$\cdots$};
-%     \node[right of=3] (5) {$\cdots$};
-%     % \node[state, below right of=1] (3) {$3$};
-%     % \node[state, below of=2] (4) {$4$};
-
-%     \path[->] (1) edge[below] node{$\varepsilon$} (2);
-%     \path[->] (1) edge[below] node{$\varepsilon$} (3);
-%     \path[->] (2) edge[below] node{} (4);
-%     \path[->] (3) edge[below] node{} (5);
-
-%     \node[label={[name=l] $N$}, draw,line width=2pt,rounded corners=5pt, fit=(2)(4)] {};
-%     \node[label={[name=l] $M$}, draw,line width=2pt,rounded corners=5pt, fit=(3)(5)] {};
-%   \end{tikzpicture}
-%   \caption{$N \oplus_{NFA} M$}
-%   \label{fig:disjunctionNFA}
-%   \end{figure}
-
-%   \Cref{fig:disjunctionNFA} shows the process of
-%   disjunctively combining NFAs. Precisely, we add a single
-%   new state and we included copies of the states from each of $N$
-%   and $M$. The new state acts as the initial state and has
-%   $\varepsilon$-transitions to the initial states of $N$ and
-%   $M$. We include all of the internal transitions from $N$
-%   and $M$, and the accepting states of $N \oplus_{NFA} M$
-%   are exactly the accepting states in each subautomaton.
-
-%   Let $g$ and $g'$ be two regular grammars such that
-%   $g \cong NFA(g)$ and $g' \cong NFA(g')$. As a matter of
-%   notation\footnote{We shall similarly abuse notation for
-%     $\otimes$ and Kleene. That is, for a regular grammar
-%     $g$, when we write $NFA(g)$ we mean the NFA inductively
-%     built up with the NFA-analogues to the constructors that
-%     built up $g$.}, we will write $NFA(g \oplus g')$ for
-%   $NFA(g) \oplus_{NFA} NFA(g')$. The traces of
-%   $NFA(g \oplus g')$ are then given by,
-
-%   \[
-%     \mathsf{Trace}_{NFA(g \oplus g')}(src , dst) = \mu
-%       \begin{pmatrix}
-%         g_{init} := g_{q_{0}} \oplus g_{r_{0}} \\
-%         g_{q_{j}} := \mathsf{Trans}_{NFA(g)}(q_{j}) \oplus \mathsf{isDst}(q_{j}) \\
-%         g_{r_{k}} := \mathsf{Trans}_{NFA(g')}(r_{k}) \oplus \mathsf{isDst}(r_{k})
-%       \end{pmatrix}.g_{src}
-%     \]
-
-%   where $\mathsf{Trans}$ is used to echo the same syntactic
-%   definitions that appear in the $NFA(g)$ and $NFA(g')$.
-%   Also, $src$ and $dst$ may take on any value in
-%   $Q := \{init\} \cup \{q_{j}\} \cup \{r_{k}\}$, and
-%   $\mathsf{isDst}(q)$ checks if $q$ is equal to $dst$. Which
-%   is all to say, the traces of $NFA(g \oplus g')$ comprise
-%   either a trace of $NFA(g)$, or a trace of $NFA(g')$, and
-%   the transition coming out of $g_{init}$ determines which
-%   subautomaton we step into.
-
-%   The parse transformer from $g \oplus g'$ checks which side
-%   of the sum type we are on, then takes the appropriate step
-%   from $g_{init}$ in the automaton.
-
-%   \[
-%     \inferrule
-%     {
-%       u : g \vdash \iota (\phi (u)) : \mathsf{AccTrace}_{NFA(g \oplus g')} \\
-%       v : g' \vdash \iota' (\psi (v)) : \mathsf{AccTrace}_{NFA(g \oplus g')}
-%     }
-%     {p : g \oplus g' \vdash \mathsf{case}~p \{ \mathsf{inl}(u) \mapsto s , \mathsf{inr}(v) \mapsto r \} : \mathsf{AccTrace}_{NFA(g \oplus g')}}
-%   \]
-
-%   with $\iota$ and $\iota'$ as embeddings from $NFA(g)$ and
-%   $NFA(g')$, respectively, into $NFA(g \oplus g')$,
-%   $\phi: g \cong \mathsf{AccTrace}_{NFA(g)}$, and $\psi: g' \cong \mathsf{AccTrace}_{NFA(g')}$. On a
-%   high level, all this construction does is turn a parse of
-%   $g$ into a parse of $NFA(g)$ and then embeds that inside
-%   of the larger automaton $NFA(g \oplus g')$. Likewise for $g'$.
-
-%   In the other direction, recall that the data of an
-%   accepting trace for $NFA(g \oplus g')$ is a pair of a
-%   trace and a proof that
-%   the end state $q'$ of that trace is accepting. By
-%   multifolding over the first part of that pair, we turn the
-%   term of type
-%   $\mathsf{Trace}_{NFA(g \oplus g')}(init , q')$ into a
-%   trace of either of the subautomata,
-
-%   \[
-%     p : \mathsf{Trace}_{NFA(g \oplus g')}(init , q') \vdash \mathsf{mfold}_{NFA(g \oplus g')} : \mathsf{Trace}_{NFA(g)}(q_{0} , q') \oplus \mathsf{Trace}_{NFA(g')}(r_{0} , q')
-%   \]
-
-%   Additionally, we leverage the fact that the only accepting
-%   states for $NFA(g \oplus g')$ are those from the
-%   subautomata to extract that $q'$ must be an accepting
-%   state from a subautomaton.
-
-%   \[
-%     x : \mathsf{acc}_{NFA(g \oplus g')}(q') \vdash M : \mathsf{acc}_{NFA(g)}(q') \oplus \mathsf{acc}_{NFA(g')}(q')
-%   \]
-
-%   We then combine the trace and proof of acceptance into an
-%   accepting trace of one of the subautomata,
-
-%   \[
-%     p : \mathsf{AccTrace}_{NFA(g \oplus g')} \vdash N : \mathsf{AccTrace}_{NFA(g)} \oplus \mathsf{AccTrace}_{NFA(g')}
-%   \]
-
-%   Lastly, we then inductively use the isomorphisms $\phi$
-%   and $\psi$ to turn the accepting traces into a parse of
-%   $g$ or $g'$,
-
-%   \[
-%      N : \mathsf{AccTrace}_{NFA(g)} \oplus \mathsf{AccTrace}_{NFA(g')} \vdash \mathsf{case}~N~\{\mathsf{inl}(n) \mapsto \phi^{-1}(n), \mathsf{inr}(n') \mapsto \psi^{-1}(n')\} : g \oplus g'
-%   \]
-
-%   When you wish to show that these actually maps invert each other to form an isomorphism\max{jumbled sentence}, we invoke the
-%   universal properties of the $\mathsf{Trace}$ types. That is,
-%   $\mathsf{Trace}_{\oplus NFA}$, $\mathsf{Trace}_{N}$, and $\mathsf{Trace}_{M}$
-%   are \emph{initial} algebras to the the functors that define them. Thus, the
-%   embedding of traces through $N$ into $\oplus NFA$ and then mapped back down to $N$
-%   compose to a map from $\mathsf{Trace}_{N}$ to itself. Such a map is unique and
-%   thus must be the identity. The same reasoning shows that the desired
-%   composite do indeed invert each other to form an isomorphism.
-
-%   As discussed above, the other cases and
-%   follow through a similar argument. This
-%   concludes the proof of our variant of Thompson's construction.
-% \end{proof}
-
-% \subsection{Other Automata}
-% \subsubsection{Pushdown Automata}
-% \label{subsubsec:pda}
-% A (nondeterministic) \emph{pushdown automaton} is an automaton that employs a
-% stack. Just like NFAs, they have transitions labeled with characters from a
-% fixed string alphabet $\Sigma$. Additionally, they maintain a stack of
-% characters drawn from a stack alphabet $\Gamma$. They are often represented
-% formally as a 7-tuple $(Q, \Sigma, \Gamma, \delta, q_{0}, Z, F)$,
-
-% \begin{itemize}
-%   \item $Q$ a finite set of states
-%   \item $\Sigma$ a fixed, finite string alphabet
-%   \item $S$ a fixed, finite stack alphabet
-%   \item
-%         $\delta \subset Q \times (\Sigma \cup \{ \varepsilon \}) \times S \to \mathcal{P}(Q \times S^{*})$
-%         the transition function
-%   \item $q_{0} \in Q$ the start state
-%   \item $Z \in S$ the initial stack symbol
-%   \item $F \subset Q$ the accepting states
-% \end{itemize}
-
-% We encode the traces of a pushdown automaton very similarly to those of an NFA,
-% except the transitions of a PDA are instead encoded via the LNL product type
-% $\bigamp$. This is because when simply transitioning via character, a PDA must
-% also pop and push characters onto a stack, which is used as the argument to
-% these dependent functions. Thus, the appropriate formation of traces through a
-% PDA is dependent on a stack.
-
-% Let $S$ be a non-linear type encoding the stack
-% alphabet, and build lists over $S$ as the (non-linear) least fixed-point
-% $\mathsf{List}(S) := \mu X . 1 + S \times X$. Then, the type of states for a
-% PDA $P$ with stack alphabet $S$ are given as a functions that takes in lists $L$,
-% and then makes a case distinction between possible transitions based on what was witnessed as $\mathsf{head}(L)$. The choice of transition
-% will then determine which character to transition by and what word $w$ should be
-% pushed onto the stack. The word that is added to the top of the stack is
-% appended to $\mathsf{tail}(L)$ and then we recursively step into another state
-% called on argument $w \cdot \mathsf{tail}(L)$.
-
-% \begin{gather*}
-%   \mathsf{State} \in {g _{q} : q \in Q} \\
-%   \mathsf{Word} \in \String \\
-%   \mathsf{Char} \in \Sigma   \\
-%   \mathsf{StackChar} \in \Gamma \\
-% \end{gather*}
-% \begin{align*}
-%   \mathsf{Trans}_{P}(q) ::=~&
-%                            \LinPiTy {(hd :: tl)} {\mathsf{List}(S)} {\left( \mathsf{Char} \otimes \mathsf{State}(\mathsf{Word} + tl) \right)}~| \\
-%   & \LinPiTy {(hd :: tl)} {\mathsf{List}(S)} {\left(  \mathsf{State}(\mathsf{Word} + tl) \right)}~|~\mathsf{Trans}_{P}(q) \oplus \mathsf{Trans}_{P}(q)
-% \end{align*}
-
-% That is to say, when transitioning a PDA pops off the head $hd$ of the stack
-
-% \subsubsection{Turing Machines}
-% \label{subsubsec:tm}
-
-% Above we gave a grammar presentation of
-% traces through a PDA by using a non-linear type $S$
-% to encode the stack. We may similarly use pairs $S \times S$
-% to encode the tape of a Turing machine. With two stacks we can simulate the behavior of the
-% infinite tape of a Turing machine. The intuition behind this correspondence is
-% that the left half of the tape is on one stack, the right
-% half of the tape the other, and we treat the tops of stacks
-% like the head of the tape.
-
-% \pedro{I like the overall structure of this section! Showing how the type theory
-% can encode and reason about different kinds of languages is very instructive!}
+Given a linear type $A$ and a nonlinear type $X$, we define a semantic action to
+be a function $\ltonl {(A \lto \Delta X)}$, where
+$\Delta X = \LinSigTy {x} {X} {\top}$. That is, a semantic action is an
+elaboration from the concrete parses of $A$ into the semantic domain $X$. As
+written, these actions are quite general. As discussed above, we may define
+folds that are still grammatical in nature --- such as folding over a concrete
+parse tree and forgetting parentheses that aren't needed in later compiler
+passes. We may also define arbitrary computation over the structure of the parse
+tree, such as calculating the maximum nesting depth of parentheses for $Dyck$,
+returning the length the input string, or
+even an interpreter that evaluates arithmetic expressions at parse-time.
+
+\steven{What can I say about what we have in the Agda development here?}
+\pedro{I'm assuming that this section is still incomplete, so I'll wait
+for it to be complete in order to make more detailed comments.}
 
@@
 that accounts for the possibility of ambiguity. The most common
-definition of a formal grammar is as a generative grammars, defined by
+definition of a formal grammar is as generative grammars, defined by
 a set of non-terminals, with a specified start symbol and set of
@@
 grammars, but what about the linear terms? These can be interpreted as
-\emph{parse transformers}
+\emph{parse transformers}, which we now define.
 \begin{definition}
@@
 \end{definition}
+\pedro{Do we actually need a definition of formal language morphism?
+In case we do, it should probably be its own definition.}
 First, note that parse transformers generalize language inclusion: if
@@
 
-Parse transformers have an obvious notion of composition and identity
-transformer. This defines a \emph{category}.
+Parse transformers can be composed: given two parse transformers $f$ and $g$,
+their composition is defined pointwise, i.e. $(f\circ g)_w = f_w \circ g_w$.
+Furthemore, given a formal grammar $A$, its identity transformer is $id_w =
+id_{A(w)}$, where $id_{A(w)}$ is the identity function on the set $A(w)$. This
+defines a \emph{category}.
 \begin{definition}
@@
 This category is equivalent to the slice category $\Set/\Sigma^*$ and
-as such is very well-behaved. It is complete, co-complete, cartesian
+as such is very well-behaved. It is complete, co-complete, Cartesian
 closed and carries a monoidal biclosed structure. We will use these
@@
 
-For reasons of brevity, we present only a selection of cases in
-\Cref{fig:semantics}. The full details are in the appendix.\max{todo:
-  put the full details in the appendix}.
-\max{TODO: how to describe $\mu$ succinctly?}\max{make sure $\lto$ and $\tol$ are right}
+For reasons of brevity, we present only the cases related to the
+linear types in \Cref{fig:semantics}. The full details are in the
+appendix.\max{TODO: put the full details in the appendix}.  The
+grammar for a literal $c$ has a single parse precisely when the input
+string consists of the single character. The grammar for the unit
+similarly has a single parse for the empty string.  A parse of the
+tensor product $A \otimes B$ consists of a \emph{splitting} of the
+empty string into a prefix $w_1$ and suffix $w_2$ along with an $A$
+parse of $w_1$ and $B$ parse of $w_2$. A parse of $\bigoplus_{x:X} A$
+is a pair of an element of the set $X$ and a parse of $A(x)$, while
+dually a parse of $\bigwith_{x:X} A$ is a \emph{function} taking any
+$x:X$ to a parse of $A(x)$. A $w$-parse of $A \lto B$ is a function
+that takes an $A$ parse of some other string $w'$ to a $B$ parse of
+$ww'$, and $B \tol A$ is the same except the $B$ parse is for the
+reversed concatenation $w'w$.
+%
+The set $\uparrow A$ is the set of parse for the empty string for
+$A$. This definition means that $\sem{\uparrow (A \lto B)}$ (or $\sem{\uparrow (B \tol A)}$) is
+equivalent to the set of parse transformers:
+\[ \sem{\uparrow (A \lto B)}\gamma = \sem{A \lto B}\gamma\,\varepsilon = \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,w'\]
+Next, a parse in the equalizer $\equalizer{a : A}{f}{g}$ is defined as
+a parse in $\sem{A}$ that is mapped to the same parse by $f$ and
+$g$. Recall that $\sem{f}\gamma$ and $\sem{g}$ are elements of
+$\sem{\uparrow(A \lto B)}$, i.e., parse transformers.
+
+The universe $L$ of linear types is interpreted as the set of all
+(small) grammars.
+
 \begin{figure}
@@
     \sem{A \otimes B}\gamma\,w &= \{ (w_1,w_2,a,b) \pipe w_1w_2 = w \wedge a \in \sem{A}\gamma\,w_1 \wedge b \in \sem{B}\gamma\,w_2 \}\\
+    \sem{A \lto B}\gamma\,w &= \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,ww'\\
+    \sem{B \tol A}\gamma\,w &= \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,w'w\\
     \sem{\bigoplus_{x:X} A}\gamma\,w &= \{ (x, a) \pipe x \in \sem{X}\gamma \wedge a \in \sem{A}(\gamma,x)w \}\\
     \sem{\bigwith_{x:X} A}\gamma\,w &= \prod_{x\in \sem{X}\gamma} \sem{A}(\gamma, x)\,w\\
-    \sem{A \lto B}\gamma\,w &= \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,ww'\\
-    \sem{B \tol A}\gamma\,w &= \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,w'w\\
     \sem{\uparrow A}\gamma &= \sem{A}\gamma\,\varepsilon\\
-    \sem{\SPF\,X}\gamma &= \{ F : \textrm{Functor}(\textrm{Grammar}^{\sem{X}\gamma}, \textrm{Grammar}) \pipe F \textrm{ preserves $\omega$-colimits.} \}\\
-    \sem{\el(F)}\gamma\,G &= \sem{A}\gamma\,G\\
-    \sem{\mu A}\gamma &= \mu (\sem{A}\gamma) \tag{\Cref{thm:adamek}}
+    \sem{\equalizer{a:A}{f}{g}}\gamma\,w &= \{ a \in \sem{A}\gamma\,w \pipe \sem{f}\gamma\,w\,a = \sem{g}\gamma\,w\,a \} \\
+    \sem{L}\gamma &= \Grammar_0\\
+    \sem{\SPF\,X}\gamma &= \textrm{DepPolyFunctor}(\sem{X}\gamma\times \Sigma^*,\Sigma^*)\\
+    \sem{\el(F)}\gamma\,G &= \sem{F}\gamma\,G\\
+    \sem{\map(F)}\gamma\,f &= \sem{F}\gamma\,f\\
+    \sem{\mu A}\gamma &= \mu (\sem{A}\gamma) \tag{\Cref{thm:indexed-container}}\\
   \end{align*}
@@
 
-\max{TODO: prose}
-
-\max{todo: spelling}
-\begin{theorem}[Adamek]
-  If a functor $F : \mathcal C \to \mathcal C$ preserves
-  $\omega$-colimits and $\mathcal C$ has an initial object and
-  $\omega$-colimits, then $F$ has an initial algebra given by the
-  colimit of the chain
-  \[ 0 \to F(0) \to F(F(0)) \to \cdots \]
+The most complex part of the semantics is the interpretation of
+strictly positive functors and indexed inductive linear types.  We
+interpret a strictly positive functor as a \emph{dependent polynomial
+functor} on the category of sets, also sometimes called an
+\emph{indexed container} \cite{dep-poly,indexed-container}.
+\begin{definition}
+  Let $I$ and $O$ be sets. A (dependent) \emph{polynomial} (of sets)
+  from $I$ to $O$ consists of a set of shapes $S$, a set of positions
+  $P$ and functions $I \xleftarrow f P \xrightarrow g S  \xrightarrow h O$. The \emph{extension}
+  of a polynomial is a functor $\Set/I \to \Set/O$ defined as the
+  composite
+  % https://q.uiver.app/#q=WzAsNCxbMCwwLCJcXFNldC9JIl0sWzMsMCwiXFxTZXQvTyJdLFsxLDAsIlxcU2V0L1AiXSxbMiwwLCJcXFNldC9TIl0sWzAsMiwiZl4qIl0sWzIsMywiXFxQaV9nIl0sWzMsMSwiXFxTaWdtYV9oIl1d
+  \[\begin{tikzcd}
+	    {\Set/I} & {\Set/P} & {\Set/S} & {\Set/O}
+	    \arrow["{f^*}", from=1-1, to=1-2]
+	    \arrow["{\Pi_g}", from=1-2, to=1-3]
+	    \arrow["{\Sigma_h}", from=1-3, to=1-4]
+  \end{tikzcd}\]
+  Where $f^*$ is the pullback functor along $f$; $\Pi_g$ is the depedent
+  product operation and $\Sigma_h$ is the dependent sum operation, which
+  are, respectively, the right and left adjoint of their pullback functors
+  $g^*$ and $h^*$.
+
+  A dependent polynomial functor from $I$ to $O$ is a functor $\Set/I
+  \to \Set/O$ that is nautrally isomorphic to the extension of a
+  dependent polynomial from $I$ to $O$.
+\end{definition}
+With this interpretation, $\el(F)$ and $\map(F)$ are just interpreted
+as the action of the functor on objects and morphisms,
+respectively. We interpret the constructors $\mathsf{K},\Var$, etc. on
+functors in the obvious way that matches the definitional behavior of
+$\el$ and $\map$. The non-trivial part of the construction is
+verifying that such functors are constructing corresponding dependent
+polynomials. The details are tedious but straightforward extension of
+prior work on dependent polynomials and indexed containers and we have
+verified the construction in Agda.
+
+The reason we use dependent polynomials functors on sets is that it is
+well-known that their initial algebras exist
+\cite{initial-algebras-of-dep-poly}. Additionally, initial algebras
+for dependent polynomials are straightforwardly constructed in Agda as
+an inductive type of ``IW'' trees, which is available as a
+construction in the cubical library of Agda \cite{iw,cubical-library}.
+\begin{theorem}
+  Every dependent polynomial functor from $I$ to $I$ has an initial
+  algebra.
 \end{theorem}
+Then an element $F \in \sem{X \to \SPF\,X}\gamma$ is an
+$\sem{X}\gamma$-indexed family of polynomial functors from
+$\sem{X}\gamma\times \Sigma^*$ to $\Sigma^*$, and taking the product
+of these constructs a polynomial functor from $\sem{X}\gamma\times
+\Sigma^*$ to itself. Then $\sem{\mu\,F}\gamma$ is defined to be the
+initial algebra of this functor, and the initial algebra structure is
+used to interpret $\roll, \fold$ and the corresponding axioms.
 
@@
 
-\steven{Should discussion of this connection to lambek calculus be spread
+\steven{Should discussion of this connection to Lambek calculus be spread
   throughout the paper?}
@@
 languages. That is, they seem to try to give a functorial semantic
-interpretation of natrual language.
+interpretation of natural language.
 
@@
 Also mention here that \cite{luoLambekCalculusDependent} seems similar at first
-glance, but isn't as similar as it might look because the judgements and
+glance, but isn't as similar as it might look because the judgments and
 dependent types seem to
@@
 \steven{TODO Edit the above}
-\steven{Even though the above sounds bad, are there lambek/categorial things that are missed?}
+\steven{Even though the above sounds bad, are there Lambek/categorial things that are missed?}
 
@@
 
+\paragraph{Inductive Linear Types}
+
 % In \cite{firsovCertifiedNormalizationContextFree2015}, Firsov and Uustalu
@@
 
-\steven{TODO add lambek calculus/categorial grammars here}
+\steven{TODO add Lambek calculus/categorial grammars here}
 
@@
 operations generalizing language composition, language union and the
-Kleene star.  More generally, they are defined as inequational theory
+Kleene star.  More generally, they are defined as an inequational theory
 where the inequality is meant to capture language containment. This
-theory is extremely successful, having found applications in algebraic
-path problems, theory of programming languages, compiler optimizations
-and more.
+theory is extremely successful, having found applications in concurrency theory
+\cite{hoare2000concurrent}, software-defined networks \cite{anderson2014netkat},
+theory of programming languages \cite{angus2001kleene}, compiler optimizations
+\cite{kozen2000certification} and more.
 
@@
 extensions of Kleene algebras, Kleene algebra with tests (KAT) being
-one of the most notable ones. Our approach is radically different from
+one of the most notable ones \cite{kozen1997kleene}. Our approach is radically different from
 most extensions, which usually aim at modifying or adding new
@@
 \steven{What can we say about internalizing Kleene algebra proofs or extensions?}
-\steven{any KA citations needed above?}
+\pedro{As a rule of thumb, I don't think that sketching these ideas
+will get us anything...}
 
@@
 
-\section{Elided Syntax}
-
-\max{TODO: add prose}
+\section{Syntax}
+
+In this section we include the elided syntactic forms, as well as
+definitions and basic properties of linear and non-linear
+substitution.
 
@@
 \end{figure}
-\steven{Do we need inverse for $\ltonl A$ in judgemental equality figure?}
+\steven{Do we need inverse for $\ltonl A$ in judgmental equality figure?}
 \steven{Do we need to say anything about the equalizer proofs being unique in
-  judgemental equality figure?}
+  judgmental equality figure?}
+
+\begin{definition}
+  The set of (non-linear) substitutions $\gamma \in
+  \textrm{Subst}(\Gamma,\Gamma')$ where $\Gamma \isCtx$ and $\Gamma'
+  \isCtx$ is defined by recursion on $\Gamma$:
+  \begin{align*}
+    \textrm{Subst}(\Gamma,\cdot) &= \{\cdot \}\\
+    \textrm{Subst}(\Gamma,\Gamma',x:A) &= \{ (\gamma,M/x) \pipe \gamma \in \textrm{Subst}(\Gamma,\Gamma') \wedge \Gamma \vdash M : A[\gamma] \}
+  \end{align*}
+  simultaneously with an action of substitution on types, terms,
+  etc. in the standard way.
+
+  \max{do we need identity and composition of substitutions and their functoriality?}
+  %% By induction we can define an identity substitution
+  %% $\textrm{Subst}(\Gamma,\Gamma)$ which maps each variable to
+  %% itself. We write this substitution simply as $\Gamma$.
+  %% \begin{align*}
+  %%   \cdot \circ \gamma' &= \cdot\\
+  %%   (\gamma,M/x) \circ \gamma' &= (\gamma\circ \gamma') , M[\gamma']/x
+  %% \end{align*}
+
+  It is straightforward, but laborious to establish that all forms in
+  the type theory that are parameterized by a non-linear context
+  $\Gamma$ support an admissible action of a substitution $\gamma \in
+  \Subst(\Gamma',\Gamma)$:
+  \begin{mathpar}
+    \inferrule
+    {\Gamma \vdash X \isTy}
+    {\Gamma' \vdash X[\gamma] \isTy}
+
+    \inferrule
+    {\Gamma \vdash X \isSmall}
+    {\Gamma' \vdash X[\gamma] \isSmall}
+
+    \inferrule
+    {\Gamma \vdash X \equiv Y}
+    {\Gamma' \vdash X[\gamma] \equiv Y[\gamma]}
+
+    \inferrule
+    {\Gamma \vdash M : X}
+    {\Gamma' \vdash M[\gamma] : X[\gamma]}
+
+    \inferrule
+    {\Gamma \vdash M \equiv N : X}
+    {\Gamma' \vdash M[\gamma] \equiv N[\gamma] : X[\gamma]}
+
+    \inferrule
+    {\Gamma \vdash \Delta \isLinCtx}
+    {\Gamma' \vdash \Delta[\gamma] \isLinCtx}
+
+    \inferrule
+    {\Gamma \vdash A \isLinTy}
+    {\Gamma' \vdash A[\gamma] \isLinTy}
+
+    \inferrule
+    {\Gamma \vdash A \equiv B}
+    {\Gamma' \vdash A[\gamma] \equiv B[\gamma]}
+
+    \inferrule
+    {\Gamma;\Delta \vdash e : A}
+    {\Gamma'; \Delta[\gamma] \vdash e[\gamma] : A[\gamma]}j
+
+    \inferrule
+    {\Gamma;\Delta \vdash e \equiv f : A}
+    {\Gamma'; \Delta[\gamma] \vdash e[\gamma] \equiv f[\gamma] : A[\gamma]}j
+  \end{mathpar}
+
+  %% Further, these all satisfy identity and composition equations M[\Gamma] = M and M[\gamma][\gamma'] = M[(\gamma \circ \gamma')]
+\end{definition}
+
+\begin{definition}
+  Let $\Gamma \vdash \Delta \isLinCtx$ and $\Gamma \vdash \Delta'
+  \isLinCtx$. The set of linear substitutions $\Subst(\Delta',\Delta)$
+  is defined by recursion on $\Delta$:
+  \begin{align*}
+    \Subst(\Delta',\cdot) &= \{ \cdot \pipe \Delta' = \cdot \}\\
+    \Subst(\Delta',(\Delta,a:A)) &= \{ (\delta, e/a) \pipe \delta \in \Subst(\Delta_1, \Delta) \and \Delta_2 \vdash e : A \and \Delta' = \Delta_1,\Delta_2\}
+  \end{align*}
+
+  Given substitutions $\delta_1 \in \Subst(\Delta_1', \Delta_1)$ and
+  $\delta_2 \in \Subst(\Delta_2', \Delta_2)$, we can define a
+  substitution $\delta_1,\delta_2 \in
+  \Subst((\Delta_1',\Delta_2'),(\Delta_1,\Delta_2))$. Furthermore, for
+  any substitution $\delta \in \Subst(\Delta,(\Delta_1,\Delta_2))$, we
+  can deconstruct $\delta = \delta_1,\delta_2$ with $\delta_1 \in
+  \Subst(\Delta_1', \Delta_1)$ and $\delta_2 \in \Subst(\Delta_2',
+  \Delta_2)$.
+  %% Again for any $\Gamma \vdash \Delta$, we can define an identity
+  %% substitution mapping each variable to itself. Additionally we can define composition of 
+\end{definition}
+
+\begin{definition}
+  Given any $\Gamma ; \Delta \vdash e : A$ and $\delta \in
+  \Subst(\Delta',\Delta)$, we define the action of the substitution on
+  $e$ as follows, frequently using the inversion principle to split
+  the substitution into constituent components.
+  \begin{align*}
+    a[e/a] &= e\\
+    (e_1,e_2)[\delta_1,\delta_2] &= (e_1[\delta_1], e_2[\delta_2])\\
+    (\letin {(a , b)} e {e'})[\delta_1,\delta_2,\delta_3] &= \letin{(a,b)} {e[\delta_2]}{e'[\delta_1,a/a,b/b,\delta_2]}\\
+    ()[\cdot] &= ()\\
+    \letin{()} e {e'}[\delta_1,\delta_2,\delta_3] &= \letin{()} {e[\delta_2]}{e'[\delta_1,a/a,b/b,\delta_2]}\\
+    (\lamblto {a} e)[\delta] &= \lamblto a {e[\delta,a/a]}\\
+    (\applto {e'} {e})[\delta_1,\delta_2] &= \applto {e'[\delta_1]} {e[\delta_2]}\\
+    (\lamblto {a} e)[\delta] &= \lamblto a {e[\delta,a/a]}\\
+    (\applto {e'} {e})[\delta_1,\delta_2] &= \applto {e'[\delta_1]} {e[\delta_2]}\\
+    (\lambtol {a} e)[\delta] &= \lamblto a {e[a/a,\delta]}\\
+    (\apptol {e'} {e})[\delta_1,\delta_2] &= \apptol {e'[\delta_1]} {e[\delta_2]}\\
+    (\dlamb x e)[\delta] &= \dlamb x {e[\delta]}\\
+    (e\,.\pi\,M)[\delta] &= (e[\delta]\,.\pi\,M)\\
+    (\sigma\,M\,e)[\delta] &= \sigma\,M\,e[\delta]\\
+    (\letin{\sigma\,x\,a}{e}{e'})[\delta_1,\delta_2,\delta_3] &= \letin{\sigma\,x\,a} {e[\delta_2]}{e'[\delta_1,\delta_3]}\\
+    (\equalizerin{e})[\delta] &= \equalizerin{e[\delta]}\\
+    (\equalizerpi{e})[\delta] &= \equalizerpi{e[\delta]}
+  \end{align*}
+  By induction on linear term and equality judgments, we establish the
+  following admissible rules for $\delta \in \Subst(\Delta',\Delta)$:
+  \begin{mathpar}
+    \inferrule
+    {\Gamma; \Delta \vdash e : A}
+    {\Gamma; \Delta' \vdash e[\delta] : A}
+
+    \inferrule
+    {\Gamma; \Delta \vdash e \equiv f : A}
+    {\Gamma; \Delta' \vdash e[\delta] \equiv f[\delta'] : A}
+  \end{mathpar}
+
+  \max{identity and composition?}
+\end{definition}
+
+\section{Denotational Semantics}
+
+\begin{definition}
+  We define the unit grammar
+  \[ I\,w = \{ ()\pipe w = \varepsilon \}\]
+  and given grammars $A$ and $B$ we define their tensor product
+  \[ (A \otimes B)\,w = \{ (w_1,w_2,a,b) \pipe w_1w_2 = w \wedge a \in A\,w_1 \wedge b \in B\,w_2 \} \]
+  This extends to a monoidal structure on the category of grammars.
+Note that we defined in the semantics of linear types $\sem{I}\gamma = I$ and $\sem{A \otimes B}\gamma = \sem{A}\gamma \otimes \sem{B} \gamma$
+\end{definition}
+
+\begin{definition}[Denotation of Linear Contexts]
+  The semantics of linear contexts $\Gamma \vdash \Delta \isLinCtx$ is
+  defined as follows:
+  \max{make sure this aligns with our interpretation of $\lto$/$\tol$}
+  \begin{align*}
+    \sem{\cdot}\,\gamma &= I \\
+    \sem{\Delta,x:A}\,\gamma &= \sem{\Delta}\,\gamma \otimes \sem{A}\gamma \\
+  \end{align*}
+\end{definition}
+
+We define denotations of linear terms in
+\Cref{fig:linear-term-semantics}. Note that the denotations interpret
+typing derivations, not raw terms, as the data of how contexts are
+split is needed in order to construct the correct associator
+functions.
+\max{todo: describe associators, define denotations, prove axioms}
+\begin{figure}
+  \begin{mathpar}
+  \inferrule
+  {\Delta_1 \vdash e_1 : A_1 \and \Delta_2 \vdash e_2 : A_2}
+  {\sem{(e_1,e_2)}\gamma\,w\,\delta
+    = \letin{(w_1,w_2,\delta_1,\delta_2)} {m(\Delta_1,\Delta_2)(\delta)} {(w_1,w_2,\sem{e_1}\gamma\,w_1\,\delta_1, \sem{e_2}\gamma\,w_2\,\delta_2)}
+  }
+
+  \inferrule
+  {\Delta_2 \vdash e : A \otimes B \and \Delta_1,a:A,b:B,\Delta_3 \vdash e' : C}
+  {\sem{\letin{(a,b)}{e}{e'}}\gamma\,w\,\delta =
+    \letin{(w_1,w_2w_3,\delta_1,(w_2,w_3,\delta_2,\delta_3))}{m_{\Delta_1,\Delta_2,\Delta_3}(\delta)}
+    \letin{(w_{2,1},w_{2,2},a,b)}{\sem{e}\gamma\,w_2\,\delta_2}
+    {\sem{e'}\gamma\,w\,(w_1,w_{2,1}w_{2,2}w_3, \delta_1, (w_{2,1},w_{2,2}w_3, a, (w_{2,2},w_3, b, \delta_3)))}
+    }
+\end{mathpar}
+  \caption{Denotations of linear terms}
+\end{figure}
+
+\begin{definition}
+  For any $\Gamma \vdash \Delta_1,\Delta_2 \isLinCtx$ and $\gamma \in
+  \sem{\Gamma}$ there is an isomorphism $m_{\Delta_1,\Delta_2} :
+  \sem{\Delta_1, \Delta_2}\gamma \cong \sem{\Delta_1}\gamma \otimes
+  \sem{\Delta_2}\gamma$.
+
+  This can be extended to a sequence of contexts of any length.
+\end{definition}
+
+\begin{lemma}[Equations for $\otimes$]
+
+\end{lemma}
+
+\begin{lemma}[Equations for $I$]
+
+\end{lemma}
+
+\begin{lemma}[Equations for $\lto$]
+
+\end{lemma}
+
+\begin{lemma}[Equations for $\tol$]
+
+\end{lemma}
+
+\begin{align*}
+  \sem{(e_1,e_2)}\gamma\,w\,\delta &=
+  \letin{(w_1,w_2,\delta_1,\delta_2)} {m(\Delta_1,\Delta_2)(\delta)} {(w_1,w_2,\sem{e_1}\gamma\,w_1\,\delta_1, \sem{e_2}\gamma\,w_2\,\delta_2)}\\
+\end{align*}
 

### 2024-11-22 paper/paper.tex
MWPLS draft presentation

@@
 % -*- fill-column: 80; -*-
-\documentclass[review,anonymous,screen,acmsmall,nonacm]{acmart}
+\documentclass[screen,acmsmall,nonacm]{acmart}
 \usepackage{mathpartir}
@@
 
+ %% \steven{TODO make the Agda logo clickable} %
+\newcommand{\Agda}{\agdalogo}
 \renewcommand{\Gamma}{\mathgtt{G}}
@@
 \newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
+\newcommand{\semg}[1]{\sem{#1}\gamma}
 \newcommand{\cat}[1]{\mathbf{#1}}
@@
 \newcommand{\String}{\textbf{String}}
+\newcommand{\Char}{\textbf{Char}}
 \newcommand{\stringg}{\texttt{String}}
@@
 \newcommand{\Grammar}{\mathbf{Gr}}
+\newcommand{\semcat}{\mathbf{C}}
 \newcommand{\Type}{\mathbf{Type}}
@@
 \newif\ifdraft
-\drafttrue
+\draftfalse
 \newcommand{\steven}[1]{\ifdraft{\color{orange}[{\bf Steven says}: #1]}\fi}
@@
 
+\newif\ifappendix
+\appendixtrue
+
 \begin{document}
 
-\pagestyle{plain}
+%\pagestyle{plain}
 
@@
 }
-\email{nvarner@umich.edu}
+\email{nmvarner@umich.edu}
 
@@
 \begin{abstract}
-  We propose \theoryname~(\theoryabbv), a domain-specific dependent type theory
-  for the definition of intrinsically verified parsers and verification of weak
-  and strong equivalences of formal grammars. The linear types in \theoryabbv
-  have a simple interpretation as formal grammars, with values of the types
-  interpreted as parse trees and general terms as a form of intrinsically
-  verified parse transformers, of which parsers being a special case. We
-  demonstrate the expressivity of this system by showing that the indexed
-  inductive linear types in this calculus can be used to encode many different
-  familiar grammar formalisms, including regular and context-free grammars but
-  also traces of various types of automata. We can then construct strong
-  equivalences between grammar and automata formalisms and construct verified
-  parsers as simple linear functional programs. All of our examples have been
-  implemented in a prototype implementation of \theoryabbv as a shallow
-  embedding in Agda.
+  We present \theoryname~(\theoryabbv), a domain-specific dependent
+  type theory for verified parsing and formal grammar theory. In
+  \theoryabbv, linear types are used as a syntax for formal grammars,
+  and parsers can be written as linear terms. The linear typing
+  restriction provides a form of intrinsic verification that a parser
+  yields only valid parse trees for the input string. We demonstrate
+  the expressivity of this system by showing that the combination of
+  inductive linear types and dependency on non-linear data can be used
+  to encode commonly used grammar formalisms such as regular and
+  context-free grammars as well as traces of various types of
+  automata. Using these encodings, we define parsers for regular
+  expressions using deterministic automata, as well as
+  examples of verified parsers of context-free grammars.
+
+  We present a denotational semantics of our type theory that
+  interprets the types as a mathematical notion of formal
+  grammars. Based on this denotational semantics, we have made a
+  prototype implementation of \theoryabbv using a shallow embedding in
+  the Agda proof assistant. All of our examples parsers have been
+  implemented in this prototype implementation.
 \end{abstract}
@@
 \label{sec:intro}
-Formal verification encompasses a broad suite of techniques one may use to
-mathematically guarantee the correctness of software systems. We can only
-reap the benefits of
-any proof effort if its underlying assumptions hold. For instance, verifying the
-functional correctness of an algorithm is no use if it is run on buggy hardware.
-Likewise, the promised guarantees of a program may be jeopardized if it is
-parsed incorrectly.
-
-Large verified developments, like CompCert \cite{leroy_formal_2009},
-invest a significant number of person-hours into proof development. In early
-versions of CompCert, the apparent
-assurances associated with their effort were at risk of invalidation by
-parser bugs \cite{yangFindingUnderstandingBugs}. Parsing is a ubiquitous task
-throughout computing and yet parsers are frequently unverified, even in
-otherwise verified systems. The CompCert team later
-developed a method for validating the correctness of LR parsers
-\cite{jourdanValidatingLRParsers2012} and then replaced their parser with one
-that is certified to be correct. Despite their success in  validating their C parser and other work in the
-generation of  verified parsers
-\cite{lasserCoStarVerifiedALL2021,firsovCertifiedNormalizationContextFree2015,danielssonTotalParserCombinators2010},
-it is still arduous and highly nontrivial to write a verified
-parser. Further, approaches are sensitive to a choice of grammar and automata formalism to serve
-as a
-foundation, which can make it difficult to relate and interoperate different
-styles of verified parsers.
-
-In \cite{elliottSymbolicAutomaticDifferentiation2021}, Elliott describes formal
-grammar as type-level predicates on strings that prove
-language membership. That is, a \emph{formal grammar} $A$ is a function
-$\String \to \Set$ such that for a string $w$, $A~w$ is the set of proofs
-showing that $w$ belongs to the language recognized by $A$. This typed approach
-to parsing serves a general setting that can simultaneuously interpret many distinct
-grammar formalisms. In this paper, we adopt Elliott's view on formal grammars and present a
-syntax that axiomatizes this semantic notion of grammar.
-
-We present \theoryname (\theoryabbv) as a domain specific
-type theory for building verified parsers. The type theory itself is a form of dependent linear-non-linear type
-theory: the linear types can be interpreted as grammars, whereas the
-non-linear types can be interpreted as ordinary sets. As in prior work
-on dependent linear-non-linear types, the linear types are allowed to
-depend on the non-linear types but not vice-versa \cite{krishnaswami_integrating_2015}.
+Parsing structured data from untrusted input is a ubiquitous task in
+computing. Any formally verified software system that interacts with
+the outside world must contain some parsing component. For example, in
+an extensive experiment finding bugs in C compilers
+\cite{yangFindingUnderstandingBugs}, an early version of the formally
+verified CompCert C compiler only contained bugs in the then
+unverified parsing component \cite{leroy_formal_2009}. Bugs in parsers
+undermine the overall correctness theorem for a verified system: an
+incorrectly parsed C program will be compiled correctly but this is
+not very useful if it did not correctly correspond to the actual
+source program. Eventually, a correct parser was implemented using an
+automaton that is formally verified to implement an LR grammar
+\cite{jourdanValidatingLRParsers2012}.
+
+It is entirely understandable from an engineering perspective
+\emph{why} verified parsing was not part of the initial releases of
+CompCert: parsing algorithms and formal grammars are a complex area,
+featuring a variety of domain-specific formalisms such as context-free
+grammars and various automata. These formalisms have little relation
+to the main components of a verified compiler. For this reason, it is
+advantageous for verified parsers to be implemented using a reusable
+verified library, just as parser generators and regular expression
+matchers have done for many decades in unverified software.
+
+Prior approaches to verified parsing focus on verification of a
+particular grammar formalism such as non-left-recursive grammars or
+LL(1) grammars
+\cite{lasserCoStarVerifiedALL2021,EdelmannZippy2020,danielssonTotalParserCombinators2010}.
+Each new grammar formalism is extended with its own independent
+verified implementation.
+
+In this work, we present the design of \theoryname (\theoryabbv), a
+domain-specific language for formal verification of parsers. A key
+property is that \theoryabbv is an \emph{extensible} framework for
+verification of parsers in that it supports the definition of grammar
+formalisms of unrestricted complexity. That is, \theoryabbv is not a
+system for verifying \emph{one} type of grammar formalism, but instead
+is a domain-specific language in which many grammar formalisms and
+their verified parsers can be implemented. For example, \theoryabbv is
+not a verified parser generator that compiles regular expressions to
+deterministic finite automata, but is instead is a domain specific
+language \emph{for writing} such a verified parser generator.
+
+The design of \theoryabbv
+is an extension of Joachim Lambek's \emph{syntactic calculus}
+\cite{lambek58}. Lambek calculus is a grammar formalism equivalent
+in expressive power to context-free grammars that in modern
+terminology would be considered a kind of \emph{non-commutative linear
+logic} --- a version of linear logic where the tensor product is not
+commutative, reflecting the obvious property that the relative
+ordering of characters is significant in parsing problems. We extend
+non-commutative linear logic with two key components that increase its
+power to support arbitrarily powerful grammar formalisms: inductive
+linear types, as well as dependency of linear types on non-linear
+data. The resulting system has two kinds of types: non-linear types
+which model sets and linear types which model formal
+grammars. Crucially, the non-linear types and linear types are allowed
+to be dependent on non-linear types, but not on linear types. This
+combination has been used previously in the ``linear-non-linear
+dependent'' type theory with \emph{commutative} linear logic to model
+imperative programming \cite{krishnaswami_integrating_2015}.
 
@@
 that derivations of a term type checking carry intrinsic proofs of correctness.
-
 Parsers written in \theoryabbv take on a linear functional style, which makes
-them familiar to write and amenable to compositional verification techniques. We
-implement \theoryabbv as a shallowly embedded domain-specific langauge in the
-Cubical Agda proof assistant. Throughout this paper, we will use \agdalogo~ to
-mark results that are mechanized in our Agda development.
-
-\paragraph{Contributions} This paper begins in \cref{sec:type-theory-examples}
-by studying small example programs from \theoryabbv to build intuition.
-From there, in \cref{sec:tt} we provide the syntax of \theoryname, a type theory
-that axiomatizes the underlying structures needed to conduct formal grammar
-theory. Later in \cref{sec:applications} we demonstrate the
-applicability of \theoryabbv for relating familiar grammar and automata
-formalisms as well as building concrete parsers.
-Then in \cref{sec:semantics-and-metatheory}, we give a denotational semantics
-that makes precise the connection between \theoryabbv syntax and formal grammars.
-
-Our contributions are then:
+them familiar to write and amenable to compositional verification techniques.
+
+To show the feasibility of our design, we have implemented \theoryabbv
+as a shallowly embedded domain-specific language in the Cubical Agda
+proof assistant \cite{VezzosiMortbergAbel2019}. We have implemented many example
+grammars and parsers in our system including regular expressions,
+non-deterministic and deterministic automata, as well as some example
+context-free grammars and parsers based on LL(1) and LR(1)
+automata. Throughout this paper, we will use \agdalogo~ to mark
+results that are mechanized in our Agda development.
+
+Our Agda prototype is based on a \emph{denotational semantics} of
+\theoryabbv. The core idea of the denotational semantics is
+\cite{elliottSymbolicAutomaticDifferentiation2021}, Elliott describes
+a formal grammar as type-level predicates on strings that prove
+language membership. That is, a \emph{formal grammar} $A$ is a
+function $\String \to \Set$ such that for a string $w$, $A~w$ is the
+set of ``proofs'' showing that $w$ belongs to the language recognized
+by $A$. We show that all linear types in \theoryabbv can be so
+interpreted as an abstract formal grammar in this sense, and that
+linear terms are a kind of \emph{parse transformer}, a function that
+takes a parse tree from one grammar to a parse tree in a different
+grammar but over the same underlying string.
+
+%% Formal verification encompasses a broad suite of techniques one may use to
+%% mathematically guarantee the correctness of software systems. We can only
+%% reap the benefits of
+%% any proof effort if its underlying assumptions hold. For instance, verifying the
+%% functional correctness of an algorithm is no use if it is run on buggy hardware.
+%% Likewise, the promised guarantees of a program may be jeopardized if it is
+%% parsed incorrectly.
+
+Our contributions are then:% \max{This contribution section needs to be punchier}
 %
 \begin{itemize}
-  \item \theoryname (\theoryabbv): A dependent
-        linear-non-linear type theory for building verified parsers
-  \item A denotational grammars semantics for \theoryabbv
-  \item An implementation of \theoryabbv in Agda
+  \item The design of \theoryname (\theoryabbv): A dependent
+    linear-non-linear type theory for building verified parsers, which
+    extends prior work on dependent linear-non-linear type theory to
+    support inductive linear types
+  \item Demonstration of how to encode many common grammar formalisms
+    (regular expressions, (non-)deterministic automata, context-free
+    grammars) and parser formalisms within our type theory.
+  \item A prototype implementation of \theoryabbv in Agda with all
+    examples mechanized.
+  \item A denotational semantics for \theoryabbv that shows that the
+    parsers are in fact verified to be correct and soundness of the
+    equational theory.
 \end{itemize}
+
+This paper begins in \cref{sec:type-theory-examples} by studying small
+example programs from \theoryabbv to build intuition.  From there, in
+\cref{sec:tt} we provide the syntax, typing and equational theory of
+\theoryname. In \cref{sec:applications} we demonstrate the
+applicability of \theoryabbv for relating familiar grammar and
+automata formalisms as well as building concrete parsers.  Then in
+\cref{sec:semantics-and-metatheory}, we give a denotational semantics
+that makes precise the connection between \theoryabbv syntax and
+formal grammars. Finally in \Cref{sec:discussion} we discuss related
+and future work.
 
@@
 \label{sec:type-theory-examples}
-To gain intuition for working in \theoryabbv, we begin with some illustrative
-examples drawn from the theory of formal languages. In particular, we
-demonstrate how to encode finite grammars, regular expressions, and finite
-automata in \theoryabbv. Each of our examples will be defined for strings over the three character alphabet
-$\Sigma = \{ \texttt{a} , \texttt{b}, \texttt{c} \}$.
+To gain intuition for working in \theoryabbv, we begin with some
+illustrative examples drawn from the theory of formal languages. Each
+of our examples will be defined for strings over the three character
+alphabet $\Sigma = \{ \texttt{a} , \texttt{b}, \texttt{c} \}$.
 
@@
 empty string, and the empty grammar.
-For instance, the grammar corresponding to the character $a$ is given by $\literal a$ which
+For each character $a$ in the alphabet we have a type $\literal a$ which
 has a single parse tree for the string
-\stringquote{a} and no parse trees at any other strings. The grammar $\I$ recognizes has a single
+\stringquote{a} and no parse trees at any other strings. The grammar $\I$ has a single
 parse tree for the empty string $\epsilon = \stringquote{}$ and no parses for any other strings.
-The final base type, the empty grammar $\bot$, has no parses for any string. We
+The final base type, the empty grammar $0$, has no parses for any string. We
 use type-theoretic syntax to represent disjunction $\oplus$ and concatenation
@@
 
-For a grammar $\A$, a parse tree of a string $\w$ can be represented as
-a term of type $\A$ in the context $\internalize \w$, where
+For a type $\A$, a parse tree of a string $\w$ is represented as a
+term of type $\A$ in the context $\internalize \w$, where
 $\internalize \w$ is a context with one variable for each character of
 $\w$. For example, to define a parse tree for \stringquote{ab}, we use
-the context $\internalize{\stringquote{ab}} = a : \literal a , b : \literal b$. Then the left of
-\cref{fig:fingram} gives a typing derivation that shows that $\inl(a
-\otimes b)$ defines a valid parse tree for the finite grammar
-$(\literal a \otimes \literal b)
-\oplus \literal c$. Underneath the derivation in \cref{fig:fingram}, we give a linear $\lambda$-term
-that implements the program given in the derivation.
+the context $\internalize{\stringquote{ab}} = a : \literal a , b :
+\literal b$. In \Cref{fig:fingram}, we give a lambda term and its
+typing derivation to define a parse for a finite grammar.
 
 \begin{figure}
-\begin{mathpar}
+\begin{lstlisting}
+f : (*@$\uparrow$@*)('a' (*@$\otimes$@*) 'b' (*@$\lto$@*) ('a' (*@$\otimes$@*) 'b') (*@$\oplus$@*) 'c')
+f (a , b) = inl (a (*@$\otimes$@*) b)
+\end{lstlisting}\begin{mathpar}
   \inferrule
@@
 \end{mathpar}
-\begin{lstlisting}
-f : (*@$\uparrow$@*)('a' (*@$\otimes$@*) 'b' (*@$\lto$@*) ('a' (*@$\otimes$@*) 'b') (*@$\oplus$@*) 'c')
-f (a , b) = inl (a (*@$\otimes$@*) b)
-\end{lstlisting}
 \caption{\stringquote{ac} matches
@@
 \end{figure}
-For this interpretation of terms as parse trees to make sense, our
+For this interpretation of parse trees as terms to make sense, our
 calculus cannot allow for \emph{any} of the usual structural rules of
-a type theory: weakening, contraction and exchange. Weakening allows
+type theory: weakening, contraction and exchange. Weakening allows
 for variables to go unused, while contraction allows for the same
@@
 $\oplus$, and $\otimes$, and the Kleene star $(\cdot)^{\ast}$.  For a
-grammar $\A$, we can define the Kleene star $\A^{*}$ as a particular
+grammar $\A$, we define the Kleene star $\A^{*}$ as a particular
 \emph{inductive linear type} of linear lists, as shown in
@@
 
-Through repeated application of the Kleene star constructors, \cref{fig:kleenestarderivation} gives a derivation that shows \stringquote{ab} matches the
-regular expression $({\literal a}^{*} \otimes \literal b) \oplus \literal c$. The leaves of the proof tree
-that mention the arrow $\uparrow$ describe a cast from a non-linear type to a linear type.
-For instance, the premise of the leaf involving $\nil$
-views $\nil : \ltonl {({\literal a}^{*})}$ as the name of a
-constructor, and a constructor should be nonlinearly valued because we may call
-it several time. However, the conclusion of this leaf views $\nil : {\literal a}^{*}$ as a
-linear value. After we call the constructor it returns a value that may only be
-used a single time. In this paper such casts are often made implicitly, but this
-does not introduce any ambiguity.
-Even though we use $\nil$ to refer to both a constructor and the
-linear value produced by it
-, we differentiate between the two views by inspecting
-the type.
-
-\steven{TODO fig 3 sizing}
+Through repeated application of the Kleene star constructors,
+\cref{fig:kleenestarderivation} gives a derivation that shows
+\stringquote{ab} matches the regular expression $({\literal a}^{*}
+\otimes \literal b) \oplus \literal c$. The leaves of the proof tree
+that mention the arrow $\uparrow$ describe a cast from a non-linear
+type to a linear type.  For instance, the premise of the leaf
+involving $\nil$ views $\nil : \ltonl {({\literal a}^{*})}$ as the
+name of a constructor, and a constructor should be nonlinearly valued
+because we may call it several times (or not at all). However, the
+conclusion of this leaf views $\nil : {\literal a}^{*}$ as a linear
+value, which in our syntax is an implicit coercion from a nonlinear
+value to a linear value. After we call the constructor it ``returns''
+a value that may only be used a single time.
+
 \begin{figure}
@@
 We may also have derivations where the term in context is not simply a
-string of base types. In \cref{fig:kleeneabstractproof} we show that every parse
+string of literals. In \cref{fig:kleeneabstractproof} we show that every parse
 of the grammar $(A \otimes A)^{*}$ induces a parse of $A^{*}$ for an
@@
 not quite appropriate
-to treat linear terms here as parse \textit{trees}.
+to think of a linear term here as a parse \textit{tree}.
 The
@@
 of \cref{fig:kleeneabstractproof} by pattern matching on the input and making an
-explicit recursive call in the body of its definition. When appropriate we will favor this code-like
-presentation of terms in our calculus, as it is more readable and more
-faithful to the implementation of \theoryabbv in Agda.
+explicit recursive call in the body of its definition.
 
@@
 Regular expressions are a compact formalism for defining a formal grammar,
-however simply defining a regular expression --- such as
-$({\literal a}^{*} \otimes \literal b) \oplus \literal c$ --- does not give any insight into how to parse
-it. Many parsing algorithms are based on a notion of automaton, and to verify
-these algorithms we need to represent their associated automata in
-\theoryabbv.
+but an expression such as
+$({\literal a}^{*} \otimes \literal b) \oplus \literal c$ does not give a very operational perspective of how to parse
+it. For this reason, most parsers are based on compiling a grammar to a corresponding notion of automaton, which is readily implemented. To implement
+these algorithms in \theoryabbv, we need a way to represent automata as types in the same way we can represent regular expressions.
 
@@
 finite automaton (NFA) for the regular expression
-$({\literal a}^{*} \otimes \literal b) \oplus \literal c$.
-\cref{fig:nfainductive} defines $\Trace$ an \emph{indexed} inductive linear type of
-traces through this automaton. Defining an indexed inductive type can be thought
-of like defining a family of mutually inductive types varying coherently over a
-type of indexing data. That is, \cref{fig:nfainductive} defines the linear types
-$\Trace~\0$,
-$\Trace~\1$, and $\Trace~\2$ mutually inductively. The index $\s : \texttt{Fin 3}$ picks out which
+$({\literal a}^{*} \otimes \literal b) \oplus \literal c$, along with a type $\Trace$, an \emph{indexed} inductive linear type of traces through this automaton.
+Defining an indexed inductive type can be thought
+of as defining a family of mutually recursive inductive types, one for each element of the indexing type. Here $\Trace$ 
+uses an index $\s : \texttt{Fin 3}$ which picks out which
 state in the automaton a trace begins at --- where $\texttt{Fin 3}$ is the
-finite type containing inhabitants $\{\0 , \1 , \2\}$.
-There are three kinds of constructors for $\Trace$:
-(1) those that terminate traces, (2) those that correspond to transitions
-labeled by a character, and (3) those that correspond to transitions labeled by
-the empty string $\epsilon$. The constructor $\texttt{stop}$ terminates a trace
-in the accepting state $\2$. The constructors $\texttt{1to1}$, $\texttt{1to2}$,
-$\texttt{0to2}$ each define a labeled transition through the NFA, and each of
-these consumes a parse of the label's character and a trace beginning at the
-destination of a transition to produce a trace beginning at the source of a
-transition. The constructor $\texttt{0to1}$ behaves similarly, except its
-transition is labeled with the empty string $\epsilon$. Therefore,
-$\texttt{0to1}$ takes in a trace beginning a state $\1$ and returns a trace
-beginning at state $\0$ corresponding to the same underlying string.
-
-Just as \cref{fig:kleenestarderivation} shows the \stringquote{ab} matches
-$({\literal a}^{*} \otimes \literal b) \oplus \literal c$, \cref{fig:nfaderivation} demonstrates that \stringquote{ab} is
-accepted by the NFA representation of this regular expression. In particular, we
-show that \stringquote{ab} matches the grammar $\Trace~0$ because $\0$ is the
-initial state of the NFA.\ We may imagine defining parse transformers from
-$({\literal a}^{*} \otimes \literal b) \oplus \literal c$ to $\Trace~\1$, and vice versa. The existence of
-such maps between the two types would demonstrate that the regular expression
-and the automaton capture the same language --- i.e.\ they recognize the same
-set of strings. As we will see in LATER SECTION, we can define parse
-transformers between regular expressions and their associated automaton.
-Moreover, we prove that these parse transformers define a bijective
-correspondence of parses over every input string. That is, in a strong sense the
-parse of a regular expression and the traces of their associated automata
-capture precisely the same data.
+finite type containing inhabitants $\{\0 , \1 , \2\}$. We can think of this as defining three mutually recursive inductive types $\Trace~\0$,
+$\Trace~\1$, and $\Trace~\2$.
+%
+There are three kinds of constructors for $\Trace$: (1) those that
+terminate traces, (2) those that correspond to transitions labeled by
+a character, and (3) those that correspond to transitions labeled by
+the empty string $\epsilon$. The constructor $\texttt{stop}$
+terminates a trace in the accepting state $\2$. The constructors
+$\texttt{1to1}$, $\texttt{1to2}$, $\texttt{0to2}$ each define a
+labeled transition through the NFA, and each of these consumes a parse
+of the label's character and a trace beginning at the destination of a
+transition to produce a trace beginning at the source of a
+transition. The constructor $\texttt{0to1}$ behaves similarly, except
+its transition is labeled with the empty string $\epsilon$. Therefore,
+$\texttt{0to1}$ takes in a trace beginning at state $\1$ and returns a
+trace beginning at state $\0$ corresponding to the same underlying
+string.
+%
+Lastly, we give a lambda term that constructs an accepting trace
+starting at the initial state for the string \stringquote{ab}.  Later
+in \Cref{sec:applications}, we will show that we can actually
+construct mutually inverse functions between the regular expression
+$({\literal a}^{*} \otimes \literal b) \oplus \literal c$ and its
+corresponding NFA traces ($\Trace~\0$) demonstrating that the regular
+expression and the automaton capture the same language. Further, since
+the functions are mutually inverse, this shows they are \emph{strongly
+equivalent} as grammars.
 
@@
   \end{tikzpicture}
-  \caption{NFA for $(\a^{*} \otimes \b) \oplus \c$}
-  \label{fig:exampleNFA}
-\end{figure}
-
-\begin{figure}
   \begin{lstlisting}
@@
   0to1 : (*@$\uparrow$@*)(Trace 1 (*@$\lto$@*) Trace 0)
-\end{lstlisting}
-\caption{$N$ as an indexed inductive type}
-\label{fig:nfainductive}
-\end{figure}
-
-\begin{figure}
-\begin{lstlisting}
+
 k : (*@$\uparrow$@*)(('a' (*@$\otimes$@*) 'b') (*@$\lto$@*)  Trace 0)
 k (a , b) = 0to1 (1to1 a (1to2 b stop))
+
 \end{lstlisting}
-\caption{\stringquote{ab} is accepted by the NFA for
-  $({\literal a}^{*} \otimes {\literal b }) \oplus \literal c$}
-\label{fig:nfaderivation}
+  \caption{NFA for $(\a^{*} \otimes \b) \oplus \c$ and its corresponding type}
+  \label{fig:exampleNFA}
 \end{figure}
@@
 
-The syntax of
-\theoryabbv is based on the dependent linear-non-linear (\lnld)
-calculus from \cite{krishnaswami_integrating_2015}. As in \lnld, \theoryabbv includes both
-non-linear dependent types, as well as linear types, which are allowed
-to depend on the non-linear types, but not on other linear types.
-%
-The main point of departure from \lnld's design in that the linear
-typing is \emph{non-commutative}, i.e., that exchange is not an
+The design of \theoryabbv is based on the dependent
+linear-non-linear calculus (\lnld) and Lambek calculus, also known as
+non-commutative linear logic.
+\cite{krishnaswami_integrating_2015,lambek58}. As in \lnld,
+\theoryabbv includes both non-linear dependent types, as well as
+linear types, which are allowed to depend on the non-linear types, but
+not on other linear types.
+%
+The main point of departure from \lnld's design is that, as in Lambek calculus \cite{lambek58}, the linear
+typing is \emph{non-commutative} --- i.e., that exchange is not an
 admissible structural rule. Furthermore, we add a general-purpose
@@
 %
-Finally, we add base types and axioms to the system specifically to
-model formal grammars and parsing.
+Finally, while \lnld was enhanced with special connectives
+inspired by separation logic to model imperative programming, we
+instead add base types and axioms to the system specifically to model
+formal grammars and parsing.
 
 The formation rules for the judgments of \theoryabbv are shown in
-\Cref{fig:formation}. $\Gamma$ stand for non-linear contexts, $X,Y,Z$
-stand for non-linear types, $M,N$ stand for non-linear terms, these
-act as in an ordinary dependent type theory. $\Delta$ stands for
-linear contexts, $A,B,C$ for linear types and $e,f,g$ for linear
+\Cref{fig:formation}. $\Gamma$ stands for non-linear contexts; $X,Y,Z$
+stand for non-linear types; $M,N$ stand for non-linear terms, these
+act as in an ordinary dependent type theory; $\Delta$ stands for
+linear contexts; $A,B,C$ for linear types; and, $e,f,g$ for linear
 terms. These contexts, types and terms are allowed to depend on an
@@
   \end{mathpar}
-  \caption{Formation Rules}
+  \caption{Formation rules}
   \label{fig:formation}
@@
 \subsection{Non-linear Typing}
-
 The non-linear types of \theoryabbv include the standard dependent
 types for $\Pi,\Sigma$, extensional equality, natural numbers,
-booleans, unit and empty types. We include these rules in the
-appendix. We present the other non-linear type constructions in
-\Cref{fig:non-linear-stuff}. First, we include universe types $U$ of
+booleans, unit and empty types, and we assume function
+extensionality\cite{Hofmann_1997}. We present the
+other non-linear type constructions in
+\Cref{fig:non-linear-types}. First, we include universe types $U$ of
 small non-linear types and $L$ of linear types. These are defined as
-``universes ala Coquand'' in that we define a judgment saying when a
+universes ``ala Coquand'' in that we define a judgment saying when a
 non-linear type is small and define the universe to internalize
-precisely this judgment. These universe types are needed so that we
-can define types by recursion on natural numbers. Next, we include a
-non-linear type $\ltonl A$ where $A$ is a linear type. The intuition
-for this type is that its elements are the linear terms that are
-``resource free'': its introduction rule says we can construct an
-$\ltonl A$ when we have a linear term of type $A$ with no free linear
-variables. This type is used extensively in our examples, playing a
-similar role to the $!$ modality of ordinary linear logic or the
-persitent modality $\square$ from separation logic.
+precisely this judgment \cite{coquandPresheafModel,lmcs:7713}. These universe
+types are needed so that we can define types by recursion on natural
+numbers. Next, we include a non-linear type $\ltonl A$ where $A$ is a
+linear type. The intuition for this type is that its elements are the
+linear terms that are ``resource free'': its introduction rule says we
+can construct an $\ltonl A$ when we have a linear term of type $A$
+with no free linear variables. This type is used extensively in our
+examples, playing a similar role to the $!$ modality of ordinary
+linear logic or the persistence modality $\square$ of separation logic
+\cite{girard_linear_1987,jung_higher-order_2016}.
 
@@
 \begin{figure}
+  \begin{footnotesize}
     \begin{mathpar}
@@
   \end{mathpar}
+  \end{footnotesize}
   \caption{Non-linear types (selection)}
@@
 
-We give an overview of the linear types in \Cref{fig:linear-types} and
-terms in \Cref{fig:linear-terms}. The equational theory for these
-types is straightforward and included in the appendix. First, the
-linear variable rule says that a linear variable can be used if it is
-the \emph{only} variable in the context.
+We give an overview of the linear types and terms in
+\Cref{fig:linear-types-and-terms}. The equational theory for these
+types is straightforward $\beta\eta$ equivalence and included in the
+appendix. First, the linear variable rule says that a linear variable
+can be used if it is the \emph{only} variable in the context.
 
 First, we cover the ``multiplicative'' connectives of non-commutative
-linear logic. The linear unit and tensor product are standard for a
-non-commutative linear logic: when we construct a linear unit we
-cannot use any variables and when we construct a tensor product, the
-two sides must use disjoint variables, and the variables the left side
-of the product uses must be to the left in the context of the
-variables used by the right side of the tensor product. Because we are
-non-commutative, there are two function types: $A \lto B$ and $B \tol
-A$. The difference between these is that the introduction rule for $A
-\lto B$ adds a variable to the right side of the context, whereas the
-introduction rule for $B \tol A$ adds a variable to the left side of
-the context. In our experience, because by convention parsing
-algorithms parse from left-to-right, we rarely need to use the $B \tol
-A$ connective. One of the main uses of linear types is that, in
-combination with the $\uparrow$ connective, we can abstract over
-linear terms, in that the non-linear type $\ltonl{A \lto B}$ is a
-non-linear type of linear terms with one linear input $A$ and output
-$B$.
-\pedro{Isn't this is a slightly incorrect desciption of $\ltonl{A \lto B}$?
-How is the description above distinct form $A \lto B$?}
-
-Next, we have the ``additive'' connectives. For these, we use the
+linear logic. The linear unit ($\I$) and tensor product ($\otimes$)
+are standard for a non-commutative linear logic: when we construct a
+linear unit we cannot use any variables and when we construct a tensor
+product, the two sides must use disjoint variables, and the variables
+the left side of the product uses must be to the left in the context
+of the variables used by the right side of the tensor product. The
+elimination rules for unit and tensor are given by pattern
+matching. The pattern matching rules split the linear context into
+three pieces $\Delta_1,\Delta_2,\Delta_3$: the middle $\Delta_2$ is
+used by the scrutinee of the pattern match, and in the continuation
+this context is replaced by the variables brought into scope by the
+pattern match. This ensures that pattern matches maintain the proper
+ordering of resource usage.
+
+Because we are non-commutative, there are two function types: $A \lto
+B$ and $B \tol A$, which have similar $\lambda$ introduction forms and
+application elimination forms. The difference between these is that
+the introduction rule for $A \lto B$ adds a variable to the right side
+of the context, whereas the introduction rule (elided) for $B \tol A$ adds a
+variable to the left side of the context. In our experience, because
+by convention parsing algorithms parse from left-to-right, we rarely
+need to use the $B \tol A$ connective. As we have already seen, the
+$\lto$ connective is frequently used in conjunction with the
+$\uparrow$ connective so that we can abstract non-linearly over linear
+functions.
+
+Next, we cover the ``additive'' connectives. First, we use the
 non-linear types to define \emph{indexed} versions of the additive
-disjunction $\oplus$ and additive conjunction $\&$ of linear
-logic. These are like a linear version of the $\Sigma$ and $\Pi$
-connectives of ordinary dependent type theory. The indexed $\&$ is
-defined by a $\lambda$ and eliminated using projection. The indexed
-$\oplus$ is like a \emph{weak} $\Sigma$ type: it has an injection
-introduction rule $\sigma$, but its elimination rule is given by
-\emph{pattern matching} rather than projections. Note that we can
-define the more typical nullary and binary versions of these
-connectives by using indexing over the empty and boolean type
-respectively. We will freely use $\bot$ to refer to this empty
-disjunction and $\top$ to refer to the empty conjunction. Lastly, we
-include a type $\equalizer {a}{f}{g}$ that we call the
-\emph{equalizer} of $f$ and $g$. We think of this type as the
-``subset'' of $A$ that satisfy the equation $f\,a\equiv g\,a$. If
-these were non-linear types, such a type could be constructed using
-the equality type as $\sum_{x:X} f\,x=_Y g\,x$, but this construction
-can't be used because it uses a \emph{dependent} equality type. We
-note that the equalizer type is not used in defining any of our
-parsers or formal grammars, but instead will be used shortly to derive
-an induction principle for proving equalities of functions whose
-domain is an inductive type.
+disjunction $\oplus$ and additive conjunction $\&$ of linear logic,
+which can be thought of as linear versions of the $\Sigma$ and $\Pi$
+connectives of ordinary dependent type theory, respectively. The
+indexed $\&$ is defined by a $\lambda$ that brings a \emph{non-linear}
+variable into scope and eliminated using projection where the index
+specified is given by a non-linear term. The rules for indexed
+$\oplus$ are analogous to a ``\emph{weak}'' $\Sigma$ type: it has an
+injection introduction rule $\sigma$, but its elimination rule is
+given by \emph{pattern matching} rather than first and second
+projections. We can define the more typical nullary and binary
+versions of these connectives by using indexing over the empty and
+boolean type respectively. We will freely use $0$ to refer to this
+empty disjunction and $\top$ to refer to the empty conjunction, and
+use infix $\oplus/\&$ for binary disjunction/conjunction.
+
+Lastly, we include a type $\equalizer {a}{f}{g}$ that we call the
+\emph{equalizer} of linear functions $f$ and $g$. We think of this
+type as the ``subtype'' of elements of $A$ that satisfy the equation
+$f\,a\equiv g\,a$. Note that it is important here that $f, g$
+themselves are non-linearly used functions, as linear values cannot be
+used in a type.  Equalizer types are not needed for non-linear types
+since they can be constructed using the equality type as $\sum_{x:X}
+f\,x=_Y g\,x$, but this construction can't be used for linear types
+because it uses a \emph{dependent} version of the equality type, which
+we cannot define as a linear type. While the equalizer type is not
+used directly in defining any of our parsers or formal grammars, it is
+used for several proofs, allowing for inductive arguments about our
+indexed inductive types.
+
+In addition to these type-theoretic principles, we need two additional
+axioms that do not generally hold in systems based on linear
+logic. First, we need that additive conjunction \emph{distributes}
+over additive disjunction --- e.g., in the finitary case that $0 \& A
+\cong 0$ and $(A + B) \& C \cong (A \& C) + (B \& C)$. In its most
+general form, the axiom says that the definable function
+${\bigoplus\limits_{f : \prod_{x:X}
+    Y(x)}\bigamp\limits_{x:X}A\,x\,(f\,x) \lto
+  \bigamp\limits_{x:X}\bigoplus\limits_{y:Y(x)}A\,x\,y}$ has an
+inverse. Second, we need that the different constructors of
+$\bigoplus$ are disjoint. We add this by adding the axiom that for any
+$A : X \to L$ and $x \neq x' : X$ there is a function
+$\uparrow(\equalizer{b}{\sigma\,x \circ \pi_1}{\sigma\,x'\circ \pi_2}
+\lto 0)$ where $b: A(x) \& A(x')$, i.e., the grammar of pairs of an
+$a:A(x)$ and an $a': A(x')$ such that $\sigma\,x\,a = \sigma\,x'\,a'$
+is empty.
+
+\pedro{imo, since we are providing an Agda implementation, we can
+  be looser with the explanation of the meaning of the disjointness axioms.
+  As it stands, it's a bit hard do follow along. Is there a simpler example
+  we could point to?}
 
 \begin{figure}
+  \begin{footnotesize}
   \begin{mathpar}
-    \inferrule{}{\Gamma \vdash I \isLinTy}\and
+    \inferrule{}{\Gamma \vdash \I \isLinTy}\and
     \inferrule{\Gamma \vdash A \isLinTy \and \Gamma \vdash B \isLinTy}{\Gamma \vdash A \otimes B \isLinTy}\and
@@
     \inferrule{\Gamma \vdash A \isLinTy \and \Gamma \vdash B \isLinTy}{\Gamma \vdash A \tol B \isLinTy}\and
-    \inferrule{\Gamma,x:X \vdash A \isLinTy}{\Gamma \vdash \bigoplus\limits_{x:X} A}\and
-    \inferrule{\Gamma,x:X \vdash A \isLinTy}{\Gamma \vdash \bigamp\limits_{x:X} A}\and
+    \inferrule{\Gamma,x:X \vdash A \isLinTy}{\Gamma \vdash \bigoplus\limits_{x:X} A \isLinTy}\and
+    \inferrule{\Gamma,x:X \vdash A \isLinTy}{\Gamma \vdash \bigamp\limits_{x:X} A \isLinTy}\and
     \inferrule
@@
   \end{mathpar}
-  \caption{Linear Types (selection)}
-\end{figure}
-
-\begin{figure}
 \begin{mathpar}
@@
     \and
-    \inferrule{\Gamma ; \Delta \vdash e : I \\ \Gamma ; \Delta_1',\Delta_2' \vdash e' : C}{\Gamma ; \Delta_1',\Delta,\Delta_2' \vdash \letin {()} e {e'} : C}
+    \inferrule{\Gamma ; \Delta_2 \vdash e : I \\ \Gamma ; \Delta_1,\Delta_3 \vdash e' : C}{\Gamma ; \Delta_1,\Delta_2,\Delta_3 \vdash \letin {()} e {e'} : C}
     %
@@
     %
-    \inferrule{\Gamma ; \Delta \vdash e : A \\ \Gamma ; \Delta' \vdash e' : B}{\Gamma ; \Delta, \Delta' \vdash e \otimes e' : A \otimes B}
+    \inferrule{\Gamma ; \Delta \vdash e : A \\ \Gamma ; \Delta' \vdash e' : B}{\Gamma ; \Delta, \Delta' \vdash (e , e') : A \otimes B}
     %
@@
     %
-    \inferrule{\Gamma ; \Delta \vdash e : A \otimes B \\ \Gamma ; \Delta'_1, a : A, b : B, \Delta'_2 \vdash e' : C}{\Gamma ;  \Delta_1', \Delta, \Delta'_2 \vdash \letin {a \otimes b} e {e'} : C}
+    \inferrule{\Gamma ; \Delta_2 \vdash e : A \otimes B \\ \Gamma ; \Delta_1, a : A, b : B, \Delta_2 \vdash e' : C}{\Gamma ;  \Delta_1, \Delta_2, \Delta_3 \vdash \letin {(a , b)} e {e'} : C}
     \\
@@
     \and
-    \inferrule{\Gamma ; \Delta \vdash e : \LinPiTy x X A \\ \Gamma \vdash M : X}{\Gamma ; \Delta \vdash e\,.\pi\,M : \subst A {e'} x}
+    \inferrule{\Gamma ; \Delta \vdash e : \LinPiTy x X A \\ \Gamma \vdash M : X}{\Gamma ; \Delta \vdash e\,.\pi\,M : \subst A {M} x}
     %
@@
     %
-    \inferrule{\Gamma \vdash M : X \quad \Gamma ; \Delta \vdash e : \subst A e x}{\Gamma ; \Delta \vdash \sigma\,M\,e : \bigoplus\limits_{x:X} A}
+    \inferrule{\Gamma \vdash M : X \quad \Gamma ; \Delta \vdash e : \subst A M x}{\Gamma ; \Delta \vdash \sigma\,M\,e : \bigoplus\limits_{x:X} A}
     %
@@
     %
-    \inferrule{\Gamma ; \Delta \vdash e : \bigoplus\limits_{x:X} A \quad \Gamma, x : X ; \Delta'_1, a : A, \Delta'_2 \vdash e' : C}{\Gamma; \Delta'_1, \Delta, \Delta'_2 \vdash \letin {\sigma\,x\,a} e {e'}: C}
+    \inferrule{\Gamma ; \Delta_2 \vdash e : \bigoplus\limits_{x:X} A \quad \Gamma, x : X ; \Delta_1, a : A, \Delta_3 \vdash e' : C}{\Gamma; \Delta_1, \Delta_2, \Delta_3 \vdash \letin {\sigma\,x\,a} e {e'}: C}
     %
@@
       \Gamma ; \Delta \vdash \applto {f}{e} \equiv \applto {g}{e}}
-    {\Gamma ; \Delta \vdash \equalizerin{e} : \equalizer {e}{f}{g}}
+    {\Gamma ; \Delta \vdash \equalizerin{e} : \equalizer {a}{f}{g}}
     %
@@
   \end{mathpar}    
-  \caption{Linear terms (selection)}
-  \label{fig:linsyntax}
+  \end{footnotesize}
+  \caption{Linear types and terms (selection)}
+  \label{fig:linear-types-and-terms}
 \end{figure}
-\pedro{Fixed tensor elimination rule (added output type)}
-\pedro{Perhaps we should modify the equalizer introduction rule.
-  As it stands, readers might be confused by the same symbol $e$
-  being used everywhere.}
 
@@
 
-Next, we introduce one of the most complex and important linear
-type constructors of our development, \emph{indexed inductive linear
-types} $\mu F$. To encode these, we first add a non-linear type
-$\SPF\,X$ of \emph{strictly-positive functorial} linear type
-expressions indexed by a non-linear type $X$. We give the rules for
-strictly-positive functors in \Cref{fig:spf}. First, we add operations
-for a recursive reference $\Var~x$ the constant functor, $K$, tensor
-products and additive conjunction and disjunction. As an example,
-\max{TODO: make an example}.  Next we add an operation $\el(F)$ which
-interprets our syntactic codes as an actual type constructor, and
-define it for each of the operations given above. Lastly, we give a
-\emph{functorial action} $\map(F)$ for each of the codes.\max{TODO:
-  find a citation for this kind of approach. Container papers work
-  this way for instance. I think there is a linear container paper
-  that uses this approach}
-
-With this notion of strictly-positive functor so defined, in
-\Cref{fig:iilt} we next add a linear type constructor $\mu F$, which
-defines a family of mutually-recursive inductive types for the family
-of type expressions $F$. The introduction rule for this is
-$\roll$, which constructs an element of $\mu F\,x$ from its
-one-level unfolding. The elimination principle is defined by a
-$\fold$ operation. The equations ensure that this makes the
-family an \emph{initial algebra} for the type expressions. That is the
-$\beta$ rule says that a $\mathsf{fold}$ applied to a $\mathsf{roll}$
-is equivalent to $\mathsf{map}$ping the $\mathsf{fold}$ over all the
-sub-expressions, which means that $\mathsf{fold}$ is an algebra
-homomorphism from $\mathsf{roll}$ to $\mathsf{f}$. Then the $\eta$
-rule says that $\mathsf{fold}$ is the \emph{unique} such homomorphism.
+Next, we introduce the most complex and important linear type
+constructors of our development, \emph{indexed inductive linear
+types}. We encode these by adding a mechanism for constructing initial
+algebras of strictly positive functorial type expressions, following
+prior work on inductive types
+\cite{nakov_quantitative_2022,altenkirch_indexed_2015}. The syntax is
+given in \Cref{fig:iilt}. First, we add a non-linear type $\SPF\,X$ of
+\emph{strictly positive functorial} linear type expressions indexed by
+a non-linear type $X$. We think of the elements of this type as
+syntactic descriptions of linear types that are parameterized by
+$X$-many variables standing for linear types that are only used in
+strictly positive positions. Accordingly, the $\SPF\,X$ type supports
+an operation $\el$ that interprets it as such a type constructor, as
+well as an operator $\map$ that defines a functorial action on parse
+transformers. The $\SPF\,X$ type supports constructors for a reference
+$\Var~x$ to one of the linear type variables, a constant expression
+that doesn't mention any type variables $K$, as well as tensor
+products and additive conjunction and disjunction of type expressions.
+We additionally add equations in the appendix that say that the
+$\el$/$\map$ operations correspond to these descriptions of the
+constructors.
+
+\begin{figure}
+  \begin{footnotesize}
+    \begin{mathpar}
+      \inferrule{\Gamma \vdash X\isTy}{\Gamma \vdash \SPF\,X \isTy}\and
+      \inferrule{\Gamma \vdash X\isSmall}{\Gamma \vdash \SPF\,X \isSmall}\and
+    \el : \prod_{X:U}\SPF\,X \to (X \to L) \to L\and
+    \map : \prod_{X:U}\prod_{F : \SPF\,X}\prod_{A,B:X\to L}{\left(\prod_{x:X}\uparrow(\unquoteTy{A\,x}\lto \unquoteTy{B\,x})\right)} \to {\uparrow(\unquoteTy{\el(F)(A)} \lto \unquoteTy{\el(F)(B)})}\and
+    \mathsf{Var} : \prod_{X:U} X \to \SPF\,X\and
+    \mathsf{K} : \prod_{X:U} L \to \SPF\,X\and
+    \mathsf{\bigoplus} : \prod_{X:U}\prod_{Y:U}(Y \to \SPF\,X) \to \SPF\,X\and
+    \mathsf{\bigamp} : \prod_{X:U}\prod_{Y:U}(Y \to \SPF\,X) \to \SPF\,X\and
+    \mathsf{\otimes} : \prod_{X:U}\SPF\,X \to \SPF\,X \to \SPF\,X\and
+    \roll : \prod_{X:U}\prod_{F:X \to \SPF\,X}\prod_{x:X}\ltonl{(\el(F\,x)(\mu\,F))}\and
+    \fold : \prod_{X:U}\prod_{F:X\to\SPF\,X}\prod_{A:X \to L}
+    \left(\prod_{x:X}\ltonl{(\unquoteTy{\el(F\,x)(\unquoteTy{A})} \lto \unquoteTy{A\,x})}\right)
+    \to \prod_{x:X}\ltonl{(\mu F\,x \lto A\,x)}\and
+
+    \inferrule*[right=Ind$\beta$]
+    {\Gamma \vdash f : \prod_{x:X}\ltonl{(\el(F\,x)(A) \lto A\,x)} \and \Gamma; \Delta \vdash e : \el(F\,x)(\mu\,F)}
+    {\Gamma;\Delta \vdash \fold\,F\,f\,x\,(\roll e) \equiv f\,x\,(\map(F\,x)\,(\fold\,F\,f)) : A\,x}
+
+    \inferrule*[right=Ind$\eta$]
+    {\Gamma \vdash f : \prod_{x:X}\ltonl{(\el(F\,x)(A) \lto A\,x)}
+     \and \Gamma \vdash e : \prod_{x:X}\ltonl{(\mu F\,x \lto A\,x)}\\\\
+      \Gamma,x:X;a:\el(F\,x)(\mu F) \vdash e\,x\,(\roll a) \equiv f\,x\,(\map(F\,x)\,e) : A\,x }
+    {\fold\,F\,f\equiv e' : \prod_{x:X}\ltonl{(\mu F\,x \lto A\,x)}}
+  \end{mathpar}
+  \end{footnotesize}
+  %% \begin{footnotesize}
+  %%   \begin{align*}
+  %%   \mathsf{Var} &: \prod_{X:U} X \to \SPF\,X\\
+  %%   \mathsf{K} &: \prod_{X:U} L \to \SPF\,X\\
+  %%   \mathsf{\bigoplus} &: \prod_{X:U}\prod_{Y:U}(Y \to \SPF\,X) \to \SPF\,X\\
+  %%   \mathsf{\bigamp} &: \prod_{X:U}\prod_{Y:U}(Y \to \SPF\,X) \to \SPF\,X\\
+  %%   \mathsf{\otimes} &: \prod_{X:U}\SPF\,X \to \SPF\,X \to \SPF\,X\\
+  %%   \el &: \prod_{X:U}\SPF\,X \to (X \to A) \to L\\
+  %%   \map &: \prod_{X:U}\prod_{F : \SPF\,X}\prod_{A,B:X\to L}{\left(\prod_{x:X}\uparrow(\unquoteTy{A\,x}\lto \unquoteTy{B\,x})\right)} \to {\uparrow(\unquoteTy{\el(F)(A)} \lto \unquoteTy{\el(F)(B)})}\\\\
+  %%   \roll &: \prod_{X:U}\prod_{F:X \to \SPF\,X}\prod_{x:X}\ltonl{(\el(F\,x)(\mu\,F))}\\
+  %%   \fold &: \prod_{X:U}\prod_{F:X\to\SPF\,X}\prod_{A:X \to L}
+  %%   \left(\prod_{x:X}\ltonl{(\el(F\,x)(A) \lto A\,x)}\right)
+  %%   \to \prod_{x:X}\ltonl{(\mu F\,x \lto A\,x)}
+  %% \end{align*}
+  %% \end{footnotesize}
+  \caption{Strictly positive functors and indexed inductive linear types}
+  \label{fig:iilt}
+\end{figure}
+
+Next, given a family of $X$-many strictly positive linear type
+expressions $F : X \to \SPF\,X$, we define a family $\mu F : X \to L$
+of $X$-many mutually recursive inductive types. The introduction rule
+for this is $\roll$, which constructs an element of $\mu F\,x$ from
+the one-level of the $x$th type expression. The elimination principle
+is defined by a mutual $\fold$ operation: given a family of output
+types $A$ indexed by $X$, we can define a family of functions from
+$\mu F\,x \multimap A\,x$ if you specify how to interpret all of the
+constructors as operations on $A$ values. We add $\beta\eta$ equations
+that specify that this makes the family $\mu F$ into an \emph{initial
+algebra} for the functor $\el(F)$. That is the $\beta$ rule says that
+a $\mathsf{fold}$ applied to a $\mathsf{roll}$ is equivalent to
+$\mathsf{map}$ping the $\mathsf{fold}$ over all the sub-expressions,
+which means that $\mathsf{fold}$ interprets all of the constructors
+homomorphically using the provided interpretation $\mathsf{f}$. Then
+the $\eta$ rule says that $\mathsf{fold}$ is the \emph{unique} such
+homomorphism, i.e. anything that satisfies the recurrence equation of
+the $\mathsf{fold}$ is equal to it.
 
@@
 directly in proofs. In dependent type theory, we would have a
-dependent \emph{elimination} principle, which justifies
-\emph{inductive proofs}. However since linear types do not support
-dependency on linear types, we cannot directly adapt this
-approach. However, if we are trying to prove that two morphisms out of
-a mutually recursive type are equal, we can use the \emph{equalizer}
-type to prove their equality by induction.\max{TODO: show how this is
-  done.}
-
-\begin{figure}
-  \begin{align*}
-    \mathsf{Var} &: X \to \SPF\,X\\
-    \mathsf{K} &: L \to \SPF\,X\\
-    \mathsf{\bigoplus} &: (Y \to \SPF\,X) \to \SPF\,X\\
-    \mathsf{\bigamp} &: (Y \to \SPF\,X) \to \SPF\,X\\
-    \mathsf{\otimes} &: \SPF\,X \to \SPF\,X \to \SPF\,X\\
-    \el &: \prod_{X:U}\SPF\,X \to (X \to L) \to L\\
-    \map &: \prod_{X:U}\prod_{F : \SPF\,X}\prod_{A,B:X\to L}(\prod_{x:X}\uparrow(A\,x\lto B\,x)) \to \uparrow(\el(F)(A) \lto \el(F)(B))\\\\
-    \el(\Var\,M) B &= B\,M\\
-    \el(\mathsf{K}\,A) B &= A\\
-    \el(\bigoplus A) B &= \bigoplus_{y:Y}\el(A y)B\\
-    \el(\bigamp A) B &= \bigamp_{y:Y}\el(A y)B\\
-    \el(A \otimes A') B &= \el(A)B \otimes \el(A')B\\\\
-    \map(\Var\,M)\,f &= f\,M\\
-    \map(\K\,A)\,f &= \lambda a. a\\
-    \map(\bigoplus A)\,f &= \lambda a. \letin{\oplusinj y {a_y}}{a}{\oplusinj y {\map(A\,y)\,f\,a_y}}\\
-    \map(\bigwith\,A)\,f &= \lambda a. \withlamb{y}{\map(A\,y)\,f\,(\withprj y a)}\\
-    \map(A \otimes A')\,f &= \lambda b. \letin{(a,a')}{b}{(\map(A)\,f\,a,\map(A')\,f\,a')}
-  \end{align*}
-  \begin{mathpar}
-  \end{mathpar}
-  \caption{Strictly Positive Functors}
-  \label{fig:spf}
-\end{figure}
-
-\begin{figure}
-  \begin{mathpar}
-    \mathsf{roll} : \prod_{X:U}\prod_{F:X \to \SPF\,X}\prod_{x:X}\ltonl{(\el(F\,x)(\mu\,F))}
-
-    %% \inferrule*[right=IndIntro]
-    %% {\Gamma \vdash A : X \to \textrm{SPF}\,X\and
-    %%  \Gamma \vdash M : X\and
-    %%  \Gamma; \Delta \vdash e : \textrm{el}(A\,M) (\mu\,A)}
-    %% {\Gamma; \Delta \vdash \mathsf{roll}\, e : \mu\,A\, M}
-
-    \mathsf{fold} : \prod_{X:U}\prod_{F:X\to\SPF\,X}\prod_{A:X \to L}
-    (\prod_{x:X}\ltonl{(\el(F\,x)(A) \lto A\,x)})
-    \to \prod_{x:X}\ltonl{(\mu F\,x \lto A\,x)}
-
-    \inferrule*[right=Ind$\beta$]
-    {}
-    {\fold\,F\,f\,x\,(\roll e) \equiv f\,x\,(\map(F\,x)\,(\fold\,F\,f))}
-
-    \inferrule*[right=Ind$\eta$]
-    {x:X;a:\el(F\,x)(\mu F) \vdash e'\,x\,(\roll a) \equiv f\,x\,(\map(F\,x)\,e')
-    }
-    {\fold\,F\,f\equiv e'}
-    %% {\Gamma,x:X; a:\mu\,A\,x \vdash e' : B\,x
-    %%   \and \Gamma;\Delta \vdash e : \mu A M\\\\
-    %%   \Gamma,x:X; a':\el(A x)(\mu A) \vdash
-    %%   e'[x/x][\roll(a')/a] = f[x/x][\map(A\,x)\,(y.a.e'[y/x][a/a])\,a/b] : B\,x
-    %% }
-    %% {\Gamma; \Delta \vdash \fold(A)(x.b.f)(M,e) = e'[M/x][e/a] : B\,M}
-  \end{mathpar}
-  \caption{Indexed Inductive Linear Types}
-  \label{fig:iilt}
-\end{figure}
+dependent \emph{elimination} principle, which can be used to
+implement functions by recursion as well as proofs by
+induction. Unfortunately, since linear types do not support dependency on
+linear types, we cannot directly adapt this approach. However, if we
+are trying to prove that two morphisms out of a mutually recursive
+type are equal, we can use the \emph{equalizer} type to prove their
+equality by induction. That is, if our goal is to prove two functions
+$f,g : \uparrow(\mu F\,x \lto A\,x)$ equal, it suffices to
+implement a function $\mathsf{ind} : \uparrow(\mu F\,x \lto \equalizer
+{a} f g)$ such that $\textrm{ind}(a) \equiv a$. Then an
+inductive-style proof can be implemented by constructing
+$\mathsf{ind}$ using a $\fold$. This can all be justified using only
+the $\beta\eta$ principles for equalizers and inductive types, and
+this is how our most complex inductive proofs are implemented in the Agda
+formalization.
+\pedro{I don't like the inductive proof explanation above, but I
+  don't know how I would phrase it}
 
@@
 
-So far, our calculus is a generic combination of dependent types with
-non-commutative linear types. In order to carry out formal grammar
-theory and define parsers, however, we add a few carefully chosen base
-types, terms and axioms that allow us to reason about formal grammars.
-\max{TODO: more prose, also are these all of our axioms?}
-\begin{enumerate}
-\item (Characters) For each character of our alphabet $c \in \Sigma$, we add a linear type $c \isLinTy$.
-\item (Strings are total) The function $\ltonl{(\String \lto \top)}$ is an isomorphism.\max{for this need to define $\String$, and isomorphism, and possibly we should define $\top$ in more detail earlier?
-idk}
-\item (Distributivity of $\oplus,\&$) The function  $\ltonl{(\bigoplus\limits_{f : \prod_{x:X} Y(x)}\bigamp\limits_{x:X}A\,x\,(f\,x) \lto \bigamp\limits_{x:X}\bigoplus\limits_{y:Y(x)}A\,x\,y)}$ is an isomorphism.
-\item ($\ltonl$ preserves $\oplus$) The function $\sum_{x:X}\ltonl{(A\,x)} \to \ltonl{(\bigoplus\limits_{x:X}A\,x)}$ is an isomorphism.
-\end{enumerate}
+So far, our calculus is a somewhat generic combination of dependent
+types with non-commutative linear types. In order to carry out formal
+grammar theory and define parsers, we need only add a few
+grammar-specific constructions. First for each character in our
+alphabet, we add a corresponding linear type $c$. We can then define a
+non-linear type $\Char$ as the disjunction of all of these characters,
+and define a type $\String$ as the Kleene star of $\Char$, i.e. as an
+inductive linear type.
+Then we add a function $\mathsf{read} :
+\ltonl{(\top\lto \String)}$ that intuitively ``reads'' the input
+string from the input and makes it available. It is important that the
+input type of $\mathsf{read}$ is $\top$, which can control any amount
+of resources, and not $\I$ which controls no resources. Further, we
+add an axiom that $\lambda s. \mathsf{read}(!(s)) \equiv \lambda
+s. s$ where $!$ is the unique function $\ltonl(\String\lto \top)$,
+i.e., that if we have a string, but then throw it away and read
+it from the input, then in fact that is equivalent to the string we
+were given originally. This ensures that the elements of the $\String$
+type always stand for the actual input string in our reasoning. In the
+next section, we will show how these basic principles are enough to
+provide a basis for verified parsing and formal grammar theory.
 
@@
 
-\paragraph{Equivalence}
-In \cite{chom1963}, Chomsky differentiates between the weak generative capacity
-of a formal grammar --- the set of strings accepted by it --- and its strong
-generative capacity --- the set of ``structural descriptions'', or parse trees,
-that it generates. Two grammars are said to be \emph{weakly equivalent} if they
-agree in their weak generative capacity, and likewise \emph{strongly equivalent}
-if they agree in their strong generative capacity. We aim to recover each of
+%% \paragraph{Weak and Strong Equivalence}
+In the theory of formal grammars, there are two different notions of
+equivalence: up to weak generative capacity, meaning just which strings are
+accepted by the grammar; and up to \emph{strong} generative capacity, when
+the parse trees of the two grammars are isomorphic
+\cite{chom1963}. Using linear types as grammars, we can define both of
 these notions of equivalence in \theoryabbv.
-
 \begin{definition}
   \label{def:weakequiv}
-  Grammars $\A$ and $\B$ are \emph{weakly equivalent} if there exists parse
-  transformers $\f : \ltonl{(\A \lto \B)}$ and $\g : \ltonl {(\B \lto \A)}$.
+  Grammars $\A$ and $\B$ are \emph{weakly equivalent} if there exists
+  parse transformers $\f : \ltonl{(\A \lto \B)}$ and $\g : \ltonl {(\B
+    \lto \A)}$. $\A$ is a \emph{retract} of $\B$ if $\A$ and $\B$ are
+  weakly equivalent and $\lambda a. g(f(a)) \equiv \lambda a.a$. They
+  are \emph{strongly equivalent} if further the other composition is
+  the identity, i.e., $\lambda b. f(g(b)) \equiv \lambda b.b$.
 \end{definition}
-\pedro{Up to this point, we have mentioned a couple of times parse transformers,
-  but we have never defined what they are. In particular, why does the definition
-  require the $\uparrow$ modality. I'm assuming that it's to properly deal with
-  linear variables, but these things are not obvious at first.}
-
-\newcommand{\lang}{\texttt{Lang}}
-A parse transformer $\f : \ltonl{(\A \lto \B)}$ takes as input parses of $\A$ for string
-$\w$ and returns a parse of $\B$ for $\w$. Therefore, $\f$ demonstrates an
-inclusion of languages $\lang(\A) \subset \lang(\B)$. Similarly, the existence of
-$\g$ shows that $\lang(\B) \subset \lang(\A)$, and thus $\lang(\A) = \lang(\B)$.
-
-\steven{Need better notation for language that is cohesive throughout the paper}
-
-\begin{definition}
-  \label{def:strongequiv}
-  Grammars $\A$ and $\B$ are \emph{strongly equivalent} if there exists parse
-  transformers $\f : \ltonl{(\A \lto \B)}$ and $\g : \ltonl {(\B \lto \A)}$ such
-  that $\f \circ \g \equiv \ident$ and $\g \circ \f \equiv \ident$.
-\end{definition}
-
-For a string $\w$, \cref{def:strongequiv} induces a bijection between the
-$\A$-parses of $\w$ and the $\B$-parses of $\w$, thus demonstrating that $\A$
-and $\B$ agree in their strong generative capacity.
-
-\paragraph{Ambiguity}
-A formal grammar $\A$ is ambiguous if there is a string $\w$ for which $\A$ has
-several parse trees of $\w$.
-For example, $\a \oplus \a$ is ambiguous because
-$\stringquote{a}$ has two parse trees --- on either side of the disjunction. On
-the other hand, a formal grammar is unambiguous when there is at most
-one parse tree for any input string. These notions of ambiguity are entirely
-semantic, so we seek to find a property expressible in the syntax of \theoryabbv
-that reifies the semantic description of ambiguity.
-
+
+%% That is, grammars are weakly equivalent if we can construct a parse
+%% tree of $B$ from a parse tree of $A$ and vice-versa, whereas they are
+%% strongly equivalent if these processes are inverse to each other. The
+%% definition of retract is a useful intermediate notion. In retrospect,
+%% our axiom about $\String$ says that $\String$ is a retract of $\top$.
+
+%% \newcommand{\lang}{\texttt{Lang}}
+%% A parse transformer $\f : \ltonl{(\A \lto \B)}$ takes as input parses of $\A$ for string
+%% $\w$ and returns a parse of $\B$ for $\w$. Therefore, $\f$ demonstrates an
+%% inclusion of languages $\lang(\A) \subset \lang(\B)$. Similarly, the existence of
+%% $\g$ shows that $\lang(\B) \subset \lang(\A)$, and thus $\lang(\A) = \lang(\B)$.
+
+%% \steven{Need better notation for language that is cohesive throughout the paper}
+
+%% \begin{definition}
+%%   \label{def:strongequiv}
+%%   Grammars $\A$ and $\B$ are \emph{strongly equivalent} if there exists parse
+%%   transformers $\f : \ltonl{(\A \lto \B)}$ and $\g : \ltonl {(\B \lto \A)}$ such
+%%   that $\f \circ \g \equiv \ident$ and $\g \circ \f \equiv \ident$.
+%% \end{definition}
+
+%% For a string $\w$, \cref{def:strongequiv} induces a bijection between the
+%% $\A$-parses of $\w$ and the $\B$-parses of $\w$, thus demonstrating that $\A$
+%% and $\B$ agree in their strong generative capacity.
+
+%% \paragraph{Ambiguity}
+A formal grammar $\A$ is ambiguous if there are multiple parse trees
+for the \emph{same} string $\w$.  For example, $\a \oplus \a$ is
+ambiguous because there are two parses of $\stringquote{a}$,
+constructed using $\inl$ and $\inr$. On the other hand, a formal
+grammar is unambiguous when there is at most one parse tree for any
+input string. We can capture this notion as a type in \theoryabbv in a
+clever way:
 \begin{definition}
@@
   $\f : \ltonl{(\B \lto \A)}$, and $\g : \ltonl{(\B \lto \A)}$
-  then $\f \equiv \g$
+  then $\f \equiv \g$.
 \end{definition}
-\cref{def:unambig} can be read more intuitively as stating that $\A$ is
-unambiguous if there is at most one way to transform parses of any other grammar
-$\B$ into parse of $\A$.
-
-Note that this definition of unambiguity is preserved by strong equivalence, so a common proof technique for showing $\A$ is unambiguous is by proving $\A$
-is strongly equivalent to an unambiguous grammar $\B$. In this style of proof, we
-build the functions $\f : \ltonl{(\A \lto \B)}$ and
-$\g : \ltonl{(\B \lto \A)}$, then we have the obligation to show
-$\f \circ \g \equiv \ident$ and $\g \circ \f \equiv \ident$. However, because $\B$ is
-unambiguous there is only one function of type $\ltonl{(\B \lto \B)}$, so
-$\f \circ \g : \ltonl{(\B \lto \B)}$ and $\ident : \ltonl{(\B \lto \B)}$ are equal.
-Thus, in this type of proof, we really only have the obligation of building each of $\f$
-and $\g$ and demonstrating that $\g \circ \f \equiv \ident$.
+\cref{def:unambig} can be read more intuitively as stating that $\A$
+is unambiguous if there is at most one way to transform parses of any
+other grammar $\B$ into parse of $\A$. This notion of an unambiguous
+type is the analog for linear types of the definition of a (homotopy)
+\emph{proposition} in the terminology of homotopy type
+theory\cite{hottbook}. The most basic unambiguous types are $\top$ and
+$0$, and in a system of classical logic all unambiguous types would
+have to be equivalent to one of these, but with our axioms we can show
+also that $\I$ and literals $\literal c$ are unambiguous. To see this, first, we
+establish two useful properties of unambiguity.
+\begin{lemma}[\agdalogo]
+  If $\B$ is unambiguous and $\A$ is a retract of $\B$ then $\A$ is unambiguous.
+  If a disjunction $\bigoplus\limits_{x:X} A(x)$ is unambiguous then
+  each $A(x)$ is unambiguous.
+\end{lemma}
+From the first principle, we can prove that $\String$ is unambiguous,
+since it is a retract of $\top$. In fact, observe that if $\A$ is a
+retract of $\B$ and $\B$ is unambiguous, then in fact $\A$ and $\B$
+are strongly equivalent, as the equation $\lambda b. f(g(b)) \equiv
+\lambda b. b$ follows because $\B$ is unambiguous. Therefore
+$\String$ is also strongly equivalent to $\top$. Next, since $\String$ is
+defined as a Kleene star, we can easily show that $\String \cong \I
+\oplus \Char \oplus (\Char \otimes \Char \otimes \String)$. Then by
+the second principle we have that $\I$ and $\Char$ and each literal
+$\literal c$ are unambiguous as well.
 
@@
 
+%% \paragraph{Partial and Total Parsers}
+
+We now turn to our main task, which is using our linear type system to
+implement verified parsers. Given a grammar defined as a linear type
+$A$, a first attempt at defining a parser would be to implement a
+function $\uparrow{(\String \lto A)}$. But since our linear functions
+must be total, this means that we can construct an $A$ parse for
+\emph{every} input string, which is impossible for most grammars of
+interest. Instead we might try to write a partial function as a
+$\ltonl{(\String \lto (A \oplus \top))}$ using the ``option''
+monad. This allows for the possibility that the input string doesn't
+parse, but is far too weak as a specification: we can trivially
+implement a parser for any type by always returning $\inr$. The
+correct notion of a parser should be one that allows for failure, but
+only in the case that a parse cannot be constructed.
+\begin{definition}
+  A parser for a linear type $A$ is a function $\uparrow{(\String \lto
+    A \oplus A_{\neg})}$, where $A_{\neg}$ is a linear type that is
+  \emph{disjoint} from $A$ in that we can implement a function
+  $\uparrow{(A \& A_{\neg} \lto 0)}$.
+\end{definition}
+Here we replace $\top$ in our partial parser type with a type
+$A_{\neg}$ that we can think of as a negation of $A$. The function
+$\uparrow{(A \& A_{\neg} \lto 0)}$ ensures that it is impossible for
+$A$ and $A_{\neg}$ to match the same input string. This means that in
+defining a parser, we will need to define a kind of negative grammar
+for strings that do not parse. Fortunately, we will see that
+deterministic automata naturally support such a notion with no
+additional effort: the negative grammar is simply the grammar for
+traces that end in a rejecting state. This follows from the following
+principle, a consequence of our axiom that $\oplus$ constructors are
+disjoint.
+\begin{lemma}[\agdalogo]
+  If $A \oplus A_{\neg}$ is unambiguous, then $A$ and $A_{\neg}$ are
+  disjoint.
+\end{lemma}
+
+Writing a parser as a linear term in this way is an intrinsic
+verification of the \emph{soundness} of the parser completely for free
+from the typing: any $\inl$ parse that we return \emph{must}
+correspond to a parse tree of the input string. Further if we verify
+the disjointness property we then also get the \emph{completeness} of
+the parser as well, that it never fails to generate an $A$ parse when
+it is possible.
+
+Our main method for constructing verified parsers is to show that a
+grammar $A$ is weakly equivalent to a grammar for a deterministic
+automaton. Parsers for deterministic automata are simple to implement
+by stepping through the states of the automaton, with the rejecting
+traces serving as the negative grammar. This is sufficient due to the
+following:
+\begin{lemma}[\agdalogo]
+  \label{lem:wk-eqv-parse}
+  If $\A$ is weakly equivalent to $\B$ then any parser for $\A$ can be
+  extended to a parser for $\B$.
+\end{lemma}
+Here we need both directions of the weak equivalence. We need $A \lto
+B$ to extend the parser from $\String \lto A \oplus A_{\neg}$ to
+$\String \lto B \oplus A_{\neg}$ but then we also need $B \lto A$ to
+establish that $A_{\neg}$ is disjoint from $B$.
+
 \subsection{Regular Expressions and Finite Automata}
 %% 2. Regex, NFA, DFA
-In this section, we make the equivalences
-between regular expressions and finite automata precise by encoding two theorems
-from classical language theory: (1) every regular expression is strongly equivalent
-to the traces through some non-deterministic finite automaton (NFA); and, (2) for
-every NFA, there is some deterministic finite automaton (DFA) such that the type
-of traces through the NFA is weakly equivalent to the type of traces through the DFA.
+
+In this section, we describe how to construct an intrinsically
+verified parser for regular expressions by compiling it to an NFA and
+then a DFA. That is, for each regular expression $A$, we construct an
+NFA $N(A)$ and a corresponding DFA $D(A)$ such that $A$ is strongly
+equivalent to the traces of $N(A)$ and weakly equivalent to the
+accepting traces of $D(A)$. Then we can easily construct a parser for
+traces of $D(A)$ and combine this using \Cref{lem:wk-eqv-parse} to get
+a verified regex parser.
 
@@
 
-\paragraph{Regular Expressions} The theory of regular expressions emerges as the
-subtheory of \theoryabbv generated by
-$\otimes$, $\oplus$, and Kleene star over the types $\bot$, $I$, and characters drawn from $\Sigma$.
-
-\paragraph{Nondeterministic Finite Automata} In \cref{sec:type-theory-examples}, we saw
-one particular NFA corresponding to the regular expression
-$({\literal a}^{*} \otimes \literal b) \oplus \literal c$. The definition of the linear type of traces
-through this
-NFA, as
-presented in \cref{fig:nfainductive}, inspects the concrete structure of the
-NFA.\ When generalizing from this example to an arbitrary NFA, we do not have the
-luxury of inspecting a concrete definition. Therefore, we abstract over the
-specification of an arbitrary NFA and define traces with respect to that
-specification.
-
-Let the data of an NFA $\N$ be given by the following nonlinear data,
-\begin{itemize}
-\item $\N.\states$ --- a finite set of states
-\item $\N.\init : \N.\states$ --- the initial state
-\item $\N.\isAcc : \N.\states \to \Bool$ --- a mapping marking which states are
-        accepting
-\item $\N.\transitions$ --- a finite set of labeled transitions
-\item $\N.\labelt : \N.\transitions \to \Sigma$ --- a mapping from $N.\transitions$ to the character labeling the transition
-\item $\N.\src , \N.\dst : \N.\transitions \to \N.\states$ --- mappings from
-        $\N.\transitions$ to the source and target states, respectively, of a labeled transition
-\item $\N.\epstransitions$ --- a finite set of $\varepsilon$-transitions
-\item $\N.\epssrc, \N.\epsdst : \N.\epstransitions \to \N.\states$ --- mappings
-        from $\N.\epstransitions$ to the source and target states of an $\epsilon$-transition
-\end{itemize}
-
-In \cref{fig:nfatrace} we define a linear type of traces
-through an NFA $\N$. $\Trace_{\N}$ is an inductive type indexed by the first
-state $\s : \N.\states$ of a trace and a Boolean $b : \Bool$ that marks if the
-trace ends in an accepting state. Traces in $\N$ may be built through one of three constructors. We
-may terminate a trace at an accepting state with the constructor $\nil$. If we
-had a trace beginning at the destination state of a transition, then we may use
-the $\cons$ constructor to linearly combine that trace with a parse of the label
-of the transition to build a trace beginning at the source of the transition.
-Finally, if we had a trace beginning at the destination of an
-$\epsilon$-transition then we may use $\epscons$ to pull it back along the
-$\epsilon$-transition and construct a trace beginning at the source of the
-$\epsilon$-transition. As a shorthand, write $Parse_{N}$ for the accepting
-traces out of $N.init$.
-
+A regular expressions in \theoryabbv is a linear type constructed
+using only the connectives $\literal c$, $0$, $\oplus$, $I$,
+$\otimes$, and Kleene star.  In \cref{sec:type-theory-examples}, we
+saw one particular NFA and its corresponding type of traces. More
+generally we define the linear type of traces as in
+\Cref{fig:nfatrace}
+
+%% The definition of the linear type of traces
+%% through this
+%% NFA, as
+%% presented in \cref{fig:nfainductive}, inspects the concrete structure of the
+%% NFA.\ When generalizing from this example to an arbitrary NFA, we do not have the
+%% luxury of inspecting a concrete definition. Therefore, we abstract over the
+%% specification of an arbitrary NFA and define traces with respect to that
+%% specification.
+
+%% Let the data of an NFA $\N$ be given by the following nonlinear data,
+%% \begin{itemize}
+%% \item $\N.\states$ --- a finite set of states
+%% \item $\N.\init : \N.\states$ --- the initial state
+%% \item $\N.\isAcc : \N.\states \to \Bool$ --- a mapping marking which states are
+%%         accepting
+%% \item $\N.\transitions$ --- a finite set of labeled transitions
+%% \item $\N.\labelt : \N.\transitions \to \Sigma$ --- a mapping from $N.\transitions$ to the character labeling the transition
+%% \item $\N.\src , \N.\dst : \N.\transitions \to \N.\states$ --- mappings from
+%%         $\N.\transitions$ to the source and target states, respectively, of a labeled transition
+%% \item $\N.\epstransitions$ --- a finite set of $\varepsilon$-transitions
+%% \item $\N.\epssrc, \N.\epsdst : \N.\epstransitions \to \N.\states$ --- mappings
+%%         from $\N.\epstransitions$ to the source and target states of an $\epsilon$-transition
+%% \end{itemize}
+
+In \cref{fig:nfatrace} we define a linear type of traces through an
+NFA $\N$. $\Trace_{\N}$ is an inductive type indexed by the starting
+state of the trace $\s : \N.\states$. Traces in $\N$ may be built
+through one of three constructors. We may terminate a trace at an
+accepting state with the constructor $\nil$. Here we use an Agda-style
+unicode syntax for $\bigamp$, as well as using the function arrow to
+mean a non-dependent version of $\bigamp$. If we had a trace beginning
+at the destination state of a transition, then we may use the $\cons$
+constructor to linearly combine that trace with a parse of the label
+of the transition to build a trace beginning at the source of the
+transition.  Finally, if we had a trace beginning at the destination
+of an $\epsilon$-transition then we may use $\epscons$ to pull it back
+along the $\epsilon$-transition and construct a trace beginning at the
+source of the $\epsilon$-transition. As a shorthand, write $Parse_{N}$
+for the accepting traces out of $N.init$.
+\newcommand{\D}{\texttt{D}}
 \begin{figure}
 \begin{lstlisting}
-data Trace(*@$_{\N}$@*) : (s : N.states) (b : Bool) (*@$\to$@*) L where
-  nil : (*@$\uparrow$@*)(&[ s (*@$\in$@*) N.states ] Trace(*@$_{\N}$@*) s (N.isAcc s))
-  cons : (*@$\uparrow$@*)(&[ t (*@$\in$@*) N.transitions ] &[ b (*@$\in$@*) Bool ]
-           ('N.label t' (*@$\lto$@*) Trace(*@$_{\N}$@*) (N.dst t) b (*@$\lto$@*) Trace(*@$_{\N}$@*) (N.src t) b))
-  (*@$\epsilon$@*)cons : (*@$\uparrow$@*)(&[ t (*@$\in$@*) N.(*@$\epsilon$@*)transitions ] &[ b (*@$\in$@*) Bool ]
-           (Trace(*@$_{\N}$@*) (N.(*@$\epsilon$@*)dst t) b (*@$\lto$@*) Trace(*@$_{\N}$@*)(N.(*@$\epsilon$@*)src t) b))
+data Trace(*@$_{\N}$@*) : (s : N.states) (*@$\to$@*) L where
+  nil : (*@$\uparrow$@*)(&[ s : N.states ] N.isAcc (*@$\to$@*) Trace(*@$_{\N}$@*) s)
+  cons : (*@$\uparrow$@*)(&[ t : N.transitions ]
+    ('N.label t' (*@$\lto$@*) Trace(*@$_{\N}$@*) (N.dst t) (*@$\lto$@*) Trace(*@$_{\N}$@*) (N.src t)))
+  (*@$\epsilon$@*)cons : (*@$\uparrow$@*)(&[ t : N.(*@$\epsilon$@*)transitions ]
+    (Trace(*@$_{\N}$@*) (N.(*@$\epsilon$@*)dst t) (*@$\lto$@*) Trace(*@$_{\N}$@*)(N.(*@$\epsilon$@*)src t)))
+
+data Trace(*@$_{\D}$@*) : (s : D.states) (b : Bool) (*@$\to$@*) L where
+  nil : (*@$\uparrow$@*)(&[ s : D.states ] Trace(*@$_{\D}$@*) s D.isAcc s)
+  cons : (*@$\uparrow$@*)(&[ c : Char ] &[ s : D.states ] &[ b : Bool ]
+  ('c' (*@$\lto$@*) Trace(*@$_{\D}$@*) (D.(*@$\delta$@*) c s) b (*@$\lto$@*) Trace(*@$_{\D}$@*) s b))
 \end{lstlisting}
-\caption{Traces through an NFA $N$ as an indexed inductive type}
+\caption{Traces of an NFA $N$ and a DFA $D$}
 \label{fig:nfatrace}
@@
 
-\paragraph{Deterministic Finite Automata}
-\newcommand{\D}{\texttt{D}}
-Just as the above defines an NFA, we provide a specification for an arbitrary
-deterministic finite automaton.\ A DFA $D$ is
-given by the following data,
-
-\begin{itemize}
-\item $\D.\states$ --- a finite set of states
-\item $\D.\init : \D.\states$ --- the initial state
-  \item $\D.\isAcc : \D.\states \to \Bool$ --- a mapping marking which states
-        are accepting
-\item $\D.\delta : \Sigma \to \D.\states \to \D.\states$ --- a deterministic transition function. Given a character $\c$ and a state
-        $\s$, $\D.\delta~\c~\s$ is the state for which there is a transition $\s \overset{\stringquote{c}}{\to} (\D.\delta~\c~\s)$
-\end{itemize}
-
-\begin{figure}
-\begin{lstlisting}
-data Trace(*@$_{\D}$@*) : (s : D.states) (b : Bool) (*@$\to$@*) L where
-  nil : (*@$\uparrow$@*)(&[ s (*@$\in$@*) D.states ] Trace(*@$_{\D}$@*) s D.isAcc s)
-  cons : (*@$\uparrow$@*)(&[ c (*@$\in$@*) (*@$\Sigma$@*) ] &[ s (*@$\in$@*) D.states ] &[ b (*@$\in$@*) Bool ]
-             ('c' (*@$\lto$@*) Trace(*@$_{\D}$@*) (D.(*@$\delta$@*) c s) b (*@$\lto$@*) Trace(*@$_{\D}$@*) s b))
-\end{lstlisting}
-\caption{Traces through the DFA $D$ as an indexed inductive type}
-\label{fig:dfatrace}
-\end{figure}
-
-$\Trace_{\D}$, the linear type of traces through $\D$, is given in
-\cref{fig:dfatrace}. These
-traces may be terminated in an accepting state $\s$ with the $\nil$ constructor.
-The $\cons$ constructor builds a trace out of state $\s$ by linearly combining
-a parse of some character $\c$ with a
-trace out of the state $\D.\delta~\c~\s$. The trace built with $\cons$ is
-accepting if and only if the trace out of $D.\delta~c~s$ is accepting.
-
-Because DFAs are deterministic, we are able to prove that their type of traces
-are unambiguous and define a parser for them. In particular we show that
-$\LinPiTy {s} {D.\states} {\LinSigTy {b} {\Bool} \Trace_{D}~s~b}$ is strongly
-equivalent to
-$\stringg$, which simultaneously constructs a parser and demonstrates that
-traces through $D$ is unambiguous. Intuitively this strong equivalence says that we may
-unambiguously build traces through $D$ beginning at any state $s$, and further
-that from any trace through $D$ we can uniquely recover the string that built
-the trace.
-
-The first half of the equivalence,
+%% \paragraph{Deterministic Finite Automata}
+%% Just as the above defines an NFA, we provide a specification for an arbitrary
+%% deterministic finite automaton.\ A DFA $D$ is
+%% given by the following data,
+
+%% \begin{itemize}
+%% \item $\D.\states$ --- a finite set of states
+%% \item $\D.\init : \D.\states$ --- the initial state
+%%   \item $\D.\isAcc : \D.\states \to \Bool$ --- a mapping marking which states
+%%         are accepting
+%% \item $\D.\delta : \Sigma \to \D.\states \to \D.\states$ --- a deterministic transition function. Given a character $\c$ and a state
+%%         $\s$, $\D.\delta~\c~\s$ is the state for which there is a transition $\s \overset{\stringquote{c}}{\to} (\D.\delta~\c~\s)$
+%% \end{itemize}
+
+%% \begin{figure}
+
+%% \caption{Traces through the DFA $D$ as an indexed inductive type}
+%% \label{fig:dfatrace}
+%% \end{figure}
+
+$\Trace_{\D}$, the linear type of traces through $\D$, is given
+next. Unlike traces for an NFA, we parameterize this type additionally
+by a boolean which says whether the trace is accepting or
+rejecting. These traces may be terminated in an accepting state $\s$
+with the $\nil$ constructor.  The $\cons$ constructor builds a trace
+out of state $\s$ by linearly combining a parse of some character $\c$
+with a trace out of the state $\D.\delta~\c~\s$. The trace built with
+$\cons$ is accepting if and only if the trace out of $D.\delta~c~s$ is
+accepting.
+
+Because DFAs are deterministic, we are able to prove that their type
+of traces are unambiguous and define a parser for them. In particular
+we show that for any start state $s$, ${\LinSigTy {b} {\Bool}
+  \Trace_{D}~s~b}$ is a retract of $\stringg$. That is, first we
+construct a function $\parse_{D}$ that is a parser for
+$\Trace_D~{s}~{true}$, with $\Trace_D~{s}~{false}$ being the disjoint
+type used. The fact that these types are disjoint follows by showing
+that this function is part of a retraction, i.e., that there is only
+one way to trace through a deterministic automaton.
+
+The parser,
 $\parse_{D}$,
-is defined by induction on strings in \cref{fig:dfaparser}. If this string is empty, then $\parse_{D}$
+is defined by recursion on strings in \cref{fig:printdfa}. If this string is empty, then $\parse_{D}$
 defines a linear function that terminates a trace at the input state $s$. If the
@@
 the character at the head of the string.
+%
+The inverse, $\print_{D}$ is defined by recursion on traces. If the trace is defined
+via $\nil$, then $\print_{D}$ returns the empty string. Otherwise, if the trace
+is defined by $\cons$ then $\parse_{D}$ appends the character from the most
+recent transition to the output
+string and recurses.
+%
+We prove this is a retraction by induction on traces.
+
+\begin{theorem}[\agdalogo]
+  $\parse_D s$ is a parser for $\Trace_D~s~\true$.
+\end{theorem}
 
@@
 \begin{lstlisting}
-parse(*@$_D$@*) : (*@$\uparrow$@*)(String (*@$\lto$@*) &[ s (*@$\in$@*) D.states ] (*@$\oplus$@*)[ b (*@$\in$@*) Bool ] Trace(*@$_{\D}$@*) s b)
+parse(*@$_D$@*) : (*@$\uparrow$@*)(String (*@$\lto$@*) &[ s : D.states ] (*@$\oplus$@*)[ b : Bool ] Trace(*@$_{\D}$@*) s b)
 parse(*@$_D$@*) String.nil s = (*@$\sigma$@*) (D.isAcc s) (Trace(*@$_{\D}$@*).nil s)
@@
    (*@$\sigma$@*) b (Trace(*@$_{\D}$@*).cons c s b a t)
-\end{lstlisting}
-\caption{Parser for DFA traces}
-\label{fig:dfaparser}
-\end{figure}
-
-The second half of the equivalence,
-$\print_{D}$,
-is given in \cref{fig:printdfa} by induction on traces. If the trace is defined
-via $\nil$, then $\print_{D}$ returns the empty string. Otherwise, if the trace
-is defined by $\cons$ then $\parse_{D}$ appends the character from the most
-recent transition to the output
-string and recurses.
-
-\begin{figure}
-\begin{lstlisting}
-print(*@$_D$@*) : (s : D.states) (*@$\to$@*) (*@$\uparrow$@*)(((*@$\oplus$@*)[ b (*@$\in$@*) Bool ] Trace(*@$_{\D}$@*) s b) (*@$\lto$@*) String)
+
+print(*@$_D$@*) : (s : D.states) (*@$\to$@*) (*@$\uparrow$@*)(((*@$\oplus$@*)[ b : Bool ] Trace(*@$_{\D}$@*) s b) (*@$\lto$@*) String)
 print(*@$_D$@*) s ((*@$\sigma$@*) b (Trace(*@$_{\D}$@*).nil .s)) = String.nil
@@
 \end{lstlisting}
-\caption{Printer for DFA traces}
+\caption{Parser/printer for DFA traces}
 \label{fig:printdfa}
@@
 
-\steven{TODO make the Agda logo clickable, and mention its meaning somewhere}
-\newcommand{\Agda}{\agdalogo}
+Working backwards, we can then show the traces of an NFA are weakly
+equivalent to the traces of a DFA implementing a variant of Rabin and
+Scott's classic powerset construction
+\cite{rabinFiniteAutomataTheir1959}. Here we note that this is
+\emph{only} a weak equivalence and not a strong equivalence, as the
+DFA is unambiguous even if the NFA is not.
 \begin{theorem}[Determinization, \Agda]
@@
 \begin{proof}
-  We present a variant of Rabin and Scott's classic powerset construction \cite{rabinFiniteAutomataTheir1959}. First, we build
-  the DFA $\D$. Define the
-  states of $\D$ to be the $\mathbb{P}_{\varepsilon}(\N.\states)$ --- the type
+  Define the states of $\D$ to be the $\mathbb{P}_{\varepsilon}(\N.\states)$ --- the type
   of $\epsilon$-closed\footnote{A subset of states $\X$ is $\epsilon$-closed if
@@
   $\epsilon$-closure of all the states reachable from $X$ via a transition
-  labelled with the character $\c$.
+  labeled with the character $\c$.
 
@@
   $\ltonl{(Parse_{N} \lto Parse_{D})}$, we strengthen our inductive hypothesis
-  and build the term
+  and build a term
 \newcommand{\NtoD}{\texttt{NtoD}}
@@
 
-\begin{figure}
-\begin{lstlisting}
-NtoD : (s : N.states) (*@$\to$@*)
-       (*@$\uparrow$@*)(Trace(*@$_{\N}$@*) s true (*@$\lto$@*)
-           &[ X (*@$\in$@*) D.states ] &[ sInX (*@$\in$@*) (X (*@$\ni$@*) s) ] Trace(*@$_{\D}$@*) X true)
-NtoD .s (nil s) X sInX = nil s
-NtoD .(N.src t) (N.cons trans .true a dstTrace) X sInX =
-  D.cons (N.label t) X true a
-    (((NtoD s dstTrace).(*@$\pi$@*) (D.(*@$\delta$@*) (N.label t) X)).(*@$\pi$@*) (N.dst t))
-NtoD .(N.(*@$\epsilon$@*)src t) ((*@$\epsilon$@*)cons (*@$\epsilon$@*)trans .true (*@$\epsilon$@*)DstTrace) X sInX =
-  ((NtoD s (*@$\epsilon$@*)DstTrace).(*@$\pi$@*) X).(*@$\pi$@*) (N.(*@$\epsilon$@*)dst t)
-\end{lstlisting}
-\caption{Parse transformer from NFA traces to Powerset DFA traces}
-\label{fig:NtoD}
-\end{figure}
-
-Because the state $\s$ is now arbitrary, we can define $\NtoD$ via
-recursion on the trace through the NFA. Code for this is given in \cref{fig:NtoD}.
-If the NFA trace is terminating at the accepting state $\s$ and $\s \in \X$,
-then we return a trace in $\D$ that terminates at
-$\X$, which is accepting because it contains the accepting state $\s$.
-
-If the NFA trace is defined via $\cons$ for a labeled transition
-$t : N.transitions$ such that $N.src~t \in \X$, then
-we first decompose the trace into a parse of $N.label~t$ and a
-trace starting from $N.dst~t$. We then recursively call $\NtoD$
-on this trace which returns a family of traces parameterized by $\epsilon$-closed
-subsets containing $N.\dst~t$. We instantiate this family at the
-set of states
-$\D.\delta~(N.label~t)~X$. Because $\N.\dst~t$ is
-reachable from $\N.\src~\texttt{trans}$ by a single transition labeled by
-$N.label~t$ and $N.src~t \in X$ we have that
-$N.dst~t \in \D.\delta~(N.label~t)~X$. Therefore
-we can linearly combine the parse of the label and the trace built from the
-recursive call to get a trace in $\D$ rooted at $X$.
-
-Lastly, if the NFA trace is defined via $\epsilon\cons$ for an $\epsilon$-transition
-$t : \N.\epstransitions$ such that
-$\N.\epssrc~t \in \X$, then the construction is quite
-similar. In this case we again recursively call $\NtoD$, except afterwards we
-instantiate the family of grammars at the set $X$ itself. We may do this because
-all states in $\D$ are taken to be $\epsilon$-closed, therefore
-$\N.\epsdst~t \in \X$ because
-$\N.\epssrc~t$ is in $X$. This concludes the construction of $\NtoD$.
+%% \begin{figure}
+%% \begin{lstlisting}
+%% NtoD : (s : N.states) (*@$\to$@*)
+%%        (*@$\uparrow$@*)(Trace(*@$_{\N}$@*) s true (*@$\lto$@*)
+%%            &[ X : D.states ] &[ sInX : (X (*@$\ni$@*) s) ] Trace(*@$_{\D}$@*) X true)
+%% NtoD .s (nil s) X sInX = nil s
+%% NtoD .(N.src t) (N.cons trans .true a dstTrace) X sInX =
+%%   D.cons (N.label t) X true a
+%%     (((NtoD s dstTrace).(*@$\pi$@*) (D.(*@$\delta$@*) (N.label t) X)).(*@$\pi$@*) (N.dst t))
+%% NtoD .(N.(*@$\epsilon$@*)src t) ((*@$\epsilon$@*)cons (*@$\epsilon$@*)trans .true (*@$\epsilon$@*)DstTrace) X sInX =
+%%   ((NtoD s (*@$\epsilon$@*)DstTrace).(*@$\pi$@*) X).(*@$\pi$@*) (N.(*@$\epsilon$@*)dst t)
+%% \end{lstlisting}
+%% \caption{Parse transformer from NFA traces to Powerset DFA traces}
+%% \label{fig:NtoD}
+%% \end{figure}
+
+%% Because the state $\s$ is now arbitrary, we can define $\NtoD$ via
+%% recursion on the trace through the NFA. Code for this is given in \cref{fig:NtoD}.
+%% If the NFA trace is terminating at the accepting state $\s$ and $\s \in \X$,
+%% then we return a trace in $\D$ that terminates at
+%% $\X$, which is accepting because it contains the accepting state $\s$.
+
+%% If the NFA trace is defined via $\cons$ for a labeled transition
+%% $t : N.transitions$ such that $N.src~t \in \X$, then
+%% we first decompose the trace into a parse of $N.label~t$ and a
+%% trace starting from $N.dst~t$. We then recursively call $\NtoD$
+%% on this trace which returns a family of traces parameterized by $\epsilon$-closed
+%% subsets containing $N.\dst~t$. We instantiate this family at the
+%% set of states
+%% $\D.\delta~(N.label~t)~X$. Because $\N.\dst~t$ is
+%% reachable from $\N.\src~\texttt{trans}$ by a single transition labeled by
+%% $N.label~t$ and $N.src~t \in X$ we have that
+%% $N.dst~t \in \D.\delta~(N.label~t)~X$. Therefore
+%% we can linearly combine the parse of the label and the trace built from the
+%% recursive call to get a trace in $\D$ rooted at $X$.
+
+%% Lastly, if the NFA trace is defined via $\epsilon\cons$ for an $\epsilon$-transition
+%% $t : \N.\epstransitions$ such that
+%% $\N.\epssrc~t \in \X$, then the construction is quite
+%% similar. In this case we again recursively call $\NtoD$, except afterwards we
+%% instantiate the family of grammars at the set $X$ itself. We may do this because
+%% all states in $\D$ are taken to be $\epsilon$-closed, therefore
+%% $\N.\epsdst~t \in \X$ because
+%% $\N.\epssrc~t$ is in $X$. This concludes the construction of $\NtoD$.
 
@@
 
-For a moment, let us turn our attention to the content behind the traces in $\D$.
-Traces from subset $\X$ to subset $\Y$ describes the \emph{existence} of some trace
-in the $\N$ from some $\x \in \X$ to some $\y \in \Y$. In general it is not clear how
-to extract out a choice of trace in $\N$, which is what $\DtoN$ is asking of us.
-However, we may further take the states and transitions of $\N$ to be \emph{ordered}.
-This allows us to
-induce an order on traces in $\D$ from $\X$ to $\Y$. From these orderings we can then
-define the necessary choice functions on each type by taking the smallest with
-respect to the ordering.
-
-We define $\DtoN$ by recursion on traces through $\D$. First if the trace stops at
-some accepting $\epsilon$-closed subset $\X$, then we may choose some state
-$\s \in X$ such that $\s$ is accepting and terminate the trace in $\N$ at $\s$.
-
-Next if the trace in $\D$ is defined via transition
-$\X \overset{\stringquote{c}}{\to} (\D.\delta~\c~\X)$ then we may choose some state in
-$\s \in \D.\delta~\c~\X$. Recursively, $\DtoN$ sends the trace rooted at
-$\D.\delta~\c~\X$ to one rooted at $\s$. By definition of $\D.\delta~\c~\X$,
-there exists some $\s' \in \X$ such that $\s'$ transitions to $\s$ by some number of
-$\epsilon$-transitions followed by a transition labeled by $\c$.
-We order all
-such traces of this form and choose the smallest one, which is then linearly
-combined with the parse of $\c$ to build an $\N$ trace. This concludes the
-definition of $\DtoN$, thus showing that $Parse_{N}$ and $Parse_{D}$ are
-weakly equivalent.
+%% For a moment, let us turn our attention to the content behind the traces in $\D$.
+%% Traces from subset $\X$ to subset $\Y$ describes the \emph{existence} of some trace
+%% in the $\N$ from some $\x \in \X$ to some $\y \in \Y$. In general it is not clear how
+%% to extract out a choice of trace in $\N$, which is what $\DtoN$ is asking of us.
+%% However, we may further take the states and transitions of $\N$ to be \emph{ordered}.
+%% This allows us to
+%% induce an order on traces in $\D$ from $\X$ to $\Y$. From these orderings we can then
+%% define the necessary choice functions on each type by taking the smallest with
+%% respect to the ordering.
+
+%% We define $\DtoN$ by recursion on traces through $\D$. First if the trace stops at
+%% some accepting $\epsilon$-closed subset $\X$, then we may choose some state
+%% $\s \in X$ such that $\s$ is accepting and terminate the trace in $\N$ at $\s$.
+
+%% Next if the trace in $\D$ is defined via transition
+%% $\X \overset{\stringquote{c}}{\to} (\D.\delta~\c~\X)$ then we may choose some state in
+%% $\s \in \D.\delta~\c~\X$. Recursively, $\DtoN$ sends the trace rooted at
+%% $\D.\delta~\c~\X$ to one rooted at $\s$. By definition of $\D.\delta~\c~\X$,
+%% there exists some $\s' \in \X$ such that $\s'$ transitions to $\s$ by some number of
+%% $\epsilon$-transitions followed by a transition labeled by $\c$.
+%% We order all
+%% such traces of this form and choose the smallest one, which is then linearly
+%% combined with the parse of $\c$ to build an $\N$ trace. This concludes the
+%% definition of $\DtoN$, thus showing that $Parse_{N}$ and $Parse_{D}$ are
+%% weakly equivalent.
 \end{proof}
 
+Finally, given any regular expression we can construct a
+\emph{strongly} equivalent NFA. While only weak equivalence is
+required to construct a parser, proving the strong equivalence shows
+that other aspects of formal grammar theory are also verifiable in
+\theoryabbv.
 \newcommand{\R}{\texttt{R}}
@@
 \end{theorem}
-
 \begin{proof}
-  \steven{TODO this proof needs to be rewritten. Likely through use of equalizers}
-  \pedro{Since this proof and the determinization one are well-known and we have
-    proved it in Agda, it's probably a good idea to include a much shorter proof sketch,
-    or move it to the appendix, though this would feel a bit superfluous.}
-We present a variant of Thompson's construction
-\cite{thompsonProgrammingTechniquesRegular1968} in which we recurse on the
-structure of $\R$ to construct a strongly equivalent NFAs. In this paper we will walk through
-some of the cases of this proof. The remaining cases
-follow
-similar reasoning and the details may be found in the formalized artifact.
-
-\begin{figure}
-  \begin{tikzpicture}[node distance = 17mm ]
-    \node[state, initial] (0) {$\0$};
-    \node[state, right of=0, accepting] (1) {$\1$};
-
-    \path[->] (0) edge[above] node{$\stringquote{c}$} (1);
-  \end{tikzpicture}
-  \caption{NFA $\N_{\c}$ for a single character $\c$}
-  \label{fig:litnfa}
-\end{figure}
-
-First handle the case where $R$ is a literal character $\literal c$. The NFA $N_{c}$ pictured in
-\cref{fig:litnfa} is strongly equivalent to the grammar $c$. To demonstrate this
-strong equivalence, we build functions between $\literal c$ and $Parse_{N_{c}}$
-and show they are mutually inverse.
-
-\begin{figure}
-\begin{lstlisting}
-A : (s : Fin 2) (*@$\to$@*) L
-A 0 = 'c'
-A 1 = I
-
-toN(*@$_c$@*) : (s : Fin 2) (*@$\to$@*) (*@$\uparrow$@*)(A s (*@$\lto$@*) Trace(*@$_{N_c}$@*) s true)
-toN(*@$_c$@*) 0 c = cons t true c (nil 1)
-toN(*@$_c$@*) 1 () = nil 1
-
-fromN(*@$_c$@*) : (s : Fin 2) (*@$\to$@*) (*@$\uparrow$@*)(Trace(*@$_{N_c}$@*) s true (*@$\lto$@*) A s)
-fromN(*@$_c$@*) 0 (cons t .true c tr) = let a (*@$\otimes$@*) () = c (*@$\otimes$@*) fromN(*@$_c$@*) 1 tr in a
-fromN(*@$_c$@*) 1 (nil .1) = ()
-\end{lstlisting}
-\caption{Parse transformer out of $Trace_{N_{c}}~s~true$}
-\label{fig:litNFAterms}
-\end{figure}
-
-In \cref{fig:litNFAterms} we define a family of grammars $A$ that serve as
-interpretations for each state in $N_{c}$. We then prove that each $A~s$ is
-equivalent to the accepting traces through $N_{c}$ out of state $s$.
-
-When
-defining $fromN_{c}$ we pattern match on the accepting traces beginning in each
-state. Out of state $0$ the only accepting traces are built via $cons$ over the
-single transition $t$, and out
-of state $1$ the only accepting trace is built via $nil$.
-
-Each $A~s$ is unambiguous, so $(fromN_{c}~s) \circ (toN_{c}~s)$ is equal to the
-identity on each $A~s$. To complete the proof of strong equivalence it remains
-to show that $(toN_{c}~s) \circ (fromN_{c}~s)$ is equal to the identity on
-$Trace~s~true$, which we show by using the equalizer type.
-
-\begin{figure}
-\begin{lstlisting}
-eqHom(*@$_{N_c}$@*) : (s : N(*@$_c$@*).state) (*@$\to$@*)
-      (*@$\uparrow$@*)(Trace(*@$_{N_c}$@*) s true (*@$\lto$@*) {tr | (toN(*@$_c$@*) s (*@$\circ$@*) fromN(*@$_{c}$@*) s) tr = id tr})
-\end{lstlisting}
-\caption{Signature for the equalizer term}
-\label{fig:litequalizer}
-\end{figure}
-
-The type signature for the family of functions, $eqHom_{N_{c}}$, into the equalizer type is given in
-\cref{fig:litequalizer}. By defining an $Trace_{N_{c}}$-algebra homomorphism
-into the equalizer, we enable a
-proof strategy for equality by induction on traces through $N_{c}$. First consider the traces
-out of state $1$. The only case to handle here is $nil 1$, and by definition
-$toN_{c}~1$ and $fromN_{c}~1$ invert each other.
-When defining $eqHom_{N_{c}}$ for traces out of state $0$, we must leverage an
-inductive hypothesis,
-
-\begin{align*}
-  (toN_{c}~0 \circ fromN_{c}~0) (cons~t~true~c~tr)
-  & \equiv (toN_{c}~0) (let~a~\otimes~()~=~c~\otimes~fromN_{c}~1~tr~in~a) \\
-  & \equiv cons~t~true~(let~a~\otimes~()~=~c~\otimes~fromN_{c}~1~tr~in~a)~(nil~1) \\
-  & \equiv cons~t~true~c~(nil~1) \\
-  & \equiv cons~t~true~c~tr \\
-\end{align*}
-
-The first two equalities are by defintion. The thrid equality holds because
-$\literal c$ is unambiguous and
-$let~a~\otimes~()~=~c~\otimes~fromN_{c}~1~tr~in~a$ and $c$ have type
-$\literal c$. The final equality holds because our inductive hypothesis states
-that $nil~1 \equiv tr$. Thus, each $A s$ and $Trace_{N_{c}}~s~true$ are strongly
-equivalent --- in particular, $\literal c$ and $Parse_{N_{c}}$.
-
-\steven{Rewrite the disjunction case with equalizers}
-Now consider the case where $R$ is the disjunction of two regular expressions
-$R_{1}$ and $R_{2}$. Inductively, we may build NFAs $N_{1}$ and $N_{2}$ that are
-strongly equivalent to $R_{1}$ and $R_{2}$, respectively. We define $N_{\oplus}$
-as a new NFA built as the disjunction of $N_{1}$ and $N_{2}$.
-$N_{\oplus}.\states$ comprises the states from $N_{1}$, the states from
-$N_{2}$, and a new marked initial state. The accepting states are precisely
-those that were accepting in the subautomata. The initial state has an
-$\varepsilon$-transition to each of the initial states of the subautomata. Once in
-$N_{1}.\init$, the downstream transitions are a copy of those in $N_{1}$,
-likewise for $N_{2}.\init$.  That is, $N_{\oplus}$ contains a copy of each
-$N_{1}$ and $N_{2}$, and we choose which copy to inhabit by deciding which
-$\varepsilon$-transition out of the initial state to take. A schematic
-representation of this automaton is given in \cref{fig:disjunctionnfa}.
-
-\begin{figure}
-  \begin{tikzpicture}[node distance = 17mm ]
-    \node[state, initial] (init) {$\init$};
-    \node[state, above right of=init] (init1) {$N_{1}.\init$};
-    \node[state, below right of=init] (init2) {$N_{2}.\init$};
-    \node[right of=init1] (dots1) {$\dots$};
-    \node[right of=init2] (dots2) {$\dots$};
-    \node[state, accepting, right of=dots1] (acc1) {$s_{1}$};
-    \node[state, accepting, right of=dots2] (acc2) {$s_{2}$};
-
-    \path[->] (init) edge[below] node{$\varepsilon$} (init1)
-              (init) edge[above] node{$\varepsilon$} (init2)
-              (init1) edge[above] node{} (dots1)
-              (init2) edge[above] node{} (dots2)
-              (dots1) edge[above] node{} (acc1)
-              (dots2) edge[above] node{} (acc2);
-  \end{tikzpicture}
-  \caption{NFA $N_{\oplus}$ as a disjunction of $N_{1}$ and $N_{2}$}
-  \label{fig:disjunctionnfa}
-\end{figure}
-
-Because $N_{1}$ and $N_{2}$ are each strongly equivalent to $R_{1}$ and $R_{2}$,
-it suffices to show that
-$Parse_{N_{1}} \oplus Parse_{N_{2}}$ is strongly
-equivalent to $Parse_{N_{\oplus}}$.
-
-Just like the case of a single character, when building the maps in either
-direction, we strengthen the induction hypothesis by mapping out of traces
-beginning with an arbitrary state then proceed by induction on traces.
-
-First, define the
-mapping out of $\Trace_{N_{\oplus}}(s)$. If $s$ is a state from $N_{1}$ then
-send the trace to a trace in $N_{1}$ beginning at $s$. Likewise, if $s$ is from
-$N_{2}$, then we map to a trace in $N_{2}$ beginning at $s$. If $s$ is
-$N_{\oplus}.\init$, then we map the trace to a disjunction of traces
-$\Trace_{N_{1}}(N_{1}.\init) \oplus \Trace_{N_{2}}(N_{2}.\init)$.
-
-In the other direction, we need to map a disjunction of traces through $N_{1}$
-or $N_{2}$ to a trace in $N_{\oplus}$. Without loss of generality, assume that
-the trace came from $N_{1}$. That is, assume we currently have a term of type
-$\Trace_{{N_{1}}}(N_{1}.\init)$ and we aim to construct a term of type
-$\Trace_{N_{\oplus}}(N_{\oplus}.\init)$. We induct on the trace $N$ trace an map
-it over to a trace beginning at the copy of $N_{1}.\init$ in $N_{\oplus}$, and
-then we conclude by using the $\epscons$ constructor to turn this trace into one
-beginning at $N_{\oplus}.\init$.
-
-We show that these maps mutually invert each other by the same sort of argument
-we used with the single character NFA.\ That is, we prove that the each map is a
-homomorphism of algebras and then appeal to the initiality of each of
-$\Trace_{N_{1}}$, $\Trace_{N_{2}}$, and $\Trace_{N_{\oplus}}$ to prove that the
-relevant maps compose to the identity.
-
-In this sketch, we elide the proof that these maps do indeed form homomorphisms.
-Further, we are eliding the needed NFAs for the cases where the regular
-expression is $I$, $\bot$, a Kleene star, or a concatenation. All of these
-elided details may be found in the Agda formalization.
+We use a variant of Thompson's construction
+\cite{thompsonProgrammingTechniquesRegular1968}, showing that NFAs are, up to strong equivalence, closed under each type operation for regular expressions.
+
+%% In this paper we will walk through
+%% some of the cases of this proof. The remaining cases
+%% follow
+%% similar reasoning and the details may be found in the formalized artifact.
+
+%% \begin{figure}
+%%   \begin{tikzpicture}[node distance = 17mm ]
+%%     \node[state, initial] (0) {$\0$};
+%%     \node[state, right of=0, accepting] (1) {$\1$};
+
+%%     \path[->] (0) edge[above] node{$\stringquote{c}$} (1);
+%%   \end{tikzpicture}
+%%   \caption{NFA $\N_{\c}$ for a single character $\c$}
+%%   \label{fig:litnfa}
+%% \end{figure}
+
+%% First handle the case where $R$ is a literal character $\literal c$. The NFA $N_{c}$ pictured in
+%% \cref{fig:litnfa} is strongly equivalent to the grammar $c$. To demonstrate this
+%% strong equivalence, we build functions between $\literal c$ and $Parse_{N_{c}}$
+%% and show they are mutually inverse.
+
+%% \begin{figure}
+%% \begin{lstlisting}
+%% A : (s : Fin 2) (*@$\to$@*) L
+%% A 0 = 'c'
+%% A 1 = I
+
+%% toN(*@$_c$@*) : (s : Fin 2) (*@$\to$@*) (*@$\uparrow$@*)(A s (*@$\lto$@*) Trace(*@$_{N_c}$@*) s true)
+%% toN(*@$_c$@*) 0 c = cons t true c (nil 1)
+%% toN(*@$_c$@*) 1 () = nil 1
+
+%% fromN(*@$_c$@*) : (s : Fin 2) (*@$\to$@*) (*@$\uparrow$@*)(Trace(*@$_{N_c}$@*) s true (*@$\lto$@*) A s)
+%% fromN(*@$_c$@*) 0 (cons t .true c tr) = let a (*@$\otimes$@*) () = c (*@$\otimes$@*) fromN(*@$_c$@*) 1 tr in a
+%% fromN(*@$_c$@*) 1 (nil .1) = ()
+%% \end{lstlisting}
+%% \caption{Parse transformer out of $Trace_{N_{c}}~s~true$}
+%% \label{fig:litNFAterms}
+%% \end{figure}
+
+%% In \cref{fig:litNFAterms} we define a family of grammars $A$ that serve as
+%% interpretations for each state in $N_{c}$. We then prove that each $A~s$ is
+%% equivalent to the accepting traces through $N_{c}$ out of state $s$.
+
+%% When
+%% defining $fromN_{c}$ we pattern match on the accepting traces beginning in each
+%% state. Out of state $0$ the only accepting traces are built via $cons$ over the
+%% single transition $t$, and out
+%% of state $1$ the only accepting trace is built via $nil$.
+
+%% Each $A~s$ is unambiguous, so $(fromN_{c}~s) \circ (toN_{c}~s)$ is equal to the
+%% identity on each $A~s$. To complete the proof of strong equivalence it remains
+%% to show that $(toN_{c}~s) \circ (fromN_{c}~s)$ is equal to the identity on
+%% $Trace~s~true$, which we show by using the equalizer type.
+
+%% \begin{figure}
+%% \begin{lstlisting}
+%% eqHom(*@$_{N_c}$@*) : (s : N(*@$_c$@*).state) (*@$\to$@*)
+%%       (*@$\uparrow$@*)(Trace(*@$_{N_c}$@*) s true (*@$\lto$@*) {tr | (toN(*@$_c$@*) s (*@$\circ$@*) fromN(*@$_{c}$@*) s) tr = id tr})
+%% \end{lstlisting}
+%% \caption{Signature for the equalizer term}
+%% \label{fig:litequalizer}
+%% \end{figure}
+
+%% The type signature for the family of functions, $eqHom_{N_{c}}$, into the equalizer type is given in
+%% \cref{fig:litequalizer}. By defining an $Trace_{N_{c}}$-algebra homomorphism
+%% into the equalizer, we enable a
+%% proof strategy for equality by induction on traces through $N_{c}$. First consider the traces
+%% out of state $1$. The only case to handle here is $nil 1$, and by definition
+%% $toN_{c}~1$ and $fromN_{c}~1$ invert each other.
+%% When defining $eqHom_{N_{c}}$ for traces out of state $0$, we must leverage an
+%% inductive hypothesis,
+
+%% \begin{align*}
+%%   (toN_{c}~0 \circ fromN_{c}~0) (cons~t~true~c~tr)
+%%   & \equiv (toN_{c}~0) (let~a~\otimes~()~=~c~\otimes~fromN_{c}~1~tr~in~a) \\
+%%   & \equiv cons~t~true~(let~a~\otimes~()~=~c~\otimes~fromN_{c}~1~tr~in~a)~(nil~1) \\
+%%   & \equiv cons~t~true~c~(nil~1) \\
+%%   & \equiv cons~t~true~c~tr \\
+%% \end{align*}
+
+%% The first two equalities are by defintion. The third equality holds because
+%% $\literal c$ is unambiguous and
+%% $let~a~\otimes~()~=~c~\otimes~fromN_{c}~1~tr~in~a$ and $c$ have type
+%% $\literal c$. The final equality holds because our inductive hypothesis states
+%% that $nil~1 \equiv tr$. Thus, each $A s$ and $Trace_{N_{c}}~s~true$ are strongly
+%% equivalent --- in particular, $\literal c$ and $Parse_{N_{c}}$.
+
+%% \steven{Rewrite the disjunction case with equalizers}
+%% Now consider the case where $R$ is the disjunction of two regular expressions
+%% $R_{1}$ and $R_{2}$. Inductively, we may build NFAs $N_{1}$ and $N_{2}$ that are
+%% strongly equivalent to $R_{1}$ and $R_{2}$, respectively. We define $N_{\oplus}$
+%% as a new NFA built as the disjunction of $N_{1}$ and $N_{2}$.
+%% $N_{\oplus}.\states$ comprises the states from $N_{1}$, the states from
+%% $N_{2}$, and a new marked initial state. The accepting states are precisely
+%% those that were accepting in the subautomata. The initial state has an
+%% $\varepsilon$-transition to each of the initial states of the subautomata. Once in
+%% $N_{1}.\init$, the downstream transitions are a copy of those in $N_{1}$,
+%% likewise for $N_{2}.\init$.  That is, $N_{\oplus}$ contains a copy of each
+%% $N_{1}$ and $N_{2}$, and we choose which copy to inhabit by deciding which
+%% $\varepsilon$-transition out of the initial state to take. A schematic
+%% representation of this automaton is given in \cref{fig:disjunctionnfa}.
+
+%% \begin{figure}
+%%   \begin{tikzpicture}[node distance = 17mm ]
+%%     \node[state, initial] (init) {$\init$};
+%%     \node[state, above right of=init] (init1) {$N_{1}.\init$};
+%%     \node[state, below right of=init] (init2) {$N_{2}.\init$};
+%%     \node[right of=init1] (dots1) {$\dots$};
+%%     \node[right of=init2] (dots2) {$\dots$};
+%%     \node[state, accepting, right of=dots1] (acc1) {$s_{1}$};
+%%     \node[state, accepting, right of=dots2] (acc2) {$s_{2}$};
+
+%%     \path[->] (init) edge[below] node{$\varepsilon$} (init1)
+%%               (init) edge[above] node{$\varepsilon$} (init2)
+%%               (init1) edge[above] node{} (dots1)
+%%               (init2) edge[above] node{} (dots2)
+%%               (dots1) edge[above] node{} (acc1)
+%%               (dots2) edge[above] node{} (acc2);
+%%   \end{tikzpicture}
+%%   \caption{NFA $N_{\oplus}$ as a disjunction of $N_{1}$ and $N_{2}$}
+%%   \label{fig:disjunctionnfa}
+%% \end{figure}
+
+%% Because $N_{1}$ and $N_{2}$ are each strongly equivalent to $R_{1}$ and $R_{2}$,
+%% it suffices to show that
+%% $Parse_{N_{1}} \oplus Parse_{N_{2}}$ is strongly
+%% equivalent to $Parse_{N_{\oplus}}$.
+
+%% Just like the case of a single character, when building the maps in either
+%% direction, we strengthen the induction hypothesis by mapping out of traces
+%% beginning with an arbitrary state then proceed by induction on traces.
+
+%% First, define the
+%% mapping out of $\Trace_{N_{\oplus}}(s)$. If $s$ is a state from $N_{1}$ then
+%% send the trace to a trace in $N_{1}$ beginning at $s$. Likewise, if $s$ is from
+%% $N_{2}$, then we map to a trace in $N_{2}$ beginning at $s$. If $s$ is
+%% $N_{\oplus}.\init$, then we map the trace to a disjunction of traces
+%% $\Trace_{N_{1}}(N_{1}.\init) \oplus \Trace_{N_{2}}(N_{2}.\init)$.
+
+%% In the other direction, we need to map a disjunction of traces through $N_{1}$
+%% or $N_{2}$ to a trace in $N_{\oplus}$. Without loss of generality, assume that
+%% the trace came from $N_{1}$. That is, assume we currently have a term of type
+%% $\Trace_{{N_{1}}}(N_{1}.\init)$ and we aim to construct a term of type
+%% $\Trace_{N_{\oplus}}(N_{\oplus}.\init)$. We induct on the trace $N$ trace an map
+%% it over to a trace beginning at the copy of $N_{1}.\init$ in $N_{\oplus}$, and
+%% then we conclude by using the $\epscons$ constructor to turn this trace into one
+%% beginning at $N_{\oplus}.\init$.
+
+%% We show that these maps mutually invert each other by the same sort of argument
+%% we used with the single character NFA.\ That is, we prove that the each map is a
+%% homomorphism of algebras and then appeal to the initiality of each of
+%% $\Trace_{N_{1}}$, $\Trace_{N_{2}}$, and $\Trace_{N_{\oplus}}$ to prove that the
+%% relevant maps compose to the identity.
+
+%% In this sketch, we elide the proof that these maps do indeed form homomorphisms.
+%% Further, we are eliding the needed NFAs for the cases where the regular
+%% expression is $I$, $0$, a Kleene star, or a concatenation. All of these
+%% elided details may be found in the Agda formalization.
 \end{proof}
 
-\subsection{The Dyck Grammar}
-%% 3. Dyck language
+\subsection{Context-free grammars}
+
+Next, we give two examples for parsing context-free
+grammars. Context-free grammars (CFG) can be encoded in our type
+theory in a similar way to regular expressions, as CFGs are equivalent
+to the formalism of \emph{$\mu$}-regular expressions, where the Kleene
+star is replaced by an arbitrary fixed point
+operation\cite{leis_towards_1992}.
+
 \newcommand{\balanced}{\mathtt{balanced}}
 \newcommand{\Dyck}{\mathtt{Dyck}}
-Moving up the Chomsky hierarchy from regular grammars, we arrive at context-free
-grammars (CFGs). An interesting nontrivial CFG is the Dyck grammar
-$\Dyck$ --- the grammar of balanced parentheses.
-
-$\Dyck$ is a grammar over the alphabet $\{ \stringquote{(}, \stringquote{)} \}$
-and is defined in \cref{fig:dyckinductive}. The $\nil$ constructor shows that
+A simple example of a CFG is the Dyck grammar of balanced
+parentheses, which we define in \Cref{fig:dyckinductive}
+$\Dyck$ is a grammar over the alphabet $\{ \stringquote{(}, \stringquote{)} \}$. The $\nil$ constructor shows that
 the empty string is balanced, and the $bal$ constructor builds a balanced
@@
 \end{lstlisting}
-\caption{The Dyck grammar as an inductive linear type}
-\label{fig:dyckinductive}
 \end{figure}
 
-\Cref{fig:DyckAutomaton} presents a sketch of the automaton $M$. The traditional
-approach to parse grammars like $\Dyck$ is to employ a stack to count how many
-open parentheses have been seen so far. We may think of
-$M$ as performing the same task as a stack machine through slightly different means. Instead of storing
-the number of open parentheses in a distinguished stack of memory, $M$ operates
-by storing this data in the current state. $M$ is defined almost identically to how we defined DFAs, except we omit the
-finiteness condition on the states of the automaton. This similarity allows
-functions $parse_{M}$ and $print_{M}$ to be defined analogously to
-\cref{fig:dfaparser,fig:printdfa}, and so
-$\LinSigTy {s} {M.states} {\LinSigTy {b}{Bool} {Trace_{M}~s~b}}$ is likewise unambiguous.
+The Dyck language is an example of an $\textrm{LL}(0)$ language, one
+that can be parsed top-down with no
+lookahead\cite{rosenkrantz_properties_1970}. This means we can
+implement it simply as an \emph{infinite} state deterministic
+automaton, in \Cref{fig:DyckAutomaton}. Here the state is a ``stack''
+counting how many open parentheses have been seen so far. Functions
+$parse_{M}$ and $print_{M}$ for this automaton can be defined
+analogously to the parser and printer for DFAs, and so $\LinSigTy {s}
+{M.states} {\LinSigTy {b}{Bool} {Trace_{M}~s~b}}$ is likewise
+unambiguous.
 
@@
   \label{thm:dyck}
-  $Dyck$ and $Parse_{M}$ are strongly equivalent.
+  $Dyck$ and $Parse_{M}$ are strongly equivalent. And therefore we can construct a parser for $\Dyck$.
 \end{theorem}
 
-We omit the details of \cref{thm:dyck} in this paper, and the ful proof is
-mechanized in our Agda development. We can build a $Dyck$ parser by utilizing
-this artifact, and the
-runtime of the $Dyck$ parser is about 30 seconds for input that is 25,000
-characters long.
-\pedro{It might be a good idea to mention explicitly in the introduction
-  that we are not claiming that our verified parsers are competitive with
-  the state of the art.}
+%% We can build a $Dyck$ parser by utilizing
+%% this artifact, and the
+%% runtime of the $Dyck$ parser is about 30 seconds for input that is 25,000
+%% characters long.
+%% \pedro{It might be a good idea to mention explicitly in the introduction
+%%   that we are not claiming that our verified parsers are competitive with
+%%   the state of the art.}
 
@@
 
+Our final example is of a simple grammar of arithmetic expressions
+with an associative operation. Here we take the alphabet to be $\{
+\stringquote{(}, \stringquote{)}, \stringquote{+} , \stringquote{NUM}
+\}$. In \Cref{fig:binop-inductive} we define it using two mutually
+recursive types, corresponding to the two non-terminals we would use
+in a CFG syntax. The syntactic structure encodes that the binary
+operation is right associative. In the same figure, we define the
+traces of an automaton with one token of lookahead. The automaton has
+four different ``states'', each with access to a natural number
+``stack''. The ``opening'' state $O$ expects either an open paren, in
+which case it increments the stack and stays in the opening state, or
+sees a number and proceeds to the $D$ state. The ``done opening''
+state $D$ is where lookahead is used: if the next token will be a
+right paren, then we proceed to $C$, otherwise we proceed to $M$. Here
+$\mathsf{NotStartsWithRP}$ is defined as
+$\I \oplus (({\literal{(}} \oplus {\literal {+}}
+\oplus \mathsf{NUM}) \otimes \top)$. In the ``closing'' state $C$ if
+we observe a close paren we decrement the count and continue to the
+$D$ state. In the ``multiplying'' state $M$, we succeed if the string
+ends and the count is $0$, and if we see a plus we continue to the $O$
+state. Additionally, since the automaton need also parse all of the
+incorrect strings, we add additional failing cases. It is
+straightforward to implement a parser for this lookahead automaton,
+generalizing the approach for deterministic automata.
+
+\begin{theorem}[\agdalogo]
+  We construct a parser for $\mathsf{Exp}$ by showing it is weakly
+  equivalent to $\mathsf{O}~0~\true$.
+\end{theorem}
+\begin{figure}
+\begin{lstlisting}
+data Exp : L where
+  done : (*@$\uparrow$@*)(Atom (*@$\lto$@*) Exp)
+  add : (*@$\uparrow$@*)(Atom (*@$\lto$@*) '+' (*@$\lto$@*) Exp (*@$\lto$@*) Exp)
+data Atom : L where
+  num : (*@$\uparrow$@*)(NUM (*@$\lto$@*) Atom)
+  parens : (*@$\uparrow$@*)('(' (*@$\lto$@*) Exp (*@$\lto$@*) ')' (*@$\lto$@*) Atom)
+
+data O : Nat (*@$\to$@*) Bool (*@$\to$@*) L where
+  left : (*@$\uparrow$@*)(&[ n ] &[ b ] '(' (*@$\lto$@*) O (n + 1) b (*@$\lto$@*) O n b)
+  num : (*@$\uparrow$@*)(&[ n ] &[ b ] NUM (*@$\lto$@*) DO n b (*@$\lto$@*) O n b)
+  done : (*@$\uparrow$@*)(&[ n ] O n false)
+  unexpected : (*@$\uparrow$@*)(&[ n ] (')' (*@$\oplus$@*) '+') (*@$\lto$@*) (*@$\top$@*) (*@$\lto$@*) O n false)
+data D : Nat (*@$\to$@*) Bool (*@$\to$@*) L where
+  lookAheadRP  : (*@$\uparrow$@*)(&[ n ] &[ b ] ((')' (*@$\otimes$@*) (*@$\top$@*)) (*@$\&$@*) C n b) (*@$\lto$@*) D n b)
+  lookAheadNot  : (*@$\uparrow$@*)(&[ n ] &[ b ] (NotStartsWithRP (*@$\&$@*) M n b) (*@$\lto$@*) D n b)
+data C : Nat (*@$\to$@*) Bool (*@$\to$@*) L where
+  closeGood : (*@$\uparrow$@*)(&[ n ] &[ b ] ')' (*@$\lto$@*) D n b (*@$\lto$@*) C (n + 1) b)
+  closeBad : (*@$\uparrow$@*)(&[ n ] ')' (*@$\lto$@*) C 0 b)
+  done : (*@$\uparrow$@*)(&[ n ] C n false)
+  unexpected : (*@$\uparrow$@*)(&[ n ] ('(' (*@$\oplus$@*) '+' (*@$\oplus$@*) NUM) (*@$\lto$@*) (*@$\top$@*) (*@$\lto$@*) C n false)
+data M : Nat (*@$\to$@*) Bool (*@$\to$@*) L where
+  doneGood : (*@$\uparrow$@*)(M 0 true)
+  doneBad : (*@$\uparrow$@*)(&[ n ] M (n + 1) false)
+  add : (*@$\uparrow$@*)(&[ n ] &[ b ] '(' (*@$\lto$@*) O n b (*@$\lto$@*) M n b)
+  unexpected : (*@$\uparrow$@*)(&[ n ] ('(' (*@$\oplus$@*) ')'(*@$\oplus$@*) NUM) (*@$\lto$@*) (*@$\top$@*) (*@$\lto$@*) M n false)
+\end{lstlisting}
+\caption{Associative arithmetic expressions and a corresponding lookahead automaton}
+\label{fig:binop-inductive}
+\end{figure}
 % \subsection{LL(1)}
@@
 \newcommand{\stringNL}{String'}
-\steven{Need to better introduce nonlinear type of strings. We haven't talked
-  about semantics nor have we included a nonlinear type of strings in the syntax}
-Define $\stringNL$ to be the non-linear list type over the alphabet $\Sigma$ and
-let $P : \stringNL \to U$ be a mapping from the set of strings to the universe of
-nonlinear types. The function $P$ describes a family of nonlinear types
-dependent on $\stringNL$.
-We view the image of $P$ as a set of parse trees for each input string, and thus
-$P$ carries the structure of a formal grammar. However, even though $P$ describes this grammatical data, it is not
-a first-class linear type in \theoryabbv. To address this gap, below we define a linear type that captures the same
-semantic grammar as $P$,
-
+
+While we have shown only examples for context-free grammars, in fact
+arbitrarily complex grammars are encodable in \theoryabbv. To
+demonstrate this, we show that for any \emph{non-linear} function $P :
+\stringNL \to U$, where here $\stringNL$ is the \emph{non-linear} type
+of strings over the alphabet, we can construct a grammar whose parses
+correspond to $P$.
+%% Define $\stringNL$ to be the non-linear list type over the alphabet $\Sigma$ and
+%% let  be a mapping from the set of strings to the universe of
+%% nonlinear types. The function $P$ describes a family of nonlinear types
+%% dependent on $\stringNL$.
+%% We view the image of $P$ as a set of parse trees for each input string, and thus
+%% $P$ carries the structure of a formal grammar. However, even though $P$ describes this grammatical data, it is not
+%% a first-class linear type in \theoryabbv. 
 \[
@@
 \]
-
-\steven{Probably put this ceiling business around the discussion of the linear
-  type of strings. Is this in section 3?}
-
+%% \steven{Probably put this ceiling business around the discussion of the linear
+%%   type of strings. Is this in section 3?}
 where $\lceil \stringquote{} \rceil = I$ and
@@
 as it allows to sidestep our linear typing connectives and utilize the whole of nonlinear
-dependent type theory to define a grammar. This power allows us to encode
-arbitrary grammars. In particular, given a
+dependent type theory to define a grammar. For example, given a
 Turing machine $Tur$ one may
-define a predicate $accepts : \stringNL \to U$ such that $accept~w$ is equal to $\top$ if
-$T$ accepts $w$ and equal to $\bot$ otherwise. Then, $Reify~accepts$ is a linear
+define a non-linear predicate $accepts : \stringNL \to U$ such that $accept~w$ is equal to $\top$ if
+$T$ accepts $w$ and equal to $0$ otherwise. Then, $Reify~accepts$ is a linear
 type that captures precisely the strings accepted by $Tur$. That is,
@@
 languages in the Chomsky hierarchy.
-
-\subsection{Semantic Actions}
-So far, each construction has been primarily concerned with the parsing of a
-string into a \emph{concrete} syntax tree. However, in a larger verified
-developement --- such as a compiler --- a parser will often emit an
-\emph{abstract} syntax tree that omits superflous syntactic details that aren't
-needed for later passes.  The procedure that takes in a
-concrete syntax tree and returns an abstract one is referred to as a
-\emph{semantic action}.
-
-Given a linear type $A$ and a nonlinear type $X$, we define a semantic action to
-be a function $\ltonl {(A \lto \Delta X)}$, where
-$\Delta X = \LinSigTy {x} {X} {\top}$. That is, a semantic action is an
-elaboration from the concrete parses of $A$ into the semantic domain $X$. As
-written, these actions are quite general. As discussed above, we may define
-folds that are still grammatical in nature --- such as folding over a concrete
-parse tree and forgetting parentheses that aren't needed in later compiler
-passes. We may also define arbitrary computation over the structure of the parse
-tree, such as calculating the maximum nesting depth of parentheses for $Dyck$,
-returning the length the input string, or
-even an interpreter that evaluates arithmetic expressions at parse-time.
-
-\steven{What can I say about what we have in the Agda development here?}
-\pedro{I'm assuming that this section is still incomplete, so I'll wait
-for it to be complete in order to make more detailed comments.}
-
-\section{Denotational Semantics}
+\begin{theorem}[\Agda]
+  For any Turing machine, we can construct a grammar in \theoryabbv
+  that accepts the same language as the Turing machine.
+\end{theorem}
+
+\section{Denotational Semantics and Implementation}
 \label{sec:semantics-and-metatheory}
@@
 grammars and parse transformers, we will now define a
-\emph{denotational semantics} that makes this mathematically
-precise. To interpret our types of codes we assume a universe of
-\emph{small} sets.
+\emph{denotational semantics} that makes this mathematically precise
+by defining a notion of formal grammar and parse transformer then
+showing that our type theory can be soundly interpreted in this
+model. We then discuss how this denotational semantics provides the
+basis for our prototype implementation in Agda.
 
@@
 
-First, we need to establish a precise definition of a formal grammar
-that accounts for the possibility of ambiguity. The most common
-definition of a formal grammar is as generative grammars, defined by
-a set of non-terminals, with a specified start symbol and set of
-production rules. We instead use a more abstract formulation that is
-closer in spirit to the standard definition of a formal language:
-
+The most common definition of a formal grammar is as generative
+grammars, defined by a set of non-terminals, a specified start symbol
+and set of production rules. We instead use a more abstract
+formulation that is closer in spirit to the standard definition of a
+formal \emph{language}\cite{elliottSymbolicAutomaticDifferentiation2021}:
 \begin{definition}
   A \emph{formal language} $L$ is a function from strings to propositions.
-  A \emph{formal grammar} $A$ is a function from strings to small sets.
+  A (small) \emph{formal grammar} $A$ is a function from strings to (small) sets.
 \end{definition}
-
 We think of the grammar $A$ as taking a string to the set of all parse
@@
 tree, but \emph{which} parse tree it has, i.e., what the structure of
-the element of $A(w)$ is.
-
-The linear types of \theoryabbv can all be interpreted as formal
-grammars, but what about the linear terms? These can be interpreted as
-\emph{parse transformers}, which we now define.
+the element of $A(w)$ is.  To interpret our universes $U , L$ we
+assume we have a universe of \emph{small} sets. In the remainder, all
+formal grammars are assumed to be small.
+
+We then interpret linear \emph{terms} as \emph{parse transformers}:
 \begin{definition}
-  Let $L_1,L_2$ be formal languages. Then we say $L_1\subseteq L_2$ if
-  for every string $w$, $L_1(w)\Rightarrow L_2(w)$.
-
   Let $A_1,A_2$ be formal grammars. Then a parse transformer $f$ from
@@
 \end{definition}
-\pedro{Do we actually need a definition of formal language morphism?
-In case we do, it should probably be its own definition.}
-First, note that parse transformers generalize language inclusion: if
-$A_1(w), A_2(w)$ are all subsets of a one-element set, then a parse
-transformer is equivalent to showing that $A_1(w) \subseteq A_2(w)$.
-In our denotational semantics, linear terms will be interpreted as
-such parse transformers, and the notions of unambiguous grammar,
-parseability, etc, introduced in \Cref{sec:concepts} can be verified
-to correspond to their intended meanings under this interpretation.
+Just as formal grammars generalize formal languages, parse
+transformers generalize formal language inclusion: if $A_1(w), A_2(w)$
+are all subsets of a one-element set, then a parse transformer is
+equivalent to showing that $A_1(w) \subseteq A_2(w)$. In our
+denotational semantics, linear terms will be interpreted as such parse
+transformers, and the notions of unambiguous grammar, parsers,
+disjointness, etc, introduced in \Cref{sec:applications} can be
+verified to correspond to their intended meanings under this
+interpretation.
 
@@
 their composition is defined pointwise, i.e. $(f\circ g)_w = f_w \circ g_w$.
-Furthemore, given a formal grammar $A$, its identity transformer is $id_w =
+Furthermore, given a formal grammar $A$, its identity transformer is $id_w =
 id_{A(w)}$, where $id_{A(w)}$ is the identity function on the set $A(w)$. This
@@
 structures to model the linear types, terms and equalities in
-\theoryabbv.
+\theoryabbv. These categorical properties are precisely what is
+required to interpret all of the linear type and term constructors.
 %% \begin{theorem}
@@
 
-We can now define our denotational semantics.
+We now define our denotational semantics.
 \begin{definition}[Grammar Semantics]
@@
   \item If $\Gamma \vdash X \isSmall$, then $\sem X \gamma$ is a small set.
-  \item If $\Gamma \vdash X \cong X'$ then for every $\gamma$, $\sem{X}\gamma = \sem{X'}\gamma$
-  \item If $\Gamma \vdash A \cong A'$ then for every $\gamma$, $\sem{A}\gamma = \sem{A'}\gamma$
-  \item If $\Gamma \vdash M \cong M' : X$ then for every $\gamma$, $\sem{M}\gamma = \sem{M'}\gamma$
-  \item If $\Gamma;\Delta \vdash e \cong e' : A$ then for every $\gamma$, $\sem{e}\gamma = \sem{e'}\gamma$
+  \item If $\Gamma \vdash X \equiv X'$ then for every $\gamma$, $\sem{X}\gamma = \sem{X'}\gamma$.
+  \item If $\Gamma \vdash A \equiv A'$ then for every $\gamma$, $\sem{A}\gamma = \sem{A'}\gamma$.
+  \item If $\Gamma \vdash M \equiv M' : X$ then for every $\gamma$, $\sem{M}\gamma = \sem{M'}\gamma$.
+  \item If $\Gamma;\Delta \vdash e \equiv e' : A$ then for every $\gamma$, $\sem{e}\gamma = \sem{e'}\gamma$.
   \end{enumerate}
@@
 
-For reasons of brevity, we present only the cases related to the
-linear types in \Cref{fig:semantics}. The full details are in the
-appendix.\max{TODO: put the full details in the appendix}.  The
-grammar for a literal $c$ has a single parse precisely when the input
-string consists of the single character. The grammar for the unit
-similarly has a single parse for the empty string.  A parse of the
-tensor product $A \otimes B$ consists of a \emph{splitting} of the
-empty string into a prefix $w_1$ and suffix $w_2$ along with an $A$
-parse of $w_1$ and $B$ parse of $w_2$. A parse of $\bigoplus_{x:X} A$
-is a pair of an element of the set $X$ and a parse of $A(x)$, while
-dually a parse of $\bigwith_{x:X} A$ is a \emph{function} taking any
-$x:X$ to a parse of $A(x)$. A $w$-parse of $A \lto B$ is a function
-that takes an $A$ parse of some other string $w'$ to a $B$ parse of
-$ww'$, and $B \tol A$ is the same except the $B$ parse is for the
-reversed concatenation $w'w$.
+The interpretation of dependent types as sets is standard
+\cite{Hofmann_1997}. We present the concrete descriptions of the
+semantics of linear types, as well as our non-standard non-linear
+types in \Cref{fig:semantics}. The grammar for a literal $c$ has a
+single parse precisely when the input string consists of the single
+character. The grammar for the unit similarly has a single parse for
+the empty string.  A parse of the tensor product $A \otimes B$
+consists of a \emph{splitting} of the empty string into a prefix $w_1$
+and suffix $w_2$ along with an $A$ parse of $w_1$ and $B$ parse of
+$w_2$. A parse of $\bigoplus_{x:X} A$ is a pair of an element of the
+set $X$ and a parse of $A(x)$, while dually a parse of $\bigwith_{x:X}
+A$ is a \emph{function} taking any $x:X$ to a parse of $A(x)$. A
+$w$-parse of $A \lto B$ is a function that takes an $A$ parse of some
+other string $w'$ to a $B$ parse of $ww'$, and $B \tol A$ is the same
+except the $B$ parse is for the reversed concatenation $w'w$.
 %
@@
 \[ \sem{\uparrow (A \lto B)}\gamma = \sem{A \lto B}\gamma\,\varepsilon = \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,w'\]
-Next, a parse in the equalizer $\equalizer{a : A}{f}{g}$ is defined as
-a parse in $\sem{A}$ that is mapped to the same parse by $f$ and
-$g$. Recall that $\sem{f}\gamma$ and $\sem{g}$ are elements of
-$\sem{\uparrow(A \lto B)}$, i.e., parse transformers.
-
+Next, a parse in the equalizer $\equalizer{a}{f}{g}$ is defined as
+a parse in $\sem{A}$ that is mapped to the same parse by the parse transformers $\sem{f}$ and
+$\sem{g}$.
+%
 The universe $L$ of linear types is interpreted as the set of all
-(small) grammars.
-
+small grammars.
 \begin{figure}
+  \begin{footnotesize}
     \begin{align*}
@@
     \sem{\uparrow A}\gamma &= \sem{A}\gamma\,\varepsilon\\
-    \sem{\equalizer{a:A}{f}{g}}\gamma\,w &= \{ a \in \sem{A}\gamma\,w \pipe \sem{f}\gamma\,w\,a = \sem{g}\gamma\,w\,a \} \\
+    \sem{\equalizer{a}{f}{g}}\gamma\,w &= \{ a \in \sem{A}\gamma\,w \pipe \sem{f}\gamma\,w\,a = \sem{g}\gamma\,w\,a \} \\
     \sem{L}\gamma &= \Grammar_0\\
@@
     \sem{\map(F)}\gamma\,f &= \sem{F}\gamma\,f\\
-    \sem{\mu A}\gamma &= \mu (\sem{A}\gamma) \tag{\Cref{thm:indexed-container}}\\
+    \sem{\mu A}\gamma &= \mu (\sem{A}\gamma)\\
   \end{align*}
+  \end{footnotesize}
   \caption{Grammar Semantics}
+  \label{fig:semantics}
 \end{figure}
@@
 functor} on the category of sets, also sometimes called an
-\emph{indexed container} \cite{dep-poly,indexed-container}.
+\emph{indexed container} \cite{gambino_wellfounded_2004,altenkirch_indexed_2015}.
 \begin{definition}
@@
   from $I$ to $O$ consists of a set of shapes $S$, a set of positions
-  $P$ and functions $I \xleftarrow f P \xrightarrow g S  \xrightarrow h O$. The \emph{extension}
+  $P$ and functions $f : P \to I$, $g : P \to S$ and $h : S \to O$. The \emph{extension}
   of a polynomial is a functor $\Set/I \to \Set/O$ defined as the
@@
   \end{tikzcd}\]
-  Where $f^*$ is the pullback functor along $f$; $\Pi_g$ is the depedent
+  Where $f^*$ is the pullback functor along $f$; $\Pi_g$ is the dependent
   product operation and $\Sigma_h$ is the dependent sum operation, which
   are, respectively, the right and left adjoint of their pullback functors
-  $g^*$ and $h^*$.
-
-  A dependent polynomial functor from $I$ to $O$ is a functor $\Set/I
-  \to \Set/O$ that is nautrally isomorphic to the extension of a
+  $g^*$ and $h^*$. A dependent polynomial functor from $I$ to $O$ is a functor $\Set/I
+  \to \Set/O$ that is naturally isomorphic to the extension of a
   dependent polynomial from $I$ to $O$.
 \end{definition}
-With this interpretation, $\el(F)$ and $\map(F)$ are just interpreted
-as the action of the functor on objects and morphisms,
-respectively. We interpret the constructors $\mathsf{K},\Var$, etc. on
-functors in the obvious way that matches the definitional behavior of
-$\el$ and $\map$. The non-trivial part of the construction is
-verifying that such functors are constructing corresponding dependent
-polynomials. The details are tedious but straightforward extension of
-prior work on dependent polynomials and indexed containers and we have
-verified the construction in Agda.
-
-The reason we use dependent polynomials functors on sets is that it is
-well-known that their initial algebras exist
-\cite{initial-algebras-of-dep-poly}. Additionally, initial algebras
-for dependent polynomials are straightforwardly constructed in Agda as
-an inductive type of ``IW'' trees, which is available as a
-construction in the cubical library of Agda \cite{iw,cubical-library}.
-\begin{theorem}
-  Every dependent polynomial functor from $I$ to $I$ has an initial
-  algebra.
-\end{theorem}
-Then an element $F \in \sem{X \to \SPF\,X}\gamma$ is an
-$\sem{X}\gamma$-indexed family of polynomial functors from
-$\sem{X}\gamma\times \Sigma^*$ to $\Sigma^*$, and taking the product
-of these constructs a polynomial functor from $\sem{X}\gamma\times
-\Sigma^*$ to itself. Then $\sem{\mu\,F}\gamma$ is defined to be the
-initial algebra of this functor, and the initial algebra structure is
-used to interpret $\roll, \fold$ and the corresponding axioms.
+With this interpretation of $F$ as a polynomial functor, $\el(F)$ and
+$\map(F)$ are just interpreted as the action of the functor on objects
+and morphisms, respectively. We interpret the constructors
+$\mathsf{K},\Var$, etc. on functors in the obvious way that matches
+the definitional behavior of $\el$ and $\map$. The non-trivial part of
+the construction is verifying that such such constructions are closed
+under being polynomial. The details are tedious but straightforward
+extension of prior work on dependent polynomials and indexed
+containers and we have verified the construction in Agda.
+
+We use dependent polynomials functors on sets as these are guaranteed
+to have initial algebras. Further, these initial algebras are readily
+constructed in our Agda implementation as an inductive type of $IW$
+trees which are already available in the cubical library of Agda
+\cite{The_Agda_Community_Cubical_Agda_Library_2024}.  Then an element $F \in \sem{X \to
+  \SPF\,X}\gamma$ is an $\sem{X}\gamma$-indexed family of polynomial
+functors from $\sem{X}\gamma\times \Sigma^*$ to $\Sigma^*$, and taking
+the product of these constructs a polynomial functor from
+$\sem{X}\gamma\times \Sigma^*$ to itself. Then $\sem{\mu\,F}\gamma$ is
+defined to be the initial algebra of this functor, and the initial
+algebra structure is used to interpret $\roll, \fold$ and the
+corresponding axioms.
+
+The remaining details of the interpretation of linear terms as parse
+transformers and verification of the equational axioms is a relatively
+straightforward extension of existing semantics of linear logic in
+monoidal categories, and is included in the appendix\cite{seely89}.
+
+\subsection{Agda Implementation}
+
+This denotational semantics in grammars and parse transformers serves
+as the basis for our prototype implementation of \theoryabbv in
+cubical Agda. The implementation is a shallow embedding, meaning that
+rather than formalizing a syntax of \theoryabbv types and terms, we
+work directly with Agda types and a definition of a formal grammar as
+a function $\String \to \Set$. Then we implement each of the type and
+term constructors of \theoryabbv as combinators on formal grammars or
+parse transformers, and use cubical Agda's equality type to model the
+term equalities. Cubical Agda is convenient for this purpose as it has
+built-in support for function extensionality which we use
+extensively. Axioms such as distributivity of $\oplus/\&$ and
+disjointness of constructors are then provable directly in Agda, and
+we are careful to only construct grammars, terms and proofs using
+constructs that are possible in \theoryabbv. The main difference
+between our shallow embedding and \theoryabbv is that our linear terms
+are written in a combinator-style, without being able to use named
+variables in linear terms. A benefit of this shallow embedding is that
+the parsers are immediately available to a larger Agda development, as
+they are just normal Agda code. In future work we will look to
+implement a type checker for a syntax closer to the presentation in
+this paper, but with the goal of still making it integrate with an
+existing proof assistant.
 
@@
 %%   c\lto c &\equiv I\\
-%%   c\lto d &\equiv \bot\\
-%%   c\lto I &\equiv \bot\\
-%%   c\lto 0 &\equiv \bot\\
+%%   c\lto d &\equiv 0\\
+%%   c\lto I &\equiv 0\\
+%%   c\lto 0 &\equiv 0\\
 %%   c\lto (A \otimes B) & \equiv (c\lto A) \otimes B + (A \amp I) \otimes (c\lto B)\\
@@
 %% \steven{Definable as nullary sum}
-%% The empty type $\bot$ has no inhabitants. The elimination for the empty type
-%% witnesses the principle of explosion --- i.e.\ from a term of type $\bot$ we may
+%% The empty type $0$ has no inhabitants. The elimination for the empty type
+%% witnesses the principle of explosion --- i.e.\ from a term of type $0$ we may
 %% introduce a term $\mathsf{absurd}_{A} : A$ for any type $A$.
 %% %
-%% \[ \sem {\bot} \gamma w = \emptyset \]
+%% \[ \sem {0} \gamma w = \emptyset \]
 %% %
@@
 \label{sec:discussion}
-\paragraph{Lambek Calculus and Categorial Grammar}
-In 1958 \cite{lambek1958mathematics}, Joachim Lambek introduced his syntactic
-calculus as a logical system for linguistic derivations. In fact, one may view
-the subtheory of our calculus generated over $\otimes, \lto, \tol$ to be a proof
-relevant presentation of Lambek's original system.
-
-\steven{Should discussion of this connection to Lambek calculus be spread
-  throughout the paper?}
-
-Lambek later explicitly connects his syntactic calculus to the structure of a
-non-commutative biclosed monoidal category \cite{lambek1988categorial}. There is
-a rich line of subsequent research after these revelations by Lambek, including
-the study of pregroup grammars
-\cite{coeckeMathematicalFoundationsCompositional2010}, the presentation of
-abstract categorial grammars \cite{degrooteAbstractCategorialParsing2015}, and
-even Lambek calculus with dependent types \cite{luoLambekCalculusDependent}. As
-far as the authors can tell, these works seem to be primarily interested in natural
-language semantics in the style of Montague grammars, rather than parsing formal
-languages. That is, they seem to try to give a functorial semantic
-interpretation of natural language.
-
-There seem to be some implementations from the abstract categorial grammar
-toolkit, but again they seem natural language focused and not up to the task of
-verified formal language parsing.
-
-Also mention here that \cite{luoLambekCalculusDependent} seems similar at first
-glance, but isn't as similar as it might look because the judgments and
-dependent types seem to
-take on a different interpretation of ours. Say something here like, ``as we
-understand it, it differs in x,y,z ways''
-
-\steven{TODO Edit the above}
-\steven{Even though the above sounds bad, are there Lambek/categorial things that are missed?}
+\paragraph{Grammars as (Linear) Types}
+Lambek's original syntactic calculus \cite{lambek58} describes a
+logical system for linguistic derivations, and it can be given
+semantics inside of a non-commutative biclosed monoidal category
+\cite{lambek1988categorial}. This led to many uses of non-commutative
+linear logic and lambda calculi in linguistics
+\cite{buszkowskiTypeLogicsGrammar2003} --- including mechanized
+categorial grammar parsers \cite{Guillaume2024,ranta-2011}. This style
+of grammar formalism has gone by the names Lambek calculus or
+\emph{categorial grammar}, and it is equal in expressivity to
+context-free grammars.  The existing works on categorial grammar are
+different in nature to our approach: they are based on non-commutative
+linear logic, but their terms do not include elimination rules, and so
+can only express parse trees, and so cannot be used to express
+verified parsers.
+
+The most similar prior work to our own is Luo's Lambek calculus with
+dependent types \cite{luo}. Their design differs from our own in
+allowing linear types to depend on other linear types and supporting
+directed $\Pi$ and $\Sigma$ types that have no analog in our
+system. They do not provide a semantics for these connectives, and it
+is unclear how to interpret their connectives in our grammar
+semantics. Further, we provide several examples showing that
+\theoryabbv is a practical system for describing grammar formalisms
+and parsers, and it is unclear if these could be implemented in their calculus.
+
+%% \subsubsection{Typed Grammars}
+The usage of a simple type system to reason about regular expressions up to weak
+equivalence was
+introduced by Frisch and Cardelli \cite{frischCardelli}. Henglein and Nielsen
+\cite{henglein_regular_2011} later adapt this approach by providing a type
+system for regular expressions whose semantics were in one-to-one correspondence
+with the set of parse trees.
+Elliott \cite{elliottSymbolicAutomaticDifferentiation2021} gives a similar
+semantic interpretation of regular grammars as type-level predicates on strings.
+The type system of \theoryabbv extends these semantic interpretations of to a broader class of grammars and also gives a formal syntax and equational theory for the parse transformers that these prior works lack.
+
+%% Krishnaswami and Yallop \cite{krishnaswami_typed_2019} provide a type system for
+%% demonstrating that carries proves of the LL(1) condition for context-free
+%% grammars. Despite being a type system over formal grammars, their interpretation
+%% of types is very different from the semantics of \theoryabbv.
 
@@
 
-\paragraph{Typed Approaches to Regular Expressions}
-The usage of a simple type system to reason about regular expressions was
-introduced by Frisch and Cardelli in 2004 \cite{frischCardelli}, and later expanded
-by Hengelin and Nielsen in 2011 \cite{henglein_regular_2011} to handle proofs of regular expression
-containment. In 1992, while investigating the addition of arbitrary recursion to Kleene
-algebra \cite{leiss},
-Lei{\ss} used a least fixed-point operator instead of the Kleene star. Although
-he did not explicitly make use of a type system, or have access to the view of Frisch and
-Cardelli, Lei{\ss} had added the full power of inductive types to a type system
-of regular expressions.
-
-Pratt's work on the residuation of action algebras in \cite{prattActionLogicPure1991} closely mirrors
-our observation that Brzozowski derivatives mirror linear function types.
-Moreover, he shows that his residuals form a Galois connection with sequential
-composition. This is precisely a posetal --- or \emph{thin} --- version of the
-adjunction between
-linear function types and tensor product in \theoryabbv.
-
-Our primary observation in this paper can be summarized as saying that all of
-these above works can embed naturally into a rich, unifying theory.
-
-\steven{I'd like to synthesize the discussion of these works with the categorial
-grammar works, but I am not sure how. It seems as though there are these
-parallel tracks of research that have missed each other for decades}
-
-\paragraph{Inductive Linear Types}
-
-% In \cite{firsovCertifiedNormalizationContextFree2015}, Firsov and Uustalu
-% constructed a verified normalization procedure to turn a
-% context-free grammar into an equivalent grammar in Chomsky normal form.
-% Combining their work with a CYK parser, they have built a verified context-free
-% grammar parser, \emph{up to weak equivalence}. To upgrade from a weak
-% equivalence to a strong equivalence, they suggest updating their grammar
-% formalism to treat parse trees as first-class proofs of language membership.
-% That is, without explicitly stating so, they suggest that they next natural step
-% is to reason about grammars in a system where parse trees are terms, as we have
-% given in this paper.
-
-% \steven{Make the above paragraph shorter. Also idk if its worth including, but I
-% like that they practically ask for a type theory}
-
-\steven{Mention Neel's typed algebraic approach paper here. Say it is similar at
-  first glance, as its parsing with types, but that he actually is representing
-  something quite different. Or is it so different that its not worth mentioning?}
-
-\steven{TODO add Lambek calculus/categorial grammars here}
-
-\paragraph{Kleene Algebra}
-Since the early works in the theory of formal languages, Kleene
-algebras have played an important role in its development. They
-generalize the operations known from regular languages by introducing
-operations generalizing language composition, language union and the
-Kleene star.  More generally, they are defined as an inequational theory
-where the inequality is meant to capture language containment. This
-theory is extremely successful, having found applications in concurrency theory
-\cite{hoare2000concurrent}, software-defined networks \cite{anderson2014netkat},
-theory of programming languages \cite{angus2001kleene}, compiler optimizations
-\cite{kozen2000certification} and more.
-
-A frequently fruitful research direction is exploring varying
-extensions of Kleene algebras, Kleene algebra with tests (KAT) being
-one of the most notable ones \cite{kozen1997kleene}. Our approach is radically different from
-most extensions, which usually aim at modifying or adding new
-operations to Kleene algebras, but still keeping it as an inequational
-theory. By adopting a category-theoretic treatment and allowing the
-``order structure'' to encode more information than merely
-inequalities, we were able to extend Kleene algebra to reason about
-parsing as well.
-
-\steven{Can we sketch here what an apt categorification of KATs might be? As a
-  poor first pass: perhaps a Kleene category with a full involutive
-  subcategory that corresponds to the Boolean algebra?.}
-\steven{What can we say about internalizing Kleene algebra proofs or extensions?}
-\pedro{As a rule of thumb, I don't think that sketching these ideas
-will get us anything...}
+%% \subsubsection{Kleene Algebra}
+%% Since the early works in the theory of formal languages, Kleene
+%% algebras have played an important role in its development. They
+%% generalize the operations known from regular languages by introducing
+%% operations generalizing language composition, language union and the
+%% Kleene star.  More generally, they are defined as inequational theory
+%% where the inequality is meant to capture language containment.  This
+%% theory is extremely successful, having found applications in concurrency theory
+%% \cite{hoare2009concurrent}, software-defined networks \cite{anderson2014netkat},
+%% theory of programming languages \cite{angus2001kleene}, compiler optimizations
+%% \cite{kozen2000certification} and more.
+
+%% A frequently fruitful research direction is exploring varying
+%% extensions of Kleene algebras, Kleene algebra with tests (KAT) being
+%% one of the most notable ones \cite{kozen1997kleene}. Our approach is
+%% radically different from most extensions, which usually aim at
+%% modifying or adding new operations to Kleene algebras, but still
+%% keeping it as an inequational theory. By adopting a category-theoretic
+%% treatment and allowing the ``order structure'' to encode more
+%% information than merely inequalities, we were able to extend Kleene
+%% algebra to reason about parsing as well.
 
@@
 
-\theoryabbv is similar in spirit to separation logic. Semantically,
-they are closely related: linear types in \theoryabbv denote families
-of sets indexed by a monoid of strings, whereas separation logic
-formulae typically denote families of predicates indexed by an ordered
-partial commutative monoid of worlds. The monoidal structure in both
-cases is are instances of the category-theoretic notion of Day
-convolution monoidal structure. From a separation-logic perspective,
-our notion of memory is very primitive: a memory shape is just a
-string of characters and the state of the memory is never allowed to
-evolve.
-
-Syntactically, our type theory can be viewed as a kind of ``separation
-type theory'' in that we care not just about provability of formulae
-but the constructive content of our linear terms, as these act as
-parsers. The design of \theoryabbv is based on the $\textrm{LNL}_D$
-calculus which explicitly was designed to be a dependent
-type-theoretic version of separation logic. Many concepts of
-separation logic then have direct analogues in our system. The
-characters $c$ are analogous to the ``points-to'' formulae $l \mapsto
-x$ in separation logic. Our $\otimes,\lto,\tol$ are analogous to the
-separating conjunction and magic-wand. The analogue of the persistence
-modality $\square$ of separation logic is $\bigoplus_{\_:\uparrow A} I$,
-which ``zeros out'' all but the parses from the empty string. If we
-think of the portions of the string we are parsing as our notion of
-resource, then these make sense as the ``resource-free'' elements of
-the type $A$.
-
-This semantic connection to separation logic also suggests an avenue
-of future work: rather than working with a dependent type theory, we
-could instead make a program logic based on non-commutative separation
-logic for verifying imperative implementations of parsers. This could
-be accomplished by modifying an existing separation logic
-implementation or embedding the logic within \theoryabbv.
-
-%% \paragraph{Linearity with Dependent Types}
-%% \steven{I don't know if it worth discussing the different influences of
-%%   categorical model here? We take Neel's model and add to it as necessary}
-
-% Vakar might be worth citing to discuss different characterizations of the
-% notion of model \cite{vakarSyntaxSemanticsLinear2015}
-
-% \cite{fu2023twolevellineardependenttype} give a compiler for a two-level
-% linear dependent type theory. This could serve as a template for a
-% compiler implementation of \theoryname that is not just inference rules
-% embedded in Agda.
-
-\paragraph{Implementations in Parsing}
-\steven{Talk about future implementation-side work}
-
-\cite{fu2023twolevellineardependenttype} give a compiler for a two-level
-linear dependent type theory. This could serve as a template for a
-compiler implementation of \theoryname that is not just inference rules
-embedded in Agda.
-
-\paragraph{Type Checking and Elaboration}
+\theoryabbv is similar in spirit to separation logic
+\cite{reynolds_separation_2002}. Semantically, they are closely
+related: linear types in \theoryabbv denote families of sets indexed
+by a monoid of strings, whereas separation logic formulae typically
+denote families of predicates indexed by an ordered partial
+commutative monoid of worlds \cite{jung_higher-order_2016}. The
+monoidal structure in both cases is are instances of the
+category-theoretic notion of Day convolution monoidal structure
+\cite{day1970construction}. From a separation-logic perspective, our
+notion of memory is very primitive: a memory shape is just a string of
+characters and the state of the memory is never allowed to evolve.
+
+%% Syntactically, our type theory can be viewed as a kind of ``separation
+%% type theory'' in that we care not just about provability of formulae
+%% but the constructive content of our linear terms, as these act as
+%% parsers. The design of \theoryabbv is based on the $\textrm{LNL}_D$
+%% calculus which explicitly was designed to be a dependent
+%% type-theoretic version of separation logic. Many concepts of
+%% separation logic then have direct analogues in our system. The
+%% characters $c$ are analogous to the ``points-to'' formulae $l \mapsto
+%% x$ in separation logic. Our $\otimes,\lto,\tol$ are analogous to the
+%% separating conjunction and magic-wand. The analogue of the persistence
+%% modality $\square$ of separation logic is $\bigoplus_{\_:\uparrow A} I$,
+%% which ``zeros out'' all but the parses from the empty string. If we
+%% think of the portions of the string we are parsing as our notion of
+%% resource, then these make sense as the ``resource-free'' elements of
+%% the type $A$.
+
+This semantic connection to separation logic suggests an avenue of
+future work: to develop a program logic based on non-commutative
+separation logic for verifying imperative implementations of
+parsers. This could be implemented by modifying an existing separation
+logic implementation or embedding the logic within \theoryabbv.
+
+\paragraph{Semantic Actions}
+
+Our verification has mainly focused on the verification that a parser
+outputs a correct concrete syntax tree for a grammar. However, in
+practice, parsers are combined with a \emph{semantic action} that
+emits an \emph{abstract} syntax tree that omits superfluous syntactic
+details that aren't needed in later stages of the overall program. We
+can define a semantic action in \theoryabbv for a linear type $A$ with
+semantic outputs in a non-linear type $X$ to be a function $\ltonl {(A
+  \lto \bigoplus_{\_:X} \top)}$. That is, a semantic action is a
+function that produces a semantic element of $X$ from the concrete
+parses of $A$. In future work, we aim to study the question of
+verifying efficient implementations of parsers with semantic actions.
+
+\paragraph{Implementation}
+
+Our Agda prototype implementation serves as a useful proof of concept
+for showing what can be implemented in \theoryabbv, but it has
+downsides we aim to address in future work. Firstly, it would be
+preferable to work with the more intuitive type theoretic syntax we
+have used in this work, rather than the combinator-style our shallow
+embedding requires. Additionally, Agda itself does not have a
+high-performance implementation, and so the parsers we implement in
+Agda do not have competitive performance to industry parser
+generators. In future work we aim to study if we can embed a proof of
+the correctness of a parser generator that produces imperative
+programs, and if the correctness of those imperative
+programs can be proven within \theoryabbv.
+
+\paragraph{Type Checking and Semantic Analysis}
 Our focus in this work has been on the verification of parsers for
@@
 definition of arbitrarily powerful grammars, the system could also be
-used for more sophisticated semantic analysis such as scope checking
-or type checking. Alternatively, we could more directly encode type
-type systems as linear types in a modified version of \theoryabbv
-whose semantics is not in families of sets indexed by \emph{strings}
-but families of sets indexed by \emph{abstract syntax trees}.
+used in principle for more sophisticated semantic analysis such as
+scope checking or type checking. Alternatively, we could more directly
+encode type type systems as linear types in a modified version of
+\theoryabbv where linear types are not grammars over strings, but
+\emph{type systems} over trees. This could analogously serve as a
+framework for verified type checking and static analysis.
 
@@
 
+\pagebreak
+
+\ifappendix
 \appendix
@@
 \section{Syntax}
+
+\begin{figure}
+  \begin{align*}
+    \el(\Var\,M) B &= B\,M\\
+    \el(\mathsf{K}\,A) B &= A\\
+    \el(\bigoplus A) B &= \bigoplus_{y:Y}\el(A y)B\\
+    \el(\bigamp A) B &= \bigamp_{y:Y}\el(A y)B\\
+    \el(A \otimes A') B &= \el(A)B \otimes \el(A')B\\\\
+    \map(\Var\,M)\,f &= f\,M\\
+    \map(\K\,A)\,f &= \lambda a. a\\
+    \map(\bigoplus A)\,f &= \lambda a. \letin{\oplusinj y {a_y}}{a}{\oplusinj y {\map(A\,y)\,f\,a_y}}\\
+    \map(\bigwith\,A)\,f &= \lambda a. \withlamb{y}{\map(A\,y)\,f\,(\withprj y a)}\\
+    \map(A \otimes A')\,f &= \lambda b. \letin{(a,a')}{b}{(\map(A)\,f\,a,\map(A')\,f\,a')}
+  \end{align*}
+  \caption{Strictly positive functors functorial actions}
+\end{figure}
 
@@
   \end{mathpar}
-  \caption{Context Rules}
+  \caption{Context rules}
 \end{figure}
 
-\max{TODO: add rules for non-linear types $\Pi,\Sigma$, equality etc}
-
-\begin{figure}
-  \begin{mathpar}
-  \boxed{\Gamma \vdash M : X}
-
-  \inferrule{~}{\Gamma, x : X, \Gamma' \vdash x : X}
-  %
-  \and
-  %
-  \inferrule{\Gamma \vdash M : Y \quad \ctxwffjdg \Gamma {X \equiv Y}}{\Gamma \vdash M : X}
-  %
-  \and
-  %
-  \inferrule{~}{\Gamma \vdash () : 1}
-  %
-  \and
-  %
-  \inferrule{\Gamma \vdash M : X \\ \Gamma \vdash N : \subst Y M x}{\Gamma \vdash (M, N) : \SigTy x X Y}
-  %
-  \and
-%
-  \inferrule{\Gamma \vdash M : \SigTy x X Y}{\Gamma \vdash \pi_1\, M : X}
-  %
-  \and
-  %
-  \inferrule{\Gamma \vdash M : \SigTy x X Y}{\Gamma \vdash \pi_2\, M : \subst Y {\pi_1\, M} x}
-  \and
-  \inferrule{\Gamma, x : X \vdash M : Y}{\Gamma \vdash \lamb x M : \PiTy x X Y}
-  %
-  \and
-  %
-  \inferrule{\Gamma \vdash M : \PiTy x X Y \\ \Gamma \vdash N : X}{\Gamma \vdash \app M {N} : \subst Y {N} x}
-  %
-  \and
-  %
-  \inferrule{\Gamma \vdash M \equiv N : X}{\Gamma \vdash \mathsf{refl} : M =_X N}
-  \and
-  \textrm{TODO: inductive types (boolean, nat) with their eliminators allowed in }
-  \end{mathpar}
-  \caption{Intuitionistic typing}
-  \label{fig:inttyping}
-\end{figure}
+%% \begin{figure}
+%%   \begin{mathpar}
+%%   \boxed{\Gamma \vdash M : X}
+
+%%   \inferrule{~}{\Gamma, x : X, \Gamma' \vdash x : X}
+%%   %
+%%   \and
+%%   %
+%%   \inferrule{\Gamma \vdash M : Y \quad \ctxwffjdg \Gamma {X \equiv Y}}{\Gamma \vdash M : X}
+%%   %
+%%   \and
+%%   %
+%%   \inferrule{~}{\Gamma \vdash () : 1}
+%%   %
+%%   \and
+%%   %
+%%   \inferrule{\Gamma \vdash M : X \\ \Gamma \vdash N : \subst Y M x}{\Gamma \vdash (M, N) : \SigTy x X Y}
+%%   %
+%%   \and
+%% %
+%%   \inferrule{\Gamma \vdash M : \SigTy x X Y}{\Gamma \vdash \pi_1\, M : X}
+%%   %
+%%   \and
+%%   %
+%%   \inferrule{\Gamma \vdash M : \SigTy x X Y}{\Gamma \vdash \pi_2\, M : \subst Y {\pi_1\, M} x}
+%%   \and
+%%   \inferrule{\Gamma, x : X \vdash M : Y}{\Gamma \vdash \lamb x M : \PiTy x X Y}
+%%   %
+%%   \and
+%%   %
+%%   \inferrule{\Gamma \vdash M : \PiTy x X Y \\ \Gamma \vdash N : X}{\Gamma \vdash \app M {N} : \subst Y {N} x}
+%%   %
+%%   \and
+%%   %
+%%   \inferrule{\Gamma \vdash M \equiv N : X}{\Gamma \vdash \mathsf{refl} : M =_X N}
+%%   \and
+%%   \textrm{TODO: inductive types (boolean, nat) with their eliminators allowed in }
+%%   \end{mathpar}
+%%   \caption{Intuitionistic typing}
+%%   \label{fig:inttyping}
+%% \end{figure}
 
@@
     \inferrule{\Gamma ; \Delta \vdash e : B \tol A \\ \Gamma ; \Delta' \vdash e' : A}{\Gamma ; \Delta', \Delta \vdash \apptol e {e'} : B}
-    %
-    \\
-
   \end{mathpar}
-  \caption{Linear Terms, extended}
+  \caption{Linear terms (extended)}
 \end{figure}
@@
   \begin{mathpar}
-    \boxed{\Gamma \vdash M \equiv N : X}
-
-    \inferrule{\Gamma \vdash p : M =_X N}{\Gamma \vdash M \equiv N : X}
+    \boxed{\Gamma ; \Delta \vdash e \equiv e' : A}
+
+    % \inferrule{~}{\Gamma; \cdot \vdash G^{-1}\, (G \, t ) \equiv t: A}
+%
+%
+    \inferrule{\Gamma ; \Delta , a : A \vdash e : C \\ \Gamma ; \Delta' \vdash e' : A }{\Gamma; \Delta,\Delta' \vdash \app {(\lamblto a e)} {e'} \equiv \subst e {e'} a : C}
 %
@@
 %
-    \inferrule{~}{\Gamma \vdash \app {(\lamb x M)} {N} \equiv \subst x M {N} : X}
+    \inferrule{\Gamma ; \Delta \vdash e : A \lto B}{\Gamma; \Delta \vdash e \equiv \lamblto a {\app e a} : A \lto B}
 %
@@
 %
-    \inferrule{~}{\Gamma \vdash m \equiv \lamb x {\app M x} : \PiTy x X Y}
+    \inferrule{\Gamma ; a : A , \Delta \vdash e : C \\ \Gamma ; \Delta' \vdash e' : A }{\Gamma; \Delta , \Delta' \vdash \app {(\lambtol a e)} {e'} \equiv \subst e {e'} a : C}
 %
@@
 %
-    \inferrule{~}{\Gamma \vdash \pi_1\, (M, N) \equiv M : X}
+    \inferrule{\Gamma ; \Delta \vdash e : B \tol A}{\Gamma; \Delta \vdash e \equiv \lambtol a {\app e a} : B \tol A}
 %
@@
 %
-    \inferrule{~}{\Gamma \vdash \pi_2\, (M , N) \equiv N : \subst x {M} Y}
+    \inferrule{\Gamma , x : X \vdash e : A \\ \Gamma \vdash M : X}{\Gamma; \Delta \vdash \app {(\dlamb x e)} {M} \equiv \subst {e} M x : C}
 %
@@
 %
-    \inferrule{~}{\Gamma \vdash M \equiv (\pi_1\, M, \pi_2\, M) : \SigTy x X Y}
-%
-    \and
-%
-    % \inferrule{~}{}
-%
-    \inferrule{~}{\Gamma \vdash M \equiv N : 1}
-%
-    % \and
-%
-    % \inferrule{~}{\Gamma \vdash G\, (G^{-1} \, t) \equiv t : G A}
-
-    \\
-
-    \boxed{\Gamma ; \Delta \vdash e \equiv e' : A}
-
-    % \inferrule{~}{\Gamma; \cdot \vdash G^{-1}\, (G \, t ) \equiv t: A}
-%
-%
-    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lamblto a e)} {e'} \equiv \subst e {e'} a : C}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lamblto a {\app e a} : A \lto B}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lambtol a e)} {e'} \equiv \subst e {e'} a : C}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lambtol a {\app e a} : B \tol A}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash \app {(\dlamb x e)} {M} \equiv \subst {e} M x : C}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \dlamb x {\app e x} : \LinPiTy x X A}
+    \inferrule{\Gamma ; \Delta \vdash e : \LinPiTy {x} {X} {A}}{\Gamma; \Delta \vdash e \equiv \dlamb x {\app e x} : \LinPiTy x X A}
 %
@@
 %
-    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} {()} e \equiv e : C}
+    \inferrule{\Gamma ; \Delta_1 , \Delta_2 \vdash e : C}{\Gamma; \Delta_1 , \Delta_2 \vdash \letin {()} {()} e \equiv e : C}
     %
@@
     %
-    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} e {\subst {e'} {()} a} \equiv \subst {e'} e a : C}
+    \inferrule{\Gamma ; \Delta_2 \vdash e : I \\ \Gamma ; \Delta_1 , a : A , \Delta_3 \vdash e' : C}{\Gamma; \Delta_1 , \Delta_3 \vdash \letin {()} e {\subst {e'} {()} a} \equiv \subst {e'} e a : C}
 %
@@
 %
-    \inferrule{~}{\Gamma; \Delta \vdash \letin {a \otimes b} {e \otimes e'} e'' \equiv e'' \{ e/a , e'/b \} : C}
+    \inferrule{\Gamma ; \Delta_2 \vdash e : A \\ \Gamma ; \Delta_3 \vdash e' : B \\ \Gamma ; \Delta_1 , a : A , b : B , \Delta_4 \vdash e'' : C}{\Gamma; \Delta_1 , \Delta_2, \Delta_3 , \Delta_4 \vdash \letin {(a, b)} {(e , e')} e'' \equiv e'' \{ e/a , e'/b \} : C}
 %
@@
     %
-    \inferrule{~}{\Gamma; \Delta \vdash \letin {a \otimes b} e {\subst {e'} {a \otimes b} c} \equiv \subst {e'} e c : C}
+    \inferrule{\Gamma ; \Delta_2 \vdash e : A \otimes B \\ \Gamma ; \Delta_1 , c : A \otimes B , \Delta_3 \vdash e' : C }{\Gamma; \Delta_1, \Delta_2, \Delta_3 \vdash \letin {(a , b)} e {\subst {e'} {(a, b)} c} \equiv \subst {e'} e c : C}
 %
@@
 %
-    \inferrule{~}{\Gamma;\Delta \vdash \letin {(x, a)} {(M, e)} {e'} \equiv e' \{ M/x , e/a \} : C}
+    \inferrule{\Gamma \vdash M : X \\ \Gamma ; \Delta_2 \vdash e : A \\ \Gamma , x : X \vdash \Delta_1 , a : A , \Delta_3 }{\Gamma;\Delta_1 , \Delta_2 , \Delta_3 \vdash \letin {\sigma~x~a} {\sigma~M~e} {e'} \equiv e' \{ M/x , e/a \} : C}
     %
@@
     %
-    \inferrule{~}{\Gamma; \Delta \vdash \letin {(x, a)} e {\subst {e'} {(x, a)} y} \equiv \subst {e'} e y : C}
+    \inferrule{\Gamma ; \Delta_1, y : \LinSigTy {x} {X} {A}, \Delta_2 \vdash e' : C \\ \Gamma ; \Delta_2 \vdash e : \LinSigTy {x} {X} {A}}{\Gamma; \Delta_1 , \Delta_2 , \Delta_3 \vdash \letin {\sigma~x~a} e {\subst {e'} {\sigma~x~a} y} \equiv \subst {e'} e y : C}
     \and
@@
     \inferrule
-    {~}
+    {\Gamma ; \Delta \vdash e : A \\ \Gamma ; \Delta \vdash f~e \equiv g~e}
     {\Gamma ; \Delta \vdash \equalizerpi {\equalizerin {e}} \equiv e : A}
@@
     \inferrule
-    {~}
-    {\Gamma ; \Delta \vdash \equalizerin {\equalizerpi {e}} \equiv e : A}
+    {\Gamma ; \Delta \vdash e : \equalizer e f g}
+    {\Gamma ; \Delta \vdash \equalizerin {\equalizerpi {e}} \equiv e :
+      \equalizer e f g}
 \end{mathpar}
@@
 \end{figure}
-\steven{Do we need inverse for $\ltonl A$ in judgmental equality figure?}
-\steven{Do we need to say anything about the equalizer proofs being unique in
-  judgmental equality figure?}
+\steven{Need to include premises for the judgmental equality rules}
 
@@
 
-  \max{do we need identity and composition of substitutions and their functoriality?}
+  %% \max{do we need identity and composition of substitutions and their functoriality?}
   %% By induction we can define an identity substitution
@@
   the type theory that are parameterized by a non-linear context
-  $\Gamma$ support an admissible action of a substitution $\gamma \in
-  \Subst(\Gamma',\Gamma)$:
+  $\Gamma$ support the admissible actions of a substitution $\gamma
+  \in \Subst(\Gamma',\Gamma)$ in \Cref{fig:non-linear-substitution}.
+  %% Further, these all satisfy identity and composition equations M[\Gamma] = M and M[\gamma][\gamma'] = M[(\gamma \circ \gamma')]
+\end{definition}
+
+\begin{figure}
   \begin{mathpar}
@@
   \end{mathpar}  
-
-  %% Further, these all satisfy identity and composition equations M[\Gamma] = M and M[\gamma][\gamma'] = M[(\gamma \circ \gamma')]
-\end{definition}
+  \caption{Non-linear substitution}
+  \label{fig:non-linear-substitution}
+\end{figure}
 
@@
     \Subst(\Delta',\cdot) &= \{ \cdot \pipe \Delta' = \cdot \}\\
-    \Subst(\Delta',(\Delta,a:A)) &= \{ (\delta, e/a) \pipe \delta \in \Subst(\Delta_1, \Delta) \and \Delta_2 \vdash e : A \and \Delta' = \Delta_1,\Delta_2\}
+    \Subst(\Delta',(\Delta,a:A)) &= \{ (\delta, e/a) \pipe \delta \in \Subst(\Delta_1, \Delta) , \Delta_2 \vdash e : A , \Delta' = (\Delta_1,\Delta_2)\}
   \end{align*}
@@
 
-\begin{definition}
-  Given any $\Gamma ; \Delta \vdash e : A$ and $\delta \in
-  \Subst(\Delta',\Delta)$, we define the action of the substitution on
-  $e$ as follows, frequently using the inversion principle to split
-  the substitution into constituent components.
+\begin{figure}
   \begin{align*}
@@
   \end{align*}
+\caption{Action of substitution on linear terms}
+\label{fig:substact}
+\end{figure}
+\steven{captilizataion should be consistent over all figures. idk if all caps or
+all lowercase is beter?}
+
+\begin{definition}
+  Given any $\Gamma ; \Delta \vdash e : A$ and $\delta \in
+  \Subst(\Delta',\Delta)$, we define the action of the substitution on
+  $e$ in \cref{fig:substact}, frequently using the inversion principle to split
+  the substitution into constituent components.
   By induction on linear term and equality judgments, we establish the
@@
 
-  \max{identity and composition?}
+  %% \max{identity and composition?}
 \end{definition}
@@
 \section{Denotational Semantics}
-
-\begin{definition}
-  We define the unit grammar
-  \[ I\,w = \{ ()\pipe w = \varepsilon \}\]
-  and given grammars $A$ and $B$ we define their tensor product
-  \[ (A \otimes B)\,w = \{ (w_1,w_2,a,b) \pipe w_1w_2 = w \wedge a \in A\,w_1 \wedge b \in B\,w_2 \} \]
-  This extends to a monoidal structure on the category of grammars.
-Note that we defined in the semantics of linear types $\sem{I}\gamma = I$ and $\sem{A \otimes B}\gamma = \sem{A}\gamma \otimes \sem{B} \gamma$
-\end{definition}
+Here we extend the denotational semantics from
+\cref{sec:semantics-and-metatheory} to cover all of \theoryname
+syntax. Here we will freely use that the category of grammars is a
+complete, co-complete biclosed monoidal category, and use categorical
+notation for the constructions in the denotational semantics. For
+example, we will use the same notation $I,\otimes,\lto,\tol$ for the
+biclosed monoidal structure of $\Grammar$ that we do for the
+corresponding syntactic notions.
 
@@
 
-We define denotations of linear terms in
-\Cref{fig:linear-term-semantics}. Note that the denotations interpret
-typing derivations, not raw terms, as the data of how contexts are
-split is needed in order to construct the correct associator
-functions.
-\max{todo: describe associators, define denotations, prove axioms}
-\begin{figure}
-  \begin{mathpar}
-  \inferrule
-  {\Delta_1 \vdash e_1 : A_1 \and \Delta_2 \vdash e_2 : A_2}
-  {\sem{(e_1,e_2)}\gamma\,w\,\delta
-    = \letin{(w_1,w_2,\delta_1,\delta_2)} {m(\Delta_1,\Delta_2)(\delta)} {(w_1,w_2,\sem{e_1}\gamma\,w_1\,\delta_1, \sem{e_2}\gamma\,w_2\,\delta_2)}
-  }
-
-  \inferrule
-  {\Delta_2 \vdash e : A \otimes B \and \Delta_1,a:A,b:B,\Delta_3 \vdash e' : C}
-  {\sem{\letin{(a,b)}{e}{e'}}\gamma\,w\,\delta =
-    \letin{(w_1,w_2w_3,\delta_1,(w_2,w_3,\delta_2,\delta_3))}{m_{\Delta_1,\Delta_2,\Delta_3}(\delta)}
-    \letin{(w_{2,1},w_{2,2},a,b)}{\sem{e}\gamma\,w_2\,\delta_2}
-    {\sem{e'}\gamma\,w\,(w_1,w_{2,1}w_{2,2}w_3, \delta_1, (w_{2,1},w_{2,2}w_3, a, (w_{2,2},w_3, b, \delta_3)))}
-    }
-\end{mathpar}
-  \caption{Denotations of linear terms}
-\end{figure}
-
-\begin{definition}
+\begin{definition}[Denotation of Linear Substitutions]
+  The semantics of a linear substitution
+  $\delta : \textrm{Subst}(\Delta' , \Delta)$ are given as maps
+  $\semg{\delta} : \semg{\Delta'} \to {\Delta}$. Define $\semg{\delta}$ by
+  recursion on $\delta$:
+  \begin{align*}
+    \semg{\cdot} &= id_{I} \\
+    \semg{\delta,e/a} &= \semg{\delta} \otimes \semg{e} \circ m_{\Delta_{1},\Delta_{2}}
+  \end{align*}
+where $\Delta_{2} \vdash e : A$ and $\Delta' = \Delta_{1},\Delta_{2}$.
+\end{definition}
+
+\begin{theorem}
   For any $\Gamma \vdash \Delta_1,\Delta_2 \isLinCtx$ and $\gamma \in
-  \sem{\Gamma}$ there is an isomorphism $m_{\Delta_1,\Delta_2} :
+  \sem{\Gamma}$ there is a natural isomorphism $m_{\Delta_1,\Delta_2} :
   \sem{\Delta_1, \Delta_2}\gamma \cong \sem{\Delta_1}\gamma \otimes
@@
   This can be extended to a sequence of contexts of any length.
-\end{definition}
-
-\begin{lemma}[Equations for $\otimes$]
-
+\end{theorem}
+\begin{proof}
+Construct $m_{\Delta_{1},\Delta_{2}}$ by recursion on $\Delta_{2}$.
+\begin{align*}
+  m_{\Delta_{1},\cdot} & = \rho^{-1} \\
+  m_{\Delta_{1},(\Delta_{2}, a : A)} & = \alpha \circ m_{\Delta_{1},\Delta_{2}} \otimes id
+\end{align*}
+\end{proof}
+
+\begin{lemma}
+  \label{lem:subst}
+  For each term $\Delta \vdash e : A$ and substitution
+  $\delta : \textrm{Subst}(\Delta', \Delta)$, the semantics of $\delta$ acting
+  on $e$ splits into the composition $\semg{e[\delta]} = \semg{e} \circ \semg{\delta}$.
 \end{lemma}
 
-\begin{lemma}[Equations for $I$]
-
-\end{lemma}
-
-\begin{lemma}[Equations for $\lto$]
-
-\end{lemma}
-
-\begin{lemma}[Equations for $\tol$]
-
-\end{lemma}
-
+\subsection{Grammar Semantics for Linear Terms}
+Here we define denotations of linear terms. Note that the denotations
+interpret typing derivations, not raw terms, as the data of how
+contexts are split is needed in order to construct the correct
+associator functions. Further, we demonstrate that the denotational
+semantics respects the equational theory of \theoryabbv. The
+correctness of the equational theory heavily relies on the
+\emph{coherence theorem} for monoidal categories. The coherence
+theorem says that any diagram in a monoidal category constructed using
+only associators $\alpha_{A,B,C} : (A \otimes B) \otimes C \cong A
+\otimes (B \otimes C)$, unitors $\rho_A : A \otimes I \cong A$ and
+$\lambda_A : I \otimes A \cong A$ and compositions and tensor products
+of these, commutes. We call a morphism built in this way a
+\emph{generalized associator}.
+
+\subsubsection{Variables}
+Note that the denotation of a singleton context $a : A$ is given as
+\[
+  \semg{a : A} = \semg{\cdot , a : A} = I \otimes \semg{A}
+\]
+
+So for $a: A \vdash a : A$, the denotation of a single variable term
+$\semg{a} : \semg{a : A} \to \semg{A}$ is given by the left unitor
+\[
+  \semg{a} = \lambda
+\]
+
+\subsubsection{Linear Unit}
+% I
+\paragraph{$I$-Introduction}
+$\semg{()} : \semg{\cdot} \to \semg{I}$.
+\[
+\semg{()} = id_{I}
+\]
+
+\paragraph{$I$-Elimination}
+$\semg{\letin {()} e e'} : \semg{\Delta'_{1},\Delta,\Delta'_{2}} \to \semg{C}$
+defined in the following diagram,
+
+\begin{center}
+\begin{tikzcd}
+  \semg{\Delta'_1,\Delta,\Delta'_2} \arrow[r , "m_{(\Delta'_1,\Delta),\Delta'_2}"] &
+  \semg{\Delta'_1,\Delta} \otimes \semg{\Delta'_2}
+    \arrow[r , "m_{\Delta'_1,\Delta} \otimes id"] &
+  \left(\semg{\Delta'_1} \otimes \semg{\Delta} \right) \otimes \semg{\Delta'_2}
+    \arrow[d , "(id \otimes \semg{e}) \otimes id"] \\
+    \semg{\Delta'_1,\Delta'_2} \arrow[d , "\semg{e'}"]
+  & \arrow[l , "m^{-1}_{\Delta'_1,\Delta'_2}"]
+    \semg{\Delta'_1} \otimes \semg{\Delta'_2}
+  & \left(\semg{\Delta'_1} \otimes I \right) \otimes \semg{\Delta'_2}
+  \arrow[l , "\rho \otimes id"] \\
+  \semg{C}
+\end{tikzcd}
+\end{center}
+
+We demonstrate that the denotations of the introduction and elimination forms
+for $I$ obey the $\beta$ and $\eta$ equalities for $I$.
+
+\paragraph{$I\beta$}
+Given
+$\Delta'_{1},\cdot,\Delta'_{2} \vdash e' : C$, the desired
+$\beta$ law is
+\[
+  \semg{\letin {()} {()} {e'}} = \semg{e}
+\]
+\begin{proof}
 \begin{align*}
-  \sem{(e_1,e_2)}\gamma\,w\,\delta &=
-  \letin{(w_1,w_2,\delta_1,\delta_2)} {m(\Delta_1,\Delta_2)(\delta)} {(w_1,w_2,\sem{e_1}\gamma\,w_1\,\delta_1, \sem{e_2}\gamma\,w_2\,\delta_2)}\\
+  \semg{\letin {()} {()} {e'}}
+  & = \semg{e'} \circ m^{-1}_{\Delta'_{1},\Delta'_{2}} \circ \rho \otimes id \circ (id \otimes \semg{()}) \otimes id \circ m_{\Delta_{1},\cdot} \otimes id \circ m_{(\Delta'_{1},\cdot),\Delta'_{2}} \\
+  & = \semg{e'} \circ m^{-1}_{\Delta'_{1},\Delta'_{2}} \circ \rho \otimes id \circ (id \otimes id) \otimes id \circ m_{\Delta_{1},\cdot} \otimes id \circ m_{\Delta'_{1},\Delta'_{2}} \\
+  & = \semg{e'} \circ m^{-1}_{\Delta'_{1},\Delta'_{2}} \circ \rho \otimes id \circ \rho^{-1} \otimes id \circ m_{\Delta'_{1},\Delta'_{2}} \\
+  & = \semg{e'} \tag{coherence}
 \end{align*}
-
+\end{proof}
+
+\paragraph{$I\eta$}
+Similarly, for $\Delta_{1}, a : A, \Delta_{3} \vdash e' : C$ and
+$\Delta_{2} \vdash e : I$ the desired $\eta$ law is
+\[
+\semg{\letin {()} {e} {e'[()/a]}} = \semg{e'[e/a]}
+\]
+However, through application of \cref{lem:subst} it suffices to handle the case
+where $e$ is a variable $a'$. That is,
+\begin{align*}
+\letin {()} {e} {e'[()/a]} &= (\letin {()} {a'} {e'[()/a]})[e/a'] \\
+e'[e/a] &= e'[a'/a][e/a]
+\end{align*}
+so without loss of generality we may take may take $e$ to be variable $a'$. We
+will additionally use this style of argumentation when necessary throughout this section.
+
+\begin{proof}
+\begin{align*}
+  \semg{\letin {()} {a'} {e'[()/a]}}
+  & = \semg{e'} \circ m^{-1}_{\Delta'_{1},\Delta'_{2}} \circ \rho \otimes id \circ (id \otimes \semg{a'}) \otimes id \circ m_{\Delta_{1},\cdot} \otimes id \circ m_{\Delta'_{1},\Delta'_{2}} \\
+  & = \semg{e'} \circ m^{-1}_{\Delta'_{1},\Delta'_{2}} \circ \rho \otimes id \circ (id \otimes \lambda) \otimes id \circ m_{\Delta_{1},\cdot} \otimes id \circ m_{\Delta'_{1},\Delta'_{2}} \\
+  & = \semg{e'} \tag{coherence}
+\end{align*}
+
+Because
+$m^{-1}_{\Delta'_{1},\Delta'_{2}} \circ \rho \otimes id \circ (id \otimes \lambda) \otimes id \circ m_{\Delta_{1},\cdot} \otimes id \circ m_{\Delta'_{1},\Delta'_{2}}$
+is a composition of generalized associators from
+$\semg{\Delta'_{1},\Delta'_{2}}$ to itself, it is equal to the identity by the coherence theorem for monoidal categories.
+
+Further by \cref{lem:subst},
+\begin{align*}
+  \semg{e'[a'/a]}
+  &= \semg{e'} \circ \semg{a'/a} \tag{\cref{lem:subst}} \\
+  &= \semg{e'} \tag{coherence}
+\end{align*}
+Again by the coherence theorem, $\semg{a'/a} = id$. Thus, $\eta$ law for $I$
+holds in the denotational semantics.
+\end{proof}
+
+%tensor
+\subsubsection{Tensor}
+\paragraph{$\otimes$-Introduction}
+$\semg{(e_{1}, e_{2})} : \semg{\Delta,\Delta'} \to \semg{A \otimes B}$ is given
+by the diagram
+
+\begin{center}
+\begin{tikzcd}
+  \semg{\Delta,\Delta'} \arrow[r, "m_{\Delta,\Delta'}"] &
+  \semg{\Delta} \otimes \semg{\Delta'} \arrow[r, "\semg{e_1} \otimes \semg{e_2}"] &
+  \semg{A} \otimes \semg{G}
+\end{tikzcd}
+\end{center}
+
+\paragraph{$\otimes$-Elimination}
+$\semg{\letin {(a,b)} e f} : \semg{\Delta'_{1},\Delta,\Delta'_{2}} \to \semg{C}$
+defined via the diagram,
+
+\begin{center}
+\begin{tikzcd}
+  \semg{\Delta'_1,\Delta,\Delta'_2} \arrow[r , "m_{(\Delta'_1,\Delta),\Delta'_2}"] &
+  \semg{\Delta'_1,\Delta} \otimes \semg{\Delta'_2}
+    \arrow[r , "m_{\Delta'_1,\Delta} \otimes id"] &
+  \left(\semg{\Delta'_1} \otimes \semg{\Delta} \right) \otimes \semg{\Delta'_2}
+    \arrow[d , "(id \otimes \semg{e}) \otimes id"] \\
+  \semg{c} & \arrow[l , "\semg{f}"]
+  \left( \left( \semg{\Delta'_1} \otimes \semg{A} \right) \otimes \semg{B}\right) \otimes \semg{\Delta'_2}
+  & \left(\semg{\Delta'_1} \otimes \left( \semg{A} \otimes \semg{B}\right) \right) \otimes \semg{\Delta'_2}
+    \arrow[l , "\alpha \otimes id"]
+\end{tikzcd}
+\end{center}
+
+\paragraph{$\otimes\beta$}
+The desired $\beta$ equality for $\otimes$ is,
+\[
+  \semg{\letin {(a,b)} {(a',b')} {f}} = \semg{f[a'/a,b'/b]}
+\]
+
+\begin{proof}
+The left hand side reduces as follows,
+\begin{align*}
+  \semg{\letin {(a,b)} {(a',b')} {f}}
+  & = \semg{f} \circ \alpha \otimes id \circ (id \otimes \semg{(a' , b')}) \otimes id \circ m_{\Delta,\Delta'} \\
+  & = \semg{f} \circ \alpha \otimes id \circ (id \otimes \lambda \otimes \lambda \circ m_{a : A, b : B})) \otimes id \circ m_{\Delta,\Delta'} \\
+  & = \semg{f} \tag{coherence}
+\end{align*}
+
+Which is equal to the right hand side,
+\begin{align*}
+  \semg{f[a'/a,b'/b]}
+  & = \semg{f} \circ \semg{a'/a,b'/b} \\
+  & = \semg{f} \tag{coherence}
+\end{align*}
+\end{proof}
+
+\paragraph{$\otimes\eta$}
+The desired $\eta$ equality for $\otimes$ is,
+\[
+  \semg{\letin {(a,b)} {c'} {f[(a,b)/c]}} = \semg{f[c'/c]}
+\]
+
+\begin{proof}
+\begin{align*}
+  \semg{\letin {(a,b)} {c'} {f[(a,b)/c]}}
+  & = \semg{f[(a,b)/c]} \circ \alpha \otimes id \circ (id \otimes \semg{c'}) \otimes id \circ m_{\Delta,\Delta'} \\
+  & = \semg{f} \circ \semg{(a,b)/c} \circ \alpha \otimes id \circ (id \otimes \semg{c'}) \otimes id \circ m_{\Delta,\Delta'} \\
+  & = \semg{f} \tag{coherence}
+\end{align*}
+
+\begin{align*}
+  \semg{f[c'/c]}
+  &= \semg{f} \circ \semg{c'/c} \tag{\cref{lem:subst}} \\
+  &= \semg{f} \tag{coherence}
+\end{align*}
+\end{proof}
+
+\subsubsection{$\lto$-Functions}
+% linear function
+\paragraph{$\lto$-Introduction}
+$\semg{\lamblto a e} : \semg{\Delta} \to \semg{A \lto B}$ is defined using the
+natural isomorphism
+$\phi : Hom(\semg{\Delta} \otimes \semg{A}, \semg{B}) \to Hom(\semg{\Delta}, \semg{A \lto B})$
+that is provided by the adjunction between $\semg{- \otimes A}$ and $\semg{A \lto -}$.
+
+\[
+\semg{\lamblto a e} = \phi\left( \semg{e} \right)
+\]
+
+\paragraph{$\lto$-Elimination}
+$\semg{\applto {e'} {e}} : \semg{\Delta,\Delta'} \to \semg{B}$ is defined by the
+diagram,
+
+\begin{center}
+\begin{tikzcd}
+  \semg{\Delta,\Delta'} \arrow[r , "m_{\Delta,\Delta'}"] &
+  \semg{\Delta} \otimes \semg{\Delta'} \arrow[r , "id \otimes \semg{e}"] &
+  \semg{\Delta} \otimes \semg{A} \arrow[r , "\phi^{-1}\left( \semg{e'} \right)"]
+  &
+  \semg{B}
+\end{tikzcd}
+\end{center}
+
+\paragraph{$\lto\beta$}
+The $\beta$ rule for $\lto$ is given by,
+\[
+  \semg{\applto {(\lamblto {a} {e})} {a'}} = \semg{e[a'/a]}
+\]
+\begin{proof}
+\begin{align*}
+  \semg{\applto {(\lamblto {a} {e})} {a'}}
+  & = \phi^{-1}(\semg{\lamblto {a} {e}}) \circ id \otimes \semg{a'} \circ m_{\Delta,\Delta'} \\
+  & = \phi^{-1}(\phi(\semg{e})) \circ id \otimes \semg{a'} \circ m_{\Delta,\Delta'} \\
+  & = \semg{e} \circ id \otimes \semg{a'} \circ m_{\Delta,\Delta'} \\
+  & = \semg{e} \circ id \otimes \lambda \circ m_{\Delta,\Delta'} \\
+  & = \semg{e} \tag{coherence}
+\end{align*}
+
+\begin{align*}
+  \semg{e[a'/a]}
+  & = \semg{e} \circ \semg{a'/a} \tag{\cref{lem:subst}} \\
+  & = \semg{e} \tag{coherence}
+\end{align*}
+\end{proof}
+
+\paragraph{$\lto\eta$}
+The $\eta$ rule for $\lto$ is given by,
+\[
+  \semg{\lamblto {a} {\applto e a}} = \semg{e}
+\]
+
+\begin{proof}
+\begin{align*}
+  \semg{\lamblto {a} {\applto e a}}
+  & = \phi(\semg{\applto e a}) \\
+  & = \phi(\phi^{-1}(\semg{e}) \circ id \otimes \semg{a} \circ m_{\Delta,a:A}) \\
+  & = \phi(\phi^{-1}(\semg{e}) \circ id \otimes \lambda \circ m_{\Delta,a:A}) \\
+  & = \phi(\phi^{-1}(\semg{e})) \tag{coherence} \\
+  & = \semg{e}
+\end{align*}
+\end{proof}
+
+\subsubsection{$\tol$-Functions}
+\paragraph{$\tol$-Introduction}
+Just as with the other linear function type, we have an adjunction between
+$\semg{A \otimes -}$ and $\semg{- \tol A}$. $\semg{\lambtol a e} : \semg{\Delta} \to \semg{B \tol A}$ is defined using the
+natural isomorphism
+$\psi : Hom(\semg{A} \otimes \semg{\Delta}, \semg{B}) \to Hom(\semg{\Delta}, \semg{B \tol A})$
+induced by this adjunction. In particular, $\semg{\lambtol a e}$ is given by $\psi$
+acting on the following diagram
+
+\begin{center}
+\begin{tikzcd}
+  \semg{A} \otimes \semg{\Delta} \arrow[r , "\lambda^{-1} \otimes id"] &
+  \semg{a : A} \otimes \semg{\Delta} \arrow[r , "m^{-1}_{a : A , \Delta}"] &
+  \semg{a : A , \Delta} \arrow[r , "\semg{e}"] &
+  \semg{B}
+\end{tikzcd}
+\end{center}
+
+\paragraph{$\tol$-Elimination}
+The application of a $\tol$-function,
+$\semg{\apptol {e} {e'}} : \semg{\Delta',\Delta} \to \semg{B}$ defined by the diagram
+
+\begin{center}
+\begin{tikzcd}
+  \semg{\Delta',\Delta} \arrow[r , "m_{\Delta',\Delta}"] &
+  \semg{\Delta'} \otimes \semg{\Delta} \arrow[r , "\semg{e'} \otimes id"] &
+  \semg{A} \otimes \semg{\Delta} \arrow[r , "\psi^{-1}\left( \semg{e} \right)"] &
+  \semg{B}
+\end{tikzcd}
+\end{center}
+
+\paragraph{$\tol\beta$}
+The $\beta$ rule for $\tol$ is given by,
+\[
+  \semg{\apptol {(\lambtol {a} {e})} {a'}} = \semg{e[a'/a]}
+\]
+\begin{proof}
+\begin{align*}
+  \semg{\apptol {(\lambtol {a} {e})} {a'}}
+  & = \psi^{{-1}}(\semg{\lambtol {a} {e}}) \circ \semg{a'} \otimes id \circ m_{a:A,\Delta} \\
+  & = \psi^{{-1}}(\semg{\lambtol {a} {e}}) \circ \lambda \otimes id \circ m_{a:A,\Delta} \\
+  & = \psi^{{-1}}(\psi(\semg{e} \circ m^{-1}_{a:A,\Delta} \circ \lambda^{-1} \otimes id)) \circ \lambda \otimes id \circ m_{\Delta',\Delta} \\
+  & = \semg{e} \circ m^{-1}_{a:A,\Delta} \circ \lambda^{-1} \otimes id \circ \lambda \otimes id \circ m_{a:A,\Delta} \\
+  & = \semg{e}
+\end{align*}
+
+\begin{align*}
+  \semg{e[a'/a]}
+  & = \semg{e} \circ \semg{a'/a} \tag{\cref{lem:subst}} \\
+  & = \semg{e} \tag{coherence}
+\end{align*}
+\end{proof}
+
+\paragraph{$\tol\eta$}
+The $\eta$ rule for $\tol$ is given by,
+\[
+  \semg{\lambtol {a} {\apptol e a}} = \semg{e}
+\]
+\begin{proof}
+\begin{align*}
+  \semg{\lambtol {a} {\apptol e a}}
+  & = \phi(\semg{\apptol e a} \circ m^{-1}_{a:A,\Delta} \circ \lambda^{-1} \otimes id) \\
+  & = \phi(\phi^{-1}(\semg{e}) \circ \semg{a} \otimes id \circ m_{a:A,\Delta} \circ m^{-1}_{a:A,\Delta} \circ \lambda^{-1} \otimes id) \\
+  & = \phi(\phi^{-1}(\semg{e}) \circ \lambda \otimes id \circ m_{a:A,\Delta} \circ m^{-1}_{a:A,\Delta} \circ \lambda^{-1} \otimes id) \\
+  & = \phi(\phi^{-1}(\semg{e})) \\
+  & = \semg{e}
+\end{align*}
+\end{proof}
+
+\subsubsection{$\bigamp$-Products}
+% Products
+\paragraph{$\bigamp$-Introduction}
+$\semg{\dlamb x e} : \semg{\Delta} \to \prod_{x : X} {\sem{A}(\gamma , x)}$ is
+defined by the universal property of the product
+\[
+\semg{\dlamb x e} = \left( \sem{e}(\gamma , x) \right)_{(x : X)}
+\]
+
+\paragraph{$\bigamp$-Elimination}
+$\semg{e.\pi~M} : \semg{\Delta} \to \sem{A}(\gamma , M)$ is defined using the
+projection out of the product,
+
+\begin{center}
+\begin{tikzcd}
+  \semg{\Delta} \arrow[r , "\semg{e}"] &
+  \prod_{x : X} {\sem{A}(\gamma , x)} \arrow[r , "\pi_M"] &
+  \sem{A}(\gamma , M)
+\end{tikzcd}
+\end{center}
+
+\paragraph{$\bigamp\beta$}
+The $\beta$ law for $\bigamp$ is given by,
+\[
+  \sem{(\dlamb x e).\pi~M}(\gamma, x) = \sem{e[M/x]}(\gamma, x)
+\]
+\begin{proof}
+\begin{align*}
+  \semg{(\dlamb x e).\pi~M}
+  & = \pi_{M} \circ \semg{\dlamb x e} \\
+  & = \pi_{M} \circ (\sem{e}(\gamma , x))_{(x : X)} \\
+  & = \sem{e}(\gamma , M)
+\end{align*}
+by the universal property of the product.
+
+\begin{align*}
+  \sem{e[M/x]}(\gamma , x)
+  & = \sem{e}(\gamma , x) \circ  \sem{M/x}(\gamma , x) \\
+  & = \sem{e}(\gamma , M)
+\end{align*}
+\end{proof}
+
+\paragraph{$\bigamp\eta$}
+The $\eta$ law for $\bigamp$ is given by,
+\[
+  \semg{(\dlamb x {e.\pi~x})} = \semg{e}
+\]
+\begin{proof}
+\begin{align*}
+  \semg{(\dlamb x {e.\pi~x})}
+  & = (\semg{e.\pi~x})_{x : X} \\
+  & = (\pi_x \circ \semg{e})_{x : X} \\
+  & = \semg{e}
+\end{align*}
+\end{proof}
+by the universal property of the product.
+
+\subsubsection{$\bigoplus$-Sums}
+% Sums
+\paragraph{$\bigoplus$-Introduction}
+$\semg{\sigma~M~e} : \semg{\Delta} \to \coprod_{x : X} \sem{A}(\gamma, x)$
+
+\begin{center}
+\begin{tikzcd}
+  \semg{\Delta} \arrow[r , "\semg{e}"] &
+  \sem{A}(\gamma, M) \arrow[r , "i_M"] &
+  \coprod_{x : X} {\sem{A}(\gamma , x)}
+\end{tikzcd}
+\end{center}
+
+\paragraph{$\bigoplus$-Elimination}
+$\semg{\letin {\sigma~x~a} {e} {e'}} : \semg{\Delta'_{1},\Delta,\Delta'_{2}} \to \semg{C}$
+is defined in the diagram
+
+\begin{center}
+% https://tikzcd.yichuanshen.de/#N4Igdg9gJgpgziAXAbVABwnAlgFyxMJZABgBpiBdUkANwEMAbAVxiRAB124YBbAc2CcAIjAY46AcgD6ARlLDR4+exFjJUgEwBfEFtLpMufIRRzK1es1aIOXXgIVrpcx+K0ACThDw94nu-yCKorq2rr6IBjYeAREcjJUtIwsbJwMMABmOAAU-tyBruoyHl4+fpz5DsFqJewATlh8ABY4AJT+3li+cHn2Qari0mF6BtHGcaQaiZYpNpwAMlhgAMqNACoAnos8uO7AAB4ewAAaRxW8-SHOpHTuiO4AgsoDoVrZnHx0PDy3pO77rR0I0ihhiJhIk2myWstkWK3WWy6uwOR1OezSmRyvR4lycsjeHy+P3cfwBHTKcHYYHcNNp5xxDwJ7E+31+-3anAazTa5K65QCuMGmiZLOJpMB4VGRliKDICQs0NS7HSWVy9KqL2ctU63X8cNWfE222Rhz2aKC+WAjPezKJbIBtS5LQ57B1-MqgtekpBYxlyDkAGYoVYlZaAMIiu3i71RaXgshBhUhuYBAQR3SJGBQPjwIigDJ1CA8JBkEA4CBIORJZMgHhSYA2zWyZ4hVotvFhagMOgAI1EAAVQeMbCqcN6C0XK9Ry0gNEnZrX64VroUtKU+T0sFAQF3ewOhzKQKPx4Xi4g52WK4gAKzzmHZLe83Xq4AwLQut2b7e7vsMQe+kwj0xE9J0QANpyvAA2O82GQbFXwkSNWRJdl3Aoetsn2O53GOCUf33AC2GPYEJzPAAWCCkHA6sFx4AA9YAAFpigw5dmzoRAHjbNjOyPPc-wPQDiIiUikFvS8kAomiYR3Pjf3-OMiOAkjTyQaCJMQAB2LQKC0IA
+\begin{tikzcd}[column sep =large]
+{\semg{\Delta'_1,\Delta,\Delta'_2}} \arrow[r, "{m_{(\Delta'_1,\Delta),\Delta'_2}}"]                                                                                                           & {\semg{\Delta'_1,\Delta} \otimes \semg{\Delta'_2}} \arrow[d, "{m_{\Delta'_1,\Delta}\otimes id}"]                             \\
+{\left( \semg{\Delta'_1} \otimes \coprod_{x:X} {\sem{A}(\gamma , x)} \right) \otimes \semg{\Delta'_2}} \arrow[d, "d"]                                                                 & \left( \semg{\Delta'_1} \otimes \semg{\Delta} \right) \otimes \semg{\Delta'_2} \arrow[l, "(id \otimes \semg{e}) \otimes id"] \\
+{\coprod_{x : X} {\left( \sem{\Delta'_1}(\gamma , x) \otimes
+      \sem{A}(\gamma , x) \right) \otimes \sem{\Delta'_2}(\gamma , x)}} \arrow[r, "\coprod_{x:X}({m^{-1}_{(\Delta'_1,a:A),\Delta'_2}})"] & {\coprod_{x : X} {\sem{\Delta'_1,a : A,\Delta'_2}(\gamma , x)}} \arrow[d, "{[ \sem{e'}(\gamma , x) ]_{(x : X)}}"]     \\
+\semg{C}                                                                                                                                                                                      & {\sem{C}(\gamma, x)} \arrow[l]
+\end{tikzcd}
+\end{center}
+
+where $d$ is the distributivity morphism, and the last morphism implicitly
+weakens $\sem{C}$.
+
+\paragraph{$\bigoplus\beta$}
+The $\beta$ rule for $\bigoplus$ is given by,
+\[
+  \semg{\letin {\sigma~x~a} {\sigma~M~a'} {e'}} = \semg{e'[M/x,a'/a]}
+\]
+\begin{proof}
+  \begin{align*}
+  \semg{\letin {\sigma~x~a} {\sigma~M~a'} {e'}}
+  & = {[ \sem{e'}(\gamma , x) ]_{(x : X)}} \circ \coprod_{x:X}(m^{-1}) \circ d \circ (id
+  \otimes \semg{\sigma~M~a'}) \otimes id \circ m \otimes id \circ m \\
+  & = {[ \sem{e'}(\gamma , x) ]_{(x : X)}} \circ \coprod_{x:X}m^{-1} \circ d \circ (id
+  \otimes (i_M \circ \semg{a'})) \otimes id \circ m \otimes id \circ m \\
+  & = {[ \sem{e'}(\gamma , x) ]_{(x : X)}} \circ \coprod_{x:X} m^{-1} \circ d \circ (id \otimes i_M) \otimes id \circ (id \otimes  \lambda) \otimes id \circ m \otimes id \circ m \\
+  & = {[ \sem{e'}(\gamma , x) ]_{(x : X)}} \circ \coprod_{x:X} m^{-1} \circ i_M \circ (id \otimes  \lambda) \otimes id \circ m \otimes id \circ m \\
+  & = \sem{e'}(\gamma , M) \circ m^{-1} (id \otimes  \lambda) \otimes id \circ m \otimes id \circ m \\
+  & = \sem{e'}(\gamma , M) \tag{coherence}
+  \end{align*}
+
+  \begin{align*}
+    \semg{e'[M/x,a'/a]}
+    &= \semg{e'} \circ \semg{M/x,a'/a} \tag{\cref{lem:subst}} \\
+    &= \sem{e'}(\gamma , x)
+  \end{align*}
+\end{proof}
+
+\paragraph{$\bigoplus\eta$}
+It suffices to show
+\[
+  \semg{\letin {\sigma~x~a} {c'} {f[\sigma~x~a/c]}} = \semg{f[c'/c]} = \semg{f}
+  \]
+First, expanding the left hand side, we have.
+\begin{proof}
+  \begin{align*}
+    \semg{\letin {\sigma~x~a} {c'} {f[\sigma~x~a/c]}}
+    &= [ \semg{f[\sigma~x~a/c]} ]_{(x : X)} \circ \coprod_{x:X}(m^{-1}) \circ d \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m\\
+    &= [ \semg{f} \circ (id \otimes i_x) \otimes id ]_{(x : X)} \circ \coprod_{x:X}(m^{-1}) \circ d \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m\\
+    &= \semg{f} \circ [ (id \otimes i_x) \otimes id \circ (m^{-1})]_{(x : X)} \circ d \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m\\
+  \end{align*}
+Since the domain has the universal property of a coproduct (due to distributivity), to prove
+this is equal to $\semg{f}$, it is sufficient to prove they are equal
+when composed with the injections:
+
+\begin{align*}
+  \semg{f} &\circ [ (id \otimes i_x) \otimes id \circ (m^{-1})]_{(x : X)} \circ d \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m \circ (id \otimes i_y) \otimes id\\
+  &= \semg{f} \circ [ (id \otimes i_x) \otimes id \circ (m^{-1})]_{(x : X)} \circ d  \circ (id \otimes i_y) \otimes id \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m \tag{naturality}\\
+  &= \semg{f} \circ [ (id \otimes i_x) \otimes id \circ (m^{-1})]_{(x : X)} \circ i_y \otimes id \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m \tag{naturality}\\
+  &= \semg{f} \circ (id \otimes i_y) \otimes id \circ (m^{-1}) \otimes id \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m \\
+  &= \semg{f} \circ (m^{-1}) \otimes id \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m \circ (id \otimes i_y) \otimes id \\
+  &= \semg{f} \circ (id \otimes i_y) \otimes id \tag{coherence}\\
+\end{align*}
+\end{proof}
+
+\subsubsection{Equalizer}
+%equalizer
+\paragraph{Equalizer Introduction}
+$\semg{\equalizerin{e}} : \semg{\Delta} \to \semg{\equalizer e f g}$ where
+$\semg{e} : \semg{\Delta} \to \semg{A}$ and
+$\semg{f} \circ \semg{e} = \semg{g} \circ \semg{e}$. By the universal property
+of the equalizer the preceding equality induces a unique morphism
+$\semg{\Delta} \to Eq(\semg{f} , \semg{g}) = \semg{\equalizer e f g}$. Define
+$\semg{\equalizerin{e}}$ to be this map.
+
+\paragraph{Equalizer Elimination}
+$\semg{\equalizerpi e} : \semg{\Delta} \to \semg{A}$ is defined using the map
+$\pi_{eq}$ from $Eq(\semg{f} , \semg{g})$ to the domain of $f$ and $g$.
+
+\[
+  \semg{\equalizerpi e} = \pi_{eq} \circ \semg{e}
+\]
+
+\paragraph{Equalizer $\beta$}
+The $\beta$ rule for $\equalizer e f g$ is given as,
+\[
+\semg{\equalizerpi {\equalizerin e}} = \semg{e}
+\]
+where $\semg{f} \circ \semg{e} = \semg{g} \circ \semg{e}$.
+\begin{proof}
+  In $\semcat$ the universal property of $Eq(\semg{f}, \semg{g})$ implies that
+  the following diagram commutes, implying the $\beta$ rule.
+
+% https://q.uiver.app/#q=WzAsNCxbMSwwLCJcXHNlbWd7QX0iXSxbMiwwLCJcXHNlbWd7Qn0iXSxbMCwwLCJFcShcXHNlbWd7Zn0sXFxzZW1ne2d9KSJdLFswLDEsIlxcc2VtZ3tcXERlbHRhfSJdLFswLDEsIlxcc2VtZ3tmfSIsMCx7Im9mZnNldCI6LTF9XSxbMCwxLCJcXHNlbWd7Z30iLDEseyJvZmZzZXQiOjF9XSxbMiwwLCJcXHBpX3tlcX0iXSxbMywwLCJcXHNlbWd7ZX0iLDJdLFszLDIsIlxcc2VtZ3tcXGVxdWFsaXplcmluIGV9Il1d
+\begin{center}
+\begin{tikzcd}
+	{Eq(\semg{f},\semg{g})} & {\semg{A}} & {\semg{B}} \\
+	{\semg{\Delta}}
+	\arrow["{\pi_{eq}}", from=1-1, to=1-2]
+	\arrow["{\semg{f}}", shift left, from=1-2, to=1-3]
+	\arrow["{\semg{g}}", shift right, swap, from=1-2, to=1-3]
+	\arrow["{\semg{\equalizerin e}}", from=2-1, to=1-1]
+	\arrow["{\semg{e}}"', from=2-1, to=1-2]
+\end{tikzcd}
+\end{center}
+\end{proof}
+
+\paragraph{Equalizer $\eta$}
+
+The $\eta$ rule for $\equalizer e f g$ is given as,
+\[
+\semg{\equalizerin {\equalizerpi e}} = \semg{e}
+\]
+\begin{proof}
+  Likewise, the universal property of $Eq(\semg{f}, \semg{g})$ implies $\eta$
+  rule via this diagram.
+
+\begin{center}
+\begin{tikzcd}
+	{Eq(\semg{f},\semg{g})} & {\semg{A}} & {\semg{B}} \\
+	{\semg{\Delta}}
+	\arrow["{\pi_{eq}}", from=1-1, to=1-2]
+	\arrow["{\semg{f}}", shift left, from=1-2, to=1-3]
+	\arrow["{\semg{g}}", shift right, swap, from=1-2, to=1-3]
+	\arrow["{\semg{e}}", from=2-1, to=1-1]
+	\arrow["{\semg{\equalizerpi e}}"', from=2-1, to=1-2]
+\end{tikzcd}
+\end{center}
+\end{proof}
+\fi
 \end{document}
